### 2018-09-06 cover-letter/cover-letter.Rmd
first commit

---
from: Nicholas Tierney
fromqualifications: BPsySc (Hons), PhD (Statistics)
fromtitle: Research Fellow
fromWWW: njtierney.com
fromemail: nicholas.tierney@gmail.com
fromphone: +61 3 9905 9352
fromsignature: sigfile.png
toname: Dr. Tyler McCormick
toaddress: JCGS
opening: Dear Dr. McCormick
closing: Regards
spacing: 1.4
output: MonashEBSTemplates::letter
---

Please consider my article "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations" for possible publication in JCGS. This article describes new work on handling missing values from a data centric approach. The methodology is implemented in the R package, `naniar`, which is available on CRAN, https://cran.r-project.org/web/packages/naniar, with the development version on github https://github.com/njtierney/naniar.

Along with the PDF is an accompanying Rmarkdown file, which contains all of the code to reproduce the results in this paper. This is the first time that this article has been submitted to a journal, and is not under consideration elsewhere. We have submitted a pre-print of this article to arxiv, and to our internal department pre-print at Monash University.

We hope that you find this paper of interest to the readers of JCGS,

### 2018-09-06 paper/tidy-missing-data-paper.Rmd
first commit

---
title: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
author:
- familyname: Tierney
  othernames: Nicholas
  address: Monash University
  email: nicholas.tierney@gmail.com
  correspondingauthor: true
- familyname: Cook
  othernames: Dianne
  address: Monash University
abstract: "Despite the large body of research on missing value distributions and imputation, there is comparatively little literature on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with a goal to integrating missing value handling as an integral part of data analysis workflows. New data structures are defined along with new functions (verbs) to perform common operations. Together these provide a cohesive framework for handling, exploring, and imputing missing values. These methods have been made available in the R package `naniar`."
keywords: "workflow, statistical computing, data science, data visualization, tidyverse, data pipeline"
wpnumber: 1/2018
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: tidy-missing-data-paper.bib
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.align = "center",
                      # dev = c("png"), 
                      # dpi = 300,
                      fig.width = 4, 
                      fig.height = 4, 
                      cache = TRUE)
library(tidyverse)
library(gridExtra)
library(kableExtra)
library(ggthemes)
library(png)
library(scales)
library(simputation)
library(naniar)
library(knitr)
library(lubridate)
library(magick)

```

# Introduction {#intro}

Tidy data [@Wickham2014] is a relatively new principle and suite of tools that
facilitate the process of converting raw data into a clean, analysis-ready
format that works efficiently in a data analysis pipeline. In tidy data, missing
values can be handled implicitly or explicitly, at the analyst's discretion.
However, when making a plot, missing values are simply dropped, albeit with a
warning (see Figure \@ref(fig:warning))). Most other analysis software will
simply drop cases with missings without warning.

```{r warning, fig.cap = "How ggplot behaves when displaying missing values:  Warning message (Left) is thrown prior to plot (Right); ggplot2 does not display missing values in the plot, but at least throws a warning message to the user, which is better than many statistical models.", fig.height = 4, fig.width = 4, out.width = "100%", warning = FALSE, fig.show = 'hold'}

p1 <- ggplot(oceanbuoys,
       aes(x = humidity,
           y = air_temp_c)) +
  geom_point(size = 0.5) +
  theme(aspect.ratio = 1)

ggsave("images/geom-point-warn.png", 
       width = 4, 
       height = 4, 
       units = c("in"))

img_gg_p <- image_read("images/geom-point-warn.png")

img_gg_warn <- image_read("images/ggplot-warn.jpeg")

image_append(c(img_gg_warn, img_gg_p)) %>% 
  image_write("images/ggplot2-warning-and-plot.png", "png")

knitr::include_graphics("images/ggplot2-warning-and-plot.png")

```

The imputation literature does not address how to handle missing data. Producing
a complete dataset for analysis, whether by case or variable deletion, or with
imputation, requires knowing something about the structure of the missing
values. To understand their structure, one needs to generate visualizations,
calculate summaries, and perform exploratory modelling. The imputation methods
literature (e.g. @Little1988, @Rubin1976, @Simon1986, @Schafer2002, 
@VanBuuren2012) focuses on ensuring valid statistical inference is made from
incomplete data. They approach this chiefly through probabilistic modelling, and
assume the mechanism of missing data is known to the analyst.

Exploring missing data to find the missingness mechanism is difficult, because it is an iterative process with many dead ends and often no clear answer. There is no silver bullet for missing data that will reveal it's mechanism to the analyst. Despite it being crucial to explore and understand missing data, there is very little research on how to make this process more efficient so that it integrates into analysis pipelines. Although decision tree models or latent group analysis can reveal structures or patterns of missingness [@Tierney2015; @Barnett2017], these are not definitive and usually require discussion with data curators to validate. The analyst must explore missing data using visualizations, summaries, and exploratory analyses, but there is a lack of tools to efficiently explore missing data.

The graphics literature [@Swayne1998; @Unwin1996] provides several solutions for exploring missings visually. These approaches incorporate missings into the plot in some way. For example, imputing values 10% below the range to display all values in a scatter plot. These methods require the data to be linked to the missing or imputed values; this link is typically hidden from the user. The approaches from the graphics literature have not been translated and integrated into tidy data. This limits the analysts capacity to efficiently explore missing values. Extending tidy data and tidy tools to account for missing values would create more efficient and robust workflows and analysis.

The paper is organized in the following way to discuss the problem of missing data. The next section (\@ref(tidy-data-concepts)) is an introduction to tidy data principles and tools. Missing value imputation (\@ref(missing-data-rep-dep) and \@ref(imputation)) and exploration literature (\@ref(exploration)) are then discussed. Section \@ref(extensions) discusses extensions to the tidy methodology that facilitates operations for exploring, visualizing, and imputing missing data. The application of these new methods is illustrated using a case study in section \@ref(case-study). Finally, section \@ref(discussion) discusses strengths, limitations, and future directions.

# Background {#background}

## Tidy data concepts and methods {#tidy-data-concepts}

Features of tidy data were formally discussed in 2014 [@Wickham2014], subsequently generating much discussion (see @Donoho2017 and commentary), and tools for data analysis. Tidy data is defined as:

> 1. Each variable forms a column.
> 2. Each observation forms a row.
> 3. Each type of observational unit forms a table.

Tidy data is easier to work with and analyze because the variables are in the same format as they would be put into modelling software. This helps the analyst works swiftly and clearly, closing up opportunities for errors.

With tidy data come tidy tools, which have the same input and output: tidy data. This consistency means multiple tools can be composed together in sequence, allowing for rapid, elegant operations. Contrasting tidy tools are messy tools. These have tidy input but messy output. Messy tools slow down analysis by shifting the focus from analysis to transforming output so it is the right shape for the next step in the analysis. This makes the work at each step harder to predict, and more complex and difficult to maintain. This disrupts workflow, and incites errors.

Tidy tools fall into three broad categories: data manipulation, visualization, and modelling. These are now discussed in turn.

### Data manipulation {#tidy-data-manip}

Data manipulation is made input and output tidy with R packages `dplyr` and `tidyr` [@dplyr; @tidyr]. These provide the five "verbs" of data manipulation: data reshaping, sorting, filtering, transforming, and aggregating. **Data reshaping** goes from long to wide formats; **sorting** arranges rows in a specific order; **filtering** removes rows based on a condition; **transforming** changes existing variables or adding new ones; **aggregating** creates a single value from many values, say for example by computing the minimum, maximum, and mean.

### Visualizations {#tidy-vis}

Visualization tools only have tidy data as their input, as the output is a graphic. The popular domain specific language `ggplot2` maps variables in a dataset to features (referred to as aesthetics) of a graphic [@ggplot2]. For example, a scatterplot can be created by mapping two variables to the x and y axes, and specifying a point geometry. The graph can then be changed to map a third variable to color of the points, and a fourth variable to size.

### Modelling {#tidy-model}

Modelling tools work well with tidy data, as they have a clear mapping from variables in the data to the formula for a model. For example in R, y regressed on x and z is: `lm(y ~ x + z)`. Modelling tools are input tidy, but their output is always messy. For example, estimated coefficients, predictions, and residuals from one model cannot be easily combined with the output of another model. The `recipes` package is a tidy tool under development to help make modelling input and output tidy [@recipes].

### The tidyverse {#the-tidyverse}

Defining tidy data and tidy tools has resulted in the growth of a set of packages known collectively as the "tidyverse" [@Ross2017]. These are designed to share similar principles in their design and behavior, and cover the breadth of an analysis - going from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @Ross2017; @Tidyverse-Manifesto]. This has led to a growth of tools for specific parts of analysis - from reading in data with `readr`, `readxl`, and `haven`, to handling character strings with `stringr`, dates with `lubridate`, and performing functional programming with `purrr` [@readr; @readxl; @haven; @stringr; @lubridate; @purrr]. It has also led to the growth of new packages for other fields that follow similar design principles and create fluid workflows for new domains. For example, the `tidytext` package for text analysis, the `tsibble` package for time series data, and `tidycensus` for working with US census and boundary data [@tidytext; @tsibble; @tidycensus].

### Tidy formats for missing data {#tidy-formats-missing-data}

Current tools for missing data are messy. Missing data tools can be used to perform imputations, missing data diagnostics, and data visualizations. However, these tools are primarily _messy_, and suffer the same problems as modelling for imputation: They use tidy input, but produce messy output. The complex, often multivariate nature of imputation methods also makes makes them difficult to represent. Visualization methods for missing data do not map data features to the aesthetics of a graphic, as in `ggplot2`, limiting expressive exploration.

Past graphics research on missing data has not yet been framed in a tidy way.
The graphics literature has explored methods for missing value exploration and imputation. Translating and expanding these to fit within tidy data and tidy tools would create more efficient workflows.

## Missing data representation and dependence {#missing-data-rep-dep}

The information on missing values in a dataset can be represented as a missingness matrix, $R$, where 0 is observed data and 1 is missing data, where for data $y$ with $i$ rows and $j$ columns, the value $r_{ij}$ is given by:

$$
r_{ij} =\begin{cases}
1, & \text{if } y_{ij} \text{ is missing} \\
0, & \text{if } y_{ij} \text{ is observed}
\end{cases}
$$

There are many ways each value can go missing. The information in $R$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), Missing at random, and missing not at random. **MCAR** is where values being missing have no association with observed or unobserved data, that is $Pr(missing | observed, unobserved) = Pr(missing)$. Essentially, the probability of an observation being missing is unrelated to anything else.  Although this is a convenient scenario, it is not actually possible to confirm, as it relies on statements on data unobserved. **MAR** refers to cases where the missingness only depends on the data observed, and not data unobserved, that is,$Pr(missing | observed, unobserved) = Pr(missing | observed)$. Some structure or dependence between the missing values and observed values is allowed, provided that this can be explained by the data observed. Finally, **MNAR** refers to the missingness being related to values unobserved: $Pr(missing | observed, unobserved)$. This assumes that conditioning on all available observed data, the data goes missing due to some phenomena unobserved. This presents a challenge in analysis, as it is difficult to verify this state, and also implies bias in analysis due to the unobserved phenomena.

Visualizations can help assess whether data is MCAR, MAR or MNAR. Imputation is recommended in most cases of missingness dependence, where the values are plausible (e.g., imputing a pulse for someone deceased), as this can help re-assess the classification of missingness.

## Imputation {#imputation}

Imputation methods are more accessible than they have ever been, and the number of methods and implementations in software continues to grow. Values can be imputed with one value (single imputation), or multiple values (multiple imputation), creating $m$ datasets. Methods and software for single and multiple imputation are now discussed.

### Single imputation {#single-imputation}

`VIM` [@VIM] efficiently implements well-used imputation methods k-nearest neighbor, regression, hot-deck, and iterative robust model-based imputation. This diversity of approaches allow the package to deal semi-continuous, continuous, count, and categorical data. VIM identifies imputed cases by adding an indicator variable with a suffix `_imp`. So `Var1` would have a sibling indicator column, `Var1_imp`, with TRUE or FALSE values to indicate imputed value. `VIM` also has a variety of visualization methods, discussed in section \@ref(exploration). The `simputation` package includes imputation methods for linear regression, decision trees, K nearest neighbors, hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977]. A key feature of its design is that it returns a regular dataframe. This approach contrasts from other imputation software, which may provide custom classes of data or additional objects, which require additional work to prepare for subsequent analysis. Providing a dataframe with imputed values included in the data reduces the friction of working with other tools, but comes at the cost identifying imputed values. Other software for imputation includes `Hmisc` [@Hmisc], which provide predictive mean matching, `imputeTS` [@imputeTS], which provide time series imputation methods, and `missMDA` [@missMDA], which imputes data using principal components analysis.

### Multiple imputation {#multiple-imputation}

Multiple imputation is often regarded as best practice for imputing values [@Schafer2002], as long as appropriate caution is taken [@Sterne2009]. Popular and robust methods for multiple imputation include the `mice`, `Amelia`, and `mi` packages [@mice; @amelia; @mi]. The `mice` package implements the method of chained equations, using a variable-wise algorithm to calculate the posterior distribution of parameters, from which uses to generate imputed values. The suggested workflow in `mice` revolves around imputing data, returning completed data, and fitting a model and pooling the results. The `Amelia` package [@amelia] assumes data are multivariate normal, and samples from the posterior using the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm [@Honaker2010] and allows for incorporation of information on the values in a prior. The `mi` package [@mi] also uses Bayesian models for imputation, providing better handling of semi-continuous values, and data with structural or perfect correlation. A collection of analysis models are also provided in `mi`, which work with the specially multiply imputed data, including linear models, generalized linear models, and their corresponding Bayesian components. This approach promotes fluid workflow, with a similar penalty to tidying up model output, which is still messy.  Another software for performing multiple imputation is `norm` [@norm; @schafer-norm], which uses methods from the NORM software from @schafer-norm, such as multiple imputations using EM for multivariate normal data. The `norm` package does not provide a framework for tracking missing values, but instead provides tools for making inference from multiple imputation. Each of these multiple imputation methods provide practical methods, but do not have consistent interfaces across their methods or implementation, which makes then inherently messy, as their output is always different. A consistent data structure that tracks missing values by default would make the data outputs tidy, streamlining subsequent analysis.

## Exploration {#exploration}

The primary focus of most missing data packages is exploring imputed values, not
on exploring relationships in missing values, and identifying possible patterns.
Texts that do cover exploration of missing data have the same problem as with modelling: the input is tidy, but the output is messy and does not work with other tools [@VanBuuren2012]; this is inefficient. Methods for exploring missing values are primarily covered in literature on interactive graphics [@Swayne1998; @Unwin1996; @Cook2007], and are picked up again in a discussion of a graphical user interface [@Cheng2015].

Missing values in a dataset can be represented as a binary matrix where 0 is present and 1 is missing [@Rubin1976; @Schafer2002]. This is often referred to as a matrix $R$, which can be used to assess missing data dependence. This idea of a matrix of missing values has also been used in interactive graphics, dubbed a "shadow matrix", to link missing and imputed values to the data, facilitating their display in graphics [@Swayne1998]. This work also focuses heavily on multivariate numeric data. This is an idea upon which this new work builds.

The approach of @Unwin1996 focussed on multivariate categorical data, where missingness is explicitly added as an additional category. MANET also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot. The approach of @Swayne1990 in the software XGobi, further developed in ggobi [@Cook2007], focussed more on multivariate quantitative data. Missingness is incorporated into plots in ggobi by setting them to be 10% below the minimum value.

MissingDataGUI provides a user interface for exploring missing data structure both numerically and visually. Using a GUI to explore missing data may make it easier to glean valuable insight into important structures for missingness. However, these insights are typically not captured or recorded, so it is difficult to incorporate them into reproducible analyses. This distracts and breaks analysis workflow, inviting mistakes.

VIM (visualizing and Imputing Missing Data) provides visualization methods that identify observed, imputed, and missing values. These include spinograms, spinoplots, missingness matrices, plotting missingness in the margins of other plots, and other summaries. These help explore missingness, however these visualizations do not map variables to graphical aesthetics, creating friction moving across workflows, making them difficult to extend to new circumstances. Additionally, data used to create the visualizations cannot be accessed, posing a barrier to further explorations.

`ggplot2` incorporates missingness into visualizations only when mapping a discrete variable to a graph aesthetic. This has some limitations. For example, for data with school grades and test scores, a boxplot visualization could have school grade mapped to the x axis, and test score to the y axis (See \@ref(fig:gg-box-na)). If there are missings in a continuous variable like test score, `ggplot2` omits the missing values and prints a warning message. However, if a discrete variable like school year has missing values, an NA category is created for school year, and the scores are placed into this.

Figure \@ref(fig:gg-box-na), shows four scenarios of missingness for test scores across school year - the first contains no missingness, the second contains missingness for school year, the third contains missingness in scores, and the fourth contains missingness in both.

```{r make-school-scores-data, echo = FALSE}

school_scores <- data.frame(score = c(rnorm(20, 70, 10),
                                 rnorm(20, 85, 7.5),
                                 rnorm(20, 95, 5),
                                 rnorm(20, 115, 10)),
                       year = rep(c("1st", "2nd", "3rd", "4th"), each = 20)) %>%
  mutate(score = as.numeric(score),
         year = as.factor(year)) %>%
  as_tibble()

school_scores_year_miss <- school_scores %>%
  mutate_at(.funs = function(x){
    x[sample(c(TRUE, NA), 
             prob = c(0.85, 0.150), 
             size = length(x), 
             replace = TRUE)]
  },
  .vars = "year")

school_scores_score_miss <- school_scores %>%
  mutate_at(.funs = function(x){
    x[sample(c(TRUE, NA), 
             prob = c(0.85, 0.150), 
             size = length(x), 
             replace = TRUE)]
  },
  .vars = "score")

school_scores_both_miss <- school_scores %>%
  mutate_all(.funs = function(x){
    x[sample(c(TRUE, NA), 
             prob = c(0.85, 0.150), 
             size = length(x), 
             replace = TRUE)]
  })

school_scores_all <- bind_rows(
  complete = school_scores,
  year_NA = school_scores_year_miss,
  score_NA = school_scores_score_miss,
  both_NA = school_scores_both_miss,
  .id = "NA_type"
)

```

```{r gg-box-na, fig.height = 6, fig.width = 6, out.width = "70%", fig.align = "center", fig.cap = "ggplot2 provides different visualizations depending on what type of data has missing values. (A) Data is complete and graphic is presented; (B) Missings are only in category variable year - an additional 'NA' boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and school year, additional missing information is shown. The missingness category is only shown when there are missings in categorical variables such as year (plots (B) and (D)). In (C), no missingness information is given on the graphic, despite there being missings in score, and a warning message is displayed about the number of missing values omitted.", fig.show = "asis"}

gg_boxplot <-  function(data){
  ggplot(data,
       aes_string(x = "year",
                  y = "score")) +
  geom_boxplot()
}

p1 <- gg_boxplot(school_scores_all %>% filter(NA_type == "complete"))
p2 <- gg_boxplot(school_scores_all %>% filter(NA_type == "year_NA"))
p3 <- gg_boxplot(school_scores_all %>% filter(NA_type == "score_NA"))
p4 <- gg_boxplot(school_scores_all %>% filter(NA_type == "both_NA"))

cowplot::plot_grid(p1,
          p2,
          p3,
          p4,
          nrow = 2,
          labels = LETTERS[1:4])
```

# Extensions to tidyverse {#extensions}

Applying tidyverse principles to the domain of missing data clarifies and
facilitates missing data exploration, visualization, and imputation. This section
discusses how these principles are applied, and proceeds in three parts. Section
\@ref(data-structure) discusses _nabular_ data, a tidy data structure for
missing values that tracks missings throughout the analysis and allows for
special missing values. Section \@ref(vis-summaries) covers visualizations to
understand missingness. Section \@ref(num-sum) discusses data summaries, and
Section \@ref(verbs) covers the common "verbs" used to solve problems that arise
when working with missing data. The design and naming of these features is an
important component that helps the analyst compose them efficiently in data pipelines, and
is discussed throughout.

## Data structure {#data-structure}

To explore missing data there needs to be a good representation of missing
values that integrate into analysis pipelines. A common representation of
missing values is a binary matrix the same dimension as the data, where 0 and 1
indicate not missing and missing. This is typically referred to as a matrix $R$,
used to describe missingness dependence patterns, (see for example @Rubin1976
and @Schafer2002). This matrix is used to explore missing values in the
interactive graphics library XGobi, where it was called a "shadow matrix",
defined as a copy of the original data with indicator values of missingness:
"NA" and "!NA" for missing and not missing, represented as 0 and 1 (see Figure
\@ref(fig:shadow-matrix-progress) parts 2 and 3 below).

```{r shadow-matrix-progress, echo = FALSE, fig.cap = "Progression of creating shadow matrix data. (1-2) Data to a binary representation of missings, (2-3) Binary format converted to a shadow matrix (3-4) New and improved shadow matrix with changed variable names. This clearly links a variable to its state of missingness.", out.width = "100%", fig.align="center"}

knitr::include_graphics("images/full-conversion-to-shadow.png")

```

Three features are added to the shadow matrix to facilitate analysis:

1. Coordinated names: Variables in the shadow matrix gain the same name as in the data, with the suffix "_NA".

2. Special missing values: Values in the shadow matrix can be "special" missing values, indicated as `NA_suffix`, where "suffix" is a very short message of the type of missings.

3. Cohesiveness: Binding the shadow matrix column-wise to the original data creates a cohesive _nabular_ data form, useful for visualization and summaries.

These additions are now discussed in turn.

### Coordinated names {#coordinated-names}

A consistent short suffix "`_NA`" for each variable keeps names coordinated throughout analysis (See Figure \@ref(fig:shadow-matrix-progress) part 4). The suffix is short and easy to remember in analysis or visualization. It also makes a clear distinction that `var_NA` is a random variable of the missingness of a variable, `var`. This subtle change is important as it changes the focus from the value of a variable to it's missingness state, making intent clear when performing analysis.

### Special missings {#special-missings}

Values in the shadow matrix are "factor" type values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing. Extending these to include "special" missing values labelled as "NA_\<suffix\>" allows for representing additional features such as instrument failure and drop-out as "NA_instr_fail", and "NA_drop_out". The underlying number representation changes from binary to integer. For example, instrument failure and drop out would be represented as 2 and 3, see Table \@ref(tab:shadow-encoding) for an example. Encoding these special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in `naniar`, shown in Section \@ref(verbs-recode).

```{r shadow-encoding}

fig_shadow_code <- tibble::tibble(temp = c(-99, NA, -1, 106),
                                  temp_NA = c("NA_instr",NA, "NA_dropout","!NA"),
                                  NA_value = c(2, 1, 3, 0))

knitr::kable(fig_shadow_code,
             caption = "Example data of temperature, and its observed value, shadow representation, and underlying value. The temperature value of -99 is represented in the shadow column as 'NA\\_instr', which in turn has the underlying numeric value of 2. This captures additional information about the data that is otherwise difficult to record.",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position"))

```

### Nabular data {#nabular-data}

_Nabular_ data binds the shadow matrix column-wise to the original data, and is so named as a portmanteau of `NA` and `tabular`. _Nabular_ data keeps corresponding rows together, removing the possibility of mismatching records, and explicitly links missing values to the data. _Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable, `var`, as `var_NA`. _Nabular_ data is a snapshot of the missingness of the data. This means when _nabular_ data is imputed, those imputed values can easily be identified in analysis by referring to the appropriate, coordinated names. _Nabular_ data is created using the command `bind_shadow()` or `nabular()` (see Figure \@ref(fig:nabularfig)). _Nabular_ data is not unlike classical data formats that have quality or flag columns associated with each measured variable, e.g. Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].

```{r nabularfig, echo = FALSE, fig.cap = "The process of creating nabular data. Data transformed to shadow matrix, and nabular data contains the shadow matrix, column bound to the data. Nabular data can be created using `bind\\_shadow` or `nabular()` functions. Nabular data provides a useful format for missing data exploration and analysis.", out.width = "100%", fig.align = "center"}

knitr::include_graphics("images/nabular.png")

```

A data structure for missing data provides key benefits over on-the-fly creating a logical matrix where TRUE means missing, and FALSE means present. Firstly, it can be confusing to users whether TRUE or FALSE means missing or present; `NA` and `!NA` are clearer. Secondly, special missing values cannot be added fluently during analysis with a logical matrix. Finally, the logical matrix cannot capture which values are imputed, if imputation has already taken place. Imputing values on _nabular_ data automatically tracks these value changes.

Using additional columns to represent missingness information follows recently published best practices for data organization, described in @Ellis2017 and @Broman2017: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column. Here they suggest to indicate censored data with an extra variable called "VariableNameCensored", which would be TRUE if censored, otherwise FALSE. This information is now represented in the shadow columns as additional integers, along with the other missing information.

Other statistical programs represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` through to `.z`. The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state.

## Visual summaries {#vis-summaries}

Visualizing missingness in a dataset is essential to understand its structure. This section discusses visualizations that are easy to remember and compose with other functions in a pipeline. Four different areas of visualization of missing data are discussed: overviews \@ref(overview-plots); univariate \@ref(univariate-plots), bivariate \@ref(bivariate-plots), and multivariate \@ref(multivariate-plots). All plots are created using `ggplot2`, giving users clearer control over the plot appearance, and customizability.

### Overview plots {#overview-plots}

The number of missings in each variable and case are visualized using `gg_miss_var` and `gg_miss_case` (see Figure \@ref(fig:gg-miss-case-var) below). These are shown using the "airquality" dataset, included in base R, which contains daily air quality measurements in New York, from May to September, 1973. Rather than highlighting the amount of complete data, these visualizations draw attention to the amount of missings, ordering by missingness.

```{r gg-miss-case-var, echo = FALSE, fig.show='hold', fig.cap = "Graphical summaries of missingness in variables and cases for the airquality data. (A) Missings in each variable and (B) in each case. There are missing values in Ozone and Solar.R, with Ozone having more missings, and not many cases have two missing values, with most missingness being from cases with one missing value.", fig.height = 3, fig.width = 6, out.width = "90%", fig.align="center"}

p1 <- gg_miss_var(airquality)

p2 <- gg_miss_case(airquality)

cowplot::plot_grid(p1,
                   p2, 
                   ncol = 2,
                   labels = "AUTO")
```

All missings in a dataset can be displayed using a heatmap style visualization. This is achieved using `vis_miss()` from the `visdat` package [@visdat], which also provides summaries of missingness overall in the legend, and for each column (see Figure \@ref(fig:vis-miss)). The user can also apply clustering to the rows, and arrange columns by missingness (Figure \@ref(fig:vis-miss)B). Similar visualizations are available in other missing data packages such as `VIM`, `mi`, `Amelia`, and `MissingDataGUI`. A key improvement is that `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns in the visualization and are named at the top, and each row is an observation. Using `ggplot2` means the plot can be customized and combined with other ggplot graphics.

```{r vis-miss, fig.height = 4, fig.width = 6, out.width = "95%", fig.show = 'hold', echo = FALSE, fig.cap = "Heatmap visualizations of missing data for the airquality dataset. (A) The default output and (B) ordered by clustering on rows and columns. There are only missings in ozone and solar radiation, and there appears to be some structure to their missingness."}

library(visdat)

p1 <- vis_miss(airquality)

p2 <- vis_miss(airquality, cluster = TRUE)

cowplot::plot_grid(p1,
          p2,
          nrow = 1,
          labels = "AUTO")

```

The number of times certain variables go missing together can be visualized using an "upset plot" [@Conway2017], a visualization technique for showing the size and features of sets in data that is similar to a venn diagram, but scales better with more variables (Figure \@ref(fig:airquality-upset)).

```{r airquality-upset, fig.cap = "The pattern of missingness in the airquality dataset shown in an upset plot. Only Ozone and Solar.R have missing values, and Ozone has the most missing values. There are 2 cases where both Solar.R and Ozone have missing values.", fig.height = 3, fig.width = 4, out.width = "80%"}

gg_miss_upset(airquality)

```

Figure \@ref(fig:airquality-upset) shows that only Ozone and Solar.R have missing values. The dots in the bottom right show the combinations of missingness, with the bottom left showing the missingness in each variable, and the top showing the missingness for the combinations. The bottom left shows that Ozone has the most missing values, and the top shows 2 cases where both Solar.R and Ozone have missing values together.

### Univariate plots {#univariate-plots}

Missing values are by default not shown for univariate visualizations such as histograms or densities. Two ways to use _nabular_ data to present univariate data with missings are discussed. The first imputes values below the range to facilitate visualizations. The second displays two plots of the same variable according to the missingness of a chosen variable.

#### Imputing values below the range {#imp-below-range}

To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed values below the range of data using `impute_below` (by default imputing 10% below the range). Visualizing this as a histogram shows missing values on its left. Ozone is shown in a different color by referring to the variable `Ozone` as `Ozone_NA` in a fill aesthetic in ggplot, shown in Figure \@ref(fig:impute-shift-histogram).

```{r impute-shift-histogram, fig.height = 3, fig.width = 5, fig.cap = "A histogram using nabular data to show the values and missings in ozone. Values are imputed below the range to show the number of missings in Ozone and colored according to missingness of ozone (`Ozone\\_NA`). There are about 35 missings in Ozone.", error = FALSE, warning = FALSE, message = FALSE, out.width = "75%"}

airquality %>%
  bind_shadow() %>%
  impute_below_all() %>%
  ggplot(aes(x = Ozone,
             fill = Ozone_NA)) + 
  geom_histogram() + 
  scale_fill_brewer(palette = "Dark2") + 
  theme_minimal()
```

#### Univariate split by missingness {#vis-split-by-miss}

Missingness of one variable can also be used to display different distributions in another. Figure \@ref(fig:bind-shadow-density) shows the values of temperature when ozone is present, and missing. Figure \@ref(fig:bind-shadow-density)A is a faceted histogram, and Figure \@ref(fig:bind-shadow-density)B is an overlaid density. This shows how values of temperature are affected by the missingness of ozone, and reveals a cluster of low temperature observations with missing ozone values.

```{r bind-shadow-density, fig.height = 3, fig.width = 5, out.width = "49%", error = FALSE, message = FALSE, fig.cap = "A visualization of temperature according to missingness in ozone from in the airquality dataset. A histogram of temperature facetted by the missingness of ozone (A), or a density of temperature colored by missingness in ozone (B). The distribution shows a cluster of low temperature observations with missing ozone values, but the temperature values are otherwise similar.", fig.show = 'hold'}

airquality %>%
  bind_shadow() %>%
ggplot(aes(x = Temp)) + 
  geom_histogram(na.rm = TRUE) + 
  facet_wrap(~Ozone_NA) + 
  labs(tag = "A")

airquality %>%
  bind_shadow() %>%
ggplot(aes(x = Temp,
           colour = Ozone_NA)) + 
  geom_density(na.rm = TRUE) +
  scale_colour_brewer(palette = "Dark2")  + 
  labs(tag = "B")

```

### Bivariate plots {#bivariate-plots}

To visualize missing values in two dimensions the missing values can be placed in the plot margins. This is achieved by imputing values below the range of the data. As in Section \@ref(imp-below-range), using _nabular_ data identifies imputed values, and color makes missingness pre-attentive [@treisman1985]. The steps of imputing and coloring have been been combined into `geom_miss_point()` from the `naniar` package (Figure \@ref(fig:geom-miss)), where there is a mostly uniform spread of missing values for Solar.R and Ozone.

```{r geom-miss, fig.height = 3, fig.width = 6, fig.cap = "Scatterplots with missings displayed at 10 percent below for the airquality dataset. Scatterplots of ozone and solar radiation (A), and ozone and temperature (B). There are missings in ozone and solar radiation, but not temperature."}

p1 <-
ggplot(data = airquality,
       aes(x = Ozone,
           y = Solar.R)) + 
  geom_miss_point() + 
  scale_colour_brewer(palette = "Dark2") + 
  theme(legend.position = "bottom") + 
  labs(tag = "A")

p2 <- 
ggplot(data = airquality,
       aes(x = Temp,
           y = Ozone)) + 
  geom_miss_point() + 
  scale_colour_brewer(palette = "Dark2") +
  theme(legend.position = "bottom")  + 
  labs(tag = "B")

gridExtra::grid.arrange(p1, p2, ncol = 2)

```

As `geom_miss_point` is a defined geometry for `ggplot2`, it provides features such as faceting and mapping other variables to aesthetics such as color, shape, or size. The shadow format by itself can also be used to visually explore some patterns of missingness.

```{r geom-miss-point-impute-shift-long, include = FALSE, out.width = "70%"}
airquality %>%
  bind_shadow() %>%
  impute_below_all() %>%
  add_label_shadow() %>%
  ggplot(aes(x = Ozone,
             y = Solar.R,
             colour = any_missing)) + 
  geom_point() + 
  scale_colour_brewer(palette = "Dark2") + 
  theme(legend.position = "bottom")
```

### Multivariate plots {#multivariate-plots}

Parallel coordinate plots can visualize missingness beyond two dimensions. They transform variables to the same scale, and typically make the data values range between 0 and 1. To showcase this visualization, the dataset `oceanbuoys` from `naniar` is used, which contains measurements of moored ocean buoys to understand and predict El Niño and El Niña. The data were collected in 1993 and 1997, and contains information on the sea and air temperature, humidity, and east west and north south wind directions.

Figure \@ref(fig:parallel-cord-plot) shows a parallel coordinate plot of the `oceanbuoys` data, with missing values imputed to be 10% below the range, and values colored according to whether humidity contained missing values. Although a 10% below imputation is not the most ideal way to display missing values in a parallel coordinate plot, it shows that humidity is missing at low air and sea temperatures, and that humidity is missing in one year, and one location.

```{r parallel-cord-plot, fig.width = 8, fig.height = 4, out.width = "100%", echo = FALSE, fig.cap = "Parallel coordinate plot shows missing values imputed 10\\% below range for the oceanbuoys dataset. Values are colored by missingness of humidity. Humidity is missing for low air and sea temperatures, and is missing for one year and one location. "}
library(naniar)
library(tidyverse)

range_01 <- function(x, ...){(x - min(x, ...)) / (max(x, ...) - min(x, ...))}

dat_paral <-  oceanbuoys %>%
  mutate(ID = 1:n(),
         ID = as.factor(ID)) %>%
  bind_shadow() %>%
  impute_below_all() %>%
  add_label_shadow() %>%
  mutate_if(is.numeric, range_01) %>%
  gather(key, value, -c(9:19)) %>%
  select(ID, key, value, everything())

ggplot(dat_paral,
       aes(x = key,
           y = value,
           group = ID,
           colour = humidity_NA)) + 
  geom_line(alpha = 0.3) + 
  theme(legend.position = "bottom") + 
  scale_colour_brewer(palette = "Dark2")

```

## Numerical summaries {#num-sum}

There are different ways to summarize the amount of missing and complete values in a dataset. This section describes approaches to summarizing missingness, implemented in the `naniar` package. The functions are easy to remember as they follow consistent naming, and their output is consistent, and returns a single number or dataframe. This means they integrate well with plotting and modelling tools. Section \@ref(single-num-sum) discusses single number summaries for all data and for cases and variables ; Section \@ref(sum-tab-missings) discusses variable- and case-wise summaries; and section \@ref(num-sum-w-group) shows how the design of these tools works with other tools in an analysis pipeline.

### Single number summaries {#single-num-sum}

The overall number of missing values in a dataset is shown with `n_miss`, and the proportion or percent missings, `prop_miss` and `pct_miss`. Complementing these are functions to reveal the number of complete values: `n_complete`, `prop_complete`, and `pct_complete`. These be extended to summarize the number/amount of variables and cases that contain a missing values by appending `_case` or `_var` to any of these summaries. Table \@ref(tab:n-prop-pct-miss-complete) shows the names and output of these functions on the dataset airquality.

```{r n-prop-pct-miss-complete, echo = FALSE}

table_of_fun <- tibble::tibble(
  "Missing Function" = c("n_miss", 
                         "prop_miss", 
                         "pct_miss",
                         "pct_miss_var",
                         "pct_miss_case"),
  "missing value" = c(n_miss(airquality),
                      prop_miss(airquality),
                      pct_miss(airquality),
                      pct_miss_case(airquality),
                      pct_miss_var(airquality)),
  "Complete function" = c("n_complete",
                          "prop_complete",
                          "pct_complete",
                          "prop_complete_var",
                          "pct_complete_case"),
  "complete value" = c(n_complete(airquality), 
                       prop_complete(airquality), 
                       pct_complete(airquality),
                       prop_complete_case(airquality), 
                       pct_complete_var(airquality))
) %>%
  mutate_if(is.numeric,round,2)

knitr::kable(
  table_of_fun, 
  digits = 2,
  caption = "Single number summaries of missing and complete data, applied to the airquality dataset. These functions follow consistent naming, to make them easy to remember, and to make their use clear.",
  booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position"))

```

### Summaries and tabulations of missing data {#sum-tab-missings}

Presenting the number and percent of missing values for each variable, or case, provides a summary that can be used by models to inform imputations or data handling. For example, potentially dropping variables or deciding to include others in an imputation model. Another useful approach is to tabulate the frequency of missing values for each variable or case; that is, the number of times there are zero, one, two, and so on, missing values. These summaries and tabulations are shown for variable in Tables \@ref(tab:miss-var-summary) and \@ref(tab:miss-var-table), using the `miss_var_summary` and `miss_var_table` functions. Functions are also provided for case-wise summaries and tabulations with `miss_case_summary` and `miss_case_table`, which order by `n_miss`, so the most missings are always shown at the top. Other summaries provided in `naniar` include summaries of missingness across a provided repeating span, `miss_var_span`, and finding the streaks or runs of missingness in a given variable in `miss_var_run`.

```{r miss-var-summary}

miss_var_summary(airquality) %>% 
  knitr::kable(
    caption = "Summary of the number and percent of missings in each variable. Only ozone and solar radiation have missing values.",
    digits = 1,
    booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position"))

```

```{r miss-var-table}

miss_var_table(airquality) %>% 
  knitr::kable(
    caption = "Tabulation of the amount of missing data in each variable. This shows the number of variables with no missings, 7 missings, and 37 missings, and the percentage of variables with those amounts of missingness. There are not many patterns of missingness.",
    digits = 1,
    booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position"))
```

### Combining numerical summaries with grouping operations {#num-sum-w-group}

These summaries and tabulations can be composed with the "pipe" operator to perform grouped summaries using the `group_by` operator from `dplyr`.
These summaries return dataframes, which may be used by other plotting devices, modelling tools, or other summaries. Table \@ref(tab:group-miss-var-summary) shows an example of the missing data summaries for the airquality dataset in each variable, for each month of the year. These datasets can be summarized in visualizations, or values input into a model.

```{r group-miss-var-summary}

airquality %>%
  group_by(Month) %>%
  miss_var_summary() %>%
  ungroup() %>%
  slice(1:10) %>%
  knitr::kable(caption = "A summary of the missingness in each variable, for each Month of the airquality dataset, with only the first 10 rows of output shown. There are more ozone missings in June than May.",
               digits = 1,
               booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position"))

```

## Verbs {#verbs}

Common operations on missing data can be considered to be verbs: scan, replace, add, shadow, impute, track, and flag. Missing values can be **scanned** to find possible missings not coded as `NA`. These values can then be **replaced** with an `NA` value. To facilitate exploration, summaries of missingness can be **added** as a summary column to the original data. Data values can be augmented with **shadow** values, which helps explore missing data, as well as facilitating the process of **imputing**, and **tracking**. Finally, unusual or specially coded missing values can be **flagged**. These verbs are described in this section.

### scan: Searching for common missing value labels  {#verbs-search}

There are many common values that mean "missing". For example, "N/A" "Not Available", "MISSING", or white space " ". Before replacing these values with missing values (`NA`), it is useful to first _scan_ the data to get a sense of the magnitude of the problem. The  `miss_scan_count()` function finds where specified values occur in data. For example, for the example data, `dat_ms`:

```{r setup-miss-scan-count}

dat_ms <- tibble::tribble(~x,  ~y,    ~z,
                         1,   "A",   -100,
                         3,   "N/A", -99,
                         NA,  NA,    -98,
                         -99, "E",   -101,
                         -98, "F",   -1)

knitr::kable(dat_ms, 
             caption = "An example dataset with variables x, y and z containing some common unusual missing values.",
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position"))
```

All occurrences of -99 can be searched for using the code:

```{r miss-scan-print, eval = FALSE, echo = TRUE}
miss_scan_count(data = dat_ms, search = -99)
```

This returns a table of the occurrences of that value in each variable (Table \@ref(tab:miss-scan-count)). A list of common NA values is provided in `naniar` in the `common_na_strings` and `common_na_numbers` for characters and numbers. These contain values like "n/a", "na ", and ".", and -9, -99.

```{r miss-scan-count}
miss_scan_count(data = dat_ms, 
                search = -99) %>% 
  knitr::kable(
    caption = "Table of the occurences of the search '-99' in the data, 'dat\\_ms'. There is one occurence if -99 in variables x and z.",
    booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position"))
```

### replace: Replacing missing value label with another {#verbs-replace-with}

Once the magnitude of the unusual missing values has been assessed, these values need to be replaced with appropriate missing values, using `replace_with_na`. This is useful when there is certainty of which values mean missing. For example "N/A", "N A", "Not Available", and "missing". Using the previous toy dataset, `dat_ms`, the value -99, or both -99 and 98 can be replaced with NA as follows:

```{r replace-with-na-code-chunk, echo = TRUE}
replace_with_na(dat_ms, replace = list(x = -99))

replace_with_na(dat_ms, replace = list(x = c(-99, -98)))
```

For flexibility, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`. This means `replace_with_na` works for **all** variables with `replace_with_na_all`, **at** selected variables with `replace_with_na_at`, and **if** variables that meet some condition with `replace_with_na_if`. The syntax for these is slightly different to `replace_with_na`, but powerful:

```{r replace-with-na-scoped-variants, echo = TRUE}
replace_with_na_at(data = dat_ms,
                   .vars = "x",
                   condition = ~.x == -99)

replace_with_na_if(data = dat_ms,
                 .predicate = is.character,
                 condition = ~.x == "N/A")

replace_with_na_all(data = dat_ms,
                    condition = ~.x == -99)

```

### add: Adding missingness summary variables {#verbs-add-cols}

Summary information such as the number of missings in a given case can be useful in understanding missingness structure, and even more so when this information is kept alongside the data. Functions starting with `add_` exist in `dplyr` to add count information for particular groups or conditions to the data. Similarly, `naniar` includes `add_` functions to add missingness summary information back to the data, such as the number or proportion of missingness, the missingness cluster, or if there are any missings, using: `add_n_miss()`
`add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively.

An example use of these features is in @Tierney2015,
where the proportion, number, or cluster of missings is used as the outcome in a
model. The variables in the dataset can then be used to predict the outcome, identifying variables and values important in predicting
missingness structures. There are also functions for adding information about shadow values or nice labels for if there are any missing values with `add_label_shadow()` and `add_label_missings()`:

```{r add-missing-info, echo = TRUE}
add_n_miss(dat_ms)
add_prop_miss(dat_ms)
add_miss_cluster(dat_ms)
```

### shadow: Creating `nabular` data {#verbs-nabular}

_Nabular_ data has the shadow matrix column-bound to the existing data. This facilitates visualization and summaries, and allows for imputed values to be appropriately tracked. _Nabular_ data can be created with `bind_shadow()` or `nabular()`. The `bind_shadow()` function is so named to borrow from existing functions named `bind` - such as `bind_rows()` and `bind_cols()` from `dplyr`, which bind rows or columns into a dataframe. An example use of `bind_shadow()` on the example data, `dat_ms` is shown below:

```{r bind-shadow, echo = TRUE}
bind_shadow(dat_ms)
```

### flag: Describing different types of  missing values {#verbs-recode}

Unusual or spurious data values are often identified and `flagged`. For example, there might be special codes to mark an individual dropping out of a study, known instrument failure in weather instruments, or for values censored in analysis. These special types of missingness can be encoded in the shadow matrix of  _nabular_  data, using the `recode_shadow()` function. This provides a fluent interface to recode, or flag, the shadow information in the shadow matrix as a special type of missing value. Using `recode_shadow()` requires specifying the variable you want to contain the flagged value, the condition you want this to occur at, and a suffix for the new type of missing value. This is then recoded as a new factor level in the shadow matrix, so that every column is aware of all possible new values of missingness. For example, the values -99 and -98 could be recoded to mean a broken machine sensor for the variable x:

```{r recode-shadow, eval = TRUE, echo = TRUE}
dat_ms %>%
  bind_shadow() %>%
  recode_shadow(x = .where(x == -99 ~ "broken_sensor"))
```

### impute: Imputing values {#verbs-impute}

`naniar` does not reinvent the wheel for imputation, as there are many robust R packages for imputation, such as `mice`, `mi`, `VIM`, and `simputation`. However, `naniar` does provide a few imputation methods to facilitate exploration and visualizations, which were not otherwise available: `impute_below` and `impute_mean` and `impute_median`.  These imputation methods are useful to explore structure in missingness, but are not recommended for use in analysis.

The `impute_below` function imputes values below the minimum of the data, with some jitter to reduce overplotting, but has options to change the default amount it is shifted below, and the amount of jitter.

```{r demonstrate-impute-below}
airquality %>%
  impute_below_at(vars(Ozone)) %>%
  select(Ozone, Solar.R) %>%
  head()
```

Similar to `simputation`, each `impute_` function returns the data with values imputed. However, `naniar` does not use a formula syntax, and instead each function implements "scoped variants" `_all`, `_at` and `_if`. Here, `_all` operates on **all** columns, `_at` operates **at** specific columns, and `_if` operates on columns **if** they meet some condition (such as `is.numeric` or `is.character`). The `impute_` functions with no scoped variant, e.g., `impute_mean`, will work on a single vector, but not a data.frame. An example usage is shown below:

```{r impute-vector, echo = TRUE}

impute_mean(airquality$Ozone) %>% head()

impute_mean_at(airquality, .vars = vars(Ozone)) %>% head()

impute_mean_if(airquality, .predicate = is.integer) %>% head()

impute_mean_all(airquality) %>% head()

```

One drawback to imputation functions in this form is that the location of imputed values is not tracked. This is covered in Section \@ref(verbs-track).

### track: Shadow and impute missing values {#verbs-track}

To evaluate imputations they need to be tracked, which is achieved by combining the verbs `bind_shadow`, `impute_`, and `add_label_shadow`. The missing values can then be referred to by their shadow variable, `_NA`. The missingness of any observation can be referred to with `any_missing` (Figure \@ref(fig:track-impute-example)). The code chunk below shows the track pattern, first using `bind_shadow`, then imputing with `impute_knn`, and adding a `label_shadow`:

```{r bind-impute-label-example, echo = TRUE}

aq_imputed <- airquality %>%
  bind_shadow() %>%
  as.data.frame() %>%
  simputation::impute_knn(Ozone ~ .) %>%
  simputation::impute_knn(Solar.R ~ .) %>%
  add_label_shadow() %>%
  as_tibble()

```

Missing values can then be shown in a scatterplot by setting the `color` aesthetic in ggplot to `any_missing` (Figure \@ref(fig:track-impute-example)A), or in a density plot looking at one variable, using the `fill = any_missing`, (Figure \@ref(fig:track-impute-example)B) and \@ref(fig:track-impute-example)C).   
```r
ggplot(aq_imputed,
       aes(x = Ozone,
           y = Solar.R,
           color = any_missing)) + 
  geom_point() +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "bottom")

ggplot(aq_imputed,
       aes(x = Ozone,
           fill = any_missing)) + 
  geom_density(alpha = 0.3) + 
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "none")

ggplot(aq_imputed,
       aes(x = Solar.R,
           fill = any_missing)) + 
  geom_density(alpha = 0.3) + 
  scale_fill_brewer(palette = "Dark2")
```

```{r track-impute-example, fig.show = "hold", fig.cap = "Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values imputed using simputations `impute\\_knn` function, with imputed values colored green and data values orange. Imputed values are similar, but slightly different to existing data.", fig.height = 2, fig.width = 5, out.width = "100%", echo = FALSE}

ggplot(aq_imputed,
       aes(x = Ozone,
           y = Solar.R,
           color = any_missing)) + 
  geom_point() +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none") + 
  labs(tag = "A")

ggplot(aq_imputed,
       aes(x = Ozone,
           fill = any_missing)) + 
  geom_density(alpha = 0.3) + 
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "none") + 
  labs(tag = "B")

ggplot(aq_imputed,
       aes(x = Solar.R,
           fill = any_missing)) + 
  geom_density(alpha = 0.3) + 
  scale_fill_brewer(palette = "Dark2") + 
  theme(legend.position = "none") + 
  labs(tag = "C")

```

Imputed values can also be compared to complete case data by grouping by `any_missing`, and then summarizing (Table \@ref(tab:impute-summary)).

```r
aq_imputed %>%
  group_by(any_missing) %>%
  summarise_at(.vars = vars(Ozone),
               .funs = funs(min, mean, median, max))
```

```{r impute-summary}
aq_imputed %>%
  group_by(any_missing) %>%
  summarise_at(.vars = vars(Ozone),
               .funs = funs(min, mean, median, max)) %>%
  knitr::kable(
    digits = 1, 
    caption = "Summary statistics of the previously imputed and complete cases. The mean and median values are similar, but the minimum and maximum values are different. The comparison of imputed values is similar to other dplyr summary workflows.",
    booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position"))
```

# Application {#case-study}

```{r data-setup, include = FALSE}

housing_raw <-
  readr::read_csv("../data/melbourne_housing_raw.csv") %>%
  janitor::clean_names() %>%
  dplyr::rename(region_name = regionname,
         property_count = propertycount) %>%
  dplyr::mutate(date = lubridate::dmy(date)) %>%
  dplyr::rename(latitude = lattitude,
                longitude = longtitude) %>%
  # let's create monthly quarters
  mutate(yr_qtr = as.ordered(lubridate::quarter(date, 
                                             with_year = TRUE,
                                             fiscal_start = 1))) %>%
  # drop price
  tidyr::drop_na(price) %>%
  # make price the log of price
    mutate(price = log10(price))

```

```{r data-clean, include = FALSE}

# compact down the seller levels
count_dat <- count(housing_raw,seller_g) %>% 
  arrange(-n) %>%
  mutate(cumu = cumsum(n),
         pct = cumu / nrow(housing_raw),
         pct_change = pct - lag(pct)) %>%
  tibble::rowid_to_column() %>%
  mutate(seller = case_when(
    pct < 0.52 ~ as.character(seller_g),
    pct > 0.52 & pct < 0.60 ~ "seller_g1",
    pct > 0.60 & pct < 0.65 ~ "seller_g2",
    pct > 0.65 & pct < 0.70 ~ "seller_g3",
    pct > 0.70 & pct < 0.75 ~ "seller_g4",
    pct > 0.75 & pct < 0.80 ~ "seller_g5",
    pct > 0.80 & pct < 0.85 ~ "seller_g6",
    pct > 0.85 & pct < 0.90 ~ "seller_g6",
    pct > 0.90 & pct < 0.95 ~ "seller_g6",
    pct > 0.95 ~ "seller_g5"
  ))

# drop these three variables, as we won't use them in analysis, they
# would need extensive recoding

housing <- housing_raw %>%
  mutate_if(is.character, as.factor) %>%
  mutate_at(vars(rooms,
                 bedroom2,
                 bathroom,
                 car),
            as.factor) %>%
  # recode / collapse factors with many levels and few obs
  mutate(bathroom = fct_other(bathroom,
                                drop = c("4", "5", "6", "7", "8", "9"),
                                other_level = "4+")) %>%
  mutate(bedroom2 = fct_other(bedroom2,
                               drop = c("5", "6", "7", "8", "9", "10", 
                                        "12", "16", "20"),
                               other_level = "5+")) %>%
  mutate(rooms = fct_other(rooms,
                             drop = c("6", "7", "8", "9", "10", 
                                      "12", "16"),
                             other_level = "6+")) %>%
  mutate(car = fct_other(car,
                           drop = c("4", "5", "6", "7", "8", "9", "10",
                                    "11", "18"),
                           other_level = "4+")) %>%
  left_join(select(count_dat, seller_g, seller), by = "seller_g")  %>%
  # drop seller_g
  select(-seller_g) %>%
  mutate(seller = as.factor(seller))

```

To illustrate the methods, data on housing for the city of Melbourne from January 28, 2016 to March 17, 2018 is used. The data was compiled by scraping weekly property clearance data [@kaggle2018]. There are `r scales::number(nrow(housing), big.mark = ",")` properties, and `r ncol(housing)` variables in the dataset. The variables include the type of real estate (town house, unit, house), suburb, method of selling, number of rooms, price, real estate agent, date of sale, and the distance from the Central Business District (CBD).

The goal in analyzing this data is to accurately predict Melbourne housing prices. The data contains many missing values. As a precursor to building a predictive model, this analysis focusses on understanding the patterns of missingness. This section shows how the methods from previous sections are used together in a data analysis workflow.

## Exploring patterns of missingness {#case-study-explore-pattern}

Figure \@ref(fig:housing-miss-case-var)A shows 9 variables with missing values. The most missings are in building area, followed by year built, and land size, with similar amounts of missingness in Car, bathroom, bedroom2, longitude, and latitude. Figure \@ref(fig:housing-miss-case-var)B reveals there are up to 50% missing values in cases, and that the majority of cases have more than 5% values missing. The variables building area and year built have more than 50% missing data, and so could perhaps be omitted from subsequent analysis, as imputed values are likely to be spurious.

```{r housing-miss-case-var, fig.cap = "The amount of missings in each variable (A) and in each case (B) for Melbourne housing data. (A) Build area and year built have more than 50\\% missing values, and car, bathroom, bedroom2 and longitude and latitude have about 25\\% missings. (B) There are between 5 and 50\\% missing values in cases. There are many missing values in the data, with the majority of missingness being in selected cases and variables.", fig.height = 4, fig.width = 8, out.width = "100%"}

q1 <- gg_miss_var(housing, show_pct = TRUE) + labs(tag = "A")

q2 <- gg_miss_case(housing, show_pct = TRUE) + labs(tag = "B")

gridExtra::grid.arrange(q1, q2, ncol = 2)

```

Possible missingness structures are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \@ref(fig:applic-vis-miss)). There are three main clusters of missing data: at the top there are many variables missing together; in the middle, building price and year built missing together, at the bottom, building area, year built, and land size are missing.

```{r applic-vis-miss, fig.height = 5, fig.width = 6.25, out.width = "85%", fig.show='hold', fig.cap = "Heatmap of clustered missingness for the housing data. Three groups of missingness are apparent, at the top for building area to longitude, the middle for building area and year built, at the end for building area, year built, and landsize. There is some structure in the missings."}

housing %>%
  # sample_frac(0.25) %>%
  vis_miss(cluster = TRUE,
           sort_miss = TRUE,
           show_perc_col = FALSE)

```

Missingness patterns can also be shown with an `upset` plot [@Conway2017], to display the 8 of intersecting sets of missing variables (see Figure \@ref(fig:applic-vis-miss). 
Two patterns stand out: two, and five variables missing. This provides further evidence of the patterns of missingness seen in Figures \@ref(fig:housing-miss-case-var) and \@ref(fig:housing-upset).

```{r housing-upset, fig.height = 5, fig.width = 8, fig.cap = "An upset plot of eight sets of missingness in the housing data. Two missingness patterns are clear, year built and building area, and lattitude through to building area.", out.width = "100%"}

gg_miss_upset(housing, 
              nsets = 8,
              order.by = "freq")

```

Tabulating the number of missings in variables (see Table \@ref(tab:housing-miss-var-case-table) (left)) shows three groupings: 6254 - 6824 missings cases in variables, 9265 in another variable, and between 15163 and 16736 missing. Tabulating missings in cases (Table \@ref(tab:housing-miss-var-case-table) (right)) shows 32% of cases have complete data, 26% have 2 missing, 22% have 8 missing, 12% have 1 missing, and 5% have 3 missing, the remaining cases with missings are less than 1%. Two variables with more than 50% missingness are omitted from analysis: building area and year built.

```{r housing-miss-var-case-table}

t1 <-  miss_var_table(housing) 
t2 <- miss_case_table(housing)

knitr::kable(list(t1, t2),
             caption = "Tabulating missingness for variables (left) and cases (right) to understand missingness patterns. 14 variables have between 0 and 3 missings, then 6 variables have 6000 - 9000 missings, and 2 variables have 15 - 16,000 missing values. About 30\\% of cases have no missing values, then 45\\% of cases have between 1 and 6 missing values, and then about 23\\% of cases have 8 or more missings. There are different patterns of missingness in variables and cases, but they can be broken down into smaller groups.",
             digits = 1,
             booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position"))

```

## Exploring missingness patterns for imputation {#case-study-explore-for-imp}

Using the information from the overview visualizations and tables, the following variables are explored for features predicting missingness: Land size, latitude, longitude, bedroom2, bathroom, car, and land size.

Missingness structure can be better understood and evaluated by using the data to predict where the data goes missing. This can be achieved by clustering the missing values into groups, and then applying classification and regression trees (CART) to find those variables and values that predict missingness clusters [@Tierney2015; @Barnett2017]. Two clusters are identified, one for each missingness cluster. The missingness clusters are predicted using all variables in the dataset with the CART package `rpart` [@rpart], and plotted using the `rpart.plot` package [@rpart-plot]. 
Importance scores from the CART model describe those variables important for predicting missingness: rooms, price, suburb, council area, distance, and region name. These variables are important for predicting missingness, so are important to include in the imputation model.

```{r housing-cluster-miss}
housing_cls <- add_miss_cluster(housing, n_clusters = 2)
```

```{r rpart-fit-cluster}
library(rpart)

miss_cls_pre_fit <- housing_cls %>%
  mutate(year = lubridate::year(date),
         quarter = lubridate::quarter(date)) %>%
  mutate_at(vars(year, quarter),
            as.factor) %>%
  select(suburb,
         council_area,
         postcode,
         method, 
         region_name,
         rooms,
         seller,
         type,
         year,
         quarter,
         price,
         property_count,
         distance,
         latitude,
         longitude,
         miss_cluster)

miss_cls_fit_rpart <- rpart(factor(miss_cluster) ~ ., 
                            data = miss_cls_pre_fit)

```

```{r rpart-plot, fig.height = 2, fig.width = 4, out.width = "100%", fig.align = "center", fig.cap = "Decision tree output predicting the clusters of missingness. Type of house, the year quarter, and year were important for predicting missingness cluster. The cluster with the most missingness was for quarters 1 and 4, for 2017 and 2018. Type of house, year, and year quarter are important features related to the missingness structure."}

rpart.plot::prp(miss_cls_fit_rpart,
                type = 4,
                extra = 2,
                fallen.leaves = TRUE,
                prefix = "cluster = ",
                suffix = " \nObservations",
                box.col = "lightgrey",
                border.col = "grey",
                branch.col = "grey40")

```

```{r vis-miss-each-node, fig.width = 3, fig.height = 3, out.width = "49%", fig.show='hold'}

data_in <- housing_cls %>%
  mutate(year = lubridate::year(date),
         month = lubridate::month(date),
         quarter = lubridate::quarter(date)) %>%
  mutate_at(vars(year, month, quarter),
            as.factor)

node_1 <- data_in %>% 
  filter(type == "h")

node_2 <- data_in %>% 
  filter(type != "h",
         quarter == 2 | quarter == 3)

node_3 <- data_in %>%
  filter(type != "h",
         quarter == 1 | quarter == 4,
         year == 2016)

node_4 <- data_in %>%
  filter(type != "h",
         quarter == 1 | quarter == 4,
         year == 2017 | year == 2018)

```

## Imputation and diagnostics {#case-study-imp-diagnosis}

Two different imputation techniques are used: simple linear regression and K nearest neighbors. Values are imputed stepwise in ascending order of missingness. The `simputation` R package [@simputation] is used to impute the values, applying the track missings pattern described in Section \@ref(verbs-track), to assess imputed values.

```{r clean-data-for-analysis, include = FALSE}

impute_knn <- simputation::impute_knn
dat_house <- housing_cls %>%
  mutate(year = lubridate::year(date),
         # month = lubridate::month(date),
         quarter = lubridate::quarter(date)) %>%
  mutate_at(vars(car, bathroom, bedroom2), as.integer)

dat_house_knn <- dat_house %>% 
  bind_shadow() %>%
  as.data.frame() %>% 
  # postcode
  impute_knn(postcode ~ type + quarter + year + rooms + price + distance) %>%
  # distance
  impute_knn(distance ~ type + quarter + year + rooms + price) %>%
  # property_count
  impute_knn(property_count ~ type + quarter + year + rooms + price + distance) %>%
  # longitude
  impute_knn(longitude ~ type + quarter + year + rooms + price + distance) %>%
  # latitude
  impute_knn(latitude ~ type + quarter + year + rooms + price + distance) %>%
  # bedroom2
  impute_knn(bedroom2 ~ type + quarter + year + rooms + price + distance) %>%
  # bathroom
  impute_knn(bathroom ~ type + quarter + year + rooms + price + distance) %>%
  # car
  impute_knn(car ~ type + quarter + year + rooms + price + distance) %>%
  # landsize
  impute_knn(landsize ~ type + quarter + year + rooms + price + distance) %>%
  add_label_shadow()

```

```{r impute-knn}

dat_house_lm <- dat_house %>%
  bind_shadow() %>%
  as.data.frame() %>%
  # postcode
  impute_knn(postcode ~ type + quarter + year + rooms + price + distance) %>%
  # distance
  impute_lm(distance ~ type + quarter + year + rooms + price) %>%
  # property_count
  impute_lm(property_count ~ type + quarter + year + rooms + price + distance) %>%
  # longitude
  impute_lm(longitude ~ type + quarter + year + rooms + price + distance) %>%
  # latitude
  impute_lm(latitude ~  type + quarter + year + rooms + price + distance) %>%
  # bedroom2
  impute_lm(bedroom2 ~  type + quarter + year + rooms + price + distance) %>%
  # bathroom
  impute_lm(bathroom ~  type + quarter + year + rooms + price + distance) %>%
  # car
  impute_lm(car ~ type + quarter + year + rooms + price + distance) %>%
  # landsize
  impute_lm(landsize ~  type + quarter + year + rooms + price + distance) %>%
  add_label_shadow() %>%
  # recode integers back
  mutate_at(vars(bedroom2, bathroom, car), as.integer)

```

```{r compare-assess-imputations}

dat_house_cc <- na.omit(dat_house)

bound_models <- bind_rows(lm = dat_house_lm,
                          knn = dat_house_knn,
                          cc = dat_house_cc,
                          .id = "imp_model") %>%
  as_tibble()

```

Imputed data are combined by row binding imputation models. The data are reshaped into long format to compare imputed values for each model for 4 variables (Figure \@ref(fig:imputed-by-model)). Compared to KNN imputed values, the linear model imputed values closer to the mean value.

```{r imputed-by-model, fig.cap = "Boxplots of complete case data, and data imputed with KNN or linear model for different variables. (A) number of bedrooms, (B) number of bathrooms, (C) number of carspots, and (D) landsize (on a log10 scale). KNN had similar results to complete case, and linear model had a lower median for cars and fewer extreme values for bedrooms.", fig.height = 2.5, fig.width = 7, out.width = "95%"}

# Now we want to look at the distribution of the values, so we need to do
# some gathering
bound_models_gather <- bound_models %>%
  select(
         bedroom2,
         bathroom,
         car,
         landsize,
         any_missing,
         imp_model) %>%
  gather(key = "key",
         value = "value",
         -any_missing,
         -imp_model)

p1 <- bound_models %>%
  ggplot(aes(x = imp_model,
             y = bedroom2)) + 
  geom_boxplot()

p2 <- bound_models %>%
  ggplot(aes(x = imp_model,
             y = bathroom)) + 
  geom_boxplot()

p3 <- bound_models %>%
  ggplot(aes(x = imp_model,
             y = car)) + 
  geom_boxplot()

p4 <- bound_models %>%
  ggplot(aes(x = imp_model,
             y = landsize)) + 
  geom_boxplot() + 
  scale_y_log10()

cowplot::plot_grid(p1,
          p2,
          p3,
          p4,
          nrow = 1,
          labels = "AUTO")

```

```{r SUPP-FIG-distribution-imputed-vals-all, fig.width = 6, fig.height = 6, out.width = "100%", eval = FALSE}

bound_models_gather %>%
  filter(any_missing == "Missing") %>%
  ggplot(aes(x = imp_model,
             y = value,
             colour = imp_model)) +
  geom_boxplot() + 
  facet_wrap(~key, scales = "free_y") + 
    theme(legend.position = "bottom")

```

## Assess model predictions {#case-study-assess-model}

```{r fit-each-model}

dat_fit <- dat_house %>%
  mutate_at(vars(car,
                 bathroom,
                 bedroom2,
                 type,
                 year,
                 quarter,
                 rooms),
            .funs = as.factor)

dat_fit_knn <-  dat_house_knn %>%
  mutate_at(vars(car,
                 bathroom,
                 bedroom2,
                 type,
                 year,
                 quarter,
                 rooms),
            .funs = as.factor)

dat_fit_lm <-  dat_house_lm %>%
  mutate_at(vars(car,
                 bathroom,
                 bedroom2,
                 type,
                 year,
                 quarter,
                 rooms),
            .funs = as.factor)

dat_house_cc <- na.omit(dat_fit)

fit_lm_dat_house_cc <- lm(price ~ landsize + 
                            car + 
                            bathroom + 
                            bedroom2 +
                            type + 
                            year + 
                            quarter +
                            rooms +
                            distance,
                          dat_house_cc)

fit_lm_dat_house_knn <- lm(price ~ landsize + 
                            car + 
                            bathroom + 
                            bedroom2 +
                            type + 
                            year + 
                            quarter +
                            rooms +
                            distance,
                           dat_fit_knn)

fit_lm_dat_house_lm <- lm(price ~ landsize + 
                            car + 
                            bathroom + 
                            bedroom2 +
                            type + 
                            year + 
                            quarter +
                            rooms +
                            distance,
                          dat_fit_lm)

```

The coefficients of the linear model of log price vary for room (Figure \@ref(fig:tidy-coefs)), for different imputed datasets. Notably, complete cases  resulted in underestimating the impact of room on log price. A partial residual plot  (Figure \@ref(fig:partial-resid)) shows there is not much variation amongst the models from the different datasets.

```{r tidy-coefs, fig.height = 2, fig.width = 7, fig.cap = "Visualization of the variation in coefficients for linear model of log price for each of the different datasets for the number of rooms. In red is complete case (cc), in green is the knn imputed dataset, and in blue the imputed by linear model. Using the complete case dataset produced smaller coefficients compared to the imputed models.", out.width = "100%"}

tidy_fit_cc <- broom::tidy(fit_lm_dat_house_cc)
tidy_fit_knn <- broom::tidy(fit_lm_dat_house_knn)
tidy_fit_lm <- broom::tidy(fit_lm_dat_house_lm)

tidy_fits <- bind_rows(cc = tidy_fit_cc,
                       knn = tidy_fit_knn,
                       lm = tidy_fit_lm,
                       .id = "model")

tidy_fits_data <-  tidy_fits %>%
  filter(term != "(Intercept)") %>%
  filter(grepl("rooms", term)) %>%
  select(model, term, estimate) %>%
  spread(key = model,
         value = estimate)

tidy_fits %>%
  filter(term != "(Intercept)") %>%
  filter(grepl("rooms", term)) %>%
  select(model, term, estimate) %>%
  ggplot(aes(x = estimate,
             y = term,
             colour = model)) + 
  geom_point() +
  theme_minimal() +
  labs(x = "Estimate",
       y = "Term") + 
  guides(colour = guide_legend(title = "Model"))

```

```{r partial-resid, fig.cap = "Partial residual plot for each data set, complete cases (cc), imputed with KNN (knn), and imputed with a linear model (lm). There is not much variation amongst the different datasets from the different imputation methods.", fig.width = 9, fig.height = 3, out.width = "100%"}

aug_fit_cc <-  broom::augment(fit_lm_dat_house_cc) %>% as_tibble()
aug_fit_knn <- broom::augment(fit_lm_dat_house_knn) %>% as_tibble()
aug_fit_lm <-  broom::augment(fit_lm_dat_house_lm)%>% as_tibble()

aug_fits <- bind_rows(cc =  aug_fit_cc,
                      knn = aug_fit_knn,
                      lm =  aug_fit_lm,
                      .id = "model")

aug_fits %>%
  ggplot(aes(x = .fitted,
             y = .resid)) + 
  geom_point(alpha = 0.1) + 
  facet_wrap(~model)

```

## Summary {#case-study-summary}

The `naniar` and `visdat` packages build on existing tidy tools and strike a compromise between automation and control that makes analysis efficient, readable, but not overly complex. Each tool has clear intent and effects - plotting or generating data or augmenting data in some way. This reduces repetition and typing for the user, making exploration of missing values easier as they follow consistent rules with a declarative interface.

# Discussion {#discussion}

This paper has described new methods for exploring, visualizing, and imputing missing data. The work was motivated by recent developments of tidy data, and extends them for better missing value handling. The methods have standard outputs, function arguments, and behavior. This provides consistent workflows centered around data analysis that integrate well with existing imputation methodology, visualization, and modelling.

The new data structures discussed in the paper could be used to create different visualizations than were shown in the paper. The analyst can use the data structures to decide on appropriate visualization for their problem. The data structures could also be used to support interactive graphics, in the manner of MANET and ggobi. Linking the plots (linked brushing) to explore missingness or animating between different sets of imputed values. New packages like `plotly` [@plotly] facilitate interactive graphics, and packages like `gganimate` facilitate animations [@gganimate].

Other data structures such as spatial data, time series, networks, and
longitudinal data would be supported by the inherently tabular,
_nabular_ data, if they are first structured as wide tidy format. Large data may need special handling, and additional features like efficient storage of purely imputed values and lazy evaluation. Special missing value codes could be improved by creating special classes, or expanding low level representation of `NA` at the source code level.

The methodology described in this paper can be used in conjunction with other approaches to understand multivariate missingness dependencies (e.g. decision trees [@Tierney2015], latent group analysis [@Barnett2017], and PCA [@FactoMineR]) . Evaluating imputed values using a testing framework like @VanBuuren2012 is also supported.

The approach meshes with the dynamic nature of data analysis, allowing the analyst to go from raw data to model data in a fluid workflow.

\newpage

# Acknowledgements

The authors would like to thank Miles McBain, for his key contributions and discussions on the `naniar` package, in particular for helping implement `geom_miss_point`, and for his feedback on ideas, implementations, and names. We also thank Colin Fay for his contributions to the `naniar` package, in particular for his assistance with the `replace_with_na` functions. We also thank Earo Wang and Mitchell O'Hara-Wild for the many useful discussions on missing data and package development, and for their assistance with creating elegant approaches that take advantage of the tidy syntax. We would also like to thank those who contributed pull requests and discussions on the `naniar` package, in particular Jim Hester and Romain François for improving the speed of key functions, Ross Gayler for discussion on special missing values, and Luke Smith for helping `naniar` be more compliant with `ggplot2`.

# References

### 2018-09-06 data/01-read-clean.R
add data

@@ -1,0 +1,13 @@
+# initial clean of the data
+library(tidyverse)
+library(naniar)
+
+housing <- read_csv("data/housing-data/raw/Melbourne_housing_FULL.csv") %>%
+  janitor::clean_names() %>%
+  rename(region_name = regionname,
+         property_count = propertycount)
+
+glimpse(dat_melb)
+
+write_csv(housing,
+          path = "data/housing-data/housing.csv")

### 2018-09-06 paper/tidy-missing-data-paper.Rmd
* Added JCGS CSL * updated gitignore to include preamble * updated preamble with pieces from Rob's Monash template

@@
 title: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
-author:
-- familyname: Tierney
-  othernames: Nicholas
-  address: Monash University
-  email: nicholas.tierney@gmail.com
-  correspondingauthor: true
-- familyname: Cook
-  othernames: Dianne
-  address: Monash University
-abstract: "Despite the large body of research on missing value distributions and imputation, there is comparatively little literature on how to make it easy to handle, explore, and impute missing values in data.
-This paper addresses this gap.
-The new methodology builds upon tidy data principles, with a goal to integrating missing value handling as an integral part of data analysis workflows.
-New data structures are defined along with new functions (verbs) to perform common operations.
-Together these provide a cohesive framework for handling, exploring, and imputing missing values.
-These methods have been made available in the R package `naniar`."
-keywords: "workflow, statistical computing, data science, data visualization, tidyverse, data pipeline"
-wpnumber: 1/2018
-jelcodes: C10,C14,C22
-blind: false
-cover: true
+author: "Nicholas Tierney, Department of Econometrics and Business Statistics, Monash University, **Corresponding author (nicholas.tierney@gmail.com)** \\newline Dianne Cook, Department of Econometrics and Business Statistics, Monash University"
+date: "September 6, 2018"
 toc: false
 bibliography: tidy-missing-data-paper.bib
-biblio-style: authoryear-comp
 output:
-  MonashEBSTemplates::workingpaper:
+  bookdown::pdf_document2:
     fig_caption: yes
@@
     number_sections: yes
-    citation_package: biblatex
+csl: jcgs.csl
 ---
+
+**Abstract**:
+
+Despite the large body of research on missing value distributions and imputation, there is comparatively little literature on how to make it easy to handle, explore, and impute missing values in data.
+This paper addresses this gap.
+The new methodology builds upon tidy data principles, with a goal to integrating missing value handling as an integral part of data analysis workflows.
+New data structures are defined along with new functions (verbs) to perform common operations.
+Together these provide a cohesive framework for handling, exploring, and imputing missing values.
+These methods have been made available in the R package `naniar`.
+
+_Keywords:_ workflow, statistical computing, data science, data visualization, tidyverse, data pipeline
+
+\newpage
 
@@
 MANET also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot.
-The approach of @Swayne1990 in the software XGobi, further developed in ggobi [@Cook2007], focussed more on multivariate quantitative data.
+The approach of @Swayne1998 in the software XGobi, further developed in ggobi [@Cook2007], focussed more on multivariate quantitative data.
 Missingness is incorporated into plots in ggobi by setting them to be 10% below the minimum value.
@@
 
-```{r rpart-plot, fig.height = 2, fig.width = 4, out.width = "100%", fig.align = "center", fig.cap = "Decision tree output predicting the clusters of missingness.
+```{r rpart-plot, fig.height = 4, fig.width = 8, out.width = "100%", fig.align = "center", fig.cap = "Decision tree output predicting the clusters of missingness.
 Type of house, the year quarter, and year were important for predicting missingness cluster.

### 2018-09-06 paper/tidy-missing-data-paper.Rmd
add installation instructions

@@
                       fig.align = "center",
-                      # dev = c("png"), 
-                      # dpi = 300,
                       fig.width = 4, 
@@
 library(simputation)
+# if paper doesn't compile, use development version of naniar
+# install.packages("remotes")
+# remotes::install_github("njtierney/naniar")
 library(naniar) 

### 2018-09-06 paper/tidy-missing-data-paper.Rmd
remove extra parenthesis
update vis_miss to use PNG to avoid strange printing behaviour

@@
 However, when making a plot, missing values are simply dropped, albeit with a
-warning (see Figure \@ref(fig:warning))). Most other analysis software will
+warning (see Figure \@ref(fig:warning)). Most other analysis software will
 simply drop cases with missings without warning.
@@
 
-```{r vis-miss, fig.height = 4, fig.width = 6, out.width = "95%", fig.show = 'hold', echo = FALSE, fig.cap = "Heatmap visualizations of missing data for the airquality dataset.
+```{r vis-miss, fig.height = 4, fig.width = 6, out.width = "95%", fig.show = 'hold', echo = FALSE, dev = "png", dpi = 300, fig.cap = "Heatmap visualizations of missing data for the airquality dataset.
 (A) The default output and (B) ordered by clustering on rows and columns.
@@
 Three groups of missingness are apparent, at the top for building area to longitude, the middle for building area and year built, at the end for building area, year built, and landsize.
-There is some structure in the missings."}
+There is some structure in the missings.", dev = "png", dpi = 300}
 

### 2018-09-07 paper/tidy-missing-data-paper.Rmd
minor changes from submission to JCGS - mostly typos / changing some devices to PNG

@@
 title: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
-author: "Nicholas Tierney, Department of Econometrics and Business Statistics, Monash University, **Corresponding author (nicholas.tierney@gmail.com)** \\newline Dianne Cook, Department of Econometrics and Business Statistics, Monash University"
-date: "September 6, 2018"
+author:
+- familyname: Tierney
+  othernames: Nicholas
+  address: Monash University
+  email: nicholas.tierney@gmail.com
+  correspondingauthor: true
+- familyname: Cook
+  othernames: Dianne
+  address: Monash University
+abstract: "Despite the large body of research on missing value distributions and imputation, there is comparatively little literature on how to make it easy to handle, explore, and impute missing values in data.
+This paper addresses this gap.
+The new methodology builds upon tidy data principles, with a goal to integrating missing value handling as an integral part of data analysis workflows.
+New data structures are defined along with new functions (verbs) to perform common operations.
+Together these provide a cohesive framework for handling, exploring, and imputing missing values.
+These methods have been made available in the R package `naniar`."
+keywords: "workflow, statistical computing, data science, data visualization, tidyverse, data pipeline"
+wpnumber: 1/2018
+jelcodes: C10,C14,C22
+blind: false
+cover: true
 toc: false
 bibliography: tidy-missing-data-paper.bib
+biblio-style: authoryear-comp
 output:
-  bookdown::pdf_document2:
+  MonashEBSTemplates::workingpaper:
     fig_caption: yes
@@
     number_sections: yes
-csl: jcgs.csl
+    citation_package: biblatex
 ---
-
-**Abstract**:
-
-Despite the large body of research on missing value distributions and imputation, there is comparatively little literature on how to make it easy to handle, explore, and impute missing values in data.
-This paper addresses this gap.
-The new methodology builds upon tidy data principles, with a goal to integrating missing value handling as an integral part of data analysis workflows.
-New data structures are defined along with new functions (verbs) to perform common operations.
-Together these provide a cohesive framework for handling, exploring, and imputing missing values.
-These methods have been made available in the R package `naniar`.
-
-_Keywords:_ workflow, statistical computing, data science, data visualization, tidyverse, data pipeline
-
-\newpage
 
@@
 This has some limitations.
-For example, for data with school grades and test scores, a boxplot visualization could have school grade mapped to the x axis, and test score to the y axis (See \@ref(fig:gg-box-na)).
+For example, for data with school grades and test scores, a boxplot visualization could have school grade mapped to the x axis, and test score to the y axis (See Figure \@ref(fig:gg-box-na)).
 If there are missings in a continuous variable like test score, `ggplot2` omits the missing values and prints a warning message.
@@
 This section discusses visualizations that are easy to remember and compose with other functions in a pipeline.
-Four different areas of visualization of missing data are discussed: overviews \@ref(overview-plots); univariate \@ref(univariate-plots), bivariate \@ref(bivariate-plots), and multivariate \@ref(multivariate-plots).
+Four different areas of visualization of missing data are discussed: overviews (Section \@ref(overview-plots)); univariate (Section \@ref(univariate-plots)), bivariate (Section \@ref(bivariate-plots)), and multivariate (Section \@ref(multivariate-plots)).
 All plots are created using `ggplot2`, giving users clearer control over the plot appearance, and customizability.
@@
 This means they integrate well with plotting and modelling tools.
-Section \@ref(single-num-sum) discusses single number summaries for all data and for cases and variables ; Section \@ref(sum-tab-missings) discusses variable- and case-wise summaries; and section \@ref(num-sum-w-group) shows how the design of these tools works with other tools in an analysis pipeline.
+Section \@ref(single-num-sum) discusses single number summaries for all data and for cases and variables ; Section \@ref(sum-tab-missings) discusses variable- and case-wise summaries; and Section \@ref(num-sum-w-group) shows how the design of these tools works with other tools in an analysis pipeline.
 
@@
 
-Missing values can then be shown in a scatterplot by setting the `color` aesthetic in ggplot to `any_missing` (Figure \@ref(fig:track-impute-example)A), or in a density plot looking at one variable, using the `fill = any_missing`, (Figure \@ref(fig:track-impute-example)B) and \@ref(fig:track-impute-example)C).
+Missing values can then be shown in a scatterplot by setting the `color` aesthetic in ggplot to `any_missing` (Figure \@ref(fig:track-impute-example)A), or in a density plot looking at one variable, using the `fill = any_missing`, (Figures \@ref(fig:track-impute-example)B) and \@ref(fig:track-impute-example)C).
 ```r

### 2018-09-07 README.Rmd
add README

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

This is the repository for the paper: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations" - you can find all the data and code to replicate the paper here.

### 2018-09-07 paper/tidy-missing-data-paper.tex
add the .tex file submitted to arxiv

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations},
            pdfauthor={Nicholas Tierney, Department of Econometrics and Business Statistics, Monash University, Corresponding author (nicholas.tierney@gmail.com) Dianne Cook, Department of Econometrics and Business Statistics, Monash University},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Expanding tidy data principles to facilitate missing data exploration,
visualization and assessment of imputations}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Nicholas Tierney, Department of Econometrics and Business Statistics,
Monash University, \textbf{Corresponding author
(\href{mailto:nicholas.tierney@gmail.com}{\nolinkurl{nicholas.tierney@gmail.com}})}
\newline Dianne Cook, Department of Econometrics and Business
Statistics, Monash University}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{September 6, 2018}

%% Any special functions or other packages can be loaded here.

%% FIGURE FLOATING taken from
%% https://stackoverflow.com/questions/16626462/figure-position-in-markdown-when-converting-to-pdf-with-knitr-and-pandoc
\usepackage{float}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] {
    \expandafter\origfigure\expandafter[H]
} {
    \endorigfigure
}

%% MONASH STUFF TAKEN FROM ROB

%% SPACING
\RequirePackage{setspace}
\spacing{1.5}

%% HEADERS AND FOOTERS
\RequirePackage{fancyhdr}
\pagestyle{fancy}
\rfoot{\Large\sffamily\raisebox{-0.1cm}{\textbf{\thepage}}}
\makeatletter
\lhead{\textsf{\expandafter{\@title}}}
\makeatother
\rhead{}
\cfoot{}
\setlength{\headheight}{15pt}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}
\fancypagestyle{plain}{%
\fancyhf{} % clear all header and footer fields
\fancyfoot[C]{\sffamily\thepage} % except the center
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}}

%% LINE AND PAGE BREAKING
\sloppy
\clubpenalty = 10000
\widowpenalty = 10000
\brokenpenalty = 10000
\RequirePackage{microtype}

%% PARAGRAPH BREAKS
\setlength{\parskip}{1.4ex}
\setlength{\parindent}{0em}

%% CAPTIONS
\RequirePackage{caption}
\DeclareCaptionStyle{italic}[justification=centering]
 {labelfont={bf},textfont={it},labelsep=colon}
\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}
\maketitle

\textbf{Abstract}:

Despite the large body of research on missing value distributions and
imputation, there is comparatively little literature on how to make it
easy to handle, explore, and impute missing values in data. This paper
addresses this gap. The new methodology builds upon tidy data
principles, with a goal to integrating missing value handling as an
integral part of data analysis workflows. New data structures are
defined along with new functions (verbs) to perform common operations.
Together these provide a cohesive framework for handling, exploring, and
imputing missing values. These methods have been made available in the R
package \texttt{naniar}.

\emph{Keywords:} workflow, statistical computing, data science, data
visualization, tidyverse, data pipeline

\newpage

\hypertarget{intro}{%
\section{Introduction}\label{intro}}

Tidy data (Wickham 2014) is a relatively new principle and suite of
tools that facilitate the process of converting raw data into a clean,
analysis-ready format that works efficiently in a data analysis
pipeline. In tidy data, missing values can be handled implicitly or
explicitly, at the analyst's discretion. However, when making a plot,
missing values are simply dropped, albeit with a warning (see Figure
\ref{fig:warning}). Most other analysis software will simply drop cases
with missings without warning.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{images/ggplot2-warning-and-plot}

}

\caption{How ggplot behaves when displaying missing values:  Warning message (Left) is thrown prior to plot (Right); ggplot2 does not display missing values in the plot, but at least throws a warning message to the user, which is better than many statistical models.}\label{fig:warning}
\end{figure}

The imputation literature does not address how to handle missing data.
Producing a complete dataset for analysis, whether by case or variable
deletion, or with imputation, requires knowing something about the
structure of the missing values. To understand their structure, one
needs to generate visualizations, calculate summaries, and perform
exploratory modelling. The imputation methods literature (e.g. Little
(1988), Rubin (1976), Simon and Simonoff (1986), Schafer and Graham
(2002), Buuren (2012)) focuses on ensuring valid statistical inference
is made from incomplete data. They approach this chiefly through
probabilistic modelling, and assume the mechanism of missing data is
known to the analyst.

Exploring missing data to find the missingness mechanism is difficult,
because it is an iterative process with many dead ends and often no
clear answer. There is no silver bullet for missing data that will
reveal it's mechanism to the analyst. Despite it being crucial to
explore and understand missing data, there is very little research on
how to make this process more efficient so that it integrates into
analysis pipelines. Although decision tree models or latent group
analysis can reveal structures or patterns of missingness (Barnett et
al. 2017; Tierney et al. 2015), these are not definitive and usually
require discussion with data curators to validate. The analyst must
explore missing data using visualizations, summaries, and exploratory
analyses, but there is a lack of tools to efficiently explore missing
data.

The graphics literature (Swayne and Buja 1998; Unwin et al. 1996)
provides several solutions for exploring missings visually. These
approaches incorporate missings into the plot in some way. For example,
imputing values 10\% below the range to display all values in a scatter
plot. These methods require the data to be linked to the missing or
imputed values; this link is typically hidden from the user. The
approaches from the graphics literature have not been translated and
integrated into tidy data. This limits the analysts capacity to
efficiently explore missing values. Extending tidy data and tidy tools
to account for missing values would create more efficient and robust
workflows and analysis.

The paper is organized in the following way to discuss the problem of
missing data. The next section (\ref{tidy-data-concepts}) is an
introduction to tidy data principles and tools. Missing value imputation
(\ref{missing-data-rep-dep} and \ref{imputation}) and exploration
literature (\ref{exploration}) are then discussed. Section
\ref{extensions} discusses extensions to the tidy methodology that
facilitates operations for exploring, visualizing, and imputing missing
data. The application of these new methods is illustrated using a case
study in section \ref{case-study}. Finally, section \ref{discussion}
discusses strengths, limitations, and future directions.

\hypertarget{background}{%
\section{Background}\label{background}}

\hypertarget{tidy-data-concepts}{%
\subsection{Tidy data concepts and methods}\label{tidy-data-concepts}}

Features of tidy data were formally discussed in 2014 (Wickham 2014),
subsequently generating much discussion (see Donoho (2017) and
commentary), and tools for data analysis. Tidy data is defined as:

\begin{quote}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Each variable forms a column.
\item
  Each observation forms a row.
\item
  Each type of observational unit forms a table.
\end{enumerate}
\end{quote}

Tidy data is easier to work with and analyze because the variables are
in the same format as they would be put into modelling software. This
helps the analyst works swiftly and clearly, closing up opportunities
for errors.

With tidy data come tidy tools, which have the same input and output:
tidy data. This consistency means multiple tools can be composed
together in sequence, allowing for rapid, elegant operations.
Contrasting tidy tools are messy tools. These have tidy input but messy
output. Messy tools slow down analysis by shifting the focus from
analysis to transforming output so it is the right shape for the next
step in the analysis. This makes the work at each step harder to
predict, and more complex and difficult to maintain. This disrupts
workflow, and incites errors.

Tidy tools fall into three broad categories: data manipulation,
visualization, and modelling. These are now discussed in turn.

\hypertarget{tidy-data-manip}{%
\subsubsection{Data manipulation}\label{tidy-data-manip}}

Data manipulation is made input and output tidy with R packages
\texttt{dplyr} and \texttt{tidyr} (Wickham et al. 2017a; Wickham and
Henry 2018). These provide the five ``verbs'' of data manipulation: data
reshaping, sorting, filtering, transforming, and aggregating.
\textbf{Data reshaping} goes from long to wide formats; \textbf{sorting}
arranges rows in a specific order; \textbf{filtering} removes rows based
on a condition; \textbf{transforming} changes existing variables or
adding new ones; \textbf{aggregating} creates a single value from many
values, say for example by computing the minimum, maximum, and mean.

\hypertarget{tidy-vis}{%
\subsubsection{Visualizations}\label{tidy-vis}}

Visualization tools only have tidy data as their input, as the output is
a graphic. The popular domain specific language \texttt{ggplot2} maps
variables in a dataset to features (referred to as aesthetics) of a
graphic (Wickham 2009). For example, a scatterplot can be created by
mapping two variables to the x and y axes, and specifying a point
geometry. The graph can then be changed to map a third variable to color
of the points, and a fourth variable to size.

\hypertarget{tidy-model}{%
\subsubsection{Modelling}\label{tidy-model}}

Modelling tools work well with tidy data, as they have a clear mapping
from variables in the data to the formula for a model. For example in R,
y regressed on x and z is: \texttt{lm(y\ \textasciitilde{}\ x\ +\ z)}.
Modelling tools are input tidy, but their output is always messy. For
example, estimated coefficients, predictions, and residuals from one
model cannot be easily combined with the output of another model. The
\texttt{recipes} package is a tidy tool under development to help make
modelling input and output tidy (Kuhn and Wickham 2018).

\hypertarget{the-tidyverse}{%
\subsubsection{The tidyverse}\label{the-tidyverse}}

Defining tidy data and tidy tools has resulted in the growth of a set of
packages known collectively as the ``tidyverse'' (Ross et al. 2017).
These are designed to share similar principles in their design and
behavior, and cover the breadth of an analysis - going from importing,
tidying, transforming, visualizing, modelling, to communicating (Ross et
al. 2017; Wickham 2017; Wickham and Grolemund 2016). This has led to a
growth of tools for specific parts of analysis - from reading in data
with \texttt{readr}, \texttt{readxl}, and \texttt{haven}, to handling
character strings with \texttt{stringr}, dates with \texttt{lubridate},
and performing functional programming with \texttt{purrr} (Grolemund and
Wickham 2011; Henry and Wickham 2017; Wickham 2018; Wickham and Bryan
2017; Wickham et al. 2017b; Wickham and Miller 2018). It has also led to
the growth of new packages for other fields that follow similar design
principles and create fluid workflows for new domains. For example, the
\texttt{tidytext} package for text analysis, the \texttt{tsibble}
package for time series data, and \texttt{tidycensus} for working with
US census and boundary data (Silge and Robinson 2016; Walker 2018; Wang
et al. 2018).

\hypertarget{tidy-formats-missing-data}{%
\subsubsection{Tidy formats for missing
data}\label{tidy-formats-missing-data}}

Current tools for missing data are messy. Missing data tools can be used
to perform imputations, missing data diagnostics, and data
visualizations. However, these tools are primarily \emph{messy}, and
suffer the same problems as modelling for imputation: They use tidy
input, but produce messy output. The complex, often multivariate nature
of imputation methods also makes makes them difficult to represent.
Visualization methods for missing data do not map data features to the
aesthetics of a graphic, as in \texttt{ggplot2}, limiting expressive
exploration.

Past graphics research on missing data has not yet been framed in a tidy
way. The graphics literature has explored methods for missing value
exploration and imputation. Translating and expanding these to fit
within tidy data and tidy tools would create more efficient workflows.

\hypertarget{missing-data-rep-dep}{%
\subsection{Missing data representation and
dependence}\label{missing-data-rep-dep}}

The information on missing values in a dataset can be represented as a
missingness matrix, \(R\), where 0 is observed data and 1 is missing
data, where for data \(y\) with \(i\) rows and \(j\) columns, the value
\(r_{ij}\) is given by:

\[
r_{ij} =\begin{cases}
1, & \text{if } y_{ij} \text{ is missing} \\
0, & \text{if } y_{ij} \text{ is observed}
\end{cases}
\]

There are many ways each value can go missing. The information in \(R\)
can be used to arrive at three categories of missing values: Missing
completely at random (MCAR), Missing at random, and missing not at
random. \textbf{MCAR} is where values being missing have no association
with observed or unobserved data, that is
\(Pr(missing | observed, unobserved) = Pr(missing)\). Essentially, the
probability of an observation being missing is unrelated to anything
else. Although this is a convenient scenario, it is not actually
possible to confirm, as it relies on statements on data unobserved.
\textbf{MAR} refers to cases where the missingness only depends on the
data observed, and not data unobserved, that
is,\(Pr(missing | observed, unobserved) = Pr(missing | observed)\). Some
structure or dependence between the missing values and observed values
is allowed, provided that this can be explained by the data observed.
Finally, \textbf{MNAR} refers to the missingness being related to values
unobserved: \(Pr(missing | observed, unobserved)\). This assumes that
conditioning on all available observed data, the data goes missing due
to some phenomena unobserved. This presents a challenge in analysis, as
it is difficult to verify this state, and also implies bias in analysis
due to the unobserved phenomena.

Visualizations can help assess whether data is MCAR, MAR or MNAR.
Imputation is recommended in most cases of missingness dependence, where
the values are plausible (e.g., imputing a pulse for someone deceased),
as this can help re-assess the classification of missingness.

\hypertarget{imputation}{%
\subsection{Imputation}\label{imputation}}

Imputation methods are more accessible than they have ever been, and the
number of methods and implementations in software continues to grow.
Values can be imputed with one value (single imputation), or multiple
values (multiple imputation), creating \(m\) datasets. Methods and
software for single and multiple imputation are now discussed.

\hypertarget{single-imputation}{%
\subsubsection{Single imputation}\label{single-imputation}}

\texttt{VIM} (Kowarik and Templ 2016) efficiently implements well-used
imputation methods k-nearest neighbor, regression, hot-deck, and
iterative robust model-based imputation. This diversity of approaches
allow the package to deal semi-continuous, continuous, count, and
categorical data. VIM identifies imputed cases by adding an indicator
variable with a suffix \texttt{\_imp}. So \texttt{Var1} would have a
sibling indicator column, \texttt{Var1\_imp}, with TRUE or FALSE values
to indicate imputed value. \texttt{VIM} also has a variety of
visualization methods, discussed in section \ref{exploration}. The
\texttt{simputation} package includes imputation methods for linear
regression, decision trees, K nearest neighbors, hotdeck imputation, and
the EM algorithm (Dempster et al. 1977; van der Loo 2017). A key feature
of its design is that it returns a regular dataframe. This approach
contrasts from other imputation software, which may provide custom
classes of data or additional objects, which require additional work to
prepare for subsequent analysis. Providing a dataframe with imputed
values included in the data reduces the friction of working with other
tools, but comes at the cost identifying imputed values. Other software
for imputation includes \texttt{Hmisc} (Harrell Jr et al. 2018), which
provide predictive mean matching, \texttt{imputeTS} (Moritz and
Bartz-Beielstein 2017), which provide time series imputation methods,
and \texttt{missMDA} (Josse and Husson 2016), which imputes data using
principal components analysis.

\hypertarget{multiple-imputation}{%
\subsubsection{Multiple imputation}\label{multiple-imputation}}

Multiple imputation is often regarded as best practice for imputing
values (Schafer and Graham 2002), as long as appropriate caution is
taken (Sterne et al. 2009). Popular and robust methods for multiple
imputation include the \texttt{mice}, \texttt{Amelia}, and \texttt{mi}
packages (Gelman and Hill 2011; Honaker et al. 2011; van Buuren and
Groothuis-Oudshoorn 2011). The \texttt{mice} package implements the
method of chained equations, using a variable-wise algorithm to
calculate the posterior distribution of parameters, from which uses to
generate imputed values. The suggested workflow in \texttt{mice}
revolves around imputing data, returning completed data, and fitting a
model and pooling the results. The \texttt{Amelia} package (Honaker et
al. 2011) assumes data are multivariate normal, and samples from the
posterior using the computationally efficient (and parallelizable)
Expectation-Maximization Bootstrap (EMB) algorithm (Honaker and King
2010) and allows for incorporation of information on the values in a
prior. The \texttt{mi} package (Gelman and Hill 2011) also uses Bayesian
models for imputation, providing better handling of semi-continuous
values, and data with structural or perfect correlation. A collection of
analysis models are also provided in \texttt{mi}, which work with the
specially multiply imputed data, including linear models, generalized
linear models, and their corresponding Bayesian components. This
approach promotes fluid workflow, with a similar penalty to tidying up
model output, which is still messy. Another software for performing
multiple imputation is \texttt{norm} (R by Alvaro A. Novo. Original by
Joseph L. Schafer \textless{}jls@stat.psu.edu\textgreater{}. 2013;
Schafer 1999), which uses methods from the NORM software from Schafer
(1999), such as multiple imputations using EM for multivariate normal
data. The \texttt{norm} package does not provide a framework for
tracking missing values, but instead provides tools for making inference
from multiple imputation. Each of these multiple imputation methods
provide practical methods, but do not have consistent interfaces across
their methods or implementation, which makes then inherently messy, as
their output is always different. A consistent data structure that
tracks missing values by default would make the data outputs tidy,
streamlining subsequent analysis.

\hypertarget{exploration}{%
\subsection{Exploration}\label{exploration}}

The primary focus of most missing data packages is exploring imputed
values, not on exploring relationships in missing values, and
identifying possible patterns. Texts that do cover exploration of
missing data have the same problem as with modelling: the input is tidy,
but the output is messy and does not work with other tools (Buuren
2012); this is inefficient. Methods for exploring missing values are
primarily covered in literature on interactive graphics (Cook and Swayne
2007; Swayne and Buja 1998; Unwin et al. 1996), and are picked up again
in a discussion of a graphical user interface (Cheng et al. 2015).

Missing values in a dataset can be represented as a binary matrix where
0 is present and 1 is missing (Rubin 1976; Schafer and Graham 2002).
This is often referred to as a matrix \(R\), which can be used to assess
missing data dependence. This idea of a matrix of missing values has
also been used in interactive graphics, dubbed a ``shadow matrix'', to
link missing and imputed values to the data, facilitating their display
in graphics (Swayne and Buja 1998). This work also focuses heavily on
multivariate numeric data. This is an idea upon which this new work
builds.

The approach of Unwin et al. (1996) focussed on multivariate categorical
data, where missingness is explicitly added as an additional category.
MANET also provided univariate visualizations of missing data using
linked brushing between a reference plot of the missingness for each
variable, and a plot of the data as a histogram or barplot. The approach
of Swayne and Buja (1998) in the software XGobi, further developed in
ggobi (Cook and Swayne 2007), focussed more on multivariate quantitative
data. Missingness is incorporated into plots in ggobi by setting them to
be 10\% below the minimum value.

MissingDataGUI provides a user interface for exploring missing data
structure both numerically and visually. Using a GUI to explore missing
data may make it easier to glean valuable insight into important
structures for missingness. However, these insights are typically not
captured or recorded, so it is difficult to incorporate them into
reproducible analyses. This distracts and breaks analysis workflow,
inviting mistakes.

VIM (visualizing and Imputing Missing Data) provides visualization
methods that identify observed, imputed, and missing values. These
include spinograms, spinoplots, missingness matrices, plotting
missingness in the margins of other plots, and other summaries. These
help explore missingness, however these visualizations do not map
variables to graphical aesthetics, creating friction moving across
workflows, making them difficult to extend to new circumstances.
Additionally, data used to create the visualizations cannot be accessed,
posing a barrier to further explorations.

\texttt{ggplot2} incorporates missingness into visualizations only when
mapping a discrete variable to a graph aesthetic. This has some
limitations. For example, for data with school grades and test scores, a
boxplot visualization could have school grade mapped to the x axis, and
test score to the y axis (See Figure \ref{fig:gg-box-na}). If there are
missings in a continuous variable like test score, \texttt{ggplot2}
omits the missing values and prints a warning message. However, if a
discrete variable like school year has missing values, an NA category is
created for school year, and the scores are placed into this.

Figure \ref{fig:gg-box-na}, shows four scenarios of missingness for test
scores across school year - the first contains no missingness, the
second contains missingness for school year, the third contains
missingness in scores, and the fourth contains missingness in both.

\begin{figure}

{\centering \includegraphics[width=0.7\linewidth]{tidy-missing-data-paper_files/figure-latex/gg-box-na-1}

}

\caption{ggplot2 provides different visualizations depending on what type of data has missing values. (A) Data is complete and graphic is presented; (B) Missings are only in category variable year - an additional 'NA' boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and school year, additional missing information is shown. The missingness category is only shown when there are missings in categorical variables such as year (plots (B) and (D)). In (C), no missingness information is given on the graphic, despite there being missings in score, and a warning message is displayed about the number of missing values omitted.}\label{fig:gg-box-na}
\end{figure}

\hypertarget{extensions}{%
\section{Extensions to tidyverse}\label{extensions}}

Applying tidyverse principles to the domain of missing data clarifies
and facilitates missing data exploration, visualization, and imputation.
This section discusses how these principles are applied, and proceeds in
three parts. Section \ref{data-structure} discusses \emph{nabular} data,
a tidy data structure for missing values that tracks missings throughout
the analysis and allows for special missing values. Section
\ref{vis-summaries} covers visualizations to understand missingness.
Section \ref{num-sum} discusses data summaries, and Section \ref{verbs}
covers the common ``verbs'' used to solve problems that arise when
working with missing data. The design and naming of these features is an
important component that helps the analyst compose them efficiently in
data pipelines, and is discussed throughout.

\hypertarget{data-structure}{%
\subsection{Data structure}\label{data-structure}}

To explore missing data there needs to be a good representation of
missing values that integrate into analysis pipelines. A common
representation of missing values is a binary matrix the same dimension
as the data, where 0 and 1 indicate not missing and missing. This is
typically referred to as a matrix \(R\), used to describe missingness
dependence patterns, (see for example Rubin (1976) and Schafer and
Graham (2002)). This matrix is used to explore missing values in the
interactive graphics library XGobi, where it was called a ``shadow
matrix'', defined as a copy of the original data with indicator values
of missingness: ``NA'' and ``!NA'' for missing and not missing,
represented as 0 and 1 (see Figure \ref{fig:shadow-matrix-progress}
parts 2 and 3 below).

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{images/full-conversion-to-shadow}

}

\caption{Progression of creating shadow matrix data. (1-2) Data to a binary representation of missings, (2-3) Binary format converted to a shadow matrix (3-4) New and improved shadow matrix with changed variable names. This clearly links a variable to its state of missingness.}\label{fig:shadow-matrix-progress}
\end{figure}

Three features are added to the shadow matrix to facilitate analysis:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Coordinated names: Variables in the shadow matrix gain the same name
  as in the data, with the suffix "\_NA".
\item
  Special missing values: Values in the shadow matrix can be ``special''
  missing values, indicated as \texttt{NA\_suffix}, where ``suffix'' is
  a very short message of the type of missings.
\item
  Cohesiveness: Binding the shadow matrix column-wise to the original
  data creates a cohesive \emph{nabular} data form, useful for
  visualization and summaries.
\end{enumerate}

These additions are now discussed in turn.

\hypertarget{coordinated-names}{%
\subsubsection{Coordinated names}\label{coordinated-names}}

A consistent short suffix ``\texttt{\_NA}'' for each variable keeps
names coordinated throughout analysis (See Figure
\ref{fig:shadow-matrix-progress} part 4). The suffix is short and easy
to remember in analysis or visualization. It also makes a clear
distinction that \texttt{var\_NA} is a random variable of the
missingness of a variable, \texttt{var}. This subtle change is important
as it changes the focus from the value of a variable to it's missingness
state, making intent clear when performing analysis.

\hypertarget{special-missings}{%
\subsubsection{Special missings}\label{special-missings}}

Values in the shadow matrix are ``factor'' type values, with text values
``NA'' and ``!NA'' and underlying number values 1 and 0, for missing and
not missing. Extending these to include ``special'' missing values
labelled as ``NA\_\textless{}suffix\textgreater{}'' allows for
representing additional features such as instrument failure and drop-out
as ``NA\_instr\_fail'', and ``NA\_drop\_out''. The underlying number
representation changes from binary to integer. For example, instrument
failure and drop out would be represented as 2 and 3, see Table
\ref{tab:shadow-encoding} for an example. Encoding these special missing
values is achieved by defining logical conditions and suffixes with the
\texttt{recode\_shadow} function in \texttt{naniar}, shown in Section
\ref{verbs-recode}.

\begin{table}[!h]

\caption{\label{tab:shadow-encoding}Example data of temperature, and its observed value, shadow representation, and underlying value. The temperature value of -99 is represented in the shadow column as 'NA\_instr', which in turn has the underlying numeric value of 2. This captures additional information about the data that is otherwise difficult to record.}
\centering
\begin{tabular}[t]{rlr}
\toprule
temp & temp\_NA & NA\_value\\
\midrule
-99 & NA\_instr & 2\\
NA & NA & 1\\
-1 & NA\_dropout & 3\\
106 & !NA & 0\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{nabular-data}{%
\subsubsection{Nabular data}\label{nabular-data}}

\emph{Nabular} data binds the shadow matrix column-wise to the original
data, and is so named as a portmanteau of \texttt{NA} and
\texttt{tabular}. \emph{Nabular} data keeps corresponding rows together,
removing the possibility of mismatching records, and explicitly links
missing values to the data. \emph{Nabular} data facilitates
visualization and summaries by allowing the user to reference the
missingness of a variable, \texttt{var}, as \texttt{var\_NA}.
\emph{Nabular} data is a snapshot of the missingness of the data. This
means when \emph{nabular} data is imputed, those imputed values can
easily be identified in analysis by referring to the appropriate,
coordinated names. \emph{Nabular} data is created using the command
\texttt{bind\_shadow()} or \texttt{nabular()} (see Figure
\ref{fig:nabularfig}). \emph{Nabular} data is not unlike classical data
formats that have quality or flag columns associated with each measured
variable, e.g.~Scripps CO2 data (Keeling et al. 2005), GHCN data (Durre
et al. 2008).

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{images/nabular}

}

\caption{The process of creating nabular data. Data transformed to shadow matrix, and nabular data contains the shadow matrix, column bound to the data. Nabular data can be created using `bind\_shadow` or `nabular()` functions. Nabular data provides a useful format for missing data exploration and analysis.}\label{fig:nabularfig}
\end{figure}

A data structure for missing data provides key benefits over on-the-fly
creating a logical matrix where TRUE means missing, and FALSE means
present. Firstly, it can be confusing to users whether TRUE or FALSE
means missing or present; \texttt{NA} and \texttt{!NA} are clearer.
Secondly, special missing values cannot be added fluently during
analysis with a logical matrix. Finally, the logical matrix cannot
capture which values are imputed, if imputation has already taken place.
Imputing values on \emph{nabular} data automatically tracks these value
changes.

Using additional columns to represent missingness information follows
recently published best practices for data organization, described in
Ellis and Leek (2017) and Broman and Woo (2017): (1) Keep one thing in a
cell and (2) Describe additional features of variables in a second
column. Here they suggest to indicate censored data with an extra
variable called ``VariableNameCensored'', which would be TRUE if
censored, otherwise FALSE. This information is now represented in the
shadow columns as additional integers, along with the other missing
information.

Other statistical programs represent missingness values as a full-stop,
\texttt{.}, and allow for recording special missing values as
\texttt{.a} through to \texttt{.z}. The special values from these
languages break the rule of ``keeping one thing in a cell'', as they
record both the value and the multivariate missingness state.

\hypertarget{vis-summaries}{%
\subsection{Visual summaries}\label{vis-summaries}}

Visualizing missingness in a dataset is essential to understand its
structure. This section discusses visualizations that are easy to
remember and compose with other functions in a pipeline. Four different
areas of visualization of missing data are discussed: overviews (Section
\ref{overview-plots}); univariate (Section \ref{univariate-plots}),
bivariate (Section \ref{bivariate-plots}), and multivariate (Section
\ref{multivariate-plots}). All plots are created using \texttt{ggplot2},
giving users clearer control over the plot appearance, and
customizability.

\hypertarget{overview-plots}{%
\subsubsection{Overview plots}\label{overview-plots}}

The number of missings in each variable and case are visualized using
\texttt{gg\_miss\_var} and \texttt{gg\_miss\_case} (see Figure
\ref{fig:gg-miss-case-var} below). These are shown using the
``airquality'' dataset, included in base R, which contains daily air
quality measurements in New York, from May to September, 1973. Rather
than highlighting the amount of complete data, these visualizations draw
attention to the amount of missings, ordering by missingness.

\begin{figure}

{\centering \includegraphics[width=0.9\linewidth]{tidy-missing-data-paper_files/figure-latex/gg-miss-case-var-1}

}

\caption{Graphical summaries of missingness in variables and cases for the airquality data. (A) Missings in each variable and (B) in each case. There are missing values in Ozone and Solar.R, with Ozone having more missings, and not many cases have two missing values, with most missingness being from cases with one missing value.}\label{fig:gg-miss-case-var}
\end{figure}

All missings in a dataset can be displayed using a heatmap style
visualization. This is achieved using \texttt{vis\_miss()} from the
\texttt{visdat} package (Tierney 2017), which also provides summaries of
missingness overall in the legend, and for each column (see Figure
\ref{fig:vis-miss}). The user can also apply clustering to the rows, and
arrange columns by missingness (Figure \ref{fig:vis-miss}B). Similar
visualizations are available in other missing data packages such as
\texttt{VIM}, \texttt{mi}, \texttt{Amelia}, and \texttt{MissingDataGUI}.
A key improvement is that \texttt{vis\_miss()} orients the visualization
analogous to a regular data structure: variables form columns in the
visualization and are named at the top, and each row is an observation.
Using \texttt{ggplot2} means the plot can be customized and combined
with other ggplot graphics.

\begin{figure}

{\centering \includegraphics[width=0.95\linewidth]{tidy-missing-data-paper_files/figure-latex/vis-miss-1}

}

\caption{Heatmap visualizations of missing data for the airquality dataset. (A) The default output and (B) ordered by clustering on rows and columns. There are only missings in ozone and solar radiation, and there appears to be some structure to their missingness.}\label{fig:vis-miss}
\end{figure}

The number of times certain variables go missing together can be
visualized using an ``upset plot'' (Conway et al. 2017), a visualization
technique for showing the size and features of sets in data that is
similar to a venn diagram, but scales better with more variables (Figure
\ref{fig:airquality-upset}).

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth]{tidy-missing-data-paper_files/figure-latex/airquality-upset-1}

}

\caption{The pattern of missingness in the airquality dataset shown in an upset plot. Only Ozone and Solar.R have missing values, and Ozone has the most missing values. There are 2 cases where both Solar.R and Ozone have missing values.}\label{fig:airquality-upset}
\end{figure}

Figure \ref{fig:airquality-upset} shows that only Ozone and Solar.R have
missing values. The dots in the bottom right show the combinations of
missingness, with the bottom left showing the missingness in each
variable, and the top showing the missingness for the combinations. The
bottom left shows that Ozone has the most missing values, and the top
shows 2 cases where both Solar.R and Ozone have missing values together.

\hypertarget{univariate-plots}{%
\subsubsection{Univariate plots}\label{univariate-plots}}

Missing values are by default not shown for univariate visualizations
such as histograms or densities. Two ways to use \emph{nabular} data to
present univariate data with missings are discussed. The first imputes
values below the range to facilitate visualizations. The second displays
two plots of the same variable according to the missingness of a chosen
variable.

\hypertarget{imp-below-range}{%
\paragraph{Imputing values below the range}\label{imp-below-range}}

To visualize the amount of missings in each variable, the data is
transformed into \emph{nabular} form, then values are imputed values
below the range of data using \texttt{impute\_below} (by default
imputing 10\% below the range). Visualizing this as a histogram shows
missing values on its left. Ozone is shown in a different color by
referring to the variable \texttt{Ozone} as \texttt{Ozone\_NA} in a fill
aesthetic in ggplot, shown in Figure \ref{fig:impute-shift-histogram}.

\begin{figure}

{\centering \includegraphics[width=0.75\linewidth]{tidy-missing-data-paper_files/figure-latex/impute-shift-histogram-1}

}

\caption{A histogram using nabular data to show the values and missings in ozone. Values are imputed below the range to show the number of missings in Ozone and colored according to missingness of ozone (`Ozone\_NA`). There are about 35 missings in Ozone.}\label{fig:impute-shift-histogram}
\end{figure}

\hypertarget{vis-split-by-miss}{%
\paragraph{Univariate split by missingness}\label{vis-split-by-miss}}

Missingness of one variable can also be used to display different
distributions in another. Figure \ref{fig:bind-shadow-density} shows the
values of temperature when ozone is present, and missing. Figure
\ref{fig:bind-shadow-density}A is a faceted histogram, and Figure
\ref{fig:bind-shadow-density}B is an overlaid density. This shows how
values of temperature are affected by the missingness of ozone, and
reveals a cluster of low temperature observations with missing ozone
values.

\begin{figure}

{\centering \includegraphics[width=0.49\linewidth]{tidy-missing-data-paper_files/figure-latex/bind-shadow-density-1} \includegraphics[width=0.49\linewidth]{tidy-missing-data-paper_files/figure-latex/bind-shadow-density-2}

}

\caption{A visualization of temperature according to missingness in ozone from in the airquality dataset. A histogram of temperature facetted by the missingness of ozone (A), or a density of temperature colored by missingness in ozone (B). The distribution shows a cluster of low temperature observations with missing ozone values, but the temperature values are otherwise similar.}\label{fig:bind-shadow-density}
\end{figure}

\hypertarget{bivariate-plots}{%
\subsubsection{Bivariate plots}\label{bivariate-plots}}

To visualize missing values in two dimensions the missing values can be
placed in the plot margins. This is achieved by imputing values below
the range of the data. As in Section \ref{imp-below-range}, using
\emph{nabular} data identifies imputed values, and color makes
missingness pre-attentive (Treisman 1985). The steps of imputing and
coloring have been been combined into \texttt{geom\_miss\_point()} from
the \texttt{naniar} package (Figure \ref{fig:geom-miss}), where there is
a mostly uniform spread of missing values for Solar.R and Ozone.

\begin{figure}

{\centering \includegraphics{tidy-missing-data-paper_files/figure-latex/geom-miss-1}

}

\caption{Scatterplots with missings displayed at 10 percent below for the airquality dataset. Scatterplots of ozone and solar radiation (A), and ozone and temperature (B). There are missings in ozone and solar radiation, but not temperature.}\label{fig:geom-miss}
\end{figure}

As \texttt{geom\_miss\_point} is a defined geometry for
\texttt{ggplot2}, it provides features such as faceting and mapping
other variables to aesthetics such as color, shape, or size. The shadow
format by itself can also be used to visually explore some patterns of
missingness.

\hypertarget{multivariate-plots}{%
\subsubsection{Multivariate plots}\label{multivariate-plots}}

Parallel coordinate plots can visualize missingness beyond two
dimensions. They transform variables to the same scale, and typically
make the data values range between 0 and 1. To showcase this
visualization, the dataset \texttt{oceanbuoys} from \texttt{naniar} is
used, which contains measurements of moored ocean buoys to understand
and predict El Niño and El Niña. The data were collected in 1993 and
1997, and contains information on the sea and air temperature, humidity,
and east west and north south wind directions.

Figure \ref{fig:parallel-cord-plot} shows a parallel coordinate plot of
the \texttt{oceanbuoys} data, with missing values imputed to be 10\%
below the range, and values colored according to whether humidity
contained missing values. Although a 10\% below imputation is not the
most ideal way to display missing values in a parallel coordinate plot,
it shows that humidity is missing at low air and sea temperatures, and
that humidity is missing in one year, and one location.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/parallel-cord-plot-1}

}

\caption{Parallel coordinate plot shows missing values imputed 10\% below range for the oceanbuoys dataset. Values are colored by missingness of humidity. Humidity is missing for low air and sea temperatures, and is missing for one year and one location. }\label{fig:parallel-cord-plot}
\end{figure}

\hypertarget{num-sum}{%
\subsection{Numerical summaries}\label{num-sum}}

There are different ways to summarize the amount of missing and complete
values in a dataset. This section describes approaches to summarizing
missingness, implemented in the \texttt{naniar} package. The functions
are easy to remember as they follow consistent naming, and their output
is consistent, and returns a single number or dataframe. This means they
integrate well with plotting and modelling tools. Section
\ref{single-num-sum} discusses single number summaries for all data and
for cases and variables ; Section \ref{sum-tab-missings} discusses
variable- and case-wise summaries; and Section \ref{num-sum-w-group}
shows how the design of these tools works with other tools in an
analysis pipeline.

\hypertarget{single-num-sum}{%
\subsubsection{Single number summaries}\label{single-num-sum}}

The overall number of missing values in a dataset is shown with
\texttt{n\_miss}, and the proportion or percent missings,
\texttt{prop\_miss} and \texttt{pct\_miss}. Complementing these are
functions to reveal the number of complete values: \texttt{n\_complete},
\texttt{prop\_complete}, and \texttt{pct\_complete}. These be extended
to summarize the number/amount of variables and cases that contain a
missing values by appending \texttt{\_case} or \texttt{\_var} to any of
these summaries. Table \ref{tab:n-prop-pct-miss-complete} shows the
names and output of these functions on the dataset airquality.

\begin{table}[!h]

\caption{\label{tab:n-prop-pct-miss-complete}Single number summaries of missing and complete data, applied to the airquality dataset. These functions follow consistent naming, to make them easy to remember, and to make their use clear.}
\centering
\begin{tabular}[t]{lrlr}
\toprule
Missing Function & missing value & Complete function & complete value\\
\midrule
n\_miss & 44.00 & n\_complete & 874.00\\
prop\_miss & 0.05 & prop\_complete & 0.95\\
pct\_miss & 4.79 & pct\_complete & 95.21\\
pct\_miss\_var & 27.45 & prop\_complete\_var & 0.73\\
pct\_miss\_case & 33.33 & pct\_complete\_case & 66.67\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{sum-tab-missings}{%
\subsubsection{Summaries and tabulations of missing
data}\label{sum-tab-missings}}

Presenting the number and percent of missing values for each variable,
or case, provides a summary that can be used by models to inform
imputations or data handling. For example, potentially dropping
variables or deciding to include others in an imputation model. Another
useful approach is to tabulate the frequency of missing values for each
variable or case; that is, the number of times there are zero, one, two,
and so on, missing values. These summaries and tabulations are shown for
variable in Tables \ref{tab:miss-var-summary} and
\ref{tab:miss-var-table}, using the \texttt{miss\_var\_summary} and
\texttt{miss\_var\_table} functions. Functions are also provided for
case-wise summaries and tabulations with \texttt{miss\_case\_summary}
and \texttt{miss\_case\_table}, which order by \texttt{n\_miss}, so the
most missings are always shown at the top. Other summaries provided in
\texttt{naniar} include summaries of missingness across a provided
repeating span, \texttt{miss\_var\_span}, and finding the streaks or
runs of missingness in a given variable in \texttt{miss\_var\_run}.

\begin{table}[!h]

\caption{\label{tab:miss-var-summary}Summary of the number and percent of missings in each variable. Only ozone and solar radiation have missing values.}
\centering
\begin{tabular}[t]{lrr}
\toprule
variable & n\_miss & pct\_miss\\
\midrule
Ozone & 37 & 24.2\\
Solar.R & 7 & 4.6\\
Wind & 0 & 0.0\\
Temp & 0 & 0.0\\
Month & 0 & 0.0\\
Day & 0 & 0.0\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[!h]

\caption{\label{tab:miss-var-table}Tabulation of the amount of missing data in each variable. This shows the number of variables with no missings, 7 missings, and 37 missings, and the percentage of variables with those amounts of missingness. There are not many patterns of missingness.}
\centering
\begin{tabular}[t]{rrr}
\toprule
n\_miss\_in\_var & n\_vars & pct\_vars\\
\midrule
0 & 4 & 66.7\\
7 & 1 & 16.7\\
37 & 1 & 16.7\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{num-sum-w-group}{%
\subsubsection{Combining numerical summaries with grouping
operations}\label{num-sum-w-group}}

These summaries and tabulations can be composed with the ``pipe''
operator to perform grouped summaries using the \texttt{group\_by}
operator from \texttt{dplyr}. These summaries return dataframes, which
may be used by other plotting devices, modelling tools, or other
summaries. Table \ref{tab:group-miss-var-summary} shows an example of
the missing data summaries for the airquality dataset in each variable,
for each month of the year. These datasets can be summarized in
visualizations, or values input into a model.

\begin{table}[!h]

\caption{\label{tab:group-miss-var-summary}A summary of the missingness in each variable, for each Month of the airquality dataset, with only the first 10 rows of output shown. There are more ozone missings in June than May.}
\centering
\begin{tabular}[t]{rlrr}
\toprule
Month & variable & n\_miss & pct\_miss\\
\midrule
5 & Ozone & 5 & 16.1\\
5 & Solar.R & 4 & 12.9\\
5 & Wind & 0 & 0.0\\
5 & Temp & 0 & 0.0\\
5 & Day & 0 & 0.0\\
\addlinespace
6 & Ozone & 21 & 70.0\\
6 & Solar.R & 0 & 0.0\\
6 & Wind & 0 & 0.0\\
6 & Temp & 0 & 0.0\\
6 & Day & 0 & 0.0\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{verbs}{%
\subsection{Verbs}\label{verbs}}

Common operations on missing data can be considered to be verbs: scan,
replace, add, shadow, impute, track, and flag. Missing values can be
\textbf{scanned} to find possible missings not coded as \texttt{NA}.
These values can then be \textbf{replaced} with an \texttt{NA} value. To
facilitate exploration, summaries of missingness can be \textbf{added}
as a summary column to the original data. Data values can be augmented
with \textbf{shadow} values, which helps explore missing data, as well
as facilitating the process of \textbf{imputing}, and \textbf{tracking}.
Finally, unusual or specially coded missing values can be
\textbf{flagged}. These verbs are described in this section.

\hypertarget{verbs-search}{%
\subsubsection{scan: Searching for common missing value
labels}\label{verbs-search}}

There are many common values that mean ``missing''. For example, ``N/A''
``Not Available'', ``MISSING'', or white space " ". Before replacing
these values with missing values (\texttt{NA}), it is useful to first
\emph{scan} the data to get a sense of the magnitude of the problem. The
\texttt{miss\_scan\_count()} function finds where specified values occur
in data. For example, for the example data, \texttt{dat\_ms}:

\begin{table}[!h]

\caption{\label{tab:setup-miss-scan-count}An example dataset with variables x, y and z containing some common unusual missing values.}
\centering
\begin{tabular}[t]{rlr}
\toprule
x & y & z\\
\midrule
1 & A & -100\\
3 & N/A & -99\\
NA & NA & -98\\
-99 & E & -101\\
-98 & F & -1\\
\bottomrule
\end{tabular}
\end{table}

All occurrences of -99 can be searched for using the code:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{miss_scan_count}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ dat_ms, }\DataTypeTok{search =} \DecValTok{-99}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

This returns a table of the occurrences of that value in each variable
(Table \ref{tab:miss-scan-count}). A list of common NA values is
provided in \texttt{naniar} in the \texttt{common\_na\_strings} and
\texttt{common\_na\_numbers} for characters and numbers. These contain
values like ``n/a'', ``na'', and ``.'', and -9, -99.

\begin{table}[!h]

\caption{\label{tab:miss-scan-count}Table of the occurences of the search '-99' in the data, 'dat\_ms'. There is one occurence if -99 in variables x and z.}
\centering
\begin{tabular}[t]{lr}
\toprule
Variable & n\\
\midrule
x & 1\\
y & 0\\
z & 1\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{verbs-replace-with}{%
\subsubsection{replace: Replacing missing value label with
another}\label{verbs-replace-with}}

Once the magnitude of the unusual missing values has been assessed,
these values need to be replaced with appropriate missing values, using
\texttt{replace\_with\_na}. This is useful when there is certainty of
which values mean missing. For example ``N/A'', ``N A'', ``Not
Available'', and ``missing''. Using the previous toy dataset,
\texttt{dat\_ms}, the value -99, or both -99 and 98 can be replaced with
NA as follows:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{replace_with_na}\NormalTok{(dat_ms, }\DataTypeTok{replace =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{x =} \DecValTok{-99}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # A tibble: 5 x 3
##       x y         z
##   <dbl> <chr> <dbl>
## 1     1 A      -100
## 2     3 N/A     -99
## 3    NA <NA>    -98
## 4    NA E      -101
## 5   -98 F        -1
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{replace_with_na}\NormalTok{(dat_ms, }\DataTypeTok{replace =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{x =} \KeywordTok{c}\NormalTok{(}\OperatorTok{-}\DecValTok{99}\NormalTok{, }\DecValTok{-98}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # A tibble: 5 x 3
##       x y         z
##   <dbl> <chr> <dbl>
## 1     1 A      -100
## 2     3 N/A     -99
## 3    NA <NA>    -98
## 4    NA E      -101
## 5    NA F        -1
\end{verbatim}

For flexibility, there are scoped variants for
\texttt{replace\_with\_na}: \texttt{\_all}, \texttt{\_if}, and
\texttt{\_at}. This means \texttt{replace\_with\_na} works for
\textbf{all} variables with \texttt{replace\_with\_na\_all}, \textbf{at}
selected variables with \texttt{replace\_with\_na\_at}, and \textbf{if}
variables that meet some condition with \texttt{replace\_with\_na\_if}.
The syntax for these is slightly different to
\texttt{replace\_with\_na}, but powerful:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{replace_with_na_at}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ dat_ms,}
                   \DataTypeTok{.vars =} \StringTok{"x"}\NormalTok{,}
                   \DataTypeTok{condition =} \OperatorTok{~}\NormalTok{.x }\OperatorTok{==}\StringTok{ }\DecValTok{-99}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # A tibble: 5 x 3
##       x y         z
##   <dbl> <chr> <dbl>
## 1     1 A      -100
## 2     3 N/A     -99
## 3    NA <NA>    -98
## 4    NA E      -101
## 5   -98 F        -1
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{replace_with_na_if}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ dat_ms,}
                 \DataTypeTok{.predicate =}\NormalTok{ is.character,}
                 \DataTypeTok{condition =} \OperatorTok{~}\NormalTok{.x }\OperatorTok{==}\StringTok{ "N/A"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # A tibble: 5 x 3
##       x y         z
##   <dbl> <chr> <dbl>
## 1     1 A      -100
## 2     3 <NA>    -99
## 3    NA <NA>    -98
## 4   -99 E      -101
## 5   -98 F        -1
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{replace_with_na_all}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ dat_ms,}
                    \DataTypeTok{condition =} \OperatorTok{~}\NormalTok{.x }\OperatorTok{==}\StringTok{ }\DecValTok{-99}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # A tibble: 5 x 3
##       x y         z
##   <dbl> <chr> <dbl>
## 1     1 A      -100
## 2     3 N/A      NA
## 3    NA <NA>    -98
## 4    NA E      -101
## 5   -98 F        -1
\end{verbatim}

\hypertarget{verbs-add-cols}{%
\subsubsection{add: Adding missingness summary
variables}\label{verbs-add-cols}}

Summary information such as the number of missings in a given case can
be useful in understanding missingness structure, and even more so when
this information is kept alongside the data. Functions starting with
\texttt{add\_} exist in \texttt{dplyr} to add count information for
particular groups or conditions to the data. Similarly, \texttt{naniar}
includes \texttt{add\_} functions to add missingness summary information
back to the data, such as the number or proportion of missingness, the
missingness cluster, or if there are any missings, using:
\texttt{add\_n\_miss()} \texttt{add\_prop\_miss()},
\texttt{add\_miss\_cluster}, and \texttt{add\_any\_miss()},
respectively.

An example use of these features is in Tierney et al. (2015), where the
proportion, number, or cluster of missings is used as the outcome in a
model. The variables in the dataset can then be used to predict the
outcome, identifying variables and values important in predicting
missingness structures. There are also functions for adding information
about shadow values or nice labels for if there are any missing values
with \texttt{add\_label\_shadow()} and \texttt{add\_label\_missings()}:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{add_n_miss}\NormalTok{(dat_ms)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # A tibble: 5 x 4
##       x y         z n_miss_all
##   <dbl> <chr> <dbl>      <int>
## 1     1 A      -100          0
## 2     3 N/A     -99          0
## 3    NA <NA>    -98          2
## 4   -99 E      -101          0
## 5   -98 F        -1          0
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{add_prop_miss}\NormalTok{(dat_ms)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # A tibble: 5 x 4
##       x y         z prop_miss_all
##   <dbl> <chr> <dbl>         <dbl>
## 1     1 A      -100         0    
## 2     3 N/A     -99         0    
## 3    NA <NA>    -98         0.667
## 4   -99 E      -101         0    
## 5   -98 F        -1         0
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{add_miss_cluster}\NormalTok{(dat_ms)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # A tibble: 5 x 4
##       x y         z miss_cluster
##   <dbl> <chr> <dbl>        <int>
## 1     1 A      -100            1
## 2     3 N/A     -99            1
## 3    NA <NA>    -98            2
## 4   -99 E      -101            1
## 5   -98 F        -1            1
\end{verbatim}

\hypertarget{verbs-nabular}{%
\subsubsection{\texorpdfstring{shadow: Creating \texttt{nabular}
data}{shadow: Creating nabular data}}\label{verbs-nabular}}

\emph{Nabular} data has the shadow matrix column-bound to the existing
data. This facilitates visualization and summaries, and allows for
imputed values to be appropriately tracked. \emph{Nabular} data can be
created with \texttt{bind\_shadow()} or \texttt{nabular()}. The
\texttt{bind\_shadow()} function is so named to borrow from existing
functions named \texttt{bind} - such as \texttt{bind\_rows()} and
\texttt{bind\_cols()} from \texttt{dplyr}, which bind rows or columns
into a dataframe. An example use of \texttt{bind\_shadow()} on the
example data, \texttt{dat\_ms} is shown below:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{bind_shadow}\NormalTok{(dat_ms)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # A tibble: 5 x 6
##       x y         z x_NA  y_NA  z_NA 
##   <dbl> <chr> <dbl> <fct> <fct> <fct>
## 1     1 A      -100 !NA   !NA   !NA  
## 2     3 N/A     -99 !NA   !NA   !NA  
## 3    NA <NA>    -98 NA    NA    !NA  
## 4   -99 E      -101 !NA   !NA   !NA  
## 5   -98 F        -1 !NA   !NA   !NA
\end{verbatim}

\hypertarget{verbs-recode}{%
\subsubsection{flag: Describing different types of missing
values}\label{verbs-recode}}

Unusual or spurious data values are often identified and
\texttt{flagged}. For example, there might be special codes to mark an
individual dropping out of a study, known instrument failure in weather
instruments, or for values censored in analysis. These special types of
missingness can be encoded in the shadow matrix of \emph{nabular} data,
using the \texttt{recode\_shadow()} function. This provides a fluent
interface to recode, or flag, the shadow information in the shadow
matrix as a special type of missing value. Using
\texttt{recode\_shadow()} requires specifying the variable you want to
contain the flagged value, the condition you want this to occur at, and
a suffix for the new type of missing value. This is then recoded as a
new factor level in the shadow matrix, so that every column is aware of
all possible new values of missingness. For example, the values -99 and
-98 could be recoded to mean a broken machine sensor for the variable x:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dat_ms }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{bind_shadow}\NormalTok{() }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{recode_shadow}\NormalTok{(}\DataTypeTok{x =} \KeywordTok{.where}\NormalTok{(x }\OperatorTok{==}\StringTok{ }\DecValTok{-99} \OperatorTok{~}\StringTok{ "broken_sensor"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # A tibble: 5 x 6
##       x y         z x_NA             y_NA  z_NA 
##   <dbl> <chr> <dbl> <fct>            <fct> <fct>
## 1     1 A      -100 !NA              !NA   !NA  
## 2     3 N/A     -99 !NA              !NA   !NA  
## 3    NA <NA>    -98 NA               NA    !NA  
## 4   -99 E      -101 NA_broken_sensor !NA   !NA  
## 5   -98 F        -1 !NA              !NA   !NA
\end{verbatim}

\hypertarget{verbs-impute}{%
\subsubsection{impute: Imputing values}\label{verbs-impute}}

\texttt{naniar} does not reinvent the wheel for imputation, as there are
many robust R packages for imputation, such as \texttt{mice},
\texttt{mi}, \texttt{VIM}, and \texttt{simputation}. However,
\texttt{naniar} does provide a few imputation methods to facilitate
exploration and visualizations, which were not otherwise available:
\texttt{impute\_below} and \texttt{impute\_mean} and
\texttt{impute\_median}. These imputation methods are useful to explore
structure in missingness, but are not recommended for use in analysis.

The \texttt{impute\_below} function imputes values below the minimum of
the data, with some jitter to reduce overplotting, but has options to
change the default amount it is shifted below, and the amount of jitter.

\begin{verbatim}
##       Ozone Solar.R
## 1  41.00000     190
## 2  36.00000     118
## 3  12.00000     149
## 4  18.00000     313
## 5 -19.72321      NA
## 6  28.00000      NA
\end{verbatim}

Similar to \texttt{simputation}, each \texttt{impute\_} function returns
the data with values imputed. However, \texttt{naniar} does not use a
formula syntax, and instead each function implements ``scoped variants''
\texttt{\_all}, \texttt{\_at} and \texttt{\_if}. Here, \texttt{\_all}
operates on \textbf{all} columns, \texttt{\_at} operates \textbf{at}
specific columns, and \texttt{\_if} operates on columns \textbf{if} they
meet some condition (such as \texttt{is.numeric} or
\texttt{is.character}). The \texttt{impute\_} functions with no scoped
variant, e.g., \texttt{impute\_mean}, will work on a single vector, but
not a data.frame. An example usage is shown below:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{impute_mean}\NormalTok{(airquality}\OperatorTok{$}\NormalTok{Ozone) }\OperatorTok{%>%}\StringTok{ }\KeywordTok{head}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 41.00000 36.00000 12.00000 18.00000 42.12931 28.00000
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{impute_mean_at}\NormalTok{(airquality, }\DataTypeTok{.vars =} \KeywordTok{vars}\NormalTok{(Ozone)) }\OperatorTok{%>%}\StringTok{ }\KeywordTok{head}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##      Ozone Solar.R Wind Temp Month Day
## 1 41.00000     190  7.4   67     5   1
## 2 36.00000     118  8.0   72     5   2
## 3 12.00000     149 12.6   74     5   3
## 4 18.00000     313 11.5   62     5   4
## 5 42.12931      NA 14.3   56     5   5
## 6 28.00000      NA 14.9   66     5   6
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{impute_mean_if}\NormalTok{(airquality, }\DataTypeTok{.predicate =}\NormalTok{ is.integer) }\OperatorTok{%>%}\StringTok{ }\KeywordTok{head}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##      Ozone  Solar.R Wind Temp Month Day
## 1 41.00000 190.0000  7.4   67     5   1
## 2 36.00000 118.0000  8.0   72     5   2
## 3 12.00000 149.0000 12.6   74     5   3
## 4 18.00000 313.0000 11.5   62     5   4
## 5 42.12931 185.9315 14.3   56     5   5
## 6 28.00000 185.9315 14.9   66     5   6
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{impute_mean_all}\NormalTok{(airquality) }\OperatorTok{%>%}\StringTok{ }\KeywordTok{head}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##      Ozone  Solar.R Wind Temp Month Day
## 1 41.00000 190.0000  7.4   67     5   1
## 2 36.00000 118.0000  8.0   72     5   2
## 3 12.00000 149.0000 12.6   74     5   3
## 4 18.00000 313.0000 11.5   62     5   4
## 5 42.12931 185.9315 14.3   56     5   5
## 6 28.00000 185.9315 14.9   66     5   6
\end{verbatim}

One drawback to imputation functions in this form is that the location
of imputed values is not tracked. This is covered in Section
\ref{verbs-track}.

\hypertarget{verbs-track}{%
\subsubsection{track: Shadow and impute missing
values}\label{verbs-track}}

To evaluate imputations they need to be tracked, which is achieved by
combining the verbs \texttt{bind\_shadow}, \texttt{impute\_}, and
\texttt{add\_label\_shadow}. The missing values can then be referred to
by their shadow variable, \texttt{\_NA}. The missingness of any
observation can be referred to with \texttt{any\_missing} (Figure
\ref{fig:track-impute-example}). The code chunk below shows the track
pattern, first using \texttt{bind\_shadow}, then imputing with
\texttt{impute\_knn}, and adding a \texttt{label\_shadow}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{aq_imputed <-}\StringTok{ }\NormalTok{airquality }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{bind_shadow}\NormalTok{() }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{as.data.frame}\NormalTok{() }\OperatorTok{%>%}
\StringTok{  }\NormalTok{simputation}\OperatorTok{::}\KeywordTok{impute_knn}\NormalTok{(Ozone }\OperatorTok{~}\StringTok{ }\NormalTok{.) }\OperatorTok{%>%}
\StringTok{  }\NormalTok{simputation}\OperatorTok{::}\KeywordTok{impute_knn}\NormalTok{(Solar.R }\OperatorTok{~}\StringTok{ }\NormalTok{.) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{add_label_shadow}\NormalTok{() }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{as_tibble}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

Missing values can then be shown in a scatterplot by setting the
\texttt{color} aesthetic in ggplot to \texttt{any\_missing} (Figure
\ref{fig:track-impute-example}A), or in a density plot looking at one
variable, using the \texttt{fill\ =\ any\_missing}, (Figures
\ref{fig:track-impute-example}B) and \ref{fig:track-impute-example}C).

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{(aq_imputed,}
       \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Ozone,}
           \DataTypeTok{y =}\NormalTok{ Solar.R,}
           \DataTypeTok{color =}\NormalTok{ any_missing)) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{geom_point}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_color_brewer}\NormalTok{(}\DataTypeTok{palette =} \StringTok{"Dark2"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{legend.position =} \StringTok{"bottom"}\NormalTok{)}

\KeywordTok{ggplot}\NormalTok{(aq_imputed,}
       \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Ozone,}
           \DataTypeTok{fill =}\NormalTok{ any_missing)) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{geom_density}\NormalTok{(}\DataTypeTok{alpha =} \FloatTok{0.3}\NormalTok{) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{scale_fill_brewer}\NormalTok{(}\DataTypeTok{palette =} \StringTok{"Dark2"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{theme}\NormalTok{(}\DataTypeTok{legend.position =} \StringTok{"none"}\NormalTok{)}

\KeywordTok{ggplot}\NormalTok{(aq_imputed,}
       \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ Solar.R,}
           \DataTypeTok{fill =}\NormalTok{ any_missing)) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{geom_density}\NormalTok{(}\DataTypeTok{alpha =} \FloatTok{0.3}\NormalTok{) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{scale_fill_brewer}\NormalTok{(}\DataTypeTok{palette =} \StringTok{"Dark2"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/track-impute-example-1} \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/track-impute-example-2} \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/track-impute-example-3}

}

\caption{Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values imputed using simputations `impute\_knn` function, with imputed values colored green and data values orange. Imputed values are similar, but slightly different to existing data.}\label{fig:track-impute-example}
\end{figure}

Imputed values can also be compared to complete case data by grouping by
\texttt{any\_missing}, and then summarizing (Table
\ref{tab:impute-summary}).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{aq_imputed }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{group_by}\NormalTok{(any_missing) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{summarise_at}\NormalTok{(}\DataTypeTok{.vars =} \KeywordTok{vars}\NormalTok{(Ozone),}
               \DataTypeTok{.funs =} \KeywordTok{funs}\NormalTok{(min, mean, median, max))}
\end{Highlighting}
\end{Shaded}

\begin{table}[!h]

\caption{\label{tab:impute-summary}Summary statistics of the previously imputed and complete cases. The mean and median values are similar, but the minimum and maximum values are different. The comparison of imputed values is similar to other dplyr summary workflows.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
any\_missing & min & mean & median & max\\
\midrule
Missing & 7 & 43.6 & 30 & 135\\
Not Missing & 1 & 42.1 & 31 & 168\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{case-study}{%
\section{Application}\label{case-study}}

To illustrate the methods, data on housing for the city of Melbourne
from January 28, 2016 to March 17, 2018 is used. The data was compiled
by scraping weekly property clearance data (Pino 2018). There are 27,247
properties, and 21 variables in the dataset. The variables include the
type of real estate (town house, unit, house), suburb, method of
selling, number of rooms, price, real estate agent, date of sale, and
the distance from the Central Business District (CBD).

The goal in analyzing this data is to accurately predict Melbourne
housing prices. The data contains many missing values. As a precursor to
building a predictive model, this analysis focusses on understanding the
patterns of missingness. This section shows how the methods from
previous sections are used together in a data analysis workflow.

\hypertarget{case-study-explore-pattern}{%
\subsection{Exploring patterns of
missingness}\label{case-study-explore-pattern}}

Figure \ref{fig:housing-miss-case-var}A shows 9 variables with missing
values. The most missings are in building area, followed by year built,
and land size, with similar amounts of missingness in Car, bathroom,
bedroom2, longitude, and latitude. Figure
\ref{fig:housing-miss-case-var}B reveals there are up to 50\% missing
values in cases, and that the majority of cases have more than 5\%
values missing. The variables building area and year built have more
than 50\% missing data, and so could perhaps be omitted from subsequent
analysis, as imputed values are likely to be spurious.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/housing-miss-case-var-1}

}

\caption{The amount of missings in each variable (A) and in each case (B) for Melbourne housing data. (A) Build area and year built have more than 50\% missing values, and car, bathroom, bedroom2 and longitude and latitude have about 25\% missings. (B) There are between 5 and 50\% missing values in cases. There are many missing values in the data, with the majority of missingness being in selected cases and variables.}\label{fig:housing-miss-case-var}
\end{figure}

Possible missingness structures are revealed by visualizing missingness
in the whole dataset, clustering and arranging the rows and columns of
the data (Figure \ref{fig:applic-vis-miss}). There are three main
clusters of missing data: at the top there are many variables missing
together; in the middle, building price and year built missing together,
at the bottom, building area, year built, and land size are missing.

\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{tidy-missing-data-paper_files/figure-latex/applic-vis-miss-1}

}

\caption{Heatmap of clustered missingness for the housing data. Three groups of missingness are apparent, at the top for building area to longitude, the middle for building area and year built, at the end for building area, year built, and landsize. There is some structure in the missings.}\label{fig:applic-vis-miss}
\end{figure}

Missingness patterns can also be shown with an \texttt{upset} plot
(Conway et al. 2017), to display the 8 of intersecting sets of missing
variables (see Figure \ref{fig:applic-vis-miss}. Two patterns stand out:
two, and five variables missing. This provides further evidence of the
patterns of missingness seen in Figures \ref{fig:housing-miss-case-var}
and \ref{fig:housing-upset}.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/housing-upset-1}

}

\caption{An upset plot of eight sets of missingness in the housing data. Two missingness patterns are clear, year built and building area, and lattitude through to building area.}\label{fig:housing-upset}
\end{figure}

Tabulating the number of missings in variables (see Table
\ref{tab:housing-miss-var-case-table} (left)) shows three groupings:
6254 - 6824 missings cases in variables, 9265 in another variable, and
between 15163 and 16736 missing. Tabulating missings in cases (Table
\ref{tab:housing-miss-var-case-table} (right)) shows 32\% of cases have
complete data, 26\% have 2 missing, 22\% have 8 missing, 12\% have 1
missing, and 5\% have 3 missing, the remaining cases with missings are
less than 1\%. Two variables with more than 50\% missingness are omitted
from analysis: building area and year built.

\begin{table}[!h]
\caption{\label{tab:housing-miss-var-case-table}Tabulating missingness for variables (left) and cases (right) to understand missingness patterns. 14 variables have between 0 and 3 missings, then 6 variables have 6000 - 9000 missings, and 2 variables have 15 - 16,000 missing values. About 30\% of cases have no missing values, then 45\% of cases have between 1 and 6 missing values, and then about 23\% of cases have 8 or more missings. There are different patterns of missingness in variables and cases, but they can be broken down into smaller groups.}

\centering
\begin{tabular}[t]{rrr}
\toprule
n\_miss\_in\_var & n\_vars & pct\_vars\\
\midrule
0 & 10 & 47.6\\
1 & 2 & 9.5\\
3 & 1 & 4.8\\
6254 & 2 & 9.5\\
6441 & 1 & 4.8\\
\addlinespace
6447 & 1 & 4.8\\
6824 & 1 & 4.8\\
9265 & 1 & 4.8\\
15163 & 1 & 4.8\\
16736 & 1 & 4.8\\
\bottomrule
\end{tabular}
\centering
\begin{tabular}[t]{rrr}
\toprule
n\_miss\_in\_case & n\_cases & pct\_cases\\
\midrule
0 & 8755 & 32.1\\
1 & 3356 & 12.3\\
2 & 7244 & 26.6\\
3 & 1370 & 5.0\\
4 & 79 & 0.3\\
\addlinespace
5 & 8 & 0.0\\
6 & 203 & 0.7\\
8 & 6229 & 22.9\\
9 & 2 & 0.0\\
11 & 1 & 0.0\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{case-study-explore-for-imp}{%
\subsection{Exploring missingness patterns for
imputation}\label{case-study-explore-for-imp}}

Using the information from the overview visualizations and tables, the
following variables are explored for features predicting missingness:
Land size, latitude, longitude, bedroom2, bathroom, car, and land size.

Missingness structure can be better understood and evaluated by using
the data to predict where the data goes missing. This can be achieved by
clustering the missing values into groups, and then applying
classification and regression trees (CART) to find those variables and
values that predict missingness clusters (Barnett et al. 2017; Tierney
et al. 2015). Two clusters are identified, one for each missingness
cluster. The missingness clusters are predicted using all variables in
the dataset with the CART package \texttt{rpart} (Therneau and Atkinson
2018), and plotted using the \texttt{rpart.plot} package (Milborrow
2018). Importance scores from the CART model describe those variables
important for predicting missingness: rooms, price, suburb, council
area, distance, and region name. These variables are important for
predicting missingness, so are important to include in the imputation
model.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/rpart-plot-1}

}

\caption{Decision tree output predicting the clusters of missingness. Type of house, the year quarter, and year were important for predicting missingness cluster. The cluster with the most missingness was for quarters 1 and 4, for 2017 and 2018. Type of house, year, and year quarter are important features related to the missingness structure.}\label{fig:rpart-plot}
\end{figure}

\hypertarget{case-study-imp-diagnosis}{%
\subsection{Imputation and diagnostics}\label{case-study-imp-diagnosis}}

Two different imputation techniques are used: simple linear regression
and K nearest neighbors. Values are imputed stepwise in ascending order
of missingness. The \texttt{simputation} R package (van der Loo 2017) is
used to impute the values, applying the track missings pattern described
in Section \ref{verbs-track}, to assess imputed values.

Imputed data are combined by row binding imputation models. The data are
reshaped into long format to compare imputed values for each model for 4
variables (Figure \ref{fig:imputed-by-model}). Compared to KNN imputed
values, the linear model imputed values closer to the mean value.

\begin{figure}

{\centering \includegraphics[width=0.95\linewidth]{tidy-missing-data-paper_files/figure-latex/imputed-by-model-1}

}

\caption{Boxplots of complete case data, and data imputed with KNN or linear model for different variables. (A) number of bedrooms, (B) number of bathrooms, (C) number of carspots, and (D) landsize (on a log10 scale). KNN had similar results to complete case, and linear model had a lower median for cars and fewer extreme values for bedrooms.}\label{fig:imputed-by-model}
\end{figure}

\hypertarget{case-study-assess-model}{%
\subsection{Assess model predictions}\label{case-study-assess-model}}

The coefficients of the linear model of log price vary for room (Figure
\ref{fig:tidy-coefs}), for different imputed datasets. Notably, complete
cases resulted in underestimating the impact of room on log price. A
partial residual plot (Figure \ref{fig:partial-resid}) shows there is
not much variation amongst the models from the different datasets.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/tidy-coefs-1}

}

\caption{Visualization of the variation in coefficients for linear model of log price for each of the different datasets for the number of rooms. In red is complete case (cc), in green is the knn imputed dataset, and in blue the imputed by linear model. Using the complete case dataset produced smaller coefficients compared to the imputed models.}\label{fig:tidy-coefs}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/partial-resid-1}

}

\caption{Partial residual plot for each data set, complete cases (cc), imputed with KNN (knn), and imputed with a linear model (lm). There is not much variation amongst the different datasets from the different imputation methods.}\label{fig:partial-resid}
\end{figure}

\hypertarget{case-study-summary}{%
\subsection{Summary}\label{case-study-summary}}

The \texttt{naniar} and \texttt{visdat} packages build on existing tidy
tools and strike a compromise between automation and control that makes
analysis efficient, readable, but not overly complex. Each tool has
clear intent and effects - plotting or generating data or augmenting
data in some way. This reduces repetition and typing for the user,
making exploration of missing values easier as they follow consistent
rules with a declarative interface.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

This paper has described new methods for exploring, visualizing, and
imputing missing data. The work was motivated by recent developments of
tidy data, and extends them for better missing value handling. The
methods have standard outputs, function arguments, and behavior. This
provides consistent workflows centered around data analysis that
integrate well with existing imputation methodology, visualization, and
modelling.

The new data structures discussed in the paper could be used to create
different visualizations than were shown in the paper. The analyst can
use the data structures to decide on appropriate visualization for their
problem. The data structures could also be used to support interactive
graphics, in the manner of MANET and ggobi. Linking the plots (linked
brushing) to explore missingness or animating between different sets of
imputed values. New packages like \texttt{plotly} (Sievert 2018)
facilitate interactive graphics, and packages like \texttt{gganimate}
facilitate animations (Pedersen and Robinson 2017).

Other data structures such as spatial data, time series, networks, and
longitudinal data would be supported by the inherently tabular,
\emph{nabular} data, if they are first structured as wide tidy format.
Large data may need special handling, and additional features like
efficient storage of purely imputed values and lazy evaluation. Special
missing value codes could be improved by creating special classes, or
expanding low level representation of \texttt{NA} at the source code
level.

The methodology described in this paper can be used in conjunction with
other approaches to understand multivariate missingness dependencies
(e.g.~decision trees (Tierney et al. 2015), latent group analysis
(Barnett et al. 2017), and PCA (Lê et al. 2008)) . Evaluating imputed
values using a testing framework like Buuren (2012) is also supported.

The approach meshes with the dynamic nature of data analysis, allowing
the analyst to go from raw data to model data in a fluid workflow.

\newpage

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

The authors would like to thank Miles McBain, for his key contributions
and discussions on the \texttt{naniar} package, in particular for
helping implement \texttt{geom\_miss\_point}, and for his feedback on
ideas, implementations, and names. We also thank Colin Fay for his
contributions to the \texttt{naniar} package, in particular for his
assistance with the \texttt{replace\_with\_na} functions. We also thank
Earo Wang and Mitchell O'Hara-Wild for the many useful discussions on
missing data and package development, and for their assistance with
creating elegant approaches that take advantage of the tidy syntax. We
would also like to thank those who contributed pull requests and
discussions on the \texttt{naniar} package, in particular Jim Hester and
Romain François for improving the speed of key functions, Ross Gayler
for discussion on special missing values, and Luke Smith for helping
\texttt{naniar} be more compliant with \texttt{ggplot2}.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Barnett2017}{}%
Barnett, A. G., McElwee, P., Nathan, A., Burton, N. W., and Turrell, G.
(2017), ``Identifying patterns of item missing survey data using latent
groups: An observational study,'' \emph{BMJ open}, bmjopen.bmj.com, 7,
e017284.

\leavevmode\hypertarget{ref-Broman2017}{}%
Broman, K. W., and Woo, K. H. (2017), \emph{Data organization in
spreadsheets}, PeerJ Preprints; PeerJ Inc.

\leavevmode\hypertarget{ref-VanBuuren2012}{}%
Buuren, S. van (2012), \emph{Flexible imputation of missing data}, CRC
Press.

\leavevmode\hypertarget{ref-Cheng2015}{}%
Cheng, X., Cook, D., and Hofmann, H. (2015), ``Visually exploring
missing values in multivariable data using a graphical user interface,''
\emph{Journal of statistical software}, 68, 1--23.

\leavevmode\hypertarget{ref-Conway2017}{}%
Conway, J. R., Lex, A., and Gehlenborg, N. (2017), ``UpSetR: An R
package for the visualization of intersecting sets and their
properties,'' \emph{Bioinformatics}, 33, 2938--2940.

\leavevmode\hypertarget{ref-Cook2007}{}%
Cook, D., and Swayne, D. F. (2007), \emph{Interactive and dynamic
graphics for data analysis with R and GGobi}, Springer Publishing
Company, Incorporated.

\leavevmode\hypertarget{ref-Dempster1977}{}%
Dempster, A. P., Laird, N. M., and Rubin, D. B. (1977), ``Maximum
likelihood from incomplete data via the EM algorithm,'' \emph{Journal of
the Royal Statistical Society. Series B, Statistical methodology},
{[}Royal Statistical Society, Wiley{]}, 39, 1--38.

\leavevmode\hypertarget{ref-Donoho2017}{}%
Donoho, D. (2017), ``50 years of data science,'' \emph{Journal of
computational and graphical statistics: a joint publication of American
Statistical Association, Institute of Mathematical Statistics, Interface
Foundation of North America}, Taylor \& Francis, 26, 745--766.

\leavevmode\hypertarget{ref-Durre2008-ghcn}{}%
Durre, I., Menne, M. J., and Vose, R. S. (2008), ``Strategies for
evaluating quality assurance procedures,'' \emph{Journal of Applied
Meteorology and Climatology}, American Meteorological Society, 47,
1785--1791.

\leavevmode\hypertarget{ref-Ellis2017}{}%
Ellis, S. E., and Leek, J. T. (2017), \emph{How to share data for
collaboration}, PeerJ Preprints; PeerJ Inc.

\leavevmode\hypertarget{ref-mi}{}%
Gelman, A., and Hill, J. (2011), ``Opening windows to the black box,''
\emph{Journal of Statistical Software}, 40.

\leavevmode\hypertarget{ref-lubridate}{}%
Grolemund, G., and Wickham, H. (2011), ``Dates and times made easy with
lubridate,'' \emph{Journal of Statistical Software}, 40, 1--25.

\leavevmode\hypertarget{ref-Hmisc}{}%
Harrell Jr, F. E., Charles Dupont, and others. (2018), \emph{Hmisc:
Harrell miscellaneous}.

\leavevmode\hypertarget{ref-purrr}{}%
Henry, L., and Wickham, H. (2017), \emph{Purrr: Functional programming
tools}.

\leavevmode\hypertarget{ref-Honaker2010}{}%
Honaker, J., and King, G. (2010), ``What to do about missing values in
Time-Series Cross-Section data,'' \emph{American journal of political
science}, 54, 561--581.

\leavevmode\hypertarget{ref-amelia}{}%
Honaker, J., King, G., and Blackwell, M. (2011), ``Amelia II: A program
for missing data,'' \emph{Journal of Statistical Software}, 45, 1--47.

\leavevmode\hypertarget{ref-missMDA}{}%
Josse, J., and Husson, F. (2016), ``MissMDA: A package for handling
missing values in multivariate data analysis,'' \emph{Journal of
Statistical Software, Articles}, 70, 1--31.
\url{https://doi.org/10.18637/jss.v070.i01}.

\leavevmode\hypertarget{ref-Keeling2005-scripps}{}%
Keeling, C. D., Piper, S. C., Bacastow, R. B., Wahlen, M., Whorf, T. P.,
Heimann, M., and Meijer, H. A. (2005), ``Atmospheric CO2 and 13CO2
exchange with the terrestrial biosphere and oceans from 1978 to 2000:
Observations and carbon cycle implications,'' in \emph{A history of
atmospheric CO2 and its effects on plants, animals, and ecosystems},
eds. I. T. Baldwin, M. M. Caldwell, G. Heldmaier, R. B. Jackson, O. L.
Lange, H. A. Mooney, E.-D. Schulze, U. Sommer, J. R. Ehleringer, M.
Denise Dearing, and T. E. Cerling, New York, NY: Springer New York, pp.
83--113.

\leavevmode\hypertarget{ref-VIM}{}%
Kowarik, A., and Templ, M. (2016), ``Imputation with the R package
VIM,'' \emph{Journal of Statistical Software}, 74, 1--16.
\url{https://doi.org/10.18637/jss.v074.i07}.

\leavevmode\hypertarget{ref-recipes}{}%
Kuhn, M., and Wickham, H. (2018), \emph{Recipes: Preprocessing tools to
create design matrices}.

\leavevmode\hypertarget{ref-FactoMineR}{}%
Lê, S., Josse, J., and Husson, F. (2008), ``FactoMineR: A package for
multivariate analysis,'' \emph{Journal of Statistical Software}, 25,
1--18. \url{https://doi.org/10.18637/jss.v025.i01}.

\leavevmode\hypertarget{ref-Little1988}{}%
Little, R. J. A. (1988), ``A test of missing completely at random for
multivariate data with missing values,'' \emph{Journal of the American
Statistical Association}, Taylor \& Francis, 83, 1198--1202.

\leavevmode\hypertarget{ref-rpart-plot}{}%
Milborrow, S. (2018), \emph{Rpart.plot: Plot 'rpart' models: An enhanced
version of 'plot.rpart'}.

\leavevmode\hypertarget{ref-imputeTS}{}%
Moritz, S., and Bartz-Beielstein, T. (2017), ``imputeTS: Time Series
Missing Value Imputation in R,'' \emph{The R Journal}, 9, 207--218.

\leavevmode\hypertarget{ref-gganimate}{}%
Pedersen, T. L., and Robinson, D. (2017), \emph{Gganimate: A grammar of
animated graphics}.

\leavevmode\hypertarget{ref-kaggle2018}{}%
Pino, T. (2018), ``Melbourne housing market,''
\url{https://www.kaggle.com/anthonypino/melbourne-housing-market/version/21}.

\leavevmode\hypertarget{ref-norm}{}%
R by Alvaro A. Novo. Original by Joseph L. Schafer
\textless{}jls@stat.psu.edu\textgreater{}., P. to (2013), \emph{Norm:
Analysis of multivariate normal datasets with missing values}.

\leavevmode\hypertarget{ref-Ross2017}{}%
Ross, Z., Wickham, H., and Robinson, D. (2017), \emph{Declutter your R
workflow with tidy tools}, PeerJ Preprints; PeerJ Inc.

\leavevmode\hypertarget{ref-Rubin1976}{}%
Rubin, D. B. (1976), ``Inference and missing data,'' \emph{Biometrika},
Oxford University Press, 63, 581--592.

\leavevmode\hypertarget{ref-schafer-norm}{}%
Schafer, J. (1999), \emph{NORM: Multiple imputation of incomplete
multivariate data under a normal model}, University Park: Pennsylvania
State University, Department of Statistics.

\leavevmode\hypertarget{ref-Schafer2002}{}%
Schafer, J. L., and Graham, J. W. (2002), ``Missing data: Our view of
the state of the art,'' \emph{Psychological methods}, 7, 147--177.

\leavevmode\hypertarget{ref-plotly}{}%
Sievert, C. (2018), \emph{Plotly for r}.

\leavevmode\hypertarget{ref-tidytext}{}%
Silge, J., and Robinson, D. (2016), ``Tidytext: Text mining and analysis
using tidy data principles in r,'' \emph{JOSS}, The Open Journal, 1.
\url{https://doi.org/10.21105/joss.00037}.

\leavevmode\hypertarget{ref-Simon1986}{}%
Simon, G. A., and Simonoff, J. S. (1986), ``Diagnostic plots for missing
data in least squares regression,'' \emph{Journal of the American
Statistical Association}, Taylor \& Francis, 81, 501--509.

\leavevmode\hypertarget{ref-Sterne2009}{}%
Sterne, J. a C., White, I. R., Carlin, J. B., Spratt, M., Royston, P.,
Kenward, M. G., Wood, A. M., and Carpenter, J. R. (2009), ``Multiple
imputation for missing data in epidemiological and clinical research:
Potential and pitfalls,'' \emph{BMJ}, 338, b2393.

\leavevmode\hypertarget{ref-Swayne1998}{}%
Swayne, D. F., and Buja, A. (1998), ``Missing data in interactive
high-dimensional data visualization,'' \emph{Computational statistics},
researchgate.net, 1--8.

\leavevmode\hypertarget{ref-rpart}{}%
Therneau, T., and Atkinson, B. (2018), \emph{Rpart: Recursive
partitioning and regression trees}.

\leavevmode\hypertarget{ref-visdat}{}%
Tierney, N. (2017), ``Visdat: Visualising whole data frames,''
\emph{JOSS}, Journal of Open Source Software, 2, 355.
\url{https://doi.org/10.21105/joss.00355}.

\leavevmode\hypertarget{ref-Tierney2015}{}%
Tierney, N. J., Harden, F. A., Harden, M. J., and Mengersen, K. L.
(2015), ``Using decision trees to understand structure in missing
data,'' \emph{BMJ open}, bmjopen.bmj.com, 5, e007450.

\leavevmode\hypertarget{ref-treisman1985}{}%
Treisman, A. (1985), ``Preattentive processing in vision,''
\emph{Computer vision, graphics, and image processing}, 31, 156--177.

\leavevmode\hypertarget{ref-Unwin1996}{}%
Unwin, A., Hawkins, G., Hofmann, H., and Siegl, B. (1996), ``Interactive
graphics for data sets with missing values: MANET,'' \emph{Journal of
computational and graphical statistics: a joint publication of American
Statistical Association, Institute of Mathematical Statistics, Interface
Foundation of North America}, {[}American Statistical Association,
Taylor \& Francis, Ltd., Institute of Mathematical Statistics, Interface
Foundation of America{]}, 5, 113--122.

\leavevmode\hypertarget{ref-mice}{}%
van Buuren, S., and Groothuis-Oudshoorn, K. (2011), ``mice: Multivariate
imputation by chained equations in r,'' \emph{Journal of Statistical
Software}, 45, 1--67.

\leavevmode\hypertarget{ref-simputation}{}%
van der Loo, M. (2017), \emph{Simputation: Simple imputation}.

\leavevmode\hypertarget{ref-tidycensus}{}%
Walker, K. (2018), \emph{Tidycensus: Load us census boundary and
attribute data as 'tidyverse' and 'sf'-ready data frames}.

\leavevmode\hypertarget{ref-tsibble}{}%
Wang, E., Cook, D., and Hyndman, R. (2018), \emph{Tsibble: Tidy temporal
data frames and tools}.

\leavevmode\hypertarget{ref-ggplot2}{}%
Wickham, H. (2009), \emph{Ggplot2: Elegant graphics for data analysis},
Springer-Verlag New York.

\leavevmode\hypertarget{ref-Wickham2014}{}%
Wickham, H. (2014), ``Tidy data,'' \emph{Journal of statistical
software}, 59, 1--23.

\leavevmode\hypertarget{ref-Tidyverse-Manifesto}{}%
Wickham, H. (2017), ``The tidy tools manifesto,''
\url{https://cran.r-project.org/web/packages/tidyverse/vignettes/manifesto.html}.

\leavevmode\hypertarget{ref-stringr}{}%
Wickham, H. (2018), \emph{Stringr: Simple, consistent wrappers for
common string operations}.

\leavevmode\hypertarget{ref-readxl}{}%
Wickham, H., and Bryan, J. (2017), \emph{Readxl: Read excel files}.

\leavevmode\hypertarget{ref-dplyr}{}%
Wickham, H., Francois, R., Henry, L., and Müller, K. (2017a),
\emph{Dplyr: A grammar of data manipulation}.

\leavevmode\hypertarget{ref-r4ds}{}%
Wickham, H., and Grolemund, G. (2016), \emph{R for data science: Import,
tidy, transform, visualize, and model data}, ``O'Reilly Media, Inc.''.

\leavevmode\hypertarget{ref-tidyr}{}%
Wickham, H., and Henry, L. (2018), \emph{Tidyr: Easily tidy data with
'spread()' and 'gather()' functions}.

\leavevmode\hypertarget{ref-readr}{}%
Wickham, H., Hester, J., and Francois, R. (2017b), \emph{Readr: Read
rectangular text data}.

\leavevmode\hypertarget{ref-haven}{}%
Wickham, H., and Miller, E. (2018), \emph{Haven: Import and export
'spss', 'stata' and 'sas' files}.

\end{document}

### 2018-09-07 paper/tidy-missing-data-paper.Rmd
* move images out of images/ folder * remove relative paths from Rmd * hopefully should compile for arxiv now

@@
 
-ggsave("images/geom-point-warn.png", 
+ggsave("geom-point-warn.png", 
        width = 4, 
@@
 
-img_gg_p <- image_read("images/geom-point-warn.png")
-
-img_gg_warn <- image_read("images/ggplot-warn.jpeg")
+img_gg_p <- image_read("geom-point-warn.png")
+
+img_gg_warn <- image_read("ggplot-warn.jpeg")
 
 image_append(c(img_gg_warn, img_gg_p)) %>% 
-  image_write("images/ggplot2-warning-and-plot.png", "png")
-
-knitr::include_graphics("images/ggplot2-warning-and-plot.png")
+  image_write("ggplot2-warning-and-plot.png", "png")
+
+knitr::include_graphics("ggplot2-warning-and-plot.png")
 
@@
 
-knitr::include_graphics("images/full-conversion-to-shadow.png")
+knitr::include_graphics("full-conversion-to-shadow.png")
 
@@
 
-knitr::include_graphics("images/nabular.png")
+knitr::include_graphics("nabular.png")
 

### 2019-02-21 paper/tidy-missing-data-paper.Rmd
just a final commit to note that this is as the paper was submitted to arxiv
add address and minor update of figures

--- 2018-09-07 paper/tidy-missing-data-paper.Rmd
@@
 title: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
-author: "Nicholas Tierney, Department of Econometrics and Business Statistics, Monash University, **Corresponding author (nicholas.tierney@gmail.com)** \\newline Dianne Cook, Department of Econometrics and Business Statistics, Monash University"
+author: "Nicholas Tierney, Department of Econometrics and Business Statistics, Monash University, E774, Menzies Building Monash University, Clayton, 3168, Victoria, Australia **Corresponding author (nicholas.tierney@gmail.com)** \\newline Dianne Cook, Department of Econometrics and Business Statistics, Monash University"
 date: "September 6, 2018"

### 2019-02-22 cover-letter/cover-letter.Rmd
update cover letter

@@
 fromsignature: sigfile.png
-toname: Dr. Tyler McCormick
-toaddress: JCGS
-opening: Dear Dr. McCormick
-closing: Regards
+toname: Dr. B.U. Park
+toaddress: CSDA
+opening: Dear Dr. D.U. Park
+closing: Regards, \newline Dr. Nicholas Tierney \newline Reseach Fellow
 spacing: 1.4
@@
 
-Please consider my article "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations" for possible publication in JCGS.
+Please consider my article "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations" for possible publication in CSDA.
 This article describes new work on handling missing values from a data centric approach.
@@
 Along with the PDF is an accompanying Rmarkdown file, which contains all of the code to reproduce the results in this paper.
-This is the first time that this article has been submitted to a journal, and is not under consideration elsewhere.
+This article is not under consideration elsewhere.
 We have submitted a pre-print of this article to arxiv, and to our internal department pre-print at Monash University.
 
-We hope that you find this paper of interest to the readers of JCGS,
+We hope that you find this paper of interest to the readers of CSDA,

### 2019-02-22 paper/tidy-missing-data-paper.Rmd
update reference data to point to zenodo dataset
update date

--- 2018-09-07 paper/tidy-missing-data-paper.Rmd
@@
 title: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
-author: "Nicholas Tierney, Department of Econometrics and Business Statistics, Monash University, **Corresponding author (nicholas.tierney@gmail.com)** \\newline Dianne Cook, Department of Econometrics and Business Statistics, Monash University"
-date: "September 6, 2018"
+author: "Nicholas Tierney, Department of Econometrics and Business Statistics, Monash University, E774, Menzies Building Monash University, Clayton, 3168, Victoria, Australia **Corresponding author (nicholas.tierney@gmail.com)** \\newline Dianne Cook, Department of Econometrics and Business Statistics, Monash University"
+date: "February 23, 2019"
 toc: false
@@
 To illustrate the methods, data on housing for the city of Melbourne from January 28, 2016 to March 17, 2018 is used.
-The data was compiled by scraping weekly property clearance data [@kaggle2018].
+The data was compiled by scraping weekly property clearance data [@Kaggle-2018-data].
 There are `r scales::number(nrow(housing), big.mark = ",")` properties, and `r ncol(housing)` variables in the dataset.

### 2020-01-02 paper/tidy-missing-data-paper.Rmd
create a JSS article that actually compiles

@@
 ---
-title: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
-author: "Nicholas Tierney, Department of Econometrics and Business Statistics, Monash University, E774, Menzies Building Monash University, Clayton, 3168, Victoria, Australia **Corresponding author (nicholas.tierney@gmail.com)** \\newline Dianne Cook, Department of Econometrics and Business Statistics, Monash University"
-date: "February 23, 2019"
-toc: false
-bibliography: tidy-missing-data-paper.bib
-output:
-  bookdown::pdf_document2:
-    fig_caption: yes
-    fig_height: 5
-    fig_width: 8
-    includes:
-      in_header: preamble.tex
-    keep_tex: yes
-    number_sections: yes
-csl: jcgs.csl
----
-
-**Abstract**:
-
+documentclass: jss
+author:
+  - name: Nicholas Tierney
+    affiliation: Monash University
+    address: >
+      Department of Econometrics and Business Statistics
+      E774, Menzies Building
+      Monash University, Clayton
+    email: \email{nicholas.tierney@gmail.com}
+    url: http://www.njtierney.com
+  - name: Dianne Cook
+    affiliation: Monsh University
+title:
+  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations with the Package \\pkg{naniar}"
+  # If you use tex in the formatted title, also supply version without
+  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations with the Package naniar"
+  # For running headers, if needed
+  short:     "\\pkg{naniar}: Explore missing data"
+abstract: >
   Despite the large body of research on missing value distributions and imputation, there is comparatively little literature on how to make it easy to handle, explore, and impute missing values in data.
@@
 These methods have been made available in the R package `naniar`.
-
-_Keywords:_ workflow, statistical computing, data science, data visualization, tidyverse, data pipeline
-
-\newpage
+keywords:
+  # at least one keyword must be supplied
+  formatted: [workflow, statistical computing, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
+  plain:     [kworkflow, statistical computing, data science, data visualization, tidyverse, data pipeline, R]
+preamble: >
+  \usepackage{amsmath}
+  \usepackage[english]{babel}
+output: 
+  rticles::jss_article:
+    fig_caption: yes
+    fig_height: 5
+    fig_width: 8
+    includes:
+      in_header: header.tex
+    keep_tex: yes
+    number_sections: yes
+    citation_package: 'natbib'
+bibliography: tidy-missing-data-paper.bib
+# csl: jcgs.csl
+---
 
@@
                       cache = TRUE)
+```
+
+```{r setup2, include=FALSE, message=FALSE, warning=FALSE}
 library(tidyverse)
 library(gridExtra)
-library(kableExtra)
+# options(kableExtra.latex.load_packages = FALSE)
+# library(kableExtra)
 library(ggthemes)
@@
 library(magick)
+options(tinytex.verbose = TRUE)
+# knit_hooks$set(
+#   document = function(x) {
+#     sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed = TRUE)
+#   }
+# )
 
@@
 
-Data manipulation is made input and output tidy with R packages `dplyr` and `tidyr` [@dplyr; @tidyr].
+Data manipulation is made input and output tidy with R packages \pkg{dplyr} and \pkg{tidyr} [@dplyr; @tidyr].
 These provide the five "verbs" of data manipulation: data reshaping, sorting, filtering, transforming, and aggregating.
@@
 Visualization tools only have tidy data as their input, as the output is a graphic.
-The popular domain specific language `ggplot2` maps variables in a dataset to features (referred to as aesthetics) of a graphic [@ggplot2].
+The popular domain specific language \pkg{ggplot2} maps variables in a dataset to features (referred to as aesthetics) of a graphic [@ggplot2].
 For example, a scatterplot can be created by mapping two variables to the x and y axes, and specifying a point geometry.
@@
 Modelling tools work well with tidy data, as they have a clear mapping from variables in the data to the formula for a model.
-For example in R, y regressed on x and z is: `lm(y ~ x + z)`.
+For example in R, y regressed on x and z is: \code{lm(y ~ x + z)}.
 Modelling tools are input tidy, but their output is always messy.
 For example, estimated coefficients, predictions, and residuals from one model cannot be easily combined with the output of another model.
-The `recipes` package is a tidy tool under development to help make modelling input and output tidy [@recipes].
+The \pkg{recipes} package is a tidy tool under development to help make modelling input and output tidy [@recipes].
 
@@
 These are designed to share similar principles in their design and behavior, and cover the breadth of an analysis - going from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @Ross2017; @Tidyverse-Manifesto].
-This has led to a growth of tools for specific parts of analysis - from reading in data with `readr`, `readxl`, and `haven`, to handling character strings with `stringr`, dates with `lubridate`, and performing functional programming with `purrr` [@readr; @readxl; @haven; @stringr; @lubridate; @purrr].
+This has led to a growth of tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr}, dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} [@readr; @readxl; @haven; @stringr; @lubridate; @purrr].
 It has also led to the growth of new packages for other fields that follow similar design principles and create fluid workflows for new domains.
-For example, the `tidytext` package for text analysis, the `tsibble` package for time series data, and `tidycensus` for working with US census and boundary data [@tidytext; @tsibble; @tidycensus].
+For example, the \pkg{tidytext} package for text analysis, the \pkg{tsibble} package for time series data, and \pkg{tidycensus} for working with US census and boundary data [@tidytext; @tsibble; @tidycensus].
 
@@
 The complex, often multivariate nature of imputation methods also makes makes them difficult to represent.
-Visualization methods for missing data do not map data features to the aesthetics of a graphic, as in `ggplot2`, limiting expressive exploration.
+Visualization methods for missing data do not map data features to the aesthetics of a graphic, as in \pkg{ggplot2}, limiting expressive exploration.
 
@@
 
-`VIM` [@VIM] efficiently implements well-used imputation methods k-nearest neighbor, regression, hot-deck, and iterative robust model-based imputation.
+\pkg{VIM} [@VIM] efficiently implements well-used imputation methods k-nearest neighbor, regression, hot-deck, and iterative robust model-based imputation.
 This diversity of approaches allow the package to deal semi-continuous, continuous, count, and categorical data.
@@
 So `Var1` would have a sibling indicator column, `Var1_imp`, with TRUE or FALSE values to indicate imputed value.
-`VIM` also has a variety of visualization methods, discussed in section \@ref(exploration).
-The `simputation` package includes imputation methods for linear regression, decision trees, K nearest neighbors, hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977].
+\pkg{VIM} also has a variety of visualization methods, discussed in section \@ref(exploration).
+The \pkg{simputation} package includes imputation methods for linear regression, decision trees, K nearest neighbors, hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977].
 A key feature of its design is that it returns a regular dataframe.
@@
 Providing a dataframe with imputed values included in the data reduces the friction of working with other tools, but comes at the cost identifying imputed values.
-Other software for imputation includes `Hmisc` [@Hmisc], which provide predictive mean matching, `imputeTS` [@imputeTS], which provide time series imputation methods, and `missMDA` [@missMDA], which imputes data using principal components analysis.
+Other software for imputation includes \pkg{Hmisc} [@Hmisc], which provide predictive mean matching, \pkg{imputeTS} [@imputeTS], which provide time series imputation methods, and \pkg{missMDA} [@missMDA], which imputes data using principal components analysis.
 
@@
 Multiple imputation is often regarded as best practice for imputing values [@Schafer2002], as long as appropriate caution is taken [@Sterne2009].
-Popular and robust methods for multiple imputation include the `mice`, `Amelia`, and `mi` packages [@mice; @amelia; @mi].
-The `mice` package implements the method of chained equations, using a variable-wise algorithm to calculate the posterior distribution of parameters, from which uses to generate imputed values.
-The suggested workflow in `mice` revolves around imputing data, returning completed data, and fitting a model and pooling the results.
-The `Amelia` package [@amelia] assumes data are multivariate normal, and samples from the posterior using the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm [@Honaker2010] and allows for incorporation of information on the values in a prior.
-The `mi` package [@mi] also uses Bayesian models for imputation, providing better handling of semi-continuous values, and data with structural or perfect correlation.
-A collection of analysis models are also provided in `mi`, which work with the specially multiply imputed data, including linear models, generalized linear models, and their corresponding Bayesian components.
+Popular and robust methods for multiple imputation include the \pkg{mice}, \pkg{Amelia}, and \pkg{mi} packages [@mice; @amelia; @mi].
+The \pkg{mice} package implements the method of chained equations, using a variable-wise algorithm to calculate the posterior distribution of parameters, from which uses to generate imputed values.
+The suggested workflow in \pkg{mice} revolves around imputing data, returning completed data, and fitting a model and pooling the results.
+The \pkg{Amelia} package [@amelia] assumes data are multivariate normal, and samples from the posterior using the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm [@Honaker2010] and allows for incorporation of information on the values in a prior.
+The \pkg{mi} package [@mi] also uses Bayesian models for imputation, providing better handling of semi-continuous values, and data with structural or perfect correlation.
+A collection of analysis models are also provided in \pkg{mi}, which work with the specially multiply imputed data, including linear models, generalized linear models, and their corresponding Bayesian components.
 This approach promotes fluid workflow, with a similar penalty to tidying up model output, which is still messy.
-Another software for performing multiple imputation is `norm` [@norm; @schafer-norm], which uses methods from the NORM software from @schafer-norm, such as multiple imputations using EM for multivariate normal data.
-The `norm` package does not provide a framework for tracking missing values, but instead provides tools for making inference from multiple imputation.
+Another software for performing multiple imputation is \pkg{norm} [@norm; @schafer-norm], which uses methods from the NORM software from @schafer-norm, such as multiple imputations using EM for multivariate normal data.
+The \pkg{norm} package does not provide a framework for tracking missing values, but instead provides tools for making inference from multiple imputation.
 Each of these multiple imputation methods provide practical methods, but do not have consistent interfaces across their methods or implementation, which makes then inherently messy, as their output is always different.
@@
 
-`ggplot2` incorporates missingness into visualizations only when mapping a discrete variable to a graph aesthetic.
+\pkg{ggplot2} incorporates missingness into visualizations only when mapping a discrete variable to a graph aesthetic.
 This has some limitations.
 For example, for data with school grades and test scores, a boxplot visualization could have school grade mapped to the x axis, and test score to the y axis (See Figure \@ref(fig:gg-box-na)).
-If there are missings in a continuous variable like test score, `ggplot2` omits the missing values and prints a warning message.
+If there are missings in a continuous variable like test score, \pkg{ggplot2} omits the missing values and prints a warning message.
 However, if a discrete variable like school year has missing values, an NA category is created for school year, and the scores are placed into this.
@@
 For example, instrument failure and drop out would be represented as 2 and 3, see Table \@ref(tab:shadow-encoding) for an example.
-Encoding these special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in `naniar`, shown in Section \@ref(verbs-recode).
+Encoding these special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in \pkg{naniar}, shown in Section \@ref(verbs-recode).
 
@@
 The temperature value of -99 is represented in the shadow column as 'NA\\_instr', which in turn has the underlying numeric value of 2.
-This captures additional information about the data that is otherwise difficult to record.",
-             booktabs = TRUE) %>%
-  kable_styling(latex_options = c("hold_position"))
+This captures additional information about the data that is otherwise difficult to record.")
+             # booktabs = TRUE) 
+  # kable_styling(latex_options = c("hold_position"))
 
@@
 Four different areas of visualization of missing data are discussed: overviews (Section \@ref(overview-plots)); univariate (Section \@ref(univariate-plots)), bivariate (Section \@ref(bivariate-plots)), and multivariate (Section \@ref(multivariate-plots)).
-All plots are created using `ggplot2`, giving users clearer control over the plot appearance, and customizability.
+All plots are created using \pkg{ggplot2}, giving users clearer control over the plot appearance, and customizability.
 
@@
 All missings in a dataset can be displayed using a heatmap style visualization.
-This is achieved using `vis_miss()` from the `visdat` package [@visdat], which also provides summaries of missingness overall in the legend, and for each column (see Figure \@ref(fig:vis-miss)).
+This is achieved using `vis_miss()` from the \pkg{visdat} package [@visdat], which also provides summaries of missingness overall in the legend, and for each column (see Figure \@ref(fig:vis-miss)).
 The user can also apply clustering to the rows, and arrange columns by missingness (Figure \@ref(fig:vis-miss)B).
-Similar visualizations are available in other missing data packages such as `VIM`, `mi`, `Amelia`, and `MissingDataGUI`.
+Similar visualizations are available in other missing data packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}.
 A key improvement is that `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns in the visualization and are named at the top, and each row is an observation.
-Using `ggplot2` means the plot can be customized and combined with other ggplot graphics.
+Using \pkg{ggplot2} means the plot can be customized and combined with other ggplot graphics.
 
@@
 As in Section \@ref(imp-below-range), using _nabular_ data identifies imputed values, and color makes missingness pre-attentive [@treisman1985].
-The steps of imputing and coloring have been been combined into `geom_miss_point()` from the `naniar` package (Figure \@ref(fig:geom-miss)), where there is a mostly uniform spread of missing values for Solar.R and Ozone.
+The steps of imputing and coloring have been been combined into `geom_miss_point()` from the \pkg{naniar} package (Figure \@ref(fig:geom-miss)), where there is a mostly uniform spread of missing values for Solar.R and Ozone.
 
@@
 
-As `geom_miss_point` is a defined geometry for `ggplot2`, it provides features such as faceting and mapping other variables to aesthetics such as color, shape, or size.
+As `geom_miss_point` is a defined geometry for \pkg{ggplot2}, it provides features such as faceting and mapping other variables to aesthetics such as color, shape, or size.
 The shadow format by itself can also be used to visually explore some patterns of missingness.
@@
 They transform variables to the same scale, and typically make the data values range between 0 and 1.
-To showcase this visualization, the dataset `oceanbuoys` from `naniar` is used, which contains measurements of moored ocean buoys to understand and predict El Niño and El Niña.
+To showcase this visualization, the dataset `oceanbuoys` from \pkg{naniary} is used, which contains measurements of moored ocean buoys to understand and predict El Niño and El Niña.
 The data were collected in 1993 and 1997, and contains information on the sea and air temperature, humidity, and east west and north south wind directions.
@@
 There are different ways to summarize the amount of missing and complete values in a dataset.
-This section describes approaches to summarizing missingness, implemented in the `naniar` package.
+This section describes approaches to summarizing missingness, implemented in the \pkg{naniar} package.
 The functions are easy to remember as they follow consistent naming, and their output is consistent, and returns a single number or dataframe.
@@
   caption = "Single number summaries of missing and complete data, applied to the airquality dataset.
-These functions follow consistent naming, to make them easy to remember, and to make their use clear.",
-  booktabs = TRUE) %>%
-  kable_styling(latex_options = c("hold_position"))
+These functions follow consistent naming, to make them easy to remember, and to make their use clear.")
+  # booktabs = TRUE)
+  # kable_styling(latex_options = c("hold_position"))
 
@@
 Functions are also provided for case-wise summaries and tabulations with `miss_case_summary` and `miss_case_table`, which order by `n_miss`, so the most missings are always shown at the top.
-Other summaries provided in `naniar` include summaries of missingness across a provided repeating span, `miss_var_span`, and finding the streaks or runs of missingness in a given variable in `miss_var_run`.
+Other summaries provided in \pkg{naniar} include summaries of missingness across a provided repeating span, `miss_var_span`, and finding the streaks or runs of missingness in a given variable in `miss_var_run`.
 
@@
     caption = "Summary of the number and percent of missings in each variable. Only ozone and solar radiation have missing values.",
-    digits = 1,
-    booktabs = TRUE) %>%
-  kable_styling(latex_options = c("hold_position"))
+    digits = 1)
+    # booktabs = TRUE) 
+  # kable_styling(latex_options = c("hold_position"))
 
@@
 There are not many patterns of missingness.",
-    digits = 1,
-    booktabs = TRUE) %>%
-  kable_styling(latex_options = c("hold_position"))
+    digits = 1)
+    # booktabs = TRUE)
+  # kable_styling(latex_options = c("hold_position"))
 ```
@@
 There are more ozone missings in June than May.",
-               digits = 1,
-               booktabs = TRUE) %>%
-  kable_styling(latex_options = c("hold_position"))
+               digits = 1)
+               # booktabs = TRUE)
+  # kable_styling(latex_options = c("hold_position"))
 
@@
 knitr::kable(dat_ms, 
-             caption = "An example dataset with variables x, y and z containing some common unusual missing values.",
-             booktabs = TRUE) %>%
-  kable_styling(latex_options = c("hold_position"))
+             caption = "An example dataset with variables x, y and z containing some common unusual missing values.")
+             # booktabs = TRUE) 
+  # kable_styling(latex_options = c("hold_position"))
 ```
@@
   knitr::kable(
-    caption = "Table of the occurences of the search '-99' in the data, 'dat\\_ms'. There is one occurence if -99 in variables x and z.",
-    booktabs = TRUE) %>%
-  kable_styling(latex_options = c("hold_position"))
+    caption = "Table of the occurences of the search '-99' in the data, 'dat\\_ms'. There is one occurence if -99 in variables x and z.")
+    # booktabs = TRUE) 
+  # kable_styling(latex_options = c("hold_position"))
 ```
@@
 
-### shadow: Creating `nabular` data {#verbs-nabular}
+### shadow: Creating nabular data {#verbs-nabular}
 
@@
   summarise_at(.vars = vars(Ozone),
-               .funs = funs(min, mean, median, max))
+               .funs = list(min, mean, median, max))
 ```
@@
   summarise_at(.vars = vars(Ozone),
-               .funs = funs(min, mean, median, max)) %>%
+               .funs = lst(min, mean, median, max)) %>%
   knitr::kable(
@@
 The mean and median values are similar, but the minimum and maximum values are different.
-The comparison of imputed values is similar to other dplyr summary workflows.",
-    booktabs = TRUE) %>%
-  kable_styling(latex_options = c("hold_position"))
+The comparison of imputed values is similar to other dplyr summary workflows.")
+    # booktabs = TRUE) 
+  # kable_styling(latex_options = c("hold_position"))
 ```
@@
 There are different patterns of missingness in variables and cases, but they can be broken down into smaller groups.",
-             digits = 1,
-             booktabs = TRUE) %>%
-  kable_styling(latex_options = c("hold_position"))
+             digits = 1)
+             # booktabs = TRUE)
+  # kable_styling(latex_options = c("hold_position"))
 

### 2020-01-02 paper/tidy-missing-data-paper.Rmd
remove unnecessary codes

@@
   rticles::jss_article:
-    fig_caption: yes
-    fig_height: 5
-    fig_width: 8
     includes:
@@
     keep_tex: yes
-    number_sections: yes
     citation_package: 'natbib'
@@
                       cache = TRUE)
-```
-
-```{r setup2, include=FALSE, message=FALSE, warning=FALSE}
+
 library(tidyverse)
@@
 options(tinytex.verbose = TRUE)
-# knit_hooks$set(
-#   document = function(x) {
-#     sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed = TRUE)
-#   }
-# )
-
 ```

### 2020-01-03 paper/tidy-missing-data-paper.Rmd
fixes section/table/figure referencing with bookdown::pdf_book: base_format: rticles::jss_article

@@
 output:
-  rticles::jss_article:
+  bookdown::pdf_book:
+    base_format: rticles::jss_article
     includes:
@@
 housing_raw <-
-  readr::read_csv("../data/melbourne_housing_raw.csv") %>%
+  readr::read_csv(here::here("data",
+                             "melbourne_housing_raw.csv")) %>%
   janitor::clean_names() %>%
@@
 ```{r partial-resid, fig.cap = "Partial residual plot for each data set, complete cases (cc), imputed with KNN (knn), and imputed with a linear model (lm).
-There is not much variation amongst the different datasets from the different imputation methods.", fig.width = 9, fig.height = 3, out.width = "100%"}
+These are plotted with hex bins, which are coloured according to the number of points in a given hexagon - the brighter the colour the more points.
+This reveals that the data sets imputed with knn and linear models have more points clustered around zero, whereas complete cases is more flat.", fig.width = 9, fig.height = 3, out.width = "100%"}
 
@@
              y = .resid)) + 
-  geom_point(alpha = 0.1) + 
+  # geom_point(alpha = 0.1) + 
+  geom_hex() +
+  labs(fill = "Count") +
+  scale_fill_viridis_c() +
   facet_wrap(~model)

### 2020-01-03 paper/tidy-missing-data-paper.Rmd
various fixes to .bib and other comments. Resolves #9

@@
   formatted: [workflow, statistical computing, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
-  plain:     [kworkflow, statistical computing, data science, data visualization, tidyverse, data pipeline, R]
+  plain:     [workflow, statistical computing, data science, data visualization, tidyverse, data pipeline, R]
 preamble: >
@@
 
-The imputation literature does not address how to handle missing data. Producing
-a complete dataset for analysis, whether by case or variable deletion, or with
-imputation, requires knowing something about the structure of the missing
-values. To understand their structure, one needs to generate visualizations,
-calculate summaries, and perform exploratory modelling. The imputation methods
-literature (e.g. @Little1988, @Rubin1976, @Simon1986, @Schafer2002, 
-@VanBuuren2012) focuses on ensuring valid statistical inference is made from
+The imputation literature (e.g. @Little1988, @Rubin1976, @Simon1986,
+@Schafer2002, @VanBuuren2012) does not address how to handle missing data.
+Instead the focus is on ensuring valid statistical inference is made from
 incomplete data. They approach this chiefly through probabilistic modelling, and
@@
 
-Exploring missing data to find the missingness mechanism is difficult, because it is an iterative process with many dead ends and often no clear answer.
-There is no silver bullet for missing data that will reveal it's mechanism to the analyst.
-Despite it being crucial to explore and understand missing data, there is very little research on how to make this process more efficient so that it integrates into analysis pipelines.
-Although decision tree models or latent group analysis can reveal structures or patterns of missingness [@Tierney2015; @Barnett2017], these are not definitive and usually require discussion with data curators to validate.
-The analyst must explore missing data using visualizations, summaries, and exploratory analyses, but there is a lack of tools to efficiently explore missing data.
-
-The graphics literature [@Swayne1998; @Unwin1996] provides several solutions for exploring missings visually.
+However, something has to be known about the missing value structure to produce
+a complete dataset for analysis. Whether by case or variable deletion, or with
+imputation, it is ideal to know what the mechanism is that generates missing 
+values.
+To understand their structure, the data must be explored, which means generating visualizations, summaries, and performing exploratory modelling.
+This process is inherently difficult, as it is iterative, and
+involves many dead ends, and there might not be a clear answer even after much
+exploratory data analysis.
+
+Although
+decision tree models or latent group analysis can reveal structures or patterns
+of missingness [@Tierney2015; @Barnett2017], these are not definitive and
+usually require discussion with data curators to validate.
+While there might be other times where the data are
+clearly missing completely at random (or missing at random), determining this is not always straightforward. There is no silver bullet for missing data that will reveal
+it's mechanism to the analyst. Missing data must be explored.
+
+While there have been many software tools for exploring missing data (VIM, mi, mice, MissingDataGUI, MANET, ggobi, norm), these tools generate outputs that make them challenging to integrate into analysis pipelines.
+Despite it being crucial to explore and understand missing data, there is very little research on the concepts that make
+this process more efficient, and integrate well into analysis pipelines.
+
+The analyst must
+explore missing data using visualizations, summaries, and exploratory analyses,
+but there is a lack of modern tools to efficiently explore missing data.
+
+The graphics literature [@Swayne1998; @Unwin1996; @VIM; @MissingDataGUI; ] provides several solutions for exploring missings visually.
 These approaches incorporate missings into the plot in some way.
@@
 
-Features of tidy data were formally discussed in 2014 [@Wickham2014], subsequently generating much discussion (see @Donoho2017 and commentary), and tools for data analysis. Tidy data is defined as:
-
-> 1. Each variable forms a column.
-> 2. Each observation forms a row.
-> 3. Each type of observational unit forms a table.
+Features of tidy data were formally discussed in 2014 [@Wickham2014], subsequently generating much discussion (see @Donoho2017 and commentary, @r4ds), and tools for data analysis.
+Tidy data is defined as:
+
+> 1. Each variable must have its own column.
+> 1. Each observation must have its own row.
+> 1. Each value must have its own cell.
 
@@
 For example in R, y regressed on x and z is: \code{lm(y ~ x + z)}.
-Modelling tools are input tidy, but their output is always messy.
+Modelling tools are input tidy, but their output is always messy - it is not in the right format for subsequent steps in analysis.
 For example, estimated coefficients, predictions, and residuals from one model cannot be easily combined with the output of another model.
@@
 
-Defining tidy data and tidy tools has resulted in the growth of a set of packages known collectively as the "tidyverse" [@Ross2017].
-These are designed to share similar principles in their design and behavior, and cover the breadth of an analysis - going from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @Ross2017; @Tidyverse-Manifesto].
+Defining tidy data and tidy tools has resulted in the growth of a set of packages known collectively as the "tidyverse" [@tidyverse].
+These are designed to share similar principles in their design and behavior, and cover the breadth of an analysis - going from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @tidyverse; @Tidyverse-Manifesto].
 This has led to a growth of tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr}, dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} [@readr; @readxl; @haven; @stringr; @lubridate; @purrr].
 It has also led to the growth of new packages for other fields that follow similar design principles and create fluid workflows for new domains.
-For example, the \pkg{tidytext} package for text analysis, the \pkg{tsibble} package for time series data, and \pkg{tidycensus} for working with US census and boundary data [@tidytext; @tsibble; @tidycensus].
+For example, the \pkg{tidytext} package for text analysis, the \pkg{tsibble} package for time series data, and \pkg{tidycensus} for working with US census and boundary data [@tidytext; @wang2020tsibble; @tidycensus].
 
@@
 Missing data tools can be used to perform imputations, missing data diagnostics, and data visualizations.
-However, these tools are primarily _messy_, and suffer the same problems as modelling for imputation: They use tidy input, but produce messy output.
+However, these tools suffer the same problems as modelling for imputation: They use tidy input, but produce messy output - their output is challenging to integrate with other steps of data analysis.
 The complex, often multivariate nature of imputation methods also makes makes them difficult to represent.
@@
 Past graphics research on missing data has not yet been framed in a tidy way.
-The graphics literature has explored methods for missing value exploration and imputation. Translating and expanding these to fit within tidy data and tidy tools would create more efficient workflows.
+The graphics literature has explored methods for missing value exploration and imputation.
+Translating and expanding these to fit within tidy data and tidy tools would create more efficient workflows.
+By laying down these concepts, it allows for the focus to be more general than just software, but rather, a framework for tidy tools to explore missing data that can be built upon, extended, and implemented in other languages.
 

### 2020-01-06 paper/tidy-missing-data-paper.Rmd
tidy up section 2

@@
 
-There are many ways each value can go missing.
-The information in $R$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), Missing at random, and missing not at random.
-**MCAR** is where values being missing have no association with observed or unobserved data, that is $Pr(missing | observed, unobserved) = Pr(missing)$.
-Essentially, the probability of an observation being missing is unrelated to anything else.
-Although this is a convenient scenario, it is not actually possible to confirm, as it relies on statements on data unobserved.
-**MAR** refers to cases where the missingness only depends on the data observed, and not data unobserved, that is,$Pr(missing | observed, unobserved) = Pr(missing | observed)$.
-Some structure or dependence between the missing values and observed values is allowed, provided that this can be explained by the data observed.
-Finally, **MNAR** refers to the missingness being related to values unobserved: $Pr(missing | observed, unobserved)$.
-This assumes that conditioning on all available observed data, the data goes missing due to some phenomena unobserved.
+There are many ways each value can go missing, we adopt the notation used in @VanBuuren2012.
+The information in $R$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), Missing at random (MAR), and missing not at random (MNAR).
+The distribution of missing values in $r_{ij}$ can depend on the entire dataset, which is represented as $Y = (Y_{obs}, Y_{miss})$.
+This relationship can be defined by the _missing data model_, of which $/psi$ contains the model parameters, and the missing data model can written as $Pr(r_{ij} | Y_{obs}, Y_{miss}, \psi)$ - the probability of missingness is conditional on the data observed, data missing, and some probability parameter of missingness.
+We can then use this to help precisely define categories of missing values.
+
+**MCAR** is where values being missing have no association with observed or unobserved data, that is $Pr(R = 1 | Y_{obs}, Y_{miss}) = Pr(R = 1 | \psi)$.
+Essentially, the probability of an observation being missing is unrelated to anything else, only the parameters $\psi$, which is the overall probability of missingness.
+Although this is a convenient scenario, it is not actually possible to confirm, or clearly distinguish from MAR, as it relies on statements on data unobserved.
+
+**MAR** refers to cases where the missingness only depends on the data observed, and not data missing, that is,$Pr(R = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(R = 1 | Y_{obs}, \psi)$.
+Some structure or dependence between the missing values and observed values is allowed, provided that this can be explained by the data observed, and some overall probability of missingness.
+
+Finally, **MNAR** refers to the missingness being related to values observed, and un unobserved; if this does not simplify  $Pr(R = 1 | Y_{obs}, Y_{miss}, \psi)$.
+This assumes that conditioning on all available observed data, the data goes missing due to some phenomena unobserved, including the structure of the missing data itself.
 This presents a challenge in analysis, as it is difficult to verify this state, and also implies bias in analysis due to the unobserved phenomena.
@@
 \pkg{VIM} also has a variety of visualization methods, discussed in section \@ref(exploration).
-The \pkg{simputation} package includes imputation methods for linear regression, decision trees, K nearest neighbors, hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977].
-A key feature of its design is that it returns a regular dataframe.
-This approach contrasts from other imputation software, which may provide custom classes of data or additional objects, which require additional work to prepare for subsequent analysis.
+The \pkg{simputation} package provides an interface into popular imputation methods for linear regression, decision trees, K nearest neighbors, hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977].
+Other software for imputation includes \pkg{Hmisc} [@Hmisc], which provide predictive mean matching, \pkg{imputeTS} [@imputeTS], which provide time series imputation methods, and \pkg{missMDA} [@missMDA], which imputes data using principal components analysis.
+
+A key feature of the design of \pkg{simputation} is that it returns a regular dataframe.
+This is different from other imputation packages, which return custom classes of data or additional objects.
+These create extra work when preparing for subsequent analysis.
 Providing a dataframe with imputed values included in the data reduces the friction of working with other tools, but comes at the cost identifying imputed values.
-Other software for imputation includes \pkg{Hmisc} [@Hmisc], which provide predictive mean matching, \pkg{imputeTS} [@imputeTS], which provide time series imputation methods, and \pkg{missMDA} [@missMDA], which imputes data using principal components analysis.
 
@@
 Popular and robust methods for multiple imputation include the \pkg{mice}, \pkg{Amelia}, and \pkg{mi} packages [@mice; @amelia; @mi].
-The \pkg{mice} package implements the method of chained equations, using a variable-wise algorithm to calculate the posterior distribution of parameters, from which uses to generate imputed values.
-The suggested workflow in \pkg{mice} revolves around imputing data, returning completed data, and fitting a model and pooling the results.
+The \pkg{mice} package implements the method of chained equations, using a variable-wise algorithm to calculate the posterior distribution of parameters to generate imputed values.
+The workflow in \pkg{mice} revolves around imputing data, returning completed data, and fitting a model and pooling the results.
+
 The \pkg{Amelia} package [@amelia] assumes data are multivariate normal, and samples from the posterior using the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm [@Honaker2010] and allows for incorporation of information on the values in a prior.
+
 The \pkg{mi} package [@mi] also uses Bayesian models for imputation, providing better handling of semi-continuous values, and data with structural or perfect correlation.
@@
 This approach promotes fluid workflow, with a similar penalty to tidying up model output, which is still messy.
+
 Another software for performing multiple imputation is \pkg{norm} [@norm; @schafer-norm], which uses methods from the NORM software from @schafer-norm, such as multiple imputations using EM for multivariate normal data.
 The \pkg{norm} package does not provide a framework for tracking missing values, but instead provides tools for making inference from multiple imputation.
-Each of these multiple imputation methods provide practical methods, but do not have consistent interfaces across their methods or implementation, which makes then inherently messy, as their output is always different.
-A consistent data structure that tracks missing values by default would make the data outputs tidy, streamlining subsequent analysis.
+
+Each of these multiple imputation methods provide practical methods, but all have different output structures, and do not have consistent interfaces in their implementation.
+This makes then inherently messy and challenging to integrate into an analysis pipeline.
+An ideal approach would have consistent data structures that work with other analysis tools, and track missing values by default.
+This would make the data outputs tidy, streamlining subsequent analysis.
 
@@
 
-The primary focus of most missing data packages is exploring imputed values, not
+The primary focus of most missing data packages is making inferences, and exploring imputed values, not
 on exploring relationships in missing values, and identifying possible patterns.
-Texts that do cover exploration of missing data have the same problem as with modelling: the input is tidy, but the output is messy and does not work with other tools [@VanBuuren2012]; this is inefficient.
+Texts that do cover exploration of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools [@VanBuuren2012]; this is inefficient.
 Methods for exploring missing values are primarily covered in literature on interactive graphics [@Swayne1998; @Unwin1996; @Cook2007], and are picked up again in a discussion of a graphical user interface [@Cheng2015].
 
-Missing values in a dataset can be represented as a binary matrix where 0 is present and 1 is missing [@Rubin1976; @Schafer2002].
-This is often referred to as a matrix $R$, which can be used to assess missing data dependence.
-This idea of a matrix of missing values has also been used in interactive graphics, dubbed a "shadow matrix", to link missing and imputed values to the data, facilitating their display in graphics [@Swayne1998].
+The missingness matrix $R$ can be used to assess missing data dependence, and has  been used in interactive graphics, dubbed a "shadow matrix", to link missing and imputed values to the data, facilitating their display in graphics [@Swayne1998].
 This work also focuses heavily on multivariate numeric data.
@@
 
-The approach of @Unwin1996 focussed on multivariate categorical data, where missingness is explicitly added as an additional category.
+The approach of the MANET software in @Unwin1996 focussed on multivariate categorical data, where missingness is explicitly added as an additional category.
 MANET also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot.
+The MANET software is no longer maintained and cannot be installed.
 The approach of @Swayne1998 in the software XGobi, further developed in ggobi [@Cook2007], focussed more on multivariate quantitative data.
@@
 MissingDataGUI provides a user interface for exploring missing data structure both numerically and visually.
-Using a GUI to explore missing data may make it easier to glean valuable insight into important structures for missingness.
-However, these insights are typically not captured or recorded, so it is difficult to incorporate them into reproducible analyses.
+Using a Graphical User Interface (GUI) to explore missing data may make it easier to glean valuable insight into important structures for missingness.
+However, this comes as a trade off with GUIs - the insights are typically not captured or recorded, so it is difficult to incorporate them into reproducible analyses.
 This distracts and breaks analysis workflow, inviting mistakes.
 
-VIM (visualizing and Imputing Missing Data) provides visualization methods that identify observed, imputed, and missing values.
+VIM (Visualizing and Imputing Missing Data) provides visualization methods that identify observed, imputed, and missing values.
 These include spinograms, spinoplots, missingness matrices, plotting missingness in the margins of other plots, and other summaries.
@@
 If there are missings in a continuous variable like test score, \pkg{ggplot2} omits the missing values and prints a warning message.
-However, if a discrete variable like school year has missing values, an NA category is created for school year, and the scores are placed into this.
-
-Figure \@ref(fig:gg-box-na), shows four scenarios of missingness for test scores across school year - the first contains no missingness, the second contains missingness for school year, the third contains missingness in scores, and the fourth contains missingness in both.
+However, if a discrete variable like school year has missing values, an NA category is created for school year, and the scores are placed into this (Figure \@ref(fig:gg-box-na)).
 
@@
           p4,
-          nrow = 2,
+          nrow = 1,
           labels = LETTERS[1:4])

### 2020-01-06 paper/tidy-missing-data-paper.Rmd
ref cheng GUI paper
update data structures section

@@
 
-The graphics literature [@Swayne1998; @Unwin1996; @VIM; @MissingDataGUI; ] provides several solutions for exploring missings visually.
+The graphics literature [@Swayne1998; @Unwin1996; @VIM; @Cheng2015] provides several solutions for exploring missings visually.
 These approaches incorporate missings into the plot in some way.
@@
 when working with missing data. The design and naming of these features is an
-important component that helps the analyst compose them efficiently in data pipelines, and
-is discussed throughout.
+important component that helps the analyst compose them efficiently in data pipelines, and is discussed throughout.
 
@@
 To explore missing data there needs to be a good representation of missing
-values that integrate into analysis pipelines. A common representation of
-missing values is a binary matrix the same dimension as the data, where 0 and 1
-indicate not missing and missing. This is typically referred to as a matrix $R$,
-used to describe missingness dependence patterns, (see for example @Rubin1976
-and @Schafer2002). This matrix is used to explore missing values in the
-interactive graphics library XGobi, where it was called a "shadow matrix",
-defined as a copy of the original data with indicator values of missingness:
-"NA" and "!NA" for missing and not missing, represented as 0 and 1 (see Figure
-\@ref(fig:shadow-matrix-progress) parts 2 and 3 below).
+values that integrate into analysis pipelines. The R matrix, where 0 and 1
+indicate not missing and missing provides a template, as it is very common in
+missing data literature.  This matrix has been used to explore missing values in
+the interactive graphics library XGobi, where it was called a "missing value
+shadow", or "shadow matrix", defined as a copy of the original data with
+indicator values of missingness, where 1s represent missing values, and 0
+represents complete values. The values 0 and 1 (or equally, FALSE and TRUE) can
+be confusing, however, as it is not immediately clear what they represent - does
+0 indicate a the absence of observation, or the presence of a missing value? The
+shadow matrix was specially linked to the data, but could not be referred to
+directly, instead they were linked with special window views. This meant that
+the data structure was abstracted from the user to facilitate data exploration.
+This made sense at the time with the state of the art for graphicaly analysis,
+but that data structure needs to be clearly presented to the user so that they
+can more broadly use it in other data analyses.
 

### 2020-01-06 paper/tidy-missing-data-paper.Rmd
relabel from 3 features to 4

@@
 
-Three features are added to the shadow matrix to facilitate analysis:
+Four features are added to the shadow matrix to facilitate analysis:
 
@@
 
-2. Special missing values: Values in the shadow matrix can be "special" missing values, indicated as `NA_suffix`, where "suffix" is a very short message of the type of missings.
-
-3. Cohesiveness: Binding the shadow matrix column-wise to the original data creates a cohesive _nabular_ data form, useful for visualization and summaries.
+1.
+Missing value label: A simple label to the shadow matrix clear, "NA" and "!NA" for missing and not missing, represented as 0 and 1 (see Figure \@ref(fig:shadow-matrix-progress) parts 2 and 3 below).
+
+1.
+Special missing values: Building on 2, the values in the shadow matrix can be "special" missing values, indicated as `NA_suffix`, where "suffix" is a very short label of the type of missings, and the underlying values are >= 2.
+
+1. Cohesiveness: Binding the shadow matrix column-wise to the original data creates a cohesive _nabular_ data form, useful for visualization and summaries.
 
 These additions are now discussed in turn.
+
+(**consider removing these longer expansions of these points?**)
 

### 2020-01-06 paper/tidy-missing-data-paper.Rmd
update missing value labels, special missings, and nabular data

@@
 It also makes a clear distinction that `var_NA` is a random variable of the missingness of a variable, `var`.
-This subtle change is important as it changes the focus from the value of a variable to it's missingness state, making intent clear when performing analysis.
+This subtle change is important as it shifts the focus from the value of a variable to its missingness state, making intent clear when performing analysis.
+
+### Missing value labels {#missing-labels}
+
+The $R$ matrix is a very common representation of missing values, and can be easily created in the "R" programming language with \code{is.na(data)}.
+The values 0 and 1 (or equally, FALSE and TRUE) can be confusing, however, as it is not immediately clear what they represent - does 0 indicate a the absence of observation, or the presence of a missing value?
+This is solved by adding a simple label to the shadow matrix: "NA" and "!NA" for missing and not missing, represented as 0 and 1 (see Figure \@ref(fig:shadow-matrix-progress) parts 2 and 3 below).
 
@@
 Values in the shadow matrix are "factor" type values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing.
-Extending these to include "special" missing values labelled as "NA_\<suffix\>" allows for representing additional features such as instrument failure and drop-out as "NA_instr_fail", and "NA_drop_out".
-The underlying number representation changes from binary to integer.
+Extending these to include "special" missing values labelled as "NA_\<suffix\>" allows for representing additional missingness features such as instrument failure and drop-out as "NA_instr_fail", and "NA_drop_out".
+The underlying number representation changes from binary to integer, allowing for unlimited missingness features.
 For example, instrument failure and drop out would be represented as 2 and 3, see Table \@ref(tab:shadow-encoding) for an example.
 Encoding these special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in \pkg{naniar}, shown in Section \@ref(verbs-recode).
+Special missing values do exist in other statistical programming languages, although are limited to 26 or 52 special types of missingness.
 
@@
 _Nabular_ data binds the shadow matrix column-wise to the original data, and is so named as a portmanteau of `NA` and `tabular`.
-_Nabular_ data keeps corresponding rows together, removing the possibility of mismatching records, and explicitly links missing values to the data.
+_Nabular_ data keeps corresponding rows together, removing the possibility of mismatching records, explicitly linking missing values to the data.
 _Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable, `var`, as `var_NA`.

### 2020-01-06 paper/tidy-missing-data-paper.Rmd
minor text changes

@@
 
-A data structure for missing data provides key benefits over on-the-fly creating a logical matrix where TRUE means missing, and FALSE means present.
+The _nabular_ data structure for missing data provides key benefits over on-the-fly creating a logical matrix where TRUE means missing, and FALSE means present.
 Firstly, it can be confusing to users whether TRUE or FALSE means missing or present; `NA` and `!NA` are clearer.
@@
 Here they suggest to indicate censored data with an extra variable called "VariableNameCensored", which would be TRUE if censored, otherwise FALSE.
-This information is now represented in the shadow columns as additional integers, along with the other missing information.
+This information can now be represented in the shadow columns as special missings.
 
@@
 The number of times certain variables go missing together can be visualized using an "upset plot" [@Conway2017], a visualization technique for showing the size and features of sets in data that is similar to a venn diagram, but scales better with more variables (Figure \@ref(fig:airquality-upset)).
+Figure \@ref(fig:airquality-upset) shows that only Ozone and Solar.R have missing values.
+The dots in the bottom right show the combinations of missingness, with the bottom left showing the missingness in each variable, and the top showing the missingness for the combinations.
+The bottom left shows that Ozone has the most missing values, and the top shows 2 cases where both Solar.R and Ozone have missing values together.
 
@@
 ```
-
-Figure \@ref(fig:airquality-upset) shows that only Ozone and Solar.R have missing values.
-The dots in the bottom right show the combinations of missingness, with the bottom left showing the missingness in each variable, and the top showing the missingness for the combinations.
-The bottom left shows that Ozone has the most missing values, and the top shows 2 cases where both Solar.R and Ozone have missing values together.
 
@@
 To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed values below the range of data using `impute_below` (by default imputing 10% below the range).
-Visualizing this as a histogram shows missing values on its left.
+Visualizing this as a histogram shows the number of missing values on its left.
 Ozone is shown in a different color by referring to the variable `Ozone` as `Ozone_NA` in a fill aesthetic in ggplot, shown in Figure \@ref(fig:impute-shift-histogram).

### 2020-01-06 paper/tidy-missing-data-paper.Rmd
fix table function ordering

@@
                          "pct_miss",
-                         "pct_miss_var",
-                         "pct_miss_case"),
+                         "pct_miss_case",
+                         "pct_miss_var"),
   "missing value" = c(n_miss(airquality),
@@
                           "pct_complete",
-                          "prop_complete_var",
-                          "pct_complete_case"),
+                          "prop_complete_case",
+                          "pct_complete_var"),
   "complete value" = c(n_complete(airquality), 
@@
                        pct_complete(airquality),
-                       prop_complete_case(airquality), 
+                       pct_complete_case(airquality), 
                        pct_complete_var(airquality))

### 2020-01-06 paper/tidy-missing-data-paper.Rmd
reference example data in text
add inline code into tables

@@
 
-```{r gg-box-na, fig.height = 6, fig.width = 6, out.width = "70%", fig.align = "center", fig.cap = "ggplot2 provides different visualizations depending on what type of data has missing values.
+```{r gg-box-na, fig.height = 3, fig.width = 9, out.width = "100%", fig.align = "center", fig.cap = "ggplot2 provides different visualizations depending on what type of data has missing values.
 (A) Data is complete and graphic is presented; (B) Missings are only in category variable year - an additional 'NA' boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and school year, additional missing information is shown.
@@
 
-```{r vis-miss, fig.height = 4, fig.width = 6, out.width = "95%", fig.show = 'hold', echo = FALSE, dev = "png", dpi = 300, fig.cap = "Heatmap visualizations of missing data for the airquality dataset.
+```{r vis-miss, fig.height = 3, fig.width = 4.5, out.width = "80%", fig.show = 'hold', echo = FALSE, dev = "png", dpi = 300, fig.cap = "Heatmap visualizations of missing data for the airquality dataset.
 (A) The default output and (B) ordered by clustering on rows and columns.
@@
   knitr::kable(
-    caption = "Summary of the number and percent of missings in each variable. Only ozone and solar radiation have missing values.",
+    caption = "The \\texttt{miss\\char`_var\\char`_summary} function provides a summary of the number and percent of missings in each variable. Only ozone and solar radiation have missing values.",
     digits = 1)
@@
   knitr::kable(
-    caption = "Tabulation of the amount of missing data in each variable.
+    caption = "The \\texttt{miss\\char`_var\\char`_table} function provides a tabulation of the amount of missing data in each variable.
 This shows the number of variables with no missings, 7 missings, and 37 missings, and the percentage of variables with those amounts of missingness.
@@
   slice(1:10) %>%
-  knitr::kable(caption = "A summary of the missingness in each variable, for each Month of the airquality dataset, with only the first 10 rows of output shown.
+  knitr::kable(caption = "The \\texttt{miss\\char`_var\\char`_summary} provides a grouped summary of the missingness in each variable, for each Month of the airquality dataset, with only the first 10 rows of output shown.
 There are more ozone missings in June than May.",
@@
 The  `miss_scan_count()` function finds where specified values occur in data.
-For example, for the example data, `dat_ms`:
+For example, for the example data, `dat_ms`  Table \@ref(tab:setup-miss-scan-count).
+All occurrences of -99 can be searched for using the code:
 
@@
 ```
-
-All occurrences of -99 can be searched for using the code:
 
@@
 
+(**considering removing these examples**)
+
 For flexibility, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`.
@@
 `naniar` does not reinvent the wheel for imputation, as there are many robust R packages for imputation, such as `mice`, `mi`, `VIM`, and `simputation`.
-However, `naniar` does provide a few imputation methods to facilitate exploration and visualizations, which were not otherwise available: `impute_below` and `impute_mean` and `impute_median`.
+However, `naniar` does provide a few imputation methods to facilitate exploration and visualizations, which were not otherwise easily available: `impute_below` and `impute_mean` and `impute_median`.
 These imputation methods are useful to explore structure in missingness, but are not recommended for use in analysis.
 
-The `impute_below` function imputes values below the minimum of the data, with some jitter to reduce overplotting, but has options to change the default amount it is shifted below, and the amount of jitter.
-
-```{r demonstrate-impute-below}
+The `impute_below` function imputes values below the minimum of the data, with some jitter (small values of random noise) to reduce overplotting, but has options to change the default amount it is shifted below, and the amount of jitter.
+
+```{r demonstrate-impute-below, eval = FALSE}
 airquality %>%
@@
 
-One drawback to imputation functions in this form is that the location of imputed values is not tracked. This is covered in Section \@ref(verbs-track).
+One drawback to imputation functions in this form is that the location of imputed values is not tracked. This issue is resolved with \code{nabular} data covered in Section \@ref(verbs-track).
 
@@
   bind_shadow() %>%
-  as.data.frame() %>%
-  simputation::impute_knn(Ozone ~ .) %>%
-  simputation::impute_knn(Solar.R ~ .) %>%
-  add_label_shadow() %>%
-  as_tibble()
-
+  simputation::impute_lm(Ozone ~ Temp + Wind) %>%
+  simputation::impute_lm(Solar.R ~ Temp + Wind) %>%
+  add_label_shadow()
+
+aq_imputed
 ```
@@
 Missing values can then be shown in a scatterplot by setting the `color` aesthetic in ggplot to `any_missing` (Figure \@ref(fig:track-impute-example)A), or in a density plot looking at one variable, using the `fill = any_missing`, (Figures \@ref(fig:track-impute-example)B) and \@ref(fig:track-impute-example)C).
-```r
-ggplot(aq_imputed,
-       aes(x = Ozone,
-           y = Solar.R,
-           color = any_missing)) + 
-  geom_point() +
-  scale_color_brewer(palette = "Dark2") +
-  theme(legend.position = "bottom")
-
-ggplot(aq_imputed,
-       aes(x = Ozone,
-           fill = any_missing)) + 
-  geom_density(alpha = 0.3) + 
-  scale_fill_brewer(palette = "Dark2") +
-  theme(legend.position = "none")
-
-ggplot(aq_imputed,
-       aes(x = Solar.R,
-           fill = any_missing)) + 
-  geom_density(alpha = 0.3) + 
-  scale_fill_brewer(palette = "Dark2")
-```
 
 ```{r track-impute-example, fig.show = "hold", fig.cap = "Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values imputed using simputations `impute\\_knn` function, with imputed values colored green and data values orange.
-Imputed values are similar, but slightly different to existing data.", fig.height = 2, fig.width = 5, out.width = "100%", echo = FALSE}
-
+Imputed values are similar, but slightly different to existing data.", fig.height = 4, fig.width = 12, out.width = "100%", echo = FALSE}
+
+p1 <- 
 ggplot(aq_imputed,
@@
 
+p2 <- 
 ggplot(aq_imputed,
@@
 
+p3 <- 
 ggplot(aq_imputed,
@@
 
+gridExtra::grid.arrange(p1, p2, p3, nrow = 1)
 ```
@@
   summarise_at(.vars = vars(Ozone),
-               .funs = list(min, mean, median, max))
+               .funs = lst(min, mean, median, max))
 ```
@@
 
-```{r applic-vis-miss, fig.height = 5, fig.width = 6.25, out.width = "85%", fig.show='hold', fig.cap = "Heatmap of clustered missingness for the housing data.
+```{r applic-vis-miss, fig.height = 4, fig.width = 6, out.width = "85%", fig.show='hold', fig.cap = "Heatmap of clustered missingness for the housing data.
 Three groups of missingness are apparent, at the top for building area to longitude, the middle for building area and year built, at the end for building area, year built, and landsize.
@@
 
-The new data structures discussed in the paper could be used to create different visualizations than were shown in the paper.
+The new data structures discussed in the paper promote flexibility, and could be used to create different visualizations than were shown in the paper.
 The analyst can use the data structures to decide on appropriate visualization for their problem.

### 2020-01-06 paper/tidy-missing-data-paper.Rmd
add inline code into tables

--- 2020-01-06 paper/tidy-missing-data-paper.Rmd
@@
 The  `miss_scan_count()` function finds where specified values occur in data.
-For example, for the example data, `dat_ms`:
+For example, for the example data, `dat_ms`  Table \@ref(tab:setup-miss-scan-count).
 
@@
 ```
+
+(**considering removing these examples**)
 

### 2020-01-06 paper/tidy-missing-data-paper.Rmd
removed many unnecessary instances of "that" and general grammar clean up.

--- 2020-01-06 paper/tidy-missing-data-paper.Rmd
@@
 abstract: >
-  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature on how to make it easy to handle, explore, and impute missing values in data.
+  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data.
 This paper addresses this gap.
-The new methodology builds upon tidy data principles, with a goal to integrating missing value handling as an integral part of data analysis workflows.
-New data structures are defined along with new functions (verbs) to perform common operations.
-Together these provide a cohesive framework for handling, exploring, and imputing missing values.
+The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as an integral part of data analysis workflows.
+We define both new data structures and new verbs (functions) to perform common operations.
+Together, these provide a cohesive framework for handling, exploring, and imputing missing values.
 These methods have been made available in the R package `naniar`.
@@
 
-Tidy data [@Wickham2014] is a relatively new principle and suite of tools that
-facilitate the process of converting raw data into a clean, analysis-ready
-format that works efficiently in a data analysis pipeline. In tidy data, missing
-values can be handled implicitly or explicitly, at the analyst's discretion.
-However, when making a plot, missing values are simply dropped, albeit with a
-warning (see Figure \@ref(fig:warning)). Most other analysis software will
-simply drop cases with missings without warning.
+As data science has become a more solid field, theories and principles have developed to describe best practices.
+One such idea is 'tidy data,' a set of guidelines for converting raw data into a clean, analysis-ready format that works efficiently in a data analysis pipeline, while another is the idea of the Grammar of Graphics, which describes how to map data values into a visualization [@Wickham2014, Wilkinson2012].
+These principles have guided analysts in their work for several years now.
+However, while these principles outline how to work with data in a general way, they do not address the problem of missing data.
+In particular, there is little to no guidance for working with missing values in the field of visualization.
+Most analysis software simply drops missing values when making a plot, although some packages provide a warning (see Figure \@ref(fig:warning)).
+
+<!-- Tidy data [@Wickham2014] is a relatively new principle and suite of tools -->
+<!-- facilitating the process of converting raw data into a clean, analysis-ready -->
+<!-- format that works efficiently in a data analysis pipeline. In tidy data, missing -->
+<!-- values can be handled implicitly or explicitly, at the analyst's discretion. -->
+<!-- However, when making a plot, missing values are simply dropped, albeit with a -->
+<!-- warning (see Figure \@ref(fig:warning)). Most other analysis software will -->
+<!-- simply drop cases with missings without warning. -->
 
@@
 The imputation literature (e.g. @Little1988, @Rubin1976, @Simon1986,
-@Schafer2002, @VanBuuren2012) does not address how to handle missing data.
+@Schafer2002, @VanBuuren2012) does not address how to explore, understand, and handle
+missing data structures and mechanisms.
 Instead the focus is on ensuring valid statistical inference is made from
@@
 a complete dataset for analysis. Whether by case or variable deletion, or with
-imputation, it is ideal to know what the mechanism is that generates missing 
+imputation, it is ideal to identify the mechanism generating missing 
 values.
@@
 While there might be other times where the data are
-clearly missing completely at random (or missing at random), determining this is not always straightforward. There is no silver bullet for missing data that will reveal
-it's mechanism to the analyst. Missing data must be explored.
-
-While there have been many software tools for exploring missing data (VIM, mi, mice, MissingDataGUI, MANET, ggobi, norm), these tools generate outputs that make them challenging to integrate into analysis pipelines.
-Despite it being crucial to explore and understand missing data, there is very little research on the concepts that make
-this process more efficient, and integrate well into analysis pipelines.
+clearly missing completely at random (or missing at random), determining this is not always straightforward. There is no silver bullet to reveal missing data
+mechanisms to the analyst. Missing data must be explored.
+
+While there have been many software tools for exploring missing data (VIM, mi, mice, MissingDataGUI, MANET, ggobi, norm), these tools generate outputs challenging to integrate into analysis pipelines.
+Despite it being crucial to explore and understand missing data, there is very little research on how to make this process more efficient and integrate it into the analysis pipeline.
 
@@
 The approaches from the graphics literature have not been translated and integrated into tidy data.
-This limits the analysts capacity to efficiently explore missing values.
+This limits the analyst's capacity to efficiently explore missing values.
 Extending tidy data and tidy tools to account for missing values would create more efficient and robust workflows and analysis.
@@
 Missing value imputation (\@ref(missing-data-rep-dep) and \@ref(imputation)) and exploration literature (\@ref(exploration)) are then discussed.
-Section \@ref(extensions) discusses extensions to the tidy methodology that facilitates operations for exploring, visualizing, and imputing missing data.
+Section \@ref(extensions) discusses tidy methodology extensions, which facilitate operations for exploring, visualizing, and imputing missing data.
 The application of these new methods is illustrated using a case study in section \@ref(case-study).
@@
 This helps the analyst works swiftly and clearly, closing up opportunities for errors.
+Tidy data principles are general, but are most robustly implemented in the R programming language, so the
+paper focusses on R.
 
@@
 
-Data manipulation is made input and output tidy with R packages \pkg{dplyr} and \pkg{tidyr} [@dplyr; @tidyr].
+Data manipulation is made input- and output-tidy with R packages \pkg{dplyr} and \pkg{tidyr} [@dplyr; @tidyr].
 These provide the five "verbs" of data manipulation: data reshaping, sorting, filtering, transforming, and aggregating.
@@
 For example, estimated coefficients, predictions, and residuals from one model cannot be easily combined with the output of another model.
-The \pkg{recipes} package is a tidy tool under development to help make modelling input and output tidy [@recipes].
+This output problem has been partially addressed with the \pkg{broom} package, which tidies up model outputs for data analysis, and the developing \pkg{recipes} package, which helps make modelling input and output tidy [@recipes].
 
@@
 These are designed to share similar principles in their design and behavior, and cover the breadth of an analysis - going from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @tidyverse; @Tidyverse-Manifesto].
-This has led to a growth of tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr}, dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} [@readr; @readxl; @haven; @stringr; @lubridate; @purrr].
-It has also led to the growth of new packages for other fields that follow similar design principles and create fluid workflows for new domains.
+This has led to more tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr} and dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} [@readr; @readxl; @haven; @stringr; @lubridate; @purrr].
+It has also led to a burgeoning of new packages for other fields following similar design principles and create fluid workflows for new domains.
 For example, the \pkg{tidytext} package for text analysis, the \pkg{tsibble} package for time series data, and \pkg{tidycensus} for working with US census and boundary data [@tidytext; @wang2020tsibble; @tidycensus].
@@
 
-Past graphics research on missing data has not yet been framed in a tidy way.
+Graphics research on missing data has not yet been framed in a tidy way.
 The graphics literature has explored methods for missing value exploration and imputation.
 Translating and expanding these to fit within tidy data and tidy tools would create more efficient workflows.
-By laying down these concepts, it allows for the focus to be more general than just software, but rather, a framework for tidy tools to explore missing data that can be built upon, extended, and implemented in other languages.
+By laying down these concepts, it allows the focus to be more general than just software, but rather, an extensible framework for tidy tools to explore missing data.
 
@@
 
-**MCAR** is where values being missing have no association with observed or unobserved data, that is $Pr(R = 1 | Y_{obs}, Y_{miss}) = Pr(R = 1 | \psi)$.
+**MCAR** is where values being missing have no association with observed or unobserved data, that is, $Pr(R = 1 | Y_{obs}, Y_{miss}) = Pr(R = 1 | \psi)$.
 Essentially, the probability of an observation being missing is unrelated to anything else, only the parameters $\psi$, which is the overall probability of missingness.
@@
 This assumes that conditioning on all available observed data, the data goes missing due to some phenomena unobserved, including the structure of the missing data itself.
-This presents a challenge in analysis, as it is difficult to verify this state, and also implies bias in analysis due to the unobserved phenomena.
+This presents a challenge in analysis, as it is difficult to verify, and implies bias in analysis due to the unobserved phenomena.
 
@@
 
-\pkg{VIM} [@VIM] efficiently implements well-used imputation methods k-nearest neighbor, regression, hot-deck, and iterative robust model-based imputation.
-This diversity of approaches allow the package to deal semi-continuous, continuous, count, and categorical data.
+\pkg{VIM} [@VIM] efficiently implements well-used imputation methods K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation.
+This diversity of approaches allows the package to deal with semi-continuous, continuous, count, and categorical data.
 VIM identifies imputed cases by adding an indicator variable with a suffix `_imp`.
@@
 These create extra work when preparing for subsequent analysis.
-Providing a dataframe with imputed values included in the data reduces the friction of working with other tools, but comes at the cost identifying imputed values.
+Providing a dataframe with imputed values included in the data reduces the friction of working with other tools, but comes at the cost of identifying imputed values.
 
@@
 on exploring relationships in missing values, and identifying possible patterns.
-Texts that do cover exploration of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools [@VanBuuren2012]; this is inefficient.
+Texts covering exploration of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools [@VanBuuren2012]; this is inefficient.
 Methods for exploring missing values are primarily covered in literature on interactive graphics [@Swayne1998; @Unwin1996; @Cook2007], and are picked up again in a discussion of a graphical user interface [@Cheng2015].
@@
 
-VIM (Visualizing and Imputing Missing Data) provides visualization methods that identify observed, imputed, and missing values.
+VIM (Visualizing and Imputing Missing Data) provides visualization methods to identify observed, imputed, and missing values.
 These include spinograms, spinoplots, missingness matrices, plotting missingness in the margins of other plots, and other summaries.
@@
 understand missingness. Section \@ref(num-sum) discusses data summaries, and
-Section \@ref(verbs) covers the common "verbs" used to solve problems that arise
+Section \@ref(verbs) covers the common "verbs" used to solve problems
 when working with missing data. The design and naming of these features is an
@@
 
-To explore missing data there needs to be a good representation of missing
+To explore missing data there needs to be a robust representation of missing
 values that integrate into analysis pipelines. The R matrix, where 0 and 1
@@
 indicator values of missingness, where 1s represent missing values, and 0
-represents complete values. The values 0 and 1 (or equally, FALSE and TRUE) can
-be confusing, however, as it is not immediately clear what they represent - does
+represents complete values. The values 0 and 1 can
+be confusing representations of missing values - does
 0 indicate a the absence of observation, or the presence of a missing value? The
 shadow matrix was specially linked to the data, but could not be referred to
-directly, instead they were linked with special window views. This meant that
+directly, and was instead linked with special window views. This means
 the data structure was abstracted from the user to facilitate data exploration.
-This made sense at the time with the state of the art for graphicaly analysis,
-but that data structure needs to be clearly presented to the user so that they
-can more broadly use it in other data analyses.
+This made sense at the time with the state of the art for graphical analysis,
+but the data structure needs to be clearly presented to the user so they
+can use it more flexibly in other data analyses.
 
@@
 The suffix is short and easy to remember in analysis or visualization.
-It also makes a clear distinction that `var_NA` is a random variable of the missingness of a variable, `var`.
+It also makes a clear distinction with `var_NA` being a random variable of the missingness of a variable, `var`.
 This subtle change is important as it shifts the focus from the value of a variable to its missingness state, making intent clear when performing analysis.
@@
 The temperature value of -99 is represented in the shadow column as 'NA\\_instr', which in turn has the underlying numeric value of 2.
-This captures additional information about the data that is otherwise difficult to record.")
+This captures additional information about the data otherwise difficult to record.")
              # booktabs = TRUE) 
@@
 _Nabular_ data is created using the command `bind_shadow()` or `nabular()` (see Figure \@ref(fig:nabularfig)).
-_Nabular_ data is not unlike classical data formats that have quality or flag columns associated with each measured variable, e.g.
+_Nabular_ data is not unlike classical data formats with quality or flag columns associated with each measured variable, e.g.
 Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].
@@
 Visualizing missingness in a dataset is essential to understand its structure.
-This section discusses visualizations that are easy to remember and compose with other functions in a pipeline.
+This section discusses easy to remember visualizations that compose with other functions in a pipeline.
 Four different areas of visualization of missing data are discussed: overviews (Section \@ref(overview-plots)); univariate (Section \@ref(univariate-plots)), bivariate (Section \@ref(bivariate-plots)), and multivariate (Section \@ref(multivariate-plots)).
@@
 Similar visualizations are available in other missing data packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}.
-A key improvement is that `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns in the visualization and are named at the top, and each row is an observation.
+A key improvement is `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns in the visualization and are named at the top, and each row is an observation.
 Using \pkg{ggplot2} means the plot can be customized and combined with other ggplot graphics.
@@
 
-The number of times certain variables go missing together can be visualized using an "upset plot" [@Conway2017], a visualization technique for showing the size and features of sets in data that is similar to a venn diagram, but scales better with more variables (Figure \@ref(fig:airquality-upset)).
-Figure \@ref(fig:airquality-upset) shows that only Ozone and Solar.R have missing values.
+The number of times certain variables go missing together can be visualized using an "upset plot" [@Conway2017], a visualization technique for showing the size and features of sets in data similar to a venn diagram, but scales better with more variables (Figure \@ref(fig:airquality-upset)).
+Figure \@ref(fig:airquality-upset) shows only Ozone and Solar.R have missing values.
 The dots in the bottom right show the combinations of missingness, with the bottom left showing the missingness in each variable, and the top showing the missingness for the combinations.
-The bottom left shows that Ozone has the most missing values, and the top shows 2 cases where both Solar.R and Ozone have missing values together.
+The bottom left shows Ozone has the most missing values, and the top shows 2 cases where both Solar.R and Ozone have missing values together.
 
@@
 airquality %>%
-  bind_shadow() %>%
+  nabular() %>%
   impute_below_all() %>%
@@
 airquality %>%
-  bind_shadow() %>%
+  nabular() %>%
 ggplot(aes(x = Temp)) + 
@@
 airquality %>%
-  bind_shadow() %>%
+  nabular() %>%
 ggplot(aes(x = Temp,
@@
 Figure \@ref(fig:parallel-cord-plot) shows a parallel coordinate plot of the `oceanbuoys` data, with missing values imputed to be 10% below the range, and values colored according to whether humidity contained missing values.
-Although a 10% below imputation is not the most ideal way to display missing values in a parallel coordinate plot, it shows that humidity is missing at low air and sea temperatures, and that humidity is missing in one year, and one location.
+Although a 10% below imputation is not the most ideal way to display missing values in a parallel coordinate plot, it shows humidity is missing at low air and sea temperatures, and humidity is missing in one year, and one location.
 
@@
          ID = as.factor(ID)) %>%
-  bind_shadow() %>%
+  nabular() %>%
   impute_below_all() %>%
@@
 Complementing these are functions to reveal the number of complete values: `n_complete`, `prop_complete`, and `pct_complete`.
-These be extended to summarize the number/amount of variables and cases that contain a missing values by appending `_case` or `_var` to any of these summaries.
+These be extended to summarize the number/amount of variables and cases containing a missing values by appending `_case` or `_var` to any of these summaries.
 Table \@ref(tab:n-prop-pct-miss-complete) shows the names and output of these functions on the dataset airquality.
@@
 
-Presenting the number and percent of missing values for each variable, or case, provides a summary that can be used by models to inform imputations or data handling.
+Presenting the number and percent of missing values for each variable, or case, provides a summary usable in models to inform imputations or data handling.
 For example, potentially dropping variables or deciding to include others in an imputation model.
@@
 
-There are many common values that mean "missing".
+There are many common values for "missing".
 For example, "N/A" "Not Available", "MISSING", or white space " ".
@@
 For flexibility, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`.
-This means `replace_with_na` works for **all** variables with `replace_with_na_all`, **at** selected variables with `replace_with_na_at`, and **if** variables that meet some condition with `replace_with_na_if`.
+This means `replace_with_na` works for **all** variables with `replace_with_na_all`, **at** selected variables with `replace_with_na_at`, and **if** variables meet some condition with `replace_with_na_if`.
 The syntax for these is slightly different to `replace_with_na`, but powerful:
@@
 This facilitates visualization and summaries, and allows for imputed values to be appropriately tracked.
-_Nabular_ data can be created with `bind_shadow()` or `nabular()`.
-The `bind_shadow()` function is so named to borrow from existing functions named `bind` - such as `bind_rows()` and `bind_cols()` from `dplyr`, which bind rows or columns into a dataframe.
-An example use of `bind_shadow()` on the example data, `dat_ms` is shown below:
+_Nabular_ data can be created with `nabular()` or `bind_shadow()`.
+The function was named `bind_shadow()` to borrow from existing functions named `bind` - such as `bind_rows()` and `bind_cols()` from `dplyr`, which bind rows or columns into a dataframe.
+An example use of `nabular()` on the example data, `dat_ms` is shown below:
 
 ```{r bind-shadow, echo = TRUE}
-bind_shadow(dat_ms)
+nabular(dat_ms)
 ```
@@
 Using `recode_shadow()` requires specifying the variable you want to contain the flagged value, the condition you want this to occur at, and a suffix for the new type of missing value.
-This is then recoded as a new factor level in the shadow matrix, so that every column is aware of all possible new values of missingness.
+This is then recoded as a new factor level in the shadow matrix, so every column is aware of all possible new values of missingness.
 For example, the values -99 and -98 could be recoded to mean a broken machine sensor for the variable x:
@@
 dat_ms %>%
-  bind_shadow() %>%
+  nabular() %>%
   recode_shadow(x = .where(x == -99 ~ "broken_sensor"))
@@
 
-One drawback to imputation functions in this form is that the location of imputed values is not tracked. This issue is resolved with \code{nabular} data covered in Section \@ref(verbs-track).
+One drawback to imputation functions in this form is imputed value locations are not tracked. This issue is resolved with \code{nabular} data covered in Section \@ref(verbs-track).
 
@@
 
-To evaluate imputations they need to be tracked, which is achieved by combining the verbs `bind_shadow`, `impute_`, and `add_label_shadow`.
+To evaluate imputations they need to be tracked, which is achieved by combining the verbs `nabular`, `impute_`, and `add_label_shadow`.
 The missing values can then be referred to by their shadow variable, `_NA`.
 The missingness of any observation can be referred to with `any_missing` (Figure \@ref(fig:track-impute-example)).
-The code chunk below shows the track pattern, first using `bind_shadow`, then imputing with `impute_knn`, and adding a `label_shadow`:
+The code chunk below shows the track pattern, first using `nabular`, then imputing with `impute_knn`, and adding a `label_shadow`:
 
@@
 aq_imputed <- airquality %>%
-  bind_shadow() %>%
+  nabular() %>%
   simputation::impute_lm(Ozone ~ Temp + Wind) %>%
@@
 The most missings are in building area, followed by year built, and land size, with similar amounts of missingness in Car, bathroom, bedroom2, longitude, and latitude.
-Figure \@ref(fig:housing-miss-case-var)B reveals there are up to 50% missing values in cases, and that the majority of cases have more than 5% values missing.
+Figure \@ref(fig:housing-miss-case-var)B reveals there are up to 50% missing values in cases, and the majority of cases have more than 5% values missing.
 The variables building area and year built have more than 50% missing data, and so could perhaps be omitted from subsequent analysis, as imputed values are likely to be spurious.
@@
 Missingness structure can be better understood and evaluated by using the data to predict where the data goes missing.
-This can be achieved by clustering the missing values into groups, and then applying classification and regression trees (CART) to find those variables and values that predict missingness clusters [@Tierney2015; @Barnett2017].
+This can be achieved by clustering the missing values into groups, and then applying classification and regression trees (CART) to find those variables and values predicting missingness clusters [@Tierney2015; @Barnett2017].
 Two clusters are identified, one for each missingness cluster.
@@
 dat_house_knn <- dat_house %>% 
-  bind_shadow() %>%
+  nabular() %>%
   as.data.frame() %>% 
@@
 dat_house_lm <- dat_house %>%
-  bind_shadow() %>%
+  nabular() %>%
   as.data.frame() %>%
@@
 These are plotted with hex bins, which are coloured according to the number of points in a given hexagon - the brighter the colour the more points.
-This reveals that the data sets imputed with knn and linear models have more points clustered around zero, whereas complete cases is more flat.", fig.width = 9, fig.height = 3, out.width = "100%"}
+This reveals data sets imputed with knn and linear models have more points clustered around zero, whereas complete cases is more flat.", fig.width = 9, fig.height = 3, out.width = "100%"}
 
@@
 
-The `naniar` and `visdat` packages build on existing tidy tools and strike a compromise between automation and control that makes analysis efficient, readable, but not overly complex.
+The `naniar` and `visdat` packages build on existing tidy tools and strike a compromise between automation and control, making analysis efficient, readable, but not overly complex.
 Each tool has clear intent and effects - plotting or generating data or augmenting data in some way.

### 2020-01-07 paper/tidy-missing-data-paper.Rmd
Restructuring headings of the paper

@@
 title:
-  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations with the Package \\pkg{naniar}"
+  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
   # If you use tex in the formatted title, also supply version without
-  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations with the Package naniar"
+  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations"
   # For running headers, if needed
-  short:     "\\pkg{naniar}: Explore missing data"
+  short:     "Explore missingness with \\pkg{naniar}"
 abstract: >
@@
 
+# Existing Software {#existing-software}
+
 ## Imputation {#imputation}
@@
 
-## Exploration {#exploration}
+### Exploration {#exploration}
 
@@
 missing values that tracks missings throughout the analysis and allows for
-special missing values. Section \@ref(vis-summaries) covers visualizations to
+special missing values. Section \@ref(graphics) covers visualizations to
 understand missingness. Section \@ref(num-sum) discusses data summaries, and
@@
 
-(**consider removing these longer expansions of these points?**)
-
 ### Coordinated names {#coordinated-names}
@@
 
-## Visual summaries {#vis-summaries}
+# Graphics {#graphics}
 
@@
 This section discusses easy to remember visualizations that compose with other functions in a pipeline.
-Four different areas of visualization of missing data are discussed: overviews (Section \@ref(overview-plots)); univariate (Section \@ref(univariate-plots)), bivariate (Section \@ref(bivariate-plots)), and multivariate (Section \@ref(multivariate-plots)).
+Four different areas of visualization of missing data are discussed: overviews (Section \@ref(overview-graphics)); univariate (Section \@ref(univariate-graphics)), bivariate (Section \@ref(bivariate-graphics)), and multivariate (Section \@ref(multivariate-graphics)).
 All plots are created using \pkg{ggplot2}, giving users clearer control over the plot appearance, and customizability.
 
-### Overview plots {#overview-plots}
+## Overview graphics {#overview-graphics}
 
@@
 
-### Univariate plots {#univariate-plots}
+## Univariate graphics {#univariate-graphics}
 
@@
 
-#### Imputing values below the range {#imp-below-range}
+### Imputing values below the range {#imp-below-range}
 
@@
 
-#### Univariate split by missingness {#vis-split-by-miss}
+### Univariate split by missingness {#vis-split-by-miss}
 
@@
 
-### Bivariate plots {#bivariate-plots}
+## Bivariate graphics {#bivariate-graphics}
 
@@
 
-### Multivariate plots {#multivariate-plots}
+## Multivariate graphics {#multivariate-graphics}
 
@@
 
-## Numerical summaries {#num-sum}
+# Numerical summaries {#num-sum}
 
@@
 
-### Single number summaries {#single-num-sum}
+## Single number summaries {#single-num-sum}
 
@@
 
-### Summaries and tabulations of missing data {#sum-tab-missings}
+## Summaries and tabulations of missing data {#sum-tab-missings}
 
@@
 
-### Combining numerical summaries with grouping operations {#num-sum-w-group}
+## Combining numerical summaries with grouping operations {#num-sum-w-group}
 

### 2020-01-07 paper/tidy-missing-data-paper.Rmd
cutting down on words + re-arranging
- verbs --> "missing data operations"
- move "missing data operations" to just after "data structures"
- adjust some heading levels

@@
 
-```{r warning, fig.cap = "How ggplot behaves when displaying missing values:  Warning message (Left) is thrown prior to plot (Right); ggplot2 does not display missing values in the plot, but at least throws a warning message to the user, which is better than many statistical models.", fig.height = 4, fig.width = 4, out.width = "100%", warning = FALSE, fig.show = 'hold'}
+```{r warning, fig.cap = "How ggplot behaves when displaying missing values:  Warning message (Left) is thrown prior to plot (Right); ggplot2 does not display missing values in the plot, but at least throws a warning message to the user, which is better than many statistical models.", fig.height = 2.5, fig.width = 2.5, out.width = "75%", warning = FALSE, fig.show = 'hold'}
 
@@
 Despite it being crucial to explore and understand missing data, there is very little research on how to make this process more efficient and integrate it into the analysis pipeline.
-
 The analyst must
@@
 This disrupts workflow, and incites errors.
-
-Tidy tools fall into three broad categories: data manipulation, visualization, and modelling. These are now discussed in turn.
+Tidy tools fall into three broad categories: data manipulation, visualization, and modelling.
+These are now discussed in turn.
 
@@
 
-Defining tidy data and tidy tools has resulted in the growth of a set of packages known collectively as the "tidyverse" [@tidyverse].
-These are designed to share similar principles in their design and behavior, and cover the breadth of an analysis - going from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @tidyverse; @Tidyverse-Manifesto].
+Defining tidy data and tidy tools has resulted in a growing set of packages known collectively as the "tidyverse" [@tidyverse].
+These are designed to share similar principles in their design and behavior, and cover the breadth of an analysis - from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @tidyverse; @Tidyverse-Manifesto].
 This has led to more tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr} and dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} [@readr; @readxl; @haven; @stringr; @lubridate; @purrr].
@@
 
-### Coordinated names {#coordinated-names}
+## Coordinated names {#coordinated-names}
 
@@
 
-### Missing value labels {#missing-labels}
+## Missing value labels {#missing-labels}
 
@@
 
-### Special missings {#special-missings}
+## Special missings {#special-missings}
 
@@
 
-### Nabular data {#nabular-data}
+## Nabular data {#nabular-data}
 
@@
 
+# Missing data operations {#verbs}
+
+Common operations on missing data can be considered to be verbs: scan, replace, add, shadow, impute, track, and flag.
+Missing values can be **scanned** to find possible missings not coded as `NA`.
+These values can then be **replaced** with an `NA` value.
+To facilitate exploration, summaries of missingness can be **added** as a summary column to the original data.
+Data values can be augmented with **shadow** values, which helps explore missing data, as well as facilitating the process of **imputing**, and **tracking**.
+Finally, unusual or specially coded missing values can be **flagged**.
+These verbs are described in this section.
+
+## scan: Searching for common missing value labels  {#verbs-search}
+
+There are many common values for "missing".
+For example, "N/A" "Not Available", "MISSING", or white space " ".
+Before replacing these values with missing values (`NA`), it is useful to first _scan_ the data to get a sense of the magnitude of the problem.
+The  `miss_scan_count()` function finds where specified values occur in data.
+For example, all occurrences of -99 can be searched for with `miss_scan_count(data, search = -99)`.
+This returns a table of the occurrences of that value in each variable.
+
+```{r setup-miss-scan-count}
+
+dat_ms <- tibble::tribble(~x,  ~y,    ~z,
+                         1,   "A",   -100,
+                         3,   "N/A", -99,
+                         NA,  NA,    -98,
+                         -99, "E",   -101,
+                         -98, "F",   -1)
+
+# knitr::kable(dat_ms, 
+#              caption = "An example dataset with variables x, y and z containing some common unusual missing values.")
+             # booktabs = TRUE) 
+  # kable_styling(latex_options = c("hold_position"))
+```
+
+```{r miss-scan-count, eval = FALSE}
+miss_scan_count(data = dat_ms, 
+                search = -99) %>% 
+  knitr::kable(
+    caption = "Table of the occurences of the search '-99' in the data, 'dat\\_ms'. There is one occurence if -99 in variables x and z.")
+    # booktabs = TRUE) 
+  # kable_styling(latex_options = c("hold_position"))
+```
+
+A list of common NA values is provided in `naniar` in the `common_na_strings` and `common_na_numbers` for characters and numbers.
+These contain values like "n/a", "na ", and ".", and -9, -99.
+These can be used with `miss_scan_count` to help identify common missing values.
+
+## replace: Replacing missing value label with another {#verbs-replace-with}
+
+Once the magnitude of the unusual missing values has been assessed, these values need to be replaced with appropriate missing values.
+This can be done with `replace_with_na()`.
+This is useful when there is certainty of which values mean missing.
+For example "N/A", "N A", "Not Available", and "missing".
+A dataset could have the values -99 in the `x` column replaced with the following code:
+`replace_with_na(dat_ms, replace = list(x = -99))`
+
+For flexibility, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`.
+This means `replace_with_na` works for **all** variables with `replace_with_na_all`, **at** selected variables with `replace_with_na_at`, and **if** variables meet some condition with `replace_with_na_if`.
+
+## add: Adding missingness summary variables {#verbs-add-cols}
+
+Summary information such as the number of missings in a given case can be useful in understanding missingness structure, and even more so when this information is kept alongside the data.
+Functions starting with `add_` exist in `dplyr` to add count information for particular groups or conditions to the data.
+Similarly, `naniar` includes `add_` functions to add missingness summary information back to the data, such as the number or proportion of missingness, the missingness cluster, or if there are any missings, using: `add_n_miss()`
+`add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (see Table \@ref(tab:add-missing-info)).
+
+```{r add-missing-info}
+
+tibble::tibble(
+  `Function` = c(
+    "add_n_miss(data)",
+    "add_prop_miss(data)",
+    "add_miss_cluster(data)"
+    ),
+  `Adds column which:` = c(
+    "Contains the number missing values in a row",
+    "Contains the proportion of missing values in a row",
+    "Contains the missing value cluster"
+  )) %>% 
+    knitr::kable(caption = "Overview of the 'add' functions in naniar")
+
+```
+
+An example use of these features is in @Tierney2015,
+where the proportion, number, or cluster of missings is used as the outcome in a
+model. The variables in the dataset can then be used to predict the outcome, identifying variables and values important in predicting
+missingness structures. There are also functions for adding information about shadow values or nice labels for if there are any missing values with `add_label_shadow()` and `add_label_missings()`.
+
+## shadow: Creating nabular data {#verbs-nabular}
+
+_Nabular_ data has the shadow matrix column-bound to the existing data.
+This facilitates visualization and summaries, and allows for imputed values to be appropriately tracked.
+_Nabular_ data can be created with `nabular()` or `bind_shadow()`.
+(The function was named `bind_shadow()` to borrow from existing functions named `bind` - such as `bind_rows()` and `bind_cols()` from `dplyr`, which bind rows or columns into a dataframe).
+An example use of `nabular()` some example data set is shown below:
+
+```{r bind-shadow, echo = TRUE}
+nabular(dat_ms)
+```
+
+## flag: Describing different types of  missing values {#verbs-recode}
+
+Unusual or spurious data values are often identified and `flagged`.
+For example, there might be special codes to mark an individual dropping out of a study, known instrument failure in weather instruments, or for values censored in analysis.
+These special types of missingness can be encoded in the shadow matrix of  _nabular_  data, using the `recode_shadow()` function.
+This provides a fluent interface to recode, or flag, the shadow information in the shadow matrix as a special type of missing value.
+Using `recode_shadow()` requires specifying the variable you want to contain the flagged value, the condition you want this to occur at, and a suffix for the new type of missing value.
+This is then recoded as a new factor level in the shadow matrix, so every column is aware of all possible new values of missingness.
+For example, the values -99 and -98 could be recoded to mean a broken machine sensor for the variable x:
+
+```{r recode-shadow, eval = TRUE, echo = TRUE}
+dat_ms %>%
+  nabular() %>%
+  recode_shadow(x = .where(x == -99 ~ "broken_sensor"))
+```
+
+## impute: Imputing values {#verbs-impute}
+
+`naniar` does not reinvent the wheel for imputation, as there are many robust R packages for imputation, such as `mice`, `mi`, `VIM`, and `simputation`.
+However, `naniar` does provide a few imputation methods to facilitate exploration and visualizations, which were not otherwise easily available: `impute_below` and `impute_mean` and `impute_median`.
+These imputation methods are useful to explore structure in missingness, but are not recommended for use in analysis.
+The `impute_below` function imputes values below the minimum of the data, with some jitter (small values of random noise) to reduce overplotting, but has options to change the default amount it is shifted below, and the amount of jitter.
+
+```{r demonstrate-impute-below, eval = FALSE}
+airquality %>%
+  impute_below_at(vars(Ozone)) %>%
+  select(Ozone, Solar.R) %>%
+  head()
+```
+
+Similar to `simputation`, each `impute_` function returns the data with values imputed.
+However, `naniar` does not use a formula syntax, and instead each function implements "scoped variants" `_all`, `_at` and `_if`.
+Here, `_all` operates on **all** columns, `_at` operates **at** specific columns, and `_if` operates on columns **if** they meet some condition (such as `is.numeric` or `is.character`).
+The `impute_` functions with no scoped variant, e.g., `impute_mean`, will work on a single vector, but not a data.frame.
+One drawback to imputation functions in this form is imputed value locations are not tracked.
+This issue is resolved with \code{nabular} data covered in Section \@ref(verbs-track).
+
+```{r impute-vector, eval = FALSE}
+
+impute_mean(airquality$Ozone) %>% head()
+
+impute_mean_at(airquality, .vars = vars(Ozone)) %>% head()
+
+impute_mean_if(airquality, .predicate = is.integer) %>% head()
+
+impute_mean_all(airquality) %>% head()
+
+```
+
+## track: Shadow and impute missing values {#verbs-track}
+
+To evaluate imputations they need to be tracked, which is achieved by combining the verbs `nabular`, `impute_`, and `add_label_shadow`.
+The missing values can then be referred to by their shadow variable, `_NA`.
+The missingness of any observation can be referred to with `any_missing` (Figure \@ref(fig:track-impute-example)).
+The code chunk below shows the track pattern, first using `nabular`, then imputing with `impute_knn`, and adding a `label_shadow`:
+
+```{r bind-impute-label-example, echo = TRUE}
+
+aq_imputed <- airquality %>%
+  nabular() %>%
+  simputation::impute_lm(Ozone ~ Temp + Wind) %>%
+  simputation::impute_lm(Solar.R ~ Temp + Wind) %>%
+  add_label_shadow()
+
+aq_imputed
+```
+
+Missing values can then be shown in a scatterplot by setting the `color` aesthetic in ggplot to `any_missing` (Figure \@ref(fig:track-impute-example)A), or in a density plot using the `fill = any_missing`, (Figures \@ref(fig:track-impute-example)B) and \@ref(fig:track-impute-example)C).
+Imputed values can also be compared to complete case data, grouping by `any_missing`, and then summarizing, similar to other dplyr summary workflows.
+This shows mean and median values are similar, but the minimum and maximum values are different.
+
+```{r track-impute-example, fig.show = "hold", fig.cap = "Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values imputed using simputations `impute\\_knn` function, with imputed values colored green and data values orange.
+Imputed values are similar, but slightly different to existing data.", fig.height = 4, fig.width = 12, out.width = "100%", echo = FALSE}
+
+p1 <- 
+ggplot(aq_imputed,
+       aes(x = Ozone,
+           y = Solar.R,
+           color = any_missing)) + 
+  geom_point() +
+  scale_color_brewer(palette = "Dark2") +
+  theme(legend.position = "none") + 
+  labs(tag = "A")
+
+p2 <- 
+ggplot(aq_imputed,
+       aes(x = Ozone,
+           fill = any_missing)) + 
+  geom_density(alpha = 0.3) + 
+  scale_fill_brewer(palette = "Dark2") +
+  theme(legend.position = "none") + 
+  labs(tag = "B")
+
+p3 <- 
+ggplot(aq_imputed,
+       aes(x = Solar.R,
+           fill = any_missing)) + 
+  geom_density(alpha = 0.3) + 
+  scale_fill_brewer(palette = "Dark2") + 
+  theme(legend.position = "none") + 
+  labs(tag = "C")
+
+gridExtra::grid.arrange(p1, p2, p3, nrow = 1)
+```
+
+```{r impute-summary, echo = TRUE}
+aq_imputed %>%
+  group_by(any_missing) %>%
+  summarise_at(.vars = vars(Ozone),
+               .funs = lst(min, mean, median, max)) 
+```
+
 # Graphics {#graphics}
@@
 
-### Imputing values below the range {#imp-below-range}
-
+**Imputing values below the range**.
 To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed values below the range of data using `impute_below` (by default imputing 10% below the range).
@@
 
-### Univariate split by missingness {#vis-split-by-miss}
-
+**Univariate split by missingness**.
 Missingness of one variable can also be used to display different distributions in another.
@@
 
-To visualize missing values in two dimensions the missing values can be placed in the plot margins.
-This is achieved by imputing values below the range of the data.
-As in Section \@ref(imp-below-range), using _nabular_ data identifies imputed values, and color makes missingness pre-attentive [@treisman1985].
-The steps of imputing and coloring have been been combined into `geom_miss_point()` from the \pkg{naniar} package (Figure \@ref(fig:geom-miss)), where there is a mostly uniform spread of missing values for Solar.R and Ozone.
+To visualize missing values in two dimensions the missing values can be placed in plot margins, by imputing values below the range of the data.
+As in Section \@ref(univariate-graphics), using _nabular_ data identifies imputed values, and color makes missingness pre-attentive [@treisman1985].
+The steps of imputing and coloring are combined into `geom_miss_point()` from the \pkg{naniar} package (Figure \@ref(fig:geom-miss)).
+There is a mostly uniform spread of missing values for Solar.R and Ozone.
 
@@
 This means they integrate well with plotting and modelling tools.
-Section \@ref(single-num-sum) discusses single number summaries for all data and for cases and variables ; Section \@ref(sum-tab-missings) discusses variable- and case-wise summaries; and Section \@ref(num-sum-w-group) shows how the design of these tools works with other tools in an analysis pipeline.
+Section \@ref(single-num-sum) discusses single number summaries; Section \@ref(sum-tab-missings) discusses variable- and case-wise summaries; and Section \@ref(num-sum-w-group) shows how the design of these tools works with other tools in an analysis pipeline.
 
@@
 
-The overall number of missing values in a dataset is shown with `n_miss`, and the proportion or percent missings, `prop_miss` and `pct_miss`.
+The overall number, proportion, or percent of missing values in a dataset is shown with `n_miss`, `prop_miss` and `pct_miss`.
 Complementing these are functions to reveal the number of complete values: `n_complete`, `prop_complete`, and `pct_complete`.
-These be extended to summarize the number/amount of variables and cases containing a missing values by appending `_case` or `_var` to any of these summaries.
-Table \@ref(tab:n-prop-pct-miss-complete) shows the names and output of these functions on the dataset airquality.
+These can summarize missings for variables and cases by appending `_case` or `_var` to any of the summaries.
+Table \@ref(tab:n-prop-pct-miss-complete) shows the names and output of these functions on the dataset 'airquality'.
 
@@
   digits = 2,
-  caption = "Single number summaries of missing and complete data, applied to the airquality dataset.
-These functions follow consistent naming, to make them easy to remember, and to make their use clear.")
+  caption = "Single number summaries of missingness and completeness of the airquality dataset. The functions follow consistent naming, making them easy to remember, and their use clear.")
   # booktabs = TRUE)
@@
 
-```
-
-## Verbs {#verbs}
-
-Common operations on missing data can be considered to be verbs: scan, replace, add, shadow, impute, track, and flag.
-Missing values can be **scanned** to find possible missings not coded as `NA`.
-These values can then be **replaced** with an `NA` value.
-To facilitate exploration, summaries of missingness can be **added** as a summary column to the original data.
-Data values can be augmented with **shadow** values, which helps explore missing data, as well as facilitating the process of **imputing**, and **tracking**.
-Finally, unusual or specially coded missing values can be **flagged**.
-These verbs are described in this section.
-
-### scan: Searching for common missing value labels  {#verbs-search}
-
-There are many common values for "missing".
-For example, "N/A" "Not Available", "MISSING", or white space " ".
-Before replacing these values with missing values (`NA`), it is useful to first _scan_ the data to get a sense of the magnitude of the problem.
-The  `miss_scan_count()` function finds where specified values occur in data.
-For example, for the example data, `dat_ms`  Table \@ref(tab:setup-miss-scan-count).
-All occurrences of -99 can be searched for using the code:
-
-```{r setup-miss-scan-count}
-
-dat_ms <- tibble::tribble(~x,  ~y,    ~z,
-                         1,   "A",   -100,
-                         3,   "N/A", -99,
-                         NA,  NA,    -98,
-                         -99, "E",   -101,
-                         -98, "F",   -1)
-
-knitr::kable(dat_ms, 
-             caption = "An example dataset with variables x, y and z containing some common unusual missing values.")
-             # booktabs = TRUE) 
-  # kable_styling(latex_options = c("hold_position"))
-```
-
-```{r miss-scan-print, eval = FALSE, echo = TRUE}
-miss_scan_count(data = dat_ms, search = -99)
-```
-
-This returns a table of the occurrences of that value in each variable (Table \@ref(tab:miss-scan-count)).
-A list of common NA values is provided in `naniar` in the `common_na_strings` and `common_na_numbers` for characters and numbers.
-These contain values like "n/a", "na ", and ".", and -9, -99.
-
-```{r miss-scan-count}
-miss_scan_count(data = dat_ms, 
-                search = -99) %>% 
-  knitr::kable(
-    caption = "Table of the occurences of the search '-99' in the data, 'dat\\_ms'. There is one occurence if -99 in variables x and z.")
-    # booktabs = TRUE) 
-  # kable_styling(latex_options = c("hold_position"))
-```
-
-### replace: Replacing missing value label with another {#verbs-replace-with}
-
-Once the magnitude of the unusual missing values has been assessed, these values need to be replaced with appropriate missing values, using `replace_with_na`.
-This is useful when there is certainty of which values mean missing.
-For example "N/A", "N A", "Not Available", and "missing".
-Using the previous toy dataset, `dat_ms`, the value -99, or both -99 and 98 can be replaced with NA as follows:
-
-```{r replace-with-na-code-chunk, echo = TRUE}
-replace_with_na(dat_ms, replace = list(x = -99))
-
-replace_with_na(dat_ms, replace = list(x = c(-99, -98)))
-```
-
-(**considering removing these examples**)
-
-For flexibility, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`.
-This means `replace_with_na` works for **all** variables with `replace_with_na_all`, **at** selected variables with `replace_with_na_at`, and **if** variables meet some condition with `replace_with_na_if`.
-The syntax for these is slightly different to `replace_with_na`, but powerful:
-
-```{r replace-with-na-scoped-variants, echo = TRUE}
-replace_with_na_at(data = dat_ms,
-                   .vars = "x",
-                   condition = ~.x == -99)
-
-replace_with_na_if(data = dat_ms,
-                 .predicate = is.character,
-                 condition = ~.x == "N/A")
-
-replace_with_na_all(data = dat_ms,
-                    condition = ~.x == -99)
-
-```
-
-### add: Adding missingness summary variables {#verbs-add-cols}
-
-Summary information such as the number of missings in a given case can be useful in understanding missingness structure, and even more so when this information is kept alongside the data.
-Functions starting with `add_` exist in `dplyr` to add count information for particular groups or conditions to the data.
-Similarly, `naniar` includes `add_` functions to add missingness summary information back to the data, such as the number or proportion of missingness, the missingness cluster, or if there are any missings, using: `add_n_miss()`
-`add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively.
-
-An example use of these features is in @Tierney2015,
-where the proportion, number, or cluster of missings is used as the outcome in a
-model. The variables in the dataset can then be used to predict the outcome, identifying variables and values important in predicting
-missingness structures. There are also functions for adding information about shadow values or nice labels for if there are any missing values with `add_label_shadow()` and `add_label_missings()`:
-
-```{r add-missing-info, echo = TRUE}
-add_n_miss(dat_ms)
-add_prop_miss(dat_ms)
-add_miss_cluster(dat_ms)
-```
-
-### shadow: Creating nabular data {#verbs-nabular}
-
-_Nabular_ data has the shadow matrix column-bound to the existing data.
-This facilitates visualization and summaries, and allows for imputed values to be appropriately tracked.
-_Nabular_ data can be created with `nabular()` or `bind_shadow()`.
-The function was named `bind_shadow()` to borrow from existing functions named `bind` - such as `bind_rows()` and `bind_cols()` from `dplyr`, which bind rows or columns into a dataframe.
-An example use of `nabular()` on the example data, `dat_ms` is shown below:
-
-```{r bind-shadow, echo = TRUE}
-nabular(dat_ms)
-```
-
-### flag: Describing different types of  missing values {#verbs-recode}
-
-Unusual or spurious data values are often identified and `flagged`.
-For example, there might be special codes to mark an individual dropping out of a study, known instrument failure in weather instruments, or for values censored in analysis.
-These special types of missingness can be encoded in the shadow matrix of  _nabular_  data, using the `recode_shadow()` function.
-This provides a fluent interface to recode, or flag, the shadow information in the shadow matrix as a special type of missing value.
-Using `recode_shadow()` requires specifying the variable you want to contain the flagged value, the condition you want this to occur at, and a suffix for the new type of missing value.
-This is then recoded as a new factor level in the shadow matrix, so every column is aware of all possible new values of missingness.
-For example, the values -99 and -98 could be recoded to mean a broken machine sensor for the variable x:
-
-```{r recode-shadow, eval = TRUE, echo = TRUE}
-dat_ms %>%
-  nabular() %>%
-  recode_shadow(x = .where(x == -99 ~ "broken_sensor"))
-```
-
-### impute: Imputing values {#verbs-impute}
-
-`naniar` does not reinvent the wheel for imputation, as there are many robust R packages for imputation, such as `mice`, `mi`, `VIM`, and `simputation`.
-However, `naniar` does provide a few imputation methods to facilitate exploration and visualizations, which were not otherwise easily available: `impute_below` and `impute_mean` and `impute_median`.
-These imputation methods are useful to explore structure in missingness, but are not recommended for use in analysis.
-
-The `impute_below` function imputes values below the minimum of the data, with some jitter (small values of random noise) to reduce overplotting, but has options to change the default amount it is shifted below, and the amount of jitter.
-
-```{r demonstrate-impute-below, eval = FALSE}
-airquality %>%
-  impute_below_at(vars(Ozone)) %>%
-  select(Ozone, Solar.R) %>%
-  head()
-```
-
-Similar to `simputation`, each `impute_` function returns the data with values imputed.
-However, `naniar` does not use a formula syntax, and instead each function implements "scoped variants" `_all`, `_at` and `_if`.
-Here, `_all` operates on **all** columns, `_at` operates **at** specific columns, and `_if` operates on columns **if** they meet some condition (such as `is.numeric` or `is.character`).
-The `impute_` functions with no scoped variant, e.g., `impute_mean`, will work on a single vector, but not a data.frame.
-An example usage is shown below:
-
-```{r impute-vector, echo = TRUE}
-
-impute_mean(airquality$Ozone) %>% head()
-
-impute_mean_at(airquality, .vars = vars(Ozone)) %>% head()
-
-impute_mean_if(airquality, .predicate = is.integer) %>% head()
-
-impute_mean_all(airquality) %>% head()
-
-```
-
-One drawback to imputation functions in this form is imputed value locations are not tracked. This issue is resolved with \code{nabular} data covered in Section \@ref(verbs-track).
-
-### track: Shadow and impute missing values {#verbs-track}
-
-To evaluate imputations they need to be tracked, which is achieved by combining the verbs `nabular`, `impute_`, and `add_label_shadow`.
-The missing values can then be referred to by their shadow variable, `_NA`.
-The missingness of any observation can be referred to with `any_missing` (Figure \@ref(fig:track-impute-example)).
-The code chunk below shows the track pattern, first using `nabular`, then imputing with `impute_knn`, and adding a `label_shadow`:
-
-```{r bind-impute-label-example, echo = TRUE}
-
-aq_imputed <- airquality %>%
-  nabular() %>%
-  simputation::impute_lm(Ozone ~ Temp + Wind) %>%
-  simputation::impute_lm(Solar.R ~ Temp + Wind) %>%
-  add_label_shadow()
-
-aq_imputed
-```
-
-Missing values can then be shown in a scatterplot by setting the `color` aesthetic in ggplot to `any_missing` (Figure \@ref(fig:track-impute-example)A), or in a density plot looking at one variable, using the `fill = any_missing`, (Figures \@ref(fig:track-impute-example)B) and \@ref(fig:track-impute-example)C).
-
-```{r track-impute-example, fig.show = "hold", fig.cap = "Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values imputed using simputations `impute\\_knn` function, with imputed values colored green and data values orange.
-Imputed values are similar, but slightly different to existing data.", fig.height = 4, fig.width = 12, out.width = "100%", echo = FALSE}
-
-p1 <- 
-ggplot(aq_imputed,
-       aes(x = Ozone,
-           y = Solar.R,
-           color = any_missing)) + 
-  geom_point() +
-  scale_color_brewer(palette = "Dark2") +
-  theme(legend.position = "none") + 
-  labs(tag = "A")
-
-p2 <- 
-ggplot(aq_imputed,
-       aes(x = Ozone,
-           fill = any_missing)) + 
-  geom_density(alpha = 0.3) + 
-  scale_fill_brewer(palette = "Dark2") +
-  theme(legend.position = "none") + 
-  labs(tag = "B")
-
-p3 <- 
-ggplot(aq_imputed,
-       aes(x = Solar.R,
-           fill = any_missing)) + 
-  geom_density(alpha = 0.3) + 
-  scale_fill_brewer(palette = "Dark2") + 
-  theme(legend.position = "none") + 
-  labs(tag = "C")
-
-gridExtra::grid.arrange(p1, p2, p3, nrow = 1)
-```
-
-Imputed values can also be compared to complete case data by grouping by `any_missing`, and then summarizing (Table \@ref(tab:impute-summary)).
-
-```r
-aq_imputed %>%
-  group_by(any_missing) %>%
-  summarise_at(.vars = vars(Ozone),
-               .funs = lst(min, mean, median, max))
-```
-
-```{r impute-summary}
-aq_imputed %>%
-  group_by(any_missing) %>%
-  summarise_at(.vars = vars(Ozone),
-               .funs = lst(min, mean, median, max)) %>%
-  knitr::kable(
-    digits = 1, 
-    caption = "Summary statistics of the previously imputed and complete cases.
-The mean and median values are similar, but the minimum and maximum values are different.
-The comparison of imputed values is similar to other dplyr summary workflows.")
-    # booktabs = TRUE) 
-  # kable_styling(latex_options = c("hold_position"))
 ```
@@
 Possible missingness structures are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \@ref(fig:applic-vis-miss)).
-There are three main clusters of missing data: at the top there are many variables missing together; in the middle, building price and year built missing together, at the bottom, building area, year built, and land size are missing.
+There are three main clusters of missing data: at the top many variables go missing together; in the middle, building price and year built, at the bottom, building area, year built, and land size.
 
@@
 
-Missingness patterns can also be shown with an `upset` plot [@Conway2017], to display the 8 of intersecting sets of missing variables (see Figure \@ref(fig:applic-vis-miss). 
-Two patterns stand out: two, and five variables missing. This provides further evidence of the patterns of missingness seen in Figures \@ref(fig:housing-miss-case-var) and \@ref(fig:housing-upset).
+Missingness patterns can also be shown with an `upset` plot [@Conway2017], to display the 8 of intersecting sets of missing variables (see Figure \@ref(fig:applic-vis-miss)).
+Two patterns stand out: two, and five variables missing, providing further evidence of the missingness patterns seen in Figures \@ref(fig:housing-miss-case-var) and \@ref(fig:housing-upset).
 
@@
 
-\newpage
-
 # Acknowledgements

### 2020-01-09 paper/tidy-missing-data-paper.Rmd
type-editing sections 1 and 2

@@
 
-```{r warning, fig.cap = "How ggplot behaves when displaying missing values:  Warning message (Left) is thrown prior to plot (Right); ggplot2 does not display missing values in the plot, but at least throws a warning message to the user, which is better than many statistical models.", fig.height = 2.5, fig.width = 2.5, out.width = "75%", warning = FALSE, fig.show = 'hold'}
+```{r warning, fig.cap = "How ggplot behaves when displaying missing values: ggplot2 does not display missing values in the plot, but at least throws a warning message to the user, which is better than many statistical models.", fig.height = 2.5, fig.width = 2.5, out.width = "75%", warning = FALSE, fig.show = 'hold'}
 
@@
 The imputation literature (e.g. @Little1988, @Rubin1976, @Simon1986,
-@Schafer2002, @VanBuuren2012) does not address how to explore, understand, and handle
-missing data structures and mechanisms.
-Instead the focus is on ensuring valid statistical inference is made from
-incomplete data. They approach this chiefly through probabilistic modelling, and
-assume the mechanism of missing data is known to the analyst.
-
-However, something has to be known about the missing value structure to produce
-a complete dataset for analysis. Whether by case or variable deletion, or with
-imputation, it is ideal to identify the mechanism generating missing 
-values.
-To understand their structure, the data must be explored, which means generating visualizations, summaries, and performing exploratory modelling.
-This process is inherently difficult, as it is iterative, and
-involves many dead ends, and there might not be a clear answer even after much
-exploratory data analysis.
-
-Although
-decision tree models or latent group analysis can reveal structures or patterns
-of missingness [@Tierney2015; @Barnett2017], these are not definitive and
-usually require discussion with data curators to validate.
-While there might be other times where the data are
-clearly missing completely at random (or missing at random), determining this is not always straightforward. There is no silver bullet to reveal missing data
-mechanisms to the analyst. Missing data must be explored.
-
-While there have been many software tools for exploring missing data (VIM, mi, mice, MissingDataGUI, MANET, ggobi, norm), these tools generate outputs challenging to integrate into analysis pipelines.
+@Schafer2002, @VanBuuren2012) focusses on ensuring valid statistical inference
+is made from incomplete data. This is approached chiefly through probabilistic
+modelling, assuming the mechanism of missing data is known to the analyst. It
+does not address how to explore, understand, and handle missing data structures
+and mechanisms.
+
+However, something must be known about the missing value structure to produce a
+complete dataset for analysis, whether by case- or variable-deletion, or with
+imputation. To understand their structure, the analyst must explore the data
+with perform visualizations, summaries, and modelling. This exploratory process
+is iterative, and often soaks up a lot of time before more pertinent questions
+can be answered.
+
+Decision tree models or latent group analysis can reveal structures or patterns
+of missingness [@Tierney2015; @Barnett2017], but are not definitive. While there
+might be other times where the missing data mechanism is obvious, determining
+this is not always straightforward. There is no silver bullet to reveal missing
+data mechanisms to the analyst. Missing data must be explored.
+
+While there have been many software tools for exploring missing data [@VIM, @mi, @mice, @Cheng2015, @MANET, @ggobi, @norm], these tools do not integrate easily into analysis pipelines.
 Despite it being crucial to explore and understand missing data, there is very little research on how to make this process more efficient and integrate it into the analysis pipeline.
@@
 For example, imputing values 10% below the range to display all values in a scatter plot.
-These methods require the data to be linked to the missing or imputed values; this link is typically hidden from the user.
+These methods require the data to be linked to the missing or imputed values, but this link is typically hidden from the user.
 The approaches from the graphics literature have not been translated and integrated into tidy data.
@@
 The next section (\@ref(tidy-data-concepts)) is an introduction to tidy data principles and tools.
-Missing value imputation (\@ref(missing-data-rep-dep) and \@ref(imputation)) and exploration literature (\@ref(exploration)) are then discussed.
+Existing missing data software is discussed in \@ref(existing-software).
 Section \@ref(extensions) discusses tidy methodology extensions, which facilitate operations for exploring, visualizing, and imputing missing data.
@@
 For example, a scatterplot can be created by mapping two variables to the x and y axes, and specifying a point geometry.
-The graph can then be changed to map a third variable to color of the points, and a fourth variable to size.
 
@@
 For example, estimated coefficients, predictions, and residuals from one model cannot be easily combined with the output of another model.
-This output problem has been partially addressed with the \pkg{broom} package, which tidies up model outputs for data analysis, and the developing \pkg{recipes} package, which helps make modelling input and output tidy [@recipes].
+Messy models have been partially addressed with the \pkg{broom} package, which tidies up model outputs for data analysis, and the developing \pkg{recipes} package, which helps make modelling input- and output-tidy [@recipes].
 
@@
 This has led to more tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr} and dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} [@readr; @readxl; @haven; @stringr; @lubridate; @purrr].
-It has also led to a burgeoning of new packages for other fields following similar design principles and create fluid workflows for new domains.
+It has also led to a burgeoning of new packages for other fields following similar design principles, creating fluid workflows for new domains.
 For example, the \pkg{tidytext} package for text analysis, the \pkg{tsibble} package for time series data, and \pkg{tidycensus} for working with US census and boundary data [@tidytext; @wang2020tsibble; @tidycensus].
@@
 Translating and expanding these to fit within tidy data and tidy tools would create more efficient workflows.
-By laying down these concepts, it allows the focus to be more general than just software, but rather, an extensible framework for tidy tools to explore missing data.
+Defining these concepts allows the focus to be more general than just software, but rather, an extensible framework for tidy tools to explore missing data.
 
@@
 
-The information on missing values in a dataset can be represented as a missingness matrix, $R$, where 0 is observed data and 1 is missing data, where for data $y$ with $i$ rows and $j$ columns, the value $r_{ij}$ is given by:
+Missingness can be represented as a matrix, $R$, for data $y$ with $i$ rows and $j$ columns:
 
@@
 The information in $R$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), Missing at random (MAR), and missing not at random (MNAR).
-The distribution of missing values in $r_{ij}$ can depend on the entire dataset, which is represented as $Y = (Y_{obs}, Y_{miss})$.
-This relationship can be defined by the _missing data model_, of which $/psi$ contains the model parameters, and the missing data model can written as $Pr(r_{ij} | Y_{obs}, Y_{miss}, \psi)$ - the probability of missingness is conditional on the data observed, data missing, and some probability parameter of missingness.
-We can then use this to help precisely define categories of missing values.
+The distribution of missing values in $r_{ij}$ can depend on the entire dataset, represented as $Y = (Y_{obs}, Y_{miss})$.
+This relationship can be defined by the _missing data model_ $Pr(r_{ij} | Y_{obs}, Y_{miss}, \psi)$, the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, $\psi$.
+This helps precisely define categories of missing values.
 
 **MCAR** is where values being missing have no association with observed or unobserved data, that is, $Pr(R = 1 | Y_{obs}, Y_{miss}) = Pr(R = 1 | \psi)$.
-Essentially, the probability of an observation being missing is unrelated to anything else, only the parameters $\psi$, which is the overall probability of missingness.
-Although this is a convenient scenario, it is not actually possible to confirm, or clearly distinguish from MAR, as it relies on statements on data unobserved.
-
-**MAR** refers to cases where the missingness only depends on the data observed, and not data missing, that is,$Pr(R = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(R = 1 | Y_{obs}, \psi)$.
-Some structure or dependence between the missing values and observed values is allowed, provided that this can be explained by the data observed, and some overall probability of missingness.
-
-Finally, **MNAR** refers to the missingness being related to values observed, and un unobserved; if this does not simplify  $Pr(R = 1 | Y_{obs}, Y_{miss}, \psi)$.
-This assumes that conditioning on all available observed data, the data goes missing due to some phenomena unobserved, including the structure of the missing data itself.
+Essentially, the probability of an observation being missing is unrelated to anything else, only the parameters $\psi$, the overall probability of missingness.
+Although a convenient scenario, it is not actually possible to confirm, or clearly distinguish from MAR, as it relies on statements on data unobserved.
+In **MAR**, missingness only depends on data observed, not data missing, that is,$Pr(R = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(R = 1 | Y_{obs}, \psi)$.
+Some structure or dependence between missing and observed values is allowed, provided it can be explained by data observed, and some overall probability of missingness.
+In **MNAR**, missingness is related to values observed, and unobserved: $Pr(R = 1 | Y_{obs}, Y_{miss}, \psi)$.
+This assumes conditioning on all observations, data goes missing due to some phenomena unobserved, including the structure of the missing data itself.
 This presents a challenge in analysis, as it is difficult to verify, and implies bias in analysis due to the unobserved phenomena.
+
+<!-- (possibly remove this next para?) -->
 
@@
 We would also like to thank those who contributed pull requests and discussions on the `naniar` package, in particular Jim Hester and Romain François for improving the speed of key functions, Ross Gayler for discussion on special missing values, and Luke Smith for helping `naniar` be more compliant with `ggplot2`.
+We would also like to thank Amelia McNamara for her assistance with proof-reading the paper.
 

### 2020-01-09 paper/tidy-missing-data-paper.Rmd
tidy up section 3

@@
 
+Methods for exploring, understanding, and imputing missing data are more accessible than they have ever been.
+In particular, the number of methods and implementations in imputation seeing continued growth.
+Values can be imputed with one value (single imputation), or multiple values (multiple imputation), creating $m$ datasets.
+This section discusses existing software for single and multiple imputation \@ref(imputation), and exploration \@ref(exploration).
+
 ## Imputation {#imputation}
 
-Imputation methods are more accessible than they have ever been, and the number of methods and implementations in software continues to grow.
-Values can be imputed with one value (single imputation), or multiple values (multiple imputation), creating $m$ datasets.
-Methods and software for single and multiple imputation are now discussed.
-
-### Single imputation {#single-imputation}
-
-\pkg{VIM} [@VIM] efficiently implements well-used imputation methods K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation.
-This diversity of approaches allows the package to deal with semi-continuous, continuous, count, and categorical data.
+\pkg{VIM} [@VIM] implements well-used imputation methods K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation.
+These diverse approaches allows for imputing with semi-continuous, continuous, count, and categorical data.
 VIM identifies imputed cases by adding an indicator variable with a suffix `_imp`.
-So `Var1` would have a sibling indicator column, `Var1_imp`, with TRUE or FALSE values to indicate imputed value.
-\pkg{VIM} also has a variety of visualization methods, discussed in section \@ref(exploration).
-The \pkg{simputation} package provides an interface into popular imputation methods for linear regression, decision trees, K nearest neighbors, hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977].
-Other software for imputation includes \pkg{Hmisc} [@Hmisc], which provide predictive mean matching, \pkg{imputeTS} [@imputeTS], which provide time series imputation methods, and \pkg{missMDA} [@missMDA], which imputes data using principal components analysis.
-
-A key feature of the design of \pkg{simputation} is that it returns a regular dataframe.
-This is different from other imputation packages, which return custom classes of data or additional objects.
-These create extra work when preparing for subsequent analysis.
-Providing a dataframe with imputed values included in the data reduces the friction of working with other tools, but comes at the cost of identifying imputed values.
-
-### Multiple imputation {#multiple-imputation}
+So `Var1` has a sibling column, `Var1_imp`, with values TRUE or FALSE indicating imputation.
+\pkg{VIM} also has a variety of visualization methods, discussed in \@ref(exploration).
+\pkg{simputation} provides an interface to imputation methods from VIM, in addition providing hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977].
+\pkg{simputation} provides a consistent formula interface to all imputation, and always returns a dataframe with the updated imputations.
+\pkg{Hmisc} [@Hmisc], provide predictive mean matching, \pkg{imputeTS} [@imputeTS], provides time series imputation methods, and \pkg{missMDA} [@missMDA], imputes data using principal components analysis.
 
@@
 Popular and robust methods for multiple imputation include the \pkg{mice}, \pkg{Amelia}, and \pkg{mi} packages [@mice; @amelia; @mi].
-The \pkg{mice} package implements the method of chained equations, using a variable-wise algorithm to calculate the posterior distribution of parameters to generate imputed values.
+\pkg{mice} implements the method of chained equations, using a variable-wise algorithm to calculate the posterior distribution of parameters to generate imputed values.
 The workflow in \pkg{mice} revolves around imputing data, returning completed data, and fitting a model and pooling the results.
 
-The \pkg{Amelia} package [@amelia] assumes data are multivariate normal, and samples from the posterior using the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm [@Honaker2010] and allows for incorporation of information on the values in a prior.
-
-The \pkg{mi} package [@mi] also uses Bayesian models for imputation, providing better handling of semi-continuous values, and data with structural or perfect correlation.
+\pkg{Amelia} [@amelia] assumes data are multivariate normal, and samples from the posterior using the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm [@Honaker2010] and allows for incorporation of information on the values in a prior.
+
+\pkg{mi} [@mi] also uses Bayesian models for imputation, providing better handling of semi-continuous values, and data with structural or perfect correlation.
 A collection of analysis models are also provided in \pkg{mi}, which work with the specially multiply imputed data, including linear models, generalized linear models, and their corresponding Bayesian components.
@@
 
-Another software for performing multiple imputation is \pkg{norm} [@norm; @schafer-norm], which uses methods from the NORM software from @schafer-norm, such as multiple imputations using EM for multivariate normal data.
-The \pkg{norm} package does not provide a framework for tracking missing values, but instead provides tools for making inference from multiple imputation.
-
-Each of these multiple imputation methods provide practical methods, but all have different output structures, and do not have consistent interfaces in their implementation.
+\pkg{norm} [@norm], provides multiple imputation using EM for multivariate normal data, drawing from methods in the NORM software [@schafer-norm].
+\pkg{norm} does not provide a framework for tracking missing values, instead providing tools for making inference from multiple imputation.
+
+### Summary {#imputation-summary}
+
+Each imputation method provides practical methods for different usecases, but most have different output structures, and do not have consistent interfaces in their implementation.
 This makes then inherently messy and challenging to integrate into an analysis pipeline.
-An ideal approach would have consistent data structures that work with other analysis tools, and track missing values by default.
-This would make the data outputs tidy, streamlining subsequent analysis.
-
-### Exploration {#exploration}
+For example, combining different imputation methods from different pieces of software is not currently straightforward.
+\pkg{simputation} is different from other imputation packages, and resolves some of these complications with a simple approach of a unified syntax for all imputation, and always returning a dataframe containing imputed values.
+This reduces the friction of working with other tools, but comes at the cost of identifying imputed values.
+An ideal approach would use consistent, simple data structures that work with other analysis tools, and help track missing values.
+This would make imputation outputs tidy, streamlining subsequent analysis.
+
+## Exploration {#exploration}
 
@@
 on exploring relationships in missing values, and identifying possible patterns.
-Texts covering exploration of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools [@VanBuuren2012]; this is inefficient.
+Texts covering the exploration phase of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools [@VanBuuren2012]; this is inefficient.
 Methods for exploring missing values are primarily covered in literature on interactive graphics [@Swayne1998; @Unwin1996; @Cook2007], and are picked up again in a discussion of a graphical user interface [@Cheng2015].
 
-The missingness matrix $R$ can be used to assess missing data dependence, and has  been used in interactive graphics, dubbed a "shadow matrix", to link missing and imputed values to the data, facilitating their display in graphics [@Swayne1998].
-This work also focuses heavily on multivariate numeric data.
+The missingness matrix $R$ can be used to assess missing data dependence It has been used in interactive graphics, dubbed a "shadow matrix", to link missing and imputed values to the data, facilitating their display [@Swayne1998], focusing heavily on multivariate numeric data.
 This is an idea upon which this new work builds.
 
-The approach of the MANET software in @Unwin1996 focussed on multivariate categorical data, where missingness is explicitly added as an additional category.
+The MANET (Missings Are Now Equally Treated) software @Unwin1996 focussed on multivariate categorical data, with missingness explicitly added as a category.
 MANET also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot.
@@
 
-MissingDataGUI provides a user interface for exploring missing data structure both numerically and visually.
-Using a Graphical User Interface (GUI) to explore missing data may make it easier to glean valuable insight into important structures for missingness.
-However, this comes as a trade off with GUIs - the insights are typically not captured or recorded, so it is difficult to incorporate them into reproducible analyses.
+MissingDataGUI provides a Graphhical User Interface (GUI) for exploring missing data structure both numerically and visually.
+Using a GUI to explore missing data facilitates rapid insight into missingness structures.
+However, this comes as a trade off, as insights are not captured or recorded with a GUI, making it challenging to incorporate into reproducible analyses.
 This distracts and breaks analysis workflow, inviting mistakes.
 
-VIM (Visualizing and Imputing Missing Data) provides visualization methods to identify observed, imputed, and missing values.
+VIM (Visualizing and Imputing Missing Data) provides visualization methods to identify and explore observed, imputed, and missing values.
 These include spinograms, spinoplots, missingness matrices, plotting missingness in the margins of other plots, and other summaries.
-These help explore missingness, however these visualizations do not map variables to graphical aesthetics, creating friction moving across workflows, making them difficult to extend to new circumstances.
-Additionally, data used to create the visualizations cannot be accessed, posing a barrier to further explorations.
+However, these visualizations do not map variables to graphical aesthetics, creating friction when moving through analysis workflows, making them difficult to extend to new circumstances.
+Additionally, data used to create the visualizations cannot be accessed, posing a barrier to further exploration.
 
@@
 This has some limitations.
-For example, for data with school grades and test scores, a boxplot visualization could have school grade mapped to the x axis, and test score to the y axis (See Figure \@ref(fig:gg-box-na)).
-If there are missings in a continuous variable like test score, \pkg{ggplot2} omits the missing values and prints a warning message.
-However, if a discrete variable like school year has missing values, an NA category is created for school year, and the scores are placed into this (Figure \@ref(fig:gg-box-na)).
+For example, data of school grades and test scores, a boxplot visualization with school grade and test score mapped to the x- and y-axis (Figure \@ref(fig:gg-box-na)).
+If there are missings in a continuous variable like test score, \pkg{ggplot2} omits the missings and prints a warning message.
+However, if a discrete variable like school year has missing values, an NA category is created for school year, where scores are placed (Figure \@ref(fig:gg-box-na)).
 
@@
 
-```{r gg-box-na, fig.height = 3, fig.width = 9, out.width = "100%", fig.align = "center", fig.cap = "ggplot2 provides different visualizations depending on what type of data has missing values.
-(A) Data is complete and graphic is presented; (B) Missings are only in category variable year - an additional 'NA' boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and school year, additional missing information is shown.
+```{r gg-box-na, fig.height = 3, fig.width = 9, out.width = "100%", fig.align = "center", fig.cap = "ggplot2 provides different visualizations depending on what type of data has missing values for data of student test scores in school year.
+(A) Data is complete; (B) Missings are only in year - an additional 'NA' boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and year, additional missing information is shown.
 The missingness category is only shown when there are missings in categorical variables such as year (plots (B) and (D)).

### 2020-01-09 paper/tidy-missing-data-paper.Rmd
tidying up to graphics section

@@
 Applying tidyverse principles to the domain of missing data clarifies and
-facilitates missing data exploration, visualization, and imputation. This section
-discusses how these principles are applied, and proceeds in three parts. Section
-\@ref(data-structure) discusses _nabular_ data, a tidy data structure for
-missing values that tracks missings throughout the analysis and allows for
-special missing values. Section \@ref(graphics) covers visualizations to
-understand missingness. Section \@ref(num-sum) discusses data summaries, and
-Section \@ref(verbs) covers the common "verbs" used to solve problems
-when working with missing data. The design and naming of these features is an
-important component that helps the analyst compose them efficiently in data pipelines, and is discussed throughout.
+facilitates missing data exploration, visualization, and imputation. This
+section discusses how these principles, in three parts: tidy data structures for missing data \@ref(data-structure), common operations (verbs) of missing data \@ref(verbs), graphics \@ref(graphics), and data summaries \@ref(num-sum).
+The design and naming of these features is discussed throughout, as they are an important component to help the analyst compose them efficiently in data pipelines.
 
@@
 To explore missing data there needs to be a robust representation of missing
-values that integrate into analysis pipelines. The R matrix, where 0 and 1
-indicate not missing and missing provides a template, as it is very common in
-missing data literature.  This matrix has been used to explore missing values in
-the interactive graphics library XGobi, where it was called a "missing value
+values that integrate into analysis pipelines.
+This means a data structure that is easily transported, and not a special class reliant on attributes that can be lost in analysis.
+The R matrix, where 0 and 1
+indicate not missing and missing provides a useful template, as it is very common in
+missing data literature.  This matrix was used to explore missing values in
+the interactive graphics library XGobi, called a "missing value
 shadow", or "shadow matrix", defined as a copy of the original data with
@@
 0 indicate a the absence of observation, or the presence of a missing value? The
-shadow matrix was specially linked to the data, but could not be referred to
-directly, and was instead linked with special window views. This means
-the data structure was abstracted from the user to facilitate data exploration.
-This made sense at the time with the state of the art for graphical analysis,
-but the data structure needs to be clearly presented to the user so they
-can use it more flexibly in other data analyses.
+shadow matrix was specially linked to the data, but could not be access
+directly, instead linked with special window views.
+The data structure was abstracted from the user to facilitate data exploration, which was consistent with the state of the art.
+However, the data structure needs to be clearly presented to the user so they
+can apply it more flexibly in data analysis.
 
@@
 
-Four features are added to the shadow matrix to facilitate analysis:
-
-1. Coordinated names: Variables in the shadow matrix gain the same name as in the data, with the suffix "_NA".
+We propose a new form for tidy representation of missing data.
+It is simple, and flexible, building off of past research.
+Four features are added to the shadow matrix to facilitate analysis of missing data:
 
@@
 
-1. Cohesiveness: Binding the shadow matrix column-wise to the original data creates a cohesive _nabular_ data form, useful for visualization and summaries.
+1. Coordinated names: Variables in the shadow matrix gain the same name as in the data, with the suffix "_NA".
+
+1. Cohesiveness: Binding the shadow matrix column-wise to the original data creates a cohesive _nabular_ data form, useful for visualization, summaries, and tracking imputed values.
 
 These additions are now discussed in turn.
+
+### Missing value labels {#missing-labels}
+
+The $R$ matrix is a very common representation of missing values.
+It can be created in the "R" language with \code{is.na(data)}.
+The values 0 and 1 (or FALSE and TRUE) can be confusing, however, as it is not immediately clear what they represent: does 0 indicate a the absence (0) of observation, or the presence (1) of a missing value?
+This is solved by adding a simple label to the shadow matrix: "NA" and "!NA" for missing and not missing.
+Values in the shadow matrix can be represented as "factor" types values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing (Figure \@ref(fig:shadow-matrix-progress) parts 1-3).
+
+### Special missings {#special-missings}
+
+Extending missing value labels to include "special" missing values labelled as "NA_\<suffix\>" allows for representing additional missingness features such as instrument failure and drop-out as "NA_instr_fail", and "NA_drop_out", and even imputations (Table \@ref(tab:shadow-encoding)).
+The underlying number representation changes from binary to integer, allowing for unlimited missingness features.
+For example, instrument failure and drop out would be represented as 2 and 3 (Table \@ref(tab:shadow-encoding)).
+Encoding special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in \pkg{naniar} (\@ref(verbs-recode)).
+
+Other statistical programs represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` through to `.z`.
+The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state.
+Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R.
+
+```{r shadow-encoding}
+
+fig_shadow_code <- tibble::tibble(temp = c(-99, NA, -1, 106),
+                                  temp_NA = c("NA_instr",NA, "NA_dropout","!NA"),
+                                  NA_value = c(2, 1, 3, 0))
+
+knitr::kable(fig_shadow_code,
+             caption = "Example data of temperature, its shadow representation, and underlying NA value.
+The temperature value of -99 is represented in the shadow column as 'NA\\_instr', which in turn has the underlying numeric value of 2.
+This captures additional information about the data otherwise difficult to record.")
+             # booktabs = TRUE) 
+  # kable_styling(latex_options = c("hold_position"))
+
+```
 
@@
 This subtle change is important as it shifts the focus from the value of a variable to its missingness state, making intent clear when performing analysis.
-
-## Missing value labels {#missing-labels}
-
-The $R$ matrix is a very common representation of missing values, and can be easily created in the "R" programming language with \code{is.na(data)}.
-The values 0 and 1 (or equally, FALSE and TRUE) can be confusing, however, as it is not immediately clear what they represent - does 0 indicate a the absence of observation, or the presence of a missing value?
-This is solved by adding a simple label to the shadow matrix: "NA" and "!NA" for missing and not missing, represented as 0 and 1 (see Figure \@ref(fig:shadow-matrix-progress) parts 2 and 3 below).
-
-## Special missings {#special-missings}
-
-Values in the shadow matrix are "factor" type values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing.
-Extending these to include "special" missing values labelled as "NA_\<suffix\>" allows for representing additional missingness features such as instrument failure and drop-out as "NA_instr_fail", and "NA_drop_out".
-The underlying number representation changes from binary to integer, allowing for unlimited missingness features.
-For example, instrument failure and drop out would be represented as 2 and 3, see Table \@ref(tab:shadow-encoding) for an example.
-Encoding these special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in \pkg{naniar}, shown in Section \@ref(verbs-recode).
-Special missing values do exist in other statistical programming languages, although are limited to 26 or 52 special types of missingness.
-
-```{r shadow-encoding}
-
-fig_shadow_code <- tibble::tibble(temp = c(-99, NA, -1, 106),
-                                  temp_NA = c("NA_instr",NA, "NA_dropout","!NA"),
-                                  NA_value = c(2, 1, 3, 0))
-
-knitr::kable(fig_shadow_code,
-             caption = "Example data of temperature, and its observed value, shadow representation, and underlying value.
-The temperature value of -99 is represented in the shadow column as 'NA\\_instr', which in turn has the underlying numeric value of 2.
-This captures additional information about the data otherwise difficult to record.")
-             # booktabs = TRUE) 
-  # kable_styling(latex_options = c("hold_position"))
-
-```
 
@@
 This means when _nabular_ data is imputed, those imputed values can easily be identified in analysis by referring to the appropriate, coordinated names.
-_Nabular_ data is created using the command `bind_shadow()` or `nabular()` (see Figure \@ref(fig:nabularfig)).
+_Nabular_ data is created with `nabular(data)` (see Figure \@ref(fig:nabularfig)).
 _Nabular_ data is not unlike classical data formats with quality or flag columns associated with each measured variable, e.g.
@@
 
-The _nabular_ data structure for missing data provides key benefits over on-the-fly creating a logical matrix where TRUE means missing, and FALSE means present.
-Firstly, it can be confusing to users whether TRUE or FALSE means missing or present; `NA` and `!NA` are clearer.
+The _nabular_ data structure for missing data provides three key benefits over on-the-fly creating a logical matrix where TRUE means missing, and FALSE means present.
+Firstly, missing value labels `NA` and `!NA` are clearer.
 Secondly, special missing values cannot be added fluently during analysis with a logical matrix.
@@
 
-Other statistical programs represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` through to `.z`.
-The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state.
-
-# Missing data operations {#verbs}
+## Missing data operations {#verbs}
 
@@
 Finally, unusual or specially coded missing values can be **flagged**.
-These verbs are described in this section.
-
-## scan: Searching for common missing value labels  {#verbs-search}
+
+### scan: Searching for common missing value labels  {#verbs-search}
 
 There are many common values for "missing".
-For example, "N/A" "Not Available", "MISSING", or white space " ".
-Before replacing these values with missing values (`NA`), it is useful to first _scan_ the data to get a sense of the magnitude of the problem.
-The  `miss_scan_count()` function finds where specified values occur in data.
-For example, all occurrences of -99 can be searched for with `miss_scan_count(data, search = -99)`.
-This returns a table of the occurrences of that value in each variable.
+For example, "N/A", "MISSING", `-99`.
+Before replacing these values with missing values (`NA`), it is useful to _scan_ the data to understand the magnitude of the problem.
+`miss_scan_count()` finds where specified values occur in data.
+For example, a scan of all occurrences of -99 is done with `miss_scan_count(data, search = -99)`, returning a table of occurrences of that value for each variable.
+`common_na_numbers` and `common_na_strings` provide a list of common NA values for numbers and characters.
+They contain values like "n/a", "na", and -99, and are designed to be used with `miss_scan_count` to identify common missing values.
 
@@
 
-A list of common NA values is provided in `naniar` in the `common_na_strings` and `common_na_numbers` for characters and numbers.
-These contain values like "n/a", "na ", and ".", and -9, -99.
-These can be used with `miss_scan_count` to help identify common missing values.
-
 ## replace: Replacing missing value label with another {#verbs-replace-with}
 
-Once the magnitude of the unusual missing values has been assessed, these values need to be replaced with appropriate missing values.
+Once the magnitude of possible missing values has been assessed, these values need to be replaced with missing values.
 This can be done with `replace_with_na()`.
-This is useful when there is certainty of which values mean missing.
-For example "N/A", "N A", "Not Available", and "missing".
 A dataset could have the values -99 in the `x` column replaced with the following code:
-`replace_with_na(dat_ms, replace = list(x = -99))`
-
-For flexibility, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`.
-This means `replace_with_na` works for **all** variables with `replace_with_na_all`, **at** selected variables with `replace_with_na_at`, and **if** variables meet some condition with `replace_with_na_if`.
+`replace_with_na(dat_ms, replace = list(x = -99))` For flexibility, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`.
+This means `replace_with_na_all` operates on **all** columns, `replace_with_na_at` operates **at** specific columns, and `replace_with_na_if` operates on columns **if** they meet some condition (such as `is.numeric` or `is.character`).
 
@@
 Functions starting with `add_` exist in `dplyr` to add count information for particular groups or conditions to the data.
-Similarly, `naniar` includes `add_` functions to add missingness summary information back to the data, such as the number or proportion of missingness, the missingness cluster, or if there are any missings, using: `add_n_miss()`
-`add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (see Table \@ref(tab:add-missing-info)).
+`naniar` includes functions to add missingness summary information to the data, such as the number or proportion of missingness, the missingness cluster, or if there are any missings, using: `add_n_miss()`, `add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (see Table \@ref(tab:add-missing-info)).
+There are also functions for adding information about shadow values or nice labels for any missing values with `add_label_shadow()` and `add_label_missings()`.
+
+An example use of these features is in @Tierney2015, where the proportion of
+missings is used as the outcome in a model, to identify variables and values
+important in predicting missingness structures.
 
@@
 
-An example use of these features is in @Tierney2015,
-where the proportion, number, or cluster of missings is used as the outcome in a
-model. The variables in the dataset can then be used to predict the outcome, identifying variables and values important in predicting
-missingness structures. There are also functions for adding information about shadow values or nice labels for if there are any missing values with `add_label_shadow()` and `add_label_missings()`.
-
 ## shadow: Creating nabular data {#verbs-nabular}
@@
 This facilitates visualization and summaries, and allows for imputed values to be appropriately tracked.
-_Nabular_ data can be created with `nabular()` or `bind_shadow()`.
-(The function was named `bind_shadow()` to borrow from existing functions named `bind` - such as `bind_rows()` and `bind_cols()` from `dplyr`, which bind rows or columns into a dataframe).
+_Nabular_ data can be created with `nabular()`.
 An example use of `nabular()` some example data set is shown below:
@@
 For example, there might be special codes to mark an individual dropping out of a study, known instrument failure in weather instruments, or for values censored in analysis.
-These special types of missingness can be encoded in the shadow matrix of  _nabular_  data, using the `recode_shadow()` function.
-This provides a fluent interface to recode, or flag, the shadow information in the shadow matrix as a special type of missing value.
+These special types of missingness can be encoded in the shadow matrix of  _nabular_  data, using `recode_shadow()`.
 Using `recode_shadow()` requires specifying the variable you want to contain the flagged value, the condition you want this to occur at, and a suffix for the new type of missing value.
@@
 `naniar` does not reinvent the wheel for imputation, as there are many robust R packages for imputation, such as `mice`, `mi`, `VIM`, and `simputation`.
-However, `naniar` does provide a few imputation methods to facilitate exploration and visualizations, which were not otherwise easily available: `impute_below` and `impute_mean` and `impute_median`.
-These imputation methods are useful to explore structure in missingness, but are not recommended for use in analysis.
-The `impute_below` function imputes values below the minimum of the data, with some jitter (small values of random noise) to reduce overplotting, but has options to change the default amount it is shifted below, and the amount of jitter.
+Instead, _nabular_ data is designed to work with these.
+`naniar` does provide a few imputation methods to facilitate exploration and visualizations: `impute_mean`, `impute_median`, and `impute_below`.
+While useful to explore structure in missingness, they are not recommended for use in analysis.
+`impute_below` imputes values below the minimum value, with some jitter (small random noise) to reduce overplotting, but has options to change the default amount it is shifted below, and the amount of jitter.
 
@@
 Similar to `simputation`, each `impute_` function returns the data with values imputed.
-However, `naniar` does not use a formula syntax, and instead each function implements "scoped variants" `_all`, `_at` and `_if`.
-Here, `_all` operates on **all** columns, `_at` operates **at** specific columns, and `_if` operates on columns **if** they meet some condition (such as `is.numeric` or `is.character`).
-The `impute_` functions with no scoped variant, e.g., `impute_mean`, will work on a single vector, but not a data.frame.
-One drawback to imputation functions in this form is imputed value locations are not tracked.
+However, `naniar` does not use a formula syntax, instead each function has "scoped variants" `_all`, `_at` and `_if` as in \@ref(verbs-replace-with).
+`impute_` functions with no scoped variant, (`impute_mean`), will work on a single vector, but not a data.frame.
+One challenge with this approach is imputed value locations are not tracked.
 This issue is resolved with \code{nabular} data covered in Section \@ref(verbs-track).
@@
 The missingness of any observation can be referred to with `any_missing` (Figure \@ref(fig:track-impute-example)).
-The code chunk below shows the track pattern, first using `nabular`, then imputing with `impute_knn`, and adding a `label_shadow`:
+The code chunk below shows the track pattern, first using `nabular`, then imputing with `impute_lm`, and adding a `label_shadow`:
 
@@
 
-Missing values can then be shown in a scatterplot by setting the `color` aesthetic in ggplot to `any_missing` (Figure \@ref(fig:track-impute-example)A), or in a density plot using the `fill = any_missing`, (Figures \@ref(fig:track-impute-example)B) and \@ref(fig:track-impute-example)C).
-Imputed values can also be compared to complete case data, grouping by `any_missing`, and then summarizing, similar to other dplyr summary workflows.
+Missing values can be mapped to a graphical element in a scatterplot by setting the `color` aesthetic in ggplot to `any_missing` (Figure \@ref(fig:track-impute-example)A), or in a density plot using the `fill = any_missing`, (Figures \@ref(fig:track-impute-example)B) and \@ref(fig:track-impute-example)C).
+Imputed values can also be compared to complete case data, grouping by `any_missing`, and then summarizing, similar to other dplyr summary workflows (shown below).
 This shows mean and median values are similar, but the minimum and maximum values are different.
 
-```{r track-impute-example, fig.show = "hold", fig.cap = "Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values imputed using simputations `impute\\_knn` function, with imputed values colored green and data values orange.
-Imputed values are similar, but slightly different to existing data.", fig.height = 4, fig.width = 12, out.width = "100%", echo = FALSE}
+```{r track-impute-example, fig.show = "hold", fig.cap = "Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values from a linear model.
+Imputed values are colored green, and data values orange.
+Imputed values are similar, but slightly trended to the mean of existing data.", fig.height = 4, fig.width = 12, out.width = "100%", echo = FALSE}
 
@@
 Visualizing missingness in a dataset is essential to understand its structure.
-This section discusses easy to remember visualizations that compose with other functions in a pipeline.
-Four different areas of visualization of missing data are discussed: overviews (Section \@ref(overview-graphics)); univariate (Section \@ref(univariate-graphics)), bivariate (Section \@ref(bivariate-graphics)), and multivariate (Section \@ref(multivariate-graphics)).
-All plots are created using \pkg{ggplot2}, giving users clearer control over the plot appearance, and customizability.
+This section discusses four different areas of visualization of missing data: overviews (\@ref(overview-graphics)), univariate (\@ref(univariate-graphics)), bivariate (\@ref(bivariate-graphics)), and multivariate (\@ref(multivariate-graphics)).
+All plots are created using \pkg{ggplot2}, giving users clearer control over the plot appearance.
 
@@
 
-```{r gg-miss-case-var, echo = FALSE, fig.show='hold', fig.cap = "Graphical summaries of missingness in variables and cases for the airquality data.
-(A) Missings in each variable and (B) in each case.
-There are missing values in Ozone and Solar.R, with Ozone having more missings, and not many cases have two missing values, with most missingness being from cases with one missing value.", fig.height = 3, fig.width = 6, out.width = "90%", fig.align="center"}
+```{r gg-miss-case-var, echo = FALSE, fig.show='hold', fig.cap = "Graphical summaries of missingness in variables (A) and cases (B) for the airquality data.
+There are missing values in Ozone and Solar.R, with Ozone having more missings.
+Not many cases have two missing values, with most missingness being from cases with one missing value.", fig.height = 3, fig.width = 6, out.width = "90%", fig.align="center"}
 

### 2020-01-09 paper/tidy-missing-data-paper.Rmd
finish improving readability / getting as close to 30 pages as possible.

@@
 In particular, there is little to no guidance for working with missing values in the field of visualization.
-Most analysis software simply drops missing values when making a plot, although some packages provide a warning (see Figure \@ref(fig:warning)).
+Most analysis software simply drops missing values when making a plot, although some packages provide a warning (Figure \@ref(fig:warning)).
 
@@
 
-Features of tidy data were formally discussed in 2014 [@Wickham2014], subsequently generating much discussion (see @Donoho2017 and commentary, @r4ds), and tools for data analysis.
-Tidy data is defined as:
+Features of tidy data were formally discussed in 2014 [@Wickham2014], subsequently generating much discussion (@Donoho2017 and commentary, @r4ds), and tools for data analysis. Tidy data is defined as:
 
@@
 \pkg{Amelia} [@amelia] assumes data are multivariate normal, and samples from the posterior using the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm [@Honaker2010] and allows for incorporation of information on the values in a prior.
+\pkg{norm} [@norm], provides multiple imputation using EM for multivariate normal data, drawing from methods in the NORM software [@schafer-norm].
+\pkg{norm} does not provide a framework for tracking missing values, instead providing tools for making inference from multiple imputation.
 
@@
 This approach promotes fluid workflow, with a similar penalty to tidying up model output, which is still messy.
-
-\pkg{norm} [@norm], provides multiple imputation using EM for multivariate normal data, drawing from methods in the NORM software [@schafer-norm].
-\pkg{norm} does not provide a framework for tracking missing values, instead providing tools for making inference from multiple imputation.
 
@@
 
-1.
-Missing value label: A simple label to the shadow matrix clear, "NA" and "!NA" for missing and not missing, represented as 0 and 1 (see Figure \@ref(fig:shadow-matrix-progress) parts 2 and 3 below).
+1. Missing value label: A simple label to the shadow matrix clear, "NA" and "!NA" for missing and not missing, represented as 0 and 1 (Figure \@ref(fig:shadow-matrix-progress) parts 2 and 3).
 
@@
 1. Cohesiveness: Binding the shadow matrix column-wise to the original data creates a cohesive _nabular_ data form, useful for visualization, summaries, and tracking imputed values.
-
-These additions are now discussed in turn.
 
@@
 
-Other statistical programs represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` through to `.z`.
+Other statistical programs represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` - `.z`.
 The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state.
@@
 
-A consistent short suffix "`_NA`" for each variable keeps names coordinated throughout analysis (See Figure \@ref(fig:shadow-matrix-progress) part 4).
+A consistent short suffix "`_NA`" for each variable keeps names coordinated throughout analysis (Figure \@ref(fig:shadow-matrix-progress) part 4).
 The suffix is short and easy to remember in analysis or visualization.
@@
 
-_Nabular_ data binds the shadow matrix column-wise to the original data, and is so named as a portmanteau of `NA` and `tabular`.
-_Nabular_ data keeps corresponding rows together, removing the possibility of mismatching records, explicitly linking missing values to the data.
+_Nabular_ data binds the shadow matrix column-wise to the original data.
+It is a portmanteau of `NA` and `tabular`.
+_Nabular_ data keeps corresponding rows together, removing the possibility of mismatching records, explicitly linking missing values to data.
 _Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable, `var`, as `var_NA`.
 _Nabular_ data is a snapshot of the missingness of the data.
-This means when _nabular_ data is imputed, those imputed values can easily be identified in analysis by referring to the appropriate, coordinated names.
-_Nabular_ data is created with `nabular(data)` (see Figure \@ref(fig:nabularfig)).
+This means when _nabular_ data is imputed, those imputed values can easily be identified in analysis by referring to the coordinated names.
+_Nabular_ data is created with `nabular(data)` (Figure \@ref(fig:nabularfig)).
 _Nabular_ data is not unlike classical data formats with quality or flag columns associated with each measured variable, e.g.
@@
 
-The _nabular_ data structure for missing data provides three key benefits over on-the-fly creating a logical matrix where TRUE means missing, and FALSE means present.
-Firstly, missing value labels `NA` and `!NA` are clearer.
-Secondly, special missing values cannot be added fluently during analysis with a logical matrix.
+The _nabular_ data structure provides three key benefits over on-the-fly creating a missingness matrix $R$.
+Firstly, missing value labels `NA` and `!NA` are clearer than TRUE or FALSE.
+Secondly, special missing values cannot be added easily during analysis with a logical matrix.
 Finally, the logical matrix cannot capture which values are imputed, if imputation has already taken place.
@@
 
-Common operations on missing data can be considered to be verbs: scan, replace, add, shadow, impute, track, and flag.
+Common missing data operations can be considered verbs: scan, replace, add, shadow, impute, track, and flag.
 Missing values can be **scanned** to find possible missings not coded as `NA`.
-These values can then be **replaced** with an `NA` value.
+These values can then be **replaced** with `NA`.
 To facilitate exploration, summaries of missingness can be **added** as a summary column to the original data.
-Data values can be augmented with **shadow** values, which helps explore missing data, as well as facilitating the process of **imputing**, and **tracking**.
+Data values can be augmented with **shadow** values, helping explore missing data, as well as facilitating the process of **imputing**, and **tracking**.
 Finally, unusual or specially coded missing values can be **flagged**.
@@
 Functions starting with `add_` exist in `dplyr` to add count information for particular groups or conditions to the data.
-`naniar` includes functions to add missingness summary information to the data, such as the number or proportion of missingness, the missingness cluster, or if there are any missings, using: `add_n_miss()`, `add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (see Table \@ref(tab:add-missing-info)).
+`naniar` includes functions to add missingness summary information to the data, such as the number or proportion of missingness, the missingness cluster, or if there are any missings, using: `add_n_miss()`, `add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (Table \@ref(tab:add-missing-info)).
 There are also functions for adding information about shadow values or nice labels for any missing values with `add_label_shadow()` and `add_label_missings()`.
-
 An example use of these features is in @Tierney2015, where the proportion of
@@
 
-_Nabular_ data has the shadow matrix column-bound to the existing data.
-This facilitates visualization and summaries, and allows for imputed values to be appropriately tracked.
-_Nabular_ data can be created with `nabular()`.
-An example use of `nabular()` some example data set is shown below:
+_Nabular_ data has the shadow matrix column-bound to existing data.
+This facilitates visualization and summaries, and allows for imputed values to be tracked.
+_Nabular_ data can be created with `nabular()`, as below:
 
@@
 ```{r recode-shadow, eval = TRUE, echo = TRUE}
-dat_ms %>%
-  nabular() %>%
+nabular(dat_ms) %>%
   recode_shadow(x = .where(x == -99 ~ "broken_sensor"))
@@
 
-`naniar` does not reinvent the wheel for imputation, as there are many robust R packages for imputation, such as `mice`, `mi`, `VIM`, and `simputation`.
-Instead, _nabular_ data is designed to work with these.
-`naniar` does provide a few imputation methods to facilitate exploration and visualizations: `impute_mean`, `impute_median`, and `impute_below`.
+`naniar` does not reinvent the wheel for imputation, as there are many robust R packages for imputation \@ref(imputation).
+Instead, _nabular_ data is designed to work with existing imeplementation.
+`naniar` does provide a few imputation methods to facilitate exploration and visualization: `impute_mean`, `impute_median`, and `impute_below`.
 While useful to explore structure in missingness, they are not recommended for use in analysis.
-`impute_below` imputes values below the minimum value, with some jitter (small random noise) to reduce overplotting, but has options to change the default amount it is shifted below, and the amount of jitter.
+`impute_below` imputes values below the minimum value, with some jitter (small random noise) to reduce overplotting.
+It has options to change the default amount of shift, and jitter.
 
@@
 ```{r bind-impute-label-example, echo = TRUE}
-
-aq_imputed <- airquality %>%
-  nabular() %>%
+aq_imputed <- nabular(airquality) %>%
   simputation::impute_lm(Ozone ~ Temp + Wind) %>%
@@
 
-aq_imputed
+head(aq_imputed)
 ```
@@
 Imputed values are colored green, and data values orange.
-Imputed values are similar, but slightly trended to the mean of existing data.", fig.height = 4, fig.width = 12, out.width = "100%", echo = FALSE}
+Imputed values are similar, but slightly trended to the mean.", fig.height = 4, fig.width = 12, out.width = "100%", echo = FALSE}
 
@@
 
-The number of missings in each variable and case are visualized using `gg_miss_var` and `gg_miss_case` (see Figure \@ref(fig:gg-miss-case-var) below).
-These are shown using the "airquality" dataset, included in base R, which contains daily air quality measurements in New York, from May to September, 1973.
-Rather than highlighting the amount of complete data, these visualizations draw attention to the amount of missings, ordering by missingness.
+Rather than highlighting the amount of complete data, missing data visualizations should draw attention to the amount of missings, and ordering by missingness.
+Overview visualisations for variables and cases are provided with `gg_miss_var` and `gg_miss_case` (Figure \@ref(fig:gg-miss-case-var)).
+These are shown using the "airquality" dataset, included in base R, containing daily air quality measurements in New York, from May to September, 1973.
 
@@
 There are missing values in Ozone and Solar.R, with Ozone having more missings.
-Not many cases have two missing values, with most missingness being from cases with one missing value.", fig.height = 3, fig.width = 6, out.width = "90%", fig.align="center"}
+Not many cases have two missings.
+Most missingness is from cases with one missing value.", fig.height = 3, fig.width = 6, out.width = "90%", fig.align="center"}
 
@@
 
-All missings in a dataset can be displayed using a heatmap style visualization.
-This is achieved using `vis_miss()` from the \pkg{visdat} package [@visdat], which also provides summaries of missingness overall in the legend, and for each column (see Figure \@ref(fig:vis-miss)).
+All missings can be displayed using a heatmap style visualization.
+This is achieved using `vis_miss()` from the \pkg{visdat} package [@visdat], which also provides summaries of missingness overall in the legend, and for each column (Figure \@ref(fig:vis-miss)).
 The user can also apply clustering to the rows, and arrange columns by missingness (Figure \@ref(fig:vis-miss)B).
@@
 A key improvement is `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns in the visualization and are named at the top, and each row is an observation.
-Using \pkg{ggplot2} means the plot can be customized and combined with other ggplot graphics.
+Using \pkg{ggplot2} makes the plot easily customizable.
 
@@
 
-The number of times certain variables go missing together can be visualized using an "upset plot" [@Conway2017], a visualization technique for showing the size and features of sets in data similar to a venn diagram, but scales better with more variables (Figure \@ref(fig:airquality-upset)).
+The number of times certain variables go missing together can be visualized using an "upset plot" [@Conway2017].
+Similar to a Venn diagram, an upset plot shows the size and features of overlapping sets, but scales well with more variables.
 Figure \@ref(fig:airquality-upset) shows only Ozone and Solar.R have missing values.
-The dots in the bottom right show the combinations of missingness, with the bottom left showing the missingness in each variable, and the top showing the missingness for the combinations.
-The bottom left shows Ozone has the most missing values, and the top shows 2 cases where both Solar.R and Ozone have missing values together.
+The bottom right show the combinations of missingness, the bottom left showing missingness in each variable, and the top showing the combinations of missingness.
+The bottom left shows Ozone has the most missing values, and the top shows 2 cases where both Solar.R and Ozone go missing together.
 
@@
 Only Ozone and Solar.R have missing values, and Ozone has the most missing values.
-There are 2 cases where both Solar.R and Ozone have missing values.", fig.height = 3, fig.width = 4, out.width = "80%"}
+There are 2 cases where both Solar.R and Ozone have missing values.", fig.height = 3, fig.width = 4, out.width = "75%"}
 
@@
 To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed values below the range of data using `impute_below` (by default imputing 10% below the range).
-Visualizing this as a histogram shows the number of missing values on its left.
-Ozone is shown in a different color by referring to the variable `Ozone` as `Ozone_NA` in a fill aesthetic in ggplot, shown in Figure \@ref(fig:impute-shift-histogram).
+Figure \@ref(fig:impute-shift-histogram) shows a histogram of Ozone, the number of missing values on its left.
+Ozone is shown in a different color by referring to the variable `Ozone` as `Ozone_NA` in a fill aesthetic in ggplot.
 
@@
 
-```{r bind-shadow-density, fig.height = 3, fig.width = 5, out.width = "49%", error = FALSE, message = FALSE, fig.cap = "A visualization of temperature according to missingness in ozone from in the airquality dataset.
+```{r bind-shadow-density, fig.height = 3, fig.width = 5, out.width = "49%", error = FALSE, message = FALSE, fig.cap = "Temperature according to missingness in ozone from in the airquality dataset.
 A histogram of temperature facetted by the missingness of ozone (A), or a density of temperature colored by missingness in ozone (B).
-The distribution shows a cluster of low temperature observations with missing ozone values, but the temperature values are otherwise similar.", fig.show = 'hold'}
+These show a cluster of low temperature observations with missing ozone values, but temperature is otherwise similar.", fig.show = 'hold'}
 
@@
 To visualize missing values in two dimensions the missing values can be placed in plot margins, by imputing values below the range of the data.
-As in Section \@ref(univariate-graphics), using _nabular_ data identifies imputed values, and color makes missingness pre-attentive [@treisman1985].
-The steps of imputing and coloring are combined into `geom_miss_point()` from the \pkg{naniar} package (Figure \@ref(fig:geom-miss)).
-There is a mostly uniform spread of missing values for Solar.R and Ozone.
+Using _nabular_ data identifies imputed values (\@ref(univariate-graphics)), and color makes missingness pre-attentive [@treisman1985].
+The steps of imputing and coloring are combined into `geom_miss_point()`.
+Figure \@ref(fig:geom-miss) shows a mostly uniform spread of missing values for Solar.R and Ozone.
+As `geom_miss_point()` is a defined \pkg{ggplot2} geometry, it works with features such as facetting and mapping other variables to graphical aesthetics.
+The shadow format by itself is also useful for exploring patterns of missingness.
 
@@
 ```
-
-As `geom_miss_point` is a defined geometry for \pkg{ggplot2}, it provides features such as faceting and mapping other variables to aesthetics such as color, shape, or size.
-The shadow format by itself can also be used to visually explore some patterns of missingness.
 
@@
 Parallel coordinate plots can visualize missingness beyond two dimensions.
-They transform variables to the same scale, and typically make the data values range between 0 and 1.
-To showcase this visualization, the dataset `oceanbuoys` from \pkg{naniary} is used, which contains measurements of moored ocean buoys to understand and predict El Niño and El Niña.
-The data were collected in 1993 and 1997, and contains information on the sea and air temperature, humidity, and east west and north south wind directions.
-
-Figure \@ref(fig:parallel-cord-plot) shows a parallel coordinate plot of the `oceanbuoys` data, with missing values imputed to be 10% below the range, and values colored according to whether humidity contained missing values.
-Although a 10% below imputation is not the most ideal way to display missing values in a parallel coordinate plot, it shows humidity is missing at low air and sea temperatures, and humidity is missing in one year, and one location.
+They transform variables to the same scale, ranging between 0 and 1.
+The `oceanbuoys` dataset from \pkg{naniar} is used for this visualisation, containing measurements of ocean buoys to understand and predict El Niño and El Niña.
+Data was collected in 1993 and 1997, and contains measurements of sea and air temperature, humidity, and east west and north south wind speeds.
+Figure \@ref(fig:parallel-cord-plot) shows a parallel coordinate plot of the `oceanbuoys`, with missing values imputed to be 10% below the range, and values colored according to where humidity was missing (humidity_NA).
+This shows humidity is missing at low air and sea temperatures, and humidity is missing in one year, and one location.
 
@@
 
-There are different ways to summarize the amount of missing and complete values in a dataset.
-This section describes approaches to summarizing missingness, implemented in the \pkg{naniar} package.
-The functions are easy to remember as they follow consistent naming, and their output is consistent, and returns a single number or dataframe.
-This means they integrate well with plotting and modelling tools.
-Section \@ref(single-num-sum) discusses single number summaries; Section \@ref(sum-tab-missings) discusses variable- and case-wise summaries; and Section \@ref(num-sum-w-group) shows how the design of these tools works with other tools in an analysis pipeline.
+This section describes approaches to summarizing missingness.
+The functions are implemented in \pkg{naniar}, and are easy to remember as they have consistent names and output, returning either a single number \@ref(single-num-sum), or a dataframe (\@ref(sum-tab-missings)), integrating well with plotting and modelling tools.
+How these work with other tools in an analysis pipeline is shown in \@ref(num-sum-w-group).
 
@@
 
-The overall number, proportion, or percent of missing values in a dataset is shown with `n_miss`, `prop_miss` and `pct_miss`.
-Complementing these are functions to reveal the number of complete values: `n_complete`, `prop_complete`, and `pct_complete`.
-These can summarize missings for variables and cases by appending `_case` or `_var` to any of the summaries.
-Table \@ref(tab:n-prop-pct-miss-complete) shows the names and output of these functions on the dataset 'airquality'.
+The overall number, proportion, or percent of missing values in a dataset is shown with `n_miss`, `prop_miss` and `pct_miss`, and their complete value complements: `n_complete`, `prop_complete`, and `pct_complete`.
+Summaries for variables and cases are made by appending `_case` or `_var` to any of the summaries.
+An overview is shown in Table \@ref(tab:n-prop-pct-miss-complete).
 
@@
 Another useful approach is to tabulate the frequency of missing values for each variable or case; that is, the number of times there are zero, one, two, and so on, missing values.
-These summaries and tabulations are shown for variable in Tables \@ref(tab:miss-var-summary) and \@ref(tab:miss-var-table), using the `miss_var_summary` and `miss_var_table` functions.
-Functions are also provided for case-wise summaries and tabulations with `miss_case_summary` and `miss_case_table`, which order by `n_miss`, so the most missings are always shown at the top.
-Other summaries provided in \pkg{naniar} include summaries of missingness across a provided repeating span, `miss_var_span`, and finding the streaks or runs of missingness in a given variable in `miss_var_run`.
+These summaries and tabulations are shown for variable in Tables \@ref(tab:miss-var-summary) and \@ref(tab:miss-var-table), using `miss_var_summary` and `miss_var_table`.
+Functions are also provided for case-wise summaries and tabulations: `miss_case_summary` and `miss_case_table`, which order rows by `n_miss`, to show the most missings at the top.
+`naniar` also shows missingness summaries across a repeating span, `miss_var_span`, and finding streaks or runs of missingness in a given variable in `miss_var_run`.
 
@@
   knitr::kable(
-    caption = "The \\texttt{miss\\char`_var\\char`_summary} function provides a summary of the number and percent of missings in each variable. Only ozone and solar radiation have missing values.",
+    caption = "\\texttt{miss\\char`_var\\char`_summary} provides the number and percent of missings in each variable in airquality. Only ozone and solar radiation have missing values.",
     digits = 1)
@@
   knitr::kable(
-    caption = "The \\texttt{miss\\char`_var\\char`_table} function provides a tabulation of the amount of missing data in each variable.
-This shows the number of variables with no missings, 7 missings, and 37 missings, and the percentage of variables with those amounts of missingness.
-There are not many patterns of missingness.",
+    caption = "\\texttt{miss\\char`_var\\char`_table} tabulates the amount of missing data in each variable in airquality.
+This shows the number of variables with 0, 7, and 37 missings, and the percentage of variables with those amounts of missingness.
+There are few missingness patterns.",
     digits = 1)
@@
 
-These summaries and tabulations can be composed with the "pipe" operator to perform grouped summaries using the `group_by` operator from `dplyr`.
-These summaries return dataframes, which may be used by other plotting devices, modelling tools, or other summaries.
-Table \@ref(tab:group-miss-var-summary) shows an example of the missing data summaries for the airquality dataset in each variable, for each month of the year.
-These datasets can be summarized in visualizations, or values input into a model.
+These summaries and tabulations can be composed with the "pipe" operator to perform grouped summaries with `dplyr`'s `group_by` function.
+These summaries return dataframes, easily used in other plotting devices, modelling tools, or other summaries.
+Table \@ref(tab:group-miss-var-summary) shows an example of missing data summaries for airquality by each month.
 
@@
   slice(1:10) %>%
-  knitr::kable(caption = "The \\texttt{miss\\char`_var\\char`_summary} provides a grouped summary of the missingness in each variable, for each Month of the airquality dataset, with only the first 10 rows of output shown.
+  knitr::kable(caption = "\\texttt{miss\\char`_var\\char`_summary} combined with \\texttt{group\\char`_by} provides a grouped summary of the missingness in each variable, for each Month of the airquality dataset.
+Only the first 10 rows are shown.
 There are more ozone missings in June than May.",
@@
 There are `r scales::number(nrow(housing), big.mark = ",")` properties, and `r ncol(housing)` variables in the dataset.
-The variables include the type of real estate (town house, unit, house), suburb, method of selling, number of rooms, price, real estate agent, date of sale, and the distance from the Central Business District (CBD).
+The variables include real estate type (town house, unit, house), suburb, selling method, number of rooms, price, real estate agent, sale date, and distance from Central Business District (CBD).
 
@@
 
-```{r housing-miss-case-var, fig.cap = "The amount of missings in each variable (A) and in each case (B) for Melbourne housing data.
-(A) Build area and year built have more than 50\\% missing values, and car, bathroom, bedroom2 and longitude and latitude have about 25\\% missings.
-(B) There are between 5 and 50\\% missing values in cases.
-There are many missing values in the data, with the majority of missingness being in selected cases and variables.", fig.height = 4, fig.width = 8, out.width = "100%"}
+```{r housing-miss-case-var, fig.cap = "The amount of missings in variables (A) and cases (B) for Melbourne housing data.
+(A) Build area and year built have more than 50\\% missing, and car, bathroom, bedroom2 and longitude and latitude have about 25\\% missings.
+(B) Cases are missing 5 - 50\\% of values.
+The majority of missingness is in selected cases and variables.", fig.height = 4, fig.width = 8, out.width = "100%"}
 
@@
 
-Possible missingness structures are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \@ref(fig:applic-vis-miss)).
-There are three main clusters of missing data: at the top many variables go missing together; in the middle, building price and year built, at the bottom, building area, year built, and land size.
-
-```{r applic-vis-miss, fig.height = 4, fig.width = 6, out.width = "85%", fig.show='hold', fig.cap = "Heatmap of clustered missingness for the housing data.
-Three groups of missingness are apparent, at the top for building area to longitude, the middle for building area and year built, at the end for building area, year built, and landsize.
-There is some structure in the missings.", dev = "png", dpi = 300}
+Three missingness clusters are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \@ref(fig:applic-vis-miss)).
+
+```{r applic-vis-miss, fig.height = 4, fig.width = 6, out.width = "85%", fig.show='hold', fig.cap = "Heatmap of clustered missingness for housing data reveals structured missingness.
+Three groups of missingness are apparent.
+At the top: building area to longitude; the middle: building area and year built; the end: building area, year built, and landsize.", dev = "png", dpi = 300}
 
@@
 
-Missingness patterns can also be shown with an `upset` plot [@Conway2017], to display the 8 of intersecting sets of missing variables (see Figure \@ref(fig:applic-vis-miss)).
+Figure \@ref(fig:applic-vis-miss) shows missingness patterns with an `upset` plot [@Conway2017], displaying 8 intersecting sets of missing variables.
 Two patterns stand out: two, and five variables missing, providing further evidence of the missingness patterns seen in Figures \@ref(fig:housing-miss-case-var) and \@ref(fig:housing-upset).
 
-```{r housing-upset, fig.height = 5, fig.width = 8, fig.cap = "An upset plot of eight sets of missingness in the housing data.
+```{r housing-upset, fig.height = 5, fig.width = 8, fig.cap = "An upset plot of 8 sets of missingness in the housing data.
+Missingness for each variable is shown on the bottom left.
+Connected dots show co-occurences of missings in variables.
 Two missingness patterns are clear, year built and building area, and lattitude through to building area.", out.width = "100%"}
@@
 
-Tabulating the number of missings in variables (see Table \@ref(tab:housing-miss-var-case-table) (left)) shows three groupings: 6254 - 6824 missings cases in variables, 9265 in another variable, and between 15163 and 16736 missing.
-Tabulating missings in cases (Table \@ref(tab:housing-miss-var-case-table) (right)) shows 32% of cases have complete data, 26% have 2 missing, 22% have 8 missing, 12% have 1 missing, and 5% have 3 missing, the remaining cases with missings are less than 1%.
-Two variables with more than 50% missingness are omitted from analysis: building area and year built.
+Tabulating the number of missings in variables in Table \@ref(tab:housing-miss-var-case-table) (left) shows three groups of missingness.
+Tabulating missings in cases (Table \@ref(tab:housing-miss-var-case-table) (right) shows six patterns of missingness.
+These overview plots lead to the removal of two variables from with more than 50% missingness from analysis: building area and year built.
 
@@
              caption = "Tabulating missingness for variables (left) and cases (right) to understand missingness patterns.
-14 variables have between 0 and 3 missings, then 6 variables have 6000 - 9000 missings, and 2 variables have 15 - 16,000 missing values.
-About 30\\% of cases have no missing values, then 45\\% of cases have between 1 and 6 missing values, and then about 23\\% of cases have 8 or more missings.
+14 variables have 0 - 3 missings, 6 variables have 6000 - 9000 missings, and 2 variables have 15 - 16,000 missings.
+About 30\\% of cases have no missings, 45\\% of cases have 1 - 6 missings, and about 23\\% of cases have 8 or more missings.
 There are different patterns of missingness in variables and cases, but they can be broken down into smaller groups.",
@@
 
-Using the information from the overview visualizations and tables, the following variables are explored for features predicting missingness: Land size, latitude, longitude, bedroom2, bathroom, car, and land size.
-
-Missingness structure can be better understood and evaluated by using the data to predict where the data goes missing.
-This can be achieved by clustering the missing values into groups, and then applying classification and regression trees (CART) to find those variables and values predicting missingness clusters [@Tierney2015; @Barnett2017].
-Two clusters are identified, one for each missingness cluster.
-The missingness clusters are predicted using all variables in the dataset with the CART package `rpart` [@rpart], and plotted using the `rpart.plot` package [@rpart-plot].
-Importance scores from the CART model describe those variables important for predicting missingness: rooms, price, suburb, council area, distance, and region name.
+Using information from \@ref(case-study-explore-pattern), the following variables are explored for features predicting missingness: Land size, latitude, longitude, bedroom2, bathroom, car, and land size.
+
+Missingness structure is explored by clustering the missing values into groups.
+Then, a  classification and regression trees (CART) model is applied to predict these missingness clusters using the remaining variables, to find variables and values that predicting missingness clusters [@Tierney2015; @Barnett2017].
+Two clusters of missingness are identified and predicted using all variables in the dataset with the CART package `rpart` [@rpart], and plotted using the `rpart.plot` package [@rpart-plot].
+Importance scores reveal variables important for predicting missingness: rooms, price, suburb, council area, distance, and region name.
 These variables are important for predicting missingness, so are important to include in the imputation model.
@@
 
-```{r rpart-plot, fig.height = 4, fig.width = 8, out.width = "100%", fig.align = "center", fig.cap = "Decision tree output predicting the clusters of missingness.
-Type of house, the year quarter, and year were important for predicting missingness cluster.
+```{r rpart-plot, fig.height = 4, fig.width = 8, out.width = "90%", fig.align = "center", fig.cap = "Decision tree output predicting missingness clusters.
+Type of house, year quarter, and year were important for predicting missingness cluster.
 The cluster with the most missingness was for quarters 1 and 4, for 2017 and 2018.
-Type of house, year, and year quarter are important features related to the missingness structure."}
+Type of house, year, and year quarter are important features related to missingness structure."}
 
@@
 
-Two different imputation techniques are used: simple linear regression and K nearest neighbors.
+\pkg{simputation} is used to implement two imputation: simple linear regression and K nearest neighbors.
 Values are imputed stepwise in ascending order of missingness.
-The `simputation` R package [@simputation] is used to impute the values, applying the track missings pattern described in Section \@ref(verbs-track), to assess imputed values.
+The  track missings pattern is applied (described in \@ref(verbs-track)), to assess imputed values.
+Imputed datasets are compared on their performance in a model predicting log house price for 4 variables (Figure \@ref(fig:imputed-by-model)).
+Compared to KNN imputed values, the linear model imputed values closer to the mean.
 
@@
 
-Imputed data are combined by row binding imputation models.
-The data are reshaped into long format to compare imputed values for each model for 4 variables (Figure \@ref(fig:imputed-by-model)).
-Compared to KNN imputed values, the linear model imputed values closer to the mean value.
-
 ```{r imputed-by-model, fig.cap = "Boxplots of complete case data, and data imputed with KNN or linear model for different variables.
@@
 
-The coefficients of the linear model of log price vary for room (Figure \@ref(fig:tidy-coefs)), for different imputed datasets.
-Notably, complete cases  resulted in underestimating the impact of room on log price.
-A partial residual plot  (Figure \@ref(fig:partial-resid)) shows there is not much variation amongst the models from the different datasets.
-
-```{r tidy-coefs, fig.height = 2, fig.width = 7, fig.cap = "Visualization of the variation in coefficients for linear model of log price for each of the different datasets for the number of rooms.
-In red is complete case (cc), in green is the knn imputed dataset, and in blue the imputed by linear model.
-Using the complete case dataset produced smaller coefficients compared to the imputed models.", out.width = "100%"}
+Coefficients of the linear model of log price vary for room for different imputed datasets (Figure \@ref(fig:tidy-coefs)).
+Notably, complete cases underestimate the impact of room on log price.
+A partial residual plot (Figure \@ref(fig:partial-resid)) shows there is not much variation amongst the models from the differently imputed datasets.
+
+```{r tidy-coefs, fig.height = 2, fig.width = 7, fig.cap = "The coefficient estimate for the number of rooms varies according to the imputed dataset.
+Complete case dataset produced lower coefficients, compared to imputed datasets", out.width = "100%"}
 
@@
 
-```{r partial-resid, fig.cap = "Partial residual plot for each data set, complete cases (cc), imputed with KNN (knn), and imputed with a linear model (lm).
-These are plotted with hex bins, which are coloured according to the number of points in a given hexagon - the brighter the colour the more points.
-This reveals data sets imputed with knn and linear models have more points clustered around zero, whereas complete cases is more flat.", fig.width = 9, fig.height = 3, out.width = "100%"}
+```{r partial-resid, fig.cap = "Partial residual plot for each data set, complete cases (cc), and imputed with KNN (knn) or linear model (lm).
+These are plotted as hex bins, colored according to the number of points in a given hexagon.
+Brighter colors mean more points.
+Compared to complete cases, imputed data has more points clustered around zero.", fig.width = 9, fig.height = 3, out.width = "100%"}
 
@@
 
-The `naniar` and `visdat` packages build on existing tidy tools and strike a compromise between automation and control, making analysis efficient, readable, but not overly complex.
+The `naniar` and `visdat` packages implement the methods discussed in the paper, building on existing tidy tools and strike a compromise between automation and control, making analysis efficient, readable, but not overly complex.
 Each tool has clear intent and effects - plotting or generating data or augmenting data in some way.
-This reduces repetition and typing for the user, making exploration of missing values easier as they follow consistent rules with a declarative interface.
+This reduces repetition and typing for the user, making exploration of missing values easier as they follow consistent, predictable rules.
 
@@
 
-The new data structures discussed in the paper promote flexibility, and could be used to create different visualizations than were shown in the paper.
+The _nabular_ data structures discussed in the paper are simple by design, to promote flexibility.
+They could be used to create different visualizations than were shown in the paper.
 The analyst can use the data structures to decide on appropriate visualization for their problem.
 The data structures could also be used to support interactive graphics, in the manner of MANET and ggobi.
-Linking the plots (linked brushing) to explore missingness or animating between different sets of imputed values.
-New packages like `plotly` [@plotly] facilitate interactive graphics, and packages like `gganimate` facilitate animations [@gganimate].
+Linking the plots (via linked brushing) could facilitate exploration of missingness, and could be implemented with \pkg{plotly} [@plotly] for added interactivity.
+Animating between different sets of imputed values could also be possible with packages like \pkg{gganimate} [@gganimate].
 

### 2020-01-14 paper/tidy-missing-data-paper.Rmd
minor text amendments
minor changes to figures + some titles + reference later sections in the paper.

@@
 library(magick)
+library(rpart)
+library(visdat)
+library(patchwork)
 options(tinytex.verbose = TRUE)
@@
 
-The imputation literature (e.g. @Little1988, @Rubin1976, @Simon1986,
-@Schafer2002, @VanBuuren2012) focusses on ensuring valid statistical inference
+The imputation literature (See \@ref(imputation)) focusses on ensuring valid statistical inference
 is made from incomplete data. This is approached chiefly through probabilistic
@@
 
-While there have been many software tools for exploring missing data [@VIM, @mi, @mice, @Cheng2015, @MANET, @ggobi, @norm], these tools do not integrate easily into analysis pipelines.
+While there have been many software tools for exploring missing data (See \@ref(exploration)), these tools do not integrate easily into analysis pipelines.
 Despite it being crucial to explore and understand missing data, there is very little research on how to make this process more efficient and integrate it into the analysis pipeline.
@@
 
-The graphics literature [@Swayne1998; @Unwin1996; @VIM; @Cheng2015] provides several solutions for exploring missings visually.
+The graphics literature (See \@ref(exploration)) provides several solutions for exploring missings visually.
 These approaches incorporate missings into the plot in some way.
@@
 
-# Extensions to tidyverse {#extensions}
+# Tidy framework for Missings {#extensions}
 
@@
 
-library(visdat)
-
 p1 <- vis_miss(airquality)
@@
 "}
-library(naniar)
-library(tidyverse)
 
@@
 The variables building area and year built have more than 50% missing data, and so could perhaps be omitted from subsequent analysis, as imputed values are likely to be spurious.
+Three missingness clusters are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \@ref(fig:applic-vis-miss)).
 
@@
 ```
-
-Three missingness clusters are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \@ref(fig:applic-vis-miss)).
 
@@
 ```{r rpart-fit-cluster}
-library(rpart)
 
@@
 
-aug_fits %>%
-  ggplot(aes(x = .fitted,
+gg_hex <- function(data,
+                   subset){
+
+  data_subset <- data %>% 
+    filter(model == subset)
+
+    ggplot(data_subset, aes(x = .fitted,
                y = .resid)) + 
-  # geom_point(alpha = 0.1) + 
     geom_hex() + 
-  labs(fill = "Count") +
     scale_fill_viridis_c() +
-  facet_wrap(~model)
+    theme(legend.position = "bottom",
+          aspect.ratio = 1) +
+      labs(tag = subset)
+}
+
+gg_hex(aug_fits, "cc") +
+gg_hex(aug_fits, "knn") +
+gg_hex(aug_fits, "lm")
 

### 2020-01-17 paper/tidy-missing-data-paper.Rmd
read over updates

@@
 The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as an integral part of data analysis workflows.
-We define both new data structures and new verbs (functions) to perform common operations.
+We define a new data structure, and a suite of new operations.
 Together, these provide a cohesive framework for handling, exploring, and imputing missing values.
-These methods have been made available in the R package `naniar`.
+These methods have been made available in the R package called `naniar`.
 keywords:
   # at least one keyword must be supplied
-  formatted: [workflow, statistical computing, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
-  plain:     [workflow, statistical computing, data science, data visualization, tidyverse, data pipeline, R]
+  formatted: [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
+  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
 preamble: >
@@
 As data science has become a more solid field, theories and principles have developed to describe best practices.
-One such idea is 'tidy data,' a set of guidelines for converting raw data into a clean, analysis-ready format that works efficiently in a data analysis pipeline, while another is the idea of the Grammar of Graphics, which describes how to map data values into a visualization [@Wickham2014, Wilkinson2012].
-These principles have guided analysts in their work for several years now.
-However, while these principles outline how to work with data in a general way, they do not address the problem of missing data.
-In particular, there is little to no guidance for working with missing values in the field of visualization.
-Most analysis software simply drops missing values when making a plot, although some packages provide a warning (Figure \@ref(fig:warning)).
+One such idea is 'tidy data,' which defines of a clean, analysis-ready format, that informs the workflow to convert raw data through a data analysis pipeline [@Wickham2014].
+The second idea is the grammar of graphics, which describes how to map data values into a visualization [@Wilkinson2012].
+These principles have been widely adopted.
+However, these principles do not address the problem of missing data.
+In particular, there is little guidance for missing value handling in the data workflow.
+Most analysis and visualisation software simply drops missing values when making a plot, although some packages like \pkg{ggplot2} provide a warning (Figure \@ref(fig:warning)).
 
@@
 
-```{r warning, fig.cap = "How ggplot behaves when displaying missing values: ggplot2 does not display missing values in the plot, but at least throws a warning message to the user, which is better than many statistical models.", fig.height = 2.5, fig.width = 2.5, out.width = "75%", warning = FALSE, fig.show = 'hold'}
+```{r warning, fig.cap = "How ggplot2 behaves when displaying missing values.
+A warning message is displayed, but missing values are not shown in the plot", fig.height = 2.5, fig.width = 2.5, out.width = "75%", warning = FALSE, fig.show = 'hold'}
 
@@
 
-The imputation literature (See \@ref(imputation)) focusses on ensuring valid statistical inference
+The imputation literature focusses on ensuring valid statistical inference
 is made from incomplete data. This is approached chiefly through probabilistic
@@
 imputation. To understand their structure, the analyst must explore the data
-with perform visualizations, summaries, and modelling. This exploratory process
+with visualizations, summaries, and modelling. This exploratory process
 is iterative, and often soaks up a lot of time before more pertinent questions
 can be answered.
+
+Understanding the missing data mechanism is an iterative procedure, to arrive at a reasonably confident state of understanding the missing data mechanism.
 

### 2020-01-17 paper/tidy-missing-data-paper.Rmd
wrap up introduction

@@
 complete dataset for analysis, whether by case- or variable-deletion, or with
-imputation. To understand their structure, the analyst must explore the data
-with visualizations, summaries, and modelling. This exploratory process
-is iterative, and often soaks up a lot of time before more pertinent questions
-can be answered.
-
-Understanding the missing data mechanism is an iterative procedure, to arrive at a reasonably confident state of understanding the missing data mechanism.
-
-Decision tree models or latent group analysis can reveal structures or patterns
+imputation. Decision tree models or latent group analysis can reveal structures or patterns
 of missingness [@Tierney2015; @Barnett2017], but are not definitive. While there
 might be other times where the missing data mechanism is obvious, determining
-this is not always straightforward. There is no silver bullet to reveal missing
-data mechanisms to the analyst. Missing data must be explored.
-
-While there have been many software tools for exploring missing data (See \@ref(exploration)), these tools do not integrate easily into analysis pipelines.
-Despite it being crucial to explore and understand missing data, there is very little research on how to make this process more efficient and integrate it into the analysis pipeline.
-The analyst must
-explore missing data using visualizations, summaries, and exploratory analyses,
-but there is a lack of modern tools to efficiently explore missing data.
-
-The graphics literature (See \@ref(exploration)) provides several solutions for exploring missings visually.
-These approaches incorporate missings into the plot in some way.
-For example, imputing values 10% below the range to display all values in a scatter plot.
-These methods require the data to be linked to the missing or imputed values, but this link is typically hidden from the user.
-The approaches from the graphics literature have not been translated and integrated into tidy data.
-This limits the analyst's capacity to efficiently explore missing values.
-Extending tidy data and tidy tools to account for missing values would create more efficient and robust workflows and analysis.
-
-The paper is organized in the following way to discuss the problem of missing data.
-The next section (\@ref(tidy-data-concepts)) is an introduction to tidy data principles and tools.
-Existing missing data software is discussed in \@ref(existing-software).
-Section \@ref(extensions) discusses tidy methodology extensions, which facilitate operations for exploring, visualizing, and imputing missing data.
-The application of these new methods is illustrated using a case study in section \@ref(case-study).
-Finally, section \@ref(discussion) discusses strengths, limitations, and future directions.
+this is not straightforward. To understand their structure, the analyst must explore the data
+with visualizations, summaries, and modelling. It is an iterative procedure, to arrive at a reasonably confident state of understanding the missing data mechanism.
+
+While there have been many software tools for exploring missing data, they do not work together.
+The graphics literature provides several solutions for exploring missings visually by incorporating them into the plot in some way.
+For example, imputing values to be 10% below the minimum will include all observations into a scatter plot.
+
+However, this is often implicit, and not transparent to the user.
+(**NOTE** include possible brief example from mice/mi/VIM).
+The ideas from the graphics literature need to be translated into tidy data tools to integrate missing data handling in a data pipeline.
+
+This paper is organized in the following way.
+The next section (\@ref(background)) provides the background to to tidy data principles and tools (\@ref(tidy-data-concepts) and missing data representations (\@ref(missing-data-rep-dep).
+Section \@ref(existing-software) summarises existing software for handling missing data.
+The new work extending the tidy tools to facilitate exploring, visualizing, and imputing missing data is discussed in Section \@ref(extensions).
+An application illustrating the use of the new methods is shown in Section \@ref(case-study).
+Section \@ref(discussion) discusses strengths, limitations, and future directions.
 

### 2020-01-27 paper/tidy-missing-data-paper.Rmd
misc grammar changes

@@
 
-Features of tidy data were formally discussed in 2014 [@Wickham2014], subsequently generating much discussion (@Donoho2017 and commentary, @r4ds), and tools for data analysis. Tidy data is defined as:
+Features of tidy data were formally discussed in @Wickham2014, and were discussed in terms of their importance for data science by @Donoho2017, and tools for data analysis.
+Tidy data is defined by @r4ds as:
 
@@
 This helps the analyst works swiftly and clearly, closing up opportunities for errors.
-Tidy data principles are general, but are most robustly implemented in the R programming language, so the
-paper focusses on R.
-
-With tidy data come tidy tools, which have the same input and output: tidy data.
-This consistency means multiple tools can be composed together in sequence, allowing for rapid, elegant operations.
+Tidy data principles are general, but are comprehensively implemented in the R programming language, so this paper focusses on R.
+
+Tidy tools require the same tidy data input and output.
+This consistency means multiple tools can be composed together into a sequence, allowing for rapid, elegant, and complex operations.
 Contrasting tidy tools are messy tools.
@@
 This makes the work at each step harder to predict, and more complex and difficult to maintain.
-This disrupts workflow, and incites errors.
+This disrupts workflow, and invites errors.
 Tidy tools fall into three broad categories: data manipulation, visualization, and modelling.
-These are now discussed in turn.
 
@@
 These provide the five "verbs" of data manipulation: data reshaping, sorting, filtering, transforming, and aggregating.
-**Data reshaping** goes from long to wide formats; **sorting** arranges rows in a specific order; **filtering** removes rows based on a condition; **transforming** changes existing variables or adding new ones; **aggregating** creates a single value from many values, say for example by computing the minimum, maximum, and mean.
+**Data reshaping** goes from long to wide formats; **sorting** arranges rows in a specific order; **filtering** removes rows based on a condition; **transforming**, changes existing variables or adds new ones; **aggregating** creates a single value from many values, say, for example, by computing the minimum, maximum, and mean.
 
@@
 Defining tidy data and tidy tools has resulted in a growing set of packages known collectively as the "tidyverse" [@tidyverse].
-These are designed to share similar principles in their design and behavior, and cover the breadth of an analysis - from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @tidyverse; @Tidyverse-Manifesto].
+These are constructed to share similar principles in their design and behavior, and cover the breadth of an analysis - from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @tidyverse; @Tidyverse-Manifesto].
 This has led to more tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr} and dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} [@readr; @readxl; @haven; @stringr; @lubridate; @purrr].
 It has also led to a burgeoning of new packages for other fields following similar design principles, creating fluid workflows for new domains.
-For example, the \pkg{tidytext} package for text analysis, the \pkg{tsibble} package for time series data, and \pkg{tidycensus} for working with US census and boundary data [@tidytext; @wang2020tsibble; @tidycensus].
+For example, the \pkg{tidytext} [@tidytext] package for text analysis, the \pkg{tsibble} [@wang2020tsibble] package for time series data, and \pkg{tidycensus} [@tidycensus] for working with US census and boundary data.
 
@@
 
-Graphics research on missing data has not yet been framed in a tidy way.
-The graphics literature has explored methods for missing value exploration and imputation.
-Translating and expanding these to fit within tidy data and tidy tools would create more efficient workflows.
+Taking existing methods from the missing data graphics literature, and translating and expanding them into tidy data and tidy tools would create more effective data visualisations.
 Defining these concepts allows the focus to be more general than just software, but rather, an extensible framework for tidy tools to explore missing data.
@@
 
-Missingness can be represented as a matrix, $R$, for data $y$ with $i$ rows and $j$ columns:
+The convention for representing missingness is a matrix, $R$, for data $y$ with $i$ rows and $j$ columns:
 
@@
 
-There are many ways each value can go missing, we adopt the notation used in @VanBuuren2012.
-The information in $R$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), Missing at random (MAR), and missing not at random (MNAR).
+There are many ways each value can be missing, we adopt the notation used in @VanBuuren2012.
+The information in $R$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR).
 The distribution of missing values in $r_{ij}$ can depend on the entire dataset, represented as $Y = (Y_{obs}, Y_{miss})$.
 This relationship can be defined by the _missing data model_ $Pr(r_{ij} | Y_{obs}, Y_{miss}, \psi)$, the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, $\psi$.
-This helps precisely define categories of missing values.
+This helps to precisely define categories of missing values.
 
 **MCAR** is where values being missing have no association with observed or unobserved data, that is, $Pr(R = 1 | Y_{obs}, Y_{miss}) = Pr(R = 1 | \psi)$.
-Essentially, the probability of an observation being missing is unrelated to anything else, only the parameters $\psi$, the overall probability of missingness.
+Essentially, the probability of an observation being missing is unrelated to anything else, only the parameter $\psi$, the overall probability of missingness.
 Although a convenient scenario, it is not actually possible to confirm, or clearly distinguish from MAR, as it relies on statements on data unobserved.
-In **MAR**, missingness only depends on data observed, not data missing, that is,$Pr(R = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(R = 1 | Y_{obs}, \psi)$.
+In **MAR**, missingness only depends on data observed, not data missing, that is, $Pr(R = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(R = 1 | Y_{obs}, \psi)$.
 Some structure or dependence between missing and observed values is allowed, provided it can be explained by data observed, and some overall probability of missingness.
 In **MNAR**, missingness is related to values observed, and unobserved: $Pr(R = 1 | Y_{obs}, Y_{miss}, \psi)$.
-This assumes conditioning on all observations, data goes missing due to some phenomena unobserved, including the structure of the missing data itself.
+This assumes conditioning on all observations: data goes missing due to some phenomena unobserved, including the structure of the missing data itself.
 This presents a challenge in analysis, as it is difficult to verify, and implies bias in analysis due to the unobserved phenomena.
@@
 
-Visualizations can help assess whether data is MCAR, MAR or MNAR.
-Imputation is recommended in most cases of missingness dependence, where the values are plausible (e.g., imputing a pulse for someone deceased), as this can help re-assess the classification of missingness.
+Visualizations can help assess whether data is MCAR, MAR or MNAR. Imputation is recommended in most cases of missingness dependence.
 
@@
 
-Methods for exploring, understanding, and imputing missing data are more accessible than they have ever been.
-In particular, the number of methods and implementations in imputation seeing continued growth.
+Methods for exploring, understanding, and imputing missing data are more accessible now than they have ever been.
+In particular, the number of methods and implementations of imputation are seeing continued growth.
 Values can be imputed with one value (single imputation), or multiple values (multiple imputation), creating $m$ datasets.
-This section discusses existing software for single and multiple imputation \@ref(imputation), and exploration \@ref(exploration).
+This section discusses existing software for single and multiple imputation, and exploration.
+
+(**read over next two sections - Nick**)
 
@@
 
-# Tidy framework for Missings {#extensions}
-
-Applying tidyverse principles to the domain of missing data clarifies and
-facilitates missing data exploration, visualization, and imputation. This
-section discusses how these principles, in three parts: tidy data structures for missing data \@ref(data-structure), common operations (verbs) of missing data \@ref(verbs), graphics \@ref(graphics), and data summaries \@ref(num-sum).
-The design and naming of these features is discussed throughout, as they are an important component to help the analyst compose them efficiently in data pipelines.
+# Tidy framework for missings {#extensions}
+
+Applying tidyverse principles has the potential to clarify missing data exploration, visualization, and imputation. This
+section discusses how these principles are applied for data structures \@ref(data-structure), common operations (verbs) \@ref(verbs), graphics \@ref(graphics), and data summaries \@ref(num-sum).
+The design and naming of these features is important, and care has been taken to make the names intuitive for their purpose.
 
@@
 To explore missing data there needs to be a robust representation of missing
-values that integrate into analysis pipelines.
+values that integrates into analysis pipelines.
 This means a data structure that is easily transported, and not a special class reliant on attributes that can be lost in analysis.
-The R matrix, where 0 and 1
+The $R$ matrix, where 0 and 1
 indicate not missing and missing provides a useful template, as it is very common in
@@
 shadow", or "shadow matrix", defined as a copy of the original data with
-indicator values of missingness, where 1s represent missing values, and 0
-represents complete values. The values 0 and 1 can
+indicator values of missingness. The values 0 and 1 can
 be confusing representations of missing values - does
-0 indicate a the absence of observation, or the presence of a missing value? The
-shadow matrix was specially linked to the data, but could not be access
-directly, instead linked with special window views.
-The data structure was abstracted from the user to facilitate data exploration, which was consistent with the state of the art.
-However, the data structure needs to be clearly presented to the user so they
-can apply it more flexibly in data analysis.
-
-```{r shadow-matrix-progress, echo = FALSE, fig.cap = "Progression of creating shadow matrix data.
+0 indicate an absence of observation, or the presence of a missing value? 
+The shadow matrix could be interactively linked to the data.
+
+```{r shadow-matrix-progress, echo = FALSE, fig.cap = "Series of steps to create shadow matrix data.
 (1-2) Data to a binary representation of missings, (2-3) Binary format converted to a shadow matrix (3-4) New and improved shadow matrix with changed variable names.
@@
 
-We propose a new form for tidy representation of missing data.
-It is simple, and flexible, building off of past research.
-Four features are added to the shadow matrix to facilitate analysis of missing data:
-
-1. Missing value label: A simple label to the shadow matrix clear, "NA" and "!NA" for missing and not missing, represented as 0 and 1 (Figure \@ref(fig:shadow-matrix-progress) parts 2 and 3).
-
-1.
-Special missing values: Building on 2, the values in the shadow matrix can be "special" missing values, indicated as `NA_suffix`, where "suffix" is a very short label of the type of missings, and the underlying values are >= 2.
-
-1. Coordinated names: Variables in the shadow matrix gain the same name as in the data, with the suffix "_NA".
-
-1. Cohesiveness: Binding the shadow matrix column-wise to the original data creates a cohesive _nabular_ data form, useful for visualization, summaries, and tracking imputed values.
-
-### Missing value labels {#missing-labels}
+We propose a new form for tidy representation of missing data based on these ideas, based on this past research.
+Four features are added to the shadow matrix to facilitate analysis of missing data, illustrated in Figure \@ref(fig:shadow-matrix-progress)
+
+(**note** - condense details below into points here.)
+
+1. **Missing value labels**. A simple label to the shadow matrix clear, "NA" and "!NA" for missing and not missing, represented as 0 and 1.
 
@@
 
+1.
+Special missing values: Building on 1, the values in the shadow matrix can be "special" missing values, indicated as `NA_instrument`, where "instrument" is a very short label indicating an instrument error resulting in a missing values.
+
+1. Coordinated names: Variables in the shadow matrix gain the same name as in the data, with the suffix "_NA".
+
+1.
+Cohesiveness: Binding the shadow matrix column-wise to the original data creates a cohesive _nabular_ data format, so that it cannot be out of sync with the data.
+useful for visualization, summaries, and tracking imputed values.
+
+(**note - change R matrix to something else**)
+
+### Missing value labels {#missing-labels}
+
 ### Special missings {#special-missings}
+
+()
 
@@
 This subtle change is important as it shifts the focus from the value of a variable to its missingness state, making intent clear when performing analysis.
+
+(make this it's own section)
 

### 2020-03-20 paper/tidy-missing-data-paper.Rmd
replace $r$ with $b$ - closes #12

@@
 
-The convention for representing missingness is a matrix, $R$, for data $y$ with $i$ rows and $j$ columns:
+The convention for representing missingness is a **b**inary matrix, $B$, for data $y$ with $i$ rows and $j$ columns:
 
 $$
-r_{ij} =\begin{cases}
+b_{ij} =\begin{cases}
 1, & \text{if } y_{ij} \text{ is missing} \\
@@
 There are many ways each value can be missing, we adopt the notation used in @VanBuuren2012.
-The information in $R$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR).
-The distribution of missing values in $r_{ij}$ can depend on the entire dataset, represented as $Y = (Y_{obs}, Y_{miss})$.
+The information in $B$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR).
+The distribution of missing values in $b_{ij}$ can depend on the entire dataset, represented as $Y = (Y_{obs}, Y_{miss})$.
 This relationship can be defined by the _missing data model_ $Pr(r_{ij} | Y_{obs}, Y_{miss}, \psi)$, the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, $\psi$.
@@
 
-The missingness matrix $R$ can be used to assess missing data dependence It has been used in interactive graphics, dubbed a "shadow matrix", to link missing and imputed values to the data, facilitating their display [@Swayne1998], focusing heavily on multivariate numeric data.
+The missingness matrix $B$ can be used to assess missing data dependence It has been used in interactive graphics, dubbed a "shadow matrix", to link missing and imputed values to the data, facilitating their display [@Swayne1998], focusing heavily on multivariate numeric data.
 This is an idea upon which this new work builds.
@@
 This means a data structure that is easily transported, and not a special class reliant on attributes that can be lost in analysis.
-The $R$ matrix, where 0 and 1
+The $B$ matrix, where 0 and 1
 indicate not missing and missing provides a useful template, as it is very common in
@@
 
-The $R$ matrix is a very common representation of missing values.
+The $B$ matrix is a very common representation of missing values.
 It can be created in the "R" language with \code{is.na(data)}.
@@
 
-The _nabular_ data structure provides three key benefits over on-the-fly creating a missingness matrix $R$.
+The _nabular_ data structure provides three key benefits over on-the-fly creating a missingness matrix $B$.
 Firstly, missing value labels `NA` and `!NA` are clearer than TRUE or FALSE.
@@
 
-Common missing data operations can be considered verbs: scan, replace, add, shadow, impute, track, and flag.
-Missing values can be **scanned** to find possible missings not coded as `NA`.
+Common missing data operations can be considered verbs, in the tidyverse sense.
+For missing data, these include: **scan**, **replace**, **add**, **shadow**, **impute**, **track**, and **flag**.
+Data can be **scanned** to find possible missings not coded as `NA`.
 These values can then be **replaced** with `NA`.
 To facilitate exploration, summaries of missingness can be **added** as a summary column to the original data.
-Data values can be augmented with **shadow** values, helping explore missing data, as well as facilitating the process of **imputing**, and **tracking**.
+Data matrix can be augmented with the **shadow matrix** values, helping explore missing data, as well as facilitating the process of **imputing**, and **tracking**.
 Finally, unusual or specially coded missing values can be **flagged**.
@@
 
-There are many common values for "missing".
-For example, "N/A", "MISSING", `-99`.
-Before replacing these values with missing values (`NA`), it is useful to _scan_ the data to understand the magnitude of the problem.
-`miss_scan_count()` finds where specified values occur in data.
-For example, a scan of all occurrences of -99 is done with `miss_scan_count(data, search = -99)`, returning a table of occurrences of that value for each variable.
-`common_na_numbers` and `common_na_strings` provide a list of common NA values for numbers and characters.
-They contain values like "n/a", "na", and -99, and are designed to be used with `miss_scan_count` to identify common missing values.
+This operation is used to search the data for specific conventional representations of missings, such as, "N/A", "MISSING", `-99`. 
+The function is called  `miss_scan_count()`, and it returns a table of occurrences of that value for each variable.
+Two more data objects, `common_na_numbers` and `common_na_strings` provide a list of common NA values for numbers and characters, to help users check for the typical representations of missing.
 
@@
                          -98, "F",   -1)
-
-# knitr::kable(dat_ms, 
-#              caption = "An example dataset with variables x, y and z containing some common unusual missing values.")
-             # booktabs = TRUE) 
-  # kable_styling(latex_options = c("hold_position"))
 ```
@@
 
-## replace: Replacing missing value label with another {#verbs-replace-with}
-
-Once the magnitude of possible missing values has been assessed, these values need to be replaced with missing values.
-This can be done with `replace_with_na()`.
-A dataset could have the values -99 in the `x` column replaced with the following code:
-`replace_with_na(dat_ms, replace = list(x = -99))` For flexibility, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`.
-This means `replace_with_na_all` operates on **all** columns, `replace_with_na_at` operates **at** specific columns, and `replace_with_na_if` operates on columns **if** they meet some condition (such as `is.numeric` or `is.character`).
-
-## add: Adding missingness summary variables {#verbs-add-cols}
+### replace: Replacing missing value label with another {#verbs-replace-with}
+
+Once the different representations of possible missing values have been identified, these values can be replaced using the function, `replace_with_na()`.
+A dataset could have the values `-99` in the `x` column replaced with the following code:
+`replace_with_na(dat_ms, replace = list(x = -99))`.
+For operating on multiple variables, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`.
+This means `replace_with_na_all` operates on **all** columns, `replace_with_na_at` operates **at** specific columns, and `replace_with_na_if` makes a conditional change on columns **if** they meet some condition (such as `is.numeric` or `is.character`).
+
+### add: Adding missingness summary variables {#verbs-add-cols}
 
@@
 
-## shadow: Creating nabular data {#verbs-nabular}
+### shadow: Creating nabular data {#verbs-nabular}
 
@@
 
-## flag: Describing different types of  missing values {#verbs-recode}
+### flag: Describing different types of  missing values {#verbs-recode}
 
@@
 
-## impute: Imputing values {#verbs-impute}
+### impute: Imputing values {#verbs-impute}
 

### 2020-03-27 paper/tidy-missing-data-paper.Rmd
condence points

@@
 
-1. **Missing value labels**. A simple label to the shadow matrix clear, "NA" and "!NA" for missing and not missing, represented as 0 and 1.
-
-The $B$ matrix is a very common representation of missing values.
-It can be created in the "R" language with \code{is.na(data)}.
-The values 0 and 1 (or FALSE and TRUE) can be confusing, however, as it is not immediately clear what they represent: does 0 indicate a the absence (0) of observation, or the presence (1) of a missing value?
-This is solved by adding a simple label to the shadow matrix: "NA" and "!NA" for missing and not missing.
-Values in the shadow matrix can be represented as "factor" types values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing (Figure \@ref(fig:shadow-matrix-progress) parts 1-3).
-
 1.
-Special missing values: Building on 1, the values in the shadow matrix can be "special" missing values, indicated as `NA_instrument`, where "instrument" is a very short label indicating an instrument error resulting in a missing values.
-
-1. Coordinated names: Variables in the shadow matrix gain the same name as in the data, with the suffix "_NA".
+**Missing value labels**.
+Simple labels, "NA" and "!NA", for missing and not missing, clearly identifies missing values for analysis and plotting.
+This is an improvement over the commonly used binary matrix ($B$), which makes it unclear if 0 indicates the absence (0) of observation, or the presence (1) of a missing value.
+Missing value labels can be represented as "factor" types values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing (Figure \@ref(fig:shadow-matrix-progress) parts 1-3).
 
 1.
-Cohesiveness: Binding the shadow matrix column-wise to the original data creates a cohesive _nabular_ data format, so that it cannot be out of sync with the data.
-useful for visualization, summaries, and tracking imputed values.
-
-(**note - change R matrix to something else**)
-
-### Missing value labels {#missing-labels}
-
-### Special missings {#special-missings}
-
-()
-
-Extending missing value labels to include "special" missing values labelled as "NA_\<suffix\>" allows for representing additional missingness features such as instrument failure and drop-out as "NA_instr_fail", and "NA_drop_out", and even imputations (Table \@ref(tab:shadow-encoding)).
+**Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA_\<suffix\>" - for example: `NA_instrument` uses a short label, "instrument", indicating instrument error resulting in missing values.
+These could be also used to indicate imputations, (Table \@ref(tab:shadow-encoding)).
 The underlying number representation changes from binary to integer, allowing for unlimited missingness features.
@@
 Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R.
+
+1.
+**Coordinated names**: Variables in the shadow matrix gain a consistent short suffix, "_NA", keeping names coordinated throughout analysis (Figure \@ref(fig:shadow-matrix-progress) part 4).
+The suffix is short and easy to remember in analysis or visualization.
+This shifts the focus from the value of a variable to its missingness state, making intent clear when performing analysis.
+
+(It makes a clear distinction with `var_NA` being a random variable of the missingness of a variable, `var`. )
+
+1.
+**Cohesiveness**: Binding the shadow matrix column-wise to the original data creates a cohesive _nabular_ data format, so that it cannot be out of sync with the data.
+useful for visualization, summaries, and tracking imputed values.
+This is discussed in more detail below in \@ref(nabular-data).
 
@@
 ```
-
-## Coordinated names {#coordinated-names}
-
-A consistent short suffix "`_NA`" for each variable keeps names coordinated throughout analysis (Figure \@ref(fig:shadow-matrix-progress) part 4).
-The suffix is short and easy to remember in analysis or visualization.
-It also makes a clear distinction with `var_NA` being a random variable of the missingness of a variable, `var`.
-This subtle change is important as it shifts the focus from the value of a variable to its missingness state, making intent clear when performing analysis.
-
-(make this it's own section)
 

### 2020-03-27 paper/tidy-missing-data-paper.Rmd
adding some notes from meeting

@@
 
+(principle: clarity of labelling - transparent what the matrix means, because these could equally be "missing" or "present" - for this. Anybody looking at these values)
+
 1.
@@
 
-Other statistical programs represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` - `.z`.
+Other statistical programs (SPSS, SAS, and Stata) represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` - `.z`.
+
+SAS, SPSS, and STATA all implement
+
 The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state.
 Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R.
+
+(discuss the principles, rather than what we've done)
+
+(in the same column you have mixed information, which breaks the tidy principle of one column having one type of value)
 
@@
 
+(change cohesiveness to connectedness)
+
 1.
-**Cohesiveness**: Binding the shadow matrix column-wise to the original data creates a cohesive _nabular_ data format, so that it cannot be out of sync with the data.
+**Connectedness**: Binding the shadow matrix column-wise to the original data creates a single connected _nabular_ data format, so that it cannot be out of sync with the data.
 useful for visualization, summaries, and tracking imputed values.
@@
 ```{r clean-data-for-analysis, include = FALSE}
-
+# add memoise to store the output data for this
 impute_knn <- simputation::impute_knn

### 2020-04-05 paper/tidy-missing-data-paper.Rmd
added a note on this section to work on
minor additions

@@
 
+(note: section to work on this week.)
+
+- re-arrange this paragraph around
+
 To explore missing data there needs to be a robust representation of missing
@@
 This means a data structure that is easily transported, and not a special class reliant on attributes that can be lost in analysis.
-The $B$ matrix, where 0 and 1
-indicate not missing and missing provides a useful template, as it is very common in
-missing data literature.  This matrix was used to explore missing values in
+The $B$ matrix, where 0 and 1 indicate not missing and missing (respectively), provides a useful template, as it is very common in missing data literature.
+This matrix was used to explore missing values in
 the interactive graphics library XGobi, called a "missing value
 shadow", or "shadow matrix", defined as a copy of the original data with
-indicator values of missingness. The values 0 and 1 can
-be confusing representations of missing values - does
+indicator values of missingness. The shadow matrix could be interactively linked to the data.
+
+There are some limitations to the shadow matrix, however. Namely, the values 0 and 1 can be confusing representations of missing values - does
 0 indicate an absence of observation, or the presence of a missing value?
-The shadow matrix could be interactively linked to the data.
 
@@
 
-We propose a new form for tidy representation of missing data based on these ideas, based on this past research.
+We propose a new form for tidy representation of missing data based on these ideas from past research.
 Four features are added to the shadow matrix to facilitate analysis of missing data, illustrated in Figure \@ref(fig:shadow-matrix-progress)
@@
 
-## track: Shadow and impute missing values {#verbs-track}
+### track: Shadow and impute missing values {#verbs-track}
 
@@
 # Graphics {#graphics}
+
+(note: next section to work on)
 
@@
 
-```{r clean-data-for-analysis, include = FALSE}
+```{r clean-data-for-analysis, include = FALSE, cache = TRUE}
 # add memoise to store the output data for this
@@
 
-```{r impute-knn}
+```{r impute-knn, cache = TRUE}
 

### 2020-04-06 paper/tidy-missing-data-paper.Rmd
notes from meeting with Di

@@
 
-- re-arrange this paragraph around
-
 To explore missing data there needs to be a robust representation of missing
@@
 shadow", or "shadow matrix", defined as a copy of the original data with
-indicator values of missingness. The shadow matrix could be interactively linked to the data.
-
-There are some limitations to the shadow matrix, however. Namely, the values 0 and 1 can be confusing representations of missing values - does
-0 indicate an absence of observation, or the presence of a missing value?
+indicator values of missingness.
+The shadow matrix could be interactively linked to the data.
+There are some limitations to the shadow matrix, however.
+Namely, the values 0 and 1 can be confusing representations of missing values - does 0 indicate an absence of observation, or the presence of a missing value?
 
@@
 We propose a new form for tidy representation of missing data based on these ideas from past research.
-Four features are added to the shadow matrix to facilitate analysis of missing data, illustrated in Figure \@ref(fig:shadow-matrix-progress)
+Four features are added to the shadow matrix to facilitate analysis of missing data, illustrated in Figure \@ref(fig:shadow-matrix-progress).
 
@@
 Missing value labels can be represented as "factor" types values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing (Figure \@ref(fig:shadow-matrix-progress) parts 1-3).
-
-(principle: clarity of labelling - transparent what the matrix means, because these could equally be "missing" or "present" - for this. Anybody looking at these values)
+(principle: clarity of labelling - transparent what the matrix means, because these could equally be "missing" or "present" - for this.
+Anybody looking at these values)
 
@@
 Encoding special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in \pkg{naniar} (\@ref(verbs-recode)).
-
-Other statistical programs (SPSS, SAS, and Stata) represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` - `.z`.
-
-SAS, SPSS, and STATA all implement
-
-The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state.
-Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R.
-
-(discuss the principles, rather than what we've done)
-
-(in the same column you have mixed information, which breaks the tidy principle of one column having one type of value)
+{Other statistical programs (SPSS, SAS, and Stata) represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` - `.z`.
+SAS, SPSS, and STATA all implement } {The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state.
+Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R.} {(discuss the principles, rather than what we've done)}.
+{(in the same column you have mixed information, which breaks the tidy principle of one column having one type of value)}.
 
@@
 
+note: First sentence needs to be more on track with the overall summary of what this section is about...
+
+- Graphics are important to understand the missingness structure.
+- The multivariate relationships
+- Working through a hierarchy of how to explore missings - from overview to 
+- Sentence linking the two.
+- How to visualise a binary matrix of missing vs not, and the dependencies between those binary variables as well.
+- It is more than that, when we visualise the complete cases, we don't want to lose track of the missings themselves. This comes back to the data structure
+- gg_miss_var and gg_miss_upset and vis_miss are all showing how we can summarise the missings.
+- but the addition of missing counts into a histogram, allows us to understand that there are many missings - and we have a skewed distribution o Ozone.
+But we want to know where these values might have been located.
+tying this back to the nabular structure, and also the terminology of MCAR MAR, etc.
+(We would expect if data were MCAR for these points to be representatively thoughout the histogram)
+
+When we define the graphics, describe which part of the nabular matrix we are going to display (overview - just binary parts), the summaries, are summaries of the binary, the histogram and points are both the missings and the complete values.
+-
+
+- Combining that with the data itself, looking at the patterns in the complete cases, 
+- from the overview through to the univariate
+
 Visualizing missingness in a dataset is essential to understand its structure.
-This section discusses four different areas of visualization of missing data: overviews (\@ref(overview-graphics)), univariate (\@ref(univariate-graphics)), bivariate (\@ref(bivariate-graphics)), and multivariate (\@ref(multivariate-graphics)).
+This section discusses four different areas of visualization of missing data: overviews (\@ref(overviews)), univariate (\@ref(univariate)), bivariate (\@ref(bivariate)), and multivariate (\@ref(multivariate)).
 All plots are created using \pkg{ggplot2}, giving users clearer control over the plot appearance.
 
-## Overview graphics {#overview-graphics}
+## Overviews {#overviews}
 
@@
 
-## Univariate graphics {#univariate-graphics}
+## Univariate {#univariate}
 
@@
 
-## Bivariate graphics {#bivariate-graphics}
+## Bivariate {#bivariate}
 
@@
 
-## Multivariate graphics {#multivariate-graphics}
+## Multivariate {#multivariate}
 

### 2020-04-29 paper/tidy-missing-data-paper.Rmd
working on graphics section

@@
 
+The acronym MANET [@Unwin1996] stands for "Missings Are Now Equally Treated".
+Missing values are often ignored when plotting data, but  plots can help to reveal the type of missing values patterns existing in the data, and even help to assess whether the fall into the MCAR, MAR or even MNAR labels.
+
+Here we summarise systematically exploring the missing patterns visually, and define  useful plots to make relative to the nabular  data structure.
+
+<!--
 (note: next section to work on)
@@
 All plots are created using \pkg{ggplot2}, giving users clearer control over the plot appearance.
+-->
 
@@
 
-Rather than highlighting the amount of complete data, missing data visualizations should draw attention to the amount of missings, and ordering by missingness.
-Overview visualisations for variables and cases are provided with `gg_miss_var` and `gg_miss_case` (Figure \@ref(fig:gg-miss-case-var)).
-These are shown using the "airquality" dataset, included in base R, containing daily air quality measurements in New York, from May to September, 1973.
+The  first step is to get an overview of the extent of the missingness.
+<!--Rather than highlighting the amount of complete data,-->  Overview visualisations for variables and cases are provided with `gg_miss_var` and `gg_miss_case` (Figure \@ref(fig:gg-miss-case-var)).
+These  missing data visualizations should draw attention to the amount of missings, and ordering by missingness.
+The "airquality" dataset (available in base R), included in base R, containing daily air quality measurements in New York, from May to September, 1973, is shown.
 

### 2020-04-29 paper/tidy-missing-data-paper.Rmd
overview of graphics

@@
 The "airquality" dataset (available in base R), included in base R, containing daily air quality measurements in New York, from May to September, 1973, is shown.
+Two of the variables contain missings and about one third of observations have one missing value.
+A tiny number of observations are missing on two variables.
+These summary plots compute proportions from the shadow matrix.
+The numerical statistics can also be reported, and is discussed in the next section.
 

### 2020-04-29 paper/tidy-missing-data-paper.Rmd
finish of Overview

@@
 
-All missings can be displayed using a heatmap style visualization.
-This is achieved using `vis_miss()` from the \pkg{visdat} package [@visdat], which also provides summaries of missingness overall in the legend, and for each column (Figure \@ref(fig:vis-miss)).
+Both variables and cases can be displayed using a heatmap style visualization,available with `vis_miss()` from the \pkg{visdat} package [@visdat].
+This also provides numerical summaries of missingness in the legend, and for each column (Figure \@ref(fig:vis-miss)).
 The user can also apply clustering to the rows, and arrange columns by missingness (Figure \@ref(fig:vis-miss)B).
@@
 Using \pkg{ggplot2} makes the plot easily customizable.
+This also uses the shadow matrix as the basis for the plot.
 
@@
 The number of times certain variables go missing together can be visualized using an "upset plot" [@Conway2017].
-Similar to a Venn diagram, an upset plot shows the size and features of overlapping sets, but scales well with more variables.
-Figure \@ref(fig:airquality-upset) shows only Ozone and Solar.R have missing values.
+An alternative to a Venn diagram, an upset plot shows the size and features of overlapping sets, but scales well with more variables.
+Figure \@ref(fig:airquality-upset) shows the airquality data, and illustrates  that only Ozone and Solar.R have missing values.
 The bottom right show the combinations of missingness, the bottom left showing missingness in each variable, and the top showing the combinations of missingness.
 The bottom left shows Ozone has the most missing values, and the top shows 2 cases where both Solar.R and Ozone go missing together.
+The upset plot is constructed from the shadow matrix.
 

### 2020-05-01 paper/tidy-missing-data-paper.Rmd
tighten up a few sections and tighten up the use of mentions to shadow matrix, flag  few more things to use in the section

@@
 
-To explore missing data there needs to be a robust representation of missing
-values that integrates into analysis pipelines.
-This means a data structure that is easily transported, and not a special class reliant on attributes that can be lost in analysis.
-The $B$ matrix, where 0 and 1 indicate not missing and missing (respectively), provides a useful template, as it is very common in missing data literature.
+If a data structure is to be added to facilitate exploring missing data, it needs to be simple to understand and transport with a data analysis.
+A useful template is the $B$ matrix, where 0 and 1 indicate not missing and missing (respectively), as it is very common in missing data literature.
 This matrix was used to explore missing values in
@@
 shadow", or "shadow matrix", defined as a copy of the original data with
-indicator values of missingness.
-The shadow matrix could be interactively linked to the data.
+indicator values of missingness. The shadow matrix could be interactively linked to the data.
+
 There are some limitations to the shadow matrix, however.
@@
 Simple labels, "NA" and "!NA", for missing and not missing, clearly identifies missing values for analysis and plotting.
-This is an improvement over the commonly used binary matrix ($B$), which makes it unclear if 0 indicates the absence (0) of observation, or the presence (1) of a missing value.
+This is an improvement over ($B$), which makes it unclear if 0 indicates the absence (0) of observation, or the presence (1) of a missing value.
 Missing value labels can be represented as "factor" types values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing (Figure \@ref(fig:shadow-matrix-progress) parts 1-3).
-(principle: clarity of labelling - transparent what the matrix means, because these could equally be "missing" or "present" - for this.
-Anybody looking at these values)
+
+(NOTE: principle: clarity of labelling - transparent what the matrix means, because these could equally be "missing" or "present" - for this. Anybody looking at these values)
 
@@
 {Other statistical programs (SPSS, SAS, and Stata) represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` - `.z`.
-SAS, SPSS, and STATA all implement } {The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state.
-Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R.} {(discuss the principles, rather than what we've done)}.
-{(in the same column you have mixed information, which breaks the tidy principle of one column having one type of value)}.
+SAS, SPSS, and STATA all implement }
+
+{NOTE: The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state.
+Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R.}
+
+{NOTE: (discuss the principles, rather than what we've done)}. {(in the same column you have mixed information, which breaks the tidy principle of one column having one type of value)}.
 
@@
 
-(change cohesiveness to connectedness)
+(NOTE: change cohesiveness to connectedness)
 
@@
 
-The acronym MANET [@Unwin1996] stands for "Missings Are Now Equally Treated".
-Missing values are often ignored when plotting data, but  plots can help to reveal the type of missing values patterns existing in the data, and even help to assess whether the fall into the MCAR, MAR or even MNAR labels.
-
-Here we summarise systematically exploring the missing patterns visually, and define  useful plots to make relative to the nabular  data structure.
-
-<!--
-(note: next section to work on)
-
-note: First sentence needs to be more on track with the overall summary of what this section is about...
-
-- Graphics are important to understand the missingness structure.
-- The multivariate relationships
-- Working through a hierarchy of how to explore missings - from overview to 
-- Sentence linking the two.
-- How to visualise a binary matrix of missing vs not, and the dependencies between those binary variables as well.
-- It is more than that, when we visualise the complete cases, we don't want to lose track of the missings themselves. This comes back to the data structure
-- gg_miss_var and gg_miss_upset and vis_miss are all showing how we can summarise the missings.
-- but the addition of missing counts into a histogram, allows us to understand that there are many missings - and we have a skewed distribution o Ozone.
-But we want to know where these values might have been located.
-tying this back to the nabular structure, and also the terminology of MCAR MAR, etc.
-(We would expect if data were MCAR for these points to be representatively thoughout the histogram)
-
-When we define the graphics, describe which part of the nabular matrix we are going to display (overview - just binary parts), the summaries, are summaries of the binary, the histogram and points are both the missings and the complete values.
--
-
-- Combining that with the data itself, looking at the patterns in the complete cases, 
-- from the overview through to the univariate
-
-Visualizing missingness in a dataset is essential to understand its structure.
-This section discusses four different areas of visualization of missing data: overviews (\@ref(overviews)), univariate (\@ref(univariate)), bivariate (\@ref(bivariate)), and multivariate (\@ref(multivariate)).
-All plots are created using \pkg{ggplot2}, giving users clearer control over the plot appearance.
--->
+Missing values are often ignored when plotting data, but plots can help to reveal the type of missing value patterns, even those of MCAR, MAR or even MNAR.
+Here we summarise systematically exploring the missing patterns visually, and define useful plots to make relative to the nabular data structure, working from coarse to fine levels of granularity in understanding missingness.
+
+- The acronym MANET [@Unwin1996] stands for "Missings Are Now Equally Treated".
 
@@
 
+- How to visualise a binary matrix of missing vs not, and the dependencies between those binary variables as well.
+- These overview plots are looking at the shadow matrix, and the summaries of each of the variables
+- In this section we  describe which part of the nabular matrix we are going to display (overview - just binary parts), the summaries are summaries of the binary, the histogram and points are both the missings and the complete values.
+
 ```{r gg-miss-case-var, echo = FALSE, fig.show='hold', fig.cap = "Graphical summaries of missingness in variables (A) and cases (B) for the airquality data.
@@
 
-Both variables and cases can be displayed using a heatmap style visualization,available with `vis_miss()` from the \pkg{visdat} package [@visdat].
+The shadow matrix allows both the variables and cases to be displayed using a heatmap style visualization, `vis_miss()` from \pkg{visdat} [@visdat].
 This also provides numerical summaries of missingness in the legend, and for each column (Figure \@ref(fig:vis-miss)).
@@
 Using \pkg{ggplot2} makes the plot easily customizable.
-This also uses the shadow matrix as the basis for the plot.
 
@@
 
+- gg_miss_var and gg_miss_upset and vis_miss are all showing how we can summarise the missings.
+
 Missing values are by default not shown for univariate visualizations such as histograms or densities.
@@
 Ozone is shown in a different color by referring to the variable `Ozone` as `Ozone_NA` in a fill aesthetic in ggplot.
+
+- but the addition of missing counts into a histogram, allows us to understand that there are many missings - and we have a skewed distribution o Ozone.
+But we want to know where these values might have been located.
+tying this back to the nabular structure, and also the terminology of MCAR MAR, etc.
+(We would expect if data were MCAR for these points to be representatively thoughout the histogram)
 

### 2020-05-04 paper/tidy-missing-data-paper.Rmd
many small changes to combine figures and text, through to section 5

@@
 We define a new data structure, and a suite of new operations.
-Together, these provide a cohesive framework for handling, exploring, and imputing missing values.
+Together, these provide a connected framework for handling, exploring, and imputing missing values.
 These methods have been made available in the R package called `naniar`.
@@
 As data science has become a more solid field, theories and principles have developed to describe best practices.
-One such idea is 'tidy data,' which defines of a clean, analysis-ready format, that informs the workflow to convert raw data through a data analysis pipeline [@Wickham2014].
-The second idea is the grammar of graphics, which describes how to map data values into a visualization [@Wilkinson2012].
-These principles have been widely adopted.
-However, these principles do not address the problem of missing data.
-In particular, there is little guidance for missing value handling in the data workflow.
-Most analysis and visualisation software simply drops missing values when making a plot, although some packages like \pkg{ggplot2} provide a warning (Figure \@ref(fig:warning)).
+One such idea is 'tidy data,' which defines a clean, analysis-ready format, informing workflow converting raw data through a data analysis pipeline [@Wickham2014].
+Another idea is the grammar of graphics, describing how to map data values into a visualization [@Wilkinson2012].
+These principles have been widely adopted, but do not address the problem of missing data.
+In particular, there is little guidance on how to handle missing values in a data analsis workflow.
+Most analysis and visualisation software simply drop missing values when making a plot, although some packages (\pkg{ggplot2}) provide a warning (Figure \@ref(fig:warning)).
 
@@
 
-(**read over next two sections - Nick**)
-
 ## Imputation {#imputation}
@@
 
-(note: section to work on this week.)
-
-If a data structure is to be added to facilitate exploring missing data, it needs to be simple to understand and transport with a data analysis.
-A useful template is the $B$ matrix, where 0 and 1 indicate not missing and missing (respectively), as it is very common in missing data literature.
-This matrix was used to explore missing values in
-the interactive graphics library XGobi, called a "missing value
-shadow", or "shadow matrix", defined as a copy of the original data with
-indicator values of missingness. The shadow matrix could be interactively linked to the data.
-
-There are some limitations to the shadow matrix, however.
-Namely, the values 0 and 1 can be confusing representations of missing values - does 0 indicate an absence of observation, or the presence of a missing value?
+If a data structure is to be added to facilitate exploring missing data, it needs to be simple to understand, and transport with a data analysis.
+A useful template common in missing data literature, is the $B$ matrix, where 0 and 1 indicate not missing and missing, respectively.
+This matrix was used to explore missing values in the interactive graphics library XGobi, called a "missing value shadow", or "shadow matrix", defined as a copy of the original data with
+indicator values of missingness.
+The shadow matrix could be interactively linked to the data.
+However, there are some limitations to the shadow matrix.
+Namely, the values 0 and 1 can be confusing representations of missing values.
+Does 0 indicate an absence of observation, or the presence of a missing value?
 
@@
 Four features are added to the shadow matrix to facilitate analysis of missing data, illustrated in Figure \@ref(fig:shadow-matrix-progress).
-
-(**note** - condense details below into points here.)
 
@@
 Missing value labels can be represented as "factor" types values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing (Figure \@ref(fig:shadow-matrix-progress) parts 1-3).
-
-(NOTE: principle: clarity of labelling - transparent what the matrix means, because these could equally be "missing" or "present" - for this. Anybody looking at these values)
+This follows a principle: "**clarity of labelling**", as it is transparent what the matrix means, because these could equally be "missing" or "present" .
+Anybody looking at these values could understand what they mean, which is not the case of binary values.
 
@@
 This shifts the focus from the value of a variable to its missingness state, making intent clear when performing analysis.
-
-(It makes a clear distinction with `var_NA` being a random variable of the missingness of a variable, `var`. )
-
-(NOTE: change cohesiveness to connectedness)
+It makes a clear distinction with `var_NA` being a random variable of the missingness of a variable, `var`.
 
@@
 aq_imputed <- nabular(airquality) %>%
+  as.data.frame() %>% 
   simputation::impute_lm(Ozone ~ Temp + Wind) %>%
@@
 
-Missing values are often ignored when plotting data, but plots can help to reveal the type of missing value patterns, even those of MCAR, MAR or even MNAR.
-Here we summarise systematically exploring the missing patterns visually, and define useful plots to make relative to the nabular data structure, working from coarse to fine levels of granularity in understanding missingness.
-
-- The acronym MANET [@Unwin1996] stands for "Missings Are Now Equally Treated".
+Missing values are often ignored when plotting data - so much so that the data visualisation software, MANET, is an acronym for "Missings Are Now Equally Treated" [@Unwin1996].
+However, plots can help to reveal the type of missing value patterns, even those of MCAR, MAR or even MNAR.
+Here we summarise how to systematically explore missing patterns visually, and define useful plots to make relative to the nabular data structure.
 
 ## Overviews {#overviews}
-
-The  first step is to get an overview of the extent of the missingness.
-<!--Rather than highlighting the amount of complete data,-->  Overview visualisations for variables and cases are provided with `gg_miss_var` and `gg_miss_case` (Figure \@ref(fig:gg-miss-case-var)).
-These  missing data visualizations should draw attention to the amount of missings, and ordering by missingness.
-The "airquality" dataset (available in base R), included in base R, containing daily air quality measurements in New York, from May to September, 1973, is shown.
-Two of the variables contain missings and about one third of observations have one missing value.
-A tiny number of observations are missing on two variables.
-These summary plots compute proportions from the shadow matrix.
-The numerical statistics can also be reported, and is discussed in the next section.
 
@@
 
-```{r gg-miss-case-var, echo = FALSE, fig.show='hold', fig.cap = "Graphical summaries of missingness in variables (A) and cases (B) for the airquality data.
+The  first step is to get an overview of the extent of the missingness.
+Overview visualisations for variables and cases are provided with `gg_miss_var` and `gg_miss_case` (Figure \@ref(fig:gg-miss-case-var)), drawing attention to the amount of missings, and ordering by missingness.
+The "airquality" dataset (available in base R), is shown, and contains daily air quality measurements in New York, from May to September, 1973.
+We learn in figure \@ref("fig:gg-miss-case-var")A-B, that two variables contain missings and approximately one third of observations have one missing value.
+A tiny number of observations are missing on two variables.
+These summary plots compute proportions from the shadow matrix.
+The numerical statistics can also be reported, and are discussed in the next section.
+
+```{r gg-miss-case-var, echo = FALSE, fig.show='hold', fig.cap = "Graphical summaries of missingness in the airquality data.
+Missings invariables (A) and cases (B), and for a birds eye view with missingness as a heatmap in (C), and with clustering applied (D).
 There are missing values in Ozone and Solar.R, with Ozone having more missings.
 Not many cases have two missings.
-Most missingness is from cases with one missing value.", fig.height = 3, fig.width = 6, out.width = "90%", fig.align="center"}
-
-p1 <- gg_miss_var(airquality)
-
-p2 <- gg_miss_case(airquality)
-
-cowplot::plot_grid(p1,
-                   p2, 
-                   ncol = 2,
-                   labels = "AUTO")
+Most missingness is from cases with one missing value.
+The default output (C) and ordered by clustering on rows and columns (D).
+There are only missings in ozone and solar radiation, and there appears to be some structure to their missingness.", fig.height = 7, fig.width = 6, out.width = "90%", fig.align="center"}
+
+library(patchwork)
+aq_miss_var <- gg_miss_var(airquality) + labs(title = "A")
+
+aq_miss_case <- gg_miss_case(airquality) + labs(title = "B")
+
+aq_vis_miss <- vis_miss(airquality) + labs(title = "C")
+
+# c("fig.height = 3", "fig.width = 4.5")
+aq_vis_miss_cluster <- vis_miss(airquality, 
+                                 cluster = TRUE) + labs(title = "D")
+
+(aq_miss_var | aq_miss_case) / (aq_vis_miss | aq_vis_miss_cluster)
+
 ```
@@
 The shadow matrix allows both the variables and cases to be displayed using a heatmap style visualization, `vis_miss()` from \pkg{visdat} [@visdat].
-This also provides numerical summaries of missingness in the legend, and for each column (Figure \@ref(fig:vis-miss)).
-The user can also apply clustering to the rows, and arrange columns by missingness (Figure \@ref(fig:vis-miss)B).
+This also provides numerical summaries of missingness in the legend, and for each column (Figure \@ref(fig:gg-miss-case-var)C-D).
+The user can also apply clustering to the rows, and arrange columns by missingness (Figure \@ref(fig:gg-miss-case-var)D).
 Similar visualizations are available in other missing data packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}.
@@
 
-```{r vis-miss, fig.height = 3, fig.width = 4.5, out.width = "80%", fig.show = 'hold', echo = FALSE, dev = "png", dpi = 300, fig.cap = "Heatmap visualizations of missing data for the airquality dataset.
-(A) The default output and (B) ordered by clustering on rows and columns.
-There are only missings in ozone and solar radiation, and there appears to be some structure to their missingness."}
-
-p1 <- vis_miss(airquality)
-
-p2 <- vis_miss(airquality, cluster = TRUE)
-
-cowplot::plot_grid(p1,
-          p2,
-          nrow = 1,
-          labels = "AUTO")
+```{r airquality-upset, fig.height = 3, fig.width = 4, fig.cap = "The pattern of missingness in the airquality dataset shown in an upset plot.
+Only Ozone and Solar.R have missing values, and Ozone has the most missing values.
+There are 2 cases where both Solar.R and Ozone have missing values."}
+# 
+gg_miss_upset(airquality)
 
@@
 An alternative to a Venn diagram, an upset plot shows the size and features of overlapping sets, but scales well with more variables.
-Figure \@ref(fig:airquality-upset) shows the airquality data, and illustrates  that only Ozone and Solar.R have missing values.
-The bottom right show the combinations of missingness, the bottom left showing missingness in each variable, and the top showing the combinations of missingness.
-The bottom left shows Ozone has the most missing values, and the top shows 2 cases where both Solar.R and Ozone go missing together.
-The upset plot is constructed from the shadow matrix.
-
-```{r airquality-upset, fig.cap = "The pattern of missingness in the airquality dataset shown in an upset plot.
-Only Ozone and Solar.R have missing values, and Ozone has the most missing values.
-There are 2 cases where both Solar.R and Ozone have missing values.", fig.height = 3, fig.width = 4, out.width = "75%"}
-
-gg_miss_upset(airquality)
-
-```
+This can be applied to missing values, and an upset plot can be constructed from the shadow matrix, Figure \@ref(fig:airquality-upset) visualises overlapping amounts of missing data in the airquality data.
+The bottom right shows the combinations of missingness, the top panel shows the size of these combinations, and the bottom left shows missingness in each variable.
+This shows us clearly that Ozone has the most missing values, and there are 2 cases where values in both Solar.R and Ozone are missing together, illustrating similar points to \@ref(fig:gg-miss-case-var).
 
 ## Univariate {#univariate}
-
-- gg_miss_var and gg_miss_upset and vis_miss are all showing how we can summarise the missings.
 
@@
 **Imputing values below the range**.
-To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed values below the range of data using `impute_below` (by default imputing 10% below the range).
-Figure \@ref(fig:impute-shift-histogram) shows a histogram of Ozone, the number of missing values on its left.
-Ozone is shown in a different color by referring to the variable `Ozone` as `Ozone_NA` in a fill aesthetic in ggplot.
-
-- but the addition of missing counts into a histogram, allows us to understand that there are many missings - and we have a skewed distribution o Ozone.
-But we want to know where these values might have been located.
-tying this back to the nabular structure, and also the terminology of MCAR MAR, etc.
-(We would expect if data were MCAR for these points to be representatively thoughout the histogram)
-
-```{r impute-shift-histogram, fig.height = 3, fig.width = 5, fig.cap = "A histogram using nabular data to show the values and missings in ozone.
+To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed below the range of data using `impute_below` (by default imputing 10% below the range).
+Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone, the number of missing values on its left.
+Ozone is shown in a different color by identifying the imputed values in _nabular_ data, by referring to the variable `Ozone` as `Ozone_NA` in a fill aesthetic in ggplot.
+The addition of the counts of missings in the histogram allows us to understand there are many missings, and that we have a skewed distribution o Ozone.
+
+```{r impute-shift-histogram, fig.height = 3, fig.width = 5, fig.cap = "Univariate summaries of missingness.
+(A) A histogram using nabular data to show the values and missings in ozone.
 Values are imputed below the range to show the number of missings in Ozone and colored according to missingness of ozone (`Ozone\\_NA`).
-There are about 35 missings in Ozone.", error = FALSE, warning = FALSE, message = FALSE, out.width = "75%"}
+There are about 35 missings in Ozone.
+Panel C shows Temperature according to missingness in ozone from in the airquality dataset.
+A histogram of temperature facetted by the missingness of ozone (B), or a density of temperature colored by missingness in ozone (C).
+These show a cluster of low temperature observations with missing ozone values, but temperature is otherwise similar.", error = FALSE, warning = FALSE, message = FALSE, out.width = "75%", fig.show = 'hold'}
 
@@
   scale_fill_brewer(palette = "Dark2") + 
-  theme_minimal()
-```
-
-**Univariate split by missingness**.
-Missingness of one variable can also be used to display different distributions in another.
-Figure \@ref(fig:bind-shadow-density) shows the values of temperature when ozone is present, and missing.
-Figure \@ref(fig:bind-shadow-density)A is a faceted histogram, and Figure \@ref(fig:bind-shadow-density)B is an overlaid density.
-This shows how values of temperature are affected by the missingness of ozone, and reveals a cluster of low temperature observations with missing ozone values.
-
-```{r bind-shadow-density, fig.height = 3, fig.width = 5, out.width = "49%", error = FALSE, message = FALSE, fig.cap = "Temperature according to missingness in ozone from in the airquality dataset.
-A histogram of temperature facetted by the missingness of ozone (A), or a density of temperature colored by missingness in ozone (B).
-These show a cluster of low temperature observations with missing ozone values, but temperature is otherwise similar.", fig.show = 'hold'}
+  theme_minimal() +
+  labs(tag = "A")
 
@@
   facet_wrap(~Ozone_NA) + 
-  labs(tag = "A")
+  labs(tag = "B")
 
@@
   scale_colour_brewer(palette = "Dark2")  + 
-  labs(tag = "B")
-
+  labs(tag = "C")
+
+```
+
+**Univariate split by missingness**.
+Missingness of one variable can also be used to display different distributions in another.
+Figure \@ref(fig:bind-shadow-density) shows the values of temperature when ozone is present, and missing.
+Figure \@ref(fig:bind-shadow-density)A is a faceted histogram, and Figure \@ref(fig:bind-shadow-density)B is an overlaid density.
+This shows how values of temperature are affected by the missingness of ozone, and reveals a cluster of low temperature observations with missing ozone values.
+The nabular structure of the data reveals the distribution of the missing values
+
+```
+To know where these values might have been located.
+tying this back to the nabular structure, and also the terminology of MCAR MAR, etc.
+(We would expect if data were MCAR for these points to be representatively thoughout the histogram)
+
+- be sure to clear up vague language (it is important to...etc., )
+- Missings are often ignored
 ```

### 2020-05-04 paper/tidy-missing-data-paper.Rmd
tidying up some wording in the nabular data section

@@
 It is a portmanteau of `NA` and `tabular`.
-_Nabular_ data keeps corresponding rows together, removing the possibility of mismatching records, explicitly linking missing values to data.
+_Nabular_ explicitly links missing values to data, keeping corresponding observations together and removing the possibility of mismatching records.
 _Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable, `var`, as `var_NA`.
@@
 This means when _nabular_ data is imputed, those imputed values can easily be identified in analysis by referring to the coordinated names.
-_Nabular_ data is created with `nabular(data)` (Figure \@ref(fig:nabularfig)).
 _Nabular_ data is not unlike classical data formats with quality or flag columns associated with each measured variable, e.g.
 Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].
+_Nabular_ data is created with `nabular(data)` (Figure \@ref(fig:nabularfig)).
 
@@
 
-The _nabular_ data structure provides three key benefits over on-the-fly creating a missingness matrix $B$.
+The _nabular_ data structure provides three key benefits over creating a missingness matrix, $B$, on-the-fly.
 Firstly, missing value labels `NA` and `!NA` are clearer than TRUE or FALSE.
@@
 Finally, the logical matrix cannot capture which values are imputed, if imputation has already taken place.
-Imputing values on _nabular_ data automatically tracks these value changes.
+Imputing values on _nabular_ data automatically tracks these imputations.
 

### 2020-05-05 paper/tidy-missing-data-paper.Rmd
tidy up abstract, and part of intro

@@
 Together, these provide a connected framework for handling, exploring, and imputing missing values.
-These methods have been made available in the R package called `naniar`.
+These methods are available in the R package `naniar`.
 keywords:
@@
 As data science has become a more solid field, theories and principles have developed to describe best practices.
-One such idea is 'tidy data,' which defines a clean, analysis-ready format, informing workflow converting raw data through a data analysis pipeline [@Wickham2014].
+One such idea is 'tidy data,' which defines a clean, analysis-ready format that informs workflows converting raw data through a data analysis pipeline [@Wickham2014].
 Another idea is the grammar of graphics, describing how to map data values into a visualization [@Wilkinson2012].
@@
 In particular, there is little guidance on how to handle missing values in a data analsis workflow.
-Most analysis and visualisation software simply drop missing values when making a plot, although some packages (\pkg{ggplot2}) provide a warning (Figure \@ref(fig:warning)).
+Most analysis and visualisation software simply drop missing values when making a plot, although some (\pkg{ggplot2}) provide a warning (Figure \@ref(fig:warning)).
 
@@
 ```{r warning, fig.cap = "How ggplot2 behaves when displaying missing values.
-A warning message is displayed, but missing values are not shown in the plot", fig.height = 2.5, fig.width = 2.5, out.width = "75%", warning = FALSE, fig.show = 'hold'}
-
+A warning message is displayed, but missing values are not shown in the plot", fig.height = 4, fig.width = 8, out.width = "75%", warning = FALSE, fig.show = 'hold'}
 p1 <- ggplot(oceanbuoys,
@@
   geom_point(size = 0.5) +
-  theme(aspect.ratio = 1)
-
-ggsave("images/geom-point-warn.png", 
-       width = 4, 
-       height = 4, 
-       units = c("in"))
-
-img_gg_p <- image_read("images/geom-point-warn.png")
-
-img_gg_warn <- image_read("images/ggplot-warn.jpeg")
-
-image_append(c(img_gg_warn, img_gg_p)) %>% 
-  image_write("images/ggplot2-warning-and-plot.png", "png")
-
-knitr::include_graphics("images/ggplot2-warning-and-plot.png")
+  theme(aspect.ratio = 1) + 
+  labs(x = "Humidity",
+       y = "Air Temperature (C)")
+
+p1 + grid::textGrob(
+  label = "Warning message:
+Removed 171 rows containing 
+missing values (geom_point).", 
+gp = grid::gpar(fontsize = 14,
+                fontfamily = "Courier",
+                col = "salmon", 
+                fontface = "bold"))
 
@@
 This presents a challenge in analysis, as it is difficult to verify, and implies bias in analysis due to the unobserved phenomena.
-
-<!-- (possibly remove this next para?) -->
-
-Visualizations can help assess whether data is MCAR, MAR or MNAR. Imputation is recommended in most cases of missingness dependence.
+Visualizations can help assess whether data is MCAR, MAR or MNAR.
+Imputation is recommended in most cases of missingness dependence.
+
+<!-- (NOTE - are the above last two sentences OK?) -->
 
@@
 However, there are some limitations to the shadow matrix.
-Namely, the values 0 and 1 can be confusing representations of missing values.
-Does 0 indicate an absence of observation, or the presence of a missing value?
+Namely, the values 0 and 1 can be confusing representations of missing values: Does 0 indicate an absence of observation, or the presence of a missing value?
 
@@
 **Missing value labels**.
-Simple labels, "NA" and "!NA", for missing and not missing, clearly identifies missing values for analysis and plotting.
-This is an improvement over ($B$), which makes it unclear if 0 indicates the absence (0) of observation, or the presence (1) of a missing value.
-Missing value labels can be represented as "factor" types values, with text values "NA" and "!NA" and underlying number values 1 and 0, for missing and not missing (Figure \@ref(fig:shadow-matrix-progress) parts 1-3).
-This follows a principle: "**clarity of labelling**", as it is transparent what the matrix means, because these could equally be "missing" or "present" .
-Anybody looking at these values could understand what they mean, which is not the case of binary values.
+Simple labels, "NA" and "!NA", for missing and not missing, clearly identify missing values for analysis and plotting (Figure \@ref(fig:shadow-matrix-progress) parts 1-3).
+This improves the 0 and 1 values in $B$, which are unclear, since 0 and 1 could interchangably indicate the absence (0) of observation, or the presence (1) of a missing value.
+These values follow a principle: "**clarity of labelling**" - the matrix's meaning is transparent, and anybody looking at these values could understand what they mean, which is not the case of binary values.
+Equally, these values could instead be "missing" or "present".
 
@@
 These could be also used to indicate imputations, (Table \@ref(tab:shadow-encoding)).
-The underlying number representation changes from binary to integer, allowing for unlimited missingness features.
-For example, instrument failure and drop out would be represented as 2 and 3 (Table \@ref(tab:shadow-encoding)).
-Encoding special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in \pkg{naniar} (\@ref(verbs-recode)).
-{Other statistical programs (SPSS, SAS, and Stata) represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` - `.z`.
-SAS, SPSS, and STATA all implement }
-
-{NOTE: The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state.
-Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R.}
-
-{NOTE: (discuss the principles, rather than what we've done)}. {(in the same column you have mixed information, which breaks the tidy principle of one column having one type of value)}.
 
@@
 **Coordinated names**: Variables in the shadow matrix gain a consistent short suffix, "_NA", keeping names coordinated throughout analysis (Figure \@ref(fig:shadow-matrix-progress) part 4).
-The suffix is short and easy to remember in analysis or visualization.
-This shifts the focus from the value of a variable to its missingness state, making intent clear when performing analysis.
+The suffix is short and easy to remember during data a analysis, and shifts the focus from the value of a variable to its missingness state.
 It makes a clear distinction with `var_NA` being a random variable of the missingness of a variable, `var`.
@@
 This is discussed in more detail below in \@ref(nabular-data).
+
+Note: I think that Figure \@ref(fig:shadow-matrix-progress), Table \@ref(tab:shadow-encoding), and Figure \@ref(fig:nabularfig) could be combined into one useful graphic.
+This could form the "key graphic" of the paper.
+I also think that the point about the shadow value (`var_NA`) being a factor and having an underlying value is not directly relevant to the paper.
 
@@
 _Nabular_ explicitly links missing values to data, keeping corresponding observations together and removing the possibility of mismatching records.
-_Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable, `var`, as `var_NA`.
+_Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable in a coordinated way - with `var`, as `var_NA` during analysis.
 _Nabular_ data is a snapshot of the missingness of the data.
-This means when _nabular_ data is imputed, those imputed values can easily be identified in analysis by referring to the coordinated names.
+This means when _nabular_ data are imputed, those imputed values can easily be identified in analysis (`var` vs `var_NA`).
 _Nabular_ data is not unlike classical data formats with quality or flag columns associated with each measured variable, e.g.
@@
 Secondly, special missing values cannot be added easily during analysis with a logical matrix.
-Finally, the logical matrix cannot capture which values are imputed, if imputation has already taken place.
+Finally, the logical matrix cannot capture which values are imputed if imputation has already taken place.
 Imputing values on _nabular_ data automatically tracks these imputations.
 
-Using additional columns to represent missingness information follows recently published best practices for data organization, described in @Ellis2017 and @Broman2017: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column.
+Using additional columns to represent missingness information follows best practices for data organization, described in @Ellis2017 and @Broman2017: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column.
 Here they suggest to indicate censored data with an extra variable called "VariableNameCensored", which would be TRUE if censored, otherwise FALSE.
-This information can now be represented in the shadow columns as special missings.
+This information can now be represented in the shadow columns as special missing values.
+In practice, encoding special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in \pkg{naniar} (\@ref(verbs-recode)).
+
+- Other statistical programs (SPSS, SAS, and Stata) represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` - `.z`.
+SAS, SPSS, and STATA all implement.
+
+- The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state.
+Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R.
+- (in the same column you have mixed information, which breaks the tidy principle of one column having one type of value)
 
@@
 Ozone is shown in a different color by identifying the imputed values in _nabular_ data, by referring to the variable `Ozone` as `Ozone_NA` in a fill aesthetic in ggplot.
-The addition of the counts of missings in the histogram allows us to understand there are many missings, and that we have a skewed distribution o Ozone.
-
-```{r impute-shift-histogram, fig.height = 3, fig.width = 5, fig.cap = "Univariate summaries of missingness.
+The addition of the counts of missings in the histogram allows us to understand there are many missings, and that we have a skewed distribution of Ozone.
+
+```{r impute-shift-histogram, fig.height = 2.5, fig.width = 5, fig.cap = "Univariate summaries of missingness.
 (A) A histogram using nabular data to show the values and missings in ozone.
@@
   theme_minimal() +
-  labs(tag = "A")
+  labs(tag = "A") + 
+  theme(legend.position = "bottom")
 
@@
   scale_colour_brewer(palette = "Dark2")  + 
+  theme(legend.position = "bottom") +
   labs(tag = "C")

### 2020-05-05 paper/tidy-missing-data-paper.Rmd
complete abstract and section 1

@@
 This paper addresses this gap.
-The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as an integral part of data analysis workflows.
+The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows.
 We define a new data structure, and a suite of new operations.
@@
 These principles have been widely adopted, but do not address the problem of missing data.
-In particular, there is little guidance on how to handle missing values in a data analsis workflow.
+In particular, there is little guidance on how to handle missing values in a data analysis workflow.
 Most analysis and visualisation software simply drop missing values when making a plot, although some (\pkg{ggplot2}) provide a warning (Figure \@ref(fig:warning)).
@@
 of missingness [@Tierney2015; @Barnett2017], but are not definitive. While there
-might be other times where the missing data mechanism is obvious, determining
-this is not straightforward. To understand their structure, the analyst must explore the data
-with visualizations, summaries, and modelling. It is an iterative procedure, to arrive at a reasonably confident state of understanding the missing data mechanism.
+are times where the missing data mechanism is obvious, determining this is typically not straightforward. To understand their structure and possible mechanisms, the analyst must explore the data
+with visualizations, summaries, and modelling, in an iterative fashion.
 
@@
 The graphics literature provides several solutions for exploring missings visually by incorporating them into the plot in some way.
-For example, imputing values to be 10% below the minimum will include all observations into a scatter plot.
-
-However, this is often implicit, and not transparent to the user.
-(**NOTE** include possible brief example from mice/mi/VIM).
-The ideas from the graphics literature need to be translated into tidy data tools to integrate missing data handling in a data pipeline.
+For example, imputing values to be 10% below the minimum to include all observations into a scatter plot [@Cook2007], or treating missing values as an equal category of data [@Unwin1996].
+The ideas from the graphics literature need to be translated into tidy data tools to integrate missing data handling in a data analysis pipeline.
 
 This paper is organized in the following way.
-The next section (\@ref(background)) provides the background to to tidy data principles and tools (\@ref(tidy-data-concepts) and missing data representations (\@ref(missing-data-rep-dep).
+The next section (\@ref(background)) provides the background to tidy data principles and tools (\@ref(tidy-data-concepts)) and missing data representations (\@ref(missing-data-rep-dep)).
 Section \@ref(existing-software) summarises existing software for handling missing data.
 The new work extending the tidy tools to facilitate exploring, visualizing, and imputing missing data is discussed in Section \@ref(extensions).
+Graphics (\@ref(graphics)) and Numerical summaries (\@ref(num-sum)) of missing values are then discussed.
 An application illustrating the use of the new methods is shown in Section \@ref(case-study).

### 2020-05-06 paper/tidy-missing-data-paper.Rmd
section two complete - add citation to R language

@@
 
-Features of tidy data were formally discussed in @Wickham2014, and were discussed in terms of their importance for data science by @Donoho2017, and tools for data analysis.
+Features of tidy data were formally described in @Wickham2014, and were discussed in terms of their importance for data science by @Donoho2017, and tools for data analysis.
 Tidy data is defined by @r4ds as:
@@
 This helps the analyst works swiftly and clearly, closing up opportunities for errors.
-Tidy data principles are general, but are comprehensively implemented in the R programming language, so this paper focusses on R.
+Tidy data principles are general, but are comprehensively implemented in the R programming language, so this paper focusses on R [@rcore].
 
@@
 These provide the five "verbs" of data manipulation: data reshaping, sorting, filtering, transforming, and aggregating.
-**Data reshaping** goes from long to wide formats; **sorting** arranges rows in a specific order; **filtering** removes rows based on a condition; **transforming**, changes existing variables or adds new ones; **aggregating** creates a single value from many values, say, for example, by computing the minimum, maximum, and mean.
+Data reshaping goes from long to wide formats; sorting arranges rows in a specific order; filtering removes rows based on a condition; transforming, changes existing variables or adds new ones; aggregating creates a single value from many values, say, for example, in computing the minimum, maximum, and mean.
 
@@
 For example, estimated coefficients, predictions, and residuals from one model cannot be easily combined with the output of another model.
-Messy models have been partially addressed with the \pkg{broom} package, which tidies up model outputs for data analysis, and the developing \pkg{recipes} package, which helps make modelling input- and output-tidy [@recipes].
+Messy models have been partially addressed with the \pkg{broom} package, which tidies up model outputs into a tidy data format for use data analysis, and the developing \pkg{recipes} package, which helps make modelling input- and output-tidy [@recipes].
 
@@
 Missing data tools can be used to perform imputations, missing data diagnostics, and data visualizations.
-However, these tools suffer the same problems as modelling for imputation: They use tidy input, but produce messy output - their output is challenging to integrate with other steps of data analysis.
+However, these tools suffer the same problems as modelling: They use tidy input, but produce messy output - their output is challenging to integrate with other steps of data analysis.
 The complex, often multivariate nature of imputation methods also makes makes them difficult to represent.
@@
 The distribution of missing values in $b_{ij}$ can depend on the entire dataset, represented as $Y = (Y_{obs}, Y_{miss})$.
-This relationship can be defined by the _missing data model_ $Pr(r_{ij} | Y_{obs}, Y_{miss}, \psi)$, the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, $\psi$.
+This relationship can be defined by the _missing data model_ $Pr(b_{ij} | Y_{obs}, Y_{miss}, \psi)$, the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, $\psi$.
 This helps to precisely define categories of missing values.
 
-**MCAR** is where values being missing have no association with observed or unobserved data, that is, $Pr(R = 1 | Y_{obs}, Y_{miss}) = Pr(R = 1 | \psi)$.
+**MCAR** is where values being missing have no association with observed or unobserved data, that is, $Pr(B = 1 | Y_{obs}, Y_{miss}) = Pr(B = 1 | \psi)$.
 Essentially, the probability of an observation being missing is unrelated to anything else, only the parameter $\psi$, the overall probability of missingness.
 Although a convenient scenario, it is not actually possible to confirm, or clearly distinguish from MAR, as it relies on statements on data unobserved.
-In **MAR**, missingness only depends on data observed, not data missing, that is, $Pr(R = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(R = 1 | Y_{obs}, \psi)$.
+In **MAR**, missingness only depends on data observed, not data missing, that is, $Pr(B = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(B = 1 | Y_{obs}, \psi)$.
 Some structure or dependence between missing and observed values is allowed, provided it can be explained by data observed, and some overall probability of missingness.
-In **MNAR**, missingness is related to values observed, and unobserved: $Pr(R = 1 | Y_{obs}, Y_{miss}, \psi)$.
+In **MNAR**, missingness is related to values observed, and unobserved: $Pr(B = 1 | Y_{obs}, Y_{miss}, \psi)$.
 This assumes conditioning on all observations: data goes missing due to some phenomena unobserved, including the structure of the missing data itself.
 This presents a challenge in analysis, as it is difficult to verify, and implies bias in analysis due to the unobserved phenomena.
-Visualizations can help assess whether data is MCAR, MAR or MNAR.
+Visualizations can help assess whether data may be MCAR, MAR or MNAR.
 Imputation is recommended in most cases of missingness dependence.
 
-<!-- (NOTE - are the above last two sentences OK?) -->
+<!-- (NOTE: Are the above last two sentences OK?) -->
+<!-- (NOTE: trying to come up with a sentence that leads from this section to the next) -->
 

### 2020-05-06 paper/tidy-missing-data-paper.Rmd
cleaning up, up to section 3.2

@@
 \pkg{simputation} provides an interface to imputation methods from VIM, in addition providing hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977].
-\pkg{simputation} provides a consistent formula interface to all imputation, and always returns a dataframe with the updated imputations.
-\pkg{Hmisc} [@Hmisc], provide predictive mean matching, \pkg{imputeTS} [@imputeTS], provides time series imputation methods, and \pkg{missMDA} [@missMDA], imputes data using principal components analysis.
+\pkg{simputation} provides a consistent formula interface for all imputation methods, and always returns a dataframe with the updated imputations.
+\pkg{Hmisc} [@Hmisc], provides predictive mean matching, \pkg{imputeTS} [@imputeTS], provides time series imputation methods, and \pkg{missMDA} [@missMDA], imputes data using principal components analysis.
 
@@
 
-\pkg{Amelia} [@amelia] assumes data are multivariate normal, and samples from the posterior using the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm [@Honaker2010] and allows for incorporation of information on the values in a prior.
+\pkg{Amelia} [@amelia] assumes data are multivariate normal, and samples from the posterior, and allows for incorporation of information on the values in a prior.
+It uses the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm [@Honaker2010].
 \pkg{norm} [@norm], provides multiple imputation using EM for multivariate normal data, drawing from methods in the NORM software [@schafer-norm].
@@
 \pkg{mi} [@mi] also uses Bayesian models for imputation, providing better handling of semi-continuous values, and data with structural or perfect correlation.
-A collection of analysis models are also provided in \pkg{mi}, which work with the specially multiply imputed data, including linear models, generalized linear models, and their corresponding Bayesian components.
+A collection of analysis models are also provided in \pkg{mi}, to work with data it has imputed.
+These include linear models, generalized linear models, and their corresponding Bayesian components.
 This approach promotes fluid workflow, with a similar penalty to tidying up model output, which is still messy.
@@
 Each imputation method provides practical methods for different usecases, but most have different output structures, and do not have consistent interfaces in their implementation.
-This makes then inherently messy and challenging to integrate into an analysis pipeline.
+This makes them inherently messy and challenging to integrate into an analysis pipeline.
 For example, combining different imputation methods from different pieces of software is not currently straightforward.

### 2020-05-07 paper/tidy-missing-data-paper.Rmd
section three (existing software) tidied up

@@
 Methods for exploring, understanding, and imputing missing data are more accessible now than they have ever been.
-In particular, the number of methods and implementations of imputation are seeing continued growth.
 Values can be imputed with one value (single imputation), or multiple values (multiple imputation), creating $m$ datasets.
-This section discusses existing software for single and multiple imputation, and exploration.
+This section discusses existing software for single and multiple imputation, and missing data exploration.
 
@@
 For example, combining different imputation methods from different pieces of software is not currently straightforward.
-\pkg{simputation} is different from other imputation packages, and resolves some of these complications with a simple approach of a unified syntax for all imputation, and always returning a dataframe containing imputed values.
+\pkg{simputation} resolves some of these complications with a simple approach of a unified syntax for all imputation, and always returns a dataframe of imputed values.
 This reduces the friction of working with other tools, but comes at the cost of identifying imputed values.
@@
 The MANET software is no longer maintained and cannot be installed.
-The approach of @Swayne1998 in the software XGobi, further developed in ggobi [@Cook2007], focussed more on multivariate quantitative data.
+The approach of @Swayne1998 in the software XGobi, further developed in ggobi [@Cook2007], focussed on multivariate quantitative data.
 Missingness is incorporated into plots in ggobi by setting them to be 10% below the minimum value.
 
-MissingDataGUI provides a Graphhical User Interface (GUI) for exploring missing data structure both numerically and visually.
+MissingDataGUI provides a Graphical User Interface (GUI) for exploring missing data structure, both numerically and visually.
 Using a GUI to explore missing data facilitates rapid insight into missingness structures.
@@
 
-\pkg{ggplot2} incorporates missingness into visualizations only when mapping a discrete variable to a graph aesthetic.
-This has some limitations.
-For example, data of school grades and test scores, a boxplot visualization with school grade and test score mapped to the x- and y-axis (Figure \@ref(fig:gg-box-na)).
+\pkg{ggplot2} removes missing values with a warning (Figure \@ref(fig:warning)), and only incorporates missingness into visualizations when mapping a discrete variable with missings to a graph aesthetic.
+This has some limitations, shown in Figure \@ref(fig:gg-box-na), a boxplot visualization of school grade and test score mapped to the x- and y-axis .
 If there are missings in a continuous variable like test score, \pkg{ggplot2} omits the missings and prints a warning message.

### 2020-05-07 paper/tidy-missing-data-paper.Rmd
moving through section 4

@@
 Applying tidyverse principles has the potential to clarify missing data exploration, visualization, and imputation. This
-section discusses how these principles are applied for data structures \@ref(data-structure), common operations (verbs) \@ref(verbs), graphics \@ref(graphics), and data summaries \@ref(num-sum).
-The design and naming of these features is important, and care has been taken to make the names intuitive for their purpose.
+section discusses how these principles are applied for data structures (\@ref(data-structure)), common operations (verbs) (\@ref(verbs)), graphics (\@ref(graphics)), and data summaries (\@ref(num-sum)).
+Care has been taken to make the names and design these features intuitive for their purpose, and is discussed throughout.
 
@@
 
-If a data structure is to be added to facilitate exploring missing data, it needs to be simple to understand, and transport with a data analysis.
+A data structure facilitating exploration of missing data needs to be simple to reason with and to transport with a data analysis, otherwise it will not be used.
 A useful template common in missing data literature, is the $B$ matrix, where 0 and 1 indicate not missing and missing, respectively.
@@
 However, there are some limitations to the shadow matrix.
-Namely, the values 0 and 1 can be confusing representations of missing values: Does 0 indicate an absence of observation, or the presence of a missing value?
+Namely, the values 0 and 1 can be confusing representations of missing values, since it is not clear if 0 indicates an absence of observation, or the presence of a missing value.
+
+<!-- since 0 and 1 could interchangably indicate the absence (0) of observation, or the presence (1) of a missing value. -->
 
@@
 **Missing value labels**.
-Simple labels, "NA" and "!NA", for missing and not missing, clearly identify missing values for analysis and plotting (Figure \@ref(fig:shadow-matrix-progress) parts 1-3).
-This improves the 0 and 1 values in $B$, which are unclear, since 0 and 1 could interchangably indicate the absence (0) of observation, or the presence (1) of a missing value.
+Simple labels for missing and not missing to clearly identify missing values for analysis and plotting.
+We propose "NA" and "!NA".
+(Figure \@ref(fig:shadow-matrix-progress) parts 1-3).
+This improves the 0 and 1 values in $B$, which do not clearly identify missingness.
 These values follow a principle: "**clarity of labelling**" - the matrix's meaning is transparent, and anybody looking at these values could understand what they mean, which is not the case of binary values.
@@
 **Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA_\<suffix\>" - for example: `NA_instrument` uses a short label, "instrument", indicating instrument error resulting in missing values.
-These could be also used to indicate imputations, (Table \@ref(tab:shadow-encoding)).
+These could be also used to indicate imputations, (Table \@ref(tab:shadow-encoding)) (note: the part of the new figure).
 
 1.
-**Coordinated names**: Variables in the shadow matrix gain a consistent short suffix, "_NA", keeping names coordinated throughout analysis (Figure \@ref(fig:shadow-matrix-progress) part 4).
-The suffix is short and easy to remember during data a analysis, and shifts the focus from the value of a variable to its missingness state.
+**Coordinated names**: Variable names in the shadow matrix gain a consistent short suffix, "_NA", keeping names coordinated throughout analysis (Figure \@ref(fig:shadow-matrix-progress) part 4 - note: update this figure).
 It makes a clear distinction with `var_NA` being a random variable of the missingness of a variable, `var`.
+This suffix is short and easy to remember during data analysis, and shifts the focus from the value of a variable, to its missingness state.
 
 1.
-**Connectedness**: Binding the shadow matrix column-wise to the original data creates a single connected _nabular_ data format, so that it cannot be out of sync with the data.
-useful for visualization, summaries, and tracking imputed values.
-This is discussed in more detail below in \@ref(nabular-data).
-
-Note: I think that Figure \@ref(fig:shadow-matrix-progress), Table \@ref(tab:shadow-encoding), and Figure \@ref(fig:nabularfig) could be combined into one useful graphic.
+**Connectedness**: Binding the shadow matrix column-wise to the original data creates a single, connected, _nabular_ data format, which cannot be out of sync with the data.
+It is useful for visualization, summaries, and tracking imputed values, discussed in more detail in \@ref(nabular-data).
+
+<!-- Note: I think that Figure \@ref(fig:shadow-matrix-progress), Table \@ref(tab:shadow-encoding), and Figure \@ref(fig:nabularfig) could be combined into one useful graphic.
 This could form the "key graphic" of the paper.
 I also think that the point about the shadow value (`var_NA`) being a factor and having an underlying value is not directly relevant to the paper.
+-->
 
@@
 It is a portmanteau of `NA` and `tabular`.
-_Nabular_ explicitly links missing values to data, keeping corresponding observations together and removing the possibility of mismatching records.
-_Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable in a coordinated way - with `var`, as `var_NA` during analysis.
+_Nabular_ data explicitly links missing values to data, keeping corresponding observations together and removing the possibility of mismatching records.
+_Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable in a coordinated way: the missingness of `var`, as `var_NA` during analysis.
 _Nabular_ data is a snapshot of the missingness of the data.
 This means when _nabular_ data are imputed, those imputed values can easily be identified in analysis (`var` vs `var_NA`).
-_Nabular_ data is not unlike classical data formats with quality or flag columns associated with each measured variable, e.g.
+_Nabular_ data is not unlike classical data formats with quality or flag variables associated with each measured variable, e.g.
 Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].
@@
 ```
+
+<!-- Note: I'm not sure why I talk about this on the fly stuff here, and I wonder if this paragraph below could be removed? -->
 
@@
 This information can now be represented in the shadow columns as special missing values.
-In practice, encoding special missing values is achieved by defining logical conditions and suffixes with the `recode_shadow` function in \pkg{naniar} (\@ref(verbs-recode)).
-
-- Other statistical programs (SPSS, SAS, and Stata) represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` - `.z`.
+Encoding special missing values is achieved by defining logical conditions and suffixes.
+This is implemented with the `recode_shadow` function in \pkg{naniar} (\@ref(verbs-recode)).
+
+<!-- Note: I'm not sure where I should put these three parts below parts in the paper: -->
+
+<!-- - Other statistical programs (SPSS, SAS, and Stata) represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` - `.z`.
 SAS, SPSS, and STATA all implement.
-
-- The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state.
+-->
+
+<!-- - The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state.
 Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R.
-- (in the same column you have mixed information, which breaks the tidy principle of one column having one type of value)
+-->
+<!-- - (in the same column you have mixed information, which breaks the tidy principle of one column having one type of value) -->
 

### 2020-05-07 paper/tidy-missing-data-paper.Rmd
completed section 4

@@
 To facilitate exploration, summaries of missingness can be **added** as a summary column to the original data.
-Data matrix can be augmented with the **shadow matrix** values, helping explore missing data, as well as facilitating the process of **imputing**, and **tracking**.
+The data can be augmented with the **shadow matrix** values, helping explore missing data, as well as facilitating the process of **imputing**, and **tracking**.
 Finally, unusual or specially coded missing values can be **flagged**.
@@
 This operation is used to search the data for specific conventional representations of missings, such as, "N/A", "MISSING", `-99`. 
-The function is called  `miss_scan_count()`, and it returns a table of occurrences of that value for each variable.
-Two more data objects, `common_na_numbers` and `common_na_strings` provide a list of common NA values for numbers and characters, to help users check for the typical representations of missing.
+This is implemented in the function `miss_scan_count()`, which returns a table of occurrences of that value for each variable.
+A list of common NA values for numbers and characters, can be provided to help check for typical representations of missings.
+These are implemented in `naniar` as `common_na_numbers` and `common_na_strings`, for numbers and characters.
 
@@
 
-### replace: Replacing missing value label with another {#verbs-replace-with}
-
-Once the different representations of possible missing values have been identified, these values can be replaced using the function, `replace_with_na()`.
-A dataset could have the values `-99` in the `x` column replaced with the following code:
+### replace: Replacing values with missing values {#verbs-replace-with}
+
+Once possible missing values have been identified, these values can be replaced.
+For example, a dataset could have the values `-99` meaning a missing value.
+`naniar` implements replacement with the function, `replace_with_na()`.
+Values -99 could be replaced in the `x` column with:
 `replace_with_na(dat_ms, replace = list(x = -99))`.
@@
 
-Summary information such as the number of missings in a given case can be useful in understanding missingness structure, and even more so when this information is kept alongside the data.
-Functions starting with `add_` exist in `dplyr` to add count information for particular groups or conditions to the data.
-`naniar` includes functions to add missingness summary information to the data, such as the number or proportion of missingness, the missingness cluster, or if there are any missings, using: `add_n_miss()`, `add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (Table \@ref(tab:add-missing-info)).
-There are also functions for adding information about shadow values or nice labels for any missing values with `add_label_shadow()` and `add_label_missings()`.
-An example use of these features is in @Tierney2015, where the proportion of
-missings is used as the outcome in a model, to identify variables and values
+Understanding the missingness structure can be improved by adding summary information alongside the data. For example, in @Tierney2015, the proportion of
+missings in a row is used as the outcome in a model to identify variables
 important in predicting missingness structures.
+`naniar` implements a series of functions to add these missingness summaries to the data, starting with `add_`.
+These are inspired by the `add_count` function in `dplyr`, which adds count information for specified groups or conditions.
+`naniar` provides operations to add the number or proportion of missingness, the missingness cluster, or the precense of any missings, with: `add_n_miss()`, `add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (Table \@ref(tab:add-missing-info)).
+There are also functions for adding information about shadow values, and readable labels for any missing values with `add_label_shadow()` and `add_label_missings()`.
 
@@
 This facilitates visualization and summaries, and allows for imputed values to be tracked.
-_Nabular_ data can be created with `nabular()`, as below:
+_Nabular_ data can be created with `nabular()`:
 
@@
 For example, there might be special codes to mark an individual dropping out of a study, known instrument failure in weather instruments, or for values censored in analysis.
-These special types of missingness can be encoded in the shadow matrix of  _nabular_  data, using `recode_shadow()`.
-Using `recode_shadow()` requires specifying the variable you want to contain the flagged value, the condition you want this to occur at, and a suffix for the new type of missing value.
+`naniar` provides tools to encode these special types of missingness in the shadow matrix of  _nabular_  data, using `recode_shadow()`.
+This requires specifying the variable to contain the flagged value, the condition for flagging, and a suffix.
 This is then recoded as a new factor level in the shadow matrix, so every column is aware of all possible new values of missingness.
-For example, the values -99 and -98 could be recoded to mean a broken machine sensor for the variable x:
+For example, the values -99 and -98 could be recoded to mean a broken machine sensor for the variable `x` with the following:
 
@@
 
-`naniar` does not reinvent the wheel for imputation, as there are many robust R packages for imputation \@ref(imputation).
-Instead, _nabular_ data is designed to work with existing imeplementation.
-`naniar` does provide a few imputation methods to facilitate exploration and visualization: `impute_mean`, `impute_median`, and `impute_below`.
+`naniar` does not reinvent the wheel for imputation, since instead working with existing methods.
+However, `naniar` provides a few imputation methods to facilitate exploration and visualization: `impute_below`, `impute_mean`, and `impute_median`.
 While useful to explore structure in missingness, they are not recommended for use in analysis.
-`impute_below` imputes values below the minimum value, with some jitter (small random noise) to reduce overplotting.
-It has options to change the default amount of shift, and jitter.
-
-```{r demonstrate-impute-below, eval = FALSE}
-airquality %>%
-  impute_below_at(vars(Ozone)) %>%
-  select(Ozone, Solar.R) %>%
-  head()
-```
+`impute_below` imputes values below the minimum value, with some controlable jitter (random noise) to reduce overplotting.
 
@@
 
-```{r impute-vector, eval = FALSE}
-
-impute_mean(airquality$Ozone) %>% head()
-
-impute_mean_at(airquality, .vars = vars(Ozone)) %>% head()
-
-impute_mean_if(airquality, .predicate = is.integer) %>% head()
-
-impute_mean_all(airquality) %>% head()
-
-```
-
 ### track: Shadow and impute missing values {#verbs-track}
 
-To evaluate imputations they need to be tracked, which is achieved by combining the verbs `nabular`, `impute_`, and `add_label_shadow`.
-The missing values can then be referred to by their shadow variable, `_NA`.
-The missingness of any observation can be referred to with `any_missing` (Figure \@ref(fig:track-impute-example)).
-The code chunk below shows the track pattern, first using `nabular`, then imputing with `impute_lm`, and adding a `label_shadow`:
+To evaluate imputations they need to be tracked.
+This is achieved by first using `nabular` data, then imputing, and imputed values can then be referred to by their shadow variable, `_NA` \@ref(fig:track-impute-example)).
+The code below shows the track pattern, first using `nabular`, then imputing with `impute_lm`.
+A label is added to facilitate identifying missings with `label_shadow`:
 
@@
 
-Missing values can be mapped to a graphical element in a scatterplot by setting the `color` aesthetic in ggplot to `any_missing` (Figure \@ref(fig:track-impute-example)A), or in a density plot using the `fill = any_missing`, (Figures \@ref(fig:track-impute-example)B) and \@ref(fig:track-impute-example)C).
-Imputed values can also be compared to complete case data, grouping by `any_missing`, and then summarizing, similar to other dplyr summary workflows (shown below).
-This shows mean and median values are similar, but the minimum and maximum values are different.
+Multiple missing or imputed values can be mapped to a graphical element in ggplot2 by setting the `color` or `fill` aesthetic in ggplot to `any_missing`, a result of the `add_label_shadow()` function.
+(Figure \@ref(fig:track-impute-example)A - C).
+Imputed values can also be compared to complete case data, grouping by `any_missing`, and then summarizing, similar to other dplyr summary workflows shown below in Table \@ref(tab:impute-summary-out), showing similarities and differences in imputation methods.
 
@@
 
-```{r impute-summary, echo = TRUE}
+```{r impute-summary, echo = TRUE, eval = FALSE}
 aq_imputed %>%
@@
                .funs = lst(min, mean, median, max)) 
+```
+
+```{r impute-summary-out, echo = FALSE}
+aq_imputed %>%
+  group_by(any_missing) %>%
+  summarise_at(.vars = vars(Ozone),
+               .funs = lst(min, mean, median, max))  %>% 
+  kable(caption = "Summarising values of imputed vs non imputed values.
+Comparing imputed values (denoted as 'Missing', since they were previously missing), the mean and median values are similar, but the minimum and maximum values are very different.")
 ```
@@
 # References
+
+```{r demonstrate-impute-below, eval = FALSE}
+airquality %>%
+  impute_below_at(vars(Ozone)) %>%
+  select(Ozone, Solar.R) %>%
+  head()
+```
+
+```{r impute-vector, eval = FALSE}
+
+impute_mean(airquality$Ozone) %>% head()
+
+impute_mean_at(airquality, .vars = vars(Ozone)) %>% head()
+
+impute_mean_if(airquality, .predicate = is.integer) %>% head()
+
+impute_mean_all(airquality) %>% head()
+
+```

### 2020-05-07 paper/tidy-missing-data-paper.Rmd
complete touching up univariate and bivariate

@@
 Missing values are often ignored when plotting data - so much so that the data visualisation software, MANET, is an acronym for "Missings Are Now Equally Treated" [@Unwin1996].
-However, plots can help to reveal the type of missing value patterns, even those of MCAR, MAR or even MNAR.
-Here we summarise how to systematically explore missing patterns visually, and define useful plots to make relative to the nabular data structure.
+However, plots can help to reveal the type of missing value patterns, even those of MCAR, MAR or MNAR.
+Here we summarise how to systematically explore missing patterns visually, and define useful plots to make, relative to the nabular data structure.
 
@@
 
-- How to visualise a binary matrix of missing vs not, and the dependencies between those binary variables as well.
-- These overview plots are looking at the shadow matrix, and the summaries of each of the variables
-- In this section we  describe which part of the nabular matrix we are going to display (overview - just binary parts), the summaries are summaries of the binary, the histogram and points are both the missings and the complete values.
-
 The  first step is to get an overview of the extent of the missingness.
-Overview visualisations for variables and cases are provided with `gg_miss_var` and `gg_miss_case` (Figure \@ref(fig:gg-miss-case-var)), drawing attention to the amount of missings, and ordering by missingness.
+Overview visualisations for variables and cases are provided with `gg_miss_var` and `gg_miss_case` (Figure \@ref(fig:gg-miss-case-var)A), drawing attention to the amount of missings, and ordering by missingness.
 The "airquality" dataset (available in base R), is shown, and contains daily air quality measurements in New York, from May to September, 1973.
-We learn in figure \@ref("fig:gg-miss-case-var")A-B, that two variables contain missings and approximately one third of observations have one missing value.
-A tiny number of observations are missing on two variables.
-These summary plots compute proportions from the shadow matrix.
-The numerical statistics can also be reported, and are discussed in the next section.
+We learn in figure \@ref(fig:gg-miss-case-var)A-B, that two variables contain missings, approximately one third of observations have one missing value, and a tiny number of observations are missing across two variables.
+These overview plots are created from the shadow matrix.
+Numerical statistics can also be reported, and are discussed in the next section.
 
@@
 This also provides numerical summaries of missingness in the legend, and for each column (Figure \@ref(fig:gg-miss-case-var)C-D).
-The user can also apply clustering to the rows, and arrange columns by missingness (Figure \@ref(fig:gg-miss-case-var)D).
-Similar visualizations are available in other missing data packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}.
-A key improvement is `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns in the visualization and are named at the top, and each row is an observation.
+Clustering can be applied to the rows, and columns arranged in order of missingness (Figure \@ref(fig:gg-miss-case-var)D).
+Similar visualizations are available in other packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}.
+A key improvement is `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns and are named at the top, and each row is an observation.
 Using \pkg{ggplot2} makes the plot easily customizable.
@@
 
-The number of times certain variables go missing together can be visualized using an "upset plot" [@Conway2017].
-An alternative to a Venn diagram, an upset plot shows the size and features of overlapping sets, but scales well with more variables.
-This can be applied to missing values, and an upset plot can be constructed from the shadow matrix, Figure \@ref(fig:airquality-upset) visualises overlapping amounts of missing data in the airquality data.
+The number of times observations are missing together is visualized using an "upset plot" [@Conway2017].
+An alternative to a Venn diagram, an upset plot shows the size and features of overlapping sets, scaling well with more variables.
+An upset plot can be constructed from the shadow matrix, with Figure  \@ref(fig:airquality-upset) visualising overlapping amounts missings in the airquality data.
 The bottom right shows the combinations of missingness, the top panel shows the size of these combinations, and the bottom left shows missingness in each variable.
-This shows us clearly that Ozone has the most missing values, and there are 2 cases where values in both Solar.R and Ozone are missing together, illustrating similar points to \@ref(fig:gg-miss-case-var).
+This provides similar information to Figure \@ref(fig:gg-miss-case-var)A-D, but more clearly illustrating overlapping missingness, where 2 cases are missing together in variables Solar.R and Ozone.
 
@@
 
+<!-- note: Di can you check these last two paragraphs -->
+
 **Imputing values below the range**.
-To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed below the range of data using `impute_below` (by default imputing 10% below the range).
-Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone, the number of missing values on its left.
-Ozone is shown in a different color by identifying the imputed values in _nabular_ data, by referring to the variable `Ozone` as `Ozone_NA` in a fill aesthetic in ggplot.
-The addition of the counts of missings in the histogram allows us to understand there are many missings, and that we have a skewed distribution of Ozone.
+To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed below the range of data using `impute_below`.
+Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone values on the right in green, and the  count of missing ozone values on the left, in orange.
+The missings in Ozone are mapped to the fill aesthetic in ggplot by referring to the variable `Ozone` as `Ozone_NA`, the values imputed below the range.
+Nabular data facilitates adding counts of missingness to a histogram, allowing not only understanding of a variables' distribution of values, but also understanding of the magnitude of missings.
+
+**Univariate split by missingness**.
+Two distributions of a variable can be shown according to the missingness of another variable.
+Figure \@ref(fig:impute-shift-histogram) shows the values of temperature when ozone is present, and missing.
+Figure \@ref(fig:impute-shift-histogram)A is a faceted histogram, and Figure \@ref(fig:impute-shift-histogram)B is an overlaid density.
+This shows how values of temperature are affected by the missingness of ozone, and reveals a cluster of low temperature observations with missing ozone values.
+The nabular structure of the data helps reveals the distribution of the missing values.
+This can facilitate exploring missing data distributions.
+For example, we would expect if data were MCAR, for values to be roughly uniformly missing thoughout the histogram or density.
 
@@
 
-**Univariate split by missingness**.
-Missingness of one variable can also be used to display different distributions in another.
-Figure \@ref(fig:bind-shadow-density) shows the values of temperature when ozone is present, and missing.
-Figure \@ref(fig:bind-shadow-density)A is a faceted histogram, and Figure \@ref(fig:bind-shadow-density)B is an overlaid density.
-This shows how values of temperature are affected by the missingness of ozone, and reveals a cluster of low temperature observations with missing ozone values.
-The nabular structure of the data reveals the distribution of the missing values
-
-```
-To know where these values might have been located.
-tying this back to the nabular structure, and also the terminology of MCAR MAR, etc.
-(We would expect if data were MCAR for these points to be representatively thoughout the histogram)
-
-- be sure to clear up vague language (it is important to...etc., )
-- Missings are often ignored
-```
-
 ## Bivariate {#bivariate}
@@
 To visualize missing values in two dimensions the missing values can be placed in plot margins, by imputing values below the range of the data.
-Using _nabular_ data identifies imputed values (\@ref(univariate-graphics)), and color makes missingness pre-attentive [@treisman1985].
+Using _nabular_ data identifies imputed values, and color makes missingness pre-attentive [@treisman1985].
 The steps of imputing and coloring are combined into `geom_miss_point()`.
@@
 As `geom_miss_point()` is a defined \pkg{ggplot2} geometry, it works with features such as facetting and mapping other variables to graphical aesthetics.
-The shadow format by itself is also useful for exploring patterns of missingness.
 

### 2020-05-07 paper/tidy-missing-data-paper.Rmd
finish graphics section

@@
 
-```{r airquality-upset, fig.height = 3, fig.width = 4, fig.cap = "The pattern of missingness in the airquality dataset shown in an upset plot.
+```{r airquality-upset, fig.height = 2, fig.width = 3, fig.cap = "The pattern of missingness in the airquality dataset shown in an upset plot.
 Only Ozone and Solar.R have missing values, and Ozone has the most missing values.
@@
 They transform variables to the same scale, ranging between 0 and 1.
-The `oceanbuoys` dataset from \pkg{naniar} is used for this visualisation, containing measurements of ocean buoys to understand and predict El Niño and El Niña.
-Data was collected in 1993 and 1997, and contains measurements of sea and air temperature, humidity, and east west and north south wind speeds.
-Figure \@ref(fig:parallel-cord-plot) shows a parallel coordinate plot of the `oceanbuoys`, with missing values imputed to be 10% below the range, and values colored according to where humidity was missing (humidity_NA).
-This shows humidity is missing at low air and sea temperatures, and humidity is missing in one year, and one location.
+The `oceanbuoys` dataset from \pkg{naniar} is used for this visualisation, containing measurements of sea and air temperature, humidity, and east west and north south wind speeds.
+Data was collected in 1993 and 1997, to understand and predict El Niño and El Niña.
+Figure \@ref(fig:parallel-cord-plot) is a parallel coordinate plot of `oceanbuoys`, with missing values imputed to be 10% below the range, and values colored according to whether humidity was missing (`humidity_NA`).
+Figure \@ref(fig:parallel-cord-plot) shows humidity is missing at low air and sea temperatures, and humidity is missing in one year, and one location.
 

### 2020-05-07 paper/tidy-missing-data-paper.Rmd
section 6 completed

@@
 
-This section describes approaches to summarizing missingness.
-The functions are implemented in \pkg{naniar}, and are easy to remember as they have consistent names and output, returning either a single number \@ref(single-num-sum), or a dataframe (\@ref(sum-tab-missings)), integrating well with plotting and modelling tools.
+This section describes approaches to summarizing missingness, and an implementation in \pkg{naniar}.
+Numerical summaries should be easy to remember with consistent names and output, returning either a single number \@ref(single-num-sum), or a dataframe \@ref(sum-tab-missings), so they integrate well with plotting and modelling tools.
 How these work with other tools in an analysis pipeline is shown in \@ref(num-sum-w-group).
@@
 
-The overall number, proportion, or percent of missing values in a dataset is shown with `n_miss`, `prop_miss` and `pct_miss`, and their complete value complements: `n_complete`, `prop_complete`, and `pct_complete`.
-Summaries for variables and cases are made by appending `_case` or `_var` to any of the summaries.
+The overall number, proportion, or percent of missing values in a dataset should be simple to calculate.
+`naniar` provides the functions `n_miss`, `prop_miss` and `pct_miss`, as well as their complements.
+Summaries for variables and cases are made by appending `_case` or `_var` to these summaries.
 An overview is shown in Table \@ref(tab:n-prop-pct-miss-complete).
@@
 
-Presenting the number and percent of missing values for each variable, or case, provides a summary usable in models to inform imputations or data handling.
-For example, potentially dropping variables or deciding to include others in an imputation model.
-Another useful approach is to tabulate the frequency of missing values for each variable or case; that is, the number of times there are zero, one, two, and so on, missing values.
-These summaries and tabulations are shown for variable in Tables \@ref(tab:miss-var-summary) and \@ref(tab:miss-var-table), using `miss_var_summary` and `miss_var_table`.
-Functions are also provided for case-wise summaries and tabulations: `miss_case_summary` and `miss_case_table`, which order rows by `n_miss`, to show the most missings at the top.
-`naniar` also shows missingness summaries across a repeating span, `miss_var_span`, and finding streaks or runs of missingness in a given variable in `miss_var_run`.
+Presenting the number and percent of missing values for each variable, or case, provides a summary usable in data handling, or in models to inform imputation.
+For example, potentially dropping variables, or deciding to include others in an imputation model.
+Another useful approach is to tabulate the frequency of missing values for each variable or case; that is, the number of times there are zero missings, one missing, two, and so on.
+These summaries and tabulations are shown for variables in Tables \@ref(tab:miss-var-summary) and \@ref(tab:miss-var-table), and implemented with `miss_var_summary` and `miss_var_table`.
+Case-wise (row-wise) summaries and tabulations are implemented with `miss_case_summary` and `miss_case_table`.
+
+These summaries order rows by the number of missings (`n_miss`), to show the most missings at the top.
+The number of missings across a repeating span, or finding "runs" or "streaks" of missings in given variables can also be useful for identifying missingness patterns, and are implemented with `miss_var_span`, and `miss_var_run`.
 
@@
 
-These summaries and tabulations can be composed with the "pipe" operator to perform grouped summaries with `dplyr`'s `group_by` function.
-These summaries return dataframes, easily used in other plotting devices, modelling tools, or other summaries.
-Table \@ref(tab:group-miss-var-summary) shows an example of missing data summaries for airquality by each month.
+It is useful to explore summaries and tabulations within groups of a dataset.
+`naniar` works with `dplyr`'s `group_by` operator to produce grouped summaries, which work well with the "pipe" operator.
+The code and table below \@ref(tab:group-miss-var-summary) show an example of missing data summaries for airquality, grouped by month.
 
@@
 
-To illustrate the methods, data on housing for the city of Melbourne from January 28, 2016 to March 17, 2018 is used.
+This section shows how the methods described so far are used together in a data analysis workflow.
+These are applied to a case study, of housing data for the city of Melbourne from January 28, 2016 to March 17, 2018.
 The data was compiled by scraping weekly property clearance data [@Kaggle-2018-data].
@@
 As a precursor to building a predictive model, this analysis focusses on understanding the patterns of missingness.
-This section shows how the methods from previous sections are used together in a data analysis workflow.
 

### 2020-05-07 paper/tidy-missing-data-paper.Rmd
done proofing the graphics section

--- 2020-05-07 paper/tidy-missing-data-paper.Rmd
@@
 
-Missing values are often ignored when plotting data - so much so that the data visualisation software, MANET, is an acronym for "Missings Are Now Equally Treated" [@Unwin1996].
+Missing values are often ignored when plotting data - which is why the data visualisation software, MANET, was named using an acronym for "Missings Are Now Equally Treated" [@Unwin1996].
 However, plots can help to reveal the type of missing value patterns, even those of MCAR, MAR or MNAR.
@@
 
-The shadow matrix allows both the variables and cases to be displayed using a heatmap style visualization, `vis_miss()` from \pkg{visdat} [@visdat].
+The shadow matrix allows both the variables and cases to be displayed using a heatmap style visualization, with `vis_miss()` from \pkg{visdat} [@visdat].
 This also provides numerical summaries of missingness in the legend, and for each column (Figure \@ref(fig:gg-miss-case-var)C-D).
@@
 A key improvement is `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns and are named at the top, and each row is an observation.
-Using \pkg{ggplot2} makes the plot easily customizable.
+Using \pkg{ggplot2}, as the foundation, makes the plot easily customizable.
 
@@
 
-The number of times observations are missing together is visualized using an "upset plot" [@Conway2017].
-An alternative to a Venn diagram, an upset plot shows the size and features of overlapping sets, scaling well with more variables.
-An upset plot can be constructed from the shadow matrix, with Figure  \@ref(fig:airquality-upset) visualising overlapping amounts missings in the airquality data.
+The number of times observations are missing together can be visualized using an "upset plot" [@Conway2017].
+An alternative to a Venn diagram, an upset plot shows the size and features of overlapping sets, and scales well with more variables.
+An upset plot can be constructed from the shadow matrix, as shown in Figure  \@ref(fig:airquality-upset) which displays the overlapping counts of  missings in the airquality data.
 The bottom right shows the combinations of missingness, the top panel shows the size of these combinations, and the bottom left shows missingness in each variable.
@@
 
-Missing values are by default not shown for univariate visualizations such as histograms or densities.
+Missing values are typically not shown for univariate visualizations such as histograms or densities.
 Two ways to use _nabular_ data to present univariate data with missings are discussed.
@@
 **Imputing values below the range**.
-To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed below the range of data using `impute_below`.
-Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone values on the right in green, and the  count of missing ozone values on the left, in orange.
-The missings in Ozone are mapped to the fill aesthetic in ggplot by referring to the variable `Ozone` as `Ozone_NA`, the values imputed below the range.
-Nabular data facilitates adding counts of missingness to a histogram, allowing not only understanding of a variables' distribution of values, but also understanding of the magnitude of missings.
+To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed below the range of data using `impute_below_all`.
+Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone values on the right in green, and the histogram of missing ozone values on the left, in orange.
+The missings in Ozone are imputed and `Ozone_NA` is  mapped to the fill aesthetic.
+Nabular data facilitates adding counts of missingness to a histogram, allowing examination of a variables' distribution of values, and also the magnitude of missings.
 
 **Univariate split by missingness**.
-Two distributions of a variable can be shown according to the missingness of another variable.
-Figure \@ref(fig:impute-shift-histogram) shows the values of temperature when ozone is present, and missing.
-Figure \@ref(fig:impute-shift-histogram)A is a faceted histogram, and Figure \@ref(fig:impute-shift-histogram)B is an overlaid density.
+The distribution of a variable can be shown according to the missingness of another variable.
+The shadow matrix part of nabular is used to handle the faceting, and color mapping.
+Figure \@ref(fig:impute-shift-histogram) shows the values of temperature when ozone is present, and missing, a faceted histogram (B), and  an overlaid density (C).
 This shows how values of temperature are affected by the missingness of ozone, and reveals a cluster of low temperature observations with missing ozone values.
-The nabular structure of the data helps reveals the distribution of the missing values.
-This can facilitate exploring missing data distributions.
+This type of plot can facilitate exploring missing data distributions.
 For example, we would expect if data were MCAR, for values to be roughly uniformly missing thoughout the histogram or density.
@@
 
-Parallel coordinate plots can visualize missingness beyond two dimensions.
+Parallel coordinate plots can help to visualize missingness beyond two dimensions.
 They transform variables to the same scale, ranging between 0 and 1.

### 2020-05-07 paper/tidy-missing-data-paper.Rmd
fixed typo
fixed grammar

@@
 The shadow matrix part of nabular is used to handle the faceting, and color mapping.
-Figure \@ref(fig:impute-shift-histogram) shows the values of temperature when ozone is present, and missing, a faceted histogram (B), and  an overlaid density (C).
+Figure \@ref(fig:impute-shift-histogram) shows the values of temperature when ozone is present, and missing, using a faceted histogram (B), and  an overlaid density (C).
 This shows how values of temperature are affected by the missingness of ozone, and reveals a cluster of low temperature observations with missing ozone values.
@@
 The methodology described in this paper can be used in conjunction with other approaches to understand multivariate missingness dependencies (e.g.
-decision trees [@Tierney2015], latent group analysis [@Barnett2017], and PCA [@FactoMineR]) .
+decision trees [@Tierney2015], latent group analysis [@Barnett2017], and PCA [@FactoMineR]).
 Evaluating imputed values using a testing framework like @VanBuuren2012 is also supported.

### 2020-05-07 paper/tidy-missing-data-paper.Rmd
section 7 completed

--- 2020-05-07 paper/tidy-missing-data-paper.Rmd
@@
 
-Missing values are often ignored when plotting data - which is why the data visualisation software, MANET, was named using an acronym for "Missings Are Now Equally Treated" [@Unwin1996].
+Missing values are often ignored when plotting data - so much so that the data visualisation software, MANET, is an acronym for "Missings Are Now Equally Treated" [@Unwin1996].
 However, plots can help to reveal the type of missing value patterns, even those of MCAR, MAR or MNAR.
@@
 
-The shadow matrix allows both the variables and cases to be displayed using a heatmap style visualization, with `vis_miss()` from \pkg{visdat} [@visdat].
+The shadow matrix allows both the variables and cases to be displayed using a heatmap style visualization, `vis_miss()` from \pkg{visdat} [@visdat].
 This also provides numerical summaries of missingness in the legend, and for each column (Figure \@ref(fig:gg-miss-case-var)C-D).
@@
 A key improvement is `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns and are named at the top, and each row is an observation.
-Using \pkg{ggplot2}, as the foundation, makes the plot easily customizable.
+Using \pkg{ggplot2} makes the plot easily customizable.
 
@@
 
-The number of times observations are missing together can be visualized using an "upset plot" [@Conway2017].
-An alternative to a Venn diagram, an upset plot shows the size and features of overlapping sets, and scales well with more variables.
-An upset plot can be constructed from the shadow matrix, as shown in Figure  \@ref(fig:airquality-upset) which displays the overlapping counts of  missings in the airquality data.
+The number of times observations are missing together is visualized using an "upset plot" [@Conway2017].
+An alternative to a Venn diagram, an upset plot shows the size and features of overlapping sets, scaling well with more variables.
+An upset plot can be constructed from the shadow matrix, with Figure  \@ref(fig:airquality-upset) visualising overlapping amounts missings in the airquality data.
 The bottom right shows the combinations of missingness, the top panel shows the size of these combinations, and the bottom left shows missingness in each variable.
@@
 
-Missing values are typically not shown for univariate visualizations such as histograms or densities.
+Missing values are by default not shown for univariate visualizations such as histograms or densities.
 Two ways to use _nabular_ data to present univariate data with missings are discussed.
@@
 **Imputing values below the range**.
-To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed below the range of data using `impute_below_all`.
-Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone values on the right in green, and the histogram of missing ozone values on the left, in orange.
-The missings in Ozone are imputed and `Ozone_NA` is  mapped to the fill aesthetic.
-Nabular data facilitates adding counts of missingness to a histogram, allowing examination of a variables' distribution of values, and also the magnitude of missings.
+To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed below the range of data using `impute_below`.
+Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone values on the right in green, and the  count of missing ozone values on the left, in orange.
+The missings in Ozone are mapped to the fill aesthetic in ggplot by referring to the variable `Ozone` as `Ozone_NA`, the values imputed below the range.
+Nabular data facilitates adding counts of missingness to a histogram, allowing not only understanding of a variables' distribution of values, but also understanding of the magnitude of missings.
 
 **Univariate split by missingness**.
-The distribution of a variable can be shown according to the missingness of another variable.
-The shadow matrix part of nabular is used to handle the faceting, and color mapping.
-Figure \@ref(fig:impute-shift-histogram) shows the values of temperature when ozone is present, and missing, a faceted histogram (B), and  an overlaid density (C).
+Two distributions of a variable can be shown according to the missingness of another variable.
+Figure \@ref(fig:impute-shift-histogram) shows the values of temperature when ozone is present, and missing.
+Figure \@ref(fig:impute-shift-histogram)A is a faceted histogram, and Figure \@ref(fig:impute-shift-histogram)B is an overlaid density.
 This shows how values of temperature are affected by the missingness of ozone, and reveals a cluster of low temperature observations with missing ozone values.
-This type of plot can facilitate exploring missing data distributions.
+The nabular structure of the data helps reveals the distribution of the missing values.
+This can facilitate exploring missing data distributions.
 For example, we would expect if data were MCAR, for values to be roughly uniformly missing thoughout the histogram or density.
@@
 
-Parallel coordinate plots can help to visualize missingness beyond two dimensions.
+Parallel coordinate plots can visualize missingness beyond two dimensions.
 They transform variables to the same scale, ranging between 0 and 1.
@@
 This section shows how the methods described so far are used together in a data analysis workflow.
-These are applied to a case study, of housing data for the city of Melbourne from January 28, 2016 to March 17, 2018.
+We analyse a case study of housing data for the city of Melbourne from January 28, 2016 to March 17, 2018.
 The data was compiled by scraping weekly property clearance data [@Kaggle-2018-data].
@@
 Figure \@ref(fig:housing-miss-case-var)A shows 9 variables with missing values.
-The most missings are in building area, followed by year built, and land size, with similar amounts of missingness in Car, bathroom, bedroom2, longitude, and latitude.
-Figure \@ref(fig:housing-miss-case-var)B reveals there are up to 50% missing values in cases, and the majority of cases have more than 5% values missing.
-The variables building area and year built have more than 50% missing data, and so could perhaps be omitted from subsequent analysis, as imputed values are likely to be spurious.
+The most missings are in "building area", followed by "year built", and "land size", with similar amounts of missingness in "Car", "bathroom", "bedroom2", "longitude", and "latitude." Figure \@ref(fig:housing-miss-case-var)B reveals there are up to 50% missing values in cases, and the majority of cases have more than 5% values missing.
+The variables "building area" and "year built" have more than 50% missing data, and so could perhaps be omitted from subsequent analysis, as imputed values are likely to be spurious.
 Three missingness clusters are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \@ref(fig:applic-vis-miss)).
@@
 
-Figure \@ref(fig:applic-vis-miss) shows missingness patterns with an `upset` plot [@Conway2017], displaying 8 intersecting sets of missing variables.
+Figure \@ref(fig:housing-upset) shows missingness patterns with an `upset` plot [@Conway2017], displaying 8 intersecting sets of missing variables.
 Two patterns stand out: two, and five variables missing, providing further evidence of the missingness patterns seen in Figures \@ref(fig:housing-miss-case-var) and \@ref(fig:housing-upset).
@@
 Tabulating missings in cases (Table \@ref(tab:housing-miss-var-case-table) (right) shows six patterns of missingness.
-These overview plots lead to the removal of two variables from with more than 50% missingness from analysis: building area and year built.
+These overview plots lead to the removal of two variables from with more than 50% missingness from analysis: "building area" and "year built".
 
@@
 
-Using information from \@ref(case-study-explore-pattern), the following variables are explored for features predicting missingness: Land size, latitude, longitude, bedroom2, bathroom, car, and land size.
+Using information from \@ref(case-study-explore-pattern), the following variables are explored for features predicting missingness: "Land size", "latitude", "longitude", "bedroom2", "bathroom", "car", and "land size".
 
 Missingness structure is explored by clustering the missing values into groups.
-Then, a  classification and regression trees (CART) model is applied to predict these missingness clusters using the remaining variables, to find variables and values that predicting missingness clusters [@Tierney2015; @Barnett2017].
+Then, a  classification and regression trees (CART) model is applied to predict these missingness clusters using the remaining variables [@Tierney2015; @Barnett2017].
 Two clusters of missingness are identified and predicted using all variables in the dataset with the CART package `rpart` [@rpart], and plotted using the `rpart.plot` package [@rpart-plot].
-Importance scores reveal variables important for predicting missingness: rooms, price, suburb, council area, distance, and region name.
-These variables are important for predicting missingness, so are important to include in the imputation model.
+Importance scores reveal the following variables as most important in predicting missingness: "rooms", "price", "suburb", "council area", "distance", and "region name".
+These variables are important for predicting missingness, so are included in the imputation model.
 
@@
 
-\pkg{simputation} is used to implement two imputation: simple linear regression and K nearest neighbors.
+\pkg{simputation} is used to implement two imputations: simple linear regression and K nearest neighbors.
 Values are imputed stepwise in ascending order of missingness.
@@
 
-## Assess model predictions {#case-study-assess-model}
+## Assessing model predictions {#case-study-assess-model}
 
@@
 The `naniar` and `visdat` packages implement the methods discussed in the paper, building on existing tidy tools and strike a compromise between automation and control, making analysis efficient, readable, but not overly complex.
-Each tool has clear intent and effects - plotting or generating data or augmenting data in some way.
-This reduces repetition and typing for the user, making exploration of missing values easier as they follow consistent, predictable rules.
+Each tool has clear intent and effects - summarising, plotting or generating data or augmenting data in some way.
+This not only reduces repetition and typing in an analysis, but most importantly, allows for clear expression of intent, making exploration of missing values fluent.
 

### 2020-05-08 paper/tidy-missing-data-paper.Rmd
touching up final section
addressing most of the notes

@@
 Visualizations can help assess whether data may be MCAR, MAR or MNAR.
-Imputation is recommended in most cases of missingness dependence.
-
-<!-- (NOTE: Are the above last two sentences OK?) -->
+
+<!-- (NOTE: Is the last sentence OK?) -->
 <!-- (NOTE: trying to come up with a sentence that leads from this section to the next) -->
@@
 
-<!-- Note: I'm not sure why I talk about this on the fly stuff here, and I wonder if this paragraph below could be removed? -->
-
-The _nabular_ data structure provides three key benefits over creating a missingness matrix, $B$, on-the-fly.
+<!-- Note: Do you think this paragraph needs to be here, about the use of _nabular_ data structure? -->
+
+Although a binary missingness matrix, $B$, could be generated during analysis instead of using a _nabular_ data structure, there are key advantages to using _nabular_ data.
 Firstly, missing value labels `NA` and `!NA` are clearer than TRUE or FALSE.
@@
 
-<!-- Note: I'm not sure where I should put these three parts below parts in the paper: -->
-
-<!-- - Other statistical programs (SPSS, SAS, and Stata) represent missingness values as a full-stop, `.`, and allow for recording special missing values as `.a` - `.z`.
-SAS, SPSS, and STATA all implement.
--->
-
-<!-- - The special values from these languages break the rule of "keeping one thing in a cell", as they record both the value and the multivariate missingness state.
-Additionally, special missing values in other statistical programming languages are encoded at a lower level in the fundamental `NA` type, which is not possible in R.
--->
-<!-- - (in the same column you have mixed information, which breaks the tidy principle of one column having one type of value) -->
+Special missing values are not a new idea, and have been implemented in other statistical programming languages, SPSS, SAS, and STATA.
+These typically represent missing values as a full-stop, `.`, and record  special missing values as `.a` - `.z`.
+These special values from these languages break the tidy principle of one column having one type of value, as they record both the value, and the multivariate missingness state.
 
@@
 The second displays two plots of the same variable according to the missingness of a chosen variable.
-
-<!-- note: Di can you check these last two paragraphs -->
 
@@
 
+<!-- Note: I think I'm basically there, but I'd like to revisit the first paragraph in this section and the last sentence -->
+
 This paper has described new methods for exploring, visualizing, and imputing missing data.
@@
 We would also like to thank those who contributed pull requests and discussions on the `naniar` package, in particular Jim Hester and Romain François for improving the speed of key functions, Ross Gayler for discussion on special missing values, and Luke Smith for helping `naniar` be more compliant with `ggplot2`.
-We would also like to thank Amelia McNamara for her assistance with proof-reading the paper.
+We would also like to thank Amelia McNamara for discussions on the paper.
 

### 2020-05-13 paper/tidy-missing-data-paper.Rmd
Use new and improved nabular figure

@@
 
-```{r shadow-matrix-progress, echo = FALSE, fig.cap = "Series of steps to create shadow matrix data.
-(1-2) Data to a binary representation of missings, (2-3) Binary format converted to a shadow matrix (3-4) New and improved shadow matrix with changed variable names.
-This clearly links a variable to its state of missingness.", out.width = "100%", fig.align="center"}
-
-knitr::include_graphics("images/full-conversion-to-shadow.png")
-
-```
-
 We propose a new form for tidy representation of missing data based on these ideas from past research.
-Four features are added to the shadow matrix to facilitate analysis of missing data, illustrated in Figure \@ref(fig:shadow-matrix-progress).
+Four features are added to the shadow matrix to facilitate analysis of missing data, illustrated in Figure \@ref(fig:nabularfig).
 
@@
 We propose "NA" and "!NA".
-(Figure \@ref(fig:shadow-matrix-progress) parts 1-3).
+(Figure \@ref(fig:nabularfig)).
 This improves the 0 and 1 values in $B$, which do not clearly identify missingness.
@@
 **Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA_\<suffix\>" - for example: `NA_instrument` uses a short label, "instrument", indicating instrument error resulting in missing values.
-These could be also used to indicate imputations, (Table \@ref(tab:shadow-encoding)) (note: the part of the new figure).
+These could be also used to indicate imputations, (Figure \@ref(fig:nabularfig)).
 
 1.
-**Coordinated names**: Variable names in the shadow matrix gain a consistent short suffix, "_NA", keeping names coordinated throughout analysis (Figure \@ref(fig:shadow-matrix-progress) part 4 - note: update this figure).
+**Coordinated names**: Variable names in the shadow matrix gain a consistent short suffix, "_NA", keeping names coordinated throughout analysis (Figure \@ref(fig:nabularfig)).
 It makes a clear distinction with `var_NA` being a random variable of the missingness of a variable, `var`.
@@
 It is useful for visualization, summaries, and tracking imputed values, discussed in more detail in \@ref(nabular-data).
-
-<!-- Note: I think that Figure \@ref(fig:shadow-matrix-progress), Table \@ref(tab:shadow-encoding), and Figure \@ref(fig:nabularfig) could be combined into one useful graphic.
-This could form the "key graphic" of the paper.
-I also think that the point about the shadow value (`var_NA`) being a factor and having an underlying value is not directly relevant to the paper.
--->
-
-```{r shadow-encoding}
-
-fig_shadow_code <- tibble::tibble(temp = c(-99, NA, -1, 106),
-                                  temp_NA = c("NA_instr",NA, "NA_dropout","!NA"),
-                                  NA_value = c(2, 1, 3, 0))
-
-knitr::kable(fig_shadow_code,
-             caption = "Example data of temperature, its shadow representation, and underlying NA value.
-The temperature value of -99 is represented in the shadow column as 'NA\\_instr', which in turn has the underlying numeric value of 2.
-This captures additional information about the data otherwise difficult to record.")
-             # booktabs = TRUE) 
-  # kable_styling(latex_options = c("hold_position"))
-
-```
 
@@
 ```{r nabularfig, echo = FALSE, fig.cap = "The process of creating nabular data.
-Data transformed to shadow matrix, and nabular data contains the shadow matrix, column bound to the data.
-Nabular data can be created using `bind\\_shadow` or `nabular()` functions.
-Nabular data provides a useful format for missing data exploration and analysis.", out.width = "100%", fig.align = "center"}
-
-knitr::include_graphics("images/nabular.png")
+Data transformed to shadow matrix, where values are either not missing or missing: '!NA' or 'NA'.
+The shadow matrix can be converted to long form to create missingness summary plots.
+Nabular data is created by binding the columns od the data and shadow matrix.
+Special missing values (such as -99) are identified as special missings, and values imputed and tracked.
+Nabular data can be used to identify imputations and explore data values alongside missings, providing a useful format for missing data exploration and analysis.", out.width = "100%", fig.align = "center"}
+
+knitr::include_graphics("images/diagram.png")
 
@@
 Multiple missing or imputed values can be mapped to a graphical element in ggplot2 by setting the `color` or `fill` aesthetic in ggplot to `any_missing`, a result of the `add_label_shadow()` function.
-(Figure \@ref(fig:track-impute-example)A - C).
+(Figure \@ref(fig:track-impute-example)).
 Imputed values can also be compared to complete case data, grouping by `any_missing`, and then summarizing, similar to other dplyr summary workflows shown below in Table \@ref(tab:impute-summary-out), showing similarities and differences in imputation methods.
@@
 Overview visualisations for variables and cases are provided with `gg_miss_var` and `gg_miss_case` (Figure \@ref(fig:gg-miss-case-var)A), drawing attention to the amount of missings, and ordering by missingness.
-The "airquality" dataset (available in base R), is shown, and contains daily air quality measurements in New York, from May to September, 1973.
-We learn in figure \@ref(fig:gg-miss-case-var)A-B, that two variables contain missings, approximately one third of observations have one missing value, and a tiny number of observations are missing across two variables.
-These overview plots are created from the shadow matrix.
+The "airquality" dataset (from base R), is shown, and contains daily air quality measurements in New York, from May to September, 1973.
+We learn from figure \@ref(fig:gg-miss-case-var)A-B, that two variables contain missings, approximately one third of observations have one missing value, and a tiny number of observations are missing across two variables.
+These overview plots are created from the shadow matrix in long form (Figure \@ref(fig:nabularfig).
 Numerical statistics can also be reported, and are discussed in the next section.
@@
 The default output (C) and ordered by clustering on rows and columns (D).
+These overviews are made possible using the shadow matrix in long form.
 There are only missings in ozone and solar radiation, and there appears to be some structure to their missingness.", fig.height = 7, fig.width = 6, out.width = "90%", fig.align="center"}
@@
   scale_fill_brewer(palette = "Dark2") + 
-  theme_minimal() +
   labs(tag = "A") + 

### 2020-05-14 cover-letter/cover-letter.Rmd
csda cover letter
update cover letter

@@
 fromqualifications: BPsySc (Hons), PhD (Statistics)
-fromtitle: Research Fellow
+fromtitle: Lecturer
 fromWWW: njtierney.com
@@
 fromsignature: sigfile.png
-toname: Dr. B.U. Park
-toaddress: CSDA
-opening: Dear Dr. D.U. Park
-closing: Regards, \newline Dr. Nicholas Tierney \newline Reseach Fellow
+toname: Editor in Chief
+toaddress: Journal of Statistical Software
+opening: Dear Editors-in-Chief
+closing: Regards, \newline Dr. Nicholas Tierney \newline Lecturer
 spacing: 1.4
@@
 
-Please consider my article "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations" for possible publication in CSDA.
+Please consider my article "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations" for possible publication in the Journal of Statistical Software.
 This article describes new work on handling missing values from a data centric approach.
 The methodology is implemented in the R package, `naniar`, which is available on CRAN, https://cran.r-project.org/web/packages/naniar, with the development version on github https://github.com/njtierney/naniar.
+`naniar` is now downloaded 14 thousand times a month, and has been downloaded over 160 thousand times.
+`naniar` has also seen wide praside, for example: [Karandeep Singh](https://twitter.com/kdpsinghlab/status/1110529278830170116):
+
+> Looking forward to teaching about missing data and mechanisms of missingness today.
+Prep work involved swapping out the mi and VIM packages from last year's #rstats code with @nj_tierney's excellent naniar package.
+
+And [Omar Wasow](https://twitter.com/owasow/status/1006743170846109697):
+
+> naniar also tipped the scale for me w/teaching.
+In past, I've skipped lecturing on missing data.
+This year, I revised lectures & covered naniar/simputation/Amelia.
+Was grateful to easily show basic issue as well as trade-offs of different forms of imputation (eg mean vs lm).
+Thx!
+
+In addition, the R package has been cited 9 times, and the pre-print of this paper has been cited 6 times since 2018.
 
@@
 This article is not under consideration elsewhere.
+This paper was submitted two years aho to the Journal of Computational and Graphical Statistics, but was rejected.
+It has now been substantially re-written.
 We have submitted a pre-print of this article to arxiv, and to our internal department pre-print at Monash University.
 
-We hope that you find this paper of interest to the readers of CSDA,
+We hope that you find this paper of interest to the readers of Journal of Statistiacl Software,

### 2020-05-14 cover-letter/cover-letter.Rmd
cover letter

@@
 fromsignature: sigfile.png
-toname: Editor in Chief
+toname: Editors-in-Chief
 toaddress: Journal of Statistical Software
@@
 
-Please consider my article "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations" for possible publication in the Journal of Statistical Software.
-This article describes new work on handling missing values from a data centric approach.
-The methodology is implemented in the R package, `naniar`, which is available on CRAN, https://cran.r-project.org/web/packages/naniar, with the development version on github https://github.com/njtierney/naniar.
-`naniar` is now downloaded 14 thousand times a month, and has been downloaded over 160 thousand times.
-`naniar` has also seen wide praside, for example: [Karandeep Singh](https://twitter.com/kdpsinghlab/status/1110529278830170116):
+Please consider my article "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations" for publication in the Journal of Statistical Software.
+This article describes new work on handling missing values using a tidy approach.
+The methods are implemented in the R package, `naniar`, which is available on CRAN (https://cran.r-project.org/web/packages/naniar), and the development version is available on github https://github.com/njtierney/naniar.
+The package, `naniar` is being downloaded at a volume of 14,000 times a month, with total lifetime downloads 160,000 times, as measured from the RStudio server using the `cranlogs` package.
+The work has very visible, and widely praised, for example: [Karandeep Singh](https://twitter.com/kdpsinghlab/status/1110529278830170116) says:
 
@@
 This article is not under consideration elsewhere.
-This paper was submitted two years aho to the Journal of Computational and Graphical Statistics, but was rejected.
-It has now been substantially re-written.
-We have submitted a pre-print of this article to arxiv, and to our internal department pre-print at Monash University.
+This paper was submitted two years ago to the Journal of Computational and Graphical Statistics, but was rejected.
+It has now been substantially re-written, based on reviewer feedback.
+We have submitted a pre-print of this article to arxiv, and to our internal department pre-print series at Monash University.
 
-We hope that you find this paper of interest to the readers of Journal of Statistiacl Software,
+We hope that you find this paper of interest to the readers of Journal of Statistical Software,

### 2020-05-14 paper/tidy-missing-data-paper.Rmd
fix some minor typos

@@
 For example, estimated coefficients, predictions, and residuals from one model cannot be easily combined with the output of another model.
-Messy models have been partially addressed with the \pkg{broom} package, which tidies up model outputs into a tidy data format for use data analysis, and the developing \pkg{recipes} package, which helps make modelling input- and output-tidy [@recipes].
+Messy models have been partially addressed with the \pkg{broom} package, which tidies up model outputs into a tidy data format for data analysis, and the developing \pkg{recipes} package, which helps make modelling input- and output-tidy [@recipes].
 
@@
 These are constructed to share similar principles in their design and behavior, and cover the breadth of an analysis - from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @tidyverse; @Tidyverse-Manifesto].
-This has led to more tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr} and dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} [@readr; @readxl; @haven; @stringr; @lubridate; @purrr].
+This has led to more tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr}, dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} [@readr; @readxl; @haven; @stringr; @lubridate; @purrr].
 It has also led to a burgeoning of new packages for other fields following similar design principles, creating fluid workflows for new domains.
@@
 
-Each imputation method provides practical methods for different usecases, but most have different output structures, and do not have consistent interfaces in their implementation.
+Each imputation method provides practical methods for different use cases, but most have different output structures, and do not have consistent interfaces in their implementation.
 This makes them inherently messy and challenging to integrate into an analysis pipeline.
@@
 
-The missingness matrix $B$ can be used to assess missing data dependence It has been used in interactive graphics, dubbed a "shadow matrix", to link missing and imputed values to the data, facilitating their display [@Swayne1998], focusing heavily on multivariate numeric data.
+The missingness matrix $B$ can be used to assess missing data dependence.
+It has been used in interactive graphics, dubbed a "shadow matrix", to link missing and imputed values to the data, facilitating their display [@Swayne1998], focusing heavily on multivariate numeric data.
 This is an idea upon which this new work builds.
@@
 \pkg{ggplot2} removes missing values with a warning (Figure \@ref(fig:warning)), and only incorporates missingness into visualizations when mapping a discrete variable with missings to a graph aesthetic.
-This has some limitations, shown in Figure \@ref(fig:gg-box-na), a boxplot visualization of school grade and test score mapped to the x- and y-axis .
+This has some limitations, shown in Figure \@ref(fig:gg-box-na), a boxplot visualization of school grade and test scores.
 If there are missings in a continuous variable like test score, \pkg{ggplot2} omits the missings and prints a warning message.
@@
 1.
-**Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA_\<suffix\>" - for example: `NA_instrument` uses a short label, "instrument", indicating instrument error resulting in missing values.
+**Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA_\<suffix\>".
+For example: `NA_instrument` uses a short label, "instrument", indicating instrument error resulting in missing values.
 These could be also used to indicate imputations, (Figure \@ref(fig:nabularfig)).
@@
 1.
-**Connectedness**: Binding the shadow matrix column-wise to the original data creates a single, connected, _nabular_ data format, which cannot be out of sync with the data.
+**Connectedness**: Binding the shadow matrix column-wise to the original data creates a single, connected, _nabular_ data format, in sync with the data.
 It is useful for visualization, summaries, and tracking imputed values, discussed in more detail in \@ref(nabular-data).
@@
 It is a portmanteau of `NA` and `tabular`.
-_Nabular_ data explicitly links missing values to data, keeping corresponding observations together and removing the possibility of mismatching records.
+_Nabular_ data explicitly links missing values to data, keeping corresponding observations together and removing the possibility of mismatching records (Figure \@ref(fig:nabularfig)).
 _Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable in a coordinated way: the missingness of `var`, as `var_NA` during analysis.
@@
 Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].
-_Nabular_ data is created with `nabular(data)` (Figure \@ref(fig:nabularfig)).
 
@@
 The shadow matrix can be converted to long form to create missingness summary plots.
-Nabular data is created by binding the columns od the data and shadow matrix.
+Nabular data is created by binding the columns of the data and shadow matrix.
 Special missing values (such as -99) are identified as special missings, and values imputed and tracked.
@@
 A list of common NA values for numbers and characters, can be provided to help check for typical representations of missings.
-These are implemented in `naniar` as `common_na_numbers` and `common_na_strings`, for numbers and characters.
+These are implemented in `naniar` as `common_na_numbers` and `common_na_strings`.
 
@@
   knitr::kable(
-    caption = "Table of the occurences of the search '-99' in the data, 'dat\\_ms'. There is one occurence if -99 in variables x and z.")
+    caption = "Table of the occurrences of the search '-99' in the data, 'dat\\_ms'. There is one occurrence if -99 in variables x and z.")
     # booktabs = TRUE) 
@@
 These are inspired by the `add_count` function in `dplyr`, which adds count information for specified groups or conditions.
-`naniar` provides operations to add the number or proportion of missingness, the missingness cluster, or the precense of any missings, with: `add_n_miss()`, `add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (Table \@ref(tab:add-missing-info)).
+`naniar` provides operations to add the number or proportion of missingness, the missingness cluster, or the presence of any missings, with: `add_n_miss()`, `add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (Table \@ref(tab:add-missing-info)).
 There are also functions for adding information about shadow values, and readable labels for any missing values with `add_label_shadow()` and `add_label_missings()`.
@@
 This is then recoded as a new factor level in the shadow matrix, so every column is aware of all possible new values of missingness.
-For example, the values -99 and -98 could be recoded to mean a broken machine sensor for the variable `x` with the following:
+For example, -99 could be recoded to indicate a broken machine sensor for the variable `x` with the following:
 
@@
 
-`naniar` does not reinvent the wheel for imputation, since instead working with existing methods.
+`naniar` does not reinvent the wheel for imputation, instead working with existing methods.
 However, `naniar` provides a few imputation methods to facilitate exploration and visualization: `impute_below`, `impute_mean`, and `impute_median`.
 While useful to explore structure in missingness, they are not recommended for use in analysis.
-`impute_below` imputes values below the minimum value, with some controlable jitter (random noise) to reduce overplotting.
+`impute_below` imputes values below the minimum value, with some controllable jitter (random noise) to reduce overplotting.
 
@@
 To evaluate imputations they need to be tracked.
-This is achieved by first using `nabular` data, then imputing, and imputed values can then be referred to by their shadow variable, `_NA` \@ref(fig:track-impute-example)).
+This is achieved by first using `nabular` data, then imputing, and imputed values can then be referred to by their shadow variable, `_NA` (Figure \@ref(fig:track-impute-example)).
 The code below shows the track pattern, first using `nabular`, then imputing with `impute_lm`.
-A label is added to facilitate identifying missings with `label_shadow`:
+`label_shadow` then adds a label to facilitate identifying missings:
 
@@
 Missing values are often ignored when plotting data - which is why the data visualisation software, MANET, was named using an acronym for "Missings Are Now Equally Treated" [@Unwin1996].
-However, plots can help to reveal the type of missing value patterns, even those of MCAR, MAR or MNAR.
+However, plots can help to identify the type of missing value patterns, even those of MCAR, MAR or MNAR.
 Here we summarise how to systematically explore missing patterns visually, and define useful plots to make, relative to the nabular data structure.
@@
 We learn from figure \@ref(fig:gg-miss-case-var)A-B, that two variables contain missings, approximately one third of observations have one missing value, and a tiny number of observations are missing across two variables.
-These overview plots are created from the shadow matrix in long form (Figure \@ref(fig:nabularfig).
-Numerical statistics can also be reported, and are discussed in the next section.
+These overview plots are created from the shadow matrix in long form (Figure \@ref(fig:nabularfig)).
+Numerical statistics can also be reported (Section \@ref(num-sum)).
 
@@
 
-The shadow matrix allows both the variables and cases to be displayed using a heatmap style visualization, with `vis_miss()` from \pkg{visdat} [@visdat].
+The shadow matrix can be put into long form, allowing both the variables and cases to be displayed using a heatmap style visualization, with `vis_miss()` from \pkg{visdat} [@visdat] (Figure \@ref(fig:nabularfig)).
 This also provides numerical summaries of missingness in the legend, and for each column (Figure \@ref(fig:gg-miss-case-var)C-D).
@@
 (A) A histogram using nabular data to show the values and missings in ozone.
-Values are imputed below the range to show the number of missings in Ozone and colored according to missingness of ozone (`Ozone\\_NA`).
+Values are imputed below the range to show the number of missings in ozone and colored according to missingness of ozone (`Ozone\\_NA`).
 There are about 35 missings in Ozone.
-Panel C shows Temperature according to missingness in ozone from in the airquality dataset.
+Panel C shows temperature according to missingness in ozone from in the airquality dataset.
 A histogram of temperature facetted by the missingness of ozone (B), or a density of temperature colored by missingness in ozone (C).
@@
 Tabulating the number of missings in variables in Table \@ref(tab:housing-miss-var-case-table) (left) shows three groups of missingness.
-Tabulating missings in cases (Table \@ref(tab:housing-miss-var-case-table) (right) shows six patterns of missingness.
+Tabulating missings in cases (Table \@ref(tab:housing-miss-var-case-table) (right)) shows six patterns of missingness.
 These overview plots lead to the removal of two variables from with more than 50% missingness from analysis: "building area" and "year built".

### 2020-05-14 paper/tidy-missing-data-paper.Rmd
added more to cover letter, and fixed some Australian spelling

@@
 In particular, there is little guidance on how to handle missing values in a data analysis workflow.
-Most analysis and visualisation software simply drop missing values when making a plot, although some (\pkg{ggplot2}) provide a warning (Figure \@ref(fig:warning)).
+Most analysis and visualization software simply drop missing values when making a plot, although some (\pkg{ggplot2}) provide a warning (Figure \@ref(fig:warning)).
 
@@
 
-The imputation literature focusses on ensuring valid statistical inference
+The imputation literature focuses on ensuring valid statistical inference
 is made from incomplete data. This is approached chiefly through probabilistic
@@
 This helps the analyst works swiftly and clearly, closing up opportunities for errors.
-Tidy data principles are general, but are comprehensively implemented in the R programming language, so this paper focusses on R [@rcore].
+Tidy data principles are general, but are comprehensively implemented in the R programming language, so this paper focuses on R [@rcore].
 
@@
 
-Taking existing methods from the missing data graphics literature, and translating and expanding them into tidy data and tidy tools would create more effective data visualisations.
+Taking existing methods from the missing data graphics literature, and translating and expanding them into tidy data and tidy tools would create more effective data visualizations.
 Defining these concepts allows the focus to be more general than just software, but rather, an extensible framework for tidy tools to explore missing data.
@@
 
-Missing values are often ignored when plotting data - which is why the data visualisation software, MANET, was named using an acronym for "Missings Are Now Equally Treated" [@Unwin1996].
+Missing values are often ignored when plotting data - which is why the data visualization software, MANET, was named and is an acronym corresponding to "Missings Are Now Equally Treated" [@Unwin1996].
 However, plots can help to identify the type of missing value patterns, even those of MCAR, MAR or MNAR.
@@
 The  first step is to get an overview of the extent of the missingness.
-Overview visualisations for variables and cases are provided with `gg_miss_var` and `gg_miss_case` (Figure \@ref(fig:gg-miss-case-var)A), drawing attention to the amount of missings, and ordering by missingness.
+Overview visualizations for variables and cases are provided with `gg_miss_var` and `gg_miss_case` (Figure \@ref(fig:gg-miss-case-var)A), drawing attention to the amount of missings, and ordering by missingness.
 The "airquality" dataset (from base R), is shown, and contains daily air quality measurements in New York, from May to September, 1973.
@@
 This type of plot can facilitate exploring missing data distributions.
-For example, we would expect if data were MCAR, for values to be roughly uniformly missing thoughout the histogram or density.
+For example, we would expect if data were MCAR, for values to be roughly uniformly missing throughout the histogram or density.
 
@@
 Figure \@ref(fig:geom-miss) shows a mostly uniform spread of missing values for Solar.R and Ozone.
-As `geom_miss_point()` is a defined \pkg{ggplot2} geometry, it works with features such as facetting and mapping other variables to graphical aesthetics.
+As `geom_miss_point()` is a defined \pkg{ggplot2} geometry, it works with features such as faceting and mapping other variables to graphical aesthetics.
 
@@
 They transform variables to the same scale, ranging between 0 and 1.
-The `oceanbuoys` dataset from \pkg{naniar} is used for this visualisation, containing measurements of sea and air temperature, humidity, and east west and north south wind speeds.
+The `oceanbuoys` dataset from \pkg{naniar} is used for this visualization, containing measurements of sea and air temperature, humidity, and east west and north south wind speeds.
 Data was collected in 1993 and 1997, to understand and predict El Niño and El Niña.
@@
 There are `r scales::number(nrow(housing), big.mark = ",")` properties, and `r ncol(housing)` variables in the dataset.
-The variables include real estate type (town house, unit, house), suburb, selling method, number of rooms, price, real estate agent, sale date, and distance from Central Business District (CBD).
+The variables include real estate type (town house, unit, house), suburb, selling method, number of rooms, price, real estate agent, sale date, and distance from the Central Business District (CBD).
 
@@
 The data contains many missing values.
-As a precursor to building a predictive model, this analysis focusses on understanding the patterns of missingness.
+As a precursor to building a predictive model, this analysis focuses on understanding the patterns of missingness.
 

### 2020-05-14 paper/tidy-missing-data-paper.Rmd

@@
 bibliography: tidy-missing-data-paper.bib
-# csl: jcgs.csl
 ---
@@
 Most analysis and visualization software simply drop missing values when making a plot, although some (\pkg{ggplot2}) provide a warning (Figure \@ref(fig:warning)).
-
-<!-- Tidy data [@Wickham2014] is a relatively new principle and suite of tools -->
-<!-- facilitating the process of converting raw data into a clean, analysis-ready -->
-<!-- format that works efficiently in a data analysis pipeline. In tidy data, missing -->
-<!-- values can be handled implicitly or explicitly, at the analyst's discretion. -->
-<!-- However, when making a plot, missing values are simply dropped, albeit with a -->
-<!-- warning (see Figure \@ref(fig:warning)). Most other analysis software will -->
-<!-- simply drop cases with missings without warning. -->
 

### 2020-05-15 paper/tidy-missing-data-paper.Rmd
render .bbl file

@@
                       cache = TRUE)
-
+options(tinytex.clean = FALSE)
 library(tidyverse) 

### 2020-08-18 paper/review/response-to-review.Rmd
add reviewer comments

---
title: "Response to Review"
---

JSS 4108:

Nicholas John Tierney, Dianne Cook

Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

---------------------------------------------------------
For further instructions on JSS style requirements please see the Style Files (in particular section 2.1 Style Checklist of jss.pdf) and FAQ at http://www.jstatsoft.org/about/submissions.

For further examples please see RECENT JSS papers.
---------------------------------------------------------

From the editorial team:

We apologize for the additional delays in processing 
your submission due to the COVID-19 pandemic.

The submission "Expanding tidy data principles to facilitate missing data
exploration, visualization and assessment of imputations" presents the R 
package naniar which provides data structures, summaries and visualizations 
for missing data.

Additional comments:

Please add some discussion on what kind of functionality is available in 
other statistical software packages than R in order to better embed the 
contribution not only within R, but the statistical software ecosystem in 
general.

Please provide a plain R script which when run reproduces all results 
presented in the manuscript. In general purling the Rmd file and editing 
the resulting R file  slightly in order to guide the user what each block 
does, removing knitr specific content,   etc. is suitable.

* Please add some discussion on what kind of functionality is available in 
other statistical software packages than R in order to better embed the 
contribution not only within R, but the statistical software ecosystem in 
general.

* Please provide a plain R script which when run reproduces all results 
presented in the manuscript. In general purling the Rmd file and editing 
the resulting R file slightly in order to guide the user what each block 
does, removing knitr specific content, etc. is suitable.

* The JSS style (https://www.jstatsoft.org/pages/view/style) should be 
followed more closely, especially using the title vs. sentence case, 
JSS-specific markup (\proglang, \pkg), and improving the BibTeX file. Some 
aspects of the style are handled incorrectly by "rticles", though (see
https://github.com/rstudio/rticles/issues/254).

* Note that the affiliation of the second author is misspelled.

Manuscript style comments:

o For the code layout in R publications, we typically distinguish input/output using Sinput/Soutput (or equivalently CodeInput/CodeOutput). Unless there are special reasons to format it differently, the input should use the text width (up to 76 or 77 characters) and be indented by two spaces, e.g.,

begin{Sinput}
R> example_model <- lm(response ~ variable1 + variable2 + variable3, 
+    weights = w, data = mydata)
\end{Sinput}

o If using "e.g." and "i.e." add a comma after the period to keep LaTeX from interpreting them as the end of a sentence, i.e.: "e.g., " and "i.e., ".

o The rule for capitalizing the starting letters of Figure, Section and Table is as follows: If you are referring to a particular figure/section/table then capitalize the first letter, otherwise use a lower-case first letter. For example, something shown in Section 4 vs. there are three sections in this paper.

o All captions should appear below the corresponding figure/table. The captions should be in sentence style and end with a period.  No additional formatting (such as \emph, f or \it) should be used for the caption.

o All table row/column headers should also be in sentence style. There should not be further footnote-style annotations in tables; these should all be placed in the caption.

o As a reminder, please make sure that:
	- \proglang, \pkg and \code have been used for highlighting throughout the paper (including titles and references), except where explicitly escaped.

References:

o Journal of the Royal Statistical Society B (not: Journal of the Royal Statistical Society, Series B)
	the EM Algorithm.” Journal of the Royal Statistical Society. Series B
the EM Algorithm.” Journal of the Royal Statistical Society. Series B
o Springer-Verlag (not: Springer)
	and GGobi. 1st edition. Springer Publishing Company, Incorporated.

o Please make sure that all software packages are \cite{}'d properly.
o All references should be in title style.
o See FAQ for specific reference instructions.

Code:

o Please make sure that the files needed to replicate all code/examples within the manuscript are included in a standalone replication script.

### 2020-09-04 paper/review/response-to-review.Rmd
fix a compilation error and start adding responses to comments

@@
 
+> We have added the following:
+
+- Missing data can be visualised in other software such as STATA, SAS, and SPSS
+- They provide basic table frequencies
+- SPSS provides visualisations on missingness patterns
+- No software provides drop in functions to specifically visualise missing values.
+- Although it is possible to visualise missing values after performing data manipulations to count and summarise the missingness.
+
 Please provide a plain R script which when run reproduces all results 
@@
 
-* Please add some discussion on what kind of functionality is available in 
-other statistical software packages than R in order to better embed the 
-contribution not only within R, but the statistical software ecosystem in 
-general.
-
-* Please provide a plain R script which when run reproduces all results 
-presented in the manuscript. In general purling the Rmd file and editing 
-the resulting R file slightly in order to guide the user what each block 
-does, removing knitr specific content, etc. is suitable.
+> We have added a file, "tidy-missingdata-paper.R" that contains the R code from the paper
 
@@
 
+>
+
 * Note that the affiliation of the second author is misspelled.
+
+> We have fixed the affiliation of the second author
 
@@
 
+> I am not sure how to address this with `rticles`, I assume I'll need to update
+to the latest version  and give that a go.
+
 o If using "e.g." and "i.e." add a comma after the period to keep LaTeX from interpreting them as the end of a sentence, i.e.: "e.g., " and "i.e., ".
+
+> Instances of `e.g.` have been replaced with `e.g.,`. No cases of `i.e.` were found.
 
@@
 
+> Appropriate instances of `section` and `figure` have been replaced with `Section` and `Figure`.
+
 o All captions should appear below the corresponding figure/table.
@@
 No additional formatting (such as \emph, f or \it) should be used for the caption.
+
+> I'm not sure it is possible to control the caption output of a kable Table, I will look into this.
 

### 2020-09-04 paper/tidy-missing-data-paper.Rmd
fix a compilation error and start adding responses to comments

@@
   - name: Dianne Cook
-    affiliation: Monsh University
+    affiliation: Monash University
 title:
@@
 
-The imputation literature focuses on ensuring valid statistical inference
-is made from incomplete data. This is approached chiefly through probabilistic
-modelling, assuming the mechanism of missing data is known to the analyst. It
-does not address how to explore, understand, and handle missing data structures
-and mechanisms.
-
-However, something must be known about the missing value structure to produce a
-complete dataset for analysis, whether by case- or variable-deletion, or with
-imputation. Decision tree models or latent group analysis can reveal structures or patterns
-of missingness [@Tierney2015; @Barnett2017], but are not definitive. While there
-are times where the missing data mechanism is obvious, determining this is typically not straightforward. To understand their structure and possible mechanisms, the analyst must explore the data
-with visualizations, summaries, and modelling, in an iterative fashion.
+The imputation literature focuses on ensuring valid statistical inference is made from incomplete data.
+This is approached chiefly through probabilistic modelling, assuming the mechanism of missing data is known to the analyst.
+It does not address how to explore, understand, and handle missing data structures and mechanisms.
+
+However, something must be known about the missing value structure to produce a complete dataset for analysis, whether by case- or variable-deletion, or with imputation.
+Decision tree models or latent group analysis can reveal structures or patterns of missingness [@Tierney2015; @Barnett2017], but are not definitive.
+While there are times where the missing data mechanism is obvious, determining this is typically not straightforward.
+To understand their structure and possible mechanisms, the analyst must explore the data with visualizations, summaries, and modelling, in an iterative fashion.
 
@@
 This paper is organized in the following way.
-The next section (\@ref(background)) provides the background to tidy data principles and tools (\@ref(tidy-data-concepts)) and missing data representations (\@ref(missing-data-rep-dep)).
+Section (\@ref(background)) provides the background to tidy data principles and tools (\@ref(tidy-data-concepts)) and missing data representations (\@ref(missing-data-rep-dep)).
 Section \@ref(existing-software) summarises existing software for handling missing data.
@@
 
-Features of tidy data were formally described in @Wickham2014, and were discussed in terms of their importance for data science by @Donoho2017, and tools for data analysis.
-Tidy data is defined by @r4ds as:
+Features of tidy data were formally described in [@Wickham2014], and were discussed in terms of their importance for data science by [@Donoho2017], and tools for data analysis.
+Tidy data is defined by [@r4ds] as:
 
 > 1.  Each variable must have its own column.
-> 1. Each observation must have its own row.
-> 1. Each value must have its own cell.
+> 2.  Each observation must have its own row.
+> 3.  Each value must have its own cell.
 
@@
 
-There are many ways each value can be missing, we adopt the notation used in @VanBuuren2012.
+There are many ways each value can be missing, we adopt the notation used in [@VanBuuren2012].
 The information in $B$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR).
 The distribution of missing values in $b_{ij}$ can depend on the entire dataset, represented as $Y = (Y_{obs}, Y_{miss})$.
-This relationship can be defined by the _missing data model_ $Pr(b_{ij} | Y_{obs}, Y_{miss}, \psi)$, the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, $\psi$.
+This relationship can be defined by the *missing data model* $Pr(b_{ij} | Y_{obs}, Y_{miss}, \psi)$, the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, $\psi$.
 This helps to precisely define categories of missing values.
@@
 <!-- (NOTE: Is the last sentence OK?) -->
+
 <!-- (NOTE: trying to come up with a sentence that leads from this section to the next) -->
@@
 
-The primary focus of most missing data packages is making inferences, and exploring imputed values, not
-on exploring relationships in missing values, and identifying possible patterns.
+The primary focus of most missing data packages is making inferences, and exploring imputed values, not on exploring relationships in missing values, and identifying possible patterns.
 Texts covering the exploration phase of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools [@VanBuuren2012]; this is inefficient.
@@
 
-The MANET (Missings Are Now Equally Treated) software @Unwin1996 focussed on multivariate categorical data, with missingness explicitly added as a category.
+The MANET (Missings Are Now Equally Treated) software [@Unwin1996] focussed on multivariate categorical data, with missingness explicitly added as a category.
 MANET also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot.
 The MANET software is no longer maintained and cannot be installed.
-The approach of @Swayne1998 in the software XGobi, further developed in ggobi [@Cook2007], focussed on multivariate quantitative data.
+The approach of [@Swayne1998] in the software XGobi, further developed in ggobi [@Cook2007], focussed on multivariate quantitative data.
 Missingness is incorporated into plots in ggobi by setting them to be 10% below the minimum value.
@@
 
-Applying tidyverse principles has the potential to clarify missing data exploration, visualization, and imputation. This
-section discusses how these principles are applied for data structures (\@ref(data-structure)), common operations (verbs) (\@ref(verbs)), graphics (\@ref(graphics)), and data summaries (\@ref(num-sum)).
+Applying tidyverse principles has the potential to clarify missing data exploration, visualization, and imputation.
+This section discusses how these principles are applied for data structures (\@ref(data-structure)), common operations (verbs) (\@ref(verbs)), graphics (\@ref(graphics)), and data summaries (\@ref(num-sum)).
 Care has been taken to make the names and design these features intuitive for their purpose, and is discussed throughout.
@@
 A useful template common in missing data literature, is the $B$ matrix, where 0 and 1 indicate not missing and missing, respectively.
-This matrix was used to explore missing values in the interactive graphics library XGobi, called a "missing value shadow", or "shadow matrix", defined as a copy of the original data with
-indicator values of missingness.
+This matrix was used to explore missing values in the interactive graphics library XGobi, called a "missing value shadow", or "shadow matrix", defined as a copy of the original data with indicator values of missingness.
 The shadow matrix could be interactively linked to the data.
@@
 
-1.
-**Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA_\<suffix\>".
+2.
+**Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA\_\<suffix\>".
 For example: `NA_instrument` uses a short label, "instrument", indicating instrument error resulting in missing values.
@@
 
-1.
-**Coordinated names**: Variable names in the shadow matrix gain a consistent short suffix, "_NA", keeping names coordinated throughout analysis (Figure \@ref(fig:nabularfig)).
+3.
+**Coordinated names**: Variable names in the shadow matrix gain a consistent short suffix, \"\_NA\", keeping names coordinated throughout analysis (Figure \@ref(fig:nabularfig)).
 It makes a clear distinction with `var_NA` being a random variable of the missingness of a variable, `var`.
@@
 
-1.
-**Connectedness**: Binding the shadow matrix column-wise to the original data creates a single, connected, _nabular_ data format, in sync with the data.
+4.
+**Connectedness**: Binding the shadow matrix column-wise to the original data creates a single, connected, *nabular* data format, in sync with the data.
 It is useful for visualization, summaries, and tracking imputed values, discussed in more detail in \@ref(nabular-data).
@@
 
-_Nabular_ data binds the shadow matrix column-wise to the original data.
+*Nabular* data binds the shadow matrix column-wise to the original data.
 It is a portmanteau of `NA` and `tabular`.
-_Nabular_ data explicitly links missing values to data, keeping corresponding observations together and removing the possibility of mismatching records (Figure \@ref(fig:nabularfig)).
-_Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable in a coordinated way: the missingness of `var`, as `var_NA` during analysis.
-_Nabular_ data is a snapshot of the missingness of the data.
-This means when _nabular_ data are imputed, those imputed values can easily be identified in analysis (`var` vs `var_NA`).
-_Nabular_ data is not unlike classical data formats with quality or flag variables associated with each measured variable, e.g.
-Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].
+*Nabular* data explicitly links missing values to data, keeping corresponding observations together and removing the possibility of mismatching records (Figure \@ref(fig:nabularfig)).
+*Nabular* data facilitates visualization and summaries by allowing the user to reference the missingness of a variable in a coordinated way: the missingness of `var`, as `var_NA` during analysis.
+*Nabular* data is a snapshot of the missingness of the data.
+This means when *nabular* data are imputed, those imputed values can easily be identified in analysis (`var` vs `var_NA`).
+*Nabular* data is not unlike classical data formats with quality or flag variables associated with each measured variable, e.g., Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].
 
@@
 
-Although a binary missingness matrix, $B$, could be generated during analysis instead of using a _nabular_ data structure, there are key advantages to using _nabular_ data.
+Although a binary missingness matrix, $B$, could be generated during analysis instead of using a *nabular* data structure, there are key advantages to using *nabular* data.
 Firstly, missing value labels `NA` and `!NA` are clearer than TRUE or FALSE.
@@
 Finally, the logical matrix cannot capture which values are imputed if imputation has already taken place.
-Imputing values on _nabular_ data automatically tracks these imputations.
-
-Using additional columns to represent missingness information follows best practices for data organization, described in @Ellis2017 and @Broman2017: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column.
+Imputing values on *nabular* data automatically tracks these imputations.
+
+Using additional columns to represent missingness information follows best practices for data organization, described in [@Ellis2017] and [@Broman2017]: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column.
 Here they suggest to indicate censored data with an extra variable called "VariableNameCensored", which would be TRUE if censored, otherwise FALSE.
@@
 `naniar` implements replacement with the function, `replace_with_na()`.
-Values -99 could be replaced in the `x` column with:
-`replace_with_na(dat_ms, replace = list(x = -99))`.
+Values -99 could be replaced in the `x` column with: `replace_with_na(dat_ms, replace = list(x = -99))`.
 For operating on multiple variables, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`.
@@
 
-Understanding the missingness structure can be improved by adding summary information alongside the data. For example, in @Tierney2015, the proportion of
-missings in a row is used as the outcome in a model to identify variables
-important in predicting missingness structures.
+Understanding the missingness structure can be improved by adding summary information alongside the data.
+For example, in [@Tierney2015], the proportion of missings in a row is used as the outcome in a model to identify variables important in predicting missingness structures.
 `naniar` implements a series of functions to add these missingness summaries to the data, starting with `add_`.
@@
 
-_Nabular_ data has the shadow matrix column-bound to existing data.
+*Nabular* data has the shadow matrix column-bound to existing data.
 This facilitates visualization and summaries, and allows for imputed values to be tracked.
-_Nabular_ data can be created with `nabular()`:
+*Nabular* data can be created with `nabular()`:
 
@@
 For example, there might be special codes to mark an individual dropping out of a study, known instrument failure in weather instruments, or for values censored in analysis.
-`naniar` provides tools to encode these special types of missingness in the shadow matrix of  _nabular_  data, using `recode_shadow()`.
+`naniar` provides tools to encode these special types of missingness in the shadow matrix of *nabular* data, using `recode_shadow()`.
 This requires specifying the variable to contain the flagged value, the condition for flagging, and a suffix.
@@
 The "airquality" dataset (from base R), is shown, and contains daily air quality measurements in New York, from May to September, 1973.
-We learn from figure \@ref(fig:gg-miss-case-var)A-B, that two variables contain missings, approximately one third of observations have one missing value, and a tiny number of observations are missing across two variables.
+We learn from Figure \@ref(fig:gg-miss-case-var)A-B, that two variables contain missings, approximately one third of observations have one missing value, and a tiny number of observations are missing across two variables.
 These overview plots are created from the shadow matrix in long form (Figure \@ref(fig:nabularfig)).
@@
 Missing values are typically not shown for univariate visualizations such as histograms or densities.
-Two ways to use _nabular_ data to present univariate data with missings are discussed.
+Two ways to use *nabular* data to present univariate data with missings are discussed.
 The first imputes values below the range to facilitate visualizations.
@@
 **Imputing values below the range**.
-To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed below the range of data using `impute_below_all`.
+To visualize the amount of missings in each variable, the data is transformed into *nabular* form, then values are imputed below the range of data using `impute_below_all`.
 Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone values on the right in green, and the histogram of missing ozone values on the left, in orange.
@@
 To visualize missing values in two dimensions the missing values can be placed in plot margins, by imputing values below the range of the data.
-Using _nabular_ data identifies imputed values, and color makes missingness pre-attentive [@treisman1985].
+Using *nabular* data identifies imputed values, and color makes missingness pre-attentive [@treisman1985].
 The steps of imputing and coloring are combined into `geom_miss_point()`.
@@
 `naniar` works with `dplyr`'s `group_by` operator to produce grouped summaries, which work well with the "pipe" operator.
-The code and table below \@ref(tab:group-miss-var-summary) show an example of missing data summaries for airquality, grouped by month.
+The code and Table \@ref(tab:group-miss-var-summary) below show an example of missing data summaries for airquality, grouped by month.
 
@@
 
-dat_house_cc <- na.omit(dat_house)
+dat_house_cc <- dat_house %>% nabular() %>% add_label_shadow() %>% na.omit()
 
@@
 
-The _nabular_ data structures discussed in the paper are simple by design, to promote flexibility.
+The *nabular* data structures discussed in the paper are simple by design, to promote flexibility.
 They could be used to create different visualizations than were shown in the paper.
@@
 
-Other data structures such as spatial data, time series, networks, and
-longitudinal data would be supported by the inherently tabular,
-_nabular_ data, if they are first structured as wide tidy format.
+Other data structures such as spatial data, time series, networks, and longitudinal data would be supported by the inherently tabular, *nabular* data, if they are first structured as wide tidy format.
 Large data may need special handling, and additional features like efficient storage of purely imputed values and lazy evaluation.
@@
 
-The methodology described in this paper can be used in conjunction with other approaches to understand multivariate missingness dependencies (e.g.
-decision trees [@Tierney2015], latent group analysis [@Barnett2017], and PCA [@FactoMineR]).
-Evaluating imputed values using a testing framework like @VanBuuren2012 is also supported.
+The methodology described in this paper can be used in conjunction with other approaches to understand multivariate missingness dependencies (e.g., decision trees [@Tierney2015], latent group analysis [@Barnett2017], and PCA [@FactoMineR]).
+Evaluating imputed values using a testing framework like [@VanBuuren2012] is also supported.
 

### 2020-09-07 paper/review/response-to-review.Rmd
finish draft response to reviewers

@@
 
-> We have added the following:
+> We have added the following at the end of section 3:
 
-- Missing data can be visualised in other software such as STATA, SAS, and SPSS
-- They provide basic table frequencies
-- SPSS provides visualisations on missingness patterns
-- No software provides drop in functions to specifically visualise missing values.
-- Although it is possible to visualise missing values after performing data manipulations to count and summarise the missingness.
+Missing data can be explored and visualised in other software such as \proglang{STATA}, \proglang{SAS}, and \proglang{SPSS}.
+They provide basic table frequencies of missing data, and basic visualisations on missingness patterns.
+However, none of these software provide functions to specifically visualise missing values in bivariate or multivariate settings, and it is only possible to visualise missing values after performing data analysis, counting and summarising missingness.
+Imputation methods from mean to regression, EM, and multiple imputation are all available, with out of the box diagnostics provided to assist comparing imputations.
+\proglang{python} also provides exploratory data analysis for missing data within \pkg{missingno} [@Bilogur2018], and the \pkg{scikit-learn} package provides imputation methods [@scikit-learn].
 
@@
 
->
+> I have updated to the latest rticles package and checked that the JSS-specific markup has been fixed in the latest submission.
 
@@
 
-> I am not sure how to address this with `rticles`, I assume I'll need to update
-to the latest version  and give that a go.
+> I believe this has been resolved with the latest update to rticles
 
@@
 
+>
+
 o As a reminder, please make sure that:

### 2020-09-07 paper/tidy-missing-data-paper.Rmd
finish draft response to reviewers

@@
 Together, these provide a connected framework for handling, exploring, and imputing missing values.
-These methods are available in the R package `naniar`.
+These methods are available in the R package \pkg{naniar}.
 keywords:
@@
 This helps the analyst works swiftly and clearly, closing up opportunities for errors.
-Tidy data principles are general, but are comprehensively implemented in the R programming language, so this paper focuses on R [@rcore].
+Tidy data principles are general, but are comprehensively implemented in the \proglang{R} programming language, so this paper focuses on \proglang{R} [@rcore].
 
@@
 
-Defining tidy data and tidy tools has resulted in a growing set of packages known collectively as the "tidyverse" [@tidyverse].
+Defining tidy data and tidy tools has resulted in a growing set of packages known collectively as the \pkg{tidyverse} [@tidyverse].
 These are constructed to share similar principles in their design and behavior, and cover the breadth of an analysis - from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @tidyverse; @Tidyverse-Manifesto].
@@
 These diverse approaches allows for imputing with semi-continuous, continuous, count, and categorical data.
-VIM identifies imputed cases by adding an indicator variable with a suffix `_imp`.
-So `Var1` has a sibling column, `Var1_imp`, with values TRUE or FALSE indicating imputation.
+VIM identifies imputed cases by adding an indicator variable with a suffix "_imp".
+So "Var1" has a sibling column, "Var1_imp", with values TRUE or FALSE indicating imputation.
 \pkg{VIM} also has a variety of visualization methods, discussed in \@ref(exploration).
@@
 
-The MANET (Missings Are Now Equally Treated) software [@Unwin1996] focussed on multivariate categorical data, with missingness explicitly added as a category.
-MANET also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot.
-The MANET software is no longer maintained and cannot be installed.
-The approach of [@Swayne1998] in the software XGobi, further developed in ggobi [@Cook2007], focussed on multivariate quantitative data.
-Missingness is incorporated into plots in ggobi by setting them to be 10% below the minimum value.
-
-MissingDataGUI provides a Graphical User Interface (GUI) for exploring missing data structure, both numerically and visually.
+The \pkg{MANET} (Missings Are Now Equally Treated) software [@Unwin1996] focussed on multivariate categorical data, with missingness explicitly added as a category.
+\pkg{MANET} also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot.
+The \pkg{MANET} software is no longer maintained and cannot be installed.
+The approach of [@Swayne1998] in the software \pkg{XGobi}, further developed in \pkg{ggobi} [@Cook2007], focussed on multivariate quantitative data.
+Missingness is incorporated into plots in \pkg{ggobi} by setting them to be 10% below the minimum value.
+
+\pkg{MissingDataGUI} provides a Graphical User Interface (GUI) for exploring missing data structure, both numerically and visually.
 Using a GUI to explore missing data facilitates rapid insight into missingness structures.
@@
 
-VIM (Visualizing and Imputing Missing Data) provides visualization methods to identify and explore observed, imputed, and missing values.
+\pkg{VIM} (Visualizing and Imputing Missing Data) provides visualization methods to identify and explore observed, imputed, and missing values.
 These include spinograms, spinoplots, missingness matrices, plotting missingness in the margins of other plots, and other summaries.
@@
 
+## Approaches in other languages
+
+Missing data can be explored and visualised in closed source software such as \proglang{STATA}, \proglang{SAS}, and \proglang{SPSS}.
+These provide basic table frequencies of missing data, and basic visualisations on missingness patterns.
+However, none of these software provide functions to specifically visualise missing values in bivariate or multivariate settings, and it is only possible to visualise missing values after performing data analysis, counting and summarising missingness.
+Imputation methods from mean to regression, EM, and multiple imputation are all available, with out of the box diagnostics provided to assist comparing imputations.
+\proglang{python} also provides exploratory data analysis for missing data within \pkg{missingno} [@Bilogur2018], and the \pkg{scikit-learn} package provides imputation methods [@scikit-learn].
+
 # Tidy framework for missings {#extensions}
@@
 A useful template common in missing data literature, is the $B$ matrix, where 0 and 1 indicate not missing and missing, respectively.
-This matrix was used to explore missing values in the interactive graphics library XGobi, called a "missing value shadow", or "shadow matrix", defined as a copy of the original data with indicator values of missingness.
+This matrix was used to explore missing values in the interactive graphics library \pkg{XGobi}, called a "missing value shadow", or "shadow matrix", defined as a copy of the original data with indicator values of missingness.
 The shadow matrix could be interactively linked to the data.
@@
 **Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA\_\<suffix\>".
-For example: `NA_instrument` uses a short label, "instrument", indicating instrument error resulting in missing values.
+For example: "NA_instrument" uses a short label, "instrument", indicating instrument error resulting in missing values.
 These could be also used to indicate imputations, (Figure \@ref(fig:nabularfig)).
@@
 **Coordinated names**: Variable names in the shadow matrix gain a consistent short suffix, \"\_NA\", keeping names coordinated throughout analysis (Figure \@ref(fig:nabularfig)).
-It makes a clear distinction with `var_NA` being a random variable of the missingness of a variable, `var`.
+It makes a clear distinction with "var_NA" being a random variable of the missingness of a variable, "var".
 This suffix is short and easy to remember during data analysis, and shifts the focus from the value of a variable, to its missingness state.
@@
 *Nabular* data binds the shadow matrix column-wise to the original data.
-It is a portmanteau of `NA` and `tabular`.
+It is a portmanteau of "NA" and "tabular".
 *Nabular* data explicitly links missing values to data, keeping corresponding observations together and removing the possibility of mismatching records (Figure \@ref(fig:nabularfig)).
-*Nabular* data facilitates visualization and summaries by allowing the user to reference the missingness of a variable in a coordinated way: the missingness of `var`, as `var_NA` during analysis.
+*Nabular* data facilitates visualization and summaries by allowing the user to reference the missingness of a variable in a coordinated way: the missingness of "var", as "var_NA" during analysis.
 *Nabular* data is a snapshot of the missingness of the data.
-This means when *nabular* data are imputed, those imputed values can easily be identified in analysis (`var` vs `var_NA`).
+This means when *nabular* data are imputed, those imputed values can easily be identified in analysis ("var" vs "var_NA").
 *Nabular* data is not unlike classical data formats with quality or flag variables associated with each measured variable, e.g., Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].
@@
 Although a binary missingness matrix, $B$, could be generated during analysis instead of using a *nabular* data structure, there are key advantages to using *nabular* data.
-Firstly, missing value labels `NA` and `!NA` are clearer than TRUE or FALSE.
+Firstly, missing value labels "NA" and "!NA" are clearer than TRUE or FALSE.
 Secondly, special missing values cannot be added easily during analysis with a logical matrix.
@@
 Encoding special missing values is achieved by defining logical conditions and suffixes.
-This is implemented with the `recode_shadow` function in \pkg{naniar} (\@ref(verbs-recode)).
-
-Special missing values are not a new idea, and have been implemented in other statistical programming languages, SPSS, SAS, and STATA.
+This is implemented with the \code{recode_shadow()} function in \pkg{naniar} (\@ref(verbs-recode)).
+
+Special missing values are not a new idea, and have been implemented in other statistical programming languages, \proglang{SPSS}, \proglang{SAS}, and \proglang{STATA}.
 These typically represent missing values as a full-stop, `.`, and record special missing values as `.a` - `.z`.
@@
 For missing data, these include: **scan**, **replace**, **add**, **shadow**, **impute**, **track**, and **flag**.
-Data can be **scanned** to find possible missings not coded as `NA`.
-These values can then be **replaced** with `NA`.
+Data can be **scanned** to find possible missings not coded as \code{NA}.
+These values can then be **replaced** with \code{NA}.
 To facilitate exploration, summaries of missingness can be **added** as a summary column to the original data.
@@
 
-This operation is used to search the data for specific conventional representations of missings, such as, "N/A", "MISSING", `-99`.
-This is implemented in the function `miss_scan_count()`, which returns a table of occurrences of that value for each variable.
+This operation is used to search the data for specific conventional representations of missings, such as, "N/A", "MISSING", "-99".
+This is implemented in the function \code{miss_scan_count()}, which returns a table of occurrences of that value for each variable.
 A list of common NA values for numbers and characters, can be provided to help check for typical representations of missings.
-These are implemented in `naniar` as `common_na_numbers` and `common_na_strings`.
+These are implemented in \pkg{naniar} as objects \code{common_na_numbers} and \code{common_na_strings}.
 
@@
 Once possible missing values have been identified, these values can be replaced.
-For example, a dataset could have the values `-99` meaning a missing value.
-`naniar` implements replacement with the function, `replace_with_na()`.
-Values -99 could be replaced in the `x` column with: `replace_with_na(dat_ms, replace = list(x = -99))`.
-For operating on multiple variables, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`.
-This means `replace_with_na_all` operates on **all** columns, `replace_with_na_at` operates **at** specific columns, and `replace_with_na_if` makes a conditional change on columns **if** they meet some condition (such as `is.numeric` or `is.character`).
+For example, a dataset could have the values "-99" meaning a missing value.
+\pkg{naniar} implements replacement with the function, \code{replace_with_na()}.
+Values "-99" could be replaced in the "x" column with: \code{replace_with_na(dat_ms, replace = list(x = -99))}.
+For operating on multiple variables, there are scoped variants for \code{replace_with_na()}: \code{_all}, \code{_if}, and \code{_at}.
+This means \code{replace_with_na_all()} operates on **all** columns, \code{replace_with_na_at()} operates **at** specific columns, and \code{replace_with_na_if()} makes a conditional change on columns **if** they meet some condition (such as \code{is.numeric()} or \code{is.character()}).
 
@@
 For example, in [@Tierney2015], the proportion of missings in a row is used as the outcome in a model to identify variables important in predicting missingness structures.
-`naniar` implements a series of functions to add these missingness summaries to the data, starting with `add_`.
-These are inspired by the `add_count` function in `dplyr`, which adds count information for specified groups or conditions.
-`naniar` provides operations to add the number or proportion of missingness, the missingness cluster, or the presence of any missings, with: `add_n_miss()`, `add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (Table \@ref(tab:add-missing-info)).
-There are also functions for adding information about shadow values, and readable labels for any missing values with `add_label_shadow()` and `add_label_missings()`.
+\pkg{naniar} implements a series of functions to add these missingness summaries to the data, starting with \code{add_}.
+These are inspired by the \code{add_count()} function in \pkg{dplyr}, which adds count information for specified groups or conditions.
+\pkg{naniar} provides operations to add the number or proportion of missingness, the missingness cluster, or the presence of any missings, with: \code{add_n_miss()}, \code{add_prop_miss()}, \code{add_miss_cluster()}, and \code{add_any_miss()}, respectively (Table \@ref(tab:add-missing-info)).
+There are also functions for adding information about shadow values, and readable labels for any missing values with \code{add_label_shadow()} and \code{add_label_missings()}.
 
@@
 This facilitates visualization and summaries, and allows for imputed values to be tracked.
-*Nabular* data can be created with `nabular()`:
+*Nabular* data can be created with \code{nabular()}:
 
@@
 
-Unusual or spurious data values are often identified and `flagged`.
+Unusual or spurious data values are often identified and "flagged".
 For example, there might be special codes to mark an individual dropping out of a study, known instrument failure in weather instruments, or for values censored in analysis.
-`naniar` provides tools to encode these special types of missingness in the shadow matrix of *nabular* data, using `recode_shadow()`.
+\pkg{naniar} provides tools to encode these special types of missingness in the shadow matrix of *nabular* data, using \code{recode_shadow()}.
 This requires specifying the variable to contain the flagged value, the condition for flagging, and a suffix.
 This is then recoded as a new factor level in the shadow matrix, so every column is aware of all possible new values of missingness.
-For example, -99 could be recoded to indicate a broken machine sensor for the variable `x` with the following:
+For example, -99 could be recoded to indicate a broken machine sensor for the variable "x" with the following:
 
@@
 
-`naniar` does not reinvent the wheel for imputation, instead working with existing methods.
-However, `naniar` provides a few imputation methods to facilitate exploration and visualization: `impute_below`, `impute_mean`, and `impute_median`.
+\pkg{naniar} does not reinvent the wheel for imputation, instead working with existing methods.
+However, \pkg{naniar} provides a few imputation methods to facilitate exploration and visualization: \code{impute_below()}, \code{impute_mean()}, and \code{impute_median()}.
 While useful to explore structure in missingness, they are not recommended for use in analysis.
-`impute_below` imputes values below the minimum value, with some controllable jitter (random noise) to reduce overplotting.
-
-Similar to `simputation`, each `impute_` function returns the data with values imputed.
-However, `naniar` does not use a formula syntax, instead each function has "scoped variants" `_all`, `_at` and `_if` as in \@ref(verbs-replace-with).
-`impute_` functions with no scoped variant, (`impute_mean`), will work on a single vector, but not a data.frame.
+\code{impute_below()} imputes values below the minimum value, with some controllable jitter (random noise) to reduce overplotting.
+
+Similar to \pkg{simputation}, each \code{impute_} function returns the data with values imputed.
+However, \pkg{naniar} does not use a formula syntax, instead each function has "scoped variants" \code{_all}, \code{_at} and \code{_if} as in \@ref(verbs-replace-with).
+\code{impute_} functions with no scoped variant, (\code{impute_mean()}), will work on a single vector, but not a data.frame.
 One challenge with this approach is imputed value locations are not tracked.
-This issue is resolved with \code{nabular} data covered in Section \@ref(verbs-track).
+This issue is resolved with *nabular* data covered in Section \@ref(verbs-track).
 
@@
 To evaluate imputations they need to be tracked.
-This is achieved by first using `nabular` data, then imputing, and imputed values can then be referred to by their shadow variable, `_NA` (Figure \@ref(fig:track-impute-example)).
-The code below shows the track pattern, first using `nabular`, then imputing with `impute_lm`.
-`label_shadow` then adds a label to facilitate identifying missings:
+This is achieved by first using *nabular* data, then imputing, and imputed values can then be referred to by their shadow variable, "_NA" (Figure \@ref(fig:track-impute-example)).
+The code below shows the track pattern, first using \code{nabular()}, then imputing with \code{impute_lm()}.
+\code{label_shadow()} then adds a label to facilitate identifying missings:
 
@@
 
-Multiple missing or imputed values can be mapped to a graphical element in ggplot2 by setting the `color` or `fill` aesthetic in ggplot to `any_missing`, a result of the `add_label_shadow()` function.
+Multiple missing or imputed values can be mapped to a graphical element in ggplot2 by setting the \code{color} or \code{fill} aesthetic in ggplot to \code{any_missing}, a result of the \code{add_label_shadow()} function.
 (Figure \@ref(fig:track-impute-example)).
-Imputed values can also be compared to complete case data, grouping by `any_missing`, and then summarizing, similar to other dplyr summary workflows shown below in Table \@ref(tab:impute-summary-out), showing similarities and differences in imputation methods.
+Imputed values can also be compared to complete case data, grouping by \code{any_missing}, and then summarizing, similar to other \pkg{dplyr} summary workflows shown below in Table \@ref(tab:impute-summary-out), showing similarities and differences in imputation methods.
 
@@
 
-Missing values are often ignored when plotting data - which is why the data visualization software, MANET, was named and is an acronym corresponding to "Missings Are Now Equally Treated" [@Unwin1996].
+Missing values are often ignored when plotting data - which is why the data visualization software, \pkg{MANET}, was named and is an acronym corresponding to "Missings Are Now Equally Treated" [@Unwin1996].
 However, plots can help to identify the type of missing value patterns, even those of MCAR, MAR or MNAR.
-Here we summarise how to systematically explore missing patterns visually, and define useful plots to make, relative to the nabular data structure.
+Here we summarise how to systematically explore missing patterns visually, and define useful plots to make, relative to the *nabular* data structure.
 
@@
 The first step is to get an overview of the extent of the missingness.
-Overview visualizations for variables and cases are provided with `gg_miss_var` and `gg_miss_case` (Figure \@ref(fig:gg-miss-case-var)A), drawing attention to the amount of missings, and ordering by missingness.
-The "airquality" dataset (from base R), is shown, and contains daily air quality measurements in New York, from May to September, 1973.
+Overview visualizations for variables and cases are provided with \code{gg_miss_var()} and \code{gg_miss_case()} (Figure \@ref(fig:gg-miss-case-var)A), drawing attention to the amount of missings, and ordering by missingness.
+The "airquality" dataset (from base \proglang{R}), is shown, and contains daily air quality measurements in New York, from May to September, 1973.
 We learn from Figure \@ref(fig:gg-miss-case-var)A-B, that two variables contain missings, approximately one third of observations have one missing value, and a tiny number of observations are missing across two variables.
@@
 
-The shadow matrix can be put into long form, allowing both the variables and cases to be displayed using a heatmap style visualization, with `vis_miss()` from \pkg{visdat} [@visdat] (Figure \@ref(fig:nabularfig)).
+The shadow matrix can be put into long form, allowing both the variables and cases to be displayed using a heatmap style visualization, with \code{vis_miss()} from \pkg{visdat} [@visdat] (Figure \@ref(fig:nabularfig)).
 This also provides numerical summaries of missingness in the legend, and for each column (Figure \@ref(fig:gg-miss-case-var)C-D).
@@
 Similar visualizations are available in other packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}.
-A key improvement is `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns and are named at the top, and each row is an observation.
+A key improvement is \code{vis_miss()} orients the visualization analogous to a regular data structure: variables form columns and are named at the top, and each row is an observation.
 Using \pkg{ggplot2}, as the foundation, makes the plot easily customizable.
@@
 The bottom right shows the combinations of missingness, the top panel shows the size of these combinations, and the bottom left shows missingness in each variable.
-This provides similar information to Figure \@ref(fig:gg-miss-case-var)A-D, but more clearly illustrating overlapping missingness, where 2 cases are missing together in variables Solar.R and Ozone.
+This provides similar information to Figure \@ref(fig:gg-miss-case-var)A-D, but more clearly illustrating overlapping missingness, where 2 cases are missing together in variables "Solar.R" and "Ozone".
 
@@
 **Imputing values below the range**.
-To visualize the amount of missings in each variable, the data is transformed into *nabular* form, then values are imputed below the range of data using `impute_below_all`.
+To visualize the amount of missings in each variable, the data is transformed into *nabular* form, then values are imputed below the range of data using \code{impute_below_all()}.
 Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone values on the right in green, and the histogram of missing ozone values on the left, in orange.
-The missings in Ozone are imputed and `Ozone_NA` is mapped to the fill aesthetic.
-Nabular data facilitates adding counts of missingness to a histogram, allowing examination of a variables' distribution of values, and also the magnitude of missings.
+The missings in Ozone are imputed and "Ozone_NA" is mapped to the fill aesthetic.
+*Nabular* data facilitates adding counts of missingness to a histogram, allowing examination of a variables' distribution of values, and also the magnitude of missings.
 
@@
 The distribution of a variable can be shown according to the missingness of another variable.
-The shadow matrix part of nabular is used to handle the faceting, and color mapping.
+The shadow matrix part of *nabular* is used to handle the faceting, and color mapping.
 Figure \@ref(fig:impute-shift-histogram) shows the values of temperature when ozone is present, and missing, using a faceted histogram (B), and an overlaid density (C).
@@
 Using *nabular* data identifies imputed values, and color makes missingness pre-attentive [@treisman1985].
-The steps of imputing and coloring are combined into `geom_miss_point()`.
+The steps of imputing and coloring are combined into \code{geom_miss_point()}.
 Figure \@ref(fig:geom-miss) shows a mostly uniform spread of missing values for Solar.R and Ozone.
-As `geom_miss_point()` is a defined \pkg{ggplot2} geometry, it works with features such as faceting and mapping other variables to graphical aesthetics.
+As \code{geom_miss_point()} is a defined \pkg{ggplot2} geometry, it works with features such as faceting and mapping other variables to graphical aesthetics.
 
@@
 They transform variables to the same scale, ranging between 0 and 1.
-The `oceanbuoys` dataset from \pkg{naniar} is used for this visualization, containing measurements of sea and air temperature, humidity, and east west and north south wind speeds.
+The \code{oceanbuoys} dataset from \pkg{naniar} is used for this visualization, containing measurements of sea and air temperature, humidity, and east west and north south wind speeds.
 Data was collected in 1993 and 1997, to understand and predict El Niño and El Niña.
-Figure \@ref(fig:parallel-cord-plot) is a parallel coordinate plot of `oceanbuoys`, with missing values imputed to be 10% below the range, and values colored according to whether humidity was missing (`humidity_NA`).
+Figure \@ref(fig:parallel-cord-plot) is a parallel coordinate plot of \code{oceanbuoys}, with missing values imputed to be 10% below the range, and values colored according to whether humidity was missing ("humidity_NA").
 Figure \@ref(fig:parallel-cord-plot) shows humidity is missing at low air and sea temperatures, and humidity is missing in one year, and one location.
@@
 The overall number, proportion, or percent of missing values in a dataset should be simple to calculate.
-`naniar` provides the functions `n_miss`, `prop_miss` and `pct_miss`, as well as their complements.
-Summaries for variables and cases are made by appending `_case` or `_var` to these summaries.
+\pkg{naniar} provides the functions \code{n_miss()}, \code{prop_miss()} and \code{pct_miss()}, as well as their complements.
+Summaries for variables and cases are made by appending \code{_case} or \code{_var} to these summaries.
 An overview is shown in Table \@ref(tab:n-prop-pct-miss-complete).
@@
 Another useful approach is to tabulate the frequency of missing values for each variable or case; that is, the number of times there are zero missings, one missing, two, and so on.
-These summaries and tabulations are shown for variables in Tables \@ref(tab:miss-var-summary) and \@ref(tab:miss-var-table), and implemented with `miss_var_summary` and `miss_var_table`.
-Case-wise (row-wise) summaries and tabulations are implemented with `miss_case_summary` and `miss_case_table`.
-
-These summaries order rows by the number of missings (`n_miss`), to show the most missings at the top.
-The number of missings across a repeating span, or finding "runs" or "streaks" of missings in given variables can also be useful for identifying missingness patterns, and are implemented with `miss_var_span`, and `miss_var_run`.
+These summaries and tabulations are shown for variables in Tables \@ref(tab:miss-var-summary) and \@ref(tab:miss-var-table), and implemented with \code{miss_var_summary()} and \code{miss_var_table()}.
+Case-wise (row-wise) summaries and tabulations are implemented with \code{miss_case_summary()} and \code{miss_case_table()}.
+
+These summaries order rows by the number of missings (\code{n_miss()}), to show the most missings at the top.
+The number of missings across a repeating span, or finding "runs" or "streaks" of missings in given variables can also be useful for identifying missingness patterns, and are implemented with \code{miss_var_span()}, and \code{miss_var_run()}.
 
@@
 It is useful to explore summaries and tabulations within groups of a dataset.
-`naniar` works with `dplyr`'s `group_by` operator to produce grouped summaries, which work well with the "pipe" operator.
+\pkg{naniar} works with \pkg{dplyr}'s \code{group_by()} operator to produce grouped summaries, which work well with the "pipe" operator.
 The code and Table \@ref(tab:group-miss-var-summary) below show an example of missing data summaries for airquality, grouped by month.
@@
 
-Figure \@ref(fig:housing-upset) shows missingness patterns with an `upset` plot [@Conway2017], displaying 8 intersecting sets of missing variables.
+Figure \@ref(fig:housing-upset) shows missingness patterns with an "upset" plot [@Conway2017], displaying 8 intersecting sets of missing variables.
 Two patterns stand out: two, and five variables missing, providing further evidence of the missingness patterns seen in Figures \@ref(fig:housing-miss-case-var) and \@ref(fig:housing-upset).
@@
 Then, a classification and regression trees (CART) model is applied to predict these missingness clusters using the remaining variables [@Tierney2015; @Barnett2017].
-Two clusters of missingness are identified and predicted using all variables in the dataset with the CART package `rpart` [@rpart], and plotted using the `rpart.plot` package [@rpart-plot].
+Two clusters of missingness are identified and predicted using all variables in the dataset with the CART package \pkg{rpart} [@rpart], and plotted using the \pkg{rpart.plot} package [@rpart-plot].
 Importance scores reveal the following variables as most important in predicting missingness: "rooms", "price", "suburb", "council area", "distance", and "region name".
@@
 
-The `naniar` and `visdat` packages implement the methods discussed in the paper, building on existing tidy tools and strike a compromise between automation and control, making analysis efficient, readable, but not overly complex.
+The \pkg{naniar} and \pkg{visdat} packages implement the methods discussed in the paper, building on existing tidy tools and strike a compromise between automation and control, making analysis efficient, readable, but not overly complex.
 Each tool has clear intent and effects - summarising, plotting or generating data or augmenting data in some way.
@@
 Large data may need special handling, and additional features like efficient storage of purely imputed values and lazy evaluation.
-Special missing value codes could be improved by creating special classes, or expanding low level representation of `NA` at the source code level.
+Special missing value codes could be improved by creating special classes, or expanding low level representation of \code{NA} at the source code level.
 
@@
 
-The authors would like to thank Miles McBain, for his key contributions and discussions on the `naniar` package, in particular for helping implement `geom_miss_point`, and for his feedback on ideas, implementations, and names.
-We also thank Colin Fay for his contributions to the `naniar` package, in particular for his assistance with the `replace_with_na` functions.
+The authors would like to thank Miles McBain, for his key contributions and discussions on the \pkg{naniar} package, in particular for helping implement \code{geom_miss_point()}, and for his feedback on ideas, implementations, and names.
+We also thank Colin Fay for his contributions to the \pkg{naniar} package, in particular for his assistance with the \code{replace_with_na()} functions.
 We also thank Earo Wang and Mitchell O'Hara-Wild for the many useful discussions on missing data and package development, and for their assistance with creating elegant approaches that take advantage of the tidy syntax.
-We would also like to thank those who contributed pull requests and discussions on the `naniar` package, in particular Jim Hester and Romain François for improving the speed of key functions, Ross Gayler for discussion on special missing values, and Luke Smith for helping `naniar` be more compliant with `ggplot2`.
+We would also like to thank those who contributed pull requests and discussions on the \pkg{naniar} package, in particular Jim Hester and Romain François for improving the speed of key functions, Ross Gayler for discussion on special missing values, and Luke Smith for helping \pkg{naniar} be more compliant with \pkg{ggplot2}.
 We would also like to thank Amelia McNamara for discussions on the paper.
@@
 # References
-
-```{r demonstrate-impute-below, eval = FALSE}
-airquality %>%
-  impute_below_at(vars(Ozone)) %>%
-  select(Ozone, Solar.R) %>%
-  head()
-```
-
-```{r impute-vector, eval = FALSE}
-
-impute_mean(airquality$Ozone) %>% head()
-
-impute_mean_at(airquality, .vars = vars(Ozone)) %>% head()
-
-impute_mean_if(airquality, .predicate = is.integer) %>% head()
-
-impute_mean_all(airquality) %>% head()
-
-```

### 2020-09-08 paper/review/response-to-review.Rmd
finish reviewer comments draft

@@
 
-> I believe this has been resolved with the latest update to rticles
+> I believe this has been resolved with the latest update to rticles.
 
@@
 
-> Instances of `e.g.` have been replaced with `e.g.,`. No cases of `i.e.` were found.
+> Instances of `e.g.` have been replaced with `e.g.,`.
 
@@
 
-> I'm not sure it is possible to control the caption output of a kable Table, I will look into this.
+> Captions are below every figure, and for each table these will be moved manually in the LaTeX output.
 
@@
 
->
+> Need to double check about headings here for tables - Di should I change things like "n_miss" to "Num. Missing" ?
 
@@
 	- \proglang, \pkg and \code have been used for highlighting throughout the paper (including titles and references), except where explicitly escaped.
+
+> This has been done except for titles and references.
 

### 2020-09-08 paper/tidy-missing-data-paper.Rmd
finish reviewer comments draft

@@
 ```{r warning, fig.cap = "How ggplot2 behaves when displaying missing values.
-A warning message is displayed, but missing values are not shown in the plot", fig.height = 4, fig.width = 8, out.width = "75%", warning = FALSE, fig.show = 'hold'}
+A warning message is displayed, but missing values are not shown in the plot.", fig.height = 4, fig.width = 8, out.width = "75%", warning = FALSE, fig.show = 'hold'}
 p1 <- ggplot(oceanbuoys,
@@
 
-### The tidyverse {#the-tidyverse}
+### The \pkg{tidyverse} {#the-tidyverse}
 
@@
 Values are colored by missingness of humidity.
-Humidity is missing for low air and sea temperatures, and is missing for one year and one location.
-"}
+Humidity is missing for low air and sea temperatures, and is missing for one year and one location."}
 
@@
 ```{r tidy-coefs, fig.height = 2, fig.width = 7, fig.cap = "The coefficient estimate for the number of rooms varies according to the imputed dataset.
-Complete case dataset produced lower coefficients, compared to imputed datasets", out.width = "100%"}
+Complete case dataset produced lower coefficients, compared to imputed datasets.", out.width = "100%"}
 

### 2020-09-11 paper/review/response-to-review.Rmd
minor adornments

@@
 o Please make sure that the files needed to replicate all code/examples within the manuscript are included in a standalone replication script.
+
+Other notes
+
+- Check references are correctly referred using \pkg and \proglang
+- Check references are correctly referred to in text
+- Read over actual references compiled in the PDF - e.g., Adrian Barnett's article has a random string at the end.
+- Check quotes (" and ') are facing the right way
+- Don't start a sentence with a capital letter for pkgs.
+- Fix tables 4 and 5, ensure all appropriate tables have
+- Check that text overflows correctly.

### 2020-09-11 paper/tidy-missing-data-paper.Rmd
minor adornments

@@
   formatted: [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
-  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
+  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
 preamble: >
@@
 
-Features of tidy data were formally described in [@Wickham2014], and were discussed in terms of their importance for data science by [@Donoho2017], and tools for data analysis.
-Tidy data is defined by [@r4ds] as:
+Features of tidy data were formally described in @Wickham2014, and were discussed in terms of their importance for data science by @Donoho2017, and tools for data analysis.
+Tidy data is defined by @r4ds as:
 
@@
 
-### The \pkg{tidyverse} {#the-tidyverse}
+### The tidyverse {#the-tidyverse}
 

### 2020-09-17 paper/tidy-missing-data-paper.Rmd
tinker with YAML headings

@@
   plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
-preamble: >
-  \usepackage{amsmath}
-  \usepackage[english]{babel}
 output:
@@
     includes:
-      in_header: header.tex
+      in_preamble: preamble.tex
+     #in_header: header.tex
     keep_tex: yes
-    citation_package: 'natbib'
+    #citation_package: 'natbib'
 bibliography: tidy-missing-data-paper.bib
+cls: jss.cls
 ---
@@
 The majority of missingness is in selected cases and variables.", fig.height = 4, fig.width = 8, out.width = "100%"}
-
+```
+
+```{r housing-miss-case-var, fig.cap = "The amount of missings in variables (A) and cases (B) for Melbourne housing data.
+(A) Build area and year built have more than 50\\% missing, and car, bathroom, bedroom2 and longitude and latitude have about 25\\% missings.
+(B) Cases are missing 5 - 50\\% of values.
+The majority of missingness is in selected cases and variables.", fig.height = 4, fig.width = 8, out.width = "100%"}
 q1 <- gg_miss_var(housing, show_pct = TRUE) + labs(tag = "A")

### 2020-09-21 paper/tidy-missing-data-paper.Rmd
paper now compiles :tada:

--- 2020-09-08 paper/tidy-missing-data-paper.Rmd
@@
   # For running headers, if needed
-  short:     "Explore missingness with \\pkg{naniar}"
+  short:     "Explore missingness with naniar"
 abstract: >
@@
   # at least one keyword must be supplied
-  formatted: [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
+  formatted: [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
   plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
@@
 In particular, there is little guidance on how to handle missing values in a data analysis workflow.
-Most analysis and visualization software simply drop missing values when making a plot, although some (\pkg{ggplot2}) provide a warning (Figure \@ref(fig:warning)).
+Most analysis and visualization software simply drop missing values when making a plot, although some (ggplot2) provide a warning (Figure \@ref(fig:warning)).
 

### 2020-09-21 paper/tidy-missing-data-paper.Rmd
still compiles now

--- 2020-09-11 paper/tidy-missing-data-paper.Rmd
@@
   formatted: [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
-  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
+  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
 preamble: >
@@
 
-Features of tidy data were formally described in @Wickham2014, and were discussed in terms of their importance for data science by @Donoho2017, and tools for data analysis.
-Tidy data is defined by @r4ds as:
+Features of tidy data were formally described in [@Wickham2014], and were discussed in terms of their importance for data science by [@Donoho2017], and tools for data analysis.
+Tidy data is defined by [@r4ds] as:
 

### 2020-09-21 paper/tidy-missing-data-paper.Rmd
still compiles once adding in some \pkg

@@
 In particular, there is little guidance on how to handle missing values in a data analysis workflow.
-Most analysis and visualization software simply drop missing values when making a plot, although some (ggplot2) provide a warning (Figure \@ref(fig:warning)).
-
-```{r warning, fig.cap = "How ggplot2 behaves when displaying missing values.
+Most analysis and visualization software simply drop missing values when making a plot, although some (\pkg{ggplot2}) provide a warning (Figure \@ref(fig:warning)).
+
+```{r warning, fig.cap = "How \\pkg{ggplot2} behaves when displaying missing values.
 A warning message is displayed, but missing values are not shown in the plot.", fig.height = 4, fig.width = 8, out.width = "75%", warning = FALSE, fig.show = 'hold'}

### 2020-09-21 paper/tidy-missing-data-paper.Rmd

@@
   formatted: [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
-  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
-preamble: >
-  \usepackage{amsmath}
-  \usepackage[english]{babel}
+  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
 output:

### 2020-09-21 paper/tidy-missing-data-paper.Rmd
add bib file with no JSS

@@
     #citation_package: 'natbib'
-bibliography: tidy-missing-data-paper.bib
+# bibliography: tidy-missing-data-paper.bib
+bibliography: "tidy-missing-data-paper-no-jss.bib"
 cls: jss.cls

### 2020-09-21 paper/tidy-missing-data-paper.Rmd
compilation adventures - removing duplicate code chunk

--- 2020-09-21 paper/tidy-missing-data-paper.Rmd
@@
     keep_tex: yes
-    #citation_package: 'natbib'
-bibliography: tidy-missing-data-paper.bib
+    citation_package: 'natbib'
+# bibliography: tidy-missing-data-paper.bib
+bibliography: "tidy-missing-data-paper-no-jss.bib"
 cls: jss.cls
@@
 The majority of missingness is in selected cases and variables.", fig.height = 4, fig.width = 8, out.width = "100%"}
-```
-
-```{r housing-miss-case-var, fig.cap = "The amount of missings in variables (A) and cases (B) for Melbourne housing data.
-(A) Build area and year built have more than 50\\% missing, and car, bathroom, bedroom2 and longitude and latitude have about 25\\% missings.
-(B) Cases are missing 5 - 50\\% of values.
-The majority of missingness is in selected cases and variables.", fig.height = 4, fig.width = 8, out.width = "100%"}
 q1 <- gg_miss_var(housing, show_pct = TRUE) + labs(tag = "A")

### 2020-09-21 paper/tidy-missing-data-paper.Rmd
paper now compiles! Missing preamble in YAML which didn't load pacakge amsmath properly

@@
   formatted: [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
-  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
+  plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
+preamble: >	
+  \usepackage{amsmath}	
+  \usepackage[english]{babel}
 output:

### 2020-09-21 paper/tidy-missing-data-paper.Rmd
final minor changes

@@
     includes:
-      in_preamble: preamble.tex
-     #in_header: header.tex
+     # in_preamble: preamble.tex
+      in_header: header.tex
     keep_tex: yes

### 2020-09-22 paper/tidy-missing-data-paper.Rmd

--- 2020-09-21 paper/tidy-missing-data-paper.Rmd
@@
   \usepackage[english]{babel}
+  \usepackage{float}
+  \usepackage{booktabs}
 output:
@@
     base_format: rticles::jss_article
-    includes:
-      in_preamble: preamble.tex
-     #in_header: header.tex
     keep_tex: yes

### 2020-09-24 paper/tidy-missing-data-paper.Rmd
more minor changes to include pkg and proglang

@@
 Together, these provide a connected framework for handling, exploring, and imputing missing values.
-These methods are available in the R package \pkg{naniar}.
+These methods are available in the \proglang{R} package \pkg{naniar}.
 keywords:
   # at least one keyword must be supplied
-  formatted: [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
+  formatted: [statistical computing, statistical graphics, data science, data visualization, "\\pkg{tidyverse}", data pipeline, "\\proglang{R}"]
   plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
@@
 This paper is organized in the following way.
-Section (\@ref(background)) provides the background to tidy data principles and tools (\@ref(tidy-data-concepts)) and missing data representations (\@ref(missing-data-rep-dep)).
+Section \@ref(background) provides the background to tidy data principles and tools \@ref(tidy-data-concepts) and missing data representations (\@ref(missing-data-rep-dep)).
 Section \@ref(existing-software) summarises existing software for handling missing data.
@@
 
-Data manipulation is made input- and output-tidy with R packages \pkg{dplyr} and \pkg{tidyr} [@dplyr; @tidyr].
+Data manipulation is made input- and output-tidy with \proglang{R} packages \pkg{dplyr} and \pkg{tidyr} [@dplyr; @tidyr].
 These provide the five "verbs" of data manipulation: data reshaping, sorting, filtering, transforming, and aggregating.
@@
 Modelling tools work well with tidy data, as they have a clear mapping from variables in the data to the formula for a model.
-For example in R, y regressed on x and z is: \code{lm(y ~ x + z)}.
+For example in \proglang{R}, y regressed on x and z is: \code{lm(y ~ x + z)}.
 Modelling tools are input tidy, but their output is always messy - it is not in the right format for subsequent steps in analysis.
@@
 These diverse approaches allows for imputing with semi-continuous, continuous, count, and categorical data.
-VIM identifies imputed cases by adding an indicator variable with a suffix "_imp".
-So "Var1" has a sibling column, "Var1_imp", with values TRUE or FALSE indicating imputation.
+VIM identifies imputed cases by adding an indicator variable with a suffix "\code{_imp}".
+So, "\code{Var1}" has a sibling column, "\code{Var1_imp}", with values TRUE or FALSE indicating imputation.
 \pkg{VIM} also has a variety of visualization methods, discussed in \@ref(exploration).
-\pkg{simputation} provides an interface to imputation methods from VIM, in addition providing hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977].
+\pkg{simputation} provides an interface to imputation methods from \pkg{VIM}, in addition providing hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977].
 \pkg{simputation} provides a consistent formula interface for all imputation methods, and always returns a dataframe with the updated imputations.
@@
 It uses the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm [@Honaker2010].
-\pkg{norm} [@norm], provides multiple imputation using EM for multivariate normal data, drawing from methods in the NORM software [@schafer-norm].
+\pkg{norm} [@norm], provides multiple imputation using EM for multivariate normal data, drawing from methods in the \proglang{NORM} software [@schafer-norm].
 \pkg{norm} does not provide a framework for tracking missing values, instead providing tools for making inference from multiple imputation.
@@
 
-```{r gg-box-na, fig.height = 3, fig.width = 9, out.width = "100%", fig.align = "center", fig.cap = "ggplot2 provides different visualizations depending on what type of data has missing values for data of student test scores in school year.
+```{r gg-box-na, fig.height = 3, fig.width = 9, out.width = "100%", fig.align = "center", fig.cap = "\\pkg{ggplot2} provides different visualizations depending on what type of data has missing values for data of student test scores in school year.
 (A) Data is complete; (B) Missings are only in year - an additional 'NA' boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and year, additional missing information is shown.
@@
 
-Applying tidyverse principles has the potential to clarify missing data exploration, visualization, and imputation.
+Applying \pkg{tidyverse} principles has the potential to clarify missing data exploration, visualization, and imputation.
 This section discusses how these principles are applied for data structures (\@ref(data-structure)), common operations (verbs) (\@ref(verbs)), graphics (\@ref(graphics)), and data summaries (\@ref(num-sum)).
@@
 Namely, the values 0 and 1 can be confusing representations of missing values, since it is not clear if 0 indicates an absence of observation, or the presence of a missing value.
-
-<!-- since 0 and 1 could interchangably indicate the absence (0) of observation, or the presence (1) of a missing value. -->
 

### 2020-09-24 paper/tidy-missing-data-paper.Rmd
improving table header

@@
 This paper is organized in the following way.
-Section \@ref(background) provides the background to tidy data principles and tools \@ref(tidy-data-concepts) and missing data representations (\@ref(missing-data-rep-dep)).
+Section \@ref(background) provides the background to tidy data principles and tools (\@ref(tidy-data-concepts)) and missing data representations (\@ref(missing-data-rep-dep)).
 Section \@ref(existing-software) summarises existing software for handling missing data.
@@
 
-### The tidyverse {#the-tidyverse}
+### Tidy data and tidy tools {#the-tidyverse}
 
@@
 table_of_fun <- tibble::tibble(
-  "Missing Function" = c("n_miss", 
+  "Missing function" = c("n_miss", 
                          "prop_miss", 
@@
                          "pct_miss_var"),
-  "missing value" = c(n_miss(airquality),
+  "Missing value" = c(n_miss(airquality),
                       prop_miss(airquality),
@@
                           "pct_complete_var"),
-  "complete value" = c(n_complete(airquality), 
+  "Complete value" = c(n_complete(airquality), 
                        prop_complete(airquality), 

### 2020-09-30 paper/review/response-to-review.Rmd
still compiles now
tidying up some of the tables

@@
 
-Missing data can be explored and visualised in other software such as \proglang{STATA}, \proglang{SAS}, and \proglang{SPSS}.
+> Missing data can be explored and visualised in other software such as \proglang{STATA}, \proglang{SAS}, and \proglang{SPSS}.
 They provide basic table frequencies of missing data, and basic visualisations on missingness patterns.
@@
 
-> I have updated to the latest rticles package and checked that the JSS-specific markup has been fixed in the latest submission.
+> I have updated to the latest rticles package and checked that the JSS-specific markup has been fixed in the latest submission. Many instances of \proglang, \pkg, and \code have been used now
 

### 2020-09-30 paper/tidy-missing-data-paper.Rmd
tidying up some of the tables

@@
 
-<!-- (NOTE: Is the last sentence OK?) -->
-
-<!-- (NOTE: trying to come up with a sentence that leads from this section to the next) -->
-
 # Existing Software {#existing-software}
@@
 
-<!-- Note: Do you think this paragraph needs to be here, about the use of _nabular_ data structure? -->
-
 Although a binary missingness matrix, $B$, could be generated during analysis instead of using a *nabular* data structure, there are key advantages to using *nabular* data.
@@
 Special missing values are not a new idea, and have been implemented in other statistical programming languages, \proglang{SPSS}, \proglang{SAS}, and \proglang{STATA}.
-These typically represent missing values as a full-stop, `.`, and record special missing values as `.a` - `.z`.
+These typically represent missing values as a full-stop, ".", and record special missing values as ".a" - ".z".
 These special values from these languages break the tidy principle of one column having one type of value, as they record both the value, and the multivariate missingness state.
@@
                .funs = lst(min, mean, median, max))  %>% 
-  kable(caption = "Summarising values of imputed vs non imputed values.
-Comparing imputed values (denoted as 'Missing', since they were previously missing), the mean and median values are similar, but the minimum and maximum values are very different.")
+  kable(caption = "Output of \\pkg{dplyr} summary statistics of imputed vs non imputed values for the variable 'Ozone'.
+The 'any\\_missing' column denotes imputed values ('Missing', since they were previously missing), and non-imputed values ('Not Missing').
+The mean and median values are similar, but the minimum and maximum values are very different.")
 ```
@@
   knitr::kable(
-    caption = "\\texttt{miss\\char`_var\\char`_table} tabulates the amount of missing data in each variable in airquality.
+    caption = "The output of \\code{miss\\char`_var\\char`_table(airquality)}, tabulating the amount of missing data in each variable in airquality.
 This shows the number of variables with 0, 7, and 37 missings, and the percentage of variables with those amounts of missingness.

### 2020-10-01 paper/tidy-missing-data-paper.Rmd
fixing a couple more table captions

@@
     citation_package: 'natbib'
-# bibliography: tidy-missing-data-paper.bib
-bibliography: "tidy-missing-data-paper-no-jss.bib"
+bibliography: "tidy-missing-data-paper.bib"
 cls: jss.cls
@@
   slice(1:10) %>%
-  knitr::kable(caption = "\\texttt{miss\\char`_var\\char`_summary} combined with \\texttt{group\\char`_by} provides a grouped summary of the missingness in each variable, for each Month of the airquality dataset.
+  knitr::kable(caption = "Output of \\code{airquality \\%>\\% group\\char`_by(Month) \\%>\\% miss\\char`_var\\char`_summary()} provides a grouped summary of the missingness in each variable, for each month of the airquality dataset.
 Only the first 10 rows are shown.
@@
 knitr::kable(list(t1, t2),
-             caption = "Tabulating missingness for variables (left) and cases (right) to understand missingness patterns.
-14 variables have 0 - 3 missings, 6 variables have 6000 - 9000 missings, and 2 variables have 15 - 16,000 missings.
+             caption = "Summary tables to help understand missingness patters.
+Output of \\code{miss\\_var\\_table(housing)} (left), tabulating missingness for variables, and output of \\code{miss\\_case\\_table(housing)}.
+There are 14 variables with 0 - 3 missings, 6 variables have 6000 - 9000 missings, and 2 variables have 15 - 16,000 missings.
 About 30\\% of cases have no missings, 45\\% of cases have 1 - 6 missings, and about 23\\% of cases have 8 or more missings.
@@
 The analyst can use the data structures to decide on appropriate visualization for their problem.
-The data structures could also be used to support interactive graphics, in the manner of MANET and ggobi.
+The data structures could also be used to support interactive graphics, in the manner of \proglang{MANET} and \proglang{ggobi}.
 Linking the plots (via linked brushing) could facilitate exploration of missingness, and could be implemented with \pkg{plotly} [@plotly] for added interactivity.

### 2020-10-06 paper/tidy-missing-data-paper.Rmd
JSS review

@@
     citation_package: 'natbib'
-bibliography: "tidy-missing-data-paper.bib"
+bibliography: "tidy-missing-data-zotero.bib"
 cls: jss.cls
@@
 Defining tidy data and tidy tools has resulted in a growing set of packages known collectively as the \pkg{tidyverse} [@tidyverse].
-These are constructed to share similar principles in their design and behavior, and cover the breadth of an analysis - from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @tidyverse; @Tidyverse-Manifesto].
+These are constructed to share similar principles in their design and behavior, and cover the breadth of an analysis - from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @tidyverse].
 This has led to more tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr}, dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} [@readr; @readxl; @haven; @stringr; @lubridate; @purrr].
@@
 
-There are many ways each value can be missing, we adopt the notation used in [@VanBuuren2012].
+There are many ways each value can be missing, we adopt the notation used in [@vanBuuren2018].
 The information in $B$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR).
@@
 So, "\code{Var1}" has a sibling column, "\code{Var1_imp}", with values TRUE or FALSE indicating imputation.
-\pkg{VIM} also has a variety of visualization methods, discussed in \@ref(exploration).
+\pkg{VIM} also has a variety of visualization methods, discussed in Section \@ref(exploration).
 \pkg{simputation} provides an interface to imputation methods from \pkg{VIM}, in addition providing hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977].
@@
 The primary focus of most missing data packages is making inferences, and exploring imputed values, not on exploring relationships in missing values, and identifying possible patterns.
-Texts covering the exploration phase of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools [@VanBuuren2012]; this is inefficient.
+Texts covering the exploration phase of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools [@vanBuuren2018]; this is inefficient.
 Methods for exploring missing values are primarily covered in literature on interactive graphics [@Swayne1998; @Unwin1996; @Cook2007], and are picked up again in a discussion of a graphical user interface [@Cheng2015].
@@
 
-The \pkg{MANET} (Missings Are Now Equally Treated) software [@Unwin1996] focussed on multivariate categorical data, with missingness explicitly added as a category.
-\pkg{MANET} also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot.
-The \pkg{MANET} software is no longer maintained and cannot be installed.
-The approach of [@Swayne1998] in the software \pkg{XGobi}, further developed in \pkg{ggobi} [@Cook2007], focussed on multivariate quantitative data.
-Missingness is incorporated into plots in \pkg{ggobi} by setting them to be 10% below the minimum value.
+The \proglang{MANET} (Missings Are Now Equally Treated) software [@Unwin1996] focussed on multivariate categorical data, with missingness explicitly added as a category.
+\proglang{MANET} also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot.
+The \proglang{MANET} software is no longer maintained and cannot be installed.
+The approach of [@Swayne1998] in the software \proglang{XGobi}, further developed in \proglang{ggobi} [@Cook2007], focussed on multivariate quantitative data.
+Missingness is incorporated into plots in \proglang{ggobi} by setting them to be 10% below the minimum value.
 
@@
 A useful template common in missing data literature, is the $B$ matrix, where 0 and 1 indicate not missing and missing, respectively.
-This matrix was used to explore missing values in the interactive graphics library \pkg{XGobi}, called a "missing value shadow", or "shadow matrix", defined as a copy of the original data with indicator values of missingness.
+This matrix was used to explore missing values in the interactive graphics library \proglang{XGobi}, called a "missing value shadow", or "shadow matrix", defined as a copy of the original data with indicator values of missingness.
 The shadow matrix could be interactively linked to the data.
@@
 
-Using additional columns to represent missingness information follows best practices for data organization, described in [@Ellis2017] and [@Broman2017]: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column.
+Using additional columns to represent missingness information follows best practices for data organization, described in [@Ellis2018] and [@Broman2018]: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column.
 Here they suggest to indicate censored data with an extra variable called "VariableNameCensored", which would be TRUE if censored, otherwise FALSE.
@@
 Encoding special missing values is achieved by defining logical conditions and suffixes.
-This is implemented with the \code{recode_shadow()} function in \pkg{naniar} (\@ref(verbs-recode)).
+This is implemented with the \code{recode_shadow()} function in \pkg{naniar} (Discussed in Section \@ref(verbs)).
 
@@
 Similar to \pkg{simputation}, each \code{impute_} function returns the data with values imputed.
-However, \pkg{naniar} does not use a formula syntax, instead each function has "scoped variants" \code{_all}, \code{_at} and \code{_if} as in \@ref(verbs-replace-with).
+However, \pkg{naniar} does not use a formula syntax, instead each function has "scoped variants" \code{_all}, \code{_at} and \code{_if}.
 \code{impute_} functions with no scoped variant, (\code{impute_mean()}), will work on a single vector, but not a data.frame.
 One challenge with this approach is imputed value locations are not tracked.
-This issue is resolved with *nabular* data covered in Section \@ref(verbs-track).
+This issue is resolved with *nabular* data covered in the following section.
 
@@
 
-Missing values are often ignored when plotting data - which is why the data visualization software, \pkg{MANET}, was named and is an acronym corresponding to "Missings Are Now Equally Treated" [@Unwin1996].
+Missing values are often ignored when plotting data - which is why the data visualization software, \proglang{MANET}, was named and is an acronym corresponding to "Missings Are Now Equally Treated" [@Unwin1996].
 However, plots can help to identify the type of missing value patterns, even those of MCAR, MAR or MNAR.
@@
   knitr::kable(
-    caption = "\\texttt{miss\\char`_var\\char`_summary} provides the number and percent of missings in each variable in airquality. Only ozone and solar radiation have missing values.",
+    caption = "\\texttt{miss\\char`_var\\char`_summary(airquality)} provides the number and percent of missings in each variable in airquality. Only ozone and solar radiation have missing values.",
     digits = 1)
@@
 Values are imputed stepwise in ascending order of missingness.
-The track missings pattern is applied (described in \@ref(verbs-track)), to assess imputed values.
+The track missings pattern is applied (described in \@ref(verbs)), to assess imputed values.
 Imputed datasets are compared on their performance in a model predicting log house price for 4 variables (Figure \@ref(fig:imputed-by-model)).
@@
 The methodology described in this paper can be used in conjunction with other approaches to understand multivariate missingness dependencies (e.g., decision trees [@Tierney2015], latent group analysis [@Barnett2017], and PCA [@FactoMineR]).
-Evaluating imputed values using a testing framework like [@VanBuuren2012] is also supported.
+Evaluating imputed values using a testing framework like [@vanBuuren2018] is also supported.
 

### 2020-10-06 paper/tidy-missing-data-paper.R

@@ -1,0 +1,870 @@
+library(tidyverse) 
+library(gridExtra)
+library(ggthemes)
+library(png)
+library(scales)
+library(simputation)
+# if code doesn't work, use development version of naniar
+# install.packages("remotes")
+# remotes::install_github("njtierney/naniar")
+library(naniar) 
+library(knitr)
+library(lubridate)
+library(magick)
+library(rpart)
+library(visdat)
+library(patchwork)
+
+## ----warning, fig.cap = "How ggplot2 behaves when displaying missing values. A warning message is displayed, but missing values are not shown in the plot", fig.height = 4, fig.width = 8, out.width = "75%", warning = FALSE, fig.show = 'hold'----
+p1 <- ggplot(oceanbuoys,
+       aes(x = humidity,
+           y = air_temp_c)) +
+  geom_point(size = 0.5) +
+  theme(aspect.ratio = 1) + 
+  labs(x = "Humidity",
+       y = "Air Temperature (C)")
+
+p1 + grid::textGrob(
+  label = "Warning message:
+Removed 171 rows containing 
+missing values (geom_point).", 
+gp = grid::gpar(fontsize = 14,
+                fontfamily = "Courier",
+                col = "salmon", 
+                fontface = "bold"))
+
+  
+
+
+## ----make-school-scores-data, echo = FALSE-----------------------------------------
+
+school_scores <- data.frame(score = c(rnorm(20, 70, 10),
+                                 rnorm(20, 85, 7.5),
+                                 rnorm(20, 95, 5),
+                                 rnorm(20, 115, 10)),
+                       year = rep(c("1st", "2nd", "3rd", "4th"), each = 20)) %>%
+  mutate(score = as.numeric(score),
+         year = as.factor(year)) %>%
+  as_tibble()
+
+school_scores_year_miss <- school_scores %>%
+  mutate_at(.funs = function(x){
+    x[sample(c(TRUE, NA), 
+             prob = c(0.85, 0.150), 
+             size = length(x), 
+             replace = TRUE)]
+  },
+  .vars = "year")
+
+
+school_scores_score_miss <- school_scores %>%
+  mutate_at(.funs = function(x){
+    x[sample(c(TRUE, NA), 
+             prob = c(0.85, 0.150), 
+             size = length(x), 
+             replace = TRUE)]
+  },
+  .vars = "score")
+
+
+school_scores_both_miss <- school_scores %>%
[…]

### 2020-10-06 paper/review/response-to-review.Rmd
JSS re-submission

--- 2020-09-08 paper/review/response-to-review.Rmd
@@
 
-> I have updated to the latest rticles package and checked that the JSS-specific markup has been fixed in the latest submission.
+> I have updated to the latest rticles package and checked that the JSS-specific markup has been fixed in the latest submission. Instances of \proglang, \pkg, and \code have been added.
 
@@
 
-> I believe this has been resolved with the latest update to rticles.
+> This has been resolved with the latest update to rticles.
 
@@
 
-> Captions are below every figure, and for each table these will be moved manually in the LaTeX output.
+> Captions are below every figure, and for each table these have been moved manually in the LaTeX output.
 
@@
 
-> Need to double check about headings here for tables - Di should I change things like "n_miss" to "Num. Missing" ?
+> Row/Column headers have been changed to sentence style, where appropriate.
+Some tables represent the direct output of a function, for example, in Table 2, 4, and 5.
+This is noted in the table caption that it is the output of a computation.
 
@@
 
-o Journal of the Royal Statistical Society B (not: Journal of the Royal Statistical Society, Series B)
-	the EM Algorithm.” Journal of the Royal Statistical Society. Series B
-the EM Algorithm.” Journal of the Royal Statistical Society. Series B
-o Springer-Verlag (not: Springer)
-	and GGobi. 1st edition. Springer Publishing Company, Incorporated.
+o Journal of the Royal Statistical Society B (not: Journal of the Royal Statistical Society, Series B) the EM Algorithm.” 
+o Springer-Verlag (not: Springer) and GGobi. 1st edition. Springer Publishing Company, Incorporated.
+
+> These have been fixed.
 
 o Please make sure that all software packages are \cite{}'d properly.
+
+> These have been altered in the .bib file to include the proper \proglang{} and \pkg{} syntax.
+
 o All references should be in title style.
+
+> References are now in title style
+
 o See FAQ for specific reference instructions.
@@
 o Please make sure that the files needed to replicate all code/examples within the manuscript are included in a standalone replication script.
+
+> This has been provided in the "tidy-missing-data-paper.R" file

### 2021-05-19 paper/tidy-missing-data-paper.Rmd
Address reviewer comments

@@
 While there are times where the missing data mechanism is obvious, determining this is typically not straightforward.
+A straightforward case is where missing values are encoded in the data in a non standard format that still indicates missingness, such as "N.A.", "N/A", or "missing".
+However, identifying these amongst all data can be problematic as it is hard to know what you are looking for, like finding a needle in a haystack.
+A more challenging case is where missing values arise due to the increasing or decreasing values of other variables (including themselves).
+For example, in weather data, temperature values might go missing due to their own high temperature.
+Or similarly, other measurements go missing due to high temperatures.
+Identifying this requires exploring the data, and the analysis path to take is obvious, as missing values are often omitted from plots and other statistical summaries.
 To understand their structure and possible mechanisms, the analyst must explore the data with visualizations, summaries, and modelling, in an iterative fashion.
@@
 
-The primary focus of most missing data packages is making inferences, and exploring imputed values, not on exploring relationships in missing values, and identifying possible patterns.
+Missing data packages are focused primarily on making inferences, and exploring imputed values, not on exploring relationships in missing values, and identifying possible patterns.
 Texts covering the exploration phase of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools [@vanBuuren2018]; this is inefficient.
@@
 Additionally, data used to create the visualizations cannot be accessed, posing a barrier to further exploration.
+
+The \pkg{inspectdf} package provides tools for summarising, comparing, and visualising dataframes [@inspectdf].
+In particular the `inspect_na()` function provides a summary of missing values for variables.
+`inspect_cat()` summarises the levels of categories in each column in a dataset.
+The `show_plot()` function shows a higher lever overview of the contents of a dataframe and its categories, as well as missing values.
+All of these functions allow for comparing missing values across two dataframes.
+For example, comparing the missing values in two dataframes with the same column names.
 
@@
 Four features are added to the shadow matrix to facilitate analysis of missing data, illustrated in Figure \@ref(fig:nabularfig).
+These are briefly described below, and discussed in further detail in section \@ref(nabular-data).
 
@@
 This suffix is short and easy to remember during data analysis, and shifts the focus from the value of a variable, to its missingness state.
+Although this might break naming conventions for lower case "snake_case" variables (e.g., @Wickham2014), this distinguishes these columns from other columns.
 
 4.
-**Connectedness**: Binding the shadow matrix column-wise to the original data creates a single, connected, *nabular* data format, in sync with the data.
+**Connectedness**: Binding the shadow matrix column-wise to the original data as a factor creates a single, connected, nabular data format, in sync with the data, at the cost of a larger file.
 It is useful for visualization, summaries, and tracking imputed values, discussed in more detail in \@ref(nabular-data).
@@
 This means when *nabular* data are imputed, those imputed values can easily be identified in analysis ("var" vs "var_NA").
-*Nabular* data is not unlike classical data formats with quality or flag variables associated with each measured variable, e.g., Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].
+Imputing values on *nabular* data automatically tracks these imputations.
+Together these features make it possible to do tasks that are not possible by simply adding a binary missing matrix, $B$ during data analysis.
 
@@
 
-Although a binary missingness matrix, $B$, could be generated during analysis instead of using a *nabular* data structure, there are key advantages to using *nabular* data.
-Firstly, missing value labels "NA" and "!NA" are clearer than TRUE or FALSE.
-Secondly, special missing values cannot be added easily during analysis with a logical matrix.
-Finally, the logical matrix cannot capture which values are imputed if imputation has already taken place.
-Imputing values on *nabular* data automatically tracks these imputations.
-
+*Nabular* data is similar to classical data formats with quality or flag variables associated with each measured variable, e.g., Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].
 Using additional columns to represent missingness information follows best practices for data organization, described in [@Ellis2018] and [@Broman2018]: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column.
@@
 
-\pkg{naniar} does not reinvent the wheel for imputation, instead working with existing methods.
-However, \pkg{naniar} provides a few imputation methods to facilitate exploration and visualization: \code{impute_below()}, \code{impute_mean()}, and \code{impute_median()}.
+\pkg{naniar} does not reinvent the wheel by creating the need for customised imputation procedures.
+Instead, the \code{nabular} format means existing methods using imputation replacement in place now automatically track missing values.
+\pkg{naniar} provides a few imputation methods to facilitate exploration and visualization: \code{impute_below()}, \code{impute_mean()}, and \code{impute_median()}.
 While useful to explore structure in missingness, they are not recommended for use in analysis.
@@
 Output of \\code{miss\\_var\\_table(housing)} (left), tabulating missingness for variables, and output of \\code{miss\\_case\\_table(housing)}.
-There are 14 variables with 0 - 3 missings, 6 variables have 6000 - 9000 missings, and 2 variables have 15 - 16,000 missings.
+There are 13 variables with 0-3 missings, 6 variables have 6000-10000 missings, 2 variables have 15000 – 17000 missings.
 About 30\\% of cases have no missings, 45\\% of cases have 1 - 6 missings, and about 23\\% of cases have 8 or more missings.
@@
 Special missing value codes could be improved by creating special classes, or expanding low level representation of \code{NA} at the source code level.
+Similarly, list columns, present a challenge for analysis as their expansion into vectors in a dataframe is not always possible as they have different lengths, but should still be flagged for further analysis.
 

### 2021-05-19 README.Rmd
tidy up and add note on how to create the tarball to the README
update renv parts

@@
 This is the repository for the paper: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations" - you can find all the data and code to replicate the paper here.
+
+- Note that you need to have the development version of `knitr` and `rmarkdown`
+- Packages used described in `renv.lock` file
+- you can install the packages used in the lockfile by running:
+
+```r
+library(renv)
+restore()
+```
+
+This will install the packgages used in the lockfile, for this project. It should not impact your other local R installations.

### 2021-06-23 paper/tidy-missing-data-paper.Rmd
misc updates to paper to prepare for new merge

@@
   plain:     [statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R]
-preamble: >
-  \usepackage{amsmath}
-  \usepackage[english]{babel}
-  \usepackage{float}
-  \usepackage{booktabs}
 output:
@@
     citation_package: 'natbib'
+    includes:
+      in_header: "preamble.tex"
 bibliography: "tidy-missing-data-zotero.bib"
@@
 As data science has become a more solid field, theories and principles have developed to describe best practices.
-One such idea is 'tidy data,' which defines a clean, analysis-ready format that informs workflows converting raw data through a data analysis pipeline [@Wickham2014].
+One such idea is "tidy data", which defines a clean, analysis-ready format that informs workflows converting raw data through a data analysis pipeline [@Wickham2014].
 Another idea is the grammar of graphics, describing how to map data values into a visualization [@Wilkinson2012].
@@
   label = "Warning message:
-Removed 171 rows containing 
+removed 171 rows containing 
 missing values (geom_point).", 
@@
 
-> 1.  Each variable must have its own column.
-> 2.  Each observation must have its own row.
-> 3.  Each value must have its own cell.
+> 1.  each variable must have its own column.
+> 2.  each observation must have its own row.
+> 3.  each value must have its own cell.
 
@@
 Missing data tools can be used to perform imputations, missing data diagnostics, and data visualizations.
-However, these tools suffer the same problems as modelling: They use tidy input, but produce messy output - their output is challenging to integrate with other steps of data analysis.
+However, these tools suffer the same problems as modelling: they use tidy input, but produce messy output - their output is challenging to integrate with other steps of data analysis.
 The complex, often multivariate nature of imputation methods also makes makes them difficult to represent.
@@
 There are many ways each value can be missing, we adopt the notation used in [@vanBuuren2018].
-The information in $B$ can be used to arrive at three categories of missing values: Missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR).
+The information in $B$ can be used to arrive at three categories of missing values: missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR).
 The distribution of missing values in $b_{ij}$ can depend on the entire dataset, represented as $Y = (Y_{obs}, Y_{miss})$.
@@
 
-```{r gg-box-na, fig.height = 3, fig.width = 9, out.width = "100%", fig.align = "center", fig.cap = "\\pkg{ggplot2} provides different visualizations depending on what type of data has missing values for data of student test scores in school year.
-(A) Data is complete; (B) Missings are only in year - an additional 'NA' boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and year, additional missing information is shown.
+```{r gg-box-na, fig.height = 3, fig.width = 9, out.width = "100%", fig.align = "center", fig.cap = '\\pkg{ggplot2} provides different visualizations depending on what type of data has missing values for data of student test scores in school year.
+(A) Data is complete; (B) Missings are only in year - an additional "NA" boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and year, additional missing information is shown.
 The missingness category is only shown when there are missings in categorical variables such as year (plots (B) and (D)).
-In (C), no missingness information is given on the graphic, despite there being missings in score, and a warning message is displayed about the number of missing values omitted.", fig.show = "asis"}
+In (C), no missingness information is given on the graphic, despite there being missings in score, and a warning message is displayed about the number of missing values omitted.', fig.show = "asis"}
 
@@
 2.
-**Special missing values**: Building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA\_\<suffix\>".
+**Special missing values**: building on **missing value labels**, the values in the shadow matrix can be "special" missing values, indicated by "NA\_\<suffix\>".
 For example: "NA_instrument" uses a short label, "instrument", indicating instrument error resulting in missing values.
@@
 3.
-**Coordinated names**: Variable names in the shadow matrix gain a consistent short suffix, \"\_NA\", keeping names coordinated throughout analysis (Figure \@ref(fig:nabularfig)).
+**Coordinated names**: variable names in the shadow matrix gain a consistent short suffix, \"\_NA\", keeping names coordinated throughout analysis (Figure \@ref(fig:nabularfig)).
 It makes a clear distinction with "var_NA" being a random variable of the missingness of a variable, "var".
@@
 4.
-**Connectedness**: Binding the shadow matrix column-wise to the original data as a factor creates a single, connected, nabular data format, in sync with the data, at the cost of a larger file.
+**Connectedness**: binding the shadow matrix column-wise to the original data as a factor creates a single, connected, nabular data format, in sync with the data, at the cost of a larger file.
 It is useful for visualization, summaries, and tracking imputed values, discussed in more detail in \@ref(nabular-data).
@@
 
-```{r nabularfig, echo = FALSE, fig.cap = "The process of creating nabular data.
-Data transformed to shadow matrix, where values are either not missing or missing: '!NA' or 'NA'.
+```{r nabularfig, echo = FALSE, fig.cap = 'The process of creating nabular data.
+Data transformed to shadow matrix, where values are either not missing or missing: "!NA" or "NA".
 The shadow matrix can be converted to long form to create missingness summary plots.
@@
 Special missing values (such as -99) are identified as special missings, and values imputed and tracked.
-Nabular data can be used to identify imputations and explore data values alongside missings, providing a useful format for missing data exploration and analysis.", out.width = "100%", fig.align = "center"}
+Nabular data can be used to identify imputations and explore data values alongside missings, providing a useful format for missing data exploration and analysis.', out.width = "100%", fig.align = "center"}
 
@@
 *Nabular* data is similar to classical data formats with quality or flag variables associated with each measured variable, e.g., Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].
-Using additional columns to represent missingness information follows best practices for data organization, described in [@Ellis2018] and [@Broman2018]: (1) Keep one thing in a cell and (2) Describe additional features of variables in a second column.
+Using additional columns to represent missingness information follows best practices for data organization, described in [@Ellis2018] and [@Broman2018]: (1) keep one thing in a cell and (2) describe additional features of variables in a second column.
 Here they suggest to indicate censored data with an extra variable called "VariableNameCensored", which would be TRUE if censored, otherwise FALSE.
@@
 
-### scan: Searching for common missing value labels {#verbs-search}
+### scan: searching for common missing value labels {#verbs-search}
 
@@
   knitr::kable(
-    caption = "Table of the occurrences of the search '-99' in the data, 'dat\\_ms'. There is one occurrence if -99 in variables x and z.")
+    caption = 'Table of the occurrences of the search "-99" in the data, "dat\\_ms". There is one occurrence if -99 in variables x and z.')
     # booktabs = TRUE) 
@@
 
-### replace: Replacing values with missing values {#verbs-replace-with}
+### replace: replacing values with missing values {#verbs-replace-with}
 
@@
 
-### add: Adding missingness summary variables {#verbs-add-cols}
+### add: adding missingness summary variables {#verbs-add-cols}
 
@@
   `Adds column which:` = c(
-    "Contains the number missing values in a row",
-    "Contains the proportion of missing values in a row",
-    "Contains the missing value cluster"
+    "contains the number missing values in a row",
+    "contains the proportion of missing values in a row",
+    "contains the missing value cluster"
   )) %>% 
-    knitr::kable(caption = "Overview of the 'add' functions in naniar")
-
-```
-
-### shadow: Creating nabular data {#verbs-nabular}
+    knitr::kable(caption = 'Overview of the "add" functions in naniar')
+
+```
+
+### shadow: creating nabular data {#verbs-nabular}
 
@@
 
-### flag: Describing different types of missing values {#verbs-recode}
+### flag: describing different types of missing values {#verbs-recode}
 
@@
 
-### impute: Imputing values {#verbs-impute}
+### impute: imputing values {#verbs-impute}
 
@@
 
-### track: Shadow and impute missing values {#verbs-track}
+### track: shadow and impute missing values {#verbs-track}
 
@@
                .funs = lst(min, mean, median, max))  %>% 
-  kable(caption = "Output of \\pkg{dplyr} summary statistics of imputed vs non imputed values for the variable 'Ozone'.
-The 'any\\_missing' column denotes imputed values ('Missing', since they were previously missing), and non-imputed values ('Not Missing').
-The mean and median values are similar, but the minimum and maximum values are very different.")
+  kable(caption = 'Output of \\pkg{dplyr} summary statistics of imputed vs non imputed values for the variable "Ozone".
+The "any\\_missing" column denotes imputed values ("Missing", since they were previously missing), and non-imputed values ("Not Missing").
+The mean and median values are similar, but the minimum and maximum values are very different.')
 ```
@@
 
-Using information from \@ref(case-study-explore-pattern), the following variables are explored for features predicting missingness: "Land size", "latitude", "longitude", "bedroom2", "bathroom", "car", and "land size".
+Using information from \@ref(case-study-explore-pattern), the following variables are explored for features predicting missingness: "latitude", "longitude", "bedroom2", "bathroom", "car", and "land size".
 
@@
 
-<!-- Note: I think I'm basically there, but I'd like to revisit the first paragraph in this section and the last sentence -->
-
 This paper has described new methods for exploring, visualizing, and imputing missing data.

### 2021-06-24 README.Rmd
first changes to use targets

--- 2018-09-07 README.Rmd
@@
 This is the repository for the paper: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations" - you can find all the data and code to replicate the paper here.
+
+<!-- badges: start -->
+<!-- badges: end -->
+
+# To recreate the paper
+
+Use `targets` and `capsule` to recreate the paper.
+
+- `targets` is the workflow system that ensures the steps required to complete the paper are completed in the right order. 
+- `capsule` locks in the R package versions, including where they were downloaded from (github or CRAN, for example), so that these can be locked in. Under the hood it uses Rstudio's `renv` R package.
+
+There are three steps to reproduce the paper
+
+## Step 1: Install `capsule`
+
+```r
+# install.packages("remotes")
+remotes::install_github("milesmcbain/capsule")
+```
+
+## Step 2: Reproduce the libraries used
+
+```r
+capsule::reproduce_lib()
+```
+
+This recreates all of the R packages used in the analysis on your computer.
+Importantly, this will not change where your existing R packages are installed.
+It is just for this repository.
+So no need to be concerned about this impacting other analyses you run.
+
+## Step 3: Run the target workflow
+
+```r
+capsule::run(targets::tar_make())
+```
+
+This runs our targets workflow using the R packages specified.
+
+This will check if the targets are written, and if they aren't, it will re-run the necessary ones.
+
+## Step 5: Add new dependencies into "packages.R"?
+
+Run:
+
+```r
+capsule::recreate("./packages.R")
+```
+
+## Step 5 - ?
+
+Make some changes to the paper and want to see them? Run the capsule again:
+
+```r
+capsule::run(targets::tar_make())
+```
+
+And the paper will be recreated.

### 2021-06-24 paper/tidy-missing-data-paper.Rmd
first changes to use targets

@@
 options(tinytex.clean = FALSE)
+options(tinytex.verbose = TRUE)
+```
+
+```{r libs}
 library(tidyverse) 
 library(gridExtra)
-# options(kableExtra.latex.load_packages = FALSE)
-# library(kableExtra)
 library(ggthemes)
@@
 library(simputation)
-# if paper doesn't compile, use development version of naniar
-# install.packages("remotes")
-# remotes::install_github("njtierney/naniar")
 library(naniar) 
@@
 library(patchwork)
-options(tinytex.verbose = TRUE)
 ```
@@
 
-knitr::include_graphics("images/diagram.png")
+include_graphics(here("paper/images/diagram.png"))
 

### 2021-06-24 R/zip_for_arxiv.R

@@ -1,0 +1,38 @@
+##' .. content for \description{} (no empty lines) ..
+##'
+##' .. content for \details{} ..
+##'
+##' @title
+
+##' @return
+##' @author Nicholas Tierney
+##' @export
+zip_for_arxiv <- function() {
+  
+  dir_create("arxiv")
+  zip_name <- "arxiv/arxiv-submission.zip"
+  files_to_zip_all <- dir_ls(path = "paper/",
+                             recurse = TRUE)
+  
+  to_ignore <- glue_ignore(c("html", 
+                             "R",
+                             "Rmd",
+                             "bib",
+                             "aux",
+                             "bldg",
+                             "pdf",
+                             "brf",
+                             "out"))
+  
+  files_to_zip <- str_subset(string = files_to_zip_all,
+                             pattern = paste0("morgue|",to_ignore),
+                             negate = TRUE)
+  
+  if (file_exists(zip_name)) {
+    file_delete(zip_name)
+  }
+  
+  zip(zipfile = zip_name,
+      files = files_to_zip)
+  
+}

### 2021-06-24 _targets.R

@@ -1,0 +1,27 @@
+## Load your packages, e.g. library(targets).
+source("./packages.R")
+source("./conflicts.R")
+
+## Load your R files
+lapply(list.files("./R", full.names = TRUE), source)
+
+## tar_plan supports drake-style targets and also tar_target()
+tar_plan(
+
+# target = function_to_make(arg), ## drake style
+  tar_target(paper_bib,
+             "paper/tidy-missing-data-zotero.bib",
+             format = "file"),
+  tar_render(paper, 
+             "paper/tidy-missing-data-paper.Rmd"),
+  # ideally this would only run if the report has been rendered,
+  # not sure how to detect that.
+  tar_target(arxiv_paper,
+             zip_for_arxiv()),
+  tar_target(rjournal,
+             zip_for_jss())
+  
+
+# tar_target(target2, function_to_make2(arg)) ## targets style
+
+)

### 2021-06-24 packages.R

@@ -1,0 +1,22 @@
+## library() calls go here
+library(conflicted)
+library(dotenv)
+library(fs)
+library(gridExtra)
+library(ggthemes)
+library(here)
+library(knitr)
+library(lubridate)
+library(naniar) 
+library(magick)
+library(patchwork)
+library(png)
+library(scales)
+library(simputation)
+library(rmarkdown)
+library(rpart)
+library(rticles)
+library(tarchetypes)
+library(targets)
+library(tidyverse) 
+library(visdat)

### 2021-06-24 paper/tidy-missing-data-paper.Rmd
add plotting functions and stuff

@@
     url: http://www.njtierney.com
+    affiliation2: Telethon Kids Institute, Perth Children's Hospital, Perth, Australia
   - name: Dianne Cook
@@
 library(patchwork)
+library(withr)
+library(here)
+library(targets)
+```
+
+```{r tar-load}
+with_dir(here(), {
+  tar_load(c(housing, vis_miss_housing))
+})
 ```
@@
 However, these tools suffer the same problems as modelling: they use tidy input, but produce messy output - their output is challenging to integrate with other steps of data analysis.
-The complex, often multivariate nature of imputation methods also makes makes them difficult to represent.
+The complex, often multivariate nature of imputation methods also makes them difficult to represent.
 Visualization methods for missing data do not map data features to the aesthetics of a graphic, as in \pkg{ggplot2}, limiting expressive exploration.
@@
 
-\pkg{VIM} [@VIM] implements well-used imputation methods K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation.
-These diverse approaches allows for imputing with semi-continuous, continuous, count, and categorical data.
+\pkg{VIM} [@VIM] implements well-used imputation methods: K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation.
+These diverse approaches allow for imputing with semi-continuous, continuous, count, and categorical data.
 VIM identifies imputed cases by adding an indicator variable with a suffix "\code{_imp}".
@@
 \pkg{simputation} provides a consistent formula interface for all imputation methods, and always returns a dataframe with the updated imputations.
-\pkg{Hmisc} [@Hmisc], provides predictive mean matching, \pkg{imputeTS} [@imputeTS], provides time series imputation methods, and \pkg{missMDA} [@missMDA], imputes data using principal components analysis.
+\pkg{Hmisc} [@Hmisc], provides predictive mean matching, \pkg{imputeTS} [@imputeTS], provides time series imputation methods, and \pkg{missMDA} [@missMDA], imputes data using principal component analysis.
 
@@
 If there are missings in a continuous variable like test score, \pkg{ggplot2} omits the missings and prints a warning message.
-However, if a discrete variable like school year has missing values, an NA category is created for school year, where scores are placed (Figure \@ref(fig:gg-box-na)).
+However, if a discrete variable like school year has missing values, an NA category is created for school year, where scores are placed.
 
@@
     "add_n_miss(data)",
+    "add_any_miss(data)",
     "add_prop_miss(data)",
@@
     "contains the number missing values in a row",
+    "contains whether there are any missing values in a row",
     "contains the proportion of missing values in a row",
@@
 ```{r gg-miss-case-var, echo = FALSE, fig.show='hold', fig.cap = "Graphical summaries of missingness in the airquality data.
-Missings invariables (A) and cases (B), and for a birds eye view with missingness as a heatmap in (C), and with clustering applied (D).
+Missings in variables (A) and cases (B), and for an overview of all missingness as a heatmap in (C), and with clustering applied (D).
 There are missing values in Ozone and Solar.R, with Ozone having more missings.
@@
 The missings in Ozone are imputed and "Ozone_NA" is mapped to the fill aesthetic.
-*Nabular* data facilitates adding counts of missingness to a histogram, allowing examination of a variables' distribution of values, and also the magnitude of missings.
+*Nabular* data facilitates adding counts of missingness to a histogram, allowing examination of a variable's distribution of values, and also the magnitude of missings.
 
@@
 \pkg{naniar} works with \pkg{dplyr}'s \code{group_by()} operator to produce grouped summaries, which work well with the "pipe" operator.
-The code and Table \@ref(tab:group-miss-var-summary) below show an example of missing data summaries for airquality, grouped by month.
+Table \@ref(tab:group-miss-var-summary) below shows an example of missing data summaries for airquality, grouped by month.
 
@@
 
-```{r data-setup, include = FALSE}
-
-housing_raw <-
-  readr::read_csv(here::here("data",
-                             "melbourne_housing_raw.csv")) %>%
-  janitor::clean_names() %>%
-  dplyr::rename(region_name = regionname,
-         property_count = propertycount) %>%
-  dplyr::mutate(date = lubridate::dmy(date)) %>%
-  dplyr::rename(latitude = lattitude,
-                longitude = longtitude) %>%
-  # let's create monthly quarters
-  mutate(yr_qtr = as.ordered(lubridate::quarter(date, 
-                                             with_year = TRUE,
-                                             fiscal_start = 1))) %>%
-  # drop price
-  tidyr::drop_na(price) %>%
-  # make price the log of price
-    mutate(price = log10(price))
-
-```
-
-```{r data-clean, include = FALSE}
-
-# compact down the seller levels
-count_dat <- count(housing_raw,seller_g) %>% 
-  arrange(-n) %>%
-  mutate(cumu = cumsum(n),
-         pct = cumu / nrow(housing_raw),
-         pct_change = pct - lag(pct)) %>%
-  tibble::rowid_to_column() %>%
-  mutate(seller = case_when(
-    pct < 0.52 ~ as.character(seller_g),
-    pct > 0.52 & pct < 0.60 ~ "seller_g1",
-    pct > 0.60 & pct < 0.65 ~ "seller_g2",
-    pct > 0.65 & pct < 0.70 ~ "seller_g3",
-    pct > 0.70 & pct < 0.75 ~ "seller_g4",
-    pct > 0.75 & pct < 0.80 ~ "seller_g5",
-    pct > 0.80 & pct < 0.85 ~ "seller_g6",
-    pct > 0.85 & pct < 0.90 ~ "seller_g6",
-    pct > 0.90 & pct < 0.95 ~ "seller_g6",
-    pct > 0.95 ~ "seller_g5"
-  ))
-
-# drop these three variables, as we won't use them in analysis, they
-# would need extensive recoding
-
-housing <- housing_raw %>%
-  mutate_if(is.character, as.factor) %>%
-  mutate_at(vars(rooms,
-                 bedroom2,
-                 bathroom,
-                 car),
-            as.factor) %>%
-  # recode / collapse factors with many levels and few obs
-  mutate(bathroom = fct_other(bathroom,
-                                drop = c("4", "5", "6", "7", "8", "9"),
-                                other_level = "4+")) %>%
-  mutate(bedroom2 = fct_other(bedroom2,
-                               drop = c("5", "6", "7", "8", "9", "10", 
-                                        "12", "16", "20"),
-                               other_level = "5+")) %>%
-  mutate(rooms = fct_other(rooms,
-                             drop = c("6", "7", "8", "9", "10", 
-                                      "12", "16"),
-                             other_level = "6+")) %>%
-  mutate(car = fct_other(car,
-                           drop = c("4", "5", "6", "7", "8", "9", "10",
-                                    "11", "18"),
-                           other_level = "4+")) %>%
-  left_join(select(count_dat, seller_g, seller), by = "seller_g")  %>%
-  # drop seller_g
-  select(-seller_g) %>%
-  mutate(seller = as.factor(seller))
-
-```
-
 This section shows how the methods described so far are used together in a data analysis workflow.
@@
 Figure \@ref(fig:housing-miss-case-var)A shows 9 variables with missing values.
-The most missings are in "building area", followed by "year built", and "land size", with similar amounts of missingness in "Car", "bathroom", "bedroom2", "longitude", and "latitude." Figure \@ref(fig:housing-miss-case-var)B reveals there are up to 50% missing values in cases, and the majority of cases have more than 5% values missing.
+The most missings are in "building area", followed by "year built", and "land size", with similar amounts of missingness in "Car", "bathroom", "bedroom2", "longitude", and "latitude".
+Figure \@ref(fig:housing-miss-case-var)B reveals there are up to 50% missing values in cases, and the majority of cases have more than 5% values missing.
 The variables "building area" and "year built" have more than 50% missing data, and so could perhaps be omitted from subsequent analysis, as imputed values are likely to be spurious.
@@
 The majority of missingness is in selected cases and variables.", fig.height = 4, fig.width = 8, out.width = "100%"}
+
 q1 <- gg_miss_var(housing, show_pct = TRUE) + labs(tag = "A")
@@
 
-housing %>%
-  # sample_frac(0.25) %>%
-  vis_miss(cluster = TRUE,
-           sort_miss = TRUE,
-           show_perc_col = FALSE)
+vis_miss_housing
 

### 2021-06-24 R/clean_housing_data.R

@@ -1,0 +1,86 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+
+#' @return
+#' @author Nicholas Tierney
+#' @export
+clean_housing_data <- function() {
+
+  
+  housing_raw <-
+    read_csv(here("data",
+                  "melbourne_housing_raw.csv")) %>%
+    clean_names() %>%
+    rename(region_name = regionname,
+                  property_count = propertycount) %>%
+    mutate(date = dmy(date)) %>%
+    rename(latitude = lattitude,
+           longitude = longtitude) %>%
+    # let's create monthly quarters
+    mutate(yr_qtr = as.ordered(quarter(date, 
+                                       with_year = TRUE,
+                                        fiscal_start = 1))) %>%
+    # drop price
+    drop_na(price) %>%
+    # make price the log of price
+    mutate(price = log10(price))
+  
+  
+  # compact down the seller levels
+  count_dat <- count(housing_raw,seller_g) %>% 
+    arrange(-n) %>%
+    mutate(cumu = cumsum(n),
+           pct = cumu / nrow(housing_raw),
+           pct_change = pct - lag(pct)) %>%
+    rowid_to_column() %>%
+    mutate(seller = case_when(
+      pct < 0.52 ~ as.character(seller_g),
+      pct > 0.52 & pct < 0.60 ~ "seller_g1",
+      pct > 0.60 & pct < 0.65 ~ "seller_g2",
+      pct > 0.65 & pct < 0.70 ~ "seller_g3",
+      pct > 0.70 & pct < 0.75 ~ "seller_g4",
+      pct > 0.75 & pct < 0.80 ~ "seller_g5",
+      pct > 0.80 & pct < 0.85 ~ "seller_g6",
+      pct > 0.85 & pct < 0.90 ~ "seller_g6",
+      pct > 0.90 & pct < 0.95 ~ "seller_g6",
+      pct > 0.95 ~ "seller_g5"
+    ))
+  
+  # drop these three variables, as we won't use them in analysis, they
+  # would need extensive recoding
+  
+  housing <- housing_raw %>%
+    mutate_if(is.character, as.factor) %>%
+    mutate_at(vars(rooms,
+                   bedroom2,
+                   bathroom,
+                   car),
+              as.factor) %>%
+    # recode / collapse factors with many levels and few obs
+    mutate(bathroom = fct_other(bathroom,
[…]

### 2021-06-24 R/gg_vis_miss_housing.R

@@ -1,0 +1,17 @@
+#' .. content for \description{} (no empty lines) ..
+#'
+#' .. content for \details{} ..
+#'
+#' @title
+#' @param housing
+#' @return
+#' @author Nicholas Tierney
+#' @export
+gg_vis_miss_housing <- function(housing) {
+
+  vis_miss(housing,
+           cluster = TRUE,
+           sort_miss = TRUE,
+           show_perc_col = FALSE)
+
+}

### 2021-06-24 _targets.R

@@ -12,6 +12,10 @@ tar_plan(
   tar_target(paper_bib,
              "paper/tidy-missing-data-zotero.bib",
              format = "file"),
+  tar_target(housing,
+             clean_housing_data()),
+  tar_target(vis_miss_housing,
+             gg_vis_miss_housing(housing)),
   tar_render(paper, 
              "paper/tidy-missing-data-paper.Rmd"),
   # ideally this would only run if the report has been rendered,

### 2021-06-24 conflicts.R

@@ -4,3 +4,4 @@ conflict_prefer("select", "dplyr")
 conflict_prefer("pluck", "purrr")
 conflict_prefer("map", "purrr")
 conflict_prefer("pull", "dplyr")
+conflict_prefer("lag", "dplyr")

### 2021-06-24 packages.R

@@ -5,6 +5,7 @@ library(fs)
 library(gridExtra)
 library(ggthemes)
 library(here)
+library(janitor)
 library(knitr)
 library(lubridate)
 library(naniar) 
@@ -18,5 +19,6 @@ library(rpart)
 library(rticles)
 library(tarchetypes)
 library(targets)
+library(tibble) 
 library(tidyverse) 
 library(visdat)

### 2021-06-25 paper/tidy-missing-data-paper.Rmd

@@
   - name: Nicholas Tierney
-    affiliation: Monash University
+    affiliation: Monash University, Telethon Kids Institute
     address: >
@@
     url: http://www.njtierney.com
-    affiliation2: Telethon Kids Institute, Perth Children's Hospital, Perth, Australia
   - name: Dianne Cook

### 2021-06-29 paper/tidy-missing-data-paper.Rmd
update main paper Rmd, remove unused address

@@
   - name: Nicholas Tierney
-    affiliation: Monash University, Telethon Kids Institute
-    address: >
-      Department of Econometrics and Business Statistics
-      E774, Menzies Building
-      Monash University, Clayton
+    affiliation: Monash University
     email: \email{nicholas.tierney@gmail.com}

### 2022-03-29 paper/tidy-missing-data-paper.Rmd
misc style fixes

@@
 
-**MCAR** is where values being missing have no association with observed or unobserved data, that is, $Pr(B = 1 | Y_{obs}, Y_{miss}) = Pr(B = 1 | \psi)$.
+MCAR is where values being missing have no association with observed or unobserved data, that is, $Pr(B = 1 | Y_{obs}, Y_{miss}) = Pr(B = 1 | \psi)$.
 Essentially, the probability of an observation being missing is unrelated to anything else, only the parameter $\psi$, the overall probability of missingness.
 Although a convenient scenario, it is not actually possible to confirm, or clearly distinguish from MAR, as it relies on statements on data unobserved.
-In **MAR**, missingness only depends on data observed, not data missing, that is, $Pr(B = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(B = 1 | Y_{obs}, \psi)$.
+In MAR, missingness only depends on data observed, not data missing, that is, $Pr(B = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(B = 1 | Y_{obs}, \psi)$.
 Some structure or dependence between missing and observed values is allowed, provided it can be explained by data observed, and some overall probability of missingness.
-In **MNAR**, missingness is related to values observed, and unobserved: $Pr(B = 1 | Y_{obs}, Y_{miss}, \psi)$.
+In MNAR, missingness is related to values observed, and unobserved: $Pr(B = 1 | Y_{obs}, Y_{miss}, \psi)$.
 This assumes conditioning on all observations: data goes missing due to some phenomena unobserved, including the structure of the missing data itself.
@@
 
-# Existing Software {#existing-software}
+# Existing software {#existing-software}
 
@@
 
-### scan: searching for common missing value labels {#verbs-search}
+### Scan: searching for common missing value labels {#verbs-search}
 
@@
 
-### replace: replacing values with missing values {#verbs-replace-with}
+### Replace: replacing values with missing values {#verbs-replace-with}
 
@@
 
-### add: adding missingness summary variables {#verbs-add-cols}
+### Add: adding missingness summary variables {#verbs-add-cols}
 
@@
 
-### shadow: creating nabular data {#verbs-nabular}
+### Shadow: creating nabular data {#verbs-nabular}
 
@@
 
-### flag: describing different types of missing values {#verbs-recode}
+### Flag: describing different types of missing values {#verbs-recode}
 
@@
 
-### impute: imputing values {#verbs-impute}
+### Impute: imputing values {#verbs-impute}
 
@@
 
-### track: shadow and impute missing values {#verbs-track}
+### Track: shadow and impute missing values {#verbs-track}
 

### 2022-03-29 R/zip_for_jss.R

@@ -7,17 +7,12 @@
 ##' @return
 ##' @author Nicholas Tierney
 ##' @export
-zip_for_jss <- function() {
+zip_for_jss <- function(paper, paper_purl) {
   
   dir_create("jss")
   zip_name <- "jss/jss-submission.zip"
-  files_to_zip_all <- dir_ls(path = "paper/",
-                             recurse = TRUE)
   
-  cover_letter <- dir_ls(path = "cover-letter",
-                         regexp = "cover-letter.pdf$")
-  
-  to_ignore <- glue_ignore(c("html", 
+  to_ignore <- glue_ignore(c("html",
                              "aux",
                              "bbl",
                              "blg",
@@ -26,10 +21,14 @@ zip_for_jss <- function() {
                              "brf",
                              "out"))
   
-  files_to_zip <- str_subset(string = c(files_to_zip_all,
-                                        cover_letter),
-                             pattern = paste0("morgue|",to_ignore),
-                             negate = TRUE)
+  files_to_zip <- dir_ls(
+    path = "paper/",
+    recurse = TRUE,
+    regexp = paste0(c("images|diagram|review|*cache|"),
+                    to_ignore),
+    invert = TRUE
+  )
+  
   
   if (file_exists(zip_name)) {
     file_delete(zip_name)

### 2022-03-29 _targets.R

@@ -18,12 +18,15 @@ tar_plan(
              gg_vis_miss_housing(housing)),
   tar_render(paper, 
              "paper/tidy-missing-data-paper.Rmd"),
+  tar_target(paper_purl,
+             knitr::purl(input = "paper/tidy-missing-data-paper.Rmd",
+                         output = "paper/tidy-missing-data-paper.R")),
   # ideally this would only run if the report has been rendered,
   # not sure how to detect that.
   tar_target(arxiv_paper,
              zip_for_arxiv()),
-  tar_target(rjournal,
-             zip_for_jss())
+  tar_target(jss,
+             zip_for_jss(paper, paper_purl))
   
 
 # tar_target(target2, function_to_make2(arg)) ## targets style

### 2022-03-29 paper/tidy-missing-data-paper.R

@@ -1,12 +1,22 @@
+## ----setup, include=FALSE, message=FALSE, warning=FALSE-----------------------
+knitr::opts_chunk$set(echo = FALSE, 
+                      message = FALSE, 
+                      warning = FALSE, 
+                      fig.align = "center",
+                      fig.width = 4, 
+                      fig.height = 4, 
+                      cache = TRUE)
+options(tinytex.clean = FALSE)
+options(tinytex.verbose = TRUE)
+
+
+## ----libs---------------------------------------------------------------------
 library(tidyverse) 
 library(gridExtra)
 library(ggthemes)
 library(png)
 library(scales)
 library(simputation)
-# if code doesn't work, use development version of naniar
-# install.packages("remotes")
-# remotes::install_github("njtierney/naniar")
 library(naniar) 
 library(knitr)
 library(lubridate)
@@ -14,8 +24,18 @@ library(magick)
 library(rpart)
 library(visdat)
 library(patchwork)
+library(withr)
+library(here)
+library(targets)
 
-## ----warning, fig.cap = "How ggplot2 behaves when displaying missing values. A warning message is displayed, but missing values are not shown in the plot", fig.height = 4, fig.width = 8, out.width = "75%", warning = FALSE, fig.show = 'hold'----
+
+## ----tar-load-----------------------------------------------------------------
+with_dir(here(), {
+  tar_load(c(housing, vis_miss_housing))
+})
+
+
+## ----warning, fig.cap = "How \\pkg{ggplot2} behaves when displaying missing values. A warning message is displayed, but missing values are not shown in the plot.", fig.height = 4, fig.width = 8, out.width = "75%", warning = FALSE, fig.show = 'hold'----
 p1 <- ggplot(oceanbuoys,
        aes(x = humidity,
            y = air_temp_c)) +
@@ -26,7 +46,7 @@ p1 <- ggplot(oceanbuoys,
 
 p1 + grid::textGrob(
   label = "Warning message:
-Removed 171 rows containing 
+removed 171 rows containing 
 missing values (geom_point).", 
 gp = grid::gpar(fontsize = 14,
                 fontfamily = "Courier",
@@ -36,7 +56,7 @@ gp = grid::gpar(fontsize = 14,
   
 
 
-## ----make-school-scores-data, echo = FALSE-----------------------------------------
+## ----make-school-scores-data, echo = FALSE------------------------------------
 
[…]

### 2022-04-01 paper/tidy-missing-data-paper.Rmd
add final changes from JSS review

@@
 options(tinytex.verbose = TRUE)
+options(xtable.comment = FALSE)
 ```
@@
                 search = -99) %>% 
-  knitr::kable(
-    caption = 'Table of the occurrences of the search "-99" in the data, "dat\\_ms". There is one occurrence if -99 in variables x and z.')
-    # booktabs = TRUE) 
-  # kable_styling(latex_options = c("hold_position"))
+  xtable::xtable(
+    caption = 'Table of the occurrences of the search "-99" in the data, "dat\\_ms". There is one occurrence if -99 in variables x and z.',
+    label = "tab:miss-scan-count"
+    )
 ```
@@
 
-```{r add-missing-info}
-
-tibble::tibble(
+```{r add-missing-info, results='asis'}
+tab_add_missing_info <- tibble::tibble(
   `Function` = c(
@@
   )) %>% 
-    knitr::kable(caption = 'Overview of the "add" functions in naniar')
-
+    xtable::xtable(
+      caption = 'Overview of the "add" functions in naniar',
+      label = "tab:add-missing-info"
+      )
+
+print(tab_add_missing_info, comment = FALSE, include.rownames = FALSE)
 ```
@@
 
-```{r impute-summary-out, echo = FALSE}
-aq_imputed %>%
+```{r impute-summary-out, echo = FALSE, results='asis'}
+tab_impute_summary_out <- aq_imputed %>%
   group_by(any_missing) %>%
@@
                .funs = lst(min, mean, median, max))  %>% 
-  kable(caption = 'Output of \\pkg{dplyr} summary statistics of imputed vs non imputed values for the variable "Ozone".
+  xtable::xtable(
+    caption = 'Output of \\pkg{dplyr} summary statistics of imputed vs non imputed values for the variable "Ozone".
 The "any\\_missing" column denotes imputed values ("Missing", since they were previously missing), and non-imputed values ("Not Missing").
-The mean and median values are similar, but the minimum and maximum values are very different.')
+The mean and median values are similar, but the minimum and maximum values are very different.',
+    label = "tab:impute-summary-out"
+    )
+print(tab_impute_summary_out, comment = FALSE, include.rownames = FALSE)
 ```
@@
 
-```{r n-prop-pct-miss-complete, echo = FALSE}
+```{r n-prop-pct-miss-complete, echo = FALSE, results='asis'}
 
@@
 
-knitr::kable(
+show_table_of_fun <- xtable::xtable(
   table_of_fun, 
   digits = 2,
-  caption = "Single number summaries of missingness and completeness of the airquality dataset. The functions follow consistent naming, making them easy to remember, and their use clear.")
-  # booktabs = TRUE)
-  # kable_styling(latex_options = c("hold_position"))
+  caption = "Single number summaries of missingness and completeness of the airquality dataset. The functions follow consistent naming, making them easy to remember, and their use clear.",
+  label = "tab:n-prop-pct-miss-complete"
+  )
+
+print(show_table_of_fun, comment = FALSE, include.rownames = FALSE)
 
@@
 
-```{r miss-var-summary}
-
-miss_var_summary(airquality) %>% 
-  knitr::kable(
+```{r miss-var-summary, results='asis'}
+tab_miss_var_summary <- miss_var_summary(airquality) %>% 
+  xtable::xtable(
     caption = "\\texttt{miss\\char`_var\\char`_summary(airquality)} provides the number and percent of missings in each variable in airquality. Only ozone and solar radiation have missing values.",
-    digits = 1)
-    # booktabs = TRUE) 
-  # kable_styling(latex_options = c("hold_position"))
-
-```
-
-```{r miss-var-table}
-
-miss_var_table(airquality) %>% 
-  knitr::kable(
+    digits = 1,
+    label = "tab:miss-var-summary"
+    )
+
+print(tab_miss_var_summary, comment = FALSE, include.rownames = FALSE)
+
+```
+
+```{r miss-var-table, results='asis'}
+
+tab_miss_var_table <- miss_var_table(airquality) %>% 
+  xtable::xtable(
     caption = "The output of \\code{miss\\char`_var\\char`_table(airquality)}, tabulating the amount of missing data in each variable in airquality.
@@
 There are few missingness patterns.",
-    digits = 1)
-    # booktabs = TRUE)
-  # kable_styling(latex_options = c("hold_position"))
+    digits = 1,
+    label = "tab:miss-var-table"
+    )
+
+print(tab_miss_var_table, comment = FALSE, include.rownames = FALSE)
 ```
@@
 
-```{r group-miss-var-summary}
-
-airquality %>%
+```{r group-miss-var-summary, results='asis'}
+
+tab_group_miss_var_summary <- airquality %>%
   group_by(Month) %>%
@@
   slice(1:10) %>%
-  knitr::kable(caption = "Output of \\code{airquality \\%>\\% group\\char`_by(Month) \\%>\\% miss\\char`_var\\char`_summary()} provides a grouped summary of the missingness in each variable, for each month of the airquality dataset.
+  xtable::xtable(
+    caption = "Output of \\code{airquality \\%>\\% group\\char`_by(Month) \\%>\\% miss\\char`_var\\char`_summary()} provides a grouped summary of the missingness in each variable, for each month of the airquality dataset.
 Only the first 10 rows are shown.
 There are more ozone missings in June than May.",
-               digits = 1)
-               # booktabs = TRUE)
-  # kable_styling(latex_options = c("hold_position"))
+    digits = 1,
+    label = "tab:group-miss-var-summary"
+    )
+
+print(tab_group_miss_var_summary, comment = FALSE, include.rownames = FALSE)
 
@@
 
-```{r housing-miss-var-case-table}
+```{r housing-miss-var-case-table, results='asis'}
 
@@
 
-knitr::kable(list(t1, t2),
+t3 <- bind_cols(t1,t2)
+
+# knitr::kable(list(t1, t2),
+
+tab_housing_miss_var_case_table <- xtable::xtable(
+  t3,
   caption = "Summary tables to help understand missingness patters.
-Output of \\code{miss\\_var\\_table(housing)} (left), tabulating missingness for variables, and output of \\code{miss\\_case\\_table(housing)}.
+Output of \\code{miss\\_var\\_table(housing)} (left), tabulating missingness for variables, and output of \\code{miss\\_case\\_table(housing)} (right).
 There are 13 variables with 0-3 missings, 6 variables have 6000-10000 missings, 2 variables have 15000 – 17000 missings.
@@
 There are different patterns of missingness in variables and cases, but they can be broken down into smaller groups.",
-             digits = 1)
-             # booktabs = TRUE)
-  # kable_styling(latex_options = c("hold_position"))
+  digits = 1,
+  label = "tab:housing-miss-var-case-table",
+  align = "llll||lll"
+  )
+
+print(tab_housing_miss_var_case_table, comment = FALSE, include.rownames = FALSE)
 

### 2022-04-01 paper/tidy-missing-data-paper.R

@@ -1,4 +1,4 @@
-## ----setup, include=FALSE, message=FALSE, warning=FALSE-----------------------
+## ----setup, include=FALSE, message=FALSE, warning=FALSE------------------------------
 knitr::opts_chunk$set(echo = FALSE, 
                       message = FALSE, 
                       warning = FALSE, 
@@ -8,9 +8,10 @@ knitr::opts_chunk$set(echo = FALSE,
                       cache = TRUE)
 options(tinytex.clean = FALSE)
 options(tinytex.verbose = TRUE)
+options(xtable.comment = FALSE)
 
 
-## ----libs---------------------------------------------------------------------
+## ----libs----------------------------------------------------------------------------
 library(tidyverse) 
 library(gridExtra)
 library(ggthemes)
@@ -29,7 +30,7 @@ library(here)
 library(targets)
 
 
-## ----tar-load-----------------------------------------------------------------
+## ----tar-load------------------------------------------------------------------------
 with_dir(here(), {
   tar_load(c(housing, vis_miss_housing))
 })
@@ -56,7 +57,7 @@ gp = grid::gpar(fontsize = 14,
   
 
 
-## ----make-school-scores-data, echo = FALSE------------------------------------
+## ----make-school-scores-data, echo = FALSE-------------------------------------------
 
 school_scores <- data.frame(score = c(rnorm(20, 70, 10),
                                  rnorm(20, 85, 7.5),
@@ -134,7 +135,7 @@ include_graphics(here("paper/images/diagram.png"))
 
 
 
-## ----setup-miss-scan-count----------------------------------------------------
+## ----setup-miss-scan-count-----------------------------------------------------------
 
 dat_ms <- tibble::tribble(~x,  ~y,    ~z,
                          1,   "A",   -100,
@@ -144,18 +145,17 @@ dat_ms <- tibble::tribble(~x,  ~y,    ~z,
                          -98, "F",   -1)
 
 
-## ----miss-scan-count, eval = FALSE--------------------------------------------
+## ----miss-scan-count, eval = FALSE---------------------------------------------------
 ## miss_scan_count(data = dat_ms,
 ##                 search = -99) %>%
-##   knitr::kable(
-##     caption = 'Table of the occurrences of the search "-99" in the data, "dat\\_ms". There is one occurrence if -99 in variables x and z.')
-##     # booktabs = TRUE)
-##   # kable_styling(latex_options = c("hold_position"))
+##   xtable::xtable(
[…]

### 2022-07-20 jss/paper/jss/tidy-missing-data-paper.R
add JSS files

@@ -1,0 +1,811 @@
+## ----setup, include=FALSE, message=FALSE, warning=FALSE-----------------------
+knitr::opts_chunk$set(echo = FALSE, 
+                      message = FALSE, 
+                      warning = FALSE, 
+                      fig.align = "center",
+                      fig.width = 4, 
+                      fig.height = 4, 
+                      cache = TRUE)
+options(tinytex.clean = FALSE)
+options(tinytex.verbose = TRUE)
+
+
+## ----libs---------------------------------------------------------------------
+library(tidyverse) 
+library(gridExtra)
+library(ggthemes)
+library(png)
+library(scales)
+library(simputation)
+library(naniar) 
+library(knitr)
+library(lubridate)
+library(magick)
+library(rpart)
+library(visdat)
+library(patchwork)
+library(withr)
+library(here)
+library(targets)
+
+
+## ----tar-load-----------------------------------------------------------------
+with_dir(here(), {
+  tar_load(c(housing, vis_miss_housing))
+})
+
+
+## ----warning, fig.cap = "How \\pkg{ggplot2} behaves when displaying missing values. A warning message is displayed, but missing values are not shown in the plot.", fig.height = 4, fig.width = 8, out.width = "75%", warning = FALSE, fig.show = 'hold'----
+p1 <- ggplot(oceanbuoys,
+       aes(x = humidity,
+           y = air_temp_c)) +
+  geom_point(size = 0.5) +
+  theme(aspect.ratio = 1) + 
+  labs(x = "Humidity",
+       y = "Air Temperature (C)")
+
+p1 + grid::textGrob(
+  label = "Warning message:
+removed 171 rows containing 
+missing values (geom_point).", 
+gp = grid::gpar(fontsize = 14,
+                fontfamily = "Courier",
+                col = "salmon", 
+                fontface = "bold"))
+
+  
+
+
+## ----make-school-scores-data, echo = FALSE------------------------------------
+
+school_scores <- data.frame(score = c(rnorm(20, 70, 10),
+                                 rnorm(20, 85, 7.5),
+                                 rnorm(20, 95, 5),
+                                 rnorm(20, 115, 10)),
+                       year = rep(c("1st", "2nd", "3rd", "4th"), each = 20)) %>%
+  mutate(score = as.numeric(score),
+         year = as.factor(year)) %>%
+  as_tibble()
+
+school_scores_year_miss <- school_scores %>%
+  mutate_at(.funs = function(x){
[…]

### 2022-07-20 jss/paper/tidy-missing-data-paper.R

@@ -1,4 +1,4 @@
-## ----setup, include=FALSE, message=FALSE, warning=FALSE------------------------------
+## ----setup, include=FALSE, message=FALSE, warning=FALSE-----------------------
 knitr::opts_chunk$set(echo = FALSE, 
                       message = FALSE, 
                       warning = FALSE, 
@@ -7,17 +7,16 @@ knitr::opts_chunk$set(echo = FALSE,
                       fig.height = 4, 
                       cache = TRUE)
 options(tinytex.clean = FALSE)
+options(tinytex.verbose = TRUE)
+
+
+## ----libs---------------------------------------------------------------------
 library(tidyverse) 
 library(gridExtra)
-# options(kableExtra.latex.load_packages = FALSE)
-# library(kableExtra)
 library(ggthemes)
 library(png)
 library(scales)
 library(simputation)
-# if paper doesn't compile, use development version of naniar
-# install.packages("remotes")
-# remotes::install_github("njtierney/naniar")
 library(naniar) 
 library(knitr)
 library(lubridate)
@@ -25,7 +24,15 @@ library(magick)
 library(rpart)
 library(visdat)
 library(patchwork)
-options(tinytex.verbose = TRUE)
+library(withr)
+library(here)
+library(targets)
+
+
+## ----tar-load-----------------------------------------------------------------
+with_dir(here(), {
+  tar_load(c(housing, vis_miss_housing))
+})
 
 
 ## ----warning, fig.cap = "How \\pkg{ggplot2} behaves when displaying missing values. A warning message is displayed, but missing values are not shown in the plot.", fig.height = 4, fig.width = 8, out.width = "75%", warning = FALSE, fig.show = 'hold'----
@@ -39,7 +46,7 @@ p1 <- ggplot(oceanbuoys,
 
 p1 + grid::textGrob(
   label = "Warning message:
-Removed 171 rows containing 
+removed 171 rows containing 
 missing values (geom_point).", 
 gp = grid::gpar(fontsize = 14,
                 fontfamily = "Courier",
@@ -49,7 +56,7 @@ gp = grid::gpar(fontsize = 14,
   
 
 
-## ----make-school-scores-data, echo = FALSE-------------------------------------------
+## ----make-school-scores-data, echo = FALSE------------------------------------
 
 school_scores <- data.frame(score = c(rnorm(20, 70, 10),
                                  rnorm(20, 85, 7.5),
@@ -99,7 +106,7 @@ school_scores_all <- bind_rows(
 
 
 
[…]

### 2022-07-20 paper/tidy-missing-data-paper.R
first resolution at resolving new changes and unifying everything into one place again.

@@ -1,4 +1,4 @@
-## ----setup, include=FALSE, message=FALSE, warning=FALSE------------------------------
+## ----setup, include=FALSE, message=FALSE, warning=FALSE-----------------------
 knitr::opts_chunk$set(echo = FALSE, 
                       message = FALSE, 
                       warning = FALSE, 
@@ -8,10 +8,9 @@ knitr::opts_chunk$set(echo = FALSE,
                       cache = TRUE)
 options(tinytex.clean = FALSE)
 options(tinytex.verbose = TRUE)
-options(xtable.comment = FALSE)
 
 
-## ----libs----------------------------------------------------------------------------
+## ----libs---------------------------------------------------------------------
 library(tidyverse) 
 library(gridExtra)
 library(ggthemes)
@@ -30,7 +29,7 @@ library(here)
 library(targets)
 
 
-## ----tar-load------------------------------------------------------------------------
+## ----tar-load-----------------------------------------------------------------
 with_dir(here(), {
   tar_load(c(housing, vis_miss_housing))
 })
@@ -57,7 +56,7 @@ gp = grid::gpar(fontsize = 14,
   
 
 
-## ----make-school-scores-data, echo = FALSE-------------------------------------------
+## ----make-school-scores-data, echo = FALSE------------------------------------
 
 school_scores <- data.frame(score = c(rnorm(20, 70, 10),
                                  rnorm(20, 85, 7.5),
@@ -135,7 +134,7 @@ include_graphics(here("paper/images/diagram.png"))
 
 
 
-## ----setup-miss-scan-count-----------------------------------------------------------
+## ----setup-miss-scan-count----------------------------------------------------
 
 dat_ms <- tibble::tribble(~x,  ~y,    ~z,
                          1,   "A",   -100,
@@ -145,17 +144,18 @@ dat_ms <- tibble::tribble(~x,  ~y,    ~z,
                          -98, "F",   -1)
 
 
-## ----miss-scan-count, eval = FALSE---------------------------------------------------
+## ----miss-scan-count, eval = FALSE--------------------------------------------
 ## miss_scan_count(data = dat_ms,
 ##                 search = -99) %>%
-##   xtable::xtable(
-##     caption = 'Table of the occurrences of the search "-99" in the data, "dat\\_ms". There is one occurrence if -99 in variables x and z.',
-##     label = "tab:miss-scan-count"
-##     )
+##   knitr::kable(
[…]

### 2022-08-15 packages.R
update to latest R

@@ -1,5 +1,6 @@
 ## library() calls go here
 library(conflicted)
+library(cowplot)
 library(dotenv)
 library(fs)
 library(gridExtra)

### 2022-08-15 paper/tidy-missing-data-paper.R
minor changes

@@ -130,7 +130,9 @@ cowplot::plot_grid(p1,
 
 ## ----nabularfig, echo = FALSE, fig.cap = 'The process of creating nabular data. Data transformed to shadow matrix, where values are either not missing or missing: "!NA" or "NA". The shadow matrix can be converted to long form to create missingness summary plots. Nabular data is created by binding the columns of the data and shadow matrix. Special missing values (such as -99) are identified as special missings, and values imputed and tracked. Nabular data can be used to identify imputations and explore data values alongside missings, providing a useful format for missing data exploration and analysis.', out.width = "100%", fig.align = "center"----
 
-include_graphics(here("paper/images/diagram.png"))
+include_graphics(here::here("paper",
+                            "images",
+                            "diagram.png"))
 
 
 

### 2022-08-16 paper/tidy-missing-data-paper.Rmd
first resolution at resolving new changes and unifying everything into one place again.
various things trying to get repo to compile

--- 2021-06-29 paper/tidy-missing-data-paper.Rmd
@@
 
-```{r nabularfig, echo = FALSE, fig.cap = 'The process of creating nabular data.
+```{r}
+# stop(fs::dir_ls(here::here("paper/images"), glob = "*.png"))
+# stop(file.exists(fs::dir_ls(here::here("paper/images"), glob = "*.png")))
+```
+
+```{r nabularfig, echo = FALSE, results='asis', fig.cap = 'The process of creating nabular data.
 Data transformed to shadow matrix, where values are either not missing or missing: "!NA" or "NA".
@@
 
-include_graphics(here("paper/images/diagram.png"))
-
+knitr::include_graphics(path = here::here("paper/images/overview-diagram.png"),
+                        rel_path = FALSE)
 ```

### 2022-08-16 packages.R

@@ -17,6 +17,7 @@ library(scales)
 library(simputation)
 library(rmarkdown)
 library(rpart)
+library(rpart.plot)
 library(rticles)
 library(tarchetypes)
 library(targets)

### 2022-08-16 paper/tidy-missing-data-paper.R

@@ -128,12 +128,13 @@ cowplot::plot_grid(p1,
           labels = LETTERS[1:4])
 
 
+## ----new-fig------------------------------------------------------------------
+knitr::include_graphics(here::here("paper/images/overview-diagram.png"))
+
+
 ## ----nabularfig, echo = FALSE, fig.cap = 'The process of creating nabular data. Data transformed to shadow matrix, where values are either not missing or missing: "!NA" or "NA". The shadow matrix can be converted to long form to create missingness summary plots. Nabular data is created by binding the columns of the data and shadow matrix. Special missing values (such as -99) are identified as special missings, and values imputed and tracked. Nabular data can be used to identify imputations and explore data values alongside missings, providing a useful format for missing data exploration and analysis.', out.width = "100%", fig.align = "center"----
 
-include_graphics(here::here("paper",
-                            "images",
-                            "diagram.png"))
-
+knitr::include_graphics(here::here("paper/images/overview-diagram.png"))
 
 
 ## ----setup-miss-scan-count----------------------------------------------------

### 2022-08-16 paper/tidy-missing-data-paper-manually-edited-final.tex

\documentclass[
]{jss}

\usepackage[utf8]{inputenc}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\author{
Nicholas Tierney\\Monash University \And Dianne Cook\\Monash University
}
\title{Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations}

\Plainauthor{Nicholas Tierney, Dianne Cook}
\Plaintitle{Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations}
\Shorttitle{Explore missingness with \pkg{naniar}}

\Abstract{
Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows. We define a new data structure, and a suite of new operations. Together, these provide a connected framework for handling, exploring, and imputing missing values. These methods are available in the \proglang{R} package \pkg{naniar}.
}

\Keywords{statistical computing, statistical graphics, data science, data visualization, \pkg{tidyverse}, data pipeline, \proglang{R}}
\Plainkeywords{statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    }

% Pandoc citation processing

% Pandoc header
\usepackage{amsmath}
\usepackage[english]{babel}

\usepackage{float}
\usepackage{booktabs}

\begin{document}

\hypertarget{intro}{%
\section{Introduction}\label{intro}}

As data science has become a more solid field, theories and principles have developed to describe best practices. One such idea is ``tidy data'', which defines a clean, analysis-ready format that informs workflows converting raw data through a data analysis pipeline \citep{Wickham2014}. Another idea is the grammar of graphics, describing how to map data values into a visualization \citep{Wilkinson2012}. These principles have been widely adopted, but do not address the problem of missing data. In particular, there is little guidance on how to handle missing values in a data analysis workflow. Most analysis and visualization software simply drop missing values when making a plot, although some (\pkg{ggplot2}) provide a warning (Figure \ref{fig:warning}).

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=0.75\linewidth]{tidy-missing-data-paper_files/figure-latex/warning-1}

}

\caption[How \pkg{ggplot2} behaves when displaying missing values]{How \pkg{ggplot2} behaves when displaying missing values. A warning message is displayed, but missing values are not shown in the plot.}\label{fig:warning}
\end{figure}
\end{CodeChunk}

The imputation literature focuses on ensuring valid statistical inference is made from incomplete data. This is approached chiefly through probabilistic modelling, assuming the mechanism of missing data is known to the analyst. It does not address how to explore, understand, and handle missing data structures and mechanisms.

However, something must be known about the missing value structure to produce a complete dataset for analysis, whether by case- or variable-deletion, or with imputation. Decision tree models or latent group analysis can reveal structures or patterns of missingness \citep{Tierney2015, Barnett2017}, but are not definitive. While there are times where the missing data mechanism is obvious, determining this is typically not straightforward. A straightforward case is where missing values are encoded in the data in a non standard format that still indicates missingness, such as ``N.A.'', ``N/A'', or ``missing''. However, identifying these amongst all data can be problematic as it is hard to know what you are looking for, like finding a needle in a haystack. A more challenging case is where missing values arise due to the increasing or decreasing values of other variables (including themselves). For example, in weather data, temperature values might go missing due to their own high temperature. Or similarly, other measurements go missing due to high temperatures. Identifying this requires exploring the data, and the analysis path to take is obvious, as missing values are often omitted from plots and other statistical summaries. To understand their structure and possible mechanisms, the analyst must explore the data with visualizations, summaries, and modelling, in an iterative fashion.

While there have been many software tools for exploring missing data, they do not work together. The graphics literature provides several solutions for exploring missings visually by incorporating them into the plot in some way. For example, imputing values to be 10\% below the minimum to include all observations into a scatter plot \citep{Cook2007}, or treating missing values as an equal category of data \citep{Unwin1996}. The ideas from the graphics literature need to be translated into tidy data tools to integrate missing data handling in a data analysis pipeline.

This paper is organized in the following way. Section \ref{background} provides the background to tidy data principles and tools (\ref{tidy-data-concepts}) and missing data representations (\ref{missing-data-rep-dep}). Section \ref{existing-software} summarises existing software for handling missing data. The new work extending the tidy tools to facilitate exploring, visualizing, and imputing missing data is discussed in Section \ref{extensions}. Graphics (\ref{graphics}) and Numerical summaries (\ref{num-sum}) of missing values are then discussed. An application illustrating the use of the new methods is shown in Section \ref{case-study}. Section \ref{discussion} discusses strengths, limitations, and future directions.

\hypertarget{background}{%
\section{Background}\label{background}}

\hypertarget{tidy-data-concepts}{%
\subsection{Tidy data concepts and methods}\label{tidy-data-concepts}}

Features of tidy data were formally described in \citet{Wickham2014}, and were discussed in terms of their importance for data science by \citet{Donoho2017}, and tools for data analysis. Tidy data is defined by \citet{r4ds} as:

\begin{quote}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  each variable must have its own column.
\item
  each observation must have its own row.
\item
  each value must have its own cell.
\end{enumerate}
\end{quote}

Tidy data is easier to work with and analyze because the variables are in the same format as they would be put into modelling software. This helps the analyst works swiftly and clearly, closing up opportunities for errors. Tidy data principles are general, but are comprehensively implemented in the \proglang{R} programming language, so this paper focuses on \proglang{R} \citep{rcore}.

Tidy tools require the same tidy data input and output. This consistency means multiple tools can be composed together into a sequence, allowing for rapid, elegant, and complex operations. Contrasting tidy tools are messy tools. These have tidy input but messy output. Messy tools slow down analysis by shifting the focus from analysis to transforming output so it is the right shape for the next step in the analysis. This makes the work at each step harder to predict, and more complex and difficult to maintain. This disrupts workflow, and invites errors. Tidy tools fall into three broad categories: data manipulation, visualization, and modelling.

\hypertarget{tidy-data-manip}{%
\subsubsection{Data manipulation}\label{tidy-data-manip}}

Data manipulation is made input- and output-tidy with \proglang{R} packages \pkg{dplyr} and \pkg{tidyr} \citep{dplyr, tidyr}. These provide the five ``verbs'' of data manipulation: data reshaping, sorting, filtering, transforming, and aggregating. Data reshaping goes from long to wide formats; sorting arranges rows in a specific order; filtering removes rows based on a condition; transforming, changes existing variables or adds new ones; aggregating creates a single value from many values, say, for example, in computing the minimum, maximum, and mean.

\hypertarget{tidy-vis}{%
\subsubsection{Visualizations}\label{tidy-vis}}

Visualization tools only have tidy data as their input, as the output is a graphic. The popular domain specific language \pkg{ggplot2} maps variables in a dataset to features (referred to as aesthetics) of a graphic \citep{ggplot2}. For example, a scatterplot can be created by mapping two variables to the x and y axes, and specifying a point geometry.

\hypertarget{tidy-model}{%
\subsubsection{Modelling}\label{tidy-model}}

Modelling tools work well with tidy data, as they have a clear mapping from variables in the data to the formula for a model. For example in \proglang{R}, y regressed on x and z is: \code{lm(y ~ x + z)}. Modelling tools are input tidy, but their output is always messy - it is not in the right format for subsequent steps in analysis. For example, estimated coefficients, predictions, and residuals from one model cannot be easily combined with the output of another model. Messy models have been partially addressed with the \pkg{broom} package, which tidies up model outputs into a tidy data format for data analysis, and the developing \pkg{recipes} package, which helps make modelling input- and output-tidy \citep{recipes}.

\hypertarget{the-tidyverse}{%
\subsubsection{Tidy data and tidy tools}\label{the-tidyverse}}

Defining tidy data and tidy tools has resulted in a growing set of packages known collectively as the \pkg{tidyverse} \citep{tidyverse}. These are constructed to share similar principles in their design and behavior, and cover the breadth of an analysis - from importing, tidying, transforming, visualizing, modelling, to communicating \citep{r4ds, tidyverse}. This has led to more tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr}, dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} \citep{readr, readxl, haven, stringr, lubridate, purrr}. It has also led to a burgeoning of new packages for other fields following similar design principles, creating fluid workflows for new domains. For example, the \pkg{tidytext} \citep{tidytext} package for text analysis, the \pkg{tsibble} \citep{wang2020tsibble} package for time series data, and \pkg{tidycensus} \citep{tidycensus} for working with US census and boundary data.

\hypertarget{tidy-formats-missing-data}{%
\subsubsection{Tidy formats for missing data}\label{tidy-formats-missing-data}}

Current tools for missing data are messy. Missing data tools can be used to perform imputations, missing data diagnostics, and data visualizations. However, these tools suffer the same problems as modelling: they use tidy input, but produce messy output - their output is challenging to integrate with other steps of data analysis. The complex, often multivariate nature of imputation methods also makes them difficult to represent. Visualization methods for missing data do not map data features to the aesthetics of a graphic, as in \pkg{ggplot2}, limiting expressive exploration.

Taking existing methods from the missing data graphics literature, and translating and expanding them into tidy data and tidy tools would create more effective data visualizations. Defining these concepts allows the focus to be more general than just software, but rather, an extensible framework for tidy tools to explore missing data.

\hypertarget{missing-data-rep-dep}{%
\subsection{Missing data representation and dependence}\label{missing-data-rep-dep}}

The convention for representing missingness is a \textbf{b}inary matrix, \(B\), for data \(y\) with \(i\) rows and \(j\) columns:

\[
b_{ij} =\begin{cases}
1, & \text{if } y_{ij} \text{ is missing} \\
0, & \text{if } y_{ij} \text{ is observed}
\end{cases}
\]

There are many ways each value can be missing, we adopt the notation used in \citep{vanBuuren2018}. The information in \(B\) can be used to arrive at three categories of missing values: missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR). The distribution of missing values in \(b_{ij}\) can depend on the entire dataset, represented as \(Y = (Y_{obs}, Y_{miss})\). This relationship can be defined by the \emph{missing data model} \(Pr(b_{ij} | Y_{obs}, Y_{miss}, \psi)\), the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, \(\psi\). This helps to precisely define categories of missing values.

\textbf{MCAR} is where values being missing have no association with observed or unobserved data, that is, \(Pr(B = 1 | Y_{obs}, Y_{miss}) = Pr(B = 1 | \psi)\). Essentially, the probability of an observation being missing is unrelated to anything else, only the parameter \(\psi\), the overall probability of missingness. Although a convenient scenario, it is not actually possible to confirm, or clearly distinguish from MAR, as it relies on statements on data unobserved. In \textbf{MAR}, missingness only depends on data observed, not data missing, that is, \(Pr(B = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(B = 1 | Y_{obs}, \psi)\). Some structure or dependence between missing and observed values is allowed, provided it can be explained by data observed, and some overall probability of missingness. In \textbf{MNAR}, missingness is related to values observed, and unobserved: \(Pr(B = 1 | Y_{obs}, Y_{miss}, \psi)\). This assumes conditioning on all observations: data goes missing due to some phenomena unobserved, including the structure of the missing data itself. This presents a challenge in analysis, as it is difficult to verify, and implies bias in analysis due to the unobserved phenomena. Visualizations can help assess whether data may be MCAR, MAR or MNAR.

\hypertarget{existing-software}{%
\section{Existing Software}\label{existing-software}}

Methods for exploring, understanding, and imputing missing data are more accessible now than they have ever been. Values can be imputed with one value (single imputation), or multiple values (multiple imputation), creating \(m\) datasets. This section discusses existing software for single and multiple imputation, and missing data exploration.

\hypertarget{imputation}{%
\subsection{Imputation}\label{imputation}}

\pkg{VIM} \citep{VIM} implements well-used imputation methods: K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation. These diverse approaches allow for imputing with semi-continuous, continuous, count, and categorical data. VIM identifies imputed cases by adding an indicator variable with a suffix ``\code{_imp}''. So, ``\code{Var1}'' has a sibling column, ``\code{Var1_imp}'', with values TRUE or FALSE indicating imputation. \pkg{VIM} also has a variety of visualization methods, discussed in Section \ref{exploration}. \pkg{simputation} provides an interface to imputation methods from \pkg{VIM}, in addition providing hotdeck imputation, and the EM algorithm \citep{simputation, Dempster1977}. \pkg{simputation} provides a consistent formula interface for all imputation methods, and always returns a dataframe with the updated imputations. \pkg{Hmisc} \citep{Hmisc}, provides predictive mean matching, \pkg{imputeTS} \citep{imputeTS}, provides time series imputation methods, and \pkg{missMDA} \citep{missMDA}, imputes data using principal component analysis.

Multiple imputation is often regarded as best practice for imputing values \citep{Schafer2002}, as long as appropriate caution is taken \citep{Sterne2009}. Popular and robust methods for multiple imputation include the \pkg{mice}, \pkg{Amelia}, and \pkg{mi} packages \citep{mice, amelia, mi}. \pkg{mice} implements the method of chained equations, using a variable-wise algorithm to calculate the posterior distribution of parameters to generate imputed values. The workflow in \pkg{mice} revolves around imputing data, returning completed data, and fitting a model and pooling the results.

\pkg{Amelia} \citep{amelia} assumes data are multivariate normal, and samples from the posterior, and allows for incorporation of information on the values in a prior. It uses the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm \citep{Honaker2010}. \pkg{norm} \citep{norm}, provides multiple imputation using EM for multivariate normal data, drawing from methods in the \proglang{NORM} software \citep{schafer-norm}. \pkg{norm} does not provide a framework for tracking missing values, instead providing tools for making inference from multiple imputation.

\pkg{mi} \citep{mi} also uses Bayesian models for imputation, providing better handling of semi-continuous values, and data with structural or perfect correlation. A collection of analysis models are also provided in \pkg{mi}, to work with data it has imputed. These include linear models, generalized linear models, and their corresponding Bayesian components. This approach promotes fluid workflow, with a similar penalty to tidying up model output, which is still messy.

\hypertarget{imputation-summary}{%
\subsubsection{Summary}\label{imputation-summary}}

Each imputation method provides practical methods for different use cases, but most have different output structures, and do not have consistent interfaces in their implementation. This makes them inherently messy and challenging to integrate into an analysis pipeline. For example, combining different imputation methods from different pieces of software is not currently straightforward. \pkg{simputation} resolves some of these complications with a simple approach of a unified syntax for all imputation, and always returns a dataframe of imputed values. This reduces the friction of working with other tools, but comes at the cost of identifying imputed values. An ideal approach would use consistent, simple data structures that work with other analysis tools, and help track missing values. This would make imputation outputs tidy, streamlining subsequent analysis.

\hypertarget{exploration}{%
\subsection{Exploration}\label{exploration}}

Missing data packages are focused primarily on making inferences, and exploring imputed values, not on exploring relationships in missing values, and identifying possible patterns. Texts covering the exploration phase of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools \citep{vanBuuren2018}; this is inefficient. Methods for exploring missing values are primarily covered in literature on interactive graphics \citep{Swayne1998, Unwin1996, Cook2007}, and are picked up again in a discussion of a graphical user interface \citep{Cheng2015}.

The missingness matrix \(B\) can be used to assess missing data dependence. It has been used in interactive graphics, dubbed a ``shadow matrix'', to link missing and imputed values to the data, facilitating their display \citep{Swayne1998}, focusing heavily on multivariate numeric data. This is an idea upon which this new work builds.

The \proglang{MANET} (Missings Are Now Equally Treated) software \citep{Unwin1996} focussed on multivariate categorical data, with missingness explicitly added as a category. \proglang{MANET} also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot. The \proglang{MANET} software is no longer maintained and cannot be installed. The approach of \citep{Swayne1998} in the software \proglang{XGobi}, further developed in \proglang{ggobi} \citep{Cook2007}, focussed on multivariate quantitative data. Missingness is incorporated into plots in \proglang{ggobi} by setting them to be 10\% below the minimum value.

\pkg{MissingDataGUI} provides a Graphical User Interface (GUI) for exploring missing data structure, both numerically and visually. Using a GUI to explore missing data facilitates rapid insight into missingness structures. However, this comes as a trade off, as insights are not captured or recorded with a GUI, making it challenging to incorporate into reproducible analyses. This distracts and breaks analysis workflow, inviting mistakes.

\pkg{VIM} (Visualizing and Imputing Missing Data) provides visualization methods to identify and explore observed, imputed, and missing values. These include spinograms, spinoplots, missingness matrices, plotting missingness in the margins of other plots, and other summaries. However, these visualizations do not map variables to graphical aesthetics, creating friction when moving through analysis workflows, making them difficult to extend to new circumstances. Additionally, data used to create the visualizations cannot be accessed, posing a barrier to further exploration.

The \pkg{inspectdf} package provides tools for summarising, comparing, and visualising dataframes \citep{inspectdf}. In particular the \texttt{inspect\_na()} function provides a summary of missing values for variables. \texttt{inspect\_cat()} summarises the levels of categories in each column in a dataset. The \texttt{show\_plot()} function shows a higher lever overview of the contents of a dataframe and its categories, as well as missing values. All of these functions allow for comparing missing values across two dataframes. For example, comparing the missing values in two dataframes with the same column names.

\pkg{ggplot2} removes missing values with a warning (Figure \ref{fig:warning}), and only incorporates missingness into visualizations when mapping a discrete variable with missings to a graph aesthetic. This has some limitations, shown in Figure \ref{fig:gg-box-na}, a boxplot visualization of school grade and test scores. If there are missings in a continuous variable like test score, \pkg{ggplot2} omits the missings and prints a warning message. However, if a discrete variable like school year has missing values, an NA category is created for school year, where scores are placed.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/gg-box-na-1}

}

\caption[\pkg{ggplot2} provides different visualizations depending on what type of data has missing values for data of student test scores in school year]{\pkg{ggplot2} provides different visualizations depending on what type of data has missing values for data of student test scores in school year. (A) Data is complete; (B) Missings are only in year - an additional "NA" boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and year, additional missing information is shown. The missingness category is only shown when there are missings in categorical variables such as year (plots (B) and (D)). In (C), no missingness information is given on the graphic, despite there being missings in score, and a warning message is displayed about the number of missing values omitted.}\label{fig:gg-box-na}
\end{figure}
\end{CodeChunk}

\hypertarget{approaches-in-other-languages}{%
\subsection{Approaches in other languages}\label{approaches-in-other-languages}}

Missing data can be explored and visualised in closed source software such as \proglang{STATA}, \proglang{SAS}, and \proglang{SPSS}. These provide basic table frequencies of missing data, and basic visualisations on missingness patterns. However, none of these software provide functions to specifically visualise missing values in bivariate or multivariate settings, and it is only possible to visualise missing values after performing data analysis, counting and summarising missingness. Imputation methods from mean to regression, EM, and multiple imputation are all available, with out of the box diagnostics provided to assist comparing imputations. \proglang{python} also provides exploratory data analysis for missing data within \pkg{missingno} \citep{Bilogur2018}, and the \pkg{scikit-learn} package provides imputation methods \citep{scikit-learn}.

\hypertarget{extensions}{%
\section{Tidy framework for missings}\label{extensions}}

Applying \pkg{tidyverse} principles has the potential to clarify missing data exploration, visualization, and imputation. This section discusses how these principles are applied for data structures (\ref{data-structure}), common operations (verbs) (\ref{verbs}), graphics (\ref{graphics}), and data summaries (\ref{num-sum}). Care has been taken to make the names and design these features intuitive for their purpose, and is discussed throughout.

\hypertarget{data-structure}{%
\subsection{Data structure}\label{data-structure}}

A data structure facilitating exploration of missing data needs to be simple to reason with and to transport with a data analysis, otherwise it will not be used. A useful template common in missing data literature, is the \(B\) matrix, where 0 and 1 indicate not missing and missing, respectively. This matrix was used to explore missing values in the interactive graphics library \proglang{XGobi}, called a ``missing value shadow'', or ``shadow matrix'', defined as a copy of the original data with indicator values of missingness. The shadow matrix could be interactively linked to the data. However, there are some limitations to the shadow matrix. Namely, the values 0 and 1 can be confusing representations of missing values, since it is not clear if 0 indicates an absence of observation, or the presence of a missing value.

We propose a new form for tidy representation of missing data based on these ideas from past research. Four features are added to the shadow matrix to facilitate analysis of missing data, illustrated in Figure \ref{fig:nabularfig}. These are briefly described below, and discussed in further detail in section \ref{nabular-data}.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \textbf{Missing value labels}. Simple labels for missing and not missing to clearly identify missing values for analysis and plotting. We propose ``NA'' and ``!NA''. (Figure \ref{fig:nabularfig}). This improves the 0 and 1 values in \(B\), which do not clearly identify missingness. These values follow a principle: ``\textbf{clarity of labelling}'' - the matrix's meaning is transparent, and anybody looking at these values could understand what they mean, which is not the case of binary values. Equally, these values could instead be ``missing'' or ``present''.
\item
  \textbf{Special missing values}: building on \textbf{missing value labels}, the values in the shadow matrix can be ``special'' missing values, indicated by ``NA\_\textless suffix\textgreater{}''. For example: ``NA\_instrument'' uses a short label, ``instrument'', indicating instrument error resulting in missing values. These could be also used to indicate imputations, (Figure \ref{fig:nabularfig}).
\item
  \textbf{Coordinated names}: variable names in the shadow matrix gain a consistent short suffix, "\_NA", keeping names coordinated throughout analysis (Figure \ref{fig:nabularfig}). It makes a clear distinction with ``var\_NA'' being a random variable of the missingness of a variable, ``var''. This suffix is short and easy to remember during data analysis, and shifts the focus from the value of a variable, to its missingness state. Although this might break naming conventions for lower case ``snake\_case'' variables (e.g., \citet{Wickham2014}), this distinguishes these columns from other columns.
\item
  \textbf{Connectedness}: binding the shadow matrix column-wise to the original data as a factor creates a single, connected, nabular data format, in sync with the data, at the cost of a larger file. It is useful for visualization, summaries, and tracking imputed values, discussed in more detail in \ref{nabular-data}.
\end{enumerate}

\hypertarget{nabular-data}{%
\subsection{Nabular data}\label{nabular-data}}

\emph{Nabular} data binds the shadow matrix column-wise to the original data. It is a portmanteau of ``NA'' and ``tabular''. \emph{Nabular} data explicitly links missing values to data, keeping corresponding observations together and removing the possibility of mismatching records (Figure \ref{fig:nabularfig}). \emph{Nabular} data facilitates visualization and summaries by allowing the user to reference the missingness of a variable in a coordinated way: the missingness of ``var'', as ``var\_NA'' during analysis. \emph{Nabular} data is a snapshot of the missingness of the data. This means when \emph{nabular} data are imputed, those imputed values can easily be identified in analysis (``var'' vs ``var\_NA''). Imputing values on \emph{nabular} data automatically tracks these imputations. Together these features make it possible to do tasks that are not possible by simply adding a binary missing matrix, \(B\) during data analysis.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/Users/njtierney/github/njtierney/papers/tidy-missing-data-paper/paper/images/diagram}

}

\caption[The process of creating nabular data]{The process of creating nabular data. Data transformed to shadow matrix, where values are either not missing or missing: "!NA" or "NA". The shadow matrix can be converted to long form to create missingness summary plots. Nabular data is created by binding the columns of the data and shadow matrix. Special missing values (such as -99) are identified as special missings, and values imputed and tracked. Nabular data can be used to identify imputations and explore data values alongside missings, providing a useful format for missing data exploration and analysis.}\label{fig:nabularfig}
\end{figure}
\end{CodeChunk}

\emph{Nabular} data is similar to classical data formats with quality or flag variables associated with each measured variable, e.g., Scripps CO2 data \citep{Keeling2005-scripps}, GHCN data \citep{Durre2008-ghcn}. Using additional columns to represent missingness information follows best practices for data organization, described in \citep{Ellis2018} and \citep{Broman2018}: (1) keep one thing in a cell and (2) describe additional features of variables in a second column. Here they suggest to indicate censored data with an extra variable called ``VariableNameCensored'', which would be TRUE if censored, otherwise FALSE. This information can now be represented in the shadow columns as special missing values. Encoding special missing values is achieved by defining logical conditions and suffixes. This is implemented with the \code{recode_shadow()} function in \pkg{naniar} (Discussed in Section \ref{verbs}).

Special missing values are not a new idea, and have been implemented in other statistical programming languages, \proglang{SPSS}, \proglang{SAS}, and \proglang{STATA}. These typically represent missing values as a full-stop, ``.'', and record special missing values as ``.a'' - ``.z''. These special values from these languages break the tidy principle of one column having one type of value, as they record both the value, and the multivariate missingness state.

\hypertarget{verbs}{%
\subsection{Missing data operations}\label{verbs}}

Common missing data operations can be considered verbs, in the tidyverse sense. For missing data, these include: \textbf{scan}, \textbf{replace}, \textbf{add}, \textbf{shadow}, \textbf{impute}, \textbf{track}, and \textbf{flag}. Data can be \textbf{scanned} to find possible missings not coded as \code{NA}. These values can then be \textbf{replaced} with \code{NA}. To facilitate exploration, summaries of missingness can be \textbf{added} as a summary column to the original data. The data can be augmented with the \textbf{shadow matrix} values, helping explore missing data, as well as facilitating the process of \textbf{imputing}, and \textbf{tracking}. Finally, unusual or specially coded missing values can be \textbf{flagged}.

\hypertarget{verbs-search}{%
\subsubsection{scan: searching for common missing value labels}\label{verbs-search}}

This operation is used to search the data for specific conventional representations of missings, such as, ``N/A'', ``MISSING'', ``-99''. This is implemented in the function \code{miss_scan_count()}, which returns a table of occurrences of that value for each variable. A list of common NA values for numbers and characters, can be provided to help check for typical representations of missings. These are implemented in \pkg{naniar} as objects \code{common_na_numbers} and \code{common_na_strings}.

\hypertarget{verbs-replace-with}{%
\subsubsection{replace: replacing values with missing values}\label{verbs-replace-with}}

Once possible missing values have been identified, these values can be replaced. For example, a dataset could have the values ``-99'' meaning a missing value. \pkg{naniar} implements replacement with the function, \code{replace_with_na()}. Values ``-99'' could be replaced in the ``x'' column with: \code{replace_with_na(dat_ms, replace = list(x = -99))}. For operating on multiple variables, there are scoped variants for \code{replace_with_na()}: \code{_all}, \code{_if}, and \code{_at}. This means \code{replace_with_na_all()} operates on \textbf{all} columns, \code{replace_with_na_at()} operates \textbf{at} specific columns, and \code{replace_with_na_if()} makes a conditional change on columns \textbf{if} they meet some condition (such as \code{is.numeric()} or \code{is.character()}).

\hypertarget{verbs-add-cols}{%
\subsubsection{add: adding missingness summary variables}\label{verbs-add-cols}}

Understanding the missingness structure can be improved by adding summary information alongside the data. For example, in \citep{Tierney2015}, the proportion of missings in a row is used as the outcome in a model to identify variables important in predicting missingness structures. \pkg{naniar} implements a series of functions to add these missingness summaries to the data, starting with \code{add_}. These are inspired by the \code{add_count()} function in \pkg{dplyr}, which adds count information for specified groups or conditions. \pkg{naniar} provides operations to add the number or proportion of missingness, the missingness cluster, or the presence of any missings, with: \code{add_n_miss()}, \code{add_prop_miss()}, \code{add_miss_cluster()}, and \code{add_any_miss()}, respectively (Table \ref{tab:add-missing-info}). There are also functions for adding information about shadow values, and readable labels for any missing values with \code{add_label_shadow()} and \code{add_label_missings()}.

\begin{CodeChunk}
\begin{table}

\caption{\label{tab:add-missing-info}Overview of the "add" functions in naniar}
\centering
\begin{tabular}[t]{l|l}
\hline
Function & Adds column which:\\
\hline
add\_n\_miss(data) & contains the number missing values in a row\\
\hline
add\_any\_miss(data) & contains whether there are any missing values in a row\\
\hline
add\_prop\_miss(data) & contains the proportion of missing values in a row\\
\hline
add\_miss\_cluster(data) & contains the missing value cluster\\
\hline
\end{tabular}
\end{table}

\end{CodeChunk}

\hypertarget{verbs-nabular}{%
\subsubsection{shadow: creating nabular data}\label{verbs-nabular}}

\emph{Nabular} data has the shadow matrix column-bound to existing data. This facilitates visualization and summaries, and allows for imputed values to be tracked. \emph{Nabular} data can be created with \code{nabular()}:

\begin{CodeChunk}
\begin{CodeInput}
R> nabular(dat_ms)
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 5 x 6
      x y         z x_NA  y_NA  z_NA 
  <dbl> <chr> <dbl> <fct> <fct> <fct>
1     1 A      -100 !NA   !NA   !NA  
2     3 N/A     -99 !NA   !NA   !NA  
3    NA <NA>    -98 NA    NA    !NA  
4   -99 E      -101 !NA   !NA   !NA  
5   -98 F        -1 !NA   !NA   !NA  
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{verbs-recode}{%
\subsubsection{flag: describing different types of missing values}\label{verbs-recode}}

Unusual or spurious data values are often identified and ``flagged''. For example, there might be special codes to mark an individual dropping out of a study, known instrument failure in weather instruments, or for values censored in analysis. \pkg{naniar} provides tools to encode these special types of missingness in the shadow matrix of \emph{nabular} data, using \code{recode_shadow()}. This requires specifying the variable to contain the flagged value, the condition for flagging, and a suffix. This is then recoded as a new factor level in the shadow matrix, so every column is aware of all possible new values of missingness. For example, -99 could be recoded to indicate a broken machine sensor for the variable ``x'' with the following:

\begin{CodeChunk}
\begin{CodeInput}
R> nabular(dat_ms) %>%
+   recode_shadow(x = .where(x == -99 ~ "broken_sensor"))
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 5 x 6
      x y         z x_NA             y_NA  z_NA 
* <dbl> <chr> <dbl> <fct>            <fct> <fct>
1     1 A      -100 !NA              !NA   !NA  
2     3 N/A     -99 !NA              !NA   !NA  
3    NA <NA>    -98 NA               NA    !NA  
4   -99 E      -101 NA_broken_sensor !NA   !NA  
5   -98 F        -1 !NA              !NA   !NA  
\end{CodeOutput}
\end{CodeChunk}

\hypertarget{verbs-impute}{%
\subsubsection{impute: imputing values}\label{verbs-impute}}

\pkg{naniar} does not reinvent the wheel by creating the need for customised imputation procedures. Instead, the \code{nabular} format means existing methods using imputation replacement in place now automatically track missing values. \pkg{naniar} provides a few imputation methods to facilitate exploration and visualization: \code{impute_below()}, \code{impute_mean()}, and \code{impute_median()}. While useful to explore structure in missingness, they are not recommended for use in analysis. \code{impute_below()} imputes values below the minimum value, with some controllable jitter (random noise) to reduce overplotting.

Similar to \pkg{simputation}, each \code{impute_} function returns the data with values imputed. However, \pkg{naniar} does not use a formula syntax, instead each function has ``scoped variants'' \code{_all}, \code{_at} and \code{_if}. \code{impute_} functions with no scoped variant, (\code{impute_mean()}), will work on a single vector, but not a data.frame. One challenge with this approach is imputed value locations are not tracked. This issue is resolved with \emph{nabular} data covered in the following section.

\hypertarget{verbs-track}{%
\subsubsection{track: shadow and impute missing values}\label{verbs-track}}

To evaluate imputations they need to be tracked. This is achieved by first using \emph{nabular} data, then imputing, and imputed values can then be referred to by their shadow variable, "\_NA" (Figure \ref{fig:track-impute-example}). The code below shows the track pattern, first using \code{nabular()}, then imputing with \code{impute_lm()}. \code{label_shadow()} then adds a label to facilitate identifying missings:

\begin{CodeChunk}
\begin{CodeInput}
R> aq_imputed <- nabular(airquality) %>%
+   as.data.frame() %>% 
+   simputation::impute_lm(Ozone ~ Temp + Wind) %>%
+   simputation::impute_lm(Solar.R ~ Temp + Wind) %>%
+   add_label_shadow()
R> 
R> head(aq_imputed)
\end{CodeInput}
\begin{CodeOutput}
      Ozone  Solar.R Wind Temp Month Day Ozone_NA Solar.R_NA Wind_NA Temp_NA
1  41.00000 190.0000  7.4   67     5   1      !NA        !NA     !NA     !NA
2  36.00000 118.0000  8.0   72     5   2      !NA        !NA     !NA     !NA
3  12.00000 149.0000 12.6   74     5   3      !NA        !NA     !NA     !NA
4  18.00000 313.0000 11.5   62     5   4      !NA        !NA     !NA     !NA
5 -11.67673 127.4317 14.3   56     5   5       NA         NA     !NA     !NA
6  28.00000 159.5042 14.9   66     5   6      !NA         NA     !NA     !NA
  Month_NA Day_NA any_missing
1      !NA    !NA Not Missing
2      !NA    !NA Not Missing
3      !NA    !NA Not Missing
4      !NA    !NA Not Missing
5      !NA    !NA     Missing
6      !NA    !NA     Missing
\end{CodeOutput}
\end{CodeChunk}

Multiple missing or imputed values can be mapped to a graphical element in ggplot2 by setting the \code{color} or \code{fill} aesthetic in ggplot to \code{any_missing}, a result of the \code{add_label_shadow()} function. (Figure \ref{fig:track-impute-example}). Imputed values can also be compared to complete case data, grouping by \code{any_missing}, and then summarizing, similar to other \pkg{dplyr} summary workflows shown below in Table \ref{tab:impute-summary-out}, showing similarities and differences in imputation methods.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/track-impute-example-1}

}

\caption[Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values from a linear model]{Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values from a linear model. Imputed values are colored green, and data values orange. Imputed values are similar, but slightly trended to the mean.}\label{fig:track-impute-example}
\end{figure}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> aq_imputed %>%
+   group_by(any_missing) %>%
+   summarise_at(.vars = vars(Ozone),
+                .funs = lst(min, mean, median, max)) 
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{table}

\caption{\label{tab:impute-summary-out}Output of \pkg{dplyr} summary statistics of imputed vs non imputed values for the variable "Ozone". The "any\_missing" column denotes imputed values ("Missing", since they were previously missing), and non-imputed values ("Not Missing"). The mean and median values are similar, but the minimum and maximum values are very different.}
\centering
\begin{tabular}[t]{l|r|r|r|r}
\hline
any\_missing & min & mean & median & max\\
\hline
Missing & -16.86418 & 41.22494 & 45.4734 & 78\\
\hline
Not Missing & 1.00000 & 42.09910 & 31.0000 & 168\\
\hline
\end{tabular}
\end{table}

\end{CodeChunk}

\hypertarget{graphics}{%
\section{Graphics}\label{graphics}}

Missing values are often ignored when plotting data - which is why the data visualization software, \proglang{MANET}, was named and is an acronym corresponding to ``Missings Are Now Equally Treated'' \citep{Unwin1996}. However, plots can help to identify the type of missing value patterns, even those of MCAR, MAR or MNAR. Here we summarise how to systematically explore missing patterns visually, and define useful plots to make, relative to the \emph{nabular} data structure.

\hypertarget{overviews}{%
\subsection{Overviews}\label{overviews}}

The first step is to get an overview of the extent of the missingness. Overview visualizations for variables and cases are provided with \code{gg_miss_var()} and \code{gg_miss_case()} (Figure \ref{fig:gg-miss-case-var}A), drawing attention to the amount of missings, and ordering by missingness. The ``airquality'' dataset (from base \proglang{R}), is shown, and contains daily air quality measurements in New York, from May to September, 1973. We learn from Figure \ref{fig:gg-miss-case-var}A-B, that two variables contain missings, approximately one third of observations have one missing value, and a tiny number of observations are missing across two variables. These overview plots are created from the shadow matrix in long form (Figure \ref{fig:nabularfig}). Numerical statistics can also be reported (Section \ref{num-sum}).

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=0.9\linewidth]{tidy-missing-data-paper_files/figure-latex/gg-miss-case-var-1}

}

\caption[Graphical summaries of missingness in the airquality data]{Graphical summaries of missingness in the airquality data. Missings in variables (A) and cases (B), and for an overview of all missingness as a heatmap in (C), and with clustering applied (D). There are missing values in Ozone and Solar.R, with Ozone having more missings. Not many cases have two missings. Most missingness is from cases with one missing value. The default output (C) and ordered by clustering on rows and columns (D). These overviews are made possible using the shadow matrix in long form. There are only missings in ozone and solar radiation, and there appears to be some structure to their missingness.}\label{fig:gg-miss-case-var}
\end{figure}
\end{CodeChunk}

The shadow matrix can be put into long form, allowing both the variables and cases to be displayed using a heatmap style visualization, with \code{vis_miss()} from \pkg{visdat} \citep{visdat} (Figure \ref{fig:nabularfig}). This also provides numerical summaries of missingness in the legend, and for each column (Figure \ref{fig:gg-miss-case-var}C-D). Clustering can be applied to the rows, and columns arranged in order of missingness (Figure \ref{fig:gg-miss-case-var}D). Similar visualizations are available in other packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}. A key improvement is \code{vis_miss()} orients the visualization analogous to a regular data structure: variables form columns and are named at the top, and each row is an observation. Using \pkg{ggplot2}, as the foundation, makes the plot easily customizable.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{tidy-missing-data-paper_files/figure-latex/airquality-upset-1}

}

\caption[The pattern of missingness in the airquality dataset shown in an upset plot]{The pattern of missingness in the airquality dataset shown in an upset plot. Only Ozone and Solar.R have missing values, and Ozone has the most missing values. There are 2 cases where both Solar.R and Ozone have missing values.}\label{fig:airquality-upset}
\end{figure}
\end{CodeChunk}

The number of times observations are missing together can be visualized using an ``upset plot'' \citep{Conway2017}. An alternative to a Venn diagram, an upset plot shows the size and features of overlapping sets, and scales well with more variables. An upset plot can be constructed from the shadow matrix, as shown in Figure \ref{fig:airquality-upset} which displays the overlapping counts of missings in the airquality data. The bottom right shows the combinations of missingness, the top panel shows the size of these combinations, and the bottom left shows missingness in each variable. This provides similar information to Figure \ref{fig:gg-miss-case-var}A-D, but more clearly illustrating overlapping missingness, where 2 cases are missing together in variables ``Solar.R'' and ``Ozone''.

\hypertarget{univariate}{%
\subsection{Univariate}\label{univariate}}

Missing values are typically not shown for univariate visualizations such as histograms or densities. Two ways to use \emph{nabular} data to present univariate data with missings are discussed. The first imputes values below the range to facilitate visualizations. The second displays two plots of the same variable according to the missingness of a chosen variable.

\textbf{Imputing values below the range}. To visualize the amount of missings in each variable, the data is transformed into \emph{nabular} form, then values are imputed below the range of data using \code{impute_below_all()}. Figure \ref{fig:impute-shift-histogram}A shows a histogram of Ozone values on the right in green, and the histogram of missing ozone values on the left, in orange. The missings in Ozone are imputed and ``Ozone\_NA'' is mapped to the fill aesthetic. \emph{Nabular} data facilitates adding counts of missingness to a histogram, allowing examination of a variable's distribution of values, and also the magnitude of missings.

\textbf{Univariate split by missingness}. The distribution of a variable can be shown according to the missingness of another variable. The shadow matrix part of \emph{nabular} is used to handle the faceting, and color mapping. Figure \ref{fig:impute-shift-histogram} shows the values of temperature when ozone is present, and missing, using a faceted histogram (B), and an overlaid density (C). This shows how values of temperature are affected by the missingness of ozone, and reveals a cluster of low temperature observations with missing ozone values. This type of plot can facilitate exploring missing data distributions. For example, we would expect if data were MCAR, for values to be roughly uniformly missing throughout the histogram or density.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=0.75\linewidth]{tidy-missing-data-paper_files/figure-latex/impute-shift-histogram-1} \includegraphics[width=0.75\linewidth]{tidy-missing-data-paper_files/figure-latex/impute-shift-histogram-2} \includegraphics[width=0.75\linewidth]{tidy-missing-data-paper_files/figure-latex/impute-shift-histogram-3}

}

\caption[Univariate summaries of missingness]{Univariate summaries of missingness. (A) A histogram using nabular data to show the values and missings in ozone. Values are imputed below the range to show the number of missings in ozone and colored according to missingness of ozone (`Ozone\_NA`). There are about 35 missings in Ozone. Panel C shows temperature according to missingness in ozone from in the airquality dataset. A histogram of temperature facetted by the missingness of ozone (B), or a density of temperature colored by missingness in ozone (C). These show a cluster of low temperature observations with missing ozone values, but temperature is otherwise similar.}\label{fig:impute-shift-histogram}
\end{figure}
\end{CodeChunk}

\hypertarget{bivariate}{%
\subsection{Bivariate}\label{bivariate}}

To visualize missing values in two dimensions the missing values can be placed in plot margins, by imputing values below the range of the data. Using \emph{nabular} data identifies imputed values, and color makes missingness pre-attentive \citep{treisman1985}. The steps of imputing and coloring are combined into \code{geom_miss_point()}. Figure \ref{fig:geom-miss} shows a mostly uniform spread of missing values for Solar.R and Ozone. As \code{geom_miss_point()} is a defined \pkg{ggplot2} geometry, it works with features such as faceting and mapping other variables to graphical aesthetics.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics{tidy-missing-data-paper_files/figure-latex/geom-miss-1}

}

\caption[Scatterplots with missings displayed at 10 percent below for the airquality dataset]{Scatterplots with missings displayed at 10 percent below for the airquality dataset. Scatterplots of ozone and solar radiation (A), and ozone and temperature (B). There are missings in ozone and solar radiation, but not temperature.}\label{fig:geom-miss}
\end{figure}
\end{CodeChunk}

\hypertarget{multivariate}{%
\subsection{Multivariate}\label{multivariate}}

Parallel coordinate plots can help to visualize missingness beyond two dimensions. They transform variables to the same scale, ranging between 0 and 1. The \code{oceanbuoys} dataset from \pkg{naniar} is used for this visualization, containing measurements of sea and air temperature, humidity, and east west and north south wind speeds. Data was collected in 1993 and 1997, to understand and predict El Niño and El Niña. Figure \ref{fig:parallel-cord-plot} is a parallel coordinate plot of \code{oceanbuoys}, with missing values imputed to be 10\% below the range, and values colored according to whether humidity was missing (``humidity\_NA''). Figure \ref{fig:parallel-cord-plot} shows humidity is missing at low air and sea temperatures, and humidity is missing in one year, and one location.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/parallel-cord-plot-1}

}

\caption[Parallel coordinate plot shows missing values imputed 10\% below range for the oceanbuoys dataset]{Parallel coordinate plot shows missing values imputed 10\% below range for the oceanbuoys dataset. Values are colored by missingness of humidity. Humidity is missing for low air and sea temperatures, and is missing for one year and one location.}\label{fig:parallel-cord-plot}
\end{figure}
\end{CodeChunk}

\hypertarget{num-sum}{%
\section{Numerical summaries}\label{num-sum}}

This section describes approaches to summarizing missingness, and an implementation in \pkg{naniar}. Numerical summaries should be easy to remember with consistent names and output, returning either a single number \ref{single-num-sum}, or a dataframe \ref{sum-tab-missings}, so they integrate well with plotting and modelling tools. How these work with other tools in an analysis pipeline is shown in \ref{num-sum-w-group}.

\hypertarget{single-num-sum}{%
\subsection{Single number summaries}\label{single-num-sum}}

The overall number, proportion, or percent of missing values in a dataset should be simple to calculate. \pkg{naniar} provides the functions \code{n_miss()}, \code{prop_miss()} and \code{pct_miss()}, as well as their complements. Summaries for variables and cases are made by appending \code{_case} or \code{_var} to these summaries. An overview is shown in Table \ref{tab:n-prop-pct-miss-complete}.

\begin{CodeChunk}
\begin{table}

\caption{\label{tab:n-prop-pct-miss-complete}Single number summaries of missingness and completeness of the airquality dataset. The functions follow consistent naming, making them easy to remember, and their use clear.}
\centering
\begin{tabular}[t]{l|r|l|r}
\hline
Missing function & Missing value & Complete function & Complete value\\
\hline
n\_miss & 44.00 & n\_complete & 874.00\\
\hline
prop\_miss & 0.05 & prop\_complete & 0.95\\
\hline
pct\_miss & 4.79 & pct\_complete & 95.21\\
\hline
pct\_miss\_case & 27.45 & prop\_complete\_case & 72.55\\
\hline
pct\_miss\_var & 33.33 & pct\_complete\_var & 66.67\\
\hline
\end{tabular}
\end{table}

\end{CodeChunk}

\hypertarget{sum-tab-missings}{%
\subsection{Summaries and tabulations of missing data}\label{sum-tab-missings}}

Presenting the number and percent of missing values for each variable, or case, provides a summary usable in data handling, or in models to inform imputation. For example, potentially dropping variables, or deciding to include others in an imputation model. Another useful approach is to tabulate the frequency of missing values for each variable or case; that is, the number of times there are zero missings, one missing, two, and so on. These summaries and tabulations are shown for variables in Tables \ref{tab:miss-var-summary} and \ref{tab:miss-var-table}, and implemented with \code{miss_var_summary()} and \code{miss_var_table()}. Case-wise (row-wise) summaries and tabulations are implemented with \code{miss_case_summary()} and \code{miss_case_table()}.

These summaries order rows by the number of missings (\code{n_miss()}), to show the most missings at the top. The number of missings across a repeating span, or finding ``runs'' or ``streaks'' of missings in given variables can also be useful for identifying missingness patterns, and are implemented with \code{miss_var_span()}, and \code{miss_var_run()}.

\begin{CodeChunk}
\begin{table}

\caption{\label{tab:miss-var-summary}\texttt{miss\char`_var\char`_summary(airquality)} provides the number and percent of missings in each variable in airquality. Only ozone and solar radiation have missing values.}
\centering
\begin{tabular}[t]{l|r|r}
\hline
variable & n\_miss & pct\_miss\\
\hline
Ozone & 37 & 24.2\\
\hline
Solar.R & 7 & 4.6\\
\hline
Wind & 0 & 0.0\\
\hline
Temp & 0 & 0.0\\
\hline
Month & 0 & 0.0\\
\hline
Day & 0 & 0.0\\
\hline
\end{tabular}
\end{table}

\end{CodeChunk}

\begin{CodeChunk}
\begin{table}

\caption{\label{tab:miss-var-table}The output of \code{miss\char`_var\char`_table(airquality)}, tabulating the amount of missing data in each variable in airquality. This shows the number of variables with 0, 7, and 37 missings, and the percentage of variables with those amounts of missingness. There are few missingness patterns.}
\centering
\begin{tabular}[t]{r|r|r}
\hline
n\_miss\_in\_var & n\_vars & pct\_vars\\
\hline
0 & 4 & 66.7\\
\hline
7 & 1 & 16.7\\
\hline
37 & 1 & 16.7\\
\hline
\end{tabular}
\end{table}

\end{CodeChunk}

\hypertarget{num-sum-w-group}{%
\subsection{Combining numerical summaries with grouping operations}\label{num-sum-w-group}}

It is useful to explore summaries and tabulations within groups of a dataset. \pkg{naniar} works with \pkg{dplyr}'s \code{group_by()} operator to produce grouped summaries, which work well with the ``pipe'' operator. Table \ref{tab:group-miss-var-summary} below shows an example of missing data summaries for airquality, grouped by month.

\begin{CodeChunk}
\begin{table}

\caption{\label{tab:group-miss-var-summary}Output of \code{airquality \%>\% group\char`_by(Month) \%>\% miss\char`_var\char`_summary()} provides a grouped summary of the missingness in each variable, for each month of the airquality dataset. Only the first 10 rows are shown. There are more ozone missings in June than May.}
\centering
\begin{tabular}[t]{r|l|r|r}
\hline
Month & variable & n\_miss & pct\_miss\\
\hline
5 & Ozone & 5 & 16.1\\
\hline
5 & Solar.R & 4 & 12.9\\
\hline
5 & Wind & 0 & 0.0\\
\hline
5 & Temp & 0 & 0.0\\
\hline
5 & Day & 0 & 0.0\\
\hline
6 & Ozone & 21 & 70.0\\
\hline
6 & Solar.R & 0 & 0.0\\
\hline
6 & Wind & 0 & 0.0\\
\hline
6 & Temp & 0 & 0.0\\
\hline
6 & Day & 0 & 0.0\\
\hline
\end{tabular}
\end{table}

\end{CodeChunk}

\hypertarget{case-study}{%
\section{Application}\label{case-study}}

This section shows how the methods described so far are used together in a data analysis workflow. We analyse a case study of housing data for the city of Melbourne from January 28, 2016 to March 17, 2018. The data was compiled by scraping weekly property clearance data \citep{Kaggle-2018-data}. There are 27,247 properties, and 21 variables in the dataset. The variables include real estate type (town house, unit, house), suburb, selling method, number of rooms, price, real estate agent, sale date, and distance from the Central Business District (CBD).

The goal in analyzing this data is to accurately predict Melbourne housing prices. The data contains many missing values. As a precursor to building a predictive model, this analysis focuses on understanding the patterns of missingness.

\hypertarget{case-study-explore-pattern}{%
\subsection{Exploring patterns of missingness}\label{case-study-explore-pattern}}

Figure \ref{fig:housing-miss-case-var}A shows 9 variables with missing values. The most missings are in ``building area'', followed by ``year built'', and ``land size'', with similar amounts of missingness in ``Car'', ``bathroom'', ``bedroom2'', ``longitude'', and ``latitude''. Figure \ref{fig:housing-miss-case-var}B reveals there are up to 50\% missing values in cases, and the majority of cases have more than 5\% values missing. The variables ``building area'' and ``year built'' have more than 50\% missing data, and so could perhaps be omitted from subsequent analysis, as imputed values are likely to be spurious. Three missingness clusters are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \ref{fig:applic-vis-miss}).

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/housing-miss-case-var-1}

}

\caption[The amount of missings in variables (A) and cases (B) for Melbourne housing data]{The amount of missings in variables (A) and cases (B) for Melbourne housing data. (A) Build area and year built have more than 50\% missing, and car, bathroom, bedroom2 and longitude and latitude have about 25\% missings. (B) Cases are missing 5 - 50\% of values. The majority of missingness is in selected cases and variables.}\label{fig:housing-miss-case-var}
\end{figure}
\end{CodeChunk}

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{tidy-missing-data-paper_files/figure-latex/applic-vis-miss-1}

}

\caption[Heatmap of clustered missingness for housing data reveals structured missingness]{Heatmap of clustered missingness for housing data reveals structured missingness. Three groups of missingness are apparent. At the top: building area to longitude; the middle: building area and year built; the end: building area, year built, and landsize.}\label{fig:applic-vis-miss}
\end{figure}
\end{CodeChunk}

Figure \ref{fig:housing-upset} shows missingness patterns with an ``upset'' plot \citep{Conway2017}, displaying 8 intersecting sets of missing variables. Two patterns stand out: two, and five variables missing, providing further evidence of the missingness patterns seen in Figures \ref{fig:housing-miss-case-var} and \ref{fig:housing-upset}.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/housing-upset-1}

}

\caption[An upset plot of 8 sets of missingness in the housing data]{An upset plot of 8 sets of missingness in the housing data. Missingness for each variable is shown on the bottom left. Connected dots show co-occurences of missings in variables. Two missingness patterns are clear, year built and building area, and lattitude through to building area.}\label{fig:housing-upset}
\end{figure}
\end{CodeChunk}

Tabulating the number of missings in variables in Table \ref{tab:housing-miss-var-case-table} (left) shows three groups of missingness. Tabulating missings in cases (Table \ref{tab:housing-miss-var-case-table} (right)) shows six patterns of missingness. These overview plots lead to the removal of two variables from with more than 50\% missingness from analysis: ``building area'' and ``year built''.

\begin{CodeChunk}
\begin{table}
\caption{\label{tab:housing-miss-var-case-table}Summary tables to help understand missingness patters. Output of \code{miss\_var\_table(housing)} (left), tabulating missingness for variables, and output of \code{miss\_case\_table(housing)}. There are 13 variables with 0-3 missings, 6 variables have 6000-10000 missings, 2 variables have 15000 – 17000 missings. About 30\% of cases have no missings, 45\% of cases have 1 - 6 missings, and about 23\% of cases have 8 or more missings. There are different patterns of missingness in variables and cases, but they can be broken down into smaller groups.}

\centering
\begin{tabular}[t]{r|r|r}
\hline
n\_miss\_in\_var & n\_vars & pct\_vars\\
\hline
0 & 10 & 47.6\\
\hline
1 & 2 & 9.5\\
\hline
3 & 1 & 4.8\\
\hline
6254 & 2 & 9.5\\
\hline
6441 & 1 & 4.8\\
\hline
6447 & 1 & 4.8\\
\hline
6824 & 1 & 4.8\\
\hline
9265 & 1 & 4.8\\
\hline
15163 & 1 & 4.8\\
\hline
16591 & 1 & 4.8\\
\hline
\end{tabular}
\centering
\begin{tabular}[t]{r|r|r}
\hline
n\_miss\_in\_case & n\_cases & pct\_cases\\
\hline
0 & 8887 & 32.6\\
\hline
1 & 3237 & 11.9\\
\hline
2 & 7231 & 26.5\\
\hline
3 & 1370 & 5.0\\
\hline
4 & 79 & 0.3\\
\hline
5 & 8 & 0.0\\
\hline
6 & 203 & 0.7\\
\hline
8 & 6229 & 22.9\\
\hline
9 & 2 & 0.0\\
\hline
11 & 1 & 0.0\\
\hline
\end{tabular}
\end{table}

\end{CodeChunk}

\hypertarget{case-study-explore-for-imp}{%
\subsection{Exploring missingness patterns for imputation}\label{case-study-explore-for-imp}}

Using information from \ref{case-study-explore-pattern}, the following variables are explored for features predicting missingness: ``latitude'', ``longitude'', ``bedroom2'', ``bathroom'', ``car'', and ``land size''.

Missingness structure is explored by clustering the missing values into groups. Then, a classification and regression trees (CART) model is applied to predict these missingness clusters using the remaining variables \citep{Tierney2015, Barnett2017}. Two clusters of missingness are identified and predicted using all variables in the dataset with the CART package \pkg{rpart} \citep{rpart}, and plotted using the \pkg{rpart.plot} package \citep{rpart-plot}. Importance scores reveal the following variables as most important in predicting missingness: ``rooms'', ``price'', ``suburb'', ``council area'', ``distance'', and ``region name''. These variables are important for predicting missingness, so are included in the imputation model.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=0.9\linewidth]{tidy-missing-data-paper_files/figure-latex/rpart-plot-1}

}

\caption[Decision tree output predicting missingness clusters]{Decision tree output predicting missingness clusters. Type of house, year quarter, and year were important for predicting missingness cluster. The cluster with the most missingness was for quarters 1 and 4, for 2017 and 2018. Type of house, year, and year quarter are important features related to missingness structure.}\label{fig:rpart-plot}
\end{figure}
\end{CodeChunk}

\hypertarget{case-study-imp-diagnosis}{%
\subsection{Imputation and diagnostics}\label{case-study-imp-diagnosis}}

\pkg{simputation} is used to implement two imputations: simple linear regression and K nearest neighbors. Values are imputed stepwise in ascending order of missingness. The track missings pattern is applied (described in \ref{verbs}), to assess imputed values. Imputed datasets are compared on their performance in a model predicting log house price for 4 variables (Figure \ref{fig:imputed-by-model}). Compared to KNN imputed values, the linear model imputed values closer to the mean.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=0.95\linewidth]{tidy-missing-data-paper_files/figure-latex/imputed-by-model-1}

}

\caption[Boxplots of complete case data, and data imputed with KNN or linear model for different variables]{Boxplots of complete case data, and data imputed with KNN or linear model for different variables. (A) number of bedrooms, (B) number of bathrooms, (C) number of carspots, and (D) landsize (on a log10 scale). KNN had similar results to complete case, and linear model had a lower median for cars and fewer extreme values for bedrooms.}\label{fig:imputed-by-model}
\end{figure}
\end{CodeChunk}

\hypertarget{case-study-assess-model}{%
\subsection{Assessing model predictions}\label{case-study-assess-model}}

Coefficients of the linear model of log price vary for room for different imputed datasets (Figure \ref{fig:tidy-coefs}). Notably, complete cases underestimate the impact of room on log price. A partial residual plot (Figure \ref{fig:partial-resid}) shows there is not much variation amongst the models from the differently imputed datasets.

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/tidy-coefs-1}

}

\caption[The coefficient estimate for the number of rooms varies according to the imputed dataset]{The coefficient estimate for the number of rooms varies according to the imputed dataset. Complete case dataset produced lower coefficients, compared to imputed datasets.}\label{fig:tidy-coefs}
\end{figure}
\end{CodeChunk}

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/partial-resid-1}

}

\caption[Partial residual plot for each data set, complete cases (cc), and imputed with KNN (knn) or linear model (lm)]{Partial residual plot for each data set, complete cases (cc), and imputed with KNN (knn) or linear model (lm). These are plotted as hex bins, colored according to the number of points in a given hexagon. Brighter colors mean more points. Compared to complete cases, imputed data has more points clustered around zero.}\label{fig:partial-resid}
\end{figure}
\end{CodeChunk}

\hypertarget{case-study-summary}{%
\subsection{Summary}\label{case-study-summary}}

The \pkg{naniar} and \pkg{visdat} packages implement the methods discussed in the paper, building on existing tidy tools and strike a compromise between automation and control, making analysis efficient, readable, but not overly complex. Each tool has clear intent and effects - summarising, plotting or generating data or augmenting data in some way. This not only reduces repetition and typing in an analysis, but most importantly, allows for clear expression of intent, making exploration of missing values fluent.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

This paper has described new methods for exploring, visualizing, and imputing missing data. The work was motivated by recent developments of tidy data, and extends them for better missing value handling. The methods have standard outputs, function arguments, and behavior. This provides consistent workflows centered around data analysis that integrate well with existing imputation methodology, visualization, and modelling.

The \emph{nabular} data structures discussed in the paper are simple by design, to promote flexibility. They could be used to create different visualizations than were shown in the paper. The analyst can use the data structures to decide on appropriate visualization for their problem. The data structures could also be used to support interactive graphics, in the manner of \proglang{MANET} and \proglang{ggobi}. Linking the plots (via linked brushing) could facilitate exploration of missingness, and could be implemented with \pkg{plotly} \citep{plotly} for added interactivity. Animating between different sets of imputed values could also be possible with packages like \pkg{gganimate} \citep{gganimate}.

Other data structures such as spatial data, time series, networks, and longitudinal data would be supported by the inherently tabular, \emph{nabular} data, if they are first structured as wide tidy format. Large data may need special handling, and additional features like efficient storage of purely imputed values and lazy evaluation. Special missing value codes could be improved by creating special classes, or expanding low level representation of \code{NA} at the source code level. Similarly, list columns, present a challenge for analysis as their expansion into vectors in a dataframe is not always possible as they have different lengths, but should still be flagged for further analysis.

The methodology described in this paper can be used in conjunction with other approaches to understand multivariate missingness dependencies (e.g., decision trees \citep{Tierney2015}, latent group analysis \citep{Barnett2017}, and PCA \citep{FactoMineR}). Evaluating imputed values using a testing framework like \citep{vanBuuren2018} is also supported.

The approach meshes with the dynamic nature of data analysis, allowing the analyst to go from raw data to model data in a fluid workflow.

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

The authors would like to thank Miles McBain, for his key contributions and discussions on the \pkg{naniar} package, in particular for helping implement \code{geom_miss_point()}, and for his feedback on ideas, implementations, and names. We also thank Colin Fay for his contributions to the \pkg{naniar} package, in particular for his assistance with the \code{replace_with_na()} functions. We also thank Earo Wang and Mitchell O'Hara-Wild for the many useful discussions on missing data and package development, and for their assistance with creating elegant approaches that take advantage of the tidy syntax. We would also like to thank those who contributed pull requests and discussions on the \pkg{naniar} package, in particular Jim Hester and Romain François for improving the speed of key functions, Ross Gayler for discussion on special missing values, and Luke Smith for helping \pkg{naniar} be more compliant with \pkg{ggplot2}. We would also like to thank Amelia McNamara for discussions on the paper.

\renewcommand\refname{References}
\bibliography{tidy-missing-data-zotero.bib}

\end{document}

### 2022-08-16 paper/tidy-missing-data-paper.Rmd

@@
     includes:
-      in_header: "preamble.tex"
+      header-includes: 
+        - \usepackage{amsmath}
+        - \usepackage[english]{babel}
+        - \usepackage{float}
+        - \usepackage{booktabs}
 bibliography: "tidy-missing-data-zotero.bib"
@@
 })
+```
+
+```{r tar-read-housing}
+housing <- readr::read_rds(here::here("paper/housing.rds"))
 ```

### 2022-08-17 packages.R
add bookdown

@@ -1,4 +1,5 @@
 ## library() calls go here
+library(bookdown)
 library(conflicted)
 library(cowplot)
 library(dotenv)

### 2022-11-30 paper/tidy-missing-data-paper.Rmd
small copyedits

@@
 Together, these provide a connected framework for handling, exploring, and imputing missing values.
-These methods are available in the \proglang{R} package \pkg{naniar}.
+These methods are available in the \proglang{R} package \pkg{naniar} \citep{pkg:naniar}, available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/package=naniar}.
 keywords:
@@
 The ideas from the graphics literature need to be translated into tidy data tools to integrate missing data handling in a data analysis pipeline.
+We have implemented these tools for exploring missing data into the \proglang{R} package \pkg{naniar} \citep{pkg:naniar}, available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/package=naniar}.
 
@@
 $$
-b_{ij} =\begin{cases}
-  1, & \text{if } y_{ij} \text{ is missing} \\
-  0, & \text{if } y_{ij} \text{ is observed}
+b_{ij} = \begin{cases}
+  1, \text{if } y_{ij} \text{ is missing} \\
+  0, \text{if } y_{ij} \text{ is observed}
 \end{cases}
@@
 ```{r make-school-scores-data, echo = FALSE}
-
+set.seed(2022-11-29)
 school_scores <- data.frame(score = c(rnorm(20, 70, 10),

### 2022-12-01 jss-copyedit/2022-01-12-4108/paper.tex
add files that have been copyedited by JSS

--- 2022-08-16 paper/tidy-missing-data-paper-manually-edited-final.tex
@@
-\documentclass[
-]{jss}
+\documentclass[article]{jss}
+
+\usepackage{thumbpdf,lmodern}
+\graphicspath{{Figures/}} % folder with figures
 
@@
 \author{
-Nicholas Tierney\\Monash University \And Dianne Cook\\Monash University
+Nicholas Tierney\\Monash University, \\Telethon Kids Institute
+\And Dianne Cook\\Monash University
 }
-\title{Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations}
+\title{Expanding Tidy Data Principles to Facilitate Missing Data Exploration,
+Visualization and Assessment of Imputations}
 
 \Plainauthor{Nicholas Tierney, Dianne Cook}
-\Plaintitle{Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations}
-\Shorttitle{Explore missingness with \pkg{naniar}}
+\Plaintitle{Expanding Tidy Data Principles to Facilitate Missing Data Exploration,
+Visualization and Assessment of Imputations}
+\Shorttitle{Explore Missingness with \pkg{naniar}}
 
 \Abstract{
-Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data.
-This paper addresses this gap.
-The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as a key part of data analysis workflows.
-We define a new data structure, and a suite of new operations.
-Together, these provide a connected framework for handling, exploring, and imputing missing values.
-These methods are available in the \proglang{R} package \pkg{naniar}.
+Despite the large body of research on missing value distributions and
+imputation, there is comparatively little literature with a focus on how to
+make it easy to handle, explore, and impute missing values in data.  This
+paper addresses this gap.  The new methodology builds upon tidy data
+principles, with the goal of integrating missing value handling as a key
+part of data analysis workflows.  We define a new data structure, and a
+suite of new operations.  Together, these provide a connected framework for
+handling, exploring, and imputing missing values.  These methods are
+available in the \proglang{R}~package \pkg{naniar}. \citep{pkg:naniar}, available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/package=naniar}.
 }
 
-\Keywords{statistical computing, statistical graphics, data science, data visualization, \pkg{tidyverse}, data pipeline, \proglang{R}}
-\Plainkeywords{statistical computing, statistical graphics, data science, data visualization, tidyverse, data pipeline, R}
+\Keywords{statistical computing, statistical graphics, data science,
+data visualization, \pkg{tidyverse}, data pipeline, \proglang{R}}
+\Plainkeywords{statistical computing, statistical graphics, data science,
+data visualization, tidyverse, data pipeline, R}
 
@@
 \Address{
+  Nicholas Tierney\\
+  Monash University\\
+  Melbourne, Victoria, Australia\\
+  \emph{and}\\
+  Telethon Kids Institute\\
+  Perth, Western Australia, Australia\\
+  nicholas.tierney\@gmail.com
+
+  Dianne Cook\\
+  Monash University\\
+  Melbourne, Victoria, Australia\\
   }
@@
 
-As data science has become a more solid field, theories and principles have developed to describe best practices.
-One such idea is ``tidy data'', which defines a clean, analysis-ready format that informs workflows converting raw data through a data analysis pipeline \citep{Wickham2014}.
-Another idea is the grammar of graphics, describing how to map data values into a visualization \citep{Wilkinson2012}.
-These principles have been widely adopted, but do not address the problem of missing data.
-In particular, there is little guidance on how to handle missing values in a data analysis workflow.
-Most analysis and visualization software simply drop missing values when making a plot, although some (\pkg{ggplot2}) provide a warning (Figure \ref{fig:warning}).
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{tidy-missing-data-paper_files/figure-latex/warning-1}
-
-}
-
-\caption[How \pkg{ggplot2} behaves when displaying missing values]{How \pkg{ggplot2} behaves when displaying missing values.
-A warning message is displayed, but missing values are not shown in the plot.}\label{fig:warning}
-\end{figure}
-\end{CodeChunk}
-
-The imputation literature focuses on ensuring valid statistical inference is made from incomplete data.
-This is approached chiefly through probabilistic modelling, assuming the mechanism of missing data is known to the analyst.
-It does not address how to explore, understand, and handle missing data structures and mechanisms.
-
-However, something must be known about the missing value structure to produce a complete dataset for analysis, whether by case- or variable-deletion, or with imputation.
-Decision tree models or latent group analysis can reveal structures or patterns of missingness \citep{Tierney2015, Barnett2017}, but are not definitive.
-While there are times where the missing data mechanism is obvious, determining this is typically not straightforward.
-A straightforward case is where missing values are encoded in the data in a non standard format that still indicates missingness, such as ``N.A.'', ``N/A'', or ``missing''.
-However, identifying these amongst all data can be problematic as it is hard to know what you are looking for, like finding a needle in a haystack.
-A more challenging case is where missing values arise due to the increasing or decreasing values of other variables (including themselves).
-For example, in weather data, temperature values might go missing due to their own high temperature.
-Or similarly, other measurements go missing due to high temperatures.
-Identifying this requires exploring the data, and the analysis path to take is obvious, as missing values are often omitted from plots and other statistical summaries.
-To understand their structure and possible mechanisms, the analyst must explore the data with visualizations, summaries, and modelling, in an iterative fashion.
-
-While there have been many software tools for exploring missing data, they do not work together.
-The graphics literature provides several solutions for exploring missings visually by incorporating them into the plot in some way.
-For example, imputing values to be 10\% below the minimum to include all observations into a scatter plot \citep{Cook2007}, or treating missing values as an equal category of data \citep{Unwin1996}.
-The ideas from the graphics literature need to be translated into tidy data tools to integrate missing data handling in a data analysis pipeline.
-
-This paper is organized in the following way.
-Section \ref{background} provides the background to tidy data principles and tools (\ref{tidy-data-concepts}) and missing data representations (\ref{missing-data-rep-dep}).
-Section \ref{existing-software} summarises existing software for handling missing data.
-The new work extending the tidy tools to facilitate exploring, visualizing, and imputing missing data is discussed in Section \ref{extensions}.
-Graphics (\ref{graphics}) and Numerical summaries (\ref{num-sum}) of missing values are then discussed.
-An application illustrating the use of the new methods is shown in Section \ref{case-study}.
-Section \ref{discussion} discusses strengths, limitations, and future directions.
+As data science has become a more solid field, theories and principles have
+developed to describe best practices.  One such idea is ``tidy data'', which
+defines a clean, analysis-ready format that informs workflows converting raw
+data through a data analysis pipeline \citep{Wickham2014}.  Another idea is
+the grammar of graphics, describing how to map data values into a
+visualization \citep{Wilkinson2012}.  These principles have been widely
+adopted, but do not address the problem of missing data.  In particular,
+there is little guidance on how to handle missing values in a data analysis
+workflow.  Most analysis and visualization software simply drop missing
+values when making a plot, although some (\pkg{ggplot2}) provide a warning
+(Figure~\ref{fig:warning}).
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=0.75\linewidth]{warning-1}
+\caption[How \pkg{ggplot2} behaves when displaying missing values]{How
+\pkg{ggplot2} behaves when displaying missing values.  A warning message is
+displayed, but missing values are not shown in the plot.}\label{fig:warning}
+\end{figure}
+
+The imputation literature focuses on ensuring valid statistical inference is
+made from incomplete data.  This is approached chiefly through probabilistic
+modelling, assuming the mechanism of missing data is known to the analyst.
+It does not address how to explore, understand, and handle missing data
+structures and mechanisms.
+
+However, something must be known about the missing value structure to
+produce a complete dataset for analysis, whether by case- or
+variable-deletion, or with imputation.  Decision tree models or latent group
+analysis can reveal structures or patterns of missingness
+\citep{Tierney2015, Barnett2017}, but are not definitive.  While there are
+times where the missing data mechanism is obvious, determining this is
+typically not straightforward.  A straightforward case is where missing
+values are encoded in the data in a non standard format that still indicates
+missingness, such as ``N.A.'', ``N/A'', or ``missing''.  However,
+identifying these amongst all data can be problematic as it is hard to know
+what you are looking for, like finding a needle in a haystack.  A more
+challenging case is where missing values arise due to the increasing or
+decreasing values of other variables (including themselves).  For example,
+in weather data, temperature values might go missing due to their own high
+temperature.  Or similarly, other measurements go missing due to high
+temperatures.  Identifying this requires exploring the data, and the
+analysis path to take is obvious, as missing values are often omitted from
+plots and other statistical summaries.  To understand their structure and
+possible mechanisms, the analyst must explore the data with visualizations,
+summaries, and modelling, in an iterative fashion.
+
+While there have been many software tools for exploring missing data, they
+do not work together.  The graphics literature provides several solutions
+for exploring missings visually by incorporating them into the plot in some
+way.  For example, imputing values to be 10\% below the minimum to include
+all observations into a scatter plot \citep{Cook2007}, or treating missing
+values as an equal category of data \citep{Unwin1996}.  The ideas from the
+graphics literature need to be translated into tidy data tools to integrate
+missing data handling in a data analysis pipeline.
+We have implemented these tools for exploring missing data into the \proglang{R} package \pkg{naniar} \citep{pkg:naniar}, available from the Comprehensive \proglang{R} Archive Network (CRAN) at \url{https://CRAN.R-project.org/package=naniar}.
+
+This paper is organized in the following way.  Section~\ref{background}
+provides the background to tidy data principles and tools
+(Section~\ref{tidy-data-concepts}) and missing data representations
+(Section~\ref{missing-data-rep-dep}).  Section~\ref{existing-software} summarizes
+existing software for handling missing data.  The new work extending the
+tidy tools to facilitate exploring, visualizing, and imputing missing data
+is discussed in Section~\ref{extensions}.  Graphics (Section~\ref{graphics}) and
+Numerical summaries (Section~\ref{num-sum}) of missing values are then discussed.
+An application illustrating the use of the new methods is shown in
+Section~\ref{case-study}.  Section~\ref{discussion} discusses strengths,
+limitations, and future directions.
 
@@
 
-Features of tidy data were formally described in \citet{Wickham2014}, and were discussed in terms of their importance for data science by \citet{Donoho2017}, and tools for data analysis.
-Tidy data is defined by \citet{r4ds} as:
-
-\begin{quote}
+Features of tidy data were formally described in \citet{Wickham2014}, and
+were discussed in terms of their importance for data science by
+\citet{Donoho2017}, and tools for data analysis.  Tidy data is defined by
+\citet{r4ds} as:
+%
+% \begin{quote}
 \begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
+% \def\labelenumi{\arabic{enumi}.}
+% \tightlist
 \item
@@
 \end{enumerate}
-\end{quote}
-
-Tidy data is easier to work with and analyze because the variables are in the same format as they would be put into modelling software.
-This helps the analyst works swiftly and clearly, closing up opportunities for errors.
-Tidy data principles are general, but are comprehensively implemented in the \proglang{R} programming language, so this paper focuses on \proglang{R} \citep{rcore}.
-
-Tidy tools require the same tidy data input and output.
-This consistency means multiple tools can be composed together into a sequence, allowing for rapid, elegant, and complex operations.
-Contrasting tidy tools are messy tools.
-These have tidy input but messy output.
-Messy tools slow down analysis by shifting the focus from analysis to transforming output so it is the right shape for the next step in the analysis.
-This makes the work at each step harder to predict, and more complex and difficult to maintain.
-This disrupts workflow, and invites errors.
-Tidy tools fall into three broad categories: data manipulation, visualization, and modelling.
+% \end{quote}
+%
+Tidy data is easier to work with and analyze because the variables are in
+the same format as they would be put into modelling software.  This helps
+the analyst works swiftly and clearly, closing up opportunities for errors.
+Tidy data principles are general, but are comprehensively implemented in the
+\proglang{R} programming language, so this paper focuses on \proglang{R}
+\citep{rcore}.
+
+Tidy tools require the same tidy data input and output.  This consistency
+means multiple tools can be composed together into a sequence, allowing for
+rapid, elegant, and complex operations.  Contrasting tidy tools are messy
+tools.  These have tidy input but messy output.  Messy tools slow down
+analysis by shifting the focus from analysis to transforming output so it is
+the right shape for the next step in the analysis.  This makes the work at
+each step harder to predict, and more complex and difficult to maintain.
+This disrupts workflow, and invites errors.  Tidy tools fall into three
+broad categories: data manipulation, visualization, and modelling.
 
@@
 
-Data manipulation is made input- and output-tidy with \proglang{R} packages \pkg{dplyr} and \pkg{tidyr} \citep{dplyr, tidyr}.
-These provide the five ``verbs'' of data manipulation: data reshaping, sorting, filtering, transforming, and aggregating.
-Data reshaping goes from long to wide formats; sorting arranges rows in a specific order; filtering removes rows based on a condition; transforming, changes existing variables or adds new ones; aggregating creates a single value from many values, say, for example, in computing the minimum, maximum, and mean.
+Data manipulation is made input- and output-tidy with \proglang{R}~packages
+\pkg{dplyr} and \pkg{tidyr} \citep{dplyr, tidyr}.  These provide the five
+``verbs'' of data manipulation: data reshaping, sorting, filtering,
+transforming, and aggregating.  Data reshaping goes from long to wide
+formats; sorting arranges rows in a specific order; filtering removes rows
+based on a condition; transforming, changes existing variables or adds new
+ones; aggregating creates a single value from many values, say, for example,
+in computing the minimum, maximum, and mean.
 
@@
 
-Visualization tools only have tidy data as their input, as the output is a graphic.
-The popular domain specific language \pkg{ggplot2} maps variables in a dataset to features (referred to as aesthetics) of a graphic \citep{ggplot2}.
-For example, a scatterplot can be created by mapping two variables to the x and y axes, and specifying a point geometry.
+Visualization tools only have tidy data as their input, as the output is a
+graphic.  The popular domain specific language \pkg{ggplot2} maps variables
+in a dataset to features (referred to as aesthetics) of a graphic
+\citep{ggplot2}.  For example, a scatterplot can be created by mapping two
+variables to the $x$ and $y$~axes, and specifying a point geometry.
 
@@
 
-Modelling tools work well with tidy data, as they have a clear mapping from variables in the data to the formula for a model.
-For example in \proglang{R}, y regressed on x and z is: \code{lm(y ~ x + z)}.
-Modelling tools are input tidy, but their output is always messy - it is not in the right format for subsequent steps in analysis.
-For example, estimated coefficients, predictions, and residuals from one model cannot be easily combined with the output of another model.
-Messy models have been partially addressed with the \pkg{broom} package, which tidies up model outputs into a tidy data format for data analysis, and the developing \pkg{recipes} package, which helps make modelling input- and output-tidy \citep{recipes}.
+Modelling tools work well with tidy data, as they have a clear mapping from
+variables in the data to the formula for a model.  For example in
+\proglang{R}, $y$ regressed on $x$ and $z$ is: \code{lm(y ~ x + z)}.  Modelling
+tools are input tidy, but their output is always messy -- it is not in the
+right format for subsequent steps in analysis.  For example, estimated
+coefficients, predictions, and residuals from one model cannot be easily
+combined with the output of another model.  Messy models have been partially
+addressed with the \pkg{broom} package, which tidies up model outputs into a
+tidy data format for data analysis, and the developing \pkg{recipes}
+package, which helps make modelling input- and output-tidy \citep{recipes}.
 
@@
 
-Defining tidy data and tidy tools has resulted in a growing set of packages known collectively as the \pkg{tidyverse} \citep{tidyverse}.
-These are constructed to share similar principles in their design and behavior, and cover the breadth of an analysis - from importing, tidying, transforming, visualizing, modelling, to communicating \citep{r4ds, tidyverse}.
-This has led to more tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr}, dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} \citep{readr, readxl, haven, stringr, lubridate, purrr}.
-It has also led to a burgeoning of new packages for other fields following similar design principles, creating fluid workflows for new domains.
-For example, the \pkg{tidytext} \citep{tidytext} package for text analysis, the \pkg{tsibble} \citep{wang2020tsibble} package for time series data, and \pkg{tidycensus} \citep{tidycensus} for working with US census and boundary data.
+Defining tidy data and tidy tools has resulted in a growing set of packages
+known collectively as the \pkg{tidyverse} \citep{tidyverse}.  These are
+constructed to share similar principles in their design and behavior, and
+cover the breadth of an analysis - from importing, tidying, transforming,
+visualizing, modelling, to communicating \citep{r4ds, tidyverse}.  This has
+led to more tools for specific parts of analysis - from reading in data with
+\pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings
+with \pkg{stringr}, dates with \pkg{lubridate}, and performing functional
+programming with \pkg{purrr} \citep{readr, readxl, haven, stringr,
+lubridate, purrr}.  It has also led to a burgeoning of new packages for
+other fields following similar design principles, creating fluid workflows
+for new domains.  For example, the \pkg{tidytext} \citep{tidytext} package
+for text analysis, the \pkg{tsibble} \citep{wang2020tsibble} package for
+time series data, and \pkg{tidycensus} \citep{tidycensus} for working with
+US census and boundary data.
 
@@
 
-Current tools for missing data are messy.
-Missing data tools can be used to perform imputations, missing data diagnostics, and data visualizations.
-However, these tools suffer the same problems as modelling: they use tidy input, but produce messy output - their output is challenging to integrate with other steps of data analysis.
-The complex, often multivariate nature of imputation methods also makes them difficult to represent.
-Visualization methods for missing data do not map data features to the aesthetics of a graphic, as in \pkg{ggplot2}, limiting expressive exploration.
-
-Taking existing methods from the missing data graphics literature, and translating and expanding them into tidy data and tidy tools would create more effective data visualizations.
-Defining these concepts allows the focus to be more general than just software, but rather, an extensible framework for tidy tools to explore missing data.
+Current tools for missing data are messy.  Missing data tools can be used to
+perform imputations, missing data diagnostics, and data visualizations.
+However, these tools suffer the same problems as modelling: they use tidy
+input, but produce messy output -- their output is challenging to integrate
+with other steps of data analysis.  The complex, often multivariate nature
+of imputation methods also makes them difficult to represent.  Visualization
+methods for missing data do not map data features to the aesthetics of a
+graphic, as in \pkg{ggplot2}, limiting expressive exploration.
+
+Taking existing methods from the missing data graphics literature, and
+translating and expanding them into tidy data and tidy tools would create
+more effective data visualizations.  Defining these concepts allows the
+focus to be more general than just software, but rather, an extensible
+framework for tidy tools to explore missing data.
 
@@
 
-The convention for representing missingness is a \textbf{b}inary matrix, \(B\), for data \(y\) with \(i\) rows and \(j\) columns:
-
+The convention for representing missingness is a \textbf{b}inary matrix,
+\(B\), for data \(y\) with \(i\) rows and \(j\) columns:
+%
 \[
@@
 \]
-
-There are many ways each value can be missing, we adopt the notation used in \citep{vanBuuren2018}.
-The information in \(B\) can be used to arrive at three categories of missing values: missing completely at random (MCAR), missing at random (MAR), and missing not at random (MNAR).
-The distribution of missing values in \(b_{ij}\) can depend on the entire dataset, represented as \(Y = (Y_{obs}, Y_{miss})\).
-This relationship can be defined by the \emph{missing data model} \(Pr(b_{ij} | Y_{obs}, Y_{miss}, \psi)\), the probability of missingness is conditional on data observed, data missing, and some probability parameter of missingness, \(\psi\).
-This helps to precisely define categories of missing values.
-
-\textbf{MCAR} is where values being missing have no association with observed or unobserved data, that is, \(Pr(B = 1 | Y_{obs}, Y_{miss}) = Pr(B = 1 | \psi)\).
-Essentially, the probability of an observation being missing is unrelated to anything else, only the parameter \(\psi\), the overall probability of missingness.
-Although a convenient scenario, it is not actually possible to confirm, or clearly distinguish from MAR, as it relies on statements on data unobserved.
-In \textbf{MAR}, missingness only depends on data observed, not data missing, that is, \(Pr(B = 1 | Y_{obs}, Y_{miss}, \psi) = Pr(B = 1 | Y_{obs}, \psi)\).
-Some structure or dependence between missing and observed values is allowed, provided it can be explained by data observed, and some overall probability of missingness.
-In \textbf{MNAR}, missingness is related to values observed, and unobserved: \(Pr(B = 1 | Y_{obs}, Y_{miss}, \psi)\).
-This assumes conditioning on all observations: data goes missing due to some phenomena unobserved, including the structure of the missing data itself.
-This presents a challenge in analysis, as it is difficult to verify, and implies bias in analysis due to the unobserved phenomena.
-Visualizations can help assess whether data may be MCAR, MAR or MNAR.
+%
+There are many ways each value can be missing, we adopt the notation used in
+\citep{vanBuuren2018}.  The information in \(B\) can be used to arrive at
+three categories of missing values: missing completely at random (MCAR),
+missing at random (MAR), and missing not at random (MNAR).  The distribution
+of missing values in \(b_{ij}\) can depend on the entire dataset,
+represented as \(Y = (Y_\text{obs}, Y_\text{miss})\).  This relationship can be
+defined by the \emph{missing data model} \(\Prob(b_{ij} \mid Y_\text{obs}, Y_\text{miss},
+\psi)\), the probability of missingness is conditional on data observed,
+data missing, and some probability parameter of missingness, \(\psi\).  This
+helps to precisely define categories of missing values.
+
+MCAR is where values being missing have no association with
+observed or unobserved data, that is, \(\Prob(B = 1 \mid Y_\text{obs}, Y_\text{miss}) = \Prob(B
+= 1 \mid \psi)\).  Essentially, the probability of an observation being missing
+is unrelated to anything else, only the parameter \(\psi\), the overall
+probability of missingness.  Although a convenient scenario, it is not
+actually possible to confirm, or clearly distinguish from MAR, as it relies
+on statements on data unobserved.  In MAR, missingness only depends
+on data observed, not data missing, that is, \(\Prob(B = 1 \mid Y_\text{obs}, Y_\text{miss},
+\psi) = \Prob(B = 1 \mid Y_\text{obs}, \psi)\).  Some structure or dependence between
+missing and observed values is allowed, provided it can be explained by data
+observed, and some overall probability of missingness.  In MNAR,
+missingness is related to values observed, and unobserved: \(\Prob(B = 1 \mid
+Y_\text{obs}, Y_\text{miss}, \psi)\).  This assumes conditioning on all observations:
+data goes missing due to some phenomena unobserved, including the structure
+of the missing data itself.  This presents a challenge in analysis, as it is
+difficult to verify, and implies bias in analysis due to the unobserved
+phenomena.  Visualizations can help assess whether data may be MCAR, MAR or
+MNAR.
 
@@
 
-Methods for exploring, understanding, and imputing missing data are more accessible now than they have ever been.
-Values can be imputed with one value (single imputation), or multiple values (multiple imputation), creating \(m\) datasets.
-This section discusses existing software for single and multiple imputation, and missing data exploration.
+Methods for exploring, understanding, and imputing missing data are more
+accessible now than they have ever been.  Values can be imputed with one
+value (single imputation), or multiple values (multiple imputation),
+creating \(m\) datasets.  This section discusses existing software for
+single and multiple imputation, and missing data exploration.
 
@@
 
-\pkg{VIM} \citep{VIM} implements well-used imputation methods: K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation.
-These diverse approaches allow for imputing with semi-continuous, continuous, count, and categorical data.
-VIM identifies imputed cases by adding an indicator variable with a suffix ``\code{_imp}''.
-So, ``\code{Var1}'' has a sibling column, ``\code{Var1_imp}'', with values TRUE or FALSE indicating imputation.
-\pkg{VIM} also has a variety of visualization methods, discussed in Section \ref{exploration}.
-\pkg{simputation} provides an interface to imputation methods from \pkg{VIM}, in addition providing hotdeck imputation, and the EM algorithm \citep{simputation, Dempster1977}.
-\pkg{simputation} provides a consistent formula interface for all imputation methods, and always returns a dataframe with the updated imputations.
-\pkg{Hmisc} \citep{Hmisc}, provides predictive mean matching, \pkg{imputeTS} \citep{imputeTS}, provides time series imputation methods, and \pkg{missMDA} \citep{missMDA}, imputes data using principal component analysis.
-
-Multiple imputation is often regarded as best practice for imputing values \citep{Schafer2002}, as long as appropriate caution is taken \citep{Sterne2009}.
-Popular and robust methods for multiple imputation include the \pkg{mice}, \pkg{Amelia}, and \pkg{mi} packages \citep{mice, amelia, mi}.
-\pkg{mice} implements the method of chained equations, using a variable-wise algorithm to calculate the posterior distribution of parameters to generate imputed values.
-The workflow in \pkg{mice} revolves around imputing data, returning completed data, and fitting a model and pooling the results.
-
-\pkg{Amelia} \citep{amelia} assumes data are multivariate normal, and samples from the posterior, and allows for incorporation of information on the values in a prior.
-It uses the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm \citep{Honaker2010}.
-\pkg{norm} \citep{norm}, provides multiple imputation using EM for multivariate normal data, drawing from methods in the \proglang{NORM} software \citep{schafer-norm}.
-\pkg{norm} does not provide a framework for tracking missing values, instead providing tools for making inference from multiple imputation.
-
-\pkg{mi} \citep{mi} also uses Bayesian models for imputation, providing better handling of semi-continuous values, and data with structural or perfect correlation.
-A collection of analysis models are also provided in \pkg{mi}, to work with data it has imputed.
-These include linear models, generalized linear models, and their corresponding Bayesian components.
-This approach promotes fluid workflow, with a similar penalty to tidying up model output, which is still messy.
+\pkg{VIM} \citep{VIM} implements well-used imputation methods: $K$ nearest
+neighbors, regression, hot-deck, and iterative robust model-based
+imputation.  These diverse approaches allow for imputing with
+semi-continuous, continuous, count, and categorical data.  \pkg{VIM} identifies
+imputed cases by adding an indicator variable with a suffix \code{_imp}.
+So, \code{Var1} has a sibling column, \code{Var1_imp}, with values
+\code{TRUE} or \code{FALSE} indicating imputation.  \pkg{VIM} also has a variety of
+visualization methods, discussed in Section~\ref{exploration}.
+\pkg{simputation} provides an interface to imputation methods from
+\pkg{VIM}, in addition providing hotdeck imputation, and the
+Expectation-maximization (EM) algorithm
+\citep{simputation, Dempster1977}.  \pkg{simputation} provides a consistent
+formula interface for all imputation methods, and always returns a dataframe
+with the updated imputations.  \pkg{Hmisc} \citep{Hmisc}, provides
+predictive mean matching, \pkg{imputeTS} \citep{imputeTS}, provides time
+series imputation methods, and \pkg{missMDA} \citep{missMDA}, imputes data
+using principal component analysis.
+
+Multiple imputation is often regarded as best practice for imputing values
+\citep{Schafer2002}, as long as appropriate caution is taken
+\citep{Sterne2009}.  Popular and robust methods for multiple imputation
+include the \pkg{mice}, \pkg{Amelia}, and \pkg{mi} packages \citep{mice,
+amelia, mi}.  \pkg{mice} implements the method of chained equations, using a
+variable-wise algorithm to calculate the posterior distribution of
+parameters to generate imputed values.  The workflow in \pkg{mice} revolves
+around imputing data, returning completed data, and fitting a model and
+pooling the results.
+
+\pkg{Amelia} \citep{amelia} assumes data are multivariate normal, and
+samples from the posterior, and allows for incorporation of information on
+the values in a prior.  It uses the computationally efficient (and
+parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm
+\citep{Honaker2010}.  \pkg{norm} \citep{norm}, provides multiple imputation
+using EM for multivariate normal data, drawing from methods in the
+\proglang{NORM} software \citep{schafer-norm}.  \pkg{norm} does not provide
+a framework for tracking missing values, instead providing tools for making
+inference from multiple imputation.
+
+\pkg{mi} \citep{mi} also uses Bayesian models for imputation, providing
+better handling of semi-continuous values, and data with structural or
+perfect correlation.  A collection of analysis models are also provided in
+\pkg{mi}, to work with data it has imputed.  These include linear models,
+generalized linear models, and their corresponding Bayesian components.
+This approach promotes fluid workflow, with a similar penalty to tidying up
+model output, which is still messy.
 
@@
 
-Each imputation method provides practical methods for different use cases, but most have different output structures, and do not have consistent interfaces in their implementation.
-This makes them inherently messy and challenging to integrate into an analysis pipeline.
-For example, combining different imputation methods from different pieces of software is not currently straightforward.
-\pkg{simputation} resolves some of these complications with a simple approach of a unified syntax for all imputation, and always returns a dataframe of imputed values.
-This reduces the friction of working with other tools, but comes at the cost of identifying imputed values.
-An ideal approach would use consistent, simple data structures that work with other analysis tools, and help track missing values.
-This would make imputation outputs tidy, streamlining subsequent analysis.
+Each imputation method provides practical methods for different use cases,
+but most have different output structures, and do not have consistent
+interfaces in their implementation.  This makes them inherently messy and
+challenging to integrate into an analysis pipeline.  For example, combining
+different imputation methods from different pieces of software is not
+currently straightforward.  \pkg{simputation} resolves some of these
+complications with a simple approach of a unified syntax for all imputation,
+and always returns a dataframe of imputed values.  This reduces the friction
+of working with other tools, but comes at the cost of identifying imputed
+values.  An ideal approach would use consistent, simple data structures that
+work with other analysis tools, and help track missing values.  This would
+make imputation outputs tidy, streamlining subsequent analysis.
 
@@
 
-Missing data packages are focused primarily on making inferences, and exploring imputed values, not on exploring relationships in missing values, and identifying possible patterns.
-Texts covering the exploration phase of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools \citep{vanBuuren2018}; this is inefficient.
-Methods for exploring missing values are primarily covered in literature on interactive graphics \citep{Swayne1998, Unwin1996, Cook2007}, and are picked up again in a discussion of a graphical user interface \citep{Cheng2015}.
+Missing data packages are focused primarily on making inferences, and
+exploring imputed values, not on exploring relationships in missing values,
+and identifying possible patterns.  Texts covering the exploration phase of
+missing data have the same problem as with modelling: the input is tidy, but
+the output does not work with other tools \citep{vanBuuren2018}; this is
+inefficient.  Methods for exploring missing values are primarily covered in
+literature on interactive graphics \citep{Swayne1998, Unwin1996, Cook2007},
+and are picked up again in a discussion of a graphical user interface
+\citep{Cheng2015}.
 
 The missingness matrix \(B\) can be used to assess missing data dependence.
-It has been used in interactive graphics, dubbed a ``shadow matrix'', to link missing and imputed values to the data, facilitating their display \citep{Swayne1998}, focusing heavily on multivariate numeric data.
-This is an idea upon which this new work builds.
-
-The \proglang{MANET} (Missings Are Now Equally Treated) software \citep{Unwin1996} focussed on multivariate categorical data, with missingness explicitly added as a category.
-\proglang{MANET} also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot.
-The \proglang{MANET} software is no longer maintained and cannot be installed.
-The approach of \citep{Swayne1998} in the software \proglang{XGobi}, further developed in \proglang{ggobi} \citep{Cook2007}, focussed on multivariate quantitative data.
-Missingness is incorporated into plots in \proglang{ggobi} by setting them to be 10\% below the minimum value.
-
-\pkg{MissingDataGUI} provides a Graphical User Interface (GUI) for exploring missing data structure, both numerically and visually.
-Using a GUI to explore missing data facilitates rapid insight into missingness structures.
-However, this comes as a trade off, as insights are not captured or recorded with a GUI, making it challenging to incorporate into reproducible analyses.
+It has been used in interactive graphics, dubbed a ``shadow matrix'', to
+link missing and imputed values to the data, facilitating their display
+\citep{Swayne1998}, focusing heavily on multivariate numeric data.  This is
+an idea upon which this new work builds.
+
+The \proglang{MANET} (missings are now equally treated) software
+\citep{Unwin1996} focused on multivariate categorical data, with
+missingness explicitly added as a category.  \proglang{MANET} also provided
+univariate visualizations of missing data using linked brushing between a
+reference plot of the missingness for each variable, and a plot of the data
+as a histogram or barplot.  The \proglang{MANET} software is no longer
+maintained and cannot be installed.  The approach of \citep{Swayne1998} in
+the software \proglang{XGobi}, further developed in \proglang{ggobi}
+\citep{Cook2007}, focused on multivariate quantitative data.  Missingness
+is incorporated into plots in \proglang{ggobi} by setting them to be 10\%
+below the minimum value.
+
+\pkg{MissingDataGUI} \citep{pkg:MissingDataGUI}
+provides a Graphical User Interface (GUI) for exploring
+missing data structure, both numerically and visually.  Using a GUI to
+explore missing data facilitates rapid insight into missingness structures.
+However, this comes as a trade off, as insights are not captured or recorded
+with a GUI, making it challenging to incorporate into reproducible analyses.
 This distracts and breaks analysis workflow, inviting mistakes.
 
-\pkg{VIM} (Visualizing and Imputing Missing Data) provides visualization methods to identify and explore observed, imputed, and missing values.
-These include spinograms, spinoplots, missingness matrices, plotting missingness in the margins of other plots, and other summaries.
-However, these visualizations do not map variables to graphical aesthetics, creating friction when moving through analysis workflows, making them difficult to extend to new circumstances.
-Additionally, data used to create the visualizations cannot be accessed, posing a barrier to further exploration.
-
-The \pkg{inspectdf} package provides tools for summarising, comparing, and visualising dataframes \citep{inspectdf}.
-In particular the \texttt{inspect\_na()} function provides a summary of missing values for variables.
-\texttt{inspect\_cat()} summarises the levels of categories in each column in a dataset.
-The \texttt{show\_plot()} function shows a higher lever overview of the contents of a dataframe and its categories, as well as missing values.
-All of these functions allow for comparing missing values across two dataframes.
-For example, comparing the missing values in two dataframes with the same column names.
-
-\pkg{ggplot2} removes missing values with a warning (Figure \ref{fig:warning}), and only incorporates missingness into visualizations when mapping a discrete variable with missings to a graph aesthetic.
-This has some limitations, shown in Figure \ref{fig:gg-box-na}, a boxplot visualization of school grade and test scores.
-If there are missings in a continuous variable like test score, \pkg{ggplot2} omits the missings and prints a warning message.
-However, if a discrete variable like school year has missing values, an NA category is created for school year, where scores are placed.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/gg-box-na-1}
-
-}
-
-\caption[\pkg{ggplot2} provides different visualizations depending on what type of data has missing values for data of student test scores in school year]{\pkg{ggplot2} provides different visualizations depending on what type of data has missing values for data of student test scores in school year.
-(A) Data is complete; (B) Missings are only in year - an additional "NA" boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and year, additional missing information is shown.
-The missingness category is only shown when there are missings in categorical variables such as year (plots (B) and (D)).
-In (C), no missingness information is given on the graphic, despite there being missings in score, and a warning message is displayed about the number of missing values omitted.}\label{fig:gg-box-na}
-\end{figure}
-\end{CodeChunk}
+\pkg{VIM} (visualizing and imputing missing data) provides visualization
+methods to identify and explore observed, imputed, and missing values.
+These include spinograms, spinoplots, missingness matrices, plotting
+missingness in the margins of other plots, and other summaries.  However,
+these visualizations do not map variables to graphical aesthetics, creating
+friction when moving through analysis workflows, making them difficult to
+extend to new circumstances.  Additionally, data used to create the
+visualizations cannot be accessed, posing a barrier to further exploration.
+
+The \pkg{inspectdf} package provides tools for summarizing, comparing, and
+visualizing dataframes \citep{inspectdf}.  In particular the
+\code{inspect\_na()} function provides a summary of missing values for
+variables.  \code{inspect\_cat()} summarizes the levels of categories in
+each column in a dataset.  The \code{show\_plot()} function shows a higher
+lever overview of the contents of a dataframe and its categories, as well as
+missing values.  All of these functions allow for comparing missing values
+across two dataframes.  For example, comparing the missing values in two
+dataframes with the same column names.
+
+\pkg{ggplot2} removes missing values with a warning
+(Figure~\ref{fig:warning}), and only incorporates missingness into
+visualizations when mapping a discrete variable with missings to a graph
+aesthetic.  This has some limitations, shown in Figure~\ref{fig:gg-box-na},
+a boxplot visualization of school grade and test scores.  If there are
+missings in a continuous variable like test score, \pkg{ggplot2} omits the
+missings and prints a warning message.  However, if a discrete variable like
+school year has missing values, an NA category is created for school year,
+where scores are placed.
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=1\linewidth]{gg-box-na-1}
+%
+\caption[\pkg{ggplot2} provides different visualizations depending on what
+type of data has missing values for data of student test scores in school
+year]{\pkg{ggplot2} provides different visualizations depending on what type
+of data has missing values for data of student test scores in school year.
+(A) Data is complete; (B) Missings are only in year -- an additional ``NA''
+boxplot is created; (C) Missings only in scores, no additional missingness
+information is shown; (D) Missings in both scores and year, additional
+missing information is shown.  The missingness category is only shown when
+there are missings in categorical variables such as year (plots (B) and
+(D)).  In (C), no missingness information is given on the graphic, despite
+there being missings in score, and a warning message is displayed about the
+number of missing values omitted.}\label{fig:gg-box-na}
+%
+\end{figure}
 
@@
 
-Missing data can be explored and visualised in closed source software such as \proglang{STATA}, \proglang{SAS}, and \proglang{SPSS}.
-These provide basic table frequencies of missing data, and basic visualisations on missingness patterns.
-However, none of these software provide functions to specifically visualise missing values in bivariate or multivariate settings, and it is only possible to visualise missing values after performing data analysis, counting and summarising missingness.
-Imputation methods from mean to regression, EM, and multiple imputation are all available, with out of the box diagnostics provided to assist comparing imputations.
-\proglang{python} also provides exploratory data analysis for missing data within \pkg{missingno} \citep{Bilogur2018}, and the \pkg{scikit-learn} package provides imputation methods \citep{scikit-learn}.
+Missing data can be explored and visualized in closed source software such
+as \proglang{STATA}, \proglang{SAS}, and \proglang{SPSS}.  These provide
+basic table frequencies of missing data, and basic visualizations on
+missingness patterns.  However, none of these software provide functions to
+specifically visualize missing values in bivariate or multivariate settings,
+and it is only possible to visualize missing values after performing data
+analysis, counting and summarizing missingness.  Imputation methods from
+mean to regression, EM, and multiple imputation are all available, with out
+of the box diagnostics provided to assist comparing imputations.
+\proglang{Python} also provides exploratory data analysis for missing data
+within \pkg{missingno} \citep{Bilogur2018}, and the \pkg{scikit-learn}
+package provides imputation methods \citep{scikit-learn}.
 
@@
 
-Applying \pkg{tidyverse} principles has the potential to clarify missing data exploration, visualization, and imputation.
-This section discusses how these principles are applied for data structures (\ref{data-structure}), common operations (verbs) (\ref{verbs}), graphics (\ref{graphics}), and data summaries (\ref{num-sum}).
-Care has been taken to make the names and design these features intuitive for their purpose, and is discussed throughout.
+Applying \pkg{tidyverse} principles has the potential to clarify missing
+data exploration, visualization, and imputation.  This section discusses how
+these principles are applied for data structures (Section~\ref{data-structure}),
+common operations (verbs) (Section~\ref{verbs}), graphics (Section~\ref{graphics}), and data
+summaries (Section~\ref{num-sum}).  Care has been taken to make the names and design
+these features intuitive for their purpose, and is discussed throughout.
 
@@
 
-A data structure facilitating exploration of missing data needs to be simple to reason with and to transport with a data analysis, otherwise it will not be used.
-A useful template common in missing data literature, is the \(B\) matrix, where 0 and 1 indicate not missing and missing, respectively.
-This matrix was used to explore missing values in the interactive graphics library \proglang{XGobi}, called a ``missing value shadow'', or ``shadow matrix'', defined as a copy of the original data with indicator values of missingness.
-The shadow matrix could be interactively linked to the data.
-However, there are some limitations to the shadow matrix.
-Namely, the values 0 and 1 can be confusing representations of missing values, since it is not clear if 0 indicates an absence of observation, or the presence of a missing value.
-
-We propose a new form for tidy representation of missing data based on these ideas from past research.
-Four features are added to the shadow matrix to facilitate analysis of missing data, illustrated in Figure \ref{fig:nabularfig}.
-These are briefly described below, and discussed in further detail in section \ref{nabular-data}.
-
+A data structure facilitating exploration of missing data needs to be simple
+to reason with and to transport with a data analysis, otherwise it will not
+be used.  A useful template common in missing data literature, is the \(B\)
+matrix, where 0 and 1 indicate not missing and missing, respectively.  This
+matrix was used to explore missing values in the interactive graphics
+library \proglang{XGobi}, called a ``missing value shadow'', or ``shadow
+matrix'', defined as a copy of the original data with indicator values of
+missingness.  The shadow matrix could be interactively linked to the data.
+However, there are some limitations to the shadow matrix.  Namely, the
+values 0 and 1 can be confusing representations of missing values, since it
+is not clear if 0 indicates an absence of observation, or the presence of a
+missing value.
+
+We propose a new form for tidy representation of missing data based on these
+ideas from past research.  Four features are added to the shadow matrix to
+facilitate analysis of missing data, illustrated in
+Figure~\ref{fig:nabularfig}.  These are briefly described below, and
+discussed in further detail in Section~\ref{nabular-data}.
+%
 \begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
+% \def\labelenumi{\arabic{enumi}.}
+%
 \item
-  \textbf{Missing value labels}.
-Simple labels for missing and not missing to clearly identify missing values for analysis and plotting.
-We propose ``NA'' and ``!NA''.
-(Figure \ref{fig:nabularfig}).
-This improves the 0 and 1 values in \(B\), which do not clearly identify missingness.
-These values follow a principle: ``\textbf{clarity of labelling}'' - the matrix's meaning is transparent, and anybody looking at these values could understand what they mean, which is not the case of binary values.
-Equally, these values could instead be ``missing'' or ``present''.
+  \textbf{Missing value labels}:  Simple labels for missing and not missing
+to clearly identify missing values for analysis and plotting.  We propose
+``NA'' and ``!NA'' (Figure~\ref{fig:nabularfig}).  This improves the 0 and
+1 values in \(B\), which do not clearly identify missingness.  These values
+follow a principle: ``clarity of labelling'' -- the matrix's meaning
+is transparent, and anybody looking at these values could understand what
+they mean, which is not the case of binary values.  Equally, these values
+could instead be ``missing'' or ``present''.
+%
 \item
-  \textbf{Special missing values}: building on \textbf{missing value labels}, the values in the shadow matrix can be ``special'' missing values, indicated by ``NA\_\textless suffix\textgreater{}''.
-For example: ``NA\_instrument'' uses a short label, ``instrument'', indicating instrument error resulting in missing values.
-These could be also used to indicate imputations, (Figure \ref{fig:nabularfig}).
+  \textbf{Special missing values}: Building on \textbf{missing value
+labels}, the values in the shadow matrix can be ``special'' missing values,
+indicated by ``NA\_\textless suffix\textgreater{}''.  For example:
+``NA\_instrument'' uses a short label, ``instrument'', indicating instrument
+error resulting in missing values.  These could be also used to indicate
+imputations (Figure~\ref{fig:nabularfig}).
+%
 \item
-  \textbf{Coordinated names}: variable names in the shadow matrix gain a consistent short suffix, "\_NA", keeping names coordinated throughout analysis (Figure \ref{fig:nabularfig}).
-It makes a clear distinction with ``var\_NA'' being a random variable of the missingness of a variable, ``var''.
-This suffix is short and easy to remember during data analysis, and shifts the focus from the value of a variable, to its missingness state.
-Although this might break naming conventions for lower case ``snake\_case'' variables (e.g., \citet{Wickham2014}), this distinguishes these columns from other columns.
+  \textbf{Coordinated names}: Variable names in the shadow matrix gain a
+consistent short suffix, ``\_NA'', keeping names coordinated throughout
+analysis (Figure~\ref{fig:nabularfig}).  It makes a clear distinction with
+``var\_NA'' being a random variable of the missingness of a variable,
+``var''.  This suffix is short and easy to remember during data analysis,
+and shifts the focus from the value of a variable, to its missingness state.
+Although this might break naming conventions for lower case ``snake\_case''
+variables \citep[e.g.,][]{Wickham2014}, this distinguishes these columns from
+other columns.
+%
 \item
-  \textbf{Connectedness}: binding the shadow matrix column-wise to the original data as a factor creates a single, connected, nabular data format, in sync with the data, at the cost of a larger file.
-It is useful for visualization, summaries, and tracking imputed values, discussed in more detail in \ref{nabular-data}.
+  \textbf{Connectedness}: Binding the shadow matrix column-wise to the
+original data as a factor creates a single, connected, nabular data format,
+in sync with the data, at the cost of a larger file.  It is useful for
+visualization, summaries, and tracking imputed values, discussed in more
+detail in Section~\ref{nabular-data}.
 \end{enumerate}
@@
 
-\emph{Nabular} data binds the shadow matrix column-wise to the original data.
-It is a portmanteau of ``NA'' and ``tabular''.
-\emph{Nabular} data explicitly links missing values to data, keeping corresponding observations together and removing the possibility of mismatching records (Figure \ref{fig:nabularfig}).
-\emph{Nabular} data facilitates visualization and summaries by allowing the user to reference the missingness of a variable in a coordinated way: the missingness of ``var'', as ``var\_NA'' during analysis.
-\emph{Nabular} data is a snapshot of the missingness of the data.
-This means when \emph{nabular} data are imputed, those imputed values can easily be identified in analysis (``var'' vs ``var\_NA'').
-Imputing values on \emph{nabular} data automatically tracks these imputations.
-Together these features make it possible to do tasks that are not possible by simply adding a binary missing matrix, \(B\) during data analysis.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{/Users/njtierney/github/njtierney/papers/tidy-missing-data-paper/paper/images/diagram}
-
-}
-
-\caption[The process of creating nabular data]{The process of creating nabular data.
-Data transformed to shadow matrix, where values are either not missing or missing: "!NA" or "NA".
-The shadow matrix can be converted to long form to create missingness summary plots.
-Nabular data is created by binding the columns of the data and shadow matrix.
-Special missing values (such as -99) are identified as special missings, and values imputed and tracked.
-Nabular data can be used to identify imputations and explore data values alongside missings, providing a useful format for missing data exploration and analysis.}\label{fig:nabularfig}
-\end{figure}
-\end{CodeChunk}
-
-\emph{Nabular} data is similar to classical data formats with quality or flag variables associated with each measured variable, e.g., Scripps CO2 data \citep{Keeling2005-scripps}, GHCN data \citep{Durre2008-ghcn}.
-Using additional columns to represent missingness information follows best practices for data organization, described in \citep{Ellis2018} and \citep{Broman2018}: (1) keep one thing in a cell and (2) describe additional features of variables in a second column.
-Here they suggest to indicate censored data with an extra variable called ``VariableNameCensored'', which would be TRUE if censored, otherwise FALSE.
-This information can now be represented in the shadow columns as special missing values.
-Encoding special missing values is achieved by defining logical conditions and suffixes.
-This is implemented with the \code{recode_shadow()} function in \pkg{naniar} (Discussed in Section \ref{verbs}).
-
-Special missing values are not a new idea, and have been implemented in other statistical programming languages, \proglang{SPSS}, \proglang{SAS}, and \proglang{STATA}.
-These typically represent missing values as a full-stop, ``.'', and record special missing values as ``.a'' - ``.z''.
-These special values from these languages break the tidy principle of one column having one type of value, as they record both the value, and the multivariate missingness state.
+\emph{Nabular} data binds the shadow matrix column-wise to the original
+data.  It is a portmanteau of ``NA'' and ``tabular''.  \emph{Nabular} data
+explicitly links missing values to data, keeping corresponding observations
+together and removing the possibility of mismatching records
+(Figure~\ref{fig:nabularfig}).  \emph{Nabular} data facilitates
+visualization and summaries by allowing the user to reference the
+missingness of a variable in a coordinated way: the missingness of ``var'',
+as ``var\_NA'' during analysis.  \emph{Nabular} data is a snapshot of the
+missingness of the data.  This means when \emph{nabular} data are imputed,
+those imputed values can easily be identified in analysis (``var'' vs
+``var\_NA'').  Imputing values on \emph{nabular} data automatically tracks
+these imputations.  Together these features make it possible to do tasks
+that are not possible by simply adding a binary missing matrix, \(B\) during
+data analysis.
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=1\linewidth]{diagram}
+\caption[The process of creating nabular data]{The process of creating
+nabular data.  Data transformed to shadow matrix, where values are either
+not missing or missing: ``!NA'' or ``NA''.  The shadow matrix can be converted
+to long form to create missingness summary plots.  Nabular data is created
+by binding the columns of the data and shadow matrix.  Special missing
+values (such as ``$-99$'') are identified as special missings, and values imputed
+and tracked.  Nabular data can be used to identify imputations and explore
+data values alongside missings, providing a useful format for missing data
+exploration and analysis.}\label{fig:nabularfig}
+\end{figure}
+
+\emph{Nabular} data is similar to classical data formats with quality or
+flag variables associated with each measured variable, e.g.,~Scripps CO2
+data \citep{Keeling2005-scripps}, GHCN data \citep{Durre2008-ghcn}.  Using
+additional columns to represent missingness information follows best
+practices for data organization, described in \citep{Ellis2018} and
+\citep{Broman2018}: (1) keep one thing in a cell and (2) describe additional
+features of variables in a second column.  Here they suggest to indicate
+censored data with an extra variable called ``VariableNameCensored'', which
+would be TRUE if censored, otherwise FALSE.  This information can now be
+represented in the shadow columns as special missing values.  Encoding
+special missing values is achieved by defining logical conditions and
+suffixes.  This is implemented with the \code{recode_shadow()} function in
+\pkg{naniar} (discussed in Section~\ref{verbs}).
+
+Special missing values are not a new idea, and have been implemented in
+other statistical programming languages, \proglang{SPSS}, \proglang{SAS},
+and \proglang{STATA}.  These typically represent missing values as a
+full-stop, ``.'', and record special missing values as ``.a'' -- ``.z''.
+These special values from these languages break the tidy principle of one
+column having one type of value, as they record both the value, and the
+multivariate missingness state.
 
@@
 
-Common missing data operations can be considered verbs, in the tidyverse sense.
-For missing data, these include: \textbf{scan}, \textbf{replace}, \textbf{add}, \textbf{shadow}, \textbf{impute}, \textbf{track}, and \textbf{flag}.
-Data can be \textbf{scanned} to find possible missings not coded as \code{NA}.
-These values can then be \textbf{replaced} with \code{NA}.
-To facilitate exploration, summaries of missingness can be \textbf{added} as a summary column to the original data.
-The data can be augmented with the \textbf{shadow matrix} values, helping explore missing data, as well as facilitating the process of \textbf{imputing}, and \textbf{tracking}.
-Finally, unusual or specially coded missing values can be \textbf{flagged}.
+Common missing data operations can be considered verbs, in the tidyverse
+sense.  For missing data, these include: \textbf{scan}, \textbf{replace},
+\textbf{add}, \textbf{shadow}, \textbf{impute}, \textbf{track}, and
+\textbf{flag}.  Data can be \textbf{scanned} to find possible missings not
+coded as NA.  These values can then be \textbf{replaced} with
+NA.  To facilitate exploration, summaries of missingness can be
+\textbf{added} as a summary column to the original data.  The data can be
+augmented with the \textbf{shadow matrix} values, helping explore missing
+data, as well as facilitating the process of \textbf{imputing}, and
+\textbf{tracking}.  Finally, unusual or specially coded missing values can
+be \textbf{flagged}.
 
 \hypertarget{verbs-search}{%
-\subsubsection{scan: searching for common missing value labels}\label{verbs-search}}
-
-This operation is used to search the data for specific conventional representations of missings, such as, ``N/A'', ``MISSING'', ``-99''.
-This is implemented in the function \code{miss_scan_count()}, which returns a table of occurrences of that value for each variable.
-A list of common NA values for numbers and characters, can be provided to help check for typical representations of missings.
-These are implemented in \pkg{naniar} as objects \code{common_na_numbers} and \code{common_na_strings}.
+\subsubsection{``scan'': Searching for common missing value labels}\label{verbs-search}}
+
+This operation is used to search the data for specific conventional
+representations of missings, such as, ``N/A'', ``MISSING'', ``$-99$''.  This
+is implemented in the function \code{miss_scan_count()}, which returns a
+table of occurrences of that value for each variable.  A list of common NA
+values for numbers and characters, can be provided to help check for typical
+representations of missings.  These are implemented in \pkg{naniar} as
+objects `\code{common_na_numbers}' and `\code{common_na_strings}'.
 
 \hypertarget{verbs-replace-with}{%
-\subsubsection{replace: replacing values with missing values}\label{verbs-replace-with}}
-
-Once possible missing values have been identified, these values can be replaced.
-For example, a dataset could have the values ``-99'' meaning a missing value.
-\pkg{naniar} implements replacement with the function, \code{replace_with_na()}.
-Values ``-99'' could be replaced in the ``x'' column with: \code{replace_with_na(dat_ms, replace = list(x = -99))}.
-For operating on multiple variables, there are scoped variants for \code{replace_with_na()}: \code{_all}, \code{_if}, and \code{_at}.
-This means \code{replace_with_na_all()} operates on \textbf{all} columns, \code{replace_with_na_at()} operates \textbf{at} specific columns, and \code{replace_with_na_if()} makes a conditional change on columns \textbf{if} they meet some condition (such as \code{is.numeric()} or \code{is.character()}).
+\subsubsection{``replace'': Replacing values with missing values}\label{verbs-replace-with}}
+
+Once possible missing values have been identified, these values can be
+replaced.  For example, a dataset could have the values ``$-99$'' meaning a
+missing value.  \pkg{naniar} implements replacement with the function,
+\code{replace_with_na()}.  Values ``$-99$'' could be replaced in the ``x''
+column with: \code{replace_with_na(dat_ms, replace = list(x = -99))}.  For
+operating on multiple variables, there are scoped variants for
+\code{replace_with_na()}: \code{_all}, \code{_if}, and \code{_at}.  This
+means \code{replace_with_na_all()} operates on \textbf{all} columns,
+\code{replace_with_na_at()} operates \textbf{at} specific columns, and
+\code{replace_with_na_if()} makes a conditional change on columns
+\textbf{if} they meet some condition (such as \code{is.numeric()} or
+\code{is.character()}).
 
 \hypertarget{verbs-add-cols}{%
-\subsubsection{add: adding missingness summary variables}\label{verbs-add-cols}}
-
-Understanding the missingness structure can be improved by adding summary information alongside the data.
-For example, in \citep{Tierney2015}, the proportion of missings in a row is used as the outcome in a model to identify variables important in predicting missingness structures.
-\pkg{naniar} implements a series of functions to add these missingness summaries to the data, starting with \code{add_}.
-These are inspired by the \code{add_count()} function in \pkg{dplyr}, which adds count information for specified groups or conditions.
-\pkg{naniar} provides operations to add the number or proportion of missingness, the missingness cluster, or the presence of any missings, with: \code{add_n_miss()}, \code{add_prop_miss()}, \code{add_miss_cluster()}, and \code{add_any_miss()}, respectively (Table \ref{tab:add-missing-info}).
-There are also functions for adding information about shadow values, and readable labels for any missing values with \code{add_label_shadow()} and \code{add_label_missings()}.
-
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:add-missing-info}Overview of the "add" functions in naniar}
-\centering
-\begin{tabular}[t]{l|l}
-\hline
-Function & Adds column which:\\
-\hline
-add\_n\_miss(data) & contains the number missing values in a row\\
-\hline
-add\_any\_miss(data) & contains whether there are any missing values in a row\\
-\hline
-add\_prop\_miss(data) & contains the proportion of missing values in a row\\
-\hline
-add\_miss\_cluster(data) & contains the missing value cluster\\
+\subsubsection{``add'': Adding missingness summary variables}\label{verbs-add-cols}}
+
+Understanding the missingness structure can be improved by adding summary
+information alongside the data.  For example, in \citep{Tierney2015}, the
+proportion of missings in a row is used as the outcome in a model to
+identify variables important in predicting missingness structures.
+\pkg{naniar} implements a series of functions to add these missingness
+summaries to the data, starting with \code{add_}.  These are inspired by the
+\code{add_count()} function in \pkg{dplyr}, which adds count information for
+specified groups or conditions.  \pkg{naniar} provides operations to add the
+number or proportion of missingness, the missingness cluster, or the
+presence of any missings, with: \code{add_n_miss()}, \code{add_prop_miss()},
+\code{add_miss_cluster()}, and \code{add_any_miss()}, respectively
+(Table~\ref{tab:add-missing-info}).  There are also functions for adding
+information about shadow values, and readable labels for any missing values
+with \code{add_label_shadow()} and \code{add_label_missings()}.
+
+\begin{table}[t!]
+\centering
+\begin{tabular}[t]{ll}
+\hline
+Function & Adds column which\\
+\hline
+\code{add\_n\_miss(data)} & contains the number missing values in a row\\
+\code{add\_any\_miss(data)} & contains whether there are any missing values in a row\\
+\code{add\_prop\_miss(data)} & contains the proportion of missing values in a row\\
+\code{add\_miss\_cluster(data)} & contains the missing value cluster\\
 \hline
 \end{tabular}
+\caption{\label{tab:add-missing-info}Overview of the \code{add} functions in
+\pkg{naniar}}.
 \end{table}
 
-\end{CodeChunk}
-
 \hypertarget{verbs-nabular}{%
-\subsubsection{shadow: creating nabular data}\label{verbs-nabular}}
+\subsubsection{``shadow'': Creating nabular data}\label{verbs-nabular}}
 
 \emph{Nabular} data has the shadow matrix column-bound to existing data.
-This facilitates visualization and summaries, and allows for imputed values to be tracked.
-\emph{Nabular} data can be created with \code{nabular()}:
+This facilitates visualization and summaries, and allows for imputed values
+to be tracked.  \emph{Nabular} data can be created with \code{nabular()}:
 
@@
 \hypertarget{verbs-recode}{%
-\subsubsection{flag: describing different types of missing values}\label{verbs-recode}}
-
-Unusual or spurious data values are often identified and ``flagged''.
-For example, there might be special codes to mark an individual dropping out of a study, known instrument failure in weather instruments, or for values censored in analysis.
-\pkg{naniar} provides tools to encode these special types of missingness in the shadow matrix of \emph{nabular} data, using \code{recode_shadow()}.
-This requires specifying the variable to contain the flagged value, the condition for flagging, and a suffix.
-This is then recoded as a new factor level in the shadow matrix, so every column is aware of all possible new values of missingness.
-For example, -99 could be recoded to indicate a broken machine sensor for the variable ``x'' with the following:
-
+\subsubsection{``flag'': Describing different types of missing values}\label{verbs-recode}}
+
+Unusual or spurious data values are often identified and ``flagged''.  For
+example, there might be special codes to mark an individual dropping out of
+a study, known instrument failure in weather instruments, or for values
+censored in analysis.  \pkg{naniar} provides tools to encode these special
+types of missingness in the shadow matrix of \emph{nabular} data, using
+\code{recode_shadow()}.  This requires specifying the variable to contain
+the flagged value, the condition for flagging, and a suffix.  This is then
+recoded as a new factor level in the shadow matrix, so every column is aware
+of all possible new values of missingness.  For example, ``$-99$'' could be
+recoded to indicate a broken machine sensor for the variable ``x'' with the
+following:
+%
 \begin{CodeChunk}
@@
 \end{CodeChunk}
-
+%
 \hypertarget{verbs-impute}{%
-\subsubsection{impute: imputing values}\label{verbs-impute}}
-
-\pkg{naniar} does not reinvent the wheel by creating the need for customised imputation procedures.
-Instead, the \code{nabular} format means existing methods using imputation replacement in place now automatically track missing values.
-\pkg{naniar} provides a few imputation methods to facilitate exploration and visualization: \code{impute_below()}, \code{impute_mean()}, and \code{impute_median()}.
-While useful to explore structure in missingness, they are not recommended for use in analysis.
-\code{impute_below()} imputes values below the minimum value, with some controllable jitter (random noise) to reduce overplotting.
-
-Similar to \pkg{simputation}, each \code{impute_} function returns the data with values imputed.
-However, \pkg{naniar} does not use a formula syntax, instead each function has ``scoped variants'' \code{_all}, \code{_at} and \code{_if}.
-\code{impute_} functions with no scoped variant, (\code{impute_mean()}), will work on a single vector, but not a data.frame.
+\subsubsection{``impute'': Imputing values}\label{verbs-impute}}
+
+\pkg{naniar} does not reinvent the wheel by creating the need for customized
+imputation procedures.  Instead, the \code{nabular} format means existing
+methods using imputation replacement in place now automatically track
+missing values.  \pkg{naniar} provides a few imputation methods to
+facilitate exploration and visualization: \code{impute_below()},
+\code{impute_mean()}, and \code{impute_median()}.  While useful to explore
+structure in missingness, they are not recommended for use in analysis.
+\code{impute_below()} imputes values below the minimum value, with some
+controllable jitter (random noise) to reduce overplotting.
+
+Similar to \pkg{simputation}, each \code{impute_} function returns the data
+with values imputed.  However, \pkg{naniar} does not use a formula syntax,
+instead each function has ``scoped variants'' \code{_all}, \code{_at} and
+\code{_if}.  \code{impute_} functions with no scoped variant,
+(\code{impute_mean()}), will work on a single vector, but not a data.frame.
 One challenge with this approach is imputed value locations are not tracked.
-This issue is resolved with \emph{nabular} data covered in the following section.
+This issue is resolved with \emph{nabular} data covered in the following
+section.
 
 \hypertarget{verbs-track}{%
-\subsubsection{track: shadow and impute missing values}\label{verbs-track}}
-
-To evaluate imputations they need to be tracked.
-This is achieved by first using \emph{nabular} data, then imputing, and imputed values can then be referred to by their shadow variable, "\_NA" (Figure \ref{fig:track-impute-example}).
-The code below shows the track pattern, first using \code{nabular()}, then imputing with \code{impute_lm()}.
-\code{label_shadow()} then adds a label to facilitate identifying missings:
-
+\subsubsection{``track'': Shadow and impute missing values}\label{verbs-track}}
+
+To evaluate imputations they need to be tracked.  This is achieved by first
+using \emph{nabular} data, then imputing, and imputed values can then be
+referred to by their shadow variable, ``\_NA''
+(Figure~\ref{fig:track-impute-example}).  The code below shows the track
+pattern, first using \code{nabular()}, then imputing with
+\code{impute_lm()}.  \code{label_shadow()} then adds a label to facilitate
+identifying missings:
+%
 \begin{CodeChunk}
@@
 +   add_label_shadow()
-R> 
 R> head(aq_imputed)
@@
 \end{CodeChunk}
-
-Multiple missing or imputed values can be mapped to a graphical element in ggplot2 by setting the \code{color} or \code{fill} aesthetic in ggplot to \code{any_missing}, a result of the \code{add_label_shadow()} function.
-(Figure \ref{fig:track-impute-example}).
-Imputed values can also be compared to complete case data, grouping by \code{any_missing}, and then summarizing, similar to other \pkg{dplyr} summary workflows shown below in Table \ref{tab:impute-summary-out}, showing similarities and differences in imputation methods.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/track-impute-example-1}
-
-}
-
-\caption[Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values from a linear model]{Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values from a linear model.
-Imputed values are colored green, and data values orange.
-Imputed values are similar, but slightly trended to the mean.}\label{fig:track-impute-example}
-\end{figure}
-\end{CodeChunk}
+%
+Multiple missing or imputed values can be mapped to a graphical element in
+\pkg{ggplot2} by setting the \code{color} or \code{fill} aesthetic in ggplot to
+\code{any_missing}, a result of the \code{add_label_shadow()} function.
+(Figure~\ref{fig:track-impute-example}).  Imputed values can also be
+compared to complete case data, grouping by \code{any_missing}, and then
+summarizing, similar to other \pkg{dplyr} summary workflows shown below in
+Table~\ref{tab:impute-summary-out}, showing similarities and differences in
+imputation methods.
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=1\linewidth]{track-impute-example-1}
+\caption[Scatterplot (A) and density plots (B and C) of ozone and solar
+radiation from the airquality dataset containing imputed values from a
+linear model]{Scatterplot (A) and density plots (B and C) of ozone and solar
+radiation from the airquality dataset containing imputed values from a
+linear model.  Imputed values are colored green, and data values orange.
+Imputed values are similar, but slightly trended to the
+mean.}\label{fig:track-impute-example}
+\end{figure}
 
@@
 +   group_by(any_missing) %>%
-+   summarise_at(.vars = vars(Ozone),
-+                .funs = lst(min, mean, median, max)) 
++   summarise_at(.vars = vars(Ozone), .funs = lst(min, mean, median, max))
 \end{CodeInput}
@@
 
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:impute-summary-out}Output of \pkg{dplyr} summary statistics of imputed vs non imputed values for the variable "Ozone".
-The "any\_missing" column denotes imputed values ("Missing", since they were previously missing), and non-imputed values ("Not Missing").
-The mean and median values are similar, but the minimum and maximum values are very different.}
-\centering
-\begin{tabular}[t]{l|r|r|r|r}
-\hline
-any\_missing & min & mean & median & max\\
-\hline
-Missing & -16.86418 & 41.22494 & 45.4734 & 78\\
-\hline
+\begin{table}[t!]
+\centering
+\begin{tabular}[t]{lrrrr}
+\hline
+\code{any\_missing} & \code{min} & \code{mean} & \code{median} & \code{max}\\
+\hline
+Missing & $-16.86418$ & 41.22494 & 45.4734 & 78\\
 Not Missing & 1.00000 & 42.09910 & 31.0000 & 168\\
@@
 \end{tabular}
+\caption{\label{tab:impute-summary-out}Output of \pkg{dplyr} summary
+statistics of imputed vs non imputed values for the variable ``Ozone''.  The
+``any\_missing'' column denotes imputed values (``Missing'', since they were
+previously missing), and non-imputed values (``Not Missing'').  The mean and
+median values are similar, but the minimum and maximum values are very
+different.}
 \end{table}
-
-\end{CodeChunk}
 
@@
 
-Missing values are often ignored when plotting data - which is why the data visualization software, \proglang{MANET}, was named and is an acronym corresponding to ``Missings Are Now Equally Treated'' \citep{Unwin1996}.
-However, plots can help to identify the type of missing value patterns, even those of MCAR, MAR or MNAR.
-Here we summarise how to systematically explore missing patterns visually, and define useful plots to make, relative to the \emph{nabular} data structure.
+Missing values are often ignored when plotting data - which is why the data
+visualization software, \proglang{MANET}, was named and is an acronym
+corresponding to ``missings are now equally treated'' \citep{Unwin1996}.
+However, plots can help to identify the type of missing value patterns, even
+those of MCAR, MAR or MNAR.  Here we summarize how to systematically explore
+missing patterns visually, and define useful plots to make, relative to the
+\emph{nabular} data structure.
 
@@
 The first step is to get an overview of the extent of the missingness.
-Overview visualizations for variables and cases are provided with \code{gg_miss_var()} and \code{gg_miss_case()} (Figure \ref{fig:gg-miss-case-var}A), drawing attention to the amount of missings, and ordering by missingness.
-The ``airquality'' dataset (from base \proglang{R}), is shown, and contains daily air quality measurements in New York, from May to September, 1973.
-We learn from Figure \ref{fig:gg-miss-case-var}A-B, that two variables contain missings, approximately one third of observations have one missing value, and a tiny number of observations are missing across two variables.
-These overview plots are created from the shadow matrix in long form (Figure \ref{fig:nabularfig}).
-Numerical statistics can also be reported (Section \ref{num-sum}).
-
-\begin{CodeChunk}
+Overview visualizations for variables and cases are provided with
+\code{gg_miss_var()} and \code{gg_miss_case()}
+(Figure~\ref{fig:gg-miss-case-var}A), drawing attention to the amount of
+missings, and ordering by missingness.  The ``airquality'' dataset (from
+base \proglang{R}), is shown, and contains daily air quality measurements in
+New York, from May to September, 1973.  We learn from
+Figure~\ref{fig:gg-miss-case-var}A--B, that two variables contain missings,
+approximately one third of observations have one missing value, and a tiny
+number of observations are missing across two variables.  These overview
+plots are created from the shadow matrix in long form
+(Figure~\ref{fig:nabularfig}).  Numerical statistics can also be reported
+(Section~\ref{num-sum}).
+
 \begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{tidy-missing-data-paper_files/figure-latex/gg-miss-case-var-1}
-
-}
-
-\caption[Graphical summaries of missingness in the airquality data]{Graphical summaries of missingness in the airquality data.
-Missings in variables (A) and cases (B), and for an overview of all missingness as a heatmap in (C), and with clustering applied (D).
-There are missing values in Ozone and Solar.R, with Ozone having more missings.
-Not many cases have two missings.
-Most missingness is from cases with one missing value.
-The default output (C) and ordered by clustering on rows and columns (D).
-These overviews are made possible using the shadow matrix in long form.
-There are only missings in ozone and solar radiation, and there appears to be some structure to their missingness.}\label{fig:gg-miss-case-var}
-\end{figure}
-\end{CodeChunk}
-
-The shadow matrix can be put into long form, allowing both the variables and cases to be displayed using a heatmap style visualization, with \code{vis_miss()} from \pkg{visdat} \citep{visdat} (Figure \ref{fig:nabularfig}).
-This also provides numerical summaries of missingness in the legend, and for each column (Figure \ref{fig:gg-miss-case-var}C-D).
-Clustering can be applied to the rows, and columns arranged in order of missingness (Figure \ref{fig:gg-miss-case-var}D).
-Similar visualizations are available in other packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}.
-A key improvement is \code{vis_miss()} orients the visualization analogous to a regular data structure: variables form columns and are named at the top, and each row is an observation.
-Using \pkg{ggplot2}, as the foundation, makes the plot easily customizable.
-
-\begin{CodeChunk}
+\centering
+\includegraphics[width=0.9\linewidth]{gg-miss-case-var-1}
+\caption[Graphical summaries of missingness in the airquality
+data]{Graphical summaries of missingness in the airquality data.  Missings
+in variables (A) and cases (B), and for an overview of all missingness as a
+heatmap in (C), and with clustering applied (D).  There are missing values
+in ``Ozone'' and ``Solar.R'', with Ozone having more missings.  Not many cases have
+two missings.  Most missingness is from cases with one missing value.  The
+default output (C) and ordered by clustering on rows and columns (D).  These
+overviews are made possible using the shadow matrix in long form.  There are
+only missings in ozone and solar radiation, and there appears to be some
+structure to their missingness.}\label{fig:gg-miss-case-var}
+\end{figure}
+
+The shadow matrix can be put into long form, allowing both the variables and
+cases to be displayed using a heatmap style visualization, with
+\code{vis_miss()} from \pkg{visdat} \citep{visdat}
+(Figure~\ref{fig:nabularfig}).  This also provides numerical summaries of
+missingness in the legend, and for each column
+(Figure~\ref{fig:gg-miss-case-var}C--D).  Clustering can be applied to the
+rows, and columns arranged in order of missingness
+(Figure~\ref{fig:gg-miss-case-var}D).  Similar visualizations are available
+in other packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and
+\pkg{MissingDataGUI}.  A key improvement is \code{vis_miss()} orients the
+visualization analogous to a regular data structure: variables form columns
+and are named at the top, and each row is an observation.  Using
+\pkg{ggplot2}, as the foundation, makes the plot easily customizable.
+
 \begin{figure}
-
-{\centering \includegraphics{tidy-missing-data-paper_files/figure-latex/airquality-upset-1}
-
-}
-
-\caption[The pattern of missingness in the airquality dataset shown in an upset plot]{The pattern of missingness in the airquality dataset shown in an upset plot.
-Only Ozone and Solar.R have missing values, and Ozone has the most missing values.
-There are 2 cases where both Solar.R and Ozone have missing values.}\label{fig:airquality-upset}
-\end{figure}
-\end{CodeChunk}
-
-The number of times observations are missing together can be visualized using an ``upset plot'' \citep{Conway2017}.
-An alternative to a Venn diagram, an upset plot shows the size and features of overlapping sets, and scales well with more variables.
-An upset plot can be constructed from the shadow matrix, as shown in Figure \ref{fig:airquality-upset} which displays the overlapping counts of missings in the airquality data.
-The bottom right shows the combinations of missingness, the top panel shows the size of these combinations, and the bottom left shows missingness in each variable.
-This provides similar information to Figure \ref{fig:gg-miss-case-var}A-D, but more clearly illustrating overlapping missingness, where 2 cases are missing together in variables ``Solar.R'' and ``Ozone''.
+\centering
+\includegraphics{airquality-upset-1}
+\caption[The pattern of missingness in the airquality dataset shown in an
+upset plot]{The pattern of missingness in the airquality dataset shown in an
+upset plot.  Only ``Ozone'' and ``Solar.R'' have missing values, and Ozone has the
+most missing values.  There are 2~cases where both ``Solar.R'' and Ozone have
+missing values.}\label{fig:airquality-upset}
+\end{figure}
+
+The number of times observations are missing together can be visualized
+using an ``upset plot'' \citep{Conway2017}.  An alternative to a Venn
+diagram, an upset plot shows the size and features of overlapping sets, and
+scales well with more variables.  An upset plot can be constructed from the
+shadow matrix, as shown in Figure~\ref{fig:airquality-upset} which displays
+the overlapping counts of missings in the airquality data.  The bottom right
+shows the combinations of missingness, the top panel shows the size of these
+combinations, and the bottom left shows missingness in each variable.  This
+provides similar information to Figure~\ref{fig:gg-miss-case-var}A--D, but
+more clearly illustrating overlapping missingness, where 2~cases are missing
+together in variables ``Solar.R'' and ``Ozone''.
 
@@
 
-Missing values are typically not shown for univariate visualizations such as histograms or densities.
-Two ways to use \emph{nabular} data to present univariate data with missings are discussed.
-The first imputes values below the range to facilitate visualizations.
-The second displays two plots of the same variable according to the missingness of a chosen variable.
-
-\textbf{Imputing values below the range}.
-To visualize the amount of missings in each variable, the data is transformed into \emph{nabular} form, then values are imputed below the range of data using \code{impute_below_all()}.
-Figure \ref{fig:impute-shift-histogram}A shows a histogram of Ozone values on the right in green, and the histogram of missing ozone values on the left, in orange.
-The missings in Ozone are imputed and ``Ozone\_NA'' is mapped to the fill aesthetic.
-\emph{Nabular} data facilitates adding counts of missingness to a histogram, allowing examination of a variable's distribution of values, and also the magnitude of missings.
-
-\textbf{Univariate split by missingness}.
-The distribution of a variable can be shown according to the missingness of another variable.
-The shadow matrix part of \emph{nabular} is used to handle the faceting, and color mapping.
-Figure \ref{fig:impute-shift-histogram} shows the values of temperature when ozone is present, and missing, using a faceted histogram (B), and an overlaid density (C).
-This shows how values of temperature are affected by the missingness of ozone, and reveals a cluster of low temperature observations with missing ozone values.
-This type of plot can facilitate exploring missing data distributions.
-For example, we would expect if data were MCAR, for values to be roughly uniformly missing throughout the histogram or density.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=0.75\linewidth]{tidy-missing-data-paper_files/figure-latex/impute-shift-histogram-1} \includegraphics[width=0.75\linewidth]{tidy-missing-data-paper_files/figure-latex/impute-shift-histogram-2} \includegraphics[width=0.75\linewidth]{tidy-missing-data-paper_files/figure-latex/impute-shift-histogram-3}
-
-}
-
-\caption[Univariate summaries of missingness]{Univariate summaries of missingness.
-(A) A histogram using nabular data to show the values and missings in ozone.
-Values are imputed below the range to show the number of missings in ozone and colored according to missingness of ozone (`Ozone\_NA`).
-There are about 35 missings in Ozone.
-Panel C shows temperature according to missingness in ozone from in the airquality dataset.
-A histogram of temperature facetted by the missingness of ozone (B), or a density of temperature colored by missingness in ozone (C).
-These show a cluster of low temperature observations with missing ozone values, but temperature is otherwise similar.}\label{fig:impute-shift-histogram}
-\end{figure}
-\end{CodeChunk}
+Missing values are typically not shown for univariate visualizations such as
+histograms or densities.  Two ways to use \emph{nabular} data to present
+univariate data with missings are discussed.  The first imputes values below
+the range to facilitate visualizations.  The second displays two plots of
+the same variable according to the missingness of a chosen variable.
+
+\subsubsection{Imputing values below the range}
+
+To visualize the amount of
+missings in each variable, the data is transformed into \emph{nabular} form,
+then values are imputed below the range of data using
+\code{impute_below_all()}.  Figure~\ref{fig:impute-shift-histogram}A shows a
+histogram of Ozone values on the right in green, and the histogram of
+missing ozone values on the left, in orange.  The missings in Ozone are
+imputed and ``Ozone\_NA'' is mapped to the fill aesthetic.  \emph{Nabular}
+data facilitates adding counts of missingness to a histogram, allowing
+examination of a variable's distribution of values, and also the magnitude
+of missings.
+
+\subsubsection{Univariate split by missingness}
+
+The distribution of a variable
+can be shown according to the missingness of another variable.  The shadow
+matrix part of \emph{nabular} is used to handle the faceting, and color
+mapping.  Figure~\ref{fig:impute-shift-histogram} shows the values of
+temperature when ozone is present, and missing, using a faceted histogram
+(B), and an overlaid density (C).  This shows how values of temperature are
+affected by the missingness of ozone, and reveals a cluster of low
+temperature observations with missing ozone values.  This type of plot can
+facilitate exploring missing data distributions.  For example, we would
+expect if data were MCAR, for values to be roughly uniformly missing
+throughout the histogram or density.
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=0.75\linewidth]{impute-shift-histogram-1}
+\includegraphics[width=0.75\linewidth]{impute-shift-histogram-2}
+\includegraphics[width=0.75\linewidth]{impute-shift-histogram-3}
+\caption[Univariate summaries of missingness]{Univariate summaries of
+missingness.  (A) A histogram using nabular data to show the values and
+missings in ozone.  Values are imputed below the range to show the number of
+missings in ozone and colored according to missingness of ozone
+(``Ozone\_NA'').  There are about 35 missings in Ozone.  Panel~C shows
+temperature according to missingness in ozone from in the airquality
+dataset.  A histogram of temperature faceted by the missingness of ozone
+(B), or a density of temperature colored by missingness in ozone (C).  These
+show a cluster of low temperature observations with missing ozone values,
+but temperature is otherwise similar.}\label{fig:impute-shift-histogram}
+\end{figure}
 
@@
 
-To visualize missing values in two dimensions the missing values can be placed in plot margins, by imputing values below the range of the data.
-Using \emph{nabular} data identifies imputed values, and color makes missingness pre-attentive \citep{treisman1985}.
-The steps of imputing and coloring are combined into \code{geom_miss_point()}.
-Figure \ref{fig:geom-miss} shows a mostly uniform spread of missing values for Solar.R and Ozone.
-As \code{geom_miss_point()} is a defined \pkg{ggplot2} geometry, it works with features such as faceting and mapping other variables to graphical aesthetics.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics{tidy-missing-data-paper_files/figure-latex/geom-miss-1}
-
-}
-
-\caption[Scatterplots with missings displayed at 10 percent below for the airquality dataset]{Scatterplots with missings displayed at 10 percent below for the airquality dataset.
-Scatterplots of ozone and solar radiation (A), and ozone and temperature (B).
-There are missings in ozone and solar radiation, but not temperature.}\label{fig:geom-miss}
-\end{figure}
-\end{CodeChunk}
+To visualize missing values in two dimensions the missing values can be
+placed in plot margins, by imputing values below the range of the data.
+Using \emph{nabular} data identifies imputed values, and color makes
+missingness pre-attentive \citep{treisman1985}.  The steps of imputing and
+coloring are combined into \code{geom_miss_point()}.
+Figure~\ref{fig:geom-miss} shows a mostly uniform spread of missing values
+for ``Solar.R'' and ``Ozone''.  As \code{geom_miss_point()} is a defined
+\pkg{ggplot2} geometry, it works with features such as faceting and mapping
+other variables to graphical aesthetics.
+
+\begin{figure}[t!]
+\centering
+\includegraphics{geom-miss-1}
+\caption[Scatterplots with missings displayed at 10\% below for the
+airquality dataset]{Scatterplots with missings displayed at 10\% below
+for the airquality dataset.  Scatterplots of ozone and solar radiation (A),
+and ozone and temperature (B).  There are missings in ozone and solar
+radiation, but not temperature.}\label{fig:geom-miss}
+\end{figure}
 
@@
 
-Parallel coordinate plots can help to visualize missingness beyond two dimensions.
-They transform variables to the same scale, ranging between 0 and 1.
-The \code{oceanbuoys} dataset from \pkg{naniar} is used for this visualization, containing measurements of sea and air temperature, humidity, and east west and north south wind speeds.
-Data was collected in 1993 and 1997, to understand and predict El Niño and El Niña.
-Figure \ref{fig:parallel-cord-plot} is a parallel coordinate plot of \code{oceanbuoys}, with missing values imputed to be 10\% below the range, and values colored according to whether humidity was missing (``humidity\_NA'').
-Figure \ref{fig:parallel-cord-plot} shows humidity is missing at low air and sea temperatures, and humidity is missing in one year, and one location.
-
-\begin{CodeChunk}
+Parallel coordinate plots can help to visualize missingness beyond two
+dimensions.  They transform variables to the same scale, ranging between 0
+and 1.  The \code{oceanbuoys} dataset from \pkg{naniar} is used for this
+visualization, containing measurements of sea and air temperature, humidity,
+and east west and north south wind speeds.  Data was collected in 1993 and
+1997, to understand and predict El Ni{\~n}o and El Ni{\~n}a.
+Figure~\ref{fig:parallel-cord-plot} is a parallel coordinate plot of
+\code{oceanbuoys}, with missing values imputed to be 10\% below the range,
+and values colored according to whether humidity was missing
+(``humidity\_NA'').  Figure~\ref{fig:parallel-cord-plot} shows humidity is
+missing at low air and sea temperatures, and humidity is missing in one
+year, and one location.
+
 \begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/parallel-cord-plot-1}
-
-}
-
-\caption[Parallel coordinate plot shows missing values imputed 10\% below range for the oceanbuoys dataset]{Parallel coordinate plot shows missing values imputed 10\% below range for the oceanbuoys dataset.
-Values are colored by missingness of humidity.
-Humidity is missing for low air and sea temperatures, and is missing for one year and one location.}\label{fig:parallel-cord-plot}
-\end{figure}
-\end{CodeChunk}
+\centering
+\includegraphics[width=1\linewidth]{parallel-cord-plot-1}
+\caption[Parallel coordinate plot shows missing values imputed 10\% below
+range for the oceanbuoys dataset]{Parallel coordinate plot shows missing
+values imputed 10\% below range for the oceanbuoys dataset.  Values are
+colored by missingness of humidity.  Humidity is missing for low air and sea
+temperatures, and is missing for one year and one
+location.}\label{fig:parallel-cord-plot}
+\end{figure}
 
@@
 
-This section describes approaches to summarizing missingness, and an implementation in \pkg{naniar}.
-Numerical summaries should be easy to remember with consistent names and output, returning either a single number \ref{single-num-sum}, or a dataframe \ref{sum-tab-missings}, so they integrate well with plotting and modelling tools.
-How these work with other tools in an analysis pipeline is shown in \ref{num-sum-w-group}.
+This section describes approaches to summarizing missingness, and an
+implementation in \pkg{naniar}.  Numerical summaries should be easy to
+remember with consistent names and output, returning either a single
+number (Section~\ref{single-num-sum}), or a dataframe (Section~\ref{sum-tab-missings}),
+so they integrate well with plotting and modelling tools.  How these work with other
+tools in an analysis pipeline is shown in (Section~\ref{num-sum-w-group}).
 
@@
 
-The overall number, proportion, or percent of missing values in a dataset should be simple to calculate.
-\pkg{naniar} provides the functions \code{n_miss()}, \code{prop_miss()} and \code{pct_miss()}, as well as their complements.
-Summaries for variables and cases are made by appending \code{_case} or \code{_var} to these summaries.
-An overview is shown in Table \ref{tab:n-prop-pct-miss-complete}.
-
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:n-prop-pct-miss-complete}Single number summaries of missingness and completeness of the airquality dataset.
-The functions follow consistent naming, making them easy to remember, and their use clear.}
-\centering
-\begin{tabular}[t]{l|r|l|r}
+The overall number, proportion, or percent of missing values in a dataset
+should be simple to calculate.  \pkg{naniar} provides the functions
+\code{n_miss()}, \code{prop_miss()} and \code{pct_miss()}, as well as their
+complements.  Summaries for variables and cases are made by appending
+\code{_case} or \code{_var} to these summaries.  An overview is shown in
+Table~\ref{tab:n-prop-pct-miss-complete}.
+
+\begin{table}[t!]
+\centering
+\begin{tabular}[t]{lrlr}
 \hline
@@
 \hline
-n\_miss & 44.00 & n\_complete & 874.00\\
-\hline
-prop\_miss & 0.05 & prop\_complete & 0.95\\
-\hline
-pct\_miss & 4.79 & pct\_complete & 95.21\\
-\hline
-pct\_miss\_case & 27.45 & prop\_complete\_case & 72.55\\
-\hline
-pct\_miss\_var & 33.33 & pct\_complete\_var & 66.67\\
+\code{n\_miss} & 44.00 & \code{n\_complete} & 874.00\\
+\code{prop\_miss} & 0.05 & \code{prop\_complete} & 0.95\\
+\code{pct\_miss} & 4.79 & \code{pct\_complete} & 95.21\\
+\code{pct\_miss\_case} & 27.45 & \code{prop\_complete\_case} & 72.55\\
+\code{pct\_miss\_var} & 33.33 & \code{pct\_complete\_var} & 66.67\\
 \hline
 \end{tabular}
+\caption{\label{tab:n-prop-pct-miss-complete}Single number summaries of
+missingness and completeness of the airquality dataset.  The functions
+follow consistent naming, making them easy to remember, and their use
+clear.}
 \end{table}
-
-\end{CodeChunk}
 
@@
 
-Presenting the number and percent of missing values for each variable, or case, provides a summary usable in data handling, or in models to inform imputation.
-For example, potentially dropping variables, or deciding to include others in an imputation model.
-Another useful approach is to tabulate the frequency of missing values for each variable or case; that is, the number of times there are zero missings, one missing, two, and so on.
-These summaries and tabulations are shown for variables in Tables \ref{tab:miss-var-summary} and \ref{tab:miss-var-table}, and implemented with \code{miss_var_summary()} and \code{miss_var_table()}.
-Case-wise (row-wise) summaries and tabulations are implemented with \code{miss_case_summary()} and \code{miss_case_table()}.
-
-These summaries order rows by the number of missings (\code{n_miss()}), to show the most missings at the top.
-The number of missings across a repeating span, or finding ``runs'' or ``streaks'' of missings in given variables can also be useful for identifying missingness patterns, and are implemented with \code{miss_var_span()}, and \code{miss_var_run()}.
-
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:miss-var-summary}\texttt{miss\char`_var\char`_summary(airquality)} provides the number and percent of missings in each variable in airquality.
+Presenting the number and percent of missing values for each variable, or
+case, provides a summary usable in data handling, or in models to inform
+imputation.  For example, potentially dropping variables, or deciding to
+include others in an imputation model.  Another useful approach is to
+tabulate the frequency of missing values for each variable or case; that is,
+the number of times there are zero missings, one missing, two, and so on.
+These summaries and tabulations are shown for variables in
+Tables~\ref{tab:miss-var-summary} and~\ref{tab:miss-var-table}, and
+implemented with \code{miss_var_summary()} and \code{miss_var_table()}.
+Case-wise (row-wise) summaries and tabulations are implemented with
+\code{miss_case_summary()} and \code{miss_case_table()}.
+
+These summaries order rows by the number of missings (\code{n_miss()}), to
+show the most missings at the top.  The number of missings across a
+repeating span, or finding ``runs'' or ``streaks'' of missings in given
+variables can also be useful for identifying missingness patterns, and are
+implemented with \code{miss_var_span()}, and \code{miss_var_run()}.
+
+\begin{table}[t!]
+\centering
+\begin{tabular}[t]{lrr}
+\hline
+\code{Variable} & \code{n\_miss} & \code{pct\_miss}\\
+\hline
+Ozone & 37 & 24.2\\
+Solar.R & 7 & 4.6\\
+Wind & 0 & 0.0\\
+Temp & 0 & 0.0\\
+Month & 0 & 0.0\\
+Day & 0 & 0.0\\
+\hline
+\end{tabular}
+\caption{\label{tab:miss-var-summary}\code{miss\char`_var\char`_summary(airquality)}
+provides the number and percent of missings in each variable in airquality.
 Only ozone and solar radiation have missing values.}
-\centering
-\begin{tabular}[t]{l|r|r}
-\hline
-variable & n\_miss & pct\_miss\\
-\hline
-Ozone & 37 & 24.2\\
-\hline
-Solar.R & 7 & 4.6\\
-\hline
-Wind & 0 & 0.0\\
-\hline
-Temp & 0 & 0.0\\
-\hline
-Month & 0 & 0.0\\
-\hline
-Day & 0 & 0.0\\
+\end{table}
+
+\begin{table}[t!]
+\centering
+\begin{tabular}[t]{rrr}
+\hline
+\code{n\_miss\_in\_var} & \code{n\_vars} & \code{pct\_vars}\\
+\hline
+0 & 4 & 66.7\\
+7 & 1 & 16.7\\
+37 & 1 & 16.7\\
 \hline
 \end{tabular}
+\caption{\label{tab:miss-var-table}The output of
+\code{miss\char`_var\char`_table(airquality)}, tabulating the amount of
+missing data in each variable in airquality.  This shows the number of
+variables with 0, 7, and 37 missings, and the percentage of variables with
+those amounts of missingness.  There are few missingness patterns.}
 \end{table}
-
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:miss-var-table}The output of \code{miss\char`_var\char`_table(airquality)}, tabulating the amount of missing data in each variable in airquality.
-This shows the number of variables with 0, 7, and 37 missings, and the percentage of variables with those amounts of missingness.
-There are few missingness patterns.}
-\centering
-\begin{tabular}[t]{r|r|r}
-\hline
-n\_miss\_in\_var & n\_vars & pct\_vars\\
-\hline
-0 & 4 & 66.7\\
-\hline
-7 & 1 & 16.7\\
-\hline
-37 & 1 & 16.7\\
-\hline
-\end{tabular}
-\end{table}
-
-\end{CodeChunk}
 
@@
 
-It is useful to explore summaries and tabulations within groups of a dataset.
-\pkg{naniar} works with \pkg{dplyr}'s \code{group_by()} operator to produce grouped summaries, which work well with the ``pipe'' operator.
-Table \ref{tab:group-miss-var-summary} below shows an example of missing data summaries for airquality, grouped by month.
-
-\begin{CodeChunk}
-\begin{table}
-
-\caption{\label{tab:group-miss-var-summary}Output of \code{airquality \%>\% group\char`_by(Month) \%>\% miss\char`_var\char`_summary()} provides a grouped summary of the missingness in each variable, for each month of the airquality dataset.
-Only the first 10 rows are shown.
-There are more ozone missings in June than May.}
-\centering
-\begin{tabular}[t]{r|l|r|r}
-\hline
-Month & variable & n\_miss & pct\_miss\\
+It is useful to explore summaries and tabulations within groups of a
+dataset.  \pkg{naniar} works with \pkg{dplyr}'s \code{group_by()} operator
+to produce grouped summaries, which work well with the ``pipe'' operator.
+Table~\ref{tab:group-miss-var-summary} below shows an example of missing
+data summaries for airquality, grouped by month.
+
+\begin{table}[t!]
+\centering
+\begin{tabular}[t]{rlrr}
+\hline
+\code{Month} & \code{Variable} & \code{n\_miss} & \code{pct\_miss}\\
 \hline
 5 & Ozone & 5 & 16.1\\
-\hline
 5 & Solar.R & 4 & 12.9\\
-\hline
 5 & Wind & 0 & 0.0\\
-\hline
 5 & Temp & 0 & 0.0\\
-\hline
 5 & Day & 0 & 0.0\\
-\hline
 6 & Ozone & 21 & 70.0\\
-\hline
 6 & Solar.R & 0 & 0.0\\
-\hline
 6 & Wind & 0 & 0.0\\
-\hline
 6 & Temp & 0 & 0.0\\
-\hline
 6 & Day & 0 & 0.0\\
@@
 \end{tabular}
+\caption{\label{tab:group-miss-var-summary}Output of \code{airquality \%>\%
+group\char`_by(Month) \%>\% miss\char`_var\char`_summary()} provides a
+grouped summary of the missingness in each variable, for each month of the
+airquality dataset.  Only the first 10 rows are shown.  There are more ozone
+missings in June than May.}
 \end{table}
-
-\end{CodeChunk}
 
@@
 
-This section shows how the methods described so far are used together in a data analysis workflow.
-We analyse a case study of housing data for the city of Melbourne from January 28, 2016 to March 17, 2018.
-The data was compiled by scraping weekly property clearance data \citep{Kaggle-2018-data}.
-There are 27,247 properties, and 21 variables in the dataset.
-The variables include real estate type (town house, unit, house), suburb, selling method, number of rooms, price, real estate agent, sale date, and distance from the Central Business District (CBD).
-
-The goal in analyzing this data is to accurately predict Melbourne housing prices.
-The data contains many missing values.
-As a precursor to building a predictive model, this analysis focuses on understanding the patterns of missingness.
+This section shows how the methods described so far are used together in a
+data analysis workflow.  We analyse a case study of housing data for the
+city of Melbourne from January 28, 2016 to March 17, 2018.  The data was
+compiled by scraping weekly property clearance data
+\citep{Kaggle-2018-data}.  There are $27,247$ properties, and 21~variables in
+the dataset.  The variables include real estate type (town house, unit,
+house), suburb, selling method, number of rooms, price, real estate agent,
+sale date, and distance from the central business district.
+
+The goal in analyzing this data is to accurately predict Melbourne housing
+prices.  The data contains many missing values.  As a precursor to building
+a predictive model, this analysis focuses on understanding the patterns of
+missingness.
 
@@
 
-Figure \ref{fig:housing-miss-case-var}A shows 9 variables with missing values.
-The most missings are in ``building area'', followed by ``year built'', and ``land size'', with similar amounts of missingness in ``Car'', ``bathroom'', ``bedroom2'', ``longitude'', and ``latitude''.
-Figure \ref{fig:housing-miss-case-var}B reveals there are up to 50\% missing values in cases, and the majority of cases have more than 5\% values missing.
-The variables ``building area'' and ``year built'' have more than 50\% missing data, and so could perhaps be omitted from subsequent analysis, as imputed values are likely to be spurious.
-Three missingness clusters are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \ref{fig:applic-vis-miss}).
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/housing-miss-case-var-1}
-
-}
-
-\caption[The amount of missings in variables (A) and cases (B) for Melbourne housing data]{The amount of missings in variables (A) and cases (B) for Melbourne housing data.
-(A) Build area and year built have more than 50\% missing, and car, bathroom, bedroom2 and longitude and latitude have about 25\% missings.
-(B) Cases are missing 5 - 50\% of values.
-The majority of missingness is in selected cases and variables.}\label{fig:housing-miss-case-var}
-\end{figure}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=0.85\linewidth]{tidy-missing-data-paper_files/figure-latex/applic-vis-miss-1}
-
-}
-
-\caption[Heatmap of clustered missingness for housing data reveals structured missingness]{Heatmap of clustered missingness for housing data reveals structured missingness.
-Three groups of missingness are apparent.
-At the top: building area to longitude; the middle: building area and year built; the end: building area, year built, and landsize.}\label{fig:applic-vis-miss}
-\end{figure}
-\end{CodeChunk}
-
-Figure \ref{fig:housing-upset} shows missingness patterns with an ``upset'' plot \citep{Conway2017}, displaying 8 intersecting sets of missing variables.
-Two patterns stand out: two, and five variables missing, providing further evidence of the missingness patterns seen in Figures \ref{fig:housing-miss-case-var} and \ref{fig:housing-upset}.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/housing-upset-1}
-
-}
-
-\caption[An upset plot of 8 sets of missingness in the housing data]{An upset plot of 8 sets of missingness in the housing data.
-Missingness for each variable is shown on the bottom left.
-Connected dots show co-occurences of missings in variables.
-Two missingness patterns are clear, year built and building area, and lattitude through to building area.}\label{fig:housing-upset}
-\end{figure}
-\end{CodeChunk}
-
-Tabulating the number of missings in variables in Table \ref{tab:housing-miss-var-case-table} (left) shows three groups of missingness.
-Tabulating missings in cases (Table \ref{tab:housing-miss-var-case-table} (right)) shows six patterns of missingness.
-These overview plots lead to the removal of two variables from with more than 50\% missingness from analysis: ``building area'' and ``year built''.
-
-\begin{CodeChunk}
+Figure~\ref{fig:housing-miss-case-var}A shows 9 variables with missing
+values.  The most missings are in ``building area'', followed by ``year
+built'', and ``land size'', with similar amounts of missingness in ``Car'',
+``bathroom'', ``bedroom2'', ``longitude'', and ``latitude''.
+Figure~\ref{fig:housing-miss-case-var}B reveals there are up to 50\% missing
+values in cases, and the majority of cases have more than 5\% values
+missing.  The variables ``building area'' and ``year built'' have more than
+50\% missing data, and so could perhaps be omitted from subsequent analysis,
+as imputed values are likely to be spurious.  Three missingness clusters are
+revealed by visualizing missingness in the whole dataset, clustering and
+arranging the rows and columns of the data
+(Figure~\ref{fig:applic-vis-miss}).
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=1\linewidth]{housing-miss-case-var-1}
+\caption[The amount of missings in variables (A) and cases (B) for Melbourne
+housing data]{The amount of missings in variables (A) and cases (B) for
+Melbourne housing data.  (A) Build area and year built have more than 50\%
+missing, and car, bathroom, bedroom2 and longitude and latitude have about
+25\% missings.  (B) Cases are missing $5 - 50\%$ of values.  The majority of
+missingness is in selected cases and
+variables.}\label{fig:housing-miss-case-var}
+\end{figure}
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=0.85\linewidth]{applic-vis-miss-1}
+\caption[Heatmap of clustered missingness for housing data reveals
+structured missingness]{Heatmap of clustered missingness for housing data
+reveals structured missingness.  Three groups of missingness are apparent.
+At the top: building area to longitude; the middle: building area and year
+built; the end: building area, year built, and
+landsize.}\label{fig:applic-vis-miss}
+\end{figure}
+
+Figure~\ref{fig:housing-upset} shows missingness patterns with an upset
+plot \citep{Conway2017}, displaying 8 intersecting sets of missing
+variables.  Two patterns stand out: two, and five variables missing,
+providing further evidence of the missingness patterns seen in
+Figures~\ref{fig:housing-miss-case-var} and~\ref{fig:housing-upset}.
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=1\linewidth]{housing-upset-1}
+\caption[An upset plot of 8 sets of missingness in the housing data]{An
+upset plot of 8 sets of missingness in the housing data.  Missingness for
+each variable is shown on the bottom left.  Connected dots show
+co-occurrences of missings in variables.  Two missingness patterns are clear,
+year built and building area, and latitude through to building
+area.}\label{fig:housing-upset}
+\end{figure}
+
+Tabulating the number of missings in variables in
+Table~\ref{tab:housing-miss-var-case-table} (left) shows three groups of
+missingness.  Tabulating missings in cases
+(Table~\ref{tab:housing-miss-var-case-table}, right) shows six patterns of
+missingness.  These overview plots lead to the removal of two variables from
+with more than 50\% missingness from analysis: ``building area'' and ``year
+built''.
+
 \begin{table}
-\caption{\label{tab:housing-miss-var-case-table}Summary tables to help understand missingness patters.
-Output of \code{miss\_var\_table(housing)} (left), tabulating missingness for variables, and output of \code{miss\_case\_table(housing)}.
-There are 13 variables with 0-3 missings, 6 variables have 6000-10000 missings, 2 variables have 15000 – 17000 missings.
-About 30\% of cases have no missings, 45\% of cases have 1 - 6 missings, and about 23\% of cases have 8 or more missings.
-There are different patterns of missingness in variables and cases, but they can be broken down into smaller groups.}
-
-\centering
-\begin{tabular}[t]{r|r|r}
-\hline
-n\_miss\_in\_var & n\_vars & pct\_vars\\
+\centering
+\begin{tabular}[t]{rrr}
+\hline
+\code{n\_miss\_in\_var} & \code{n\_vars} & \code{pct\_vars}\\
 \hline
 0 & 10 & 47.6\\
-\hline
 1 & 2 & 9.5\\
-\hline
 3 & 1 & 4.8\\
-\hline
 6254 & 2 & 9.5\\
-\hline
 6441 & 1 & 4.8\\
-\hline
 6447 & 1 & 4.8\\
-\hline
 6824 & 1 & 4.8\\
-\hline
 9265 & 1 & 4.8\\
-\hline
 15163 & 1 & 4.8\\
-\hline
 16591 & 1 & 4.8\\
@@
 \centering
-\begin{tabular}[t]{r|r|r}
-\hline
-n\_miss\_in\_case & n\_cases & pct\_cases\\
+\begin{tabular}[t]{rrr}
+\hline
+\code{n\_miss\_in\_case} & \code{n\_cases} & \code{pct\_cases}\\
 \hline
 0 & 8887 & 32.6\\
-\hline
 1 & 3237 & 11.9\\
-\hline
 2 & 7231 & 26.5\\
-\hline
 3 & 1370 & 5.0\\
-\hline
 4 & 79 & 0.3\\
-\hline
 5 & 8 & 0.0\\
-\hline
 6 & 203 & 0.7\\
-\hline
 8 & 6229 & 22.9\\
-\hline
 9 & 2 & 0.0\\
-\hline
 11 & 1 & 0.0\\
@@
 \end{tabular}
+\caption{\label{tab:housing-miss-var-case-table}Summary tables to help
+understand missingness patters.  Output of \code{miss\_var\_table(housing)}
+(left), tabulating missingness for variables, and output of
+\code{miss\_case\_table(housing)}.  There are 13 variables with $0-3$
+missings, 6 variables have $6,000-10,000$ missings, 2 variables have $15,000 -
+17,000$ missings.  About 30\% of cases have no missings, 45\% of cases have $1
+- 6$ missings, and about 23\% of cases have 8 or more missings.  There are
+different patterns of missingness in variables and cases, but they can be
+broken down into smaller groups.}
 \end{table}
-
-\end{CodeChunk}
 
@@
 
-Using information from \ref{case-study-explore-pattern}, the following variables are explored for features predicting missingness: ``latitude'', ``longitude'', ``bedroom2'', ``bathroom'', ``car'', and ``land size''.
-
-Missingness structure is explored by clustering the missing values into groups.
-Then, a classification and regression trees (CART) model is applied to predict these missingness clusters using the remaining variables \citep{Tierney2015, Barnett2017}.
-Two clusters of missingness are identified and predicted using all variables in the dataset with the CART package \pkg{rpart} \citep{rpart}, and plotted using the \pkg{rpart.plot} package \citep{rpart-plot}.
-Importance scores reveal the following variables as most important in predicting missingness: ``rooms'', ``price'', ``suburb'', ``council area'', ``distance'', and ``region name''.
-These variables are important for predicting missingness, so are included in the imputation model.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{tidy-missing-data-paper_files/figure-latex/rpart-plot-1}
-
-}
-
-\caption[Decision tree output predicting missingness clusters]{Decision tree output predicting missingness clusters.
-Type of house, year quarter, and year were important for predicting missingness cluster.
-The cluster with the most missingness was for quarters 1 and 4, for 2017 and 2018.
-Type of house, year, and year quarter are important features related to missingness structure.}\label{fig:rpart-plot}
-\end{figure}
-\end{CodeChunk}
+Using information from Section~\ref{case-study-explore-pattern}, the following
+variables are explored for features predicting missingness: ``latitude'',
+``longitude'', ``bedroom2'', ``bathroom'', ``car'', and ``land size''.
+
+Missingness structure is explored by clustering the missing values into
+groups.  Then, a classification and regression trees (CART) model is applied
+to predict these missingness clusters using the remaining variables
+\citep{Tierney2015, Barnett2017}.  Two clusters of missingness are
+identified and predicted using all variables in the dataset with the CART
+package \pkg{rpart} \citep{rpart}, and plotted using the \pkg{rpart.plot}
+package \citep{rpart-plot}.  Importance scores reveal the following
+variables as most important in predicting missingness: ``rooms'', ``price'',
+``suburb'', ``council area'', ``distance'', and ``region name''.  These
+variables are important for predicting missingness, so are included in the
+imputation model.
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=0.9\linewidth]{rpart-plot-1}
+\caption[Decision tree output predicting missingness clusters]{Decision tree
+output predicting missingness clusters.  Type of house, year quarter, and
+year were important for predicting missingness cluster.  The cluster with
+the most missingness was for quarters 1 and 4, for 2017 and 2018.  Type of
+house, year, and year quarter are important features related to missingness
+structure.}\label{fig:rpart-plot}
+\end{figure}
 
@@
 
-\pkg{simputation} is used to implement two imputations: simple linear regression and K nearest neighbors.
-Values are imputed stepwise in ascending order of missingness.
-The track missings pattern is applied (described in \ref{verbs}), to assess imputed values.
-Imputed datasets are compared on their performance in a model predicting log house price for 4 variables (Figure \ref{fig:imputed-by-model}).
-Compared to KNN imputed values, the linear model imputed values closer to the mean.
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=0.95\linewidth]{tidy-missing-data-paper_files/figure-latex/imputed-by-model-1}
-
-}
-
-\caption[Boxplots of complete case data, and data imputed with KNN or linear model for different variables]{Boxplots of complete case data, and data imputed with KNN or linear model for different variables.
-(A) number of bedrooms, (B) number of bathrooms, (C) number of carspots, and (D) landsize (on a log10 scale).
-KNN had similar results to complete case, and linear model had a lower median for cars and fewer extreme values for bedrooms.}\label{fig:imputed-by-model}
-\end{figure}
-\end{CodeChunk}
+\pkg{simputation} is used to implement two imputations: simple linear
+regression and $K$ nearest neighbors.  Values are imputed stepwise in
+ascending order of missingness.  The track missings pattern is applied
+(described in Section~\ref{verbs}), to assess imputed values.  Imputed datasets are
+compared on their performance in a model predicting log house price for 4~variables
+(Figure~\ref{fig:imputed-by-model}).  Compared to $k$-nearest neighbors (KNN) imputed
+values, the linear model imputed values closer to the mean.
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=0.95\linewidth]{imputed-by-model-1}
+\caption[Boxplots of complete case data, and data imputed with KNN or linear
+model for different variables]{Boxplots of complete case data, and data
+imputed with KNN or linear model for different variables.  (A) number of
+bedrooms, (B) number of bathrooms, (C) number of carspots, and (D) landsize
+(on a log10 scale).  KNN had similar results to complete case, and linear
+model had a lower median for cars and fewer extreme values for
+bedrooms.}\label{fig:imputed-by-model}
+\end{figure}
 
@@
 
-Coefficients of the linear model of log price vary for room for different imputed datasets (Figure \ref{fig:tidy-coefs}).
-Notably, complete cases underestimate the impact of room on log price.
-A partial residual plot (Figure \ref{fig:partial-resid}) shows there is not much variation amongst the models from the differently imputed datasets.
-
-\begin{CodeChunk}
+Coefficients of the linear model of log price vary for room for different
+imputed datasets (Figure~\ref{fig:tidy-coefs}).  Notably, complete cases
+underestimate the impact of room on log price.  A partial residual plot
+(Figure~\ref{fig:partial-resid}) shows there is not much variation amongst
+the models from the differently imputed datasets.
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=1\linewidth]{tidy-coefs-1}
+\caption[The coefficient estimate for the number of rooms varies according
+to the imputed dataset]{The coefficient estimate for the number of rooms
+varies according to the imputed dataset.  Complete case dataset produced
+lower coefficients, compared to imputed datasets.}\label{fig:tidy-coefs}
+\end{figure}
+
 \begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/tidy-coefs-1}
-
-}
-
-\caption[The coefficient estimate for the number of rooms varies according to the imputed dataset]{The coefficient estimate for the number of rooms varies according to the imputed dataset.
-Complete case dataset produced lower coefficients, compared to imputed datasets.}\label{fig:tidy-coefs}
-\end{figure}
-\end{CodeChunk}
-
-\begin{CodeChunk}
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{tidy-missing-data-paper_files/figure-latex/partial-resid-1}
-
-}
-
-\caption[Partial residual plot for each data set, complete cases (cc), and imputed with KNN (knn) or linear model (lm)]{Partial residual plot for each data set, complete cases (cc), and imputed with KNN (knn) or linear model (lm).
-These are plotted as hex bins, colored according to the number of points in a given hexagon.
-Brighter colors mean more points.
-Compared to complete cases, imputed data has more points clustered around zero.}\label{fig:partial-resid}
-\end{figure}
-\end{CodeChunk}
+\centering
+\includegraphics[width=1\linewidth]{partial-resid-1}
+\caption[Partial residual plot for each data set, complete cases (cc), and
+imputed with KNN (knn) or linear model (lm)]{Partial residual plot for each
+data set, complete cases (cc), and imputed with KNN (knn) or linear model
+(lm).  These are plotted as hex bins, colored according to the number of
+points in a given hexagon.  Brighter colors mean more points.  Compared to
+complete cases, imputed data has more points clustered around
+zero.}\label{fig:partial-resid}
+\end{figure}
 
@@
 
-The \pkg{naniar} and \pkg{visdat} packages implement the methods discussed in the paper, building on existing tidy tools and strike a compromise between automation and control, making analysis efficient, readable, but not overly complex.
-Each tool has clear intent and effects - summarising, plotting or generating data or augmenting data in some way.
-This not only reduces repetition and typing in an analysis, but most importantly, allows for clear expression of intent, making exploration of missing values fluent.
+The \pkg{naniar} and \pkg{visdat} packages implement the methods discussed
+in the paper, building on existing tidy tools and strike a compromise
+between automation and control, making analysis efficient, readable, but not
+overly complex.  Each tool has clear intent and effects -- summarizing,
+plotting or generating data or augmenting data in some way.  This not only
+reduces repetition and typing in an analysis, but most importantly, allows
+for clear expression of intent, making exploration of missing values fluent.
 
@@
 
-This paper has described new methods for exploring, visualizing, and imputing missing data.
-The work was motivated by recent developments of tidy data, and extends them for better missing value handling.
-The methods have standard outputs, function arguments, and behavior.
-This provides consistent workflows centered around data analysis that integrate well with existing imputation methodology, visualization, and modelling.
-
-The \emph{nabular} data structures discussed in the paper are simple by design, to promote flexibility.
-They could be used to create different visualizations than were shown in the paper.
-The analyst can use the data structures to decide on appropriate visualization for their problem.
-The data structures could also be used to support interactive graphics, in the manner of \proglang{MANET} and \proglang{ggobi}.
-Linking the plots (via linked brushing) could facilitate exploration of missingness, and could be implemented with \pkg{plotly} \citep{plotly} for added interactivity.
-Animating between different sets of imputed values could also be possible with packages like \pkg{gganimate} \citep{gganimate}.
-
-Other data structures such as spatial data, time series, networks, and longitudinal data would be supported by the inherently tabular, \emph{nabular} data, if they are first structured as wide tidy format.
-Large data may need special handling, and additional features like efficient storage of purely imputed values and lazy evaluation.
-Special missing value codes could be improved by creating special classes, or expanding low level representation of \code{NA} at the source code level.
-Similarly, list columns, present a challenge for analysis as their expansion into vectors in a dataframe is not always possible as they have different lengths, but should still be flagged for further analysis.
-
-The methodology described in this paper can be used in conjunction with other approaches to understand multivariate missingness dependencies (e.g., decision trees \citep{Tierney2015}, latent group analysis \citep{Barnett2017}, and PCA \citep{FactoMineR}).
-Evaluating imputed values using a testing framework like \citep{vanBuuren2018} is also supported.
-
-The approach meshes with the dynamic nature of data analysis, allowing the analyst to go from raw data to model data in a fluid workflow.
+This paper has described new methods for exploring, visualizing, and
+imputing missing data.  The work was motivated by recent developments of
+tidy data, and extends them for better missing value handling.  The methods
+have standard outputs, function arguments, and behavior.  This provides
+consistent workflows centered around data analysis that integrate well with
+existing imputation methodology, visualization, and modelling.
+
+The \emph{nabular} data structures discussed in the paper are simple by
+design, to promote flexibility.  They could be used to create different
+visualizations than were shown in the paper.  The analyst can use the data
+structures to decide on appropriate visualization for their problem.  The
+data structures could also be used to support interactive graphics, in the
+manner of \proglang{MANET} and \proglang{ggobi}.  Linking the plots (via
+linked brushing) could facilitate exploration of missingness, and could be
+implemented with \pkg{plotly} \citep{plotly} for added interactivity.
+Animating between different sets of imputed values could also be possible
+with packages like \pkg{gganimate} \citep{gganimate}.
+
+Other data structures such as spatial data, time series, networks, and
+longitudinal data would be supported by the inherently tabular,
+\emph{nabular} data, if they are first structured as wide tidy format.
+Large data may need special handling, and additional features like efficient
+storage of purely imputed values and lazy evaluation.  Special missing value
+codes could be improved by creating special classes, or expanding low level
+representation of NA at the source code level.  Similarly, list
+columns, present a challenge for analysis as their expansion into vectors in
+a dataframe is not always possible as they have different lengths, but
+should still be flagged for further analysis.
+
+The methodology described in this paper can be used in conjunction with
+other approaches to understand multivariate missingness dependencies (e.g.,~decision
+trees \citep{Tierney2015}, latent group analysis
+\citep{Barnett2017}, and PCA \citep{FactoMineR}).  Evaluating imputed values
+using a testing framework like \citep{vanBuuren2018} is also supported.
+
+The approach meshes with the dynamic nature of data analysis, allowing the
+analyst to go from raw data to model data in a fluid workflow.
 
 \hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
-
-The authors would like to thank Miles McBain, for his key contributions and discussions on the \pkg{naniar} package, in particular for helping implement \code{geom_miss_point()}, and for his feedback on ideas, implementations, and names.
-We also thank Colin Fay for his contributions to the \pkg{naniar} package, in particular for his assistance with the \code{replace_with_na()} functions.
-We also thank Earo Wang and Mitchell O'Hara-Wild for the many useful discussions on missing data and package development, and for their assistance with creating elegant approaches that take advantage of the tidy syntax.
-We would also like to thank those who contributed pull requests and discussions on the \pkg{naniar} package, in particular Jim Hester and Romain François for improving the speed of key functions, Ross Gayler for discussion on special missing values, and Luke Smith for helping \pkg{naniar} be more compliant with \pkg{ggplot2}.
-We would also like to thank Amelia McNamara for discussions on the paper.
-
-\renewcommand\refname{References}
-\bibliography{tidy-missing-data-zotero.bib}
+\section*{Acknowledgments}\label{acknowledgements}}
+
+The authors would like to thank Miles McBain, for his key contributions and
+discussions on the \pkg{naniar} package, in particular for helping implement
+\code{geom_miss_point()}, and for his feedback on ideas, implementations,
+and names.  We also thank Colin Fay for his contributions to the
+\pkg{naniar} package, in particular for his assistance with the
+\code{replace_with_na()} functions.  We also thank Earo Wang and Mitchell
+O'Hara-Wild for the many useful discussions on missing data and package
+development, and for their assistance with creating elegant approaches that
+take advantage of the tidy syntax.  We would also like to thank those who
+contributed pull requests and discussions on the \pkg{naniar} package, in
+particular Jim Hester and Romain Fran{\c{c}}ois for improving the speed of key
+functions, Ross Gayler for discussion on special missing values, and Luke
+Smith for helping \pkg{naniar} be more compliant with \pkg{ggplot2}.  We
+would also like to thank Amelia McNamara for discussions on the paper.
+
+\bibliography{ref}
 

### 2022-12-01 jss-copyedit/2022-01-12-4108/Code/code.R

@@ -1,0 +1,758 @@
+## ----libs---------------------------------------------------------------------
+library("tidyverse") 
+library("gridExtra")
+library("ggthemes")
+library("png")
+library("scales")
+library("simputation")
+library("naniar") 
+library("lubridate")
+library("magick")
+library("rpart")
+library("visdat")
+library("patchwork")
+library("withr")
+library("targets")
+library("tidyverse") 
+library("janitor")
+library("knitr")
+library("cowplot")
+
+
+## ----data-----------------------------------------------------------------
+housing_raw <-
+  read_csv("data/melbourne_housing_raw.csv") %>%
+  clean_names() %>%
+  rename(region_name = regionname,
+         property_count = propertycount) %>%
+  mutate(date = dmy(date)) %>%
+  rename(latitude = lattitude,
+         longitude = longtitude) %>%
+  # let's create monthly quarters
+  mutate(yr_qtr = as.ordered(quarter(date, 
+                                     with_year = TRUE,
+                                     fiscal_start = 1))) %>%
+  # drop price
+  drop_na(price) %>%
+  # make price the log of price
+  mutate(price = log10(price))
+
+
+# compact down the seller levels
+count_dat <- count(housing_raw,seller_g) %>% 
+  arrange(-n) %>%
+  mutate(cumu = cumsum(n),
+         pct = cumu / nrow(housing_raw),
+         pct_change = pct - lag(pct)) %>%
+  rowid_to_column() %>%
+  mutate(seller = case_when(
+    pct < 0.52 ~ as.character(seller_g),
+    pct > 0.52 & pct < 0.60 ~ "seller_g1",
+    pct > 0.60 & pct < 0.65 ~ "seller_g2",
+    pct > 0.65 & pct < 0.70 ~ "seller_g3",
+    pct > 0.70 & pct < 0.75 ~ "seller_g4",
+    pct > 0.75 & pct < 0.80 ~ "seller_g5",
+    pct > 0.80 & pct < 0.85 ~ "seller_g6",
+    pct > 0.85 & pct < 0.90 ~ "seller_g6",
+    pct > 0.90 & pct < 0.95 ~ "seller_g6",
+    pct > 0.95 ~ "seller_g5"
+  ))
+
+# drop these three variables, as we won't use them in analysis, they
+# would need extensive recoding
+
+housing <- housing_raw %>%
+  mutate_if(is.character, as.factor) %>%
+  mutate_at(vars(rooms,
+                 bedroom2,
[…]

### 2023-02-02 code.tex
final jss changes

\documentclass{article}\usepackage[]{graphicx}\usepackage[]{xcolor}
% maxwidth is the original width if it is less than linewidth
% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
\let\hlipl\hlkwb

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage[sc]{mathpazo}
\renewcommand{\sfdefault}{lmss}
\renewcommand{\ttdefault}{lmtt}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2.5cm,bmargin=2.5cm,lmargin=2.5cm,rmargin=2.5cm}
\setcounter{secnumdepth}{2}
\setcounter{tocdepth}{2}
\usepackage[unicode=true,pdfusetitle,
 bookmarks=true,bookmarksnumbered=true,bookmarksopen=true,bookmarksopenlevel=2,
 breaklinks=false,pdfborder={0 0 1},backref=false,colorlinks=false]
 {hyperref}
\hypersetup{
 pdfstartview={XYZ null null 1}}

\makeatletter
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\renewcommand{\textfraction}{0.05}
\renewcommand{\topfraction}{0.8}
\renewcommand{\bottomfraction}{0.8}
\renewcommand{\floatpagefraction}{0.75}

\makeatother
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

The results below are generated from an R script.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{library}\hlstd{(}\hlstr{"tidyverse"}\hlstd{)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# -- Attaching packages ------------------------------------------------- tidyverse 1.3.2 --\\\#\# v ggplot2 3.4.0 \ \ \ \ \ v purrr \ \ 0.3.5 \\\#\# v tibble \ 3.1.8 \ \ \ \ \ v dplyr \ \ 1.0.10\\\#\# v tidyr \ \ 1.2.1 \ \ \ \ \ v stringr 1.4.1 \\\#\# v readr \ \ 2.1.3 \ \ \ \ \ v forcats 0.5.2 \\\#\# -- Conflicts ---------------------------------------------------- tidyverse\_conflicts() --\\\#\# x readr::edition\_get() \ \ masks testthat::edition\_get()\\\#\# x dplyr::filter() \ \ \ \ \ \ \ masks stats::filter()\\\#\# x purrr::is\_null() \ \ \ \ \ \ masks testthat::is\_null()\\\#\# x dplyr::lag() \ \ \ \ \ \ \ \ \ \ masks stats::lag()\\\#\# x readr::local\_edition() masks testthat::local\_edition()\\\#\# x dplyr::matches() \ \ \ \ \ \ masks tidyr::matches(), testthat::matches()}}\begin{alltt}
\hlkwd{library}\hlstd{(}\hlstr{"gridExtra"}\hlstd{)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# \\\#\# Attaching package: 'gridExtra'\\\#\# \\\#\# The following object is masked from 'package:dplyr':\\\#\# \\\#\# \ \ \ \ combine}}\begin{alltt}
\hlkwd{library}\hlstd{(}\hlstr{"ggthemes"}\hlstd{)}
\hlkwd{library}\hlstd{(}\hlstr{"png"}\hlstd{)}
\hlkwd{library}\hlstd{(}\hlstr{"scales"}\hlstd{)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# \\\#\# Attaching package: 'scales'\\\#\# \\\#\# The following object is masked from 'package:purrr':\\\#\# \\\#\# \ \ \ \ discard\\\#\# \\\#\# The following object is masked from 'package:readr':\\\#\# \\\#\# \ \ \ \ col\_factor}}\begin{alltt}
\hlkwd{library}\hlstd{(}\hlstr{"simputation"}\hlstd{)}
\hlkwd{library}\hlstd{(}\hlstr{"naniar"}\hlstd{)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# \\\#\# Attaching package: 'naniar'\\\#\# \\\#\# The following object is masked from 'package:simputation':\\\#\# \\\#\# \ \ \ \ impute\_median}}\begin{alltt}
\hlkwd{library}\hlstd{(}\hlstr{"lubridate"}\hlstd{)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Loading required package: timechange\\\#\# \\\#\# Attaching package: 'lubridate'\\\#\# \\\#\# The following objects are masked from 'package:base':\\\#\# \\\#\# \ \ \ \ date, intersect, setdiff, union}}\begin{alltt}
\hlkwd{library}\hlstd{(}\hlstr{"magick"}\hlstd{)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Linking to ImageMagick 6.9.12.3\\\#\# Enabled features: cairo, fontconfig, freetype, heic, lcms, pango, raw, rsvg, webp\\\#\# Disabled features: fftw, ghostscript, x11}}\begin{alltt}
\hlkwd{library}\hlstd{(}\hlstr{"rpart"}\hlstd{)}
\hlkwd{library}\hlstd{(}\hlstr{"visdat"}\hlstd{)}
\hlkwd{library}\hlstd{(}\hlstr{"patchwork"}\hlstd{)}
\hlkwd{library}\hlstd{(}\hlstr{"withr"}\hlstd{)}
\hlkwd{library}\hlstd{(}\hlstr{"targets"}\hlstd{)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# \\\#\# Attaching package: 'targets'\\\#\# \\\#\# The following object is masked from 'package:testthat':\\\#\# \\\#\# \ \ \ \ matches}}\begin{alltt}
\hlkwd{library}\hlstd{(}\hlstr{"tidyverse"}\hlstd{)}
\hlkwd{library}\hlstd{(}\hlstr{"janitor"}\hlstd{)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# \\\#\# Attaching package: 'janitor'\\\#\# \\\#\# The following objects are masked from 'package:stats':\\\#\# \\\#\# \ \ \ \ chisq.test, fisher.test}}\begin{alltt}
\hlkwd{library}\hlstd{(}\hlstr{"knitr"}\hlstd{)}
\hlkwd{library}\hlstd{(}\hlstr{"cowplot"}\hlstd{)}
\end{alltt}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# \\\#\# Attaching package: 'cowplot'\\\#\# \\\#\# The following object is masked from 'package:patchwork':\\\#\# \\\#\# \ \ \ \ align\_plots\\\#\# \\\#\# The following object is masked from 'package:lubridate':\\\#\# \\\#\# \ \ \ \ stamp\\\#\# \\\#\# The following object is masked from 'package:ggthemes':\\\#\# \\\#\# \ \ \ \ theme\_map}}\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{housing_raw} \hlkwb{<-}
  \hlkwd{read_csv}\hlstd{(}\hlstr{"data/melbourne_housing_raw.csv"}\hlstd{)} \hlopt{%>%}
  \hlkwd{clean_names}\hlstd{()} \hlopt{%>%}
  \hlkwd{rename}\hlstd{(}\hlkwc{region_name} \hlstd{= regionname,}
         \hlkwc{property_count} \hlstd{= propertycount)} \hlopt{%>%}
  \hlkwd{mutate}\hlstd{(}\hlkwc{date} \hlstd{=} \hlkwd{dmy}\hlstd{(date))} \hlopt{%>%}
  \hlkwd{rename}\hlstd{(}\hlkwc{latitude} \hlstd{= lattitude,}
         \hlkwc{longitude} \hlstd{= longtitude)} \hlopt{%>%}
  \hlcom{# let's create monthly quarters}
  \hlkwd{mutate}\hlstd{(}\hlkwc{yr_qtr} \hlstd{=} \hlkwd{as.ordered}\hlstd{(}\hlkwd{quarter}\hlstd{(date,}
                                     \hlkwc{with_year} \hlstd{=} \hlnum{TRUE}\hlstd{,}
                                     \hlkwc{fiscal_start} \hlstd{=} \hlnum{1}\hlstd{)))} \hlopt{%>%}
  \hlcom{# drop price}
  \hlkwd{drop_na}\hlstd{(price)} \hlopt{%>%}
  \hlcom{# make price the log of price}
  \hlkwd{mutate}\hlstd{(}\hlkwc{price} \hlstd{=} \hlkwd{log10}\hlstd{(price))}
\end{alltt}

{\ttfamily\noindent\color{warningcolor}{\#\# Warning: One or more parsing issues, call `problems()` on your data frame for details, e.g.:\\\#\# \ \ dat <- vroom(...)\\\#\# \ \ problems(dat)}}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Rows: 34857 Columns: 20\\\#\# -- Column specification ------------------------------------------------------------------\\\#\# Delimiter: "{},"{}\\\#\# chr \ (7): Suburb, Type, Method, SellerG, Date, CouncilArea, Regionname\\\#\# dbl (13): Rooms, Price, Distance, Postcode, Bedroom2, Bathroom, Car, Landsize, Buildin...\\\#\# \\\#\# i Use `spec()` to retrieve the full column specification for this data.\\\#\# i Specify the column types or set `show\_col\_types = FALSE` to quiet this message.}}\begin{alltt}
\hlcom{# compact down the seller levels}
\hlstd{count_dat} \hlkwb{<-} \hlkwd{count}\hlstd{(housing_raw,seller_g)} \hlopt{%>%}
  \hlkwd{arrange}\hlstd{(}\hlopt{-}\hlstd{n)} \hlopt{%>%}
  \hlkwd{mutate}\hlstd{(}\hlkwc{cumu} \hlstd{=} \hlkwd{cumsum}\hlstd{(n),}
         \hlkwc{pct} \hlstd{= cumu} \hlopt{/} \hlkwd{nrow}\hlstd{(housing_raw),}
         \hlkwc{pct_change} \hlstd{= pct} \hlopt{-} \hlkwd{lag}\hlstd{(pct))} \hlopt{%>%}
  \hlkwd{rowid_to_column}\hlstd{()} \hlopt{%>%}
  \hlkwd{mutate}\hlstd{(}\hlkwc{seller} \hlstd{=} \hlkwd{case_when}\hlstd{(}
    \hlstd{pct} \hlopt{<} \hlnum{0.52} \hlopt{~} \hlkwd{as.character}\hlstd{(seller_g),}
    \hlstd{pct} \hlopt{>} \hlnum{0.52} \hlopt{&} \hlstd{pct} \hlopt{<} \hlnum{0.60} \hlopt{~} \hlstr{"seller_g1"}\hlstd{,}
    \hlstd{pct} \hlopt{>} \hlnum{0.60} \hlopt{&} \hlstd{pct} \hlopt{<} \hlnum{0.65} \hlopt{~} \hlstr{"seller_g2"}\hlstd{,}
    \hlstd{pct} \hlopt{>} \hlnum{0.65} \hlopt{&} \hlstd{pct} \hlopt{<} \hlnum{0.70} \hlopt{~} \hlstr{"seller_g3"}\hlstd{,}
    \hlstd{pct} \hlopt{>} \hlnum{0.70} \hlopt{&} \hlstd{pct} \hlopt{<} \hlnum{0.75} \hlopt{~} \hlstr{"seller_g4"}\hlstd{,}
    \hlstd{pct} \hlopt{>} \hlnum{0.75} \hlopt{&} \hlstd{pct} \hlopt{<} \hlnum{0.80} \hlopt{~} \hlstr{"seller_g5"}\hlstd{,}
    \hlstd{pct} \hlopt{>} \hlnum{0.80} \hlopt{&} \hlstd{pct} \hlopt{<} \hlnum{0.85} \hlopt{~} \hlstr{"seller_g6"}\hlstd{,}
    \hlstd{pct} \hlopt{>} \hlnum{0.85} \hlopt{&} \hlstd{pct} \hlopt{<} \hlnum{0.90} \hlopt{~} \hlstr{"seller_g6"}\hlstd{,}
    \hlstd{pct} \hlopt{>} \hlnum{0.90} \hlopt{&} \hlstd{pct} \hlopt{<} \hlnum{0.95} \hlopt{~} \hlstr{"seller_g6"}\hlstd{,}
    \hlstd{pct} \hlopt{>} \hlnum{0.95} \hlopt{~} \hlstr{"seller_g5"}
  \hlstd{))}

\hlcom{# drop these three variables, as we won't use them in analysis, they}
\hlcom{# would need extensive recoding}

\hlstd{housing} \hlkwb{<-} \hlstd{housing_raw} \hlopt{%>%}
  \hlkwd{mutate_if}\hlstd{(is.character, as.factor)} \hlopt{%>%}
  \hlkwd{mutate_at}\hlstd{(}\hlkwd{vars}\hlstd{(rooms,}
                 \hlstd{bedroom2,}
                 \hlstd{bathroom,}
                 \hlstd{car),}
            \hlstd{as.factor)} \hlopt{%>%}
  \hlcom{# recode / collapse factors with many levels and few obs}
  \hlkwd{mutate}\hlstd{(}\hlkwc{bathroom} \hlstd{=} \hlkwd{fct_other}\hlstd{(bathroom,}
                              \hlkwc{drop} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"4"}\hlstd{,} \hlstr{"5"}\hlstd{,} \hlstr{"6"}\hlstd{,} \hlstr{"7"}\hlstd{,} \hlstr{"8"}\hlstd{,} \hlstr{"9"}\hlstd{),}
                              \hlkwc{other_level} \hlstd{=} \hlstr{"4+"}\hlstd{))} \hlopt{%>%}
  \hlkwd{mutate}\hlstd{(}\hlkwc{bedroom2} \hlstd{=} \hlkwd{fct_other}\hlstd{(bedroom2,}
                              \hlkwc{drop} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"5"}\hlstd{,} \hlstr{"6"}\hlstd{,} \hlstr{"7"}\hlstd{,} \hlstr{"8"}\hlstd{,} \hlstr{"9"}\hlstd{,} \hlstr{"10"}\hlstd{,}
                                       \hlstr{"12"}\hlstd{,} \hlstr{"16"}\hlstd{,} \hlstr{"20"}\hlstd{),}
                              \hlkwc{other_level} \hlstd{=} \hlstr{"5+"}\hlstd{))} \hlopt{%>%}
  \hlkwd{mutate}\hlstd{(}\hlkwc{rooms} \hlstd{=} \hlkwd{fct_other}\hlstd{(rooms,}
                           \hlkwc{drop} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"6"}\hlstd{,} \hlstr{"7"}\hlstd{,} \hlstr{"8"}\hlstd{,} \hlstr{"9"}\hlstd{,} \hlstr{"10"}\hlstd{,}
                                    \hlstr{"12"}\hlstd{,} \hlstr{"16"}\hlstd{),}
                           \hlkwc{other_level} \hlstd{=} \hlstr{"6+"}\hlstd{))} \hlopt{%>%}
  \hlkwd{mutate}\hlstd{(}\hlkwc{car} \hlstd{=} \hlkwd{fct_other}\hlstd{(car,}
                         \hlkwc{drop} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"4"}\hlstd{,} \hlstr{"5"}\hlstd{,} \hlstr{"6"}\hlstd{,} \hlstr{"7"}\hlstd{,} \hlstr{"8"}\hlstd{,} \hlstr{"9"}\hlstd{,} \hlstr{"10"}\hlstd{,}
                                  \hlstr{"11"}\hlstd{,} \hlstr{"18"}\hlstd{),}
                         \hlkwc{other_level} \hlstd{=} \hlstr{"4+"}\hlstd{))} \hlopt{%>%}
  \hlkwd{left_join}\hlstd{(}\hlkwd{select}\hlstd{(count_dat, seller_g, seller),} \hlkwc{by} \hlstd{=} \hlstr{"seller_g"}\hlstd{)}  \hlopt{%>%}
  \hlcom{# drop seller_g}
  \hlkwd{select}\hlstd{(}\hlopt{-}\hlstd{seller_g)} \hlopt{%>%}
  \hlkwd{mutate}\hlstd{(}\hlkwc{seller} \hlstd{=} \hlkwd{as.factor}\hlstd{(seller))}
\end{alltt}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{p1} \hlkwb{<-} \hlkwd{ggplot}\hlstd{(oceanbuoys,}
       \hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= humidity,}
           \hlkwc{y} \hlstd{= air_temp_c))} \hlopt{+}
  \hlkwd{geom_point}\hlstd{(}\hlkwc{size} \hlstd{=} \hlnum{0.5}\hlstd{)} \hlopt{+}
  \hlkwd{theme}\hlstd{(}\hlkwc{aspect.ratio} \hlstd{=} \hlnum{1}\hlstd{)} \hlopt{+}
  \hlkwd{labs}\hlstd{(}\hlkwc{x} \hlstd{=} \hlstr{"Humidity"}\hlstd{,}
       \hlkwc{y} \hlstd{=} \hlstr{"Air Temperature (C)"}\hlstd{)}

\hlstd{p1} \hlopt{+} \hlstd{grid}\hlopt{::}\hlkwd{textGrob}\hlstd{(}
  \hlkwc{label} \hlstd{=} \hlstr{"Warning message:
removed 171 rows containing 
missing values (geom_point)."}\hlstd{,}
\hlkwc{gp} \hlstd{= grid}\hlopt{::}\hlkwd{gpar}\hlstd{(}\hlkwc{fontsize} \hlstd{=} \hlnum{14}\hlstd{,}
                \hlkwc{fontfamily} \hlstd{=} \hlstr{"Courier"}\hlstd{,}
                \hlkwc{col} \hlstd{=} \hlstr{"salmon"}\hlstd{,}
                \hlkwc{fontface} \hlstd{=} \hlstr{"bold"}\hlstd{))}
\end{alltt}
\end{kframe}\begin{figure}

{\centering \includegraphics[width=0.75\linewidth]{figure/code-Rnwwarning-1}

}

\caption[How \pkg{ggplot2} behaves when displaying missing values]{How \pkg{ggplot2} behaves when displaying missing values. A warning message is displayed, but missing values are not shown in the plot.}\label{fig:warning}
\end{figure}

\end{knitrout}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{gg_boxplot} \hlkwb{<-}  \hlkwa{function}\hlstd{(}\hlkwc{data}\hlstd{)\{}
  \hlkwd{ggplot}\hlstd{(data,}
       \hlkwd{aes_string}\hlstd{(}\hlkwc{x} \hlstd{=} \hlstr{"year"}\hlstd{,}
                  \hlkwc{y} \hlstd{=} \hlstr{"score"}\hlstd{))} \hlopt{+}
  \hlkwd{geom_boxplot}\hlstd{()}
\hlstd{\}}

\hlstd{p1} \hlkwb{<-} \hlkwd{gg_boxplot}\hlstd{(school_scores_all} \hlopt{%>%} \hlkwd{filter}\hlstd{(NA_type} \hlopt{==} \hlstr{"complete"}\hlstd{))}
\end{alltt}

{\ttfamily\noindent\color{warningcolor}{\#\# Warning: `aes\_string()` was deprecated in ggplot2 3.0.0.\\\#\# i Please use tidy evaluation ideoms with `aes()`}}\begin{alltt}
\hlstd{p2} \hlkwb{<-} \hlkwd{gg_boxplot}\hlstd{(school_scores_all} \hlopt{%>%} \hlkwd{filter}\hlstd{(NA_type} \hlopt{==} \hlstr{"year_NA"}\hlstd{))}
\hlstd{p3} \hlkwb{<-} \hlkwd{gg_boxplot}\hlstd{(school_scores_all} \hlopt{%>%} \hlkwd{filter}\hlstd{(NA_type} \hlopt{==} \hlstr{"score_NA"}\hlstd{))}
\hlstd{p4} \hlkwb{<-} \hlkwd{gg_boxplot}\hlstd{(school_scores_all} \hlopt{%>%} \hlkwd{filter}\hlstd{(NA_type} \hlopt{==} \hlstr{"both_NA"}\hlstd{))}

\hlkwd{plot_grid}\hlstd{(p1, p2, p3, p4,} \hlkwc{nrow} \hlstd{=} \hlnum{1}\hlstd{,} \hlkwc{labels} \hlstd{= LETTERS[}\hlnum{1}\hlopt{:}\hlnum{4}\hlstd{])}
\end{alltt}

{\ttfamily\noindent\color{warningcolor}{\#\# Warning: Removed 11 rows containing non-finite values (`stat\_boxplot()`).}}

{\ttfamily\noindent\color{warningcolor}{\#\# Warning: Removed 13 rows containing non-finite values (`stat\_boxplot()`).}}\end{kframe}\begin{figure}

{\centering \includegraphics[width=1\linewidth]{figure/code-Rnwgg-box-na-1}

}

\caption[\pkg{ggplot2} provides different visualizations depending on what type of data has missing values for data of student test scores in school year]{\pkg{ggplot2} provides different visualizations depending on what type of data has missing values for data of student test scores in school year. (A) Data is complete; (B) Missings are only in year - an additional "NA" boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and year, additional missing information is shown. The missingness category is only shown when there are missings in categorical variables such as year (plots (B) and (D)). In (C), no missingness information is given on the graphic, despite there being missings in score, and a warning message is displayed about the number of missing values omitted.}\label{fig:gg-box-na}
\end{figure}

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{dat_ms} \hlkwb{<-} \hlkwd{tribble}\hlstd{(}\hlopt{~}\hlstd{x,}    \hlopt{~}\hlstd{y,}     \hlopt{~}\hlstd{z,}
                   \hlnum{1}\hlstd{,}   \hlstr{"A"}\hlstd{,}   \hlopt{-}\hlnum{100}\hlstd{,}
                   \hlnum{3}\hlstd{,} \hlstr{"N/A"}\hlstd{,}    \hlopt{-}\hlnum{99}\hlstd{,}
                  \hlnum{NA}\hlstd{,}   \hlnum{NA}\hlstd{,}     \hlopt{-}\hlnum{98}\hlstd{,}
                 \hlopt{-}\hlnum{99}\hlstd{,}  \hlstr{"E"}\hlstd{,}    \hlopt{-}\hlnum{101}\hlstd{,}
                 \hlopt{-}\hlnum{98}\hlstd{,}  \hlstr{"F"}\hlstd{,}      \hlopt{-}\hlnum{1}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{tibble}\hlstd{(}
  \hlkwc{`Function`} \hlstd{=} \hlkwd{c}\hlstd{(}
    \hlstr{"add_n_miss(data)"}\hlstd{,}
    \hlstr{"add_any_miss(data)"}\hlstd{,}
    \hlstr{"add_prop_miss(data)"}\hlstd{,}
    \hlstr{"add_miss_cluster(data)"}
    \hlstd{),}
  \hlkwc{`Adds column which:`} \hlstd{=} \hlkwd{c}\hlstd{(}
    \hlstr{"contains the number missing values in a row"}\hlstd{,}
    \hlstr{"contains whether there are any missing values in a row"}\hlstd{,}
    \hlstr{"contains the proportion of missing values in a row"}\hlstd{,}
    \hlstr{"contains the missing value cluster"}
  \hlstd{))} \hlopt{%>%}
    \hlkwd{kable}\hlstd{(}\hlkwc{caption} \hlstd{=} \hlstr{'Overview of the "add" functions in naniar'}\hlstd{)}
\end{alltt}
\end{kframe}\begin{table}

\caption{\label{tab:add-missing-info}Overview of the "add" functions in naniar}
\centering
\begin{tabular}[t]{l|l}
\hline
Function & Adds column which:\\
\hline
add\_n\_miss(data) & contains the number missing values in a row\\
\hline
add\_any\_miss(data) & contains whether there are any missing values in a row\\
\hline
add\_prop\_miss(data) & contains the proportion of missing values in a row\\
\hline
add\_miss\_cluster(data) & contains the missing value cluster\\
\hline
\end{tabular}
\end{table}

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{nabular}\hlstd{(dat_ms)}
\end{alltt}
\begin{verbatim}
## # A tibble: 5 x 6
##       x y         z x_NA  y_NA  z_NA 
##   <dbl> <chr> <dbl> <fct> <fct> <fct>
## 1     1 A      -100 !NA   !NA   !NA  
## 2     3 N/A     -99 !NA   !NA   !NA  
## 3    NA <NA>    -98 NA    NA    !NA  
## 4   -99 E      -101 !NA   !NA   !NA  
## 5   -98 F        -1 !NA   !NA   !NA
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{nabular}\hlstd{(dat_ms)} \hlopt{%>%}
  \hlkwd{recode_shadow}\hlstd{(}\hlkwc{x} \hlstd{=} \hlkwd{.where}\hlstd{(x} \hlopt{== -}\hlnum{99} \hlopt{~} \hlstr{"broken_sensor"}\hlstd{))}
\end{alltt}
\begin{verbatim}
## # A tibble: 5 x 6
##       x y         z x_NA             y_NA  z_NA 
##   <dbl> <chr> <dbl> <fct>            <fct> <fct>
## 1     1 A      -100 !NA              !NA   !NA  
## 2     3 N/A     -99 !NA              !NA   !NA  
## 3    NA <NA>    -98 NA               NA    !NA  
## 4   -99 E      -101 NA_broken_sensor !NA   !NA  
## 5   -98 F        -1 !NA              !NA   !NA
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{aq_imputed} \hlkwb{<-} \hlkwd{nabular}\hlstd{(airquality)} \hlopt{%>%}
  \hlkwd{as.data.frame}\hlstd{()} \hlopt{%>%}
  \hlstd{simputation}\hlopt{::}\hlkwd{impute_lm}\hlstd{(Ozone} \hlopt{~} \hlstd{Temp} \hlopt{+} \hlstd{Wind)} \hlopt{%>%}
  \hlstd{simputation}\hlopt{::}\hlkwd{impute_lm}\hlstd{(Solar.R} \hlopt{~} \hlstd{Temp} \hlopt{+} \hlstd{Wind)} \hlopt{%>%}
  \hlkwd{add_label_shadow}\hlstd{()}

\hlkwd{head}\hlstd{(aq_imputed)}
\end{alltt}
\begin{verbatim}
##       Ozone  Solar.R Wind Temp Month Day Ozone_NA Solar.R_NA Wind_NA Temp_NA Month_NA
## 1  41.00000 190.0000  7.4   67     5   1      !NA        !NA     !NA     !NA      !NA
## 2  36.00000 118.0000  8.0   72     5   2      !NA        !NA     !NA     !NA      !NA
## 3  12.00000 149.0000 12.6   74     5   3      !NA        !NA     !NA     !NA      !NA
## 4  18.00000 313.0000 11.5   62     5   4      !NA        !NA     !NA     !NA      !NA
## 5 -11.67673 127.4317 14.3   56     5   5       NA         NA     !NA     !NA      !NA
## 6  28.00000 159.5042 14.9   66     5   6      !NA         NA     !NA     !NA      !NA
##   Day_NA any_missing
## 1    !NA Not Missing
## 2    !NA Not Missing
## 3    !NA Not Missing
## 4    !NA Not Missing
## 5    !NA     Missing
## 6    !NA     Missing
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}

{\centering \includegraphics[width=1\linewidth]{figure/code-Rnwtrack-impute-example-1}

}

\caption[Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values from a linear model]{Scatterplot (A) and density plots (B and C) of ozone and solar radiation from the airquality dataset containing imputed values from a linear model. Imputed values are colored green, and data values orange. Imputed values are similar, but slightly trended to the mean.}\label{fig:track-impute-example}
\end{figure}

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{table}

\caption{\label{tab:impute-summary-out}Output of \pkg{dplyr} summary statistics of imputed vs non imputed values for the variable "Ozone". The "any\_missing" column denotes imputed values ("Missing", since they were previously missing), and non-imputed values ("Not Missing"). The mean and median values are similar, but the minimum and maximum values are very different.}
\centering
\begin{tabular}[t]{l|r|r|r|r}
\hline
any\_missing & min & mean & median & max\\
\hline
Missing & -16.86418 & 41.22494 & 45.4734 & 78\\
\hline
Not Missing & 1.00000 & 42.09910 & 31.0000 & 168\\
\hline
\end{tabular}
\end{table}

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}

{\ttfamily\noindent\color{warningcolor}{\#\# Warning: `gather\_()` was deprecated in tidyr 1.2.0.\\\#\# i Please use `gather()` instead.\\\#\# i The deprecated feature was likely used in the visdat package.\\\#\# \ \ Please report the issue at <]8;;https://github.com/ropensci/visdat/issueshttps://github.com/ropensci/visdat/issues]8;;>.}}

{\ttfamily\noindent\color{warningcolor}{\#\# Warning: The `guide` argument in `scale\_*()` cannot be `FALSE`. This was deprecated in ggplot2\\\#\# 3.3.4.\\\#\# i Please use "{}none"{} instead.\\\#\# i The deprecated feature was likely used in the naniar package.\\\#\# \ \ Please report the issue at <]8;;https://github.com/njtierney/naniar/issueshttps://github.com/njtierney/naniar/issues]8;;>.}}\end{kframe}\begin{figure}

{\centering \includegraphics[width=0.9\linewidth]{figure/code-Rnwgg-miss-case-var-1}

}

\caption[Graphical summaries of missingness in the airquality data]{Graphical summaries of missingness in the airquality data. Missings in variables (A) and cases (B), and for an overview of all missingness as a heatmap in (C), and with clustering applied (D). There are missing values in Ozone and Solar.R, with Ozone having more missings. Not many cases have two missings. Most missingness is from cases with one missing value. The default output (C) and ordered by clustering on rows and columns (D). These overviews are made possible using the shadow matrix in long form. There are only missings in ozone and solar radiation, and there appears to be some structure to their missingness.}\label{fig:gg-miss-case-var}
\end{figure}

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{gg_miss_upset}\hlstd{(airquality)}
\end{alltt}
\end{kframe}\begin{figure}

{\centering \includegraphics[width=.6\linewidth]{figure/code-Rnwairquality-upset-1}

}

\caption[The pattern of missingness in the airquality dataset shown in an upset plot]{The pattern of missingness in the airquality dataset shown in an upset plot. Only Ozone and Solar.R have missing values, and Ozone has the most missing values. There are 2 cases where both Solar.R and Ozone have missing values.}\label{fig:airquality-upset}
\end{figure}

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{airquality} \hlopt{%>%}
  \hlkwd{nabular}\hlstd{()} \hlopt{%>%}
  \hlkwd{impute_below_all}\hlstd{()} \hlopt{%>%}
  \hlkwd{ggplot}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= Ozone,}
             \hlkwc{fill} \hlstd{= Ozone_NA))} \hlopt{+}
  \hlkwd{geom_histogram}\hlstd{()} \hlopt{+}
  \hlkwd{scale_fill_brewer}\hlstd{(}\hlkwc{palette} \hlstd{=} \hlstr{"Dark2"}\hlstd{)} \hlopt{+}
  \hlkwd{labs}\hlstd{(}\hlkwc{tag} \hlstd{=} \hlstr{"A"}\hlstd{)} \hlopt{+}
  \hlkwd{theme}\hlstd{(}\hlkwc{legend.position} \hlstd{=} \hlstr{"bottom"}\hlstd{)}

\hlstd{airquality} \hlopt{%>%}
  \hlkwd{nabular}\hlstd{()} \hlopt{%>%}
\hlkwd{ggplot}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= Temp))} \hlopt{+}
  \hlkwd{geom_histogram}\hlstd{(}\hlkwc{na.rm} \hlstd{=} \hlnum{TRUE}\hlstd{)} \hlopt{+}
  \hlkwd{facet_wrap}\hlstd{(}\hlopt{~}\hlstd{Ozone_NA)} \hlopt{+}
  \hlkwd{labs}\hlstd{(}\hlkwc{tag} \hlstd{=} \hlstr{"B"}\hlstd{)}

\hlstd{airquality} \hlopt{%>%}
  \hlkwd{nabular}\hlstd{()} \hlopt{%>%}
\hlkwd{ggplot}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= Temp,}
           \hlkwc{colour} \hlstd{= Ozone_NA))} \hlopt{+}
  \hlkwd{geom_density}\hlstd{(}\hlkwc{na.rm} \hlstd{=} \hlnum{TRUE}\hlstd{)} \hlopt{+}
  \hlkwd{scale_colour_brewer}\hlstd{(}\hlkwc{palette} \hlstd{=} \hlstr{"Dark2"}\hlstd{)}  \hlopt{+}
  \hlkwd{theme}\hlstd{(}\hlkwc{legend.position} \hlstd{=} \hlstr{"bottom"}\hlstd{)} \hlopt{+}
  \hlkwd{labs}\hlstd{(}\hlkwc{tag} \hlstd{=} \hlstr{"C"}\hlstd{)}
\end{alltt}
\end{kframe}\begin{figure}

{\centering \includegraphics[width=0.75\linewidth]{figure/code-Rnwimpute-shift-histogram-1} 
\includegraphics[width=0.75\linewidth]{figure/code-Rnwimpute-shift-histogram-2} 
\includegraphics[width=0.75\linewidth]{figure/code-Rnwimpute-shift-histogram-3}

}

\caption[Univariate summaries of missingness]{Univariate summaries of missingness. (A) A histogram using nabular data to show the values and missings in ozone. Values are imputed below the range to show the number of missings in ozone and colored according to missingness of ozone (`Ozone\_NA`). There are about 35 missings in Ozone. Panel C shows temperature according to missingness in ozone from in the airquality dataset. A histogram of temperature facetted by the missingness of ozone (B), or a density of temperature colored by missingness in ozone (C). These show a cluster of low temperature observations with missing ozone values, but temperature is otherwise similar.}\label{fig:impute-shift-histogram}
\end{figure}

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{p1} \hlkwb{<-}
\hlkwd{ggplot}\hlstd{(}\hlkwc{data} \hlstd{= airquality,}
       \hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= Ozone,}
           \hlkwc{y} \hlstd{= Solar.R))} \hlopt{+}
  \hlkwd{geom_miss_point}\hlstd{()} \hlopt{+}
  \hlkwd{scale_colour_brewer}\hlstd{(}\hlkwc{palette} \hlstd{=} \hlstr{"Dark2"}\hlstd{)} \hlopt{+}
  \hlkwd{theme}\hlstd{(}\hlkwc{legend.position} \hlstd{=} \hlstr{"bottom"}\hlstd{)} \hlopt{+}
  \hlkwd{labs}\hlstd{(}\hlkwc{tag} \hlstd{=} \hlstr{"A"}\hlstd{)}

\hlstd{p2} \hlkwb{<-}
\hlkwd{ggplot}\hlstd{(}\hlkwc{data} \hlstd{= airquality,}
       \hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= Temp,}
           \hlkwc{y} \hlstd{= Ozone))} \hlopt{+}
  \hlkwd{geom_miss_point}\hlstd{()} \hlopt{+}
  \hlkwd{scale_colour_brewer}\hlstd{(}\hlkwc{palette} \hlstd{=} \hlstr{"Dark2"}\hlstd{)} \hlopt{+}
  \hlkwd{theme}\hlstd{(}\hlkwc{legend.position} \hlstd{=} \hlstr{"bottom"}\hlstd{)}  \hlopt{+}
  \hlkwd{labs}\hlstd{(}\hlkwc{tag} \hlstd{=} \hlstr{"B"}\hlstd{)}

\hlstd{gridExtra}\hlopt{::}\hlkwd{grid.arrange}\hlstd{(p1, p2,} \hlkwc{ncol} \hlstd{=} \hlnum{2}\hlstd{)}
\end{alltt}
\end{kframe}\begin{figure}

{\centering \includegraphics[width=.6\linewidth]{figure/code-Rnwgeom-miss-1}

}

\caption[Scatterplots with missings displayed at 10 percent below for the airquality dataset]{Scatterplots with missings displayed at 10 percent below for the airquality dataset. Scatterplots of ozone and solar radiation (A), and ozone and temperature (B). There are missings in ozone and solar radiation, but not temperature.}\label{fig:geom-miss}
\end{figure}

\end{knitrout}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}

{\centering \includegraphics[width=1\linewidth]{figure/code-Rnwparallel-cord-plot-1}

}

\caption[Parallel coordinate plot shows missing values imputed 10\% below range for the oceanbuoys dataset]{Parallel coordinate plot shows missing values imputed 10\% below range for the oceanbuoys dataset. Values are colored by missingness of humidity. Humidity is missing for low air and sea temperatures, and is missing for one year and one location.}\label{fig:parallel-cord-plot}
\end{figure}

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{table}

\caption{\label{tab:n-prop-pct-miss-complete}Single number summaries of missingness and completeness of the airquality dataset. The functions follow consistent naming, making them easy to remember, and their use clear.}
\centering
\begin{tabular}[t]{l|r|l|r}
\hline
Missing function & Missing value & Complete function & Complete value\\
\hline
n\_miss & 44.00 & n\_complete & 874.00\\
\hline
prop\_miss & 0.05 & prop\_complete & 0.95\\
\hline
pct\_miss & 4.79 & pct\_complete & 95.21\\
\hline
pct\_miss\_case & 27.45 & prop\_complete\_case & 72.55\\
\hline
pct\_miss\_var & 33.33 & pct\_complete\_var & 66.67\\
\hline
\end{tabular}
\end{table}

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{miss_var_summary}\hlstd{(airquality)} \hlopt{%>%}
  \hlkwd{kable}\hlstd{(}
    \hlkwc{caption} \hlstd{=} \hlstr{"\textbackslash{}\textbackslash{}texttt\{miss\textbackslash{}\textbackslash{}char`_var\textbackslash{}\textbackslash{}char`_summary(airquality)\} provides the number and percent of missings in each variable in airquality. Only ozone and solar radiation have missing values."}\hlstd{,}
    \hlkwc{digits} \hlstd{=} \hlnum{1}\hlstd{)}
\end{alltt}
\end{kframe}\begin{table}

\caption{\label{tab:miss-var-summary}\texttt{miss\char`_var\char`_summary(airquality)} provides the number and percent of missings in each variable in airquality. Only ozone and solar radiation have missing values.}
\centering
\begin{tabular}[t]{l|r|r}
\hline
variable & n\_miss & pct\_miss\\
\hline
Ozone & 37 & 24.2\\
\hline
Solar.R & 7 & 4.6\\
\hline
Wind & 0 & 0.0\\
\hline
Temp & 0 & 0.0\\
\hline
Month & 0 & 0.0\\
\hline
Day & 0 & 0.0\\
\hline
\end{tabular}
\end{table}

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{miss_var_table}\hlstd{(airquality)} \hlopt{%>%}
  \hlkwd{kable}\hlstd{(}
    \hlkwc{caption} \hlstd{=} \hlstr{"The output of \textbackslash{}\textbackslash{}code\{miss\textbackslash{}\textbackslash{}char`_var\textbackslash{}\textbackslash{}char`_table(airquality)\}, tabulating the amount of missing data in each variable in airquality. This shows the number of variables with 0, 7, and 37 missings, and the percentage of variables with those amounts of missingness. There are few missingness patterns."}\hlstd{,}
    \hlkwc{digits} \hlstd{=} \hlnum{1}\hlstd{)}
\end{alltt}
\end{kframe}\begin{table}

\caption{\label{tab:miss-var-table}The output of \code{miss\char`_var\char`_table(airquality)}, tabulating the amount of missing data in each variable in airquality. This shows the number of variables with 0, 7, and 37 missings, and the percentage of variables with those amounts of missingness. There are few missingness patterns.}
\centering
\begin{tabular}[t]{r|r|r}
\hline
n\_miss\_in\_var & n\_vars & pct\_vars\\
\hline
0 & 4 & 66.7\\
\hline
7 & 1 & 16.7\\
\hline
37 & 1 & 16.7\\
\hline
\end{tabular}
\end{table}

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{airquality} \hlopt{%>%}
  \hlkwd{group_by}\hlstd{(Month)} \hlopt{%>%}
  \hlkwd{miss_var_summary}\hlstd{()} \hlopt{%>%}
  \hlkwd{ungroup}\hlstd{()} \hlopt{%>%}
  \hlkwd{slice}\hlstd{(}\hlnum{1}\hlopt{:}\hlnum{10}\hlstd{)} \hlopt{%>%}
  \hlkwd{kable}\hlstd{(}\hlkwc{caption} \hlstd{=} \hlstr{"Output of \textbackslash{}\textbackslash{}code\{airquality \textbackslash{}\textbackslash{}%>\textbackslash{}\textbackslash{}% group\textbackslash{}\textbackslash{}char`_by(Month) \textbackslash{}\textbackslash{}%>\textbackslash{}\textbackslash{}% miss\textbackslash{}\textbackslash{}char`_var\textbackslash{}\textbackslash{}char`_summary()\} provides a grouped summary of the missingness in each variable, for each month of the airquality dataset. Only the first 10 rows are shown. There are more ozone missings in June than May."}\hlstd{,}
               \hlkwc{digits} \hlstd{=} \hlnum{1}\hlstd{)}
\end{alltt}
\end{kframe}\begin{table}

\caption{\label{tab:group-miss-var-summary}Output of \code{airquality \%>\% group\char`_by(Month) \%>\% miss\char`_var\char`_summary()} provides a grouped summary of the missingness in each variable, for each month of the airquality dataset. Only the first 10 rows are shown. There are more ozone missings in June than May.}
\centering
\begin{tabular}[t]{r|l|r|r}
\hline
Month & variable & n\_miss & pct\_miss\\
\hline
5 & Ozone & 5 & 16.1\\
\hline
5 & Solar.R & 4 & 12.9\\
\hline
5 & Wind & 0 & 0.0\\
\hline
5 & Temp & 0 & 0.0\\
\hline
5 & Day & 0 & 0.0\\
\hline
6 & Ozone & 21 & 70.0\\
\hline
6 & Solar.R & 0 & 0.0\\
\hline
6 & Wind & 0 & 0.0\\
\hline
6 & Temp & 0 & 0.0\\
\hline
6 & Day & 0 & 0.0\\
\hline
\end{tabular}
\end{table}

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{q1} \hlkwb{<-} \hlkwd{gg_miss_var}\hlstd{(housing,} \hlkwc{show_pct} \hlstd{=} \hlnum{TRUE}\hlstd{)} \hlopt{+} \hlkwd{labs}\hlstd{(}\hlkwc{tag} \hlstd{=} \hlstr{"A"}\hlstd{)}

\hlstd{q2} \hlkwb{<-} \hlkwd{gg_miss_case}\hlstd{(housing,} \hlkwc{show_pct} \hlstd{=} \hlnum{TRUE}\hlstd{)} \hlopt{+} \hlkwd{labs}\hlstd{(}\hlkwc{tag} \hlstd{=} \hlstr{"B"}\hlstd{)}

\hlstd{gridExtra}\hlopt{::}\hlkwd{grid.arrange}\hlstd{(q1, q2,} \hlkwc{ncol} \hlstd{=} \hlnum{2}\hlstd{)}
\end{alltt}
\end{kframe}\begin{figure}

{\centering \includegraphics[width=1\linewidth]{figure/code-Rnwhousing-miss-case-var-1}

}

\caption[The amount of missings in variables (A) and cases (B) for Melbourne housing data]{The amount of missings in variables (A) and cases (B) for Melbourne housing data. (A) Build area and year built have more than 50\% missing, and car, bathroom, bedroom2 and longitude and latitude have about 25\% missings. (B) Cases are missing 5 - 50\% of values. The majority of missingness is in selected cases and variables.}\label{fig:housing-miss-case-var}
\end{figure}

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{vis_miss}\hlstd{(housing,} \hlkwc{cluster} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{sort_miss} \hlstd{=} \hlnum{TRUE}\hlstd{,} \hlkwc{show_perc_col} \hlstd{=} \hlnum{FALSE}\hlstd{)}
\end{alltt}
\end{kframe}\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{figure/code-Rnwapplic-vis-miss-1}

}

\caption[Heatmap of clustered missingness for housing data reveals structured missingness]{Heatmap of clustered missingness for housing data reveals structured missingness. Three groups of missingness are apparent. At the top: building area to longitude; the middle: building area and year built; the end: building area, year built, and landsize.}\label{fig:applic-vis-miss}
\end{figure}

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{gg_miss_upset}\hlstd{(housing,}
              \hlkwc{nsets} \hlstd{=} \hlnum{8}\hlstd{,}
              \hlkwc{order.by} \hlstd{=} \hlstr{"freq"}\hlstd{)}
\end{alltt}
\end{kframe}\begin{figure}

{\centering \includegraphics[width=1\linewidth]{figure/code-Rnwhousing-upset-1}

}

\caption[An upset plot of 8 sets of missingness in the housing data]{An upset plot of 8 sets of missingness in the housing data. Missingness for each variable is shown on the bottom left. Connected dots show co-occurences of missings in variables. Two missingness patterns are clear, year built and building area, and lattitude through to building area.}\label{fig:housing-upset}
\end{figure}

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{t1} \hlkwb{<-}  \hlkwd{miss_var_table}\hlstd{(housing)}
\hlstd{t2} \hlkwb{<-} \hlkwd{miss_case_table}\hlstd{(housing)}

\hlkwd{kable}\hlstd{(}\hlkwd{list}\hlstd{(t1, t2),}
             \hlkwc{caption} \hlstd{=} \hlstr{"Summary tables to help understand missingness patters. Output of \textbackslash{}\textbackslash{}code\{miss\textbackslash{}\textbackslash{}_var\textbackslash{}\textbackslash{}_table(housing)\} (left), tabulating missingness for variables, and output of \textbackslash{}\textbackslash{}code\{miss\textbackslash{}\textbackslash{}_case\textbackslash{}\textbackslash{}_table(housing)\}. There are 13 variables with 0-3 missings, 6 variables have 6000-10000 missings, 2 variables have 15000 – 17000 missings. About 30\textbackslash{}\textbackslash{}% of cases have no missings, 45\textbackslash{}\textbackslash{}% of cases have 1 - 6 missings, and about 23\textbackslash{}\textbackslash{}% of cases have 8 or more missings. There are different patterns of missingness in variables and cases, but they can be broken down into smaller groups."}\hlstd{,}
             \hlkwc{digits} \hlstd{=} \hlnum{1}\hlstd{)}
\end{alltt}
\end{kframe}\begin{table}
\caption{\label{tab:housing-miss-var-case-table}Summary tables to help understand missingness patters. Output of \code{miss\_var\_table(housing)} (left), tabulating missingness for variables, and output of \code{miss\_case\_table(housing)}. There are 13 variables with 0-3 missings, 6 variables have 6000-10000 missings, 2 variables have 15000 – 17000 missings. About 30\% of cases have no missings, 45\% of cases have 1 - 6 missings, and about 23\% of cases have 8 or more missings. There are different patterns of missingness in variables and cases, but they can be broken down into smaller groups.}

\centering
\begin{tabular}[t]{r|r|r}
\hline
n\_miss\_in\_var & n\_vars & pct\_vars\\
\hline
0 & 10 & 47.6\\
\hline
1 & 2 & 9.5\\
\hline
3 & 1 & 4.8\\
\hline
6254 & 2 & 9.5\\
\hline
6441 & 1 & 4.8\\
\hline
6447 & 1 & 4.8\\
\hline
6824 & 1 & 4.8\\
\hline
9265 & 1 & 4.8\\
\hline
15163 & 1 & 4.8\\
\hline
16591 & 1 & 4.8\\
\hline
\end{tabular}
\centering
\begin{tabular}[t]{r|r|r}
\hline
n\_miss\_in\_case & n\_cases & pct\_cases\\
\hline
0 & 8887 & 32.6\\
\hline
1 & 3237 & 11.9\\
\hline
2 & 7231 & 26.5\\
\hline
3 & 1370 & 5.0\\
\hline
4 & 79 & 0.3\\
\hline
5 & 8 & 0.0\\
\hline
6 & 203 & 0.7\\
\hline
8 & 6229 & 22.9\\
\hline
9 & 2 & 0.0\\
\hline
11 & 1 & 0.0\\
\hline
\end{tabular}
\end{table}

\begin{kframe}\begin{alltt}
             \hlcom{# booktabs = TRUE)}
  \hlcom{# kable_styling(latex_options = c("hold_position"))}
\end{alltt}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{housing_cls} \hlkwb{<-} \hlkwd{add_miss_cluster}\hlstd{(housing,} \hlkwc{n_clusters} \hlstd{=} \hlnum{2}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{miss_cls_pre_fit} \hlkwb{<-} \hlstd{housing_cls} \hlopt{%>%}
  \hlkwd{mutate}\hlstd{(}\hlkwc{year} \hlstd{= lubridate}\hlopt{::}\hlkwd{year}\hlstd{(date),}
         \hlkwc{quarter} \hlstd{= lubridate}\hlopt{::}\hlkwd{quarter}\hlstd{(date))} \hlopt{%>%}
  \hlkwd{mutate_at}\hlstd{(}\hlkwd{vars}\hlstd{(year, quarter),}
            \hlstd{as.factor)} \hlopt{%>%}
  \hlkwd{select}\hlstd{(suburb,}
         \hlstd{council_area,}
         \hlstd{postcode,}
         \hlstd{method,}
         \hlstd{region_name,}
         \hlstd{rooms,}
         \hlstd{seller,}
         \hlstd{type,}
         \hlstd{year,}
         \hlstd{quarter,}
         \hlstd{price,}
         \hlstd{property_count,}
         \hlstd{distance,}
         \hlstd{latitude,}
         \hlstd{longitude,}
         \hlstd{miss_cluster)}
\hlstd{miss_cls_fit_rpart} \hlkwb{<-} \hlkwd{rpart}\hlstd{(}\hlkwd{factor}\hlstd{(miss_cluster)} \hlopt{~} \hlstd{.,}
                            \hlkwc{data} \hlstd{= miss_cls_pre_fit)}
\end{alltt}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{rpart.plot}\hlopt{::}\hlkwd{prp}\hlstd{(miss_cls_fit_rpart,}
                \hlkwc{type} \hlstd{=} \hlnum{4}\hlstd{,}
                \hlkwc{extra} \hlstd{=} \hlnum{2}\hlstd{,}
                \hlkwc{fallen.leaves} \hlstd{=} \hlnum{TRUE}\hlstd{,}
                \hlkwc{prefix} \hlstd{=} \hlstr{"cluster = "}\hlstd{,}
                \hlkwc{suffix} \hlstd{=} \hlstr{" \textbackslash{}nObservations"}\hlstd{,}
                \hlkwc{box.col} \hlstd{=} \hlstr{"lightgrey"}\hlstd{,}
                \hlkwc{border.col} \hlstd{=} \hlstr{"grey"}\hlstd{,}
                \hlkwc{branch.col} \hlstd{=} \hlstr{"grey40"}\hlstd{)}
\end{alltt}
\end{kframe}\begin{figure}

{\centering \includegraphics[width=0.9\linewidth]{figure/code-Rnwrpart-plot-1}

}

\caption[Decision tree output predicting missingness clusters]{Decision tree output predicting missingness clusters. Type of house, year quarter, and year were important for predicting missingness cluster. The cluster with the most missingness was for quarters 1 and 4, for 2017 and 2018. Type of house, year, and year quarter are important features related to missingness structure.}\label{fig:rpart-plot}
\end{figure}

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{data_in} \hlkwb{<-} \hlstd{housing_cls} \hlopt{%>%}
  \hlkwd{mutate}\hlstd{(}\hlkwc{year} \hlstd{= lubridate}\hlopt{::}\hlkwd{year}\hlstd{(date),}
         \hlkwc{month} \hlstd{= lubridate}\hlopt{::}\hlkwd{month}\hlstd{(date),}
         \hlkwc{quarter} \hlstd{= lubridate}\hlopt{::}\hlkwd{quarter}\hlstd{(date))} \hlopt{%>%}
  \hlkwd{mutate_at}\hlstd{(}\hlkwd{vars}\hlstd{(year, month, quarter),}
            \hlstd{as.factor)}

\hlstd{node_1} \hlkwb{<-} \hlstd{data_in} \hlopt{%>%}
  \hlkwd{filter}\hlstd{(type} \hlopt{==} \hlstr{"h"}\hlstd{)}

\hlstd{node_2} \hlkwb{<-} \hlstd{data_in} \hlopt{%>%}
  \hlkwd{filter}\hlstd{(type} \hlopt{!=} \hlstr{"h"}\hlstd{,}
         \hlstd{quarter} \hlopt{==} \hlnum{2} \hlopt{|} \hlstd{quarter} \hlopt{==} \hlnum{3}\hlstd{)}

\hlstd{node_3} \hlkwb{<-} \hlstd{data_in} \hlopt{%>%}
  \hlkwd{filter}\hlstd{(type} \hlopt{!=} \hlstr{"h"}\hlstd{,}
         \hlstd{quarter} \hlopt{==} \hlnum{1} \hlopt{|} \hlstd{quarter} \hlopt{==} \hlnum{4}\hlstd{,}
         \hlstd{year} \hlopt{==} \hlnum{2016}\hlstd{)}

\hlstd{node_4} \hlkwb{<-} \hlstd{data_in} \hlopt{%>%}
  \hlkwd{filter}\hlstd{(type} \hlopt{!=} \hlstr{"h"}\hlstd{,}
         \hlstd{quarter} \hlopt{==} \hlnum{1} \hlopt{|} \hlstd{quarter} \hlopt{==} \hlnum{4}\hlstd{,}
         \hlstd{year} \hlopt{==} \hlnum{2017} \hlopt{|} \hlstd{year} \hlopt{==} \hlnum{2018}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{set.seed}\hlstd{(}\hlnum{2022}\hlopt{-}\hlnum{12}\hlopt{-}\hlnum{1}\hlstd{)}
\hlstd{dat_house_lm} \hlkwb{<-} \hlstd{dat_house} \hlopt{%>%}
  \hlkwd{nabular}\hlstd{()} \hlopt{%>%}
  \hlkwd{as.data.frame}\hlstd{()} \hlopt{%>%}
  \hlcom{# postcode}
  \hlkwd{impute_knn}\hlstd{(postcode} \hlopt{~} \hlstd{type} \hlopt{+} \hlstd{quarter} \hlopt{+} \hlstd{year} \hlopt{+} \hlstd{rooms} \hlopt{+} \hlstd{price} \hlopt{+} \hlstd{distance)} \hlopt{%>%}
  \hlcom{# distance}
  \hlkwd{impute_lm}\hlstd{(distance} \hlopt{~} \hlstd{type} \hlopt{+} \hlstd{quarter} \hlopt{+} \hlstd{year} \hlopt{+} \hlstd{rooms} \hlopt{+} \hlstd{price)} \hlopt{%>%}
  \hlcom{# property_count}
  \hlkwd{impute_lm}\hlstd{(property_count} \hlopt{~} \hlstd{type} \hlopt{+} \hlstd{quarter} \hlopt{+} \hlstd{year} \hlopt{+} \hlstd{rooms} \hlopt{+} \hlstd{price} \hlopt{+} \hlstd{distance)} \hlopt{%>%}
  \hlcom{# longitude}
  \hlkwd{impute_lm}\hlstd{(longitude} \hlopt{~} \hlstd{type} \hlopt{+} \hlstd{quarter} \hlopt{+} \hlstd{year} \hlopt{+} \hlstd{rooms} \hlopt{+} \hlstd{price} \hlopt{+} \hlstd{distance)} \hlopt{%>%}
  \hlcom{# latitude}
  \hlkwd{impute_lm}\hlstd{(latitude} \hlopt{~}  \hlstd{type} \hlopt{+} \hlstd{quarter} \hlopt{+} \hlstd{year} \hlopt{+} \hlstd{rooms} \hlopt{+} \hlstd{price} \hlopt{+} \hlstd{distance)} \hlopt{%>%}
  \hlcom{# bedroom2}
  \hlkwd{impute_lm}\hlstd{(bedroom2} \hlopt{~}  \hlstd{type} \hlopt{+} \hlstd{quarter} \hlopt{+} \hlstd{year} \hlopt{+} \hlstd{rooms} \hlopt{+} \hlstd{price} \hlopt{+} \hlstd{distance)} \hlopt{%>%}
  \hlcom{# bathroom}
  \hlkwd{impute_lm}\hlstd{(bathroom} \hlopt{~}  \hlstd{type} \hlopt{+} \hlstd{quarter} \hlopt{+} \hlstd{year} \hlopt{+} \hlstd{rooms} \hlopt{+} \hlstd{price} \hlopt{+} \hlstd{distance)} \hlopt{%>%}
  \hlcom{# car}
  \hlkwd{impute_lm}\hlstd{(car} \hlopt{~} \hlstd{type} \hlopt{+} \hlstd{quarter} \hlopt{+} \hlstd{year} \hlopt{+} \hlstd{rooms} \hlopt{+} \hlstd{price} \hlopt{+} \hlstd{distance)} \hlopt{%>%}
  \hlcom{# landsize}
  \hlkwd{impute_lm}\hlstd{(landsize} \hlopt{~}  \hlstd{type} \hlopt{+} \hlstd{quarter} \hlopt{+} \hlstd{year} \hlopt{+} \hlstd{rooms} \hlopt{+} \hlstd{price} \hlopt{+} \hlstd{distance)} \hlopt{%>%}
  \hlkwd{add_label_shadow}\hlstd{()} \hlopt{%>%}
  \hlcom{# recode integers back}
  \hlkwd{mutate_at}\hlstd{(}\hlkwd{vars}\hlstd{(bedroom2, bathroom, car), as.integer)}
\end{alltt}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{dat_house_cc} \hlkwb{<-} \hlstd{dat_house} \hlopt{%>%} \hlkwd{nabular}\hlstd{()} \hlopt{%>%} \hlkwd{add_label_shadow}\hlstd{()} \hlopt{%>%} \hlkwd{na.omit}\hlstd{()}

\hlstd{bound_models} \hlkwb{<-} \hlkwd{bind_rows}\hlstd{(}\hlkwc{lm} \hlstd{= dat_house_lm,}
                          \hlkwc{knn} \hlstd{= dat_house_knn,}
                          \hlkwc{cc} \hlstd{= dat_house_cc,}
                          \hlkwc{.id} \hlstd{=} \hlstr{"imp_model"}\hlstd{)} \hlopt{%>%}
  \hlkwd{as_tibble}\hlstd{()}
\end{alltt}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{## Now we want to look at the distribution of the values, so we need to do}
\hlcom{## some gathering}
\hlstd{bound_models_gather} \hlkwb{<-} \hlstd{bound_models} \hlopt{%>%}
  \hlkwd{select}\hlstd{(}
         \hlstd{bedroom2,}
         \hlstd{bathroom,}
         \hlstd{car,}
         \hlstd{landsize,}
         \hlstd{any_missing,}
         \hlstd{imp_model)} \hlopt{%>%}
  \hlkwd{gather}\hlstd{(}\hlkwc{key} \hlstd{=} \hlstr{"key"}\hlstd{,}
         \hlkwc{value} \hlstd{=} \hlstr{"value"}\hlstd{,}
         \hlopt{-}\hlstd{any_missing,}
         \hlopt{-}\hlstd{imp_model)}

\hlstd{p1} \hlkwb{<-} \hlstd{bound_models} \hlopt{%>%}
  \hlkwd{ggplot}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= imp_model,}
             \hlkwc{y} \hlstd{= bedroom2))} \hlopt{+}
  \hlkwd{geom_boxplot}\hlstd{()}

\hlstd{p2} \hlkwb{<-} \hlstd{bound_models} \hlopt{%>%}
  \hlkwd{ggplot}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= imp_model,}
             \hlkwc{y} \hlstd{= bathroom))} \hlopt{+}
  \hlkwd{geom_boxplot}\hlstd{()}

\hlstd{p3} \hlkwb{<-} \hlstd{bound_models} \hlopt{%>%}
  \hlkwd{ggplot}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= imp_model,}
             \hlkwc{y} \hlstd{= car))} \hlopt{+}
  \hlkwd{geom_boxplot}\hlstd{()}

\hlstd{p4} \hlkwb{<-} \hlstd{bound_models} \hlopt{%>%}
  \hlkwd{ggplot}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= imp_model,}
             \hlkwc{y} \hlstd{= landsize))} \hlopt{+}
  \hlkwd{geom_boxplot}\hlstd{()} \hlopt{+}
  \hlkwd{scale_y_log10}\hlstd{()}

\hlstd{cowplot}\hlopt{::}\hlkwd{plot_grid}\hlstd{(p1,}
          \hlstd{p2,}
          \hlstd{p3,}
          \hlstd{p4,}
          \hlkwc{nrow} \hlstd{=} \hlnum{1}\hlstd{,}
          \hlkwc{labels} \hlstd{=} \hlstr{"AUTO"}\hlstd{)}
\end{alltt}

{\ttfamily\noindent\color{warningcolor}{\#\# Warning in self\$trans\$transform(x): NaNs produced}}

{\ttfamily\noindent\color{warningcolor}{\#\# Warning: Transformation introduced infinite values in continuous y-axis}}

{\ttfamily\noindent\color{warningcolor}{\#\# Warning: Removed 5891 rows containing non-finite values (`stat\_boxplot()`).}}\end{kframe}\begin{figure}

{\centering \includegraphics[width=0.95\linewidth]{figure/code-Rnwimputed-by-model-1}

}

\caption[Boxplots of complete case data, and data imputed with KNN or linear model for different variables]{Boxplots of complete case data, and data imputed with KNN or linear model for different variables. (A) number of bedrooms, (B) number of bathrooms, (C) number of carspots, and (D) landsize (on a log10 scale). KNN had similar results to complete case, and linear model had a lower median for cars and fewer extreme values for bedrooms.}\label{fig:imputed-by-model}
\end{figure}

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{set.seed}\hlstd{(}\hlnum{2022}\hlopt{-}\hlnum{12}\hlopt{-}\hlnum{1}\hlstd{)}
\hlstd{dat_fit} \hlkwb{<-} \hlstd{dat_house} \hlopt{%>%}
  \hlkwd{mutate_at}\hlstd{(}\hlkwd{vars}\hlstd{(car,}
                 \hlstd{bathroom,}
                 \hlstd{bedroom2,}
                 \hlstd{type,}
                 \hlstd{year,}
                 \hlstd{quarter,}
                 \hlstd{rooms),}
            \hlkwc{.funs} \hlstd{= as.factor)}

\hlstd{dat_fit_knn} \hlkwb{<-}  \hlstd{dat_house_knn} \hlopt{%>%}
  \hlkwd{mutate_at}\hlstd{(}\hlkwd{vars}\hlstd{(car,}
                 \hlstd{bathroom,}
                 \hlstd{bedroom2,}
                 \hlstd{type,}
                 \hlstd{year,}
                 \hlstd{quarter,}
                 \hlstd{rooms),}
            \hlkwc{.funs} \hlstd{= as.factor)}

\hlstd{dat_fit_lm} \hlkwb{<-}  \hlstd{dat_house_lm} \hlopt{%>%}
  \hlkwd{mutate_at}\hlstd{(}\hlkwd{vars}\hlstd{(car,}
                 \hlstd{bathroom,}
                 \hlstd{bedroom2,}
                 \hlstd{type,}
                 \hlstd{year,}
                 \hlstd{quarter,}
                 \hlstd{rooms),}
            \hlkwc{.funs} \hlstd{= as.factor)}

\hlstd{dat_house_cc} \hlkwb{<-} \hlkwd{na.omit}\hlstd{(dat_fit)}

\hlstd{fit_lm_dat_house_cc} \hlkwb{<-} \hlkwd{lm}\hlstd{(price} \hlopt{~} \hlstd{landsize} \hlopt{+}
                            \hlstd{car} \hlopt{+}
                            \hlstd{bathroom} \hlopt{+}
                            \hlstd{bedroom2} \hlopt{+}
                            \hlstd{type} \hlopt{+}
                            \hlstd{year} \hlopt{+}
                            \hlstd{quarter} \hlopt{+}
                            \hlstd{rooms} \hlopt{+}
                            \hlstd{distance,}
                          \hlstd{dat_house_cc)}

\hlstd{fit_lm_dat_house_knn} \hlkwb{<-} \hlkwd{lm}\hlstd{(price} \hlopt{~} \hlstd{landsize} \hlopt{+}
                            \hlstd{car} \hlopt{+}
                            \hlstd{bathroom} \hlopt{+}
                            \hlstd{bedroom2} \hlopt{+}
                            \hlstd{type} \hlopt{+}
                            \hlstd{year} \hlopt{+}
                            \hlstd{quarter} \hlopt{+}
                            \hlstd{rooms} \hlopt{+}
                            \hlstd{distance,}
                           \hlstd{dat_fit_knn)}

\hlstd{fit_lm_dat_house_lm} \hlkwb{<-} \hlkwd{lm}\hlstd{(price} \hlopt{~} \hlstd{landsize} \hlopt{+}
                            \hlstd{car} \hlopt{+}
                            \hlstd{bathroom} \hlopt{+}
                            \hlstd{bedroom2} \hlopt{+}
                            \hlstd{type} \hlopt{+}
                            \hlstd{year} \hlopt{+}
                            \hlstd{quarter} \hlopt{+}
                            \hlstd{rooms} \hlopt{+}
                            \hlstd{distance,}
                          \hlstd{dat_fit_lm)}
\end{alltt}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{tidy_fit_cc} \hlkwb{<-} \hlstd{broom}\hlopt{::}\hlkwd{tidy}\hlstd{(fit_lm_dat_house_cc)}
\hlstd{tidy_fit_knn} \hlkwb{<-} \hlstd{broom}\hlopt{::}\hlkwd{tidy}\hlstd{(fit_lm_dat_house_knn)}
\hlstd{tidy_fit_lm} \hlkwb{<-} \hlstd{broom}\hlopt{::}\hlkwd{tidy}\hlstd{(fit_lm_dat_house_lm)}

\hlstd{tidy_fits} \hlkwb{<-} \hlkwd{bind_rows}\hlstd{(}\hlkwc{cc} \hlstd{= tidy_fit_cc,}
                       \hlkwc{knn} \hlstd{= tidy_fit_knn,}
                       \hlkwc{lm} \hlstd{= tidy_fit_lm,}
                       \hlkwc{.id} \hlstd{=} \hlstr{"model"}\hlstd{)}

\hlstd{tidy_fits_data} \hlkwb{<-}  \hlstd{tidy_fits} \hlopt{%>%}
  \hlkwd{filter}\hlstd{(term} \hlopt{!=} \hlstr{"(Intercept)"}\hlstd{)} \hlopt{%>%}
  \hlkwd{filter}\hlstd{(}\hlkwd{grepl}\hlstd{(}\hlstr{"rooms"}\hlstd{, term))} \hlopt{%>%}
  \hlkwd{select}\hlstd{(model, term, estimate)} \hlopt{%>%}
  \hlkwd{spread}\hlstd{(}\hlkwc{key} \hlstd{= model,}
         \hlkwc{value} \hlstd{= estimate)}

\hlstd{tidy_fits} \hlopt{%>%}
  \hlkwd{filter}\hlstd{(term} \hlopt{!=} \hlstr{"(Intercept)"}\hlstd{)} \hlopt{%>%}
  \hlkwd{filter}\hlstd{(}\hlkwd{grepl}\hlstd{(}\hlstr{"rooms"}\hlstd{, term))} \hlopt{%>%}
  \hlkwd{select}\hlstd{(model, term, estimate)} \hlopt{%>%}
  \hlkwd{ggplot}\hlstd{(}\hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= estimate,}
             \hlkwc{y} \hlstd{= term,}
             \hlkwc{colour} \hlstd{= model))} \hlopt{+}
  \hlkwd{geom_point}\hlstd{()} \hlopt{+}
  \hlkwd{theme_minimal}\hlstd{()} \hlopt{+}
  \hlkwd{labs}\hlstd{(}\hlkwc{x} \hlstd{=} \hlstr{"Estimate"}\hlstd{,}
       \hlkwc{y} \hlstd{=} \hlstr{"Term"}\hlstd{)} \hlopt{+}
  \hlkwd{guides}\hlstd{(}\hlkwc{colour} \hlstd{=} \hlkwd{guide_legend}\hlstd{(}\hlkwc{title} \hlstd{=} \hlstr{"Model"}\hlstd{))}
\end{alltt}
\end{kframe}\begin{figure}

{\centering \includegraphics[width=1\linewidth]{figure/code-Rnwtidy-coefs-1}

}

\caption[The coefficient estimate for the number of rooms varies according to the imputed dataset]{The coefficient estimate for the number of rooms varies according to the imputed dataset. Complete case dataset produced lower coefficients, compared to imputed datasets.}\label{fig:tidy-coefs}
\end{figure}

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{aug_fit_cc} \hlkwb{<-}  \hlstd{broom}\hlopt{::}\hlkwd{augment}\hlstd{(fit_lm_dat_house_cc)} \hlopt{%>%} \hlkwd{as_tibble}\hlstd{()}
\hlstd{aug_fit_knn} \hlkwb{<-} \hlstd{broom}\hlopt{::}\hlkwd{augment}\hlstd{(fit_lm_dat_house_knn)} \hlopt{%>%} \hlkwd{as_tibble}\hlstd{()}
\hlstd{aug_fit_lm} \hlkwb{<-}  \hlstd{broom}\hlopt{::}\hlkwd{augment}\hlstd{(fit_lm_dat_house_lm)}\hlopt{%>%} \hlkwd{as_tibble}\hlstd{()}

\hlstd{aug_fits} \hlkwb{<-} \hlkwd{bind_rows}\hlstd{(}\hlkwc{cc} \hlstd{=  aug_fit_cc,}
                      \hlkwc{knn} \hlstd{= aug_fit_knn,}
                      \hlkwc{lm} \hlstd{=  aug_fit_lm,}
                      \hlkwc{.id} \hlstd{=} \hlstr{"model"}\hlstd{)}

\hlkwd{ggplot}\hlstd{(aug_fits,} \hlkwd{aes}\hlstd{(}\hlkwc{x} \hlstd{= .fitted,}
                     \hlkwc{y} \hlstd{= .resid))} \hlopt{+}
  \hlkwd{stat_density_2d}\hlstd{(}
    \hlkwc{geom} \hlstd{=} \hlstr{"polygon"}\hlstd{,}
    \hlkwc{contour} \hlstd{=} \hlnum{TRUE}\hlstd{,}
    \hlkwd{aes}\hlstd{(}\hlkwc{fill} \hlstd{=} \hlkwd{after_stat}\hlstd{(level)),}
    \hlkwc{bins} \hlstd{=} \hlnum{7}
  \hlstd{)} \hlopt{+}
  \hlkwd{facet_wrap}\hlstd{(}\hlkwd{vars}\hlstd{(model))} \hlopt{+}
  \hlkwd{scale_fill_viridis_c}\hlstd{()} \hlopt{+}
  \hlkwd{theme_bw}\hlstd{()} \hlopt{+}
  \hlkwd{scale_x_continuous}\hlstd{(}\hlkwc{n.breaks} \hlstd{=} \hlnum{4}\hlstd{)} \hlopt{+}
  \hlkwd{scale_y_continuous}\hlstd{(}\hlkwc{n.breaks} \hlstd{=} \hlnum{4}\hlstd{)} \hlopt{+}
  \hlkwd{theme}\hlstd{(}\hlkwc{legend.position} \hlstd{=} \hlstr{"none"}\hlstd{,}
        \hlkwc{aspect.ratio} \hlstd{=} \hlnum{1}\hlstd{)}
\end{alltt}
\end{kframe}\begin{figure}

{\centering \includegraphics[width=1\linewidth]{figure/code-Rnwpartial-resid-1}

}

\caption[Partial residual plot for each data set, complete cases (cc), and imputed with KNN (knn) or linear model (lm)]{Partial residual plot for each data set, complete cases (cc), and imputed with KNN (knn) or linear model (lm). These are plotted as two dimensional filled density plots with 7 bins estimated. The bins are colored according to concentration of points in that area - brighter colors indicated higher concentration. Compared to complete cases, imputed data has more points clustered closer to 0 for residuals, and and around 6 fitted values.}\label{fig:partial-resid}
\end{figure}

\end{knitrout}

The R session information (including the OS info, R version and all
packages used):

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{sessionInfo}\hlstd{()}
\end{alltt}
\begin{verbatim}
## R version 4.2.1 (2022-06-23)
## Platform: aarch64-apple-darwin20 (64-bit)
## Running under: macOS Monterey 12.3.1
## 
## Matrix products: default
## LAPACK: /Library/Frameworks/R.framework/Versions/4.2-arm64/Resources/lib/libRlapack.dylib
## 
## locale:
## [1] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8
## 
## attached base packages:
## [1] stats     graphics  grDevices utils     datasets  methods   base     
## 
## other attached packages:
##  [1] cowplot_1.1.1     knitr_1.41        janitor_2.1.0     targets_0.14.0   
##  [5] withr_2.5.0       patchwork_1.1.2   visdat_0.5.3      rpart_4.1.19     
##  [9] magick_2.7.3      lubridate_1.9.0   timechange_0.1.1  naniar_0.6.1     
## [13] simputation_0.2.8 scales_1.2.1      png_0.1-7         ggthemes_4.2.4   
## [17] gridExtra_2.3     forcats_0.5.2     stringr_1.4.1     dplyr_1.0.10     
## [21] purrr_0.3.5       readr_2.1.3       tidyr_1.2.1       tibble_3.1.8     
## [25] ggplot2_3.4.0     tidyverse_1.3.2   gert_1.9.1        clipr_0.8.0      
## [29] pak_0.3.1         fs_1.5.2          testthat_3.1.5    devtools_2.4.5   
## [33] usethis_2.1.6    
## 
## loaded via a namespace (and not attached):
##  [1] googledrive_2.0.0   colorspace_2.0-3    ellipsis_0.3.2      snakecase_0.11.0   
##  [5] rstudioapi_0.14     farver_2.1.1        remotes_2.4.2       bit64_4.0.5        
##  [9] fansi_1.0.3         xml2_1.3.3          codetools_0.2-18    cachem_1.0.6       
## [13] pkgload_1.3.2       jsonlite_1.8.3      broom_1.0.1         dbplyr_2.2.1       
## [17] shiny_1.7.3         compiler_4.2.1      httr_1.4.4          backports_1.4.1    
## [21] assertthat_0.2.1    fastmap_1.1.0       gargle_1.2.1        cli_3.4.1          
## [25] later_1.3.0         htmltools_0.5.3     prettyunits_1.1.1   tools_4.2.1        
## [29] igraph_1.3.5        gtable_0.3.1        glue_1.6.2          Rcpp_1.0.9         
## [33] cellranger_1.1.0    vctrs_0.5.1         xfun_0.35           gower_1.0.0        
## [37] ps_1.7.2            brio_1.1.3          rvest_1.0.3         mime_0.12          
## [41] miniUI_0.1.1.1      lifecycle_1.0.3     sys_3.4.1           googlesheets4_1.0.1
## [45] MASS_7.3-58.1       vroom_1.6.0         hms_1.1.2           promises_1.2.0.1   
## [49] credentials_1.3.2   parallel_4.2.1      RColorBrewer_1.1-3  rpart.plot_3.1.1   
## [53] yaml_2.3.6          curl_4.3.3          memoise_2.0.1       UpSetR_1.4.0       
## [57] stringi_1.7.8       highr_0.9           pkgbuild_1.4.0      rlang_1.0.6        
## [61] pkgconfig_2.0.3     evaluate_0.18       labeling_0.4.2      htmlwidgets_1.5.4  
## [65] bit_4.0.5           processx_3.8.0      tidyselect_1.2.0    plyr_1.8.8         
## [69] magrittr_2.0.3      R6_2.5.1            generics_0.1.3      profvis_0.3.7      
## [73] base64url_1.4       DBI_1.1.3           pillar_1.8.1        haven_2.5.1        
## [77] modelr_0.1.10       crayon_1.5.2        utf8_1.2.2          tzdb_0.3.0         
## [81] urlchecker_1.0.1    isoband_0.2.6       grid_4.2.1          readxl_1.4.1       
## [85] data.table_1.14.6   callr_3.7.3         reprex_2.0.2        digest_0.6.30      
## [89] xtable_1.8-4        httpuv_1.6.6        openssl_2.0.4       munsell_0.5.0      
## [93] viridisLite_0.4.1   sessioninfo_1.2.2   askpass_1.1
\end{verbatim}
\begin{alltt}
\hlkwd{Sys.time}\hlstd{()}
\end{alltt}
\begin{verbatim}
## [1] "2022-12-02 12:28:29 AWST"
\end{verbatim}
\end{kframe}
\end{knitrout}

\end{document}

### 2023-02-02 jss-copyedit/2022-01-12-4108/paper.tex
final jss changes

@@
 \includegraphics[width=1\linewidth]{partial-resid-1}
-\caption[Partial residual plot for each data set, complete cases (cc), and
-imputed with KNN (knn) or linear model (lm)]{Partial residual plot for each
-data set, complete cases (cc), and imputed with KNN (knn) or linear model
-(lm).  These are plotted as hex bins, colored according to the number of
-points in a given hexagon.  Brighter colors mean more points.  Compared to
-complete cases, imputed data has more points clustered around
-zero.}\label{fig:partial-resid}
+\caption[Partial residual plot for each data set, complete cases (cc), and imputed with KNN (knn) or linear model (lm).
+These are plotted as two dimensional filled density plots with 7 bins estimated.
+The bins are colored according to concentration of points in that area - brighter colors indicated higher concentration.
+Compared to complete cases, imputed data has more points clustered closer to 0 for residuals, and and around 6 fitted values.}\label{fig:partial-resid}
 \end{figure}

### 2023-02-02 jss-copyedit/2022-01-12-4108/Code/code.R

@@ -728,7 +728,7 @@ tidy_fits %>%
   guides(colour = guide_legend(title = "Model"))
 
 
-## ----partial-resid, fig.cap = "Partial residual plot for each data set, complete cases (cc), and imputed with KNN (knn) or linear model (lm). These are plotted as hex bins, colored according to the number of points in a given hexagon. Brighter colors mean more points. Compared to complete cases, imputed data has more points clustered around zero.", fig.width = 9, fig.height = 3, out.width = "100%"----
+## ----partial-resid, fig.cap = "Partial residual plot for each data set, complete cases (cc), and imputed with KNN (knn) or linear model (lm). These are plotted as two dimensional filled density plots with 7 bins estimated. The bins are colored according to concentration of points in that area - brighter colors indicated higher concentration. Compared to complete cases, imputed data has more points clustered closer to 0 for residuals, and and around 6 fitted values.", fig.width = 9, fig.height = 3, out.width = "100%"----
 aug_fit_cc <-  broom::augment(fit_lm_dat_house_cc) %>% as_tibble()
 aug_fit_knn <- broom::augment(fit_lm_dat_house_knn) %>% as_tibble()
 aug_fit_lm <-  broom::augment(fit_lm_dat_house_lm)%>% as_tibble()
@@ -738,21 +738,19 @@ aug_fits <- bind_rows(cc =  aug_fit_cc,
                       lm =  aug_fit_lm,
                       .id = "model")
 
-gg_hex <- function(data,
-                   subset){
+ggplot(aug_fits, aes(x = .fitted,
+                     y = .resid)) + 
+  stat_density_2d(
+    geom = "polygon",
+    contour = TRUE,
+    aes(fill = after_stat(level)),
+    bins = 7
+  ) +
+  facet_wrap(vars(model)) +
+  scale_fill_viridis_c() + 
+  theme_bw() +
+  scale_x_continuous(n.breaks = 4) +
+  scale_y_continuous(n.breaks = 4) +
+  theme(legend.position = "none",
+        aspect.ratio = 1)
   
-  data_subset <- data %>% 
-    filter(model == subset)
-  
-    ggplot(data_subset, aes(x = .fitted,
-               y = .resid)) + 
-    geom_hex() + 
-    scale_fill_viridis_c() +
-    theme(legend.position = "bottom",
-          aspect.ratio = 1) +
-      labs(tag = subset)
-}
-
[…]
